### 2014-01-03 isit_template.tex
First Commit
This copies over the initial ISIT paper

%% Template Paper ISIT 2013
%%
%% October 2012, Stefan M. Moser
%% based on various earlier templates
%%
%% Please note that your paper must be no more than five pages in
%% the IEEEtran conference style as presented here (including figures,
%% references, etc.!)

\documentclass[conference,a4paper]{IEEEtran}

%% Conference papers do not typically use \thanks and this command
%% is locked out in conference mode. If really needed, such as for
%% the acknowledgment of grants, uncomment the following:
%\IEEEoverridecommandlockouts

\begin{document}

\sloppy

%% Paper Title
%% You can use linebreaks \\ within to get better formatting as
%% desired. 
\title{ISIT 2013 Paper Template\\Please Capitalize Important Words in Title}

%% Author names and affiliations:
%%
%% Avoiding spaces at the end of the author lines is not a problem with
%% conference papers because we don't use \thanks or \IEEEmembership.
%%
%% For several authors with only one affiliation:
%%
% \author{
%   \IEEEauthorblockN{Hui-Ting Chang and Stefan M.~Moser}
%   \IEEEauthorblockA{Department of Electrical and Computer Engineering\\
%     National Chiao Tung University (NCTU)\\
%     Hsinchu, Taiwan\\
%     Email: \{email-of-hui-ting,email-of-stefan\}@ieee.org} 
% }
%%
%% For up to three affiliations:
%%
\author{
  \IEEEauthorblockN{Stefan M.~Moser}
  \IEEEauthorblockA{Dep. of Electrical \& Computer Eng.\\
    National Chiao Tung University\\
    Hsinchu, Taiwan\\
    Email: email-of-stefan@ieee.org} 
  \and
  \IEEEauthorblockN{Erdal Ar\i{}kan}
  \IEEEauthorblockA{Dep. of Electrical \& Electronics Eng.\\
    Bilkent University\\
    Ankara, Turkey\\
    Email: email\_erdal@university.edu}
  \and
  \IEEEauthorblockN{Elza Erkip}
  \IEEEauthorblockA{ECE Department\\ 
    Polytechnic Institute of New York University\\
    New York, USA\\
    Email: elza\_s-email-address@ieee.org}
}
%%
%% For over three affiliations, or if they all won't fit within the width
%% of the page, use this alternative format:
%%
% \author{
%   \IEEEauthorblockN{
%     Michael Shell\IEEEauthorrefmark{1},
%     Homer Simpson\IEEEauthorrefmark{2},
%     James Kirk\IEEEauthorrefmark{3}, 
%     Montgomery Scott\IEEEauthorrefmark{3} and
%     Eldon Tyrell\IEEEauthorrefmark{4}}
%   \IEEEauthorblockA{
%     \IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%     Georgia Institute of Technology, Atlanta, Georgia 30332--0250\\ 
%     Email: see http://www.michaelshell.org/contact.html}
%   \IEEEauthorblockA{
%     \IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%     Email: homer@thesimpsons.com}
%   \IEEEauthorblockA{
%     \IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%     Telephone: (800) 555--1212, Fax: (888) 555--1212}
%   \IEEEauthorblockA{
%     \IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}
% }

%% Use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

%% To balance the two columns, you should reduce the text-height of
%% the last page using the following command:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\addtolength{\textheight}{-9.35cm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% with an appropriate value. This command must be place on the second
%% last page, i.e., for a one-page abstract here, for a two-page
%% abstract right after the \maketitle command.

%% Create the title:
\maketitle

%% Abstract: 
%% For the final version of the accepted paper, please make sure you
%% remove the comment "THIS PAPER IS ELIGIBLE FOR THE STUDENT PAPER
%% AWARD."
%%
\begin{abstract}
  This paper provides the instructions for the preparation of paper
  submissions and of final manuscripts for the Proceedings of ISIT
  2013.
\end{abstract}

\section{Introduction}

The 2013 IEEE International Symposium on Information Theory will be
held at ICEC Convention Center near Taksim Square at the heart of
Istanbul, Turkey, from Sunday, July 7 to Friday, July 12, 2013.

\section{Submission}

Paper submission is handled online using the EDAS system:
\begin{center}
  https://edas.info/newPaper.php?c=13285
\end{center}
The papers are restricted in length to \textbf{five pages} in the
IEEEtran-conference style as presented here (including figures,
references, etc.).

Each paper must be classified as ``Eligible for student paper award''
or ``Not eligible for student paper award''. Paper that are selected
to be eligible for the student paper award should also contain
\emph{``THIS PAPER IS ELIGIBLE FOR THE STUDENT PAPER AWARD''} as a
first line in the abstract of the submission. Note that this reference
must be removed again in the final manuscript!

The deadline for the submission is \textbf{January 27, 2013}.

\section{Proceedings}

Accepted papers will be published in full (up to five pages in
length). A hard-copy book of abstracts will also be distributed at the
Symposium to serve as a guide to the sessions.

The deadline for the submission of the final camera-ready paper is
\textbf{May 15, 2013}.  Accepted papers not submitted by that date
will not appear in the ISIT Proceedings and will not be included in
the technical program of the ISIT.

\section{Preparation of the Paper}

Only electronic submissions in form of a PDF file will be accepted.
The paper should be formatted in A4-format with sufficient margins at
the top and the bottom.  We highly encourage the authors to prepare
their papers with \LaTeX{} using the \LaTeX{} style file
\verb+IEEEtran.cls+, which can be easily obtained from
\begin{center}
  \small
  http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
\end{center}
and using the \LaTeX{} template file \verb+isit_template.tex+ that
produced this document and that may be downloaded from the ISIT 2013
web site
\begin{center}
  http://www.isit2013.org/
\end{center}
Authors using other means to prepare their manuscripts should attempt
to duplicate the style of this example as closely as possible.

The style of references (e.g., \cite{shannon1948}), equations,
figures, tables, etc., should be the same as for the \emph{IEEE
  Transactions on Information Theory}. Example codes for figures and
tables can be found in the template file (they are commented out).

%% An example of a floating figure using the graphicx package.
%% Note that \label must occur AFTER (or within) \caption.
%% For figures, \caption should occur after the \includegraphics.
%%
% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.3\textwidth]{myfigure}
%   % where an .eps filename suffix will be assumed under latex,
%   % and a .pdf suffix will be assumed for pdflatex
%   \caption{Simulation results.}
%   \label{fig:sim}
% \end{figure}

%% An example of a double column floating figure using two subfigures.
%% (The subfigure.sty package must be loaded for this to work.)  The
%% subfigure \label commands are set within each subfigure command,
%% the \label for the overall fgure must come after \caption.  
%% \hfil must be used as a separator to get equal spacing
%%
% \begin{figure*}[htbp]
%   \centerline{\subfigure[Case I]{\includegraphics[width=2.5in]{subfigcase1}
%       % where an .eps filename suffix will be assumed under latex,
%       % and a .pdf suffix will be assumed for pdflatex
%       \label{fig:first_case}}
%     \hfil
%     \subfigure[Case II]{\includegraphics[width=2.5in]{subfigcase2}
%       % where an .eps filename suffix will be assumed under latex,
%       % and a .pdf suffix will be assumed for pdflatex
%       \label{fig:second_case}}}
%   \caption{Simulation results.}
%   \label{fig:sim}
% \end{figure*}

%% An example of a floating table. 
%% Note that, for IEEE style tables, the \caption command should come
%% BEFORE the table. Table text will default to \footnotesize as IEEE
%% normally uses this smaller font for tables.  The \label must come
%% after \caption as always.
%%
% \begin{table}[htbp]
%   % increase table row spacing, adjust to taste
%   \renewcommand{\arraystretch}{1.3}
%   \caption{An Example of a Table}
%   \label{tab:table_example}
%   \centering
%   % Some packages, such as MDW tools, offer better commands for making tables
%   % than the plain LaTeX2e tabular which is used here.
%   \begin{tabular}{|c||c|}
%     \hline
%     One & Two\\
%     \hline
%     Three & Four\\
%     \hline
%   \end{tabular}
% \end{table}

For proper typesetting of equations, the environment
\verb+\begin{IEEEeqnarray}+ is highly recommended. A short
introduction on how to use this tool can be found online at
\begin{center}
  http://moser.cm.nctu.edu.tw/manuals.html\#eqlatex
\end{center}

The affiliation shown for authors should constitute a sufficient
mailing address for persons who wish to write for more details about
the paper.

\section{Conclusion}

The conclusion goes here.

For an appendix see comments in the template file.

%% Appendix:
%% If needed a single appendix is created by
%\appendix
%% If several appendices are needed, then the command
%\appendices
%% in combination with further \section-commands can be used.

%% Use \section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank various sponsors for supporting 
their research.

%% References:
%% We recommend the usage of BibTeX:
%%
%\bibliographystyle{IEEEtran}
%\bibliography{definitions,bibliofile}
%%
%% where we here have assume the existence of the files
%% definitions.bib and bibliofile.bib.
%% BibTeX documentation can be obtained at:
%% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
%%
%%
%%
%% Or manual references (pay attention to consistency!):
\begin{thebibliography}{1}
\bibitem{shannon1948}
  C.~E. Shannon, ``A mathematical theory of communication,''
  \emph{Bell System Techn. J.}, vol.~27, pp. 379--423 and 623--656,
  Jul. and Oct. 1948. 
\end{thebibliography}

\end{document}

### 2014-01-03 peISIT2013.tex
First Commit
This copies over the initial ISIT paper

\documentclass[conference,a4paper]{IEEEtran}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{subfigure}
\usepackage{amsthm}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newtheorem*{mydef}{Definition}
\title{Determinism, Complexity, and Predictability in Computer Performance}

%Version 1
%\author{
%  \IEEEauthorblockN{Joshua Garland}
%  \IEEEauthorblockA{Dept. of Computer Science\\
%    University of Colorado at Boulder\\
%    Colorado, USA\\
%    Email: joshua.garland@colorado.edu}
%  \and
%  \IEEEauthorblockN{Ryan G.~James}
%  \IEEEauthorblockA{Complexity Sciences Center \& Dept. of Physics\\
%    University of California at Davis\\
%    California, USA\\
%    Email: rgjames@ucdavis.edu}
%    \and
%      \IEEEauthorblockN{Elizabeth Bradley}
%         \IEEEauthorblockA{Santa Fe Institute \\
%    New Mexico, USA\\
%    }
%  \IEEEauthorblockA{Dept. of Computer Science\\
%    University of Colorado at Boulder\\
%    Colorado, USA\\
%    Email: lizb@colorado.edu}
%  }

 \author{
   \IEEEauthorblockN{
     Joshua Garland\IEEEauthorrefmark{1},
     Ryan G.~James\IEEEauthorrefmark{1} and
     Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
   \IEEEauthorblockA{
     \IEEEauthorrefmark{1}Dept. of Computer Science
     University of Colorado, Boulder, Colorado 80309-0430 USA\\
     Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
   \IEEEauthorblockA{
     \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
     Email: lizb@colorado.edu}
   % \IEEEauthorblockA{
   %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, California 95616 USA\\
   %   Email: rgjames@ucdavis.edu}
 }

\begin{document}

\maketitle

\begin{abstract}
  Computers are deterministic dynamical systems \cite{mytkowicz09}.
  Among other things, that implies that one should be able to use
  deterministic forecast rules to predict aspects of their behavior.
  That statement is sometimes---but not always---true. The memory and
  processor loads of some simple programs are easy to predict, for
  example, but those of more-complex programs like {\tt gcc} are not.
%%%%%%%%%%%%%%%%%%%%%%
%% I had to change all the \verb|blah| entries because they caused
%% latex to barf if they were in figure captions.  Odd bug.
%%%%%%%%%%%%%%%%%%%%%%
  The goal of this paper is to determine why that is the case. We
  conjecture that, in practice, complexity can effectively overwhelm
  the predictive power of deterministic forecast models. To explore
  that, we build models of a number of performance traces from
  different programs running on different Intel-based computers. We
  then calculate the \emph{permutation entropy}---a temporal entropy
  metric that uses ordinal analysis---of those traces and correlate
  those values against the prediction success.
\end{abstract}

\section{Introduction}
%\begin{it}
%Paragraph on computer performance, including citations to Todd paper
%and summary of the results that indicate that they're deterministic
%nonlinear dynamical systems.  Given that, we should be able to
%predict.  What benefits would accrue if we could do so: power mgmt,
%end world hunger [[this is my primary goal everyday :)]], etc.
%\end{it}

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:cache}, and chaos places fundamental limits on
predictability.
%
\begin{figure}[htbp]
   \centering
   \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
   % where an .eps filename suffix will be assumed under latex,
   % and a .pdf suffix will be assumed for pdflatex
   \caption{A small snippet of the L2 cache miss rate of {\tt
       col\_major}, a three-line C program that repeatedly initializes
     a matrix in column-major order, running on an Intel Core
     Duo\textsuperscript{\textregistered}-based machine.  Even this
     simple program exhibits chaotic performance dynamics.}
   \label{fig:cache}
 \end{figure}

The computer systems community has applied a variety of prediction
strategies to traces like this, most of which employ regression.  An
appealing alternative builds on the recently established fact that
computers can be effectively modeled as deterministic nonlinear
dynamical systems \cite{mytkowicz09}.  This result implies the
existence of a deterministic forecast rule for those dynamics.  In
particular, one can use \emph{delay-coordinate embedding} to
reconstruct the underlying dynamics of computer performance, then use
the resulting model to forecast the future values of computer
performance metrics such as memory or processor loads
\cite{josh-ida2011}.  In the case of simple microkernels like the one
that produced the trace in Figure~\ref{fig:cache}, this deterministic
modeling and forecast strategy works very well.  In more-complicated
programs, however, such as speech recognition software or compilers,
this forecast strategy---as well as the traditional methods---break
down quickly.

This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
nonlinearity, and nonstationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  We study four
examples---two simple microkernels and two complex programs from the
SPEC benchmark suite---running on different Intel-based machines.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle) and memory-use efficiency (cache-miss
rates), then compare that to the prediction accuracy attainable for
that trace using a simple deterministic model.

% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU: instructions executed per second, cache misses per 100,000
instructions, and similar statistics. Prediction of these quantities
at some finite time in the future, even with perfect accuracy, does
not violate the Rice-Shapiro theorem.

% The rest of the paper is organized as follows.
% Section~\ref{sec:compModel} describes the experimental setup, as well
% as the nonlinear modeling and forecast strategies.  In
% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
% its value for a number of different computer performance traces, and
% compare the results to the prediction accuracy.  In
% Section~\ref{sec:conc}, we discuss these results and their
% implications in regard to our conjecture, and consider future areas of
% research.

\section{Modeling Computer Performance}\label{sec:compModel}

% took out for space
% \subsection{Reconstructing hidden dynamics}

Delay-coordinate embedding allows one to reconstruct a system's full
state-space dynamics from a \emph{single} scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
scalar value $x$ that one is measuring---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
\]
from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
where $t$ is time, is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of the modeling approach that is used in this paper.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
plot of the data from Figure~\ref{fig:cache}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.
\begin{figure}
  \centering
    \includegraphics[width=0.5\textwidth]{figs/3dcolcache}
    \caption{A 3D projection of a delay-coordinate embedding of the trace
from Figure~\ref{fig:cache} with a delay ($\tau$) of 100,000 instructions.
}
\label{fig:embedding}
\end{figure}
%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \verb|col_major| L2 cache miss rate time series
$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
$y(t+2\tau)$ on the third, and so on.
%% ...down to here.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism\footnote{A deeper analysis of
  Figure~\ref{fig:embedding}---as alluded to on the previous
  page---supports that diagnosis, confirming the presence of a chaotic
  attractor in these cache-miss dynamics, with largest Lyapunov
  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
structure can also be used to build a forecast model.

% took out for space
% \subsection{LMA: Using dynamics in forecasting}

Given a nonlinear model of a deterministic dynamical system in the
form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
one can build deterministic forecast algorithms by capturing and
exploiting the geometry of the embedding.  Many techniques have been
developed by the dynamical systems community for this purpose
(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
is the ``Lorenz method of analogues'' (LMA), which is essentially
nearest-neighbor prediction in the embedded state
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
forecast---works quite well on the trace in Figure~\ref{fig:cache}, as
shown in Figure~\ref{fig:cachePredTS}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
    \caption{A forecast of the last 4,000 points of the signal in
      Figure~\ref{fig:cache} using an LMA-based strategy on the
      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
      $\times$s are the true and predicted values, respectively;
      vertical bars show where these values differ. }
\label{fig:cachePredTS}
\end{figure}
%
On the other hand, if we use the same approach to forecast the
processor load\footnote{Instructions per cycle, or IPC} of the {\tt
  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
on an Intel i7\textsuperscript{\textregistered}-based machine, the
prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
     \caption{An LMA-based forecast of the last 4,000 points of a
       processor-load performance trace from the {\tt 482.sphinx3}
       benchmark.  Red circles and blue $\times$s are the true and
       predicted values, respectively; vertical bars show where these
       values differ.}
\label{fig:predsphinx}
\end{figure}

Table~\ref{tab:PredError} presents detailed results about the
prediction accuracy of this algorithm on four different examples: the
{\tt col\_major} and {\tt 482.sphinx3} programs in
Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as
another simple microkernel that initializes the same matrix as {\tt
  col\_major}, but in row-major order, and another complex program
({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.  Both
microkernels were run on the Intel Core
Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks
were run on the Intel i7\textsuperscript{\textregistered} machine.  We
calculated a figure of merit for each prediction as follows.  We held
back the last $k$ elements\footnote{Several different prediction
  horizons were analyzed in our experiment; the results reported in
  this paper are for $k$=4000} of the $N$ points in each measured time
series, built the forecast model by embedding the first $N-k$ points,
used that embedding and the LMA method to predict the next $k$ points,
then computed the Root Mean Squared Error (RMSE) between the true and
predicted signals:
$$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$
%
To compare the success of predictions across signals with different
units, we normalized RMSE as follows:
$$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
%
% The smaller the nRMSE, obviously, the more accurate the prediction.

  \begin{table}[htbp]
   % increase table row spacing, adjust to taste
   \renewcommand{\arraystretch}{1.3}
   \caption{Normalized root mean squared error (nRMSE) of 4000-point
     predictions of memory \& processor performance from different
     programs.}
   \label{tab:PredError}
   \centering
   % Some packages, such as MDW tools, offer better commands for making tables
   % than the plain LaTeX2e tabular which is used here.
%   \begin{tabular}{|c|c|c|c|c|c|}
%     \hline
%      & cache misses NRMSE & IPC NRMSE \\
%     \hline
%     \tt{row\_major}  & 0.0102  & 0.0095 \\
%          \hline
%     \tt{col\_major}  &0.0055& 0.0071 \\
%     \hline
%     \tt{403.gcc}  & 0.0865& 0.1805 \\
%          \hline
%     \tt{482.sphinx3} & 0.1142& 0.2946 \\
%      \hline
%   \end{tabular}
% \end{table}
   \begin{tabular}{|c|c|c|c|c|c|}
     \hline
      & cache miss rate & instrs per cycle \\
     \hline
     \tt{row\_major}  & 0.0324  & 0.0778 \\
          \hline
     \tt{col\_major}  &0.0080& 0.0161 \\
     \hline
     \tt{403.gcc}  & 0.1416& 0.2033 \\
          \hline
     \tt{482.sphinx3} & 0.2032& 0.3670 \\
      \hline
   \end{tabular}
 \end{table}

The results in Table~\ref{tab:PredError} show a clear distinction
between the two microkernels, whose future behavior can be
predicted effectively using this simple deterministic modeling
strategy, and the more-complex SPEC benchmarks, for which this
prediction strategy does not work nearly as well.
% Removed for space
% For both processor load (IPC) and memory usage (cache-miss rate),
% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
% for \verb|col_major| or \verb|row_major|.
%
This begs the question: If these traces all come from deterministic
systems---computers---then why are they not equally predictable?  Our
conjecture is that the sheer complexity of the dynamics of the SPEC
benchmarks running on the Intel i7\textsuperscript{\textregistered}
machine make them effectively impossible to predict.

%% [[If space: Add a segue sentence: next section uses permutation
%% entropy to explore that conjecture.]]

\section{Measuring Complexity}\label{sec:meaComplex}

For the purposes of this paper, one can view entropy as a measure of
complexity and predictability in a time series.  A high-entropy time
series is almost completely unpredictable---and conversely.  This can
be made more rigorous: Pesin's relation \cite{pesin77} states that in
chaotic dynamical systems, the Shannon entropy rate is equal to the
sum of the positive Lyapunov exponents, $\lambda_i$. The Lyapunov
exponents directly quantify the rate at which nearby states of the
system will diverge with time: $\left| \Delta x(t) \right| \approx
e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence,
the more difficult prediction becomes.

Utilizing entropy as a measure of temporal complexity is by no means a new idea
\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage
requires categorical data: $x_t \in \mathcal{S}$ for some finite or
countably infinite \emph{alphabet} $\mathcal{S}$, and data taken from
real-world systems is effectively real-valued.  To get around this,
one must discretize the data---typically by binning.  Unfortunately,
this is rarely a good solution to the problem, as the binning of the
values introduces an additional dynamic on top of the intrinsic
dynamics whose entropy is desired.  The field of symbolic dynamics
studies how to discretize a time series in such a way that the
intrinsic behavior is not perverted, but these methods are fragile in
the face of noise and require further understanding of the underlying
system, which defeats the purpose of measuring the entropy in the
first place.

Bandt and Pompe introduced \emph{permutation entropy} (PE)
% also called permutation complexity
as a ``natural complexity measure for time series"
\cite{bandt2002per}.  Permutation entropy employs a method of
discretizing real-valued time series that follows the intrinsic
behavior of the system under examination.  Rather than looking at the
statistics of sequences of values, as is done when computing the
Shannon entropy, permutation entropy looks at the statistics of the
\emph{orderings} of sequences of values using ordinal analysis.
Ordinal analysis of a time series is the process of mapping successive
time-ordered elements of a time series to their value-ordered
permutation of the same size.  By way of example, if $(x_1, x_2, x_3)
= (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many
features; among other things, it is robust to noise and requires no
knowledge of the underlying mechanisms.

% It has been shown [amigo?] that the
% normalized permutation entropy \ref{eq:permEntropy} is equal to the
% Kolmogorov-Sinai entropy and therefore is a proxy for the rate at
% which a system becomes unpredictable.

% This spectrum is outlined very nicely in \cite{amigo2012permutation}:
% ``Periodic or quasiperiodic sequences have vanishing or negligible
% complexity. At the opposite end, independent and identically
% distributed random sequences (white noise) have asymptotically
% divergent permutation entropies, owing to the fact that the number of
% allowed (or ``admissible'') ordinal patterns grows superexponentially
% with length. Between both ends lie the kind of sequences we are
% interested in.''

\begin{mydef}[Permutation Entropy]
  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define
  $\mathcal{S}_n$ as all $n!$ permutations $\pi$ of order $n$. For
  each $\pi \in \mathcal{S}_n$ we determine the relative frequency of
  that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$: $$ p(\pi) =
  \frac{\left|\{t|t\le T-n,\phi(x_{t+1},\dots,x_{t+n}) =
      \pi\}\right|}{T-n+1}$$ Where $|\cdot|$ is set cardinality. The
  \emph{permutation entropy} of order $n\ge 2$ is defined
  as$$H(n) = - \sum_{\pi \in
      \mathcal{S}_n} p(\pi)\log_2p(\pi)$$
\end{mydef}
Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this
in mind, it is common in the literature to normalize permutation
entropy as follows: $\frac{H(n)}{\log_2(n!)}$.  With this convention,
``low'' entropy is close to 0 and ``high'' entropy is close to 1.
Finally, it should be noted that the permutation entropy has been
shown to be identical to the Shannon entropy for many large classes of
systems \cite{amigo2012permutation}.

In practice, calculating permutation entropy involves choosing a good
value for the wordlength $n$. The key consideration
here is that the value be large enough that forbidden ordinals are
discovered, yet small enough that reasonable statistics over the
ordinals are gathered: e.g., $\displaystyle n =
\operatornamewithlimits{argmax}_\ell\{T > 100 \ell!\}$, assuming an
average of 100 counts per ordinal. In the literature, $3\le n \le 6$
is a standard choice---generally without any formal justification. In
theory, the permutation entropy should reach an asymptote with
increasing $n$, but that requires an arbitrarily long time series. In
practice, what one should do is calculate the \emph{persistent}
permutation entropy by increasing $n$ until the result converges, but
data length issues can intrude before that convergence is reached.

% Tables~\ref{tab:TCMpe} \& \ref{tab:IPCpe}
Table~\ref{tab:pe} shows the permutation entropy results for the
examples considered in this paper, with the nRMSPE prediction
accuracies from the previous section included alongside for easy
comparison.
%
% \begin{table}[htbp]
%   % increase table row spacing, adjust to taste
%   \renewcommand{\arraystretch}{1.3}
%   \caption{TCM NRMSE and PE NORMS}
%   \label{tab:table_example}
%   \centering
   % Some packages, such as MDW tools, offer better commands for making tables
   % than the plain LaTeX2e tabular which is used here.
%   \begin{tabular}{|c|c|c|c|c|c|}
%     \hline
%     TCM & NRMSE & $m=3$  &$m=4$&$m=5$&$m=6$\\
%     \hline
%     \tt{row\_major} & 0.0102 & 0.7900  &0.6751&0.5458&0.4491\\
%     \hline
%     \tt{col\_major} & 0.0055 & 0.6411  &0.5029&0.4515&0.3955 \\
%     \hline
%     \tt{403.gcc} & 0.0865 & 0.9954  &0.9916&0.9880&0.9835\\
%     \hline
%     \tt{482.sphinx3} & 0.1142 & 0.9958  &0.9913&0.9866&0.9802 \\
%      \hline
%   \end{tabular}
% \end{table}
%
%%  \begin{table}[htbp]
%%   % increase table row spacing, adjust to taste
%%   %\renewcommand{\arraystretch}{1.3}
%%   \caption{Cache-miss rate performance traces: prediction error
%%     (nRMSPE) and permutation entropy}
%%   \label{tab:TCMpe}
%%   \centering
%%   % Some packages, such as MDW tools, offer better commands for making tables
%%   % than the plain LaTeX2e tabular which is used here.
%%   \begin{tabular}{|c|c|c|c|c|c|}
%%     \hline
%%      & nRMSE   &$n=4$&$n=5$&$n=6$\\
%%     \hline
%%     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
%%     \hline
%%     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
%%     \hline
%%     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
%%     \hline
%%     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
%%      \hline
%%   \end{tabular}
%% \end{table}
%
% \begin{table}[htbp]
%   % increase table row spacing, adjust to taste
%   \renewcommand{\arraystretch}{1.3}
%   \caption{IPC NRMSE and PE NORMS}
%   \label{tab:table_example}
%   \centering
%   % Some packages, such as MDW tools, offer better commands for making tables
%   % than the plain LaTeX2e tabular which is used here.
%   \begin{tabular}{|c|c|c|c|c|c|}
%     \hline
%     ipc & NRMSE & $n=3$  &$n=4$&$n=5$&$n=6$\\
%     \hline
%      \tt{row\_major} & 0.0095 & 0.9912  &0.9723&0.9354&0.8876\\
%     \hline
%      \tt{col\_major} & 0.0071 & 0.9428  &0.8356&0.7601&0.6880\\
%     \hline
%     \tt{403.gcc} & 0.1805 & 0.9918  &0.9862&0.9814&0.9764\\
%     \hline
%     \tt{482.sphinx3} & 0.2946 & 0.9978  &0.9951&0.9914&0.9849 \\
%      \hline
%   \end{tabular}
% \end{table}
%
%%  \begin{table}[htbp]
%%   % increase table row spacing, adjust to taste
%%   %\renewcommand{\arraystretch}{1.1}
%%   \caption{Processor load performance traces: prediction error
%%     (nRMSPE) and permutation entropy}
%%   \label{tab:IPCpe}
%%   \centering
%%   % Some packages, such as MDW tools, offer better commands for making tables
%%   % than the plain LaTeX2e tabular which is used here.
%%   \begin{tabular}{|c|c|c|c|c|}
%%     \hline
%%      & nRMSPE   &$n=4$&$n=5$&$n=6$\\
%%     \hline
%%      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
%%     \hline
%%      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
%%     \hline
%%     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
%%     \hline
%%     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
%%      \hline
%%   \end{tabular}
%% \end{table}
%
  \begin{table}[htbp]
   \caption{Prediction error (in nRMSPE) and permutation entropy (for
     different wordlengths $n$}
   \label{tab:pe}
   \centering
   \begin{tabular}{|c|c|c|c|c|c|}
     \hline
      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
     \hline
     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
     \hline
     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
     \hline
     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
     \hline
     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
      \hline
      \hline
      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
     \hline
      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
     \hline
      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
     \hline
     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
     \hline
     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
      \hline
   \end{tabular}
 \end{table}
The relationship between prediction accuracy and the permutation
entropy (PE) is as we conjectured: performance traces with high
PE---those whose temporal complexity is high, in the sense that little
information is being propagated forward in time---are indeed harder to
predict using the simple deterministic forecast model described in the
previous section.  The effects of changing $n$ are also interesting:
using a longer wordlength generally lowers the PE---a natural
consequence of finite-length data---but the falloff is less rapid in
some traces than in others, suggesting that those values are closer to
the theoretical asymptote that exists for perfect data.  The
persistent PE values of 0.5--0.6 for the {\tt row\_major} and {\tt
  col\_major} cache-miss traces are consistent with dynamical chaos,
further corroborating the results of~\cite{mytkowicz09}.  (PE values
above 0.97 are consistent with white noise.)  Interestingly, the
processor-load traces for these two microkernels exhibit more temporal
complexity than the cache-miss traces.  This may be a consequence of
the lower baseline value of this time series.

% In Table~\ref{tab:TCMpe} \& \ref{tab:IPCpe} the behavior of the PEs as
% they limit to their asymptotic values is observed. A feature of note
% is that some of the signals drop off and some do not.  For example at
% $m=4$ {\tt col\_major} is at 0.8356, a fairly high entropy, however as
% $m$ is increased to 6 the PE plummets to 0.6880: a value consistent
% with a chaotic time series. Notice this also corresponds to a very low
% nRMSE. In contrast, consider the permutation entropy of the processor
% load of {\tt 482.sphinx3}: with $m=4$ it is 0.9951 and only drops to
% 0.9849 at $m=6$. This is consistent with the signal also having the
% highest nRMSE.

\section{ Conclusions \& Future Work}\label{sec:conc}

The results presented here suggest that permutation entropy---a
ordinal calculation of forward information transfer in a time
series---is an effective metric for predictability of computer
performance traces. Experimentally, traces with a persistent PE
$\gtrapprox 0.97$ have a natural level of complexity that may
overshadow the inherent determinism in the system dynamics, whereas
traces with PE $\lessapprox 0.7$ seem to be highly predictable (viz.,
at least an order of magnitude improvement in nRMSPE).

If information is the limit, then gathering and using more information
is an obvious next step.  There is an equally obvious tension here
between data length and prediction speed: a forecast that requires
half a second to compute is not useful for the purposes of real-time
control of a computer system with a MHz clock rate.  Another
alternative is to sample several system variables simultaneously and
build multivariate delay-coordinate embeddings.  Existing approaches
to that are computationally prohibitive
\cite{cao-multivariate-embedding}.  We are working on alternative
methods that sidestep that complexity.

%%This use of permutaiotn entropy is a useful application of information
%%theory to computer performance modeling because it gives you a simple
%%and fast way to decide whether building a model and trying to forecast
%%the future is worthwhile or if guessing the mean is just as effective.
%%End with a sentence tying back to power management and world peace.

%\begin{it}
%Summarize the results.

%Paragraphs on the issues that come up, including one about the "amount
%of info" one: if one could sample more variables, for instance, one
%might be able to do a better job of predicting more-complex traces.
%Segue to some handwaving about multivariable LMA models; tie this back
%to the "computers are NLD systems" stuff in the intro.  This is a real
%challenge; current approaches to this modelling problem have the major
%issue of taking way too long to build.  And that's a big issue if
%you're trying not just to classify, but to predict.  In a system that
%runs at MHz speeds, a prediction that takes milliseconds to compute is
%not useful.
%\end{it}

%[[Joshua: It may be interesting to look at the PE as a function of
%prediction horizon. It is known to me that certain start and end
%locations in the prediction strategy result in different prediction
%accuracy. Is it the case that when the algorithm tanks it is also the
%case that PE of the **TEST**(the 10\% we cut off )is high? This may
%be a great way to illustrate the concept if it works... OR maybe do
%complexity of forward path in embedding space?]]  some notes about
%this looking at row Cache 1000 very good RMSPE 21.4167 m = 3 4 5 6
%1.4192 2.1286 2.5942 2.9230 row Cache 2000 very good RMSPE 19.0990 m
%= 3 4 5 6 1.4145 2.1305 2.5825 2.9095 so even better row Cache 2010
%RMSPE 19.9716 m = 3 4 5 6 1.4149 2.1311 2.5830 2.9096 row Cache 1990
%RMSPE 19.1465 So prediction Horizons ranked in order of best RMSPE
%RMSPE L 3 4 5 6 1st 19.0990 2000 1.4145 2.1305 2.5825 2.9095 2nd
%19.1465 1990 1.4151 2.1293 2.5803 2.9067 3rd 19.9716 2010 1.4149
%2.1311 2.5830 2.9096 3rd 21.4167 1000 1.4192 2.1286 2.5942 2.9230
%ipcgcc m = 3 pe = 1.7770 penorm = 0.9918 m= 4 pe = 3.1341 penorm =
%0.9862 m = 5 pe = 4.6985 penorm = 0.9814 m=6 pe = 6.4242 penorm =
%0.9764 m=7 pe = 8.2450 penorm = 0.9671 >> [pe penorm] =
%permCalc(gcc100TCM,5,1) pe = 4.7299 penorm = 0.9880 >> [pe penorm] =
%permCalc(gcc100TCM,6,1) pe = 6.4707 penorm = 0.9835

%****BEYOND THIS POINT IS A DIFFERNT PAPER*** JOSHUA WILL CLEAN THIS OUT, JUST HERE FOR REFERNCE****
%\section{Introduction: What is Permutation Entropy}

%%\begin{it}
%%Here I want to talk about the definition and algorithm for computing it. Maybe briefly talk about the asymptotic %properties. (i.e., same as metric topological in limit. Converges to dominate leap if not normalized etc. )
%%\end{it}
%
%When analyzing a time series there are several standard characterizations of complexity, including (but not limited to) entropy, Lyapunov exponents and fractal dimensions. Rigorous definitions of all these quantities exist if the observation is a \emph{noise-free} trajectory of a \emph{single} dynamical system. To an experimental-time-series analyst these are luxuries that are rarely (if ever) realized. Many times experimental observation have noise involved such as sensor error, or very commonly in the computer world finite-precision error.
%
%Unfortunately the standard algorithms for calculating many of these complexity measures are highly sensitive to noise and give spurious results in the presence of noisy observations. Even worse, if a time-series observation measures multiple dynamical systems or dynamical systems that undergo bifurcations\footnote{A drastic change in the behavior of the dynamical system, generally associated with parameter drift.} then all of the standard algorithms for estimating complexity, (e.g., correlation dimension) go out the window. Bandt and Pompe first introduced \emph{permutation entropy} as a ``natural complexity measure for time series" in 2002 \cite{bandt2002per}.
%
%
%As explained in \cite{bandt2002per}, permutation entropy is a simple complexity measure that is easily calculated for any time series, be it stochastic, periodic, quasi-periodic, chaotic, noisy or any combination of these. Most importantly for analysis of computer performance, permutation entropy as defined in \cite{bandt2002per} yields meaningful results even if observational or dynamical noise is present \cite{bandt2002per}.
%
%
%
%
%Analyzing entropies as  a complexity measure is not a new concept, however the way that the entropy is estimated in \cite{bandt2002per} is novel and particularly useful when little is known about the underlying process.   As discussed in \cite{bandt2002per}, the standard entropy one considers is Shannon entropy.
%
%
%As a quick review of Shannon entropy, we follow the treatment in \cite{amigo2012permutation}. Given a finite alphabet $\mathcal{S} = \{s_1,....,s_{|\mathcal{S}|}\}$, define an information source as a discrete-time, stationary stochastic process $X = \{X_n\}_{n \in \mathbb{N}_0}$ where $X_n$ are random variables on a common probability space taking on values from $\mathcal{S}$. A realization of $X$ is a one-sided sequence $x_0^{\infty} = (x_n)_{n \in \mathbb{N}_0}$ called a \emph{message}. The elements $x_n$ of $\mathcal{S}$ are the \emph{symbols}(a.k.a letters) of the language. A finite segment of the message, say $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$ is called a \emph{word} (of length m). We then define $p(x_0^{m-1})$ to be the probability of the word $x_0^{m-1}$ to appear some where in the message. The shannon entropy of the data source $X$ is defined as: \begin{equation}\label{eq:shannonentropy}
%h(X) = - \lim_{m\to\infty}\frac{1}{m}\sum p(x_0^{m-1})\log p(x_0^{m-1})
%\end{equation}
%
%For the purposes of this paper once can view entropy as a measure of complexity and predictability in a time series. A high entropy almost completely unpredictable and a small entropy is completely predictable. This spectrum is outlined very nicely in \cite{amigo2012permutation}: ``Periodic or quasiperiodic sequences have vanishing or negligible complexity. At the opposite end, independent and identically distributed random sequences (white noise) have asymptotically divergent permutation entropies, owing to the fact that the number of allowed (or ``admissibleÓ) ordinal patterns grows superexponentially with length. Between both ends lie the kind of sequences we are interested in."
%
%This illustrates why utilizing entropy as a measure for complexity is, in theory, a great candidate. There are several fundamental drawbacks with this approach when analyzing real-world time series. The two most fundamental problems are as follows:
%
%\begin{enumerate}
%\item Elements of an experimental time series $x_t \in \mathbb{R}$ which implies $x_t \notin \mathcal{S}$, for a finite alphabet $\mathcal{S}$.
%\item Time series are always finite in length so the limit in \ref{eq:shannonentropy} never exists in practice.
%\end{enumerate}
%
%
%The common practice to address item 1 is to perform \emph{symbolization} on the time series you want to analyze. Symbolization is the process of defining a dynamic preserving map $\phi: \mathbb{R}^m \to \mathcal{S}$, i.e., for each $x_t$ (or block $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$) we must assign a symbol from $\mathcal{S}$ in such a way that the time series has the same fundamental properties.  There are several approaches to symbolization, the standard approach is partitioning. Assume, $x_t$ has the capability of taking on infinitely many values, defining the space $X$ to be the space where these $x_t$ are drawn. We then define a partition on this space $X = P_1 \cup...\cup P_m$, and define $\mathcal{S} = \{1,\dots,m\}$. Then $\phi(x_t) = i$ if $x_t \in P_i$. We can then calculate the Shannon entropy of the transformed time series, which in the limit of $m$ converges to the Kolmogorov-Sinai entropy \cite{bandt2002per}. This approach to symbolization is computationally infeasable as we need an infinite number of partitions of an infinite time series. For \emph{extremely} rare cases, there exists a \emph{generating} partition where this limit is not necessary. However, finding such a partition even for a simple two-dimensional map like Henon is non-trivial \cite{bandt2002per}. Finding a generating partition for an arbitrary time series is for all intensive purposes impossible.
%
%[[[Maybe also talk about Mischaikow's approach to this with isolating neighborhoods and homology...but might be off topic. ]]]
%
%
%.....
%
%
%
%Instead of this approach to symbolization, Bandt et al. take the approach that ``the symbol sequence must come naturally for the time series, without any further model assumptions" \cite{bandt2002per}. To do this they perform symbolization through ordinal analysis. Ordinal analysis of a time series is the process of mapping the ordered successive elements of a time series to a permutation of the same size based on their original time-based ordering.[[Need a better way of wording this]]. This is also referred to in the literature as the permutation complexity \cite{amigo2012permutation}  For example, if $(x_1,x_2,x_3) = (9,1,7)$ then $\phi$ maps $(x_1,x_2,x_3)$ to the permutation $231$ since $x_2 < x_3<x_1$. A great explanation of why this is an interesting symbolization is found in \cite{amigo2012permutation} :
%
%\begin{quote}The study of permutation complexity, which we call ordinal analysis, can be envisioned as a new kind of symbolic dynamics whose basic blocks are ordinal patterns. Interesting enough, it turns out that under some mild mathematical assumptions, not all ordinal patterns can be materialized by the orbits of a given one- or multi-dimensional deterministic dynamics, not even if this dynamic is chaoticÑ contrarily to what happens with the symbol patterns. As a result, the existence of ÒforbiddenÓ (i.e., not occurring) ordinal patterns is always a persistent dynamical feature, in opposition to properties such as proximity and correlation which die out with time in a chaotic dynamic. Moreover, if an ordinal pattern is forbidden, its absence pervades all longer patterns in form of more missing ordinal patterns, called outgrowth forbidden patterns. Admissible ordinal patterns grow exponentially with length, while forbidden patterns do superexponentially. Since random (unconstrained) dynamics has no forbidden patterns with probability 1, their existence can be used as a fingerprint of deterministic orbit generation.
%\end{quote} Essentially by analyzing the permutation entropy of a time series we are gaining information as to the complexity of the system without any other information available, whether the time series is generated by a random process or the trajectory of a deterministic dynamical system. Now that we have justified why the study of permutation entropy is interesting let us define it. For this we follow \cite{bandt2002per}.
%
%
%\begin{mydef}[Permutation Entropy]
%Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$: \begin{equation}\label{eq:permprob}
%p(\pi) = \frac{|\{t|t\le T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}|}{T-n+1}\end{equation}
%Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n\ge 2$ is defined as $$H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi)\log_2p(\pi)$$ Notice that $0\le H(n) \le \log_2(n!)$,\cite{bandt2002per}. With this in mind, it is common in the literature to present permutation entropy normalized, i.e., $\frac{H(n)}{\log_2(n!)}$. With this convention, \emph{low} entropy is close to 0 and \emph{high} entropy is entropy close to 1.
%\end{mydef}
%
%Notes: According to \cite{bandt2002per} equation (\ref{eq:permprob}) estimates the frequency of $\pi$ as good as possible for a finite series of values. To determine $p(\pi)$ exactly, we have to assume an infinite time series and take the limit as $T\to \infty$ in (\ref{eq:permprob}). Bandt et al., also point out that this limit exists under a very weak stationarity condition: for $k\le n$, the probability for $x_t<x_{t+k}$ should not depend on $t$.
%
%It should also be noted that this is discrete-metric (or shannon) permutation entropy, to be distinguished from topological permutation entropy (a.k.a. permutation capacity). Topological permutation entropy is defined in \cite{amigo2012permutation} as: $$h(\{x_t\}_{t = 1,\dots,\infty}) =  \lim_{L\to\infty}\frac{1}{L}logN(L)$$ where $N(L)$ is the number of distinct ordinal patterns that occur of order $L$. This is a measure of how many distinct permutations of length $L$ are realized as opposed to counting the frequency at which each occurs as is the case in metric permutation entropy \cite{amigo2012permutation}. It can be shown that metric and topological entropy both converge to the true value with probability 1 in the limit.  (citation for this). When we say permutation entropy in this paper we are referring to that defined in \ref{eq:permprob}.
%
%\section{Detecting changes in a time series}
%
%Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}. To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}. Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block. (A window shift of 1.)} (or nonoverlapping) blocks. Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.  According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
%
%In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy. This is identical to the lag used in delay coordinate embedding and often the same language is used. Cao et al. describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space. When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$. So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$. Cao et al. argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1. Cao et al. then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter. The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
%
%\subsection{Example: Concatenated Time Series}\label{sec:concatenate}
%As an example of this detection, consider Figure \ref{fig:ts-banded} (a). This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise. Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points. The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population. Each Logistic map trajectory begins at $x=0.6$. The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
%
%The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$). Visually the bands of noise and the chaotic trajectory are indistinguishable. If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise. If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
%
%In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves). The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively). Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}. On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands. This is in fact the case when comparing to the way the time series was constructed.
%
%The main advantage we can see to the use of lags other than 1 is for analysis of periodicity. This is also the major advantage listed in \cite{cao2004det}. Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero. This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}. If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
%
%%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
%
%\begin{figure}
%  \centering
%  \subfigure{\includegraphics[width=\columnwidth]{figs/ts-banded-alone}}
%    {(a)}
%
%  \subfigure{\includegraphics[width=\columnwidth]{figs/ts-banded-pe}}
%  {(b)}
%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}. (b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves). Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
%\end{figure}
%
%
%
%
%
%
%and what is it good for with forward pointer to the other sections
%
%
%
%\subsection{Example: The transient Logistic map}
%In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed. As we showed, permutation was a great proxy to detect this change. Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time. To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}. Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$. The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map. The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
%
% As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well. An interesting feature of block calculations are the spikes that occur at bifurcation points. As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes. This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy. The dips you see in the second half correspond to bands of periodicity which occur between the chaos. It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range. This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
%
% This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
%
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\columnwidth]{figs/tstranslog-perm}
%  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
%  \label{fig:translogistic}
%\end{figure}
%%\begin{enumerate}
%%\item do with henon bifurcation. what happens near the broken chaotic window
%
%%\end{enumerate}
%
%
%\section{Possible Applications to Computer Systems Performance}
%
%
%\begin{enumerate}
%\item In this section I want to discuss the symbolization of performance traces and why this is advantageous over generating partitions and similar approaches.
%\item No model is assumed, some traces do appear to be stochastic some appear to be deterministic this provides ``proof" thereof. albeit not mathematical proof but CS proof for sure.
%\item We have essentially infinite data. So limits like $m!$ where $m$ is word length isn't really an issue for us.
%\end{enumerate}
%
%
%\subsection{Predictability and model change detection}
%
%Talk about how some explanatory models may be spinning their wheels because no transferable information. That is if you are introducing pure entropy at every time step really no use in using past values and statistically the mean is the most accurate.
%
%\begin{enumerate}
%\item Comp Performance goes under constant bifurcations. This allows autonomous detection of when a model is no longer valid. (need to be careful here with use of valid as to not be confused with model validation.
%\item Maybe can tell us if we should use a linear-stochastic, nonlinear-deterministic model or just guess the mean
%\end{enumerate}
%
%
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\columnwidth]{figs/ts-rowcol-pe}
%  \caption{A time series of cache misses during the execution of the row-col microkernel. The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=5,6$ (respectively). }\label{fig:rowcol}
%\end{figure}
%
%
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\columnwidth]{figs/ts-gcc-pe}
%  \caption{A time series of instructions per cycle for the gcc spec benchmark. The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=3,4$ (respectively). The lack of structure in such a signal illustrates that the use of deterministic models is most likely not the most advantageous.}\label{fig:gcc}
%\end{figure}
%
%
%
%\section{Pro and Cons}
%\subsection{Advantages of Permutation Entropy}
%Permutation entropy allows for symbolization of a time series without any further assumptions being made about the underlying system. In fact, virtually nothing needs to be known before applying this techniques. Other methods of symbolization require expert knowledge of the underlying system, such as generating partitions. These techniques have been applied to small data sets, this is truly an advantage especially when dealing with a lot of biological systems whose large datasets can be in the tens of thousands. The algorithm for computing permutation entropy is very straight forward and easy to explain. The actual code takes longer than I would like to run, but this seems to be an issue with my implementation and not the algorithm, as others state they have exceptionally fast algorithms. Although none of these algorithms are ever provided in the literature. The major advantages of permutation entropy is that it can quickly give us a rough estimate of where drastic changes in the time series which may have implications in model selection techniques. Permutation entropy also can give you a quick idea about if a system has deterministic structure. This has direct implications on what prediction strategies to implement.
%%\begin{enumerate}
%%\item Allows for symbolization without a model. Really emphasize we can obtain a symbolization without generating partitions which is the traditional approach here.
%%\item Small amounts of data not really a problem.
%%\item Maybe talk about how much data is needed per word length for results that mean anything.
%%\item Fairly easy to calculate, although need a more efficient method.
%%\item Gives a rough estimation of where drastic changes have occurred in a TS
%%\item Allows you to see if there is actually something predictable in the signal.
%%\end{enumerate}
%
%\subsection{Disadvantages of PE}
%There are three disadvantages with permutation entropy we can immediately see that need further thought to be overcome. The first is that it is sometimes hard to tell when a change in the dynamical system is not drastic enough to have a huge change on the entropy. For example, consider a dynamical system switching from a chaotic to a (different) chaotic regime. The permutation entropy may be very similar while the model is slightly different. Permutation entropy seems to be ill-suited for fine grain detection of this nature. Our current implementation is too slow for on-the-fly analysis. We do not believe this is an issue with permutation entropy itself but with our implementation. The literature all claims that fast algorithms exist.
%
%The biggest disadvantage for us in using permutation entropy is selection of parameters. To calculate permutation entropy, we have to choose an embedding dimension, a lag, and a block size. In all the literature we have found thus far, parameter selection is made based on persistence or matching the permutation entropy to known values. The problem comes when analyzing real time series these parameters seem to have a real impact on the results. Finding (or creating) a rigorous method for parameter selection is paramount in the use of this method for the analysis of real world time series.
%%\begin{enumerate}
%%\item Paramaters!
%%\item Currently numerically expensive although the literature disagrees. Think this might just be my implementation. Need to read the numerics section of the PE book.
%%\item Sometimes hard to tell a regime shift has occurred if the transition is chaotic to chaotic. or low period to low period.
%%\end{enumerate}

\section*{Acknowledgment}
This work was partially supported by NSF grant \#CMMI-1245947 and ARO
grant \#W911NF-12-1-0288.

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,bibliofile}

\end{document}

### 2014-01-04 pecomplexity.tex
Removing ISIT Specific Formating

--- 2014-01-03 peISIT2013.tex
@@
-\documentclass[conference,a4paper]{IEEEtran}
+\documentclass{article}
 \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
@@
 
- \author{
-   \IEEEauthorblockN{
-     Joshua Garland\IEEEauthorrefmark{1},
-     Ryan G.~James\IEEEauthorrefmark{1} and
-     Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
-   \IEEEauthorblockA{
-     \IEEEauthorrefmark{1}Dept. of Computer Science
-     University of Colorado, Boulder, Colorado 80309-0430 USA\\
-     Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
-   \IEEEauthorblockA{
-     \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
-     Email: lizb@colorado.edu}
+ %\author{
+ %  \IEEEauthorblockN{
+ %    Joshua Garland\IEEEauthorrefmark{1},
+ %    Ryan G.~James\IEEEauthorrefmark{1} and
+ %    Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
  %  \IEEEauthorblockA{
-   %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, California 95616 USA\\
-   %   Email: rgjames@ucdavis.edu}
- }
+ %    \IEEEauthorrefmark{1}Dept. of Computer Science
+ %    University of Colorado, Boulder, Colorado 80309-0430 USA\\
+ %    Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
+ %  \IEEEauthorblockA{
+ %    \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
+ %    Email: lizb@colorado.edu}
+ %  % \IEEEauthorblockA{
+ %  %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, %California 95616 USA\\
+%   %   Email: rgjames@ucdavis.edu}
+% }
 
@@
 
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,bibliofile}
+%\bibliographystyle{}
+\bibliography{bibliofile}
 

### 2014-01-04 pecomplexity.tex
Added Figures and Errors

@@
 
+\section{New Figures and Tables}
+
+\begin{table}[htdp]
+\caption{ Embedding Parameters for reference if needed. }
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+&$\tau$ & $m$ \\
+\hline
+gcc & 10& 13\\
+col\_major  &2&12\\
+SVD\_Full &  10  &  12 \\
+SVD\_IPC\_Regime1 & 5 &14  \\
+SVD\_IPC\_Regime2 & 10   &12  \\
+SVD\_IPC\_Regime3 & 2   &9  \\
+SVD\_IPC\_Regime4 &  3  & 11 \\
+SVD\_IPC\_Regime5 &  23  & 10 \\
+SVD\_IPC\_Regime6 &  30  & 12  \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
+
+\begin{table}[htdp]
+\caption{Average nRMSE over 15 runs for each signal and average (?) pe [[Ryan:is this average over all 15 and is this weighted]] at word length 5 and 6 for each signal. }
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+& Horizon&nRMSE LMA & nRMSE mean & $l=5$& $l=6$\\
+\hline
+gcc &4542&  $0.1407\pm0.0063 $   & $0.1487\pm0.0066$ & - & - \\
+col\_major &14,709&  $0.0252\pm0.0061$   & $0.1975\pm0.0436$ & - & -\\
+
+SVD\_Full &22,072&  $0.0323\pm0.0019$  & $ 0.1784\pm0.0040$ & - & -  \\
+SVD\_IPC\_Regime1 &2,158&  $0.1680 \pm 0.0317$  & $0.3517\pm0.5223$ &0.9786 &0.9675  \\
+SVD\_IPC\_Regime2 &6,902& $0.1716 \pm0.0043 $  &$0.1762 \pm0.0012$ &0.8800&0.8493 \\
+SVD\_IPC\_Regime3 &947& $0.0507 \pm 0.0011$   &$0.5413\pm0.0005$  &0.7646&0.7029\\
+SVD\_IPC\_Regime4 &2,929&  $0.1288\pm0.0471$  & $0.2308\pm0.0867$ &0.8735&0.7770\\
+SVD\_IPC\_Regime5 &3,255&  $0.0235\pm0.0022$ & $0.1306\pm0.0003$&0.7281&0.6724 \\
+SVD\_IPC\_Regime6 &5,804&  $0.0196\pm0.0022$  & $0.0508\pm0.0003$ &0.8316&0.7615 \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/SVD1RegimesColored}\caption{SVD IPC Time Series with Regimes Colored}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/svd_mutuals}\caption{SVD mutuals for full and regimes}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/full}\caption{SVD IPC Full Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime1}\caption{SVD IPC Regime 1 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime2}\caption{SVD IPC Regime 2 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{figs/regime3}\caption{SVD IPC Regime 3 Time Series}
+%\label{default}
+%\end{center}
+%\end{figure}
+
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{figs/regime4}\caption{SVD IPC Regime 4 Time Series}
+%\label{default}
+%\end{center}
+%\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime5}\caption{SVD IPC Regime 5 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime6}\caption{SVD IPC Regime 6 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
 \section*{Acknowledgment}

### 2014-01-04 pecomplexity.tex
finish off the table of results

@@
 \begin{table}[htdp]
-\caption{Average nRMSE over 15 runs for each signal and average (?) pe [[Ryan:is this average over all 15 and is this weighted]] at word length 5 and 6 for each signal. }
+\caption{Average nRMSE over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
 \begin{center}
@@
 \hline
-& Horizon&nRMSE LMA & nRMSE mean & $l=5$& $l=6$\\
-\hline
-gcc &4542&  $0.1407\pm0.0063 $   & $0.1487\pm0.0066$ & - & - \\
-col\_major &14,709&  $0.0252\pm0.0061$   & $0.1975\pm0.0436$ & - & -\\
-
-SVD\_Full &22,072&  $0.0323\pm0.0019$  & $ 0.1784\pm0.0040$ & - & -  \\
-SVD\_IPC\_Regime1 &2,158&  $0.1680 \pm 0.0317$  & $0.3517\pm0.5223$ &0.9786 &0.9675  \\
-SVD\_IPC\_Regime2 &6,902& $0.1716 \pm0.0043 $  &$0.1762 \pm0.0012$ &0.8800&0.8493 \\
-SVD\_IPC\_Regime3 &947& $0.0507 \pm 0.0011$   &$0.5413\pm0.0005$  &0.7646&0.7029\\
-SVD\_IPC\_Regime4 &2,929&  $0.1288\pm0.0471$  & $0.2308\pm0.0867$ &0.8735&0.7770\\
-SVD\_IPC\_Regime5 &3,255&  $0.0235\pm0.0022$ & $0.1306\pm0.0003$&0.7281&0.6724 \\
-SVD\_IPC\_Regime6 &5,804&  $0.0196\pm0.0022$  & $0.0508\pm0.0003$ &0.8316&0.7615 \\
+                  & Horizon & nRMSE LMA           & nRMSE mean          & $l=5$                   & $l=6$                   \\
+\hline
+gcc               & 4,542   & $0.1407 \pm 0.0063$ & $0.1487 \pm 0.0066$ & $0.950991 \pm 0.001073$ & $0.942964 \pm 0.001256$ \\
+
+col\_major        & 14,709  & $0.0252 \pm 0.0061$ & $0.1975 \pm 0.0436$ & $0.563640 \pm 0.003110$ & $0.513101 \pm 0.003365$ \\
+
+SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.876517 \pm 0.004354$ & $0.850805 \pm 0.003454$ \\
+SVD\_IPC\_Regime1 & 2,158   & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ & $0.976109 \pm 0.008359$ & $0.957204 \pm 0.015623$ \\
+SVD\_IPC\_Regime2 & 6,902   & $0.1716 \pm 0.0043$ & $0.1762 \pm 0.0012$ & $0.875984 \pm 0.005161$ & $0.846360 \pm 0.004420$ \\
+SVD\_IPC\_Regime3 & 947     & $0.0507 \pm 0.0011$ & $0.5413 \pm 0.0005$ & $0.776841 \pm 0.007290$ & $0.715747 \pm 0.005605$ \\
+SVD\_IPC\_Regime4 & 2,929   & $0.1288 \pm 0.0471$ & $0.2308 \pm 0.0867$ & $0.907348 \pm 0.008018$ & $0.824626 \pm 0.007707$ \\
+SVD\_IPC\_Regime5 & 3,255   & $0.0235 \pm 0.0022$ & $0.1306 \pm 0.0003$ & $0.733272 \pm 0.007562$ & $0.677575 \pm 0.006792$ \\
+SVD\_IPC\_Regime6 & 5,804   & $0.0196 \pm 0.0022$ & $0.0508 \pm 0.0003$ & $0.810087 \pm 0.013468$ & $0.747545 \pm 0.010607$ \\
 \hline

### 2014-01-05 pecomplexity.tex
add stuff about weighted permutation entropy

@@
 
-For the purposes of this paper, one can view entropy as a measure of
-complexity and predictability in a time series.  A high-entropy time
-series is almost completely unpredictable---and conversely.  This can
-be made more rigorous: Pesin's relation \cite{pesin77} states that in
-chaotic dynamical systems, the Shannon entropy rate is equal to the
-sum of the positive Lyapunov exponents, $\lambda_i$. The Lyapunov
-exponents directly quantify the rate at which nearby states of the
-system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence,
-the more difficult prediction becomes.
+For the purposes of this paper, one can view entropy as a measure of complexity
+and predictability in a time series.  A high-entropy time series is almost
+completely unpredictable---and conversely.  This can be made more rigorous:
+Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
+Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
+$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
+states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
+e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
+difficult prediction becomes.
 
 Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage
-requires categorical data: $x_t \in \mathcal{S}$ for some finite or
-countably infinite \emph{alphabet} $\mathcal{S}$, and data taken from
-real-world systems is effectively real-valued.  To get around this,
-one must discretize the data---typically by binning.  Unfortunately,
-this is rarely a good solution to the problem, as the binning of the
-values introduces an additional dynamic on top of the intrinsic
-dynamics whose entropy is desired.  The field of symbolic dynamics
-studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in
-the face of noise and require further understanding of the underlying
-system, which defeats the purpose of measuring the entropy in the
-first place.
-
-Bandt and Pompe introduced \emph{permutation entropy} (PE)
-% also called permutation complexity
-as a ``natural complexity measure for time series"
-\cite{bandt2002per}.  Permutation entropy employs a method of
-discretizing real-valued time series that follows the intrinsic
-behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the
-Shannon entropy, permutation entropy looks at the statistics of the
-\emph{orderings} of sequences of values using ordinal analysis.
-Ordinal analysis of a time series is the process of mapping successive
-time-ordered elements of a time series to their value-ordered
-permutation of the same size.  By way of example, if $(x_1, x_2, x_3)
-= (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many
-features; among other things, it is robust to noise and requires no
+\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
+categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
+\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
+effectively real-valued.  To get around this, one must discretize the data---
+typically achieved by binning.  Unfortunately, this is rarely a good solution to
+the problem, as the binning of the values introduces an additional dynamic on
+top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
+dynamics studies how to discretize a time series in such a way that the
+intrinsic behavior is not perverted, but these methods are fragile in the face
+of noise and require further understanding of the underlying system, which
+defeats the purpose of measuring the entropy in the first place.
+
+Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
+complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
+employs a method of discretizing real-valued time series that follows the
+intrinsic behavior of the system under examination.  Rather than looking at the
+statistics of sequences of values, as is done when computing the Shannon
+entropy, permutation entropy looks at the statistics of the \emph{orderings} of
+sequences of values using ordinal analysis. Ordinal analysis of a time series is
+the process of mapping successive time-ordered elements of a time series to
+their value-ordered permutation of the same size.  By way of example, if $(x_1,
+x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
+is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
+other things, it is generally robust to observational noise and requires no
 knowledge of the underlying mechanisms.
 
-% It has been shown [amigo?] that the
-% normalized permutation entropy \ref{eq:permEntropy} is equal to the
-% Kolmogorov-Sinai entropy and therefore is a proxy for the rate at
-% which a system becomes unpredictable.
-
-% This spectrum is outlined very nicely in \cite{amigo2012permutation}:
-% ``Periodic or quasiperiodic sequences have vanishing or negligible
-% complexity. At the opposite end, independent and identically
-% distributed random sequences (white noise) have asymptotically
-% divergent permutation entropies, owing to the fact that the number of
-% allowed (or ``admissible'') ordinal patterns grows superexponentially
-% with length. Between both ends lie the kind of sequences we are
-% interested in.''
-
 \begin{mydef}[Permutation Entropy]
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define
-  $\mathcal{S}_n$ as all $n!$ permutations $\pi$ of order $n$. For
-  each $\pi \in \mathcal{S}_n$ we determine the relative frequency of
-  that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$: $$ p(\pi) =
-  \frac{\left|\{t|t\le T-n,\phi(x_{t+1},\dots,x_{t+n}) =
-      \pi\}\right|}{T-n+1}$$ Where $|\cdot|$ is set cardinality. The
-  \emph{permutation entropy} of order $n\ge 2$ is defined
-  as$$H(n) = - \sum_{\pi \in
-      \mathcal{S}_n} p(\pi)\log_2p(\pi)$$
+
+  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
+  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
+  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
+  = 1,\dots,T}$:
+  \begin{align*}
+    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
+  \end{align*}
+  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
+  \ge 2$ is defined as
+  \begin{align*}
+  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
+  \end{align*}
+
 \end{mydef}
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this
-in mind, it is common in the literature to normalize permutation
-entropy as follows: $\frac{H(n)}{\log_2(n!)}$.  With this convention,
-``low'' entropy is close to 0 and ``high'' entropy is close to 1.
-Finally, it should be noted that the permutation entropy has been
-shown to be identical to the Shannon entropy for many large classes of
-systems \cite{amigo2012permutation}.
-
-In practice, calculating permutation entropy involves choosing a good
-value for the wordlength $n$. The key consideration
-here is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the
-ordinals are gathered: e.g., $\displaystyle n =
-\operatornamewithlimits{argmax}_\ell\{T > 100 \ell!\}$, assuming an
-average of 100 counts per ordinal. In the literature, $3\le n \le 6$
-is a standard choice---generally without any formal justification. In
-theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In
-practice, what one should do is calculate the \emph{persistent}
-permutation entropy by increasing $n$ until the result converges, but
-data length issues can intrude before that convergence is reached.
-
-% Tables~\ref{tab:TCMpe} \& \ref{tab:IPCpe}
+
+Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
+it is common in the literature to normalize permutation entropy as follows:
+$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
+and ``high'' entropy is close to 1. Finally, it should be noted that the
+permutation entropy has been shown to be identical to the Shannon entropy for
+many large classes of systems \cite{amigo2012permutation}.
+
+Here we will be utilizing a variation of the permutation entropy, the
+\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
+permutation entropy attempts to correct for observational noise which is larger
+than some trends in the data, but smaller than the larger scale features --- for
+example, a signal that switches between two fixed points with noise about those
+fixed points. The weighted permutation entropy would be dominated by the
+switching rather than by the stochastic fluctuation. To accomplish this, the
+\emph{weight} of a permutation is taken into account:
+\begin{align*}
+  w(x_{t+1:t+n}) = \frac{1}{n}
+                 \sum_{x_i \in \{x_{t+1:t+n}\}}
+                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
+\end{align*}
+where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
+$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
+
+The weighted probability of a permutation is then:
+\begin{align*}
+  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
+\end{align*}
+where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
+probability enhances permutations involved in ``large'' features and demotes
+permutations which are small in amplitude relative to the features of the time
+series. The weighted permutation entropy is then:
+\begin{align*}
+  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi)
+\end{align*}
+
+In practice, calculating permutation entropy involves choosing a good value for
+the wordlength $n$. The key consideration here is that the value be large enough
+that forbidden ordinals are discovered, yet small enough that reasonable
+statistics over the ordinals are gathered: e.g., $\displaystyle n =
+\operatornamewithlimits{argmax}_\ell\{T > 100 \ell!\}$, assuming an average of
+100 counts per ordinal. In the literature, $3\le n \le 6$ is a standard choice
+---generally without any formal justification. In theory, the permutation
+entropy should reach an asymptote with increasing $n$, but that requires an
+arbitrarily long time series. In practice, what one should do is calculate the
+\emph{persistent} permutation entropy by increasing $n$ until the result
+converges, but data length issues can intrude before that convergence is
+reached.
+
 Table~\ref{tab:pe} shows the permutation entropy results for the

### 2014-01-06 pecomplexity.tex
Adding cleaning up Figures

@@
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{amssymb}
@@
 \usepackage{epstopdf}
-\usepackage{subfigure}
 \usepackage{amsthm}
@@
 
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdipcfull}
+                \caption{SVD IPC without coloring}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdipcregimescolored}
+                \caption{SVD IPC with coloring}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{SVD Full Time Series }\label{fig:svdFull}
+\end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colfullts}
+                \caption{col IPC  Full Time Series}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colshortts}
+                \caption{col IPC  Short Time Series for structure illustration}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{col  Time Series }\label{fig:svdFull}
+\end{figure}
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics[width=\textwidth]{figs/SVD1RegimesColored}\caption{SVD IPC Time Series with Regimes Colored}
+\includegraphics[width=\textwidth]{figs/gccfullts}\caption{403.gcc full time series}
 \label{default}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figs/full}\caption{SVD IPC Full Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
 \includegraphics[width=\textwidth]{figs/regime1}\caption{SVD IPC Regime 1 Time Series}

### 2014-01-06 pecomplexity.tex
Updating some Prediction Figures
add author info

@@
 \usepackage{amsthm}
+\usepackage{authblk}
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
@@
 \title{Determinism, Complexity, and Predictability in Computer Performance}
+
+\author[1]{Joshua Garland \thanks{joshua.garland@colorado.edu}}
+\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
+\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
+\affil[1]{Department of Computer Science \\
+  University of Colorado at Boulder \\
+  Colorado, USA
+}
+\affil[2]{Santa Fe Institute \\
+  New Mexico, USA
+}
 
@@
         \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdPrediction}
+                \caption{SVD LMA Prediction}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdPredictionFull}
+                \caption{SVD LMA Prediction with Full Time Series}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{SVD LMA Prediction Figures }\label{fig:svdFull}
+\end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
                 \includegraphics[width=\textwidth]{figs/colfullts}
@@
         \end{subfigure}
+
         ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
@@
          \caption{col  Time Series }\label{fig:svdFull}
+\end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colPredShortTS}
+                \caption{col IPC  Full Time Series}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colPredfullTS}
+                \caption{col IPC  Short Time Series for structure illustration}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{col\_major Predictions }\label{fig:svdFull}
 \end{figure}
@@
 
-%\bibliographystyle{}
+\bibliographystyle{unsrt}
 \bibliography{bibliofile}

### 2014-01-06 pecomplexity.tex
fix some spacing
updating table a bit

@@
 \author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-\affil[1]{Department of Computer Science \\
-  University of Colorado at Boulder \\
+\affil[1]{Department of Computer Science\\
+  University of Colorado at Boulder\\
   Colorado, USA
 }
-\affil[2]{Santa Fe Institute \\
+\affil[2]{Santa Fe Institute\\
   New Mexico, USA
@@
 \hline
-                  & Horizon & nRMSE LMA           & nRMSE mean          & $l=5$                   & $l=6$                   \\
+                  & Horizon & nRMSE LMA           & nRMSE na\"{i}ve          & $l=5$                   & $l=6$                   \\
 \hline
@@
 
-SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.876517 \pm 0.004354$ & $0.850805 \pm 0.003454$ \\
+%SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.876517 \pm 0.004354$ & $0.850805 \pm 0.003454$ \\
 SVD\_IPC\_Regime1 & 2,158   & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ & $0.976109 \pm 0.008359$ & $0.957204 \pm 0.015623$ \\

### 2014-01-10 pecomplexity.tex
add two new images

@@
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.95\textwidth]{figs/SVD_wwpe}
+  \caption{The weighted permutation entropy of one run of SVD. The gray bands
+    are regions where the window overlaps regimes. The window size used is
+    $5,000 \times 100,000$ instructions and the word length is $4$.}
+  \label{fig:wwpe}
+\end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.95\textwidth]{figs/prediction_vs_entropy}
+  \caption{The improvement of LMA over na\"{i}ve vs weighted permutation entropy.
+  For each of these, the word length used is $5$.}
+  \label{fig:pred_vs_ent}
+\end{figure}
+
 \section*{Acknowledgment}

### 2014-01-10 pecomplexity.tex
update some about word length

@@
 
-In practice, calculating permutation entropy involves choosing a good value for
-the wordlength $n$. The key consideration here is that the value be large enough
-that forbidden ordinals are discovered, yet small enough that reasonable
-statistics over the ordinals are gathered: e.g., $\displaystyle n =
-\operatornamewithlimits{argmax}_\ell\{T > 100 \ell!\}$, assuming an average of
-100 counts per ordinal. In the literature, $3\le n \le 6$ is a standard choice
----generally without any formal justification. In theory, the permutation
-entropy should reach an asymptote with increasing $n$, but that requires an
-arbitrarily long time series. In practice, what one should do is calculate the
-\emph{persistent} permutation entropy by increasing $n$ until the result
-converges, but data length issues can intrude before that convergence is
-reached.
-
-Table~\ref{tab:pe} shows the permutation entropy results for the
-examples considered in this paper, with the nRMSPE prediction
-accuracies from the previous section included alongside for easy
-comparison.
+In practice, calculating permutation entropy and weighted permutation entropy
+involves choosing a good value for the word length $n$. The primary
+consideration is that the value be large enough that forbidden ordinals are
+discovered, yet small enough that reasonable statistics over the ordinals are
+gathered: e.g.:
+\begin{align*}
+n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
+\end{align*}
+assuming an average of 100 counts per ordinal is sufficient. In the literature,
+$3 \le n \le 6$ is a standard choice --- generally without any formal
+justification. In theory, the permutation entropy should reach an asymptote with
+increasing $n$, but that requires an arbitrarily long time series. In practice,
+what one should do is calculate the \emph{persistent} permutation entropy by
+increasing $n$ until the result converges, but data length issues can intrude
+before that convergence is reached.
+
+% Table~\ref{tab:pe} shows the permutation entropy results for the
+% examples considered in this paper, with the nRMSPE prediction
+% accuracies from the previous section included alongside for easy
+% comparison.
 %
@@
 %
-  \begin{table}[htbp]
-   \caption{Prediction error (in nRMSPE) and permutation entropy (for
-     different wordlengths $n$}
-   \label{tab:pe}
-   \centering
-   \begin{tabular}{|c|c|c|c|c|c|}
-     \hline
-      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
-     \hline
-     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
-     \hline
-     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
-     \hline
-     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
-     \hline
-     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
-      \hline
-      \hline
-      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
-     \hline
-      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
-     \hline
-      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
-     \hline
-     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
-     \hline
-     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
-      \hline
-   \end{tabular}
- \end{table}
+ %  \begin{table}[htbp]
+ %   \caption{Prediction error (in nRMSPE) and permutation entropy (for
+ %     different wordlengths $n$}
+ %   \label{tab:pe}
+ %   \centering
+ %   \begin{tabular}{|c|c|c|c|c|c|}
+ %     \hline
+ %      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
+ %     \hline
+ %     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
+ %     \hline
+ %     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
+ %     \hline
+ %     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
+ %     \hline
+ %     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
+ %      \hline
+ %      \hline
+ %      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
+ %     \hline
+ %      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
+ %     \hline
+ %      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
+ %     \hline
+ %     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
+ %     \hline
+ %     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
+ %      \hline
+ %   \end{tabular}
+ % \end{table}
+
 The relationship between prediction accuracy and the permutation

### 2014-01-11 pecomplexity.tex
clean up text some

@@
 \begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi)
+  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
 \end{align*}
+which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
+results of this paper.
 
@@
 
-The relationship between prediction accuracy and the permutation
-entropy (PE) is as we conjectured: performance traces with high
-PE---those whose temporal complexity is high, in the sense that little
-information is being propagated forward in time---are indeed harder to
-predict using the simple deterministic forecast model described in the
-previous section.  The effects of changing $n$ are also interesting:
-using a longer wordlength generally lowers the PE---a natural
-consequence of finite-length data---but the falloff is less rapid in
-some traces than in others, suggesting that those values are closer to
-the theoretical asymptote that exists for perfect data.  The
-persistent PE values of 0.5--0.6 for the {\tt row\_major} and {\tt
-  col\_major} cache-miss traces are consistent with dynamical chaos,
-further corroborating the results of~\cite{mytkowicz09}.  (PE values
-above 0.97 are consistent with white noise.)  Interestingly, the
-processor-load traces for these two microkernels exhibit more temporal
-complexity than the cache-miss traces.  This may be a consequence of
-the lower baseline value of this time series.
+The relationship between prediction accuracy and the weighted permutation
+entropy (WPE) is much as we conjectured: performance traces with high WPE ---
+those whose temporal complexity is high, in the sense that the entropy
+generation per value is approximately equal to the total entropy per value thus
+little information is being propagated forward in time --- are indeed harder to
+predict using the simple deterministic forecast model described in the previous
+section.  The effects of changing the word length $n$ are also interesting:
+using a longer word length generally lowers the WPE --- a natural consequence of
+finite-length data --- but the falloff is less rapid in some traces than in
+others, suggesting that those values are closer to the theoretical asymptote
+that exists for perfect data. The persistent WPE values of 0.5--0.6 for the {\tt
+col\_major} trace are consistent with dynamical chaos, further corroborating the
+results of~\cite{mytkowicz09}.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.95\textwidth]{figs/prediction_vs_entropy}
+  \caption{The improvement of LMA over na\"{i}ve vs weighted permutation entropy.
+  For each of these, the word length used is $5$.}
+  \label{fig:pred_vs_ent}
+\end{figure}
 
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.95\textwidth]{figs/prediction_vs_entropy}
-  \caption{The improvement of LMA over na\"{i}ve vs weighted permutation entropy.
-  For each of these, the word length used is $5$.}
-  \label{fig:pred_vs_ent}
-\end{figure}
-
 \section*{Acknowledgment}

### 2014-01-11 pecomplexity.tex
use fewer sigfigs in the table

@@
 \hline
-gcc               & 4,542   & $0.1407 \pm 0.0063$ & $0.1487 \pm 0.0066$ & $0.950991 \pm 0.001073$ & $0.942964 \pm 0.001256$ \\
-
-col\_major        & 14,709  & $0.0252 \pm 0.0061$ & $0.1975 \pm 0.0436$ & $0.563640 \pm 0.003110$ & $0.513101 \pm 0.003365$ \\
-
-%SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.876517 \pm 0.004354$ & $0.850805 \pm 0.003454$ \\
-SVD\_IPC\_Regime1 & 2,158   & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ & $0.976109 \pm 0.008359$ & $0.957204 \pm 0.015623$ \\
-SVD\_IPC\_Regime2 & 6,902   & $0.1716 \pm 0.0043$ & $0.1762 \pm 0.0012$ & $0.875984 \pm 0.005161$ & $0.846360 \pm 0.004420$ \\
-SVD\_IPC\_Regime3 & 947     & $0.0507 \pm 0.0011$ & $0.5413 \pm 0.0005$ & $0.776841 \pm 0.007290$ & $0.715747 \pm 0.005605$ \\
-SVD\_IPC\_Regime4 & 2,929   & $0.1288 \pm 0.0471$ & $0.2308 \pm 0.0867$ & $0.907348 \pm 0.008018$ & $0.824626 \pm 0.007707$ \\
-SVD\_IPC\_Regime5 & 3,255   & $0.0235 \pm 0.0022$ & $0.1306 \pm 0.0003$ & $0.733272 \pm 0.007562$ & $0.677575 \pm 0.006792$ \\
-SVD\_IPC\_Regime6 & 5,804   & $0.0196 \pm 0.0022$ & $0.0508 \pm 0.0003$ & $0.810087 \pm 0.013468$ & $0.747545 \pm 0.010607$ \\
+gcc               & 4,542   & $0.1407 \pm 0.0063$ & $0.1487 \pm 0.0066$ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+col\_major        & 14,709  & $0.0252 \pm 0.0061$ & $0.1975 \pm 0.0436$ & $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
+
+%SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.8765 \pm 0.0044$ & $0.8508 \pm 0.0035$ \\
+SVD\_IPC\_Regime1 & 2,158   & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ & $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
+SVD\_IPC\_Regime2 & 6,902   & $0.1716 \pm 0.0043$ & $0.1762 \pm 0.0012$ & $0.8760 \pm 0.0052$ & $0.8464 \pm 0.0044$ \\
+SVD\_IPC\_Regime3 & 947     & $0.0507 \pm 0.0011$ & $0.5413 \pm 0.0005$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
+SVD\_IPC\_Regime4 & 2,929   & $0.1288 \pm 0.0471$ & $0.2308 \pm 0.0867$ & $0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
+SVD\_IPC\_Regime5 & 3,255   & $0.0235 \pm 0.0022$ & $0.1306 \pm 0.0003$ & $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
+SVD\_IPC\_Regime6 & 5,804   & $0.0196 \pm 0.0022$ & $0.0508 \pm 0.0003$ & $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
 \hline

### 2014-01-12 pecomplexity.tex
more text

@@
 
+The weighted permutation entropy for the {\tt SVD} program is given in
+Fig.~\cite{fig:wwpe}. To generate this image a window of 5,000 values slid over
+the time series. Within each of those windows, the statistics over words of
+length 4 are computed and the WPE is calculated. The gray bands denote regions
+where the 5,000 value window overlapped visually-distinct regimes. It can be
+seen that the behaviors of the weighted permutation entropy vary between
+regimes.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
+  \caption{The weighted permutation entropy of one run of SVD. The gray bands
+    are regions where the window overlaps regimes. The window size used is
+    $5,000 \times 100,000$ instructions and the word length is $4$.}
+  \label{fig:wwpe}
+\end{figure}
+
 % Table~\ref{tab:pe} shows the permutation entropy results for the
@@
 predict using the simple deterministic forecast model described in the previous
-section.  The effects of changing the word length $n$ are also interesting:
-using a longer word length generally lowers the WPE --- a natural consequence of
-finite-length data --- but the falloff is less rapid in some traces than in
-others, suggesting that those values are closer to the theoretical asymptote
-that exists for perfect data. The persistent WPE values of 0.5--0.6 for the {\tt
-col\_major} trace are consistent with dynamical chaos, further corroborating the
-results of~\cite{mytkowicz09}.
+section. See Fig.~\cite{fig:pred_vs_ent} for the distribution. The effects of
+changing the word length $n$ are also interesting: using a longer word length
+generally lowers the WPE --- a natural consequence of finite-length data --- but
+the falloff is less rapid in some traces than in others, suggesting that those
+values are closer to the theoretical asymptote that exists for perfect data. The
+persistent WPE values of 0.5--0.6 for the {\tt col\_major} trace are consistent
+with dynamical chaos, further corroborating the results of~\cite{mytkowicz09}.
 
@@
   \centering
-  \includegraphics[width=0.95\textwidth]{figs/prediction_vs_entropy}
+  \includegraphics[width=1.0\textwidth]{figs/prediction_vs_entropy}
   \caption{The improvement of LMA over na\"{i}ve vs weighted permutation entropy.
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.95\textwidth]{figs/SVD_wwpe}
-  \caption{The weighted permutation entropy of one run of SVD. The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    $5,000 \times 100,000$ instructions and the word length is $4$.}
-  \label{fig:wwpe}
-\end{figure}
-
 \section*{Acknowledgment}

### 2014-02-04 unusedFigsTables.tex
adding an unused file for tables and figures

--- 2014-01-11 pecomplexity.tex
@@
-\documentclass{article}
+\documentclass[11pt]{amsart}
 \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
+\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
 %\geometry{landscape}                % Activate for for rotated page geometry
-%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
+%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
 \usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
 \usepackage{amssymb}
-\usepackage{amsmath}
 \usepackage{epstopdf}
-\usepackage{amsthm}
-\usepackage{authblk}
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
 
-\newtheorem*{mydef}{Definition}
-\title{Determinism, Complexity, and Predictability in Computer Performance}
-
-\author[1]{Joshua Garland \thanks{joshua.garland@colorado.edu}}
-\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
-\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-\affil[1]{Department of Computer Science\\
-  University of Colorado at Boulder\\
-  Colorado, USA
-}
-\affil[2]{Santa Fe Institute\\
-  New Mexico, USA
-}
-
-%Version 1
-%\author{
-%  \IEEEauthorblockN{Joshua Garland}
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: joshua.garland@colorado.edu}
-%  \and
-%  \IEEEauthorblockN{Ryan G.~James}
-%  \IEEEauthorblockA{Complexity Sciences Center \& Dept. of Physics\\
-%    University of California at Davis\\
-%    California, USA\\
-%    Email: rgjames@ucdavis.edu}
-%    \and
-%      \IEEEauthorblockN{Elizabeth Bradley}
-%         \IEEEauthorblockA{Santa Fe Institute \\
-%    New Mexico, USA\\
-%    }
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: lizb@colorado.edu}
-%  }
-
- %\author{
- %  \IEEEauthorblockN{
- %    Joshua Garland\IEEEauthorrefmark{1},
- %    Ryan G.~James\IEEEauthorrefmark{1} and
- %    Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{1}Dept. of Computer Science
- %    University of Colorado, Boulder, Colorado 80309-0430 USA\\
- %    Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
- %    Email: lizb@colorado.edu}
- %  % \IEEEauthorblockA{
- %  %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, %California 95616 USA\\
-%   %   Email: rgjames@ucdavis.edu}
-% }
+                                    % Activate to display a given date or no date
 
 \begin{document}
+\section{Unused Figures}
 
-\maketitle
-
-\begin{abstract}
-  Computers are deterministic dynamical systems \cite{mytkowicz09}.
-  Among other things, that implies that one should be able to use
-  deterministic forecast rules to predict aspects of their behavior.
-  That statement is sometimes---but not always---true. The memory and
-  processor loads of some simple programs are easy to predict, for
-  example, but those of more-complex programs like {\tt gcc} are not.
-%%%%%%%%%%%%%%%%%%%%%%
-%% I had to change all the \verb|blah| entries because they caused
-%% latex to barf if they were in figure captions.  Odd bug.
-%%%%%%%%%%%%%%%%%%%%%%
-  The goal of this paper is to determine why that is the case. We
-  conjecture that, in practice, complexity can effectively overwhelm
-  the predictive power of deterministic forecast models. To explore
-  that, we build models of a number of performance traces from
-  different programs running on different Intel-based computers. We
-  then calculate the \emph{permutation entropy}---a temporal entropy
-  metric that uses ordinal analysis---of those traces and correlate
-  those values against the prediction success.
-\end{abstract}
-
-\section{Introduction}
-%\begin{it}
-%Paragraph on computer performance, including citations to Todd paper
-%and summary of the results that indicate that they're deterministic
-%nonlinear dynamical systems.  Given that, we should be able to
-%predict.  What benefits would accrue if we could do so: power mgmt,
-%end world hunger [[this is my primary goal everyday :)]], etc.
-%\end{it}
-
-Computers are among the most complex engineered artifacts in current
-use.  Modern microprocessor chips contain multiple processing units
-and multi-layer memories, for instance, and they use complicated
-hardware/software strategies to move data and threads of computation
-across those resources.  These features---along with all the others
-that go into the design of these chips---make the patterns of their
-processor loads and memory accesses highly complex and hard to
-predict.  Accurate forecasts of these quantities, if one could
-construct them, could be used to improve computer design.  If one
-could predict that a particular computational thread would be bogged
-down for the next 0.6 seconds waiting for data from main memory, for
-instance, one could save power by putting that thread on hold for that
-time period (e.g., by migrating it to a processing unit whose clock
-speed is scaled back).  Computer performance traces are, however, very
-complex.  Even a simple ``microkernel,'' like a three-line loop that
-repeatedly initializes a matrix in column-major order, can produce
-{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:cache}, and chaos places fundamental limits on
-predictability.
-%
-\begin{figure}[htbp]
-   \centering
-   \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
-   % where an .eps filename suffix will be assumed under latex,
-   % and a .pdf suffix will be assumed for pdflatex
-   \caption{A small snippet of the L2 cache miss rate of {\tt
-       col\_major}, a three-line C program that repeatedly initializes
-     a matrix in column-major order, running on an Intel Core
-     Duo\textsuperscript{\textregistered}-based machine.  Even this
-     simple program exhibits chaotic performance dynamics.}
-   \label{fig:cache}
- \end{figure}
-
-The computer systems community has applied a variety of prediction
-strategies to traces like this, most of which employ regression.  An
-appealing alternative builds on the recently established fact that
-computers can be effectively modeled as deterministic nonlinear
-dynamical systems \cite{mytkowicz09}.  This result implies the
-existence of a deterministic forecast rule for those dynamics.  In
-particular, one can use \emph{delay-coordinate embedding} to
-reconstruct the underlying dynamics of computer performance, then use
-the resulting model to forecast the future values of computer
-performance metrics such as memory or processor loads
-\cite{josh-ida2011}.  In the case of simple microkernels like the one
-that produced the trace in Figure~\ref{fig:cache}, this deterministic
-modeling and forecast strategy works very well.  In more-complicated
-programs, however, such as speech recognition software or compilers,
-this forecast strategy---as well as the traditional methods---break
-down quickly.
-
-This paper is a first step in understanding when, why, and how
-deterministic forecast strategies fail when they are applied to
-deterministic systems.  We focus here on the specific example of
-computer performance.  We conjecture that the complexity of traces
-from these systems---which results from the inherent dimension,
-nonlinearity, and nonstationarity of the dynamics, as well as from
-measurement issues like noise, aggregation, and finite data
-length---can make those deterministic signals \emph{effectively}
-unpredictable.  We argue that \emph{permutation entropy}
-\cite{bandt2002per}, a method for measuring the entropy of a
-real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study four
-examples---two simple microkernels and two complex programs from the
-SPEC benchmark suite---running on different Intel-based machines.  For
-each program, we calculate the permutation entropy of the processor
-load (instructions per cycle) and memory-use efficiency (cache-miss
-rates), then compare that to the prediction accuracy attainable for
-that trace using a simple deterministic model.
-
-% paragraph to appease the theoretician in me
-It is worth taking a moment to consider the theoretical possibility of
-this task. We are not attempting to predict the state of the CPU at an
-arbitrary point in the future --- this, at least with perfect
-accuracy, would be tantamount to solving the halting problem. What we
-are attempting is to predict aspects or functions of the running of
-the CPU: instructions executed per second, cache misses per 100,000
-instructions, and similar statistics. Prediction of these quantities
-at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem.
-
-% The rest of the paper is organized as follows.
-% Section~\ref{sec:compModel} describes the experimental setup, as well
-% as the nonlinear modeling and forecast strategies.  In
-% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
-% its value for a number of different computer performance traces, and
-% compare the results to the prediction accuracy.  In
-% Section~\ref{sec:conc}, we discuss these results and their
-% implications in regard to our conjecture, and consider future areas of
-% research.
-
-\section{Modeling Computer Performance}\label{sec:compModel}
-
-% took out for space
-% \subsection{Reconstructing hidden dynamics}
-
-Delay-coordinate embedding allows one to reconstruct a system's full
-state-space dynamics from a \emph{single} scalar time-series
-measurement---provided that some conditions hold regarding that data.
-Specifically, if the underlying dynamics and the measurement
-function---the mapping from the unknown state vector $\vec{X}$ to the
-scalar value $x$ that one is measuring---are both smooth and generic,
-Takens~\cite{takens} formally proves that the delay-coordinate map
-\[
-F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
-\]
-from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
-where $t$ is time, is a diffeomorphism on $M$---in other words, that
-the reconstructed dynamics and the true (hidden) dynamics have the
-same topology.
-
-This is an extremely powerful result: among other things, it means
-that one can build a formal model of the full system dynamics without
-measuring (or even knowing) every one of its state variables.  This is
-the foundation of the modeling approach that is used in this paper.
-The first step in the process is to estimate values for the two free
-parameters in the delay-coordinate map: the delay $\tau$ and the
-dimension $m$.  We follow standard procedures for this, choosing the
-first minimum in the average mutual information as an estimate of
-$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
-plot of the data from Figure~\ref{fig:cache}, embedded following this
-procedure, is shown in Figure~\ref{fig:embedding}.
-\begin{figure}
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/3dcolcache}
-    \caption{A 3D projection of a delay-coordinate embedding of the trace
-from Figure~\ref{fig:cache} with a delay ($\tau$) of 100,000 instructions.
-}
-\label{fig:embedding}
-\end{figure}
-%% can cut for space if need be:
-The coordinates of each point on this plot are differently delayed
-elements of the \verb|col_major| L2 cache miss rate time series
-$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
-$y(t+2\tau)$ on the third, and so on.
-%% ...down to here.
-Structure in these kinds of plots---clearly visible in
-Figure~\ref{fig:embedding}---is an indication of
-determinism\footnote{A deeper analysis of
-  Figure~\ref{fig:embedding}---as alluded to on the previous
-  page---supports that diagnosis, confirming the presence of a chaotic
-  attractor in these cache-miss dynamics, with largest Lyapunov
-  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
-  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
-structure can also be used to build a forecast model.
-
-% took out for space
-% \subsection{LMA: Using dynamics in forecasting}
-
-Given a nonlinear model of a deterministic dynamical system in the
-form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
-one can build deterministic forecast algorithms by capturing and
-exploiting the geometry of the embedding.  Many techniques have been
-developed by the dynamical systems community for this purpose
-(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
-is the ``Lorenz method of analogues'' (LMA), which is essentially
-nearest-neighbor prediction in the embedded state
-space~\cite{lorenz-analogues}.  Even this simple algorithm---which
-builds predictions by finding the nearest neighbor in the embedded
-space of the given point, then taking that neighbor's path as the
-forecast---works quite well on the trace in Figure~\ref{fig:cache}, as
-shown in Figure~\ref{fig:cachePredTS}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
-    \caption{A forecast of the last 4,000 points of the signal in
-      Figure~\ref{fig:cache} using an LMA-based strategy on the
-      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
-      $\times$s are the true and predicted values, respectively;
-      vertical bars show where these values differ. }
-\label{fig:cachePredTS}
-\end{figure}
-%
-On the other hand, if we use the same approach to forecast the
-processor load\footnote{Instructions per cycle, or IPC} of the {\tt
-  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
-on an Intel i7\textsuperscript{\textregistered}-based machine, the
-prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
-     \caption{An LMA-based forecast of the last 4,000 points of a
-       processor-load performance trace from the {\tt 482.sphinx3}
-       benchmark.  Red circles and blue $\times$s are the true and
-       predicted values, respectively; vertical bars show where these
-       values differ.}
-\label{fig:predsphinx}
-\end{figure}
-
-Table~\ref{tab:PredError} presents detailed results about the
-prediction accuracy of this algorithm on four different examples: the
-{\tt col\_major} and {\tt 482.sphinx3} programs in
-Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as
-another simple microkernel that initializes the same matrix as {\tt
-  col\_major}, but in row-major order, and another complex program
-({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.  Both
-microkernels were run on the Intel Core
-Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks
-were run on the Intel i7\textsuperscript{\textregistered} machine.  We
-calculated a figure of merit for each prediction as follows.  We held
-back the last $k$ elements\footnote{Several different prediction
-  horizons were analyzed in our experiment; the results reported in
-  this paper are for $k$=4000} of the $N$ points in each measured time
-series, built the forecast model by embedding the first $N-k$ points,
-used that embedding and the LMA method to predict the next $k$ points,
-then computed the Root Mean Squared Error (RMSE) between the true and
-predicted signals:
-$$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$
-%
-To compare the success of predictions across signals with different
-units, we normalized RMSE as follows:
-$$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
-%
-% The smaller the nRMSE, obviously, the more accurate the prediction.
-
-  \begin{table}[htbp]
-   % increase table row spacing, adjust to taste
-   \renewcommand{\arraystretch}{1.3}
-   \caption{Normalized root mean squared error (nRMSE) of 4000-point
-     predictions of memory \& processor performance from different
-     programs.}
-   \label{tab:PredError}
-   \centering
-   % Some packages, such as MDW tools, offer better commands for making tables
-   % than the plain LaTeX2e tabular which is used here.
-%   \begin{tabular}{|c|c|c|c|c|c|}
-%     \hline
-%      & cache misses NRMSE & IPC NRMSE \\
-%     \hline
-%     \tt{row\_major}  & 0.0102  & 0.0095 \\
-%          \hline
-%     \tt{col\_major}  &0.0055& 0.0071 \\
-%     \hline
-%     \tt{403.gcc}  & 0.0865& 0.1805 \\
-%          \hline
-%     \tt{482.sphinx3} & 0.1142& 0.2946 \\
-%      \hline
-%   \end{tabular}
-% \end{table}
-   \begin{tabular}{|c|c|c|c|c|c|}
-     \hline
-      & cache miss rate & instrs per cycle \\
-     \hline
-     \tt{row\_major}  & 0.0324  & 0.0778 \\
-          \hline
-     \tt{col\_major}  &0.0080& 0.0161 \\
-     \hline
-     \tt{403.gcc}  & 0.1416& 0.2033 \\
-          \hline
-     \tt{482.sphinx3} & 0.2032& 0.3670 \\
-      \hline
-   \end{tabular}
- \end{table}
-
-The results in Table~\ref{tab:PredError} show a clear distinction
-between the two microkernels, whose future behavior can be
-predicted effectively using this simple deterministic modeling
-strategy, and the more-complex SPEC benchmarks, for which this
-prediction strategy does not work nearly as well.
-% Removed for space
-% For both processor load (IPC) and memory usage (cache-miss rate),
-% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
-% for \verb|col_major| or \verb|row_major|.
-%
-This begs the question: If these traces all come from deterministic
-systems---computers---then why are they not equally predictable?  Our
-conjecture is that the sheer complexity of the dynamics of the SPEC
-benchmarks running on the Intel i7\textsuperscript{\textregistered}
-machine make them effectively impossible to predict.
-
-%% [[If space: Add a segue sentence: next section uses permutation
-%% entropy to explore that conjecture.]]
-
-\section{Measuring Complexity}\label{sec:meaComplex}
-
-For the purposes of this paper, one can view entropy as a measure of complexity
-and predictability in a time series.  A high-entropy time series is almost
-completely unpredictable---and conversely.  This can be made more rigorous:
-Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
-Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
-$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
-states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
-difficult prediction becomes.
-
-Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
-categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
-\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
-effectively real-valued.  To get around this, one must discretize the data---
-typically achieved by binning.  Unfortunately, this is rarely a good solution to
-the problem, as the binning of the values introduces an additional dynamic on
-top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
-dynamics studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in the face
-of noise and require further understanding of the underlying system, which
-defeats the purpose of measuring the entropy in the first place.
-
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
-complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
-employs a method of discretizing real-valued time series that follows the
-intrinsic behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the Shannon
-entropy, permutation entropy looks at the statistics of the \emph{orderings} of
-sequences of values using ordinal analysis. Ordinal analysis of a time series is
-the process of mapping successive time-ordered elements of a time series to
-their value-ordered permutation of the same size.  By way of example, if $(x_1,
-x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
-other things, it is generally robust to observational noise and requires no
-knowledge of the underlying mechanisms.
-
-\begin{mydef}[Permutation Entropy]
-
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
-  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
-  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
-  = 1,\dots,T}$:
-  \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
-  \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
-  \ge 2$ is defined as
-  \begin{align*}
-  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
-  \end{align*}
-
-\end{mydef}
-
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
-it is common in the literature to normalize permutation entropy as follows:
-$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
-and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
-many large classes of systems \cite{amigo2012permutation}.
-
-Here we will be utilizing a variation of the permutation entropy, the
-\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
-permutation entropy attempts to correct for observational noise which is larger
-than some trends in the data, but smaller than the larger scale features --- for
-example, a signal that switches between two fixed points with noise about those
-fixed points. The weighted permutation entropy would be dominated by the
-switching rather than by the stochastic fluctuation. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
-\begin{align*}
-  w(x_{t+1:t+n}) = \frac{1}{n}
-                 \sum_{x_i \in \{x_{t+1:t+n}\}}
-                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
-\end{align*}
-where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
-$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
-
-The weighted probability of a permutation is then:
-\begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
-\end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
-probability enhances permutations involved in ``large'' features and demotes
-permutations which are small in amplitude relative to the features of the time
-series. The weighted permutation entropy is then:
-\begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
-\end{align*}
-which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
-results of this paper.
-
-In practice, calculating permutation entropy and weighted permutation entropy
-involves choosing a good value for the word length $n$. The primary
-consideration is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the ordinals are
-gathered: e.g.:
-\begin{align*}
-  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
-\end{align*}
-assuming an average of 100 counts per ordinal is sufficient. In the literature,
-$3 \le n \le 6$ is a standard choice --- generally without any formal
-justification. In theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In practice,
-what one should do is calculate the \emph{persistent} permutation entropy by
-increasing $n$ until the result converges, but data length issues can intrude
-before that convergence is reached.
-
-% Table~\ref{tab:pe} shows the permutation entropy results for the
-% examples considered in this paper, with the nRMSPE prediction
-% accuracies from the previous section included alongside for easy
-% comparison.
-%
-% \begin{table}[htbp]
-%   % increase table row spacing, adjust to taste
-%   \renewcommand{\arraystretch}{1.3}
-%   \caption{TCM NRMSE and PE NORMS}
-%   \label{tab:table_example}
-%   \centering
-   % Some packages, such as MDW tools, offer better commands for making tables
-   % than the plain LaTeX2e tabular which is used here.
-%   \begin{tabular}{|c|c|c|c|c|c|}
-%     \hline
-%     TCM & NRMSE & $m=3$  &$m=4$&$m=5$&$m=6$\\
-%     \hline
-%     \tt{row\_major} & 0.0102 & 0.7900  &0.6751&0.5458&0.4491\\
-%     \hline
-%     \tt{col\_major} & 0.0055 & 0.6411  &0.5029&0.4515&0.3955 \\
-%     \hline
-%     \tt{403.gcc} & 0.0865 & 0.9954  &0.9916&0.9880&0.9835\\
-%     \hline
-%     \tt{482.sphinx3} & 0.1142 & 0.9958  &0.9913&0.9866&0.9802 \\
-%      \hline
-%   \end{tabular}
-% \end{table}
-%
-%%  \begin{table}[htbp]
-%%   % increase table row spacing, adjust to taste
-%%   %\renewcommand{\arraystretch}{1.3}
-%%   \caption{Cache-miss rate performance traces: prediction error
-%%     (nRMSPE) and permutation entropy}
-%%   \label{tab:TCMpe}
-%%   \centering
-%%   % Some packages, such as MDW tools, offer better commands for making tables
-%%   % than the plain LaTeX2e tabular which is used here.
-%%   \begin{tabular}{|c|c|c|c|c|c|}
-%%     \hline
-%%      & nRMSE   &$n=4$&$n=5$&$n=6$\\
-%%     \hline
-%%     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
-%%     \hline
-%%     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
-%%     \hline
-%%     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
-%%     \hline
-%%     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
-%%      \hline
-%%   \end{tabular}
-%% \end{table}
-%
-% \begin{table}[htbp]
-%   % increase table row spacing, adjust to taste
-%   \renewcommand{\arraystretch}{1.3}
-%   \caption{IPC NRMSE and PE NORMS}
-%   \label{tab:table_example}
-%   \centering
-%   % Some packages, such as MDW tools, offer better commands for making tables
-%   % than the plain LaTeX2e tabular which is used here.
-%   \begin{tabular}{|c|c|c|c|c|c|}
-%     \hline
-%     ipc & NRMSE & $n=3$  &$n=4$&$n=5$&$n=6$\\
-%     \hline
-%      \tt{row\_major} & 0.0095 & 0.9912  &0.9723&0.9354&0.8876\\
-%     \hline
-%      \tt{col\_major} & 0.0071 & 0.9428  &0.8356&0.7601&0.6880\\
-%     \hline
-%     \tt{403.gcc} & 0.1805 & 0.9918  &0.9862&0.9814&0.9764\\
-%     \hline
-%     \tt{482.sphinx3} & 0.2946 & 0.9978  &0.9951&0.9914&0.9849 \\
-%      \hline
-%   \end{tabular}
-% \end{table}
-%
-%%  \begin{table}[htbp]
-%%   % increase table row spacing, adjust to taste
-%%   %\renewcommand{\arraystretch}{1.1}
-%%   \caption{Processor load performance traces: prediction error
-%%     (nRMSPE) and permutation entropy}
-%%   \label{tab:IPCpe}
-%%   \centering
-%%   % Some packages, such as MDW tools, offer better commands for making tables
-%%   % than the plain LaTeX2e tabular which is used here.
-%%   \begin{tabular}{|c|c|c|c|c|}
-%%     \hline
-%%      & nRMSPE   &$n=4$&$n=5$&$n=6$\\
-%%     \hline
-%%      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
-%%     \hline
-%%      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
-%%     \hline
-%%     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
-%%     \hline
-%%     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
-%%      \hline
-%%   \end{tabular}
-%% \end{table}
-%
- %  \begin{table}[htbp]
- %   \caption{Prediction error (in nRMSPE) and permutation entropy (for
- %     different wordlengths $n$}
- %   \label{tab:pe}
- %   \centering
- %   \begin{tabular}{|c|c|c|c|c|c|}
- %     \hline
- %      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
- %     \hline
- %     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
- %     \hline
- %     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
- %     \hline
- %     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
- %     \hline
- %     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
- %      \hline
- %      \hline
- %      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
- %     \hline
- %      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
- %     \hline
- %      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
- %     \hline
- %     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
- %     \hline
- %     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
- %      \hline
- %   \end{tabular}
- % \end{table}
-
-The relationship between prediction accuracy and the weighted permutation
-entropy (WPE) is much as we conjectured: performance traces with high WPE ---
-those whose temporal complexity is high, in the sense that the entropy
-generation per value is approximately equal to the total entropy per value thus
-little information is being propagated forward in time --- are indeed harder to
-predict using the simple deterministic forecast model described in the previous
-section.  The effects of changing the word length $n$ are also interesting:
-using a longer word length generally lowers the WPE --- a natural consequence of
-finite-length data --- but the falloff is less rapid in some traces than in
-others, suggesting that those values are closer to the theoretical asymptote
-that exists for perfect data. The persistent WPE values of 0.5--0.6 for the {\tt
-col\_major} trace are consistent with dynamical chaos, further corroborating the
-results of~\cite{mytkowicz09}.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.95\textwidth]{figs/prediction_vs_entropy}
-  \caption{The improvement of LMA over na\"{i}ve vs weighted permutation entropy.
-  For each of these, the word length used is $5$.}
-  \label{fig:pred_vs_ent}
-\end{figure}
-
-% In Table~\ref{tab:TCMpe} \& \ref{tab:IPCpe} the behavior of the PEs as
-% they limit to their asymptotic values is observed. A feature of note
-% is that some of the signals drop off and some do not.  For example at
-% $m=4$ {\tt col\_major} is at 0.8356, a fairly high entropy, however as
-% $m$ is increased to 6 the PE plummets to 0.6880: a value consistent
-% with a chaotic time series. Notice this also corresponds to a very low
-% nRMSE. In contrast, consider the permutation entropy of the processor
-% load of {\tt 482.sphinx3}: with $m=4$ it is 0.9951 and only drops to
-% 0.9849 at $m=6$. This is consistent with the signal also having the
-% highest nRMSE.
-
-\section{ Conclusions \& Future Work}\label{sec:conc}
-
-The results presented here suggest that permutation entropy---a
-ordinal calculation of forward information transfer in a time
-series---is an effective metric for predictability of computer
-performance traces. Experimentally, traces with a persistent PE
-$\gtrapprox 0.97$ have a natural level of complexity that may
-overshadow the inherent determinism in the system dynamics, whereas
-traces with PE $\lessapprox 0.7$ seem to be highly predictable (viz.,
-at least an order of magnitude improvement in nRMSPE).
-
-If information is the limit, then gathering and using more information
-is an obvious next step.  There is an equally obvious tension here
-between data length and prediction speed: a forecast that requires
-half a second to compute is not useful for the purposes of real-time
-control of a computer system with a MHz clock rate.  Another
-alternative is to sample several system variables simultaneously and
-build multivariate delay-coordinate embeddings.  Existing approaches
-to that are computationally prohibitive
-\cite{cao-multivariate-embedding}.  We are working on alternative
-methods that sidestep that complexity.
-
-%%This use of permutaiotn entropy is a useful application of information
-%%theory to computer performance modeling because it gives you a simple
-%%and fast way to decide whether building a model and trying to forecast
-%%the future is worthwhile or if guessing the mean is just as effective.
-%%End with a sentence tying back to power management and world peace.
-
-%\begin{it}
-%Summarize the results.
-
-%Paragraphs on the issues that come up, including one about the "amount
-%of info" one: if one could sample more variables, for instance, one
-%might be able to do a better job of predicting more-complex traces.
-%Segue to some handwaving about multivariable LMA models; tie this back
-%to the "computers are NLD systems" stuff in the intro.  This is a real
-%challenge; current approaches to this modelling problem have the major
-%issue of taking way too long to build.  And that's a big issue if
-%you're trying not just to classify, but to predict.  In a system that
-%runs at MHz speeds, a prediction that takes milliseconds to compute is
-%not useful.
-%\end{it}
-
-%[[Joshua: It may be interesting to look at the PE as a function of
-%prediction horizon. It is known to me that certain start and end
-%locations in the prediction strategy result in different prediction
-%accuracy. Is it the case that when the algorithm tanks it is also the
-%case that PE of the **TEST**(the 10\% we cut off )is high? This may
-%be a great way to illustrate the concept if it works... OR maybe do
-%complexity of forward path in embedding space?]]  some notes about
-%this looking at row Cache 1000 very good RMSPE 21.4167 m = 3 4 5 6
-%1.4192 2.1286 2.5942 2.9230 row Cache 2000 very good RMSPE 19.0990 m
-%= 3 4 5 6 1.4145 2.1305 2.5825 2.9095 so even better row Cache 2010
-%RMSPE 19.9716 m = 3 4 5 6 1.4149 2.1311 2.5830 2.9096 row Cache 1990
-%RMSPE 19.1465 So prediction Horizons ranked in order of best RMSPE
-%RMSPE L 3 4 5 6 1st 19.0990 2000 1.4145 2.1305 2.5825 2.9095 2nd
-%19.1465 1990 1.4151 2.1293 2.5803 2.9067 3rd 19.9716 2010 1.4149
-%2.1311 2.5830 2.9096 3rd 21.4167 1000 1.4192 2.1286 2.5942 2.9230
-%ipcgcc m = 3 pe = 1.7770 penorm = 0.9918 m= 4 pe = 3.1341 penorm =
-%0.9862 m = 5 pe = 4.6985 penorm = 0.9814 m=6 pe = 6.4242 penorm =
-%0.9764 m=7 pe = 8.2450 penorm = 0.9671 >> [pe penorm] =
-%permCalc(gcc100TCM,5,1) pe = 4.7299 penorm = 0.9880 >> [pe penorm] =
-%permCalc(gcc100TCM,6,1) pe = 6.4707 penorm = 0.9835
-
-%****BEYOND THIS POINT IS A DIFFERNT PAPER*** JOSHUA WILL CLEAN THIS OUT, JUST HERE FOR REFERNCE****
-%\section{Introduction: What is Permutation Entropy}
-
-%%\begin{it}
-%%Here I want to talk about the definition and algorithm for computing it.
-Maybe briefly talk about the asymptotic %properties.
-(i.e., same as metric topological in limit.
-Converges to dominate leap if not normalized etc.
-)
-%%\end{it}
-%
-%When analyzing a time series there are several standard characterizations of complexity, including (but not limited to) entropy, Lyapunov exponents and fractal dimensions.
-Rigorous definitions of all these quantities exist if the observation is a \emph{noise-free} trajectory of a \emph{single} dynamical system.
-To an experimental-time-series analyst these are luxuries that are rarely (if ever) realized.
-Many times experimental observation have noise involved such as sensor error, or very commonly in the computer world finite-precision error.
-%
-%Unfortunately the standard algorithms for calculating many of these complexity measures are highly sensitive to noise and give spurious results in the presence of noisy observations.
-Even worse, if a time-series observation measures multiple dynamical systems or dynamical systems that undergo bifurcations\footnote{A drastic change in the behavior of the dynamical system, generally associated with parameter drift.} then all of the standard algorithms for estimating complexity, (e.g., correlation dimension) go out the window.
-Bandt and Pompe first introduced \emph{permutation entropy} as a ``natural complexity measure for time series" in 2002 \cite{bandt2002per}.
-%
-%
-%As explained in \cite{bandt2002per}, permutation entropy is a simple complexity measure that is easily calculated for any time series, be it stochastic, periodic, quasi-periodic, chaotic, noisy or any combination of these.
-Most importantly for analysis of computer performance, permutation entropy as defined in \cite{bandt2002per} yields meaningful results even if observational or dynamical noise is present \cite{bandt2002per}.
-%
-%
-%
-%
-%Analyzing entropies as  a complexity measure is not a new concept, however the way that the entropy is estimated in \cite{bandt2002per} is novel and particularly useful when little is known about the underlying process.
-As discussed in \cite{bandt2002per}, the standard entropy one considers is Shannon entropy.
-%
-%
-%As a quick review of Shannon entropy, we follow the treatment in \cite{amigo2012permutation}.
-Given a finite alphabet $\mathcal{S} = \{s_1,....,s_{|\mathcal{S}|}\}$, define an information source as a discrete-time, stationary stochastic process $X = \{X_n\}_{n \in \mathbb{N}_0}$ where $X_n$ are random variables on a common probability space taking on values from $\mathcal{S}$.
-A realization of $X$ is a one-sided sequence $x_0^{\infty} = (x_n)_{n \in \mathbb{N}_0}$ called a \emph{message}.
-The elements $x_n$ of $\mathcal{S}$ are the \emph{symbols}(a.k.a letters) of the language.
-A finite segment of the message, say $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$ is called a \emph{word} (of length m).
-We then define $p(x_0^{m-1})$ to be the probability of the word $x_0^{m-1}$ to appear some where in the message.
-The shannon entropy of the data source $X$ is defined as: \begin{equation}\label{eq:shannonentropy}
-%h(X) = - \lim_{m\to\infty}\frac{1}{m}\sum p(x_0^{m-1})\log p(x_0^{m-1})
-%\end{equation}
-%
-%For the purposes of this paper once can view entropy as a measure of complexity and predictability in a time series.
-A high entropy almost completely unpredictable and a small entropy is completely predictable.
-This spectrum is outlined very nicely in \cite{amigo2012permutation}: ``Periodic or quasiperiodic sequences have vanishing or negligible complexity.
-At the opposite end, independent and identically distributed random sequences (white noise) have asymptotically divergent permutation entropies, owing to the fact that the number of allowed (or ``admissibleÓ) ordinal patterns grows superexponentially with length.
-Between both ends lie the kind of sequences we are interested in."
-%
-%This illustrates why utilizing entropy as a measure for complexity is, in theory, a great candidate.
-There are several fundamental drawbacks with this approach when analyzing real-world time series.
-The two most fundamental problems are as follows:
-%
-%\begin{enumerate}
-%\item Elements of an experimental time series $x_t \in \mathbb{R}$ which implies $x_t \notin \mathcal{S}$, for a finite alphabet $\mathcal{S}$.
-%\item Time series are always finite in length so the limit in \ref{eq:shannonentropy} never exists in practice.
-%\end{enumerate}
-%
-%
-%The common practice to address item 1 is to perform \emph{symbolization} on the time series you want to analyze.
-Symbolization is the process of defining a dynamic preserving map $\phi: \mathbb{R}^m \to \mathcal{S}$, i.e., for each $x_t$ (or block $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$) we must assign a symbol from $\mathcal{S}$ in such a way that the time series has the same fundamental properties.
-There are several approaches to symbolization, the standard approach is partitioning.
-Assume, $x_t$ has the capability of taking on infinitely many values, defining the space $X$ to be the space where these $x_t$ are drawn.
-We then define a partition on this space $X = P_1 \cup...\cup P_m$, and define $\mathcal{S} = \{1,\dots,m\}$.
-Then $\phi(x_t) = i$ if $x_t \in P_i$.
-We can then calculate the Shannon entropy of the transformed time series, which in the limit of $m$ converges to the Kolmogorov-Sinai entropy \cite{bandt2002per}.
-This approach to symbolization is computationally infeasable as we need an infinite number of partitions of an infinite time series.
-For \emph{extremely} rare cases, there exists a \emph{generating} partition where this limit is not necessary.
-However, finding such a partition even for a simple two-dimensional map like Henon is non-trivial \cite{bandt2002per}.
-Finding a generating partition for an arbitrary time series is for all intensive purposes impossible.
-%
-%[[[Maybe also talk about Mischaikow's approach to this with isolating neighborhoods and homology...but might be off topic. ]]]
-%
-%
-%.....
-%
-%
-%
-%Instead of this approach to symbolization, Bandt et al.
-take the approach that ``the symbol sequence must come naturally for the time series, without any further model assumptions" \cite{bandt2002per}.
-To do this they perform symbolization through ordinal analysis.
-Ordinal analysis of a time series is the process of mapping the ordered successive elements of a time series to a permutation of the same size based on their original time-based ordering.[[Need a better way of wording this]].
-This is also referred to in the literature as the permutation complexity \cite{amigo2012permutation}  For example, if $(x_1,x_2,x_3) = (9,1,7)$ then $\phi$ maps $(x_1,x_2,x_3)$ to the permutation $231$ since $x_2 < x_3<x_1$.
-A great explanation of why this is an interesting symbolization is found in \cite{amigo2012permutation} :
-%
-%\begin{quote}The study of permutation complexity, which we call ordinal analysis, can be envisioned as a new kind of symbolic dynamics whose basic blocks are ordinal patterns.
-Interesting enough, it turns out that under some mild mathematical assumptions, not all ordinal patterns can be materialized by the orbits of a given one- or multi-dimensional deterministic dynamics, not even if this dynamic is chaoticÑ contrarily to what happens with the symbol patterns.
-As a result, the existence of ÒforbiddenÓ (i.e., not occurring) ordinal patterns is always a persistent dynamical feature, in opposition to properties such as proximity and correlation which die out with time in a chaotic dynamic.
-Moreover, if an ordinal pattern is forbidden, its absence pervades all longer patterns in form of more missing ordinal patterns, called outgrowth forbidden patterns.
-Admissible ordinal patterns grow exponentially with length, while forbidden patterns do superexponentially.
-Since random (unconstrained) dynamics has no forbidden patterns with probability 1, their existence can be used as a fingerprint of deterministic orbit generation.
-%\end{quote} Essentially by analyzing the permutation entropy of a time series we are gaining information as to the complexity of the system without any other information available, whether the time series is generated by a random process or the trajectory of a deterministic dynamical system.
-Now that we have justified why the study of permutation entropy is interesting let us define it.
-For this we follow \cite{bandt2002per}.
-%
-%
-%\begin{mydef}[Permutation Entropy]
-%Given a time series $\{x_t\}_{t = 1,\dots,T}$.
-Define $\mathcal{S}_n$ as all $n!$ permutations $\pi$ of order $n$.
-For each $\pi \in \mathcal{S}_n$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$: \begin{equation}\label{eq:permprob}
-%p(\pi) = \frac{|\{t|t\le T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}|}{T-n+1}\end{equation}
-%Where $|\cdot|$ is set cardinality.
-The \emph{permutation entropy} of order $n\ge 2$ is defined as $$H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi)\log_2p(\pi)$$ Notice that $0\le H(n) \le \log_2(n!)$,\cite{bandt2002per}.
-With this in mind, it is common in the literature to present permutation entropy normalized, i.e., $\frac{H(n)}{\log_2(n!)}$.
-With this convention, \emph{low} entropy is close to 0 and \emph{high} entropy is entropy close to 1.
-%\end{mydef}
-%
-%Notes: According to \cite{bandt2002per} equation (\ref{eq:permprob}) estimates the frequency of $\pi$ as good as possible for a finite series of values.
-To determine $p(\pi)$ exactly, we have to assume an infinite time series and take the limit as $T\to \infty$ in (\ref{eq:permprob}).
-Bandt et al., also point out that this limit exists under a very weak stationarity condition: for $k\le n$, the probability for $x_t<x_{t+k}$ should not depend on $t$.
-%
-%It should also be noted that this is discrete-metric (or shannon) permutation entropy, to be distinguished from topological permutation entropy (a.k.a.
-permutation capacity).
-Topological permutation entropy is defined in \cite{amigo2012permutation} as: $$h(\{x_t\}_{t = 1,\dots,\infty}) =  \lim_{L\to\infty}\frac{1}{L}logN(L)$$ where $N(L)$ is the number of distinct ordinal patterns that occur of order $L$.
-This is a measure of how many distinct permutations of length $L$ are realized as opposed to counting the frequency at which each occurs as is the case in metric permutation entropy \cite{amigo2012permutation}.
-It can be shown that metric and topological entropy both converge to the true value with probability 1 in the limit.
-(citation for this).
-When we say permutation entropy in this paper we are referring to that defined in \ref{eq:permprob}.
-%
-%\section{Detecting changes in a time series}
-%
-%Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}.
-To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}.
-Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block.
-(A window shift of 1.)} (or nonoverlapping) blocks.
-Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.
-According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
-%
-%In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy.
-This is identical to the lag used in delay coordinate embedding and often the same language is used.
-Cao et al.
-describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space.
-When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$.
-So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$.
-Cao et al.
-argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1.
-Cao et al.
-then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter.
-The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
-%
-%\subsection{Example: Concatenated Time Series}\label{sec:concatenate}
-%As an example of this detection, consider Figure \ref{fig:ts-banded} (a).
-This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise.
-Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points.
-The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population.
-Each Logistic map trajectory begins at $x=0.6$.
-The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
-%
-%The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$).
-Visually the bands of noise and the chaotic trajectory are indistinguishable.
-If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise.
-If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
-%
-%In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves).
-The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively).
-Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}.
-On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands.
-This is in fact the case when comparing to the way the time series was constructed.
-%
-%The main advantage we can see to the use of lags other than 1 is for analysis of periodicity.
-This is also the major advantage listed in \cite{cao2004det}.
-Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero.
-This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}.
-If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
-%
-%%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
-%
-%\begin{figure}
-%  \centering
-%  \subfigure{\includegraphics[width=\columnwidth]{figs/ts-banded-alone}}
-%    {(a)}
-%
-%  \subfigure{\includegraphics[width=\columnwidth]{figs/ts-banded-pe}}
-%  {(b)}
-%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}.
-(b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves).
-Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
-%\end{figure}
-%
-%
-%
-%
-%
-%
-%and what is it good for with forward pointer to the other sections
-%
-%
-%
-%\subsection{Example: The transient Logistic map}
-%In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed.
-As we showed, permutation was a great proxy to detect this change.
-Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time.
-To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}.
-Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$.
-The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map.
-The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
-%
-% As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well.
-An interesting feature of block calculations are the spikes that occur at bifurcation points.
-As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes.
-This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy.
-The dips you see in the second half correspond to bands of periodicity which occur between the chaos.
-It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range.
-This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
-%
-% This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\columnwidth]{figs/tstranslog-perm}
-%  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
-%  \label{fig:translogistic}
-%\end{figure}
-%%\begin{enumerate}
-%%\item do with henon bifurcation. what happens near the broken chaotic window
-%
-%%\end{enumerate}
-%
-%
-%\section{Possible Applications to Computer Systems Performance}
-%
-%
-%\begin{enumerate}
-%\item In this section I want to discuss the symbolization of performance traces and why this is advantageous over generating partitions and similar approaches.
-%\item No model is assumed, some traces do appear to be stochastic some appear to be deterministic this provides ``proof" thereof. albeit not mathematical proof but CS proof for sure.
-%\item We have essentially infinite data. So limits like $m!$ where $m$ is word length isn't really an issue for us.
-%\end{enumerate}
-%
-%
-%\subsection{Predictability and model change detection}
-%
-%Talk about how some explanatory models may be spinning their wheels because no transferable information.
-That is if you are introducing pure entropy at every time step really no use in using past values and statistically the mean is the most accurate.
-%
-%\begin{enumerate}
-%\item Comp Performance goes under constant bifurcations.
-This allows autonomous detection of when a model is no longer valid.
-(need to be careful here with use of valid as to not be confused with model validation.
-%\item Maybe can tell us if we should use a linear-stochastic, nonlinear-deterministic model or just guess the mean
-%\end{enumerate}
-%
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\columnwidth]{figs/ts-rowcol-pe}
-%  \caption{A time series of cache misses during the execution of the row-col microkernel.
-The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=5,6$ (respectively).
-}\label{fig:rowcol}
-%\end{figure}
-%
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\columnwidth]{figs/ts-gcc-pe}
-%  \caption{A time series of instructions per cycle for the gcc spec benchmark.
-The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=3,4$ (respectively).
-The lack of structure in such a signal illustrates that the use of deterministic models is most likely not the most advantageous.}\label{fig:gcc}
-%\end{figure}
-%
-%
-%
-%\section{Pro and Cons}
-%\subsection{Advantages of Permutation Entropy}
-%Permutation entropy allows for symbolization of a time series without any further assumptions being made about the underlying system.
-In fact, virtually nothing needs to be known before applying this techniques.
-Other methods of symbolization require expert knowledge of the underlying system, such as generating partitions.
-These techniques have been applied to small data sets, this is truly an advantage especially when dealing with a lot of biological systems whose large datasets can be in the tens of thousands.
-The algorithm for computing permutation entropy is very straight forward and easy to explain.
-The actual code takes longer than I would like to run, but this seems to be an issue with my implementation and not the algorithm, as others state they have exceptionally fast algorithms.
-Although none of these algorithms are ever provided in the literature.
-The major advantages of permutation entropy is that it can quickly give us a rough estimate of where drastic changes in the time series which may have implications in model selection techniques.
-Permutation entropy also can give you a quick idea about if a system has deterministic structure.
-This has direct implications on what prediction strategies to implement.
-%%\begin{enumerate}
-%%\item Allows for symbolization without a model. Really emphasize we can obtain a symbolization without generating partitions which is the traditional approach here.
-%%\item Small amounts of data not really a problem.
-%%\item Maybe talk about how much data is needed per word length for results that mean anything.
-%%\item Fairly easy to calculate, although need a more efficient method.
-%%\item Gives a rough estimation of where drastic changes have occurred in a TS
-%%\item Allows you to see if there is actually something predictable in the signal.
-%%\end{enumerate}
-%
-%\subsection{Disadvantages of PE}
-%There are three disadvantages with permutation entropy we can immediately see that need further thought to be overcome.
-The first is that it is sometimes hard to tell when a change in the dynamical system is not drastic enough to have a huge change on the entropy.
-For example, consider a dynamical system switching from a chaotic to a (different) chaotic regime.
-The permutation entropy may be very similar while the model is slightly different.
-Permutation entropy seems to be ill-suited for fine grain detection of this nature.
-Our current implementation is too slow for on-the-fly analysis.
-We do not believe this is an issue with permutation entropy itself but with our implementation.
-The literature all claims that fast algorithms exist.
-%
-%The biggest disadvantage for us in using permutation entropy is selection of parameters.
-To calculate permutation entropy, we have to choose an embedding dimension, a lag, and a block size.
-In all the literature we have found thus far, parameter selection is made based on persistence or matching the permutation entropy to known values.
-The problem comes when analyzing real time series these parameters seem to have a real impact on the results.
-Finding (or creating) a rigorous method for parameter selection is paramount in the use of this method for the analysis of real world time series.
-%%\begin{enumerate}
-%%\item Paramaters!
-%%\item Currently numerically expensive although the literature disagrees. Think this might just be my implementation. Need to read the numerics section of the PE book.
-%%\item Sometimes hard to tell a regime shift has occurred if the transition is chaotic to chaotic. or low period to low period.
-%%\end{enumerate}
-
-\section{New Figures and Tables}
-
-\begin{table}[htdp]
-\caption{ Embedding Parameters for reference if needed. }
-\begin{center}
-\begin{tabular}{|c|c|c|}
-\hline
-&$\tau$ & $m$ \\
-\hline
-gcc & 10& 13\\
-col\_major  &2&12\\
-SVD\_Full &  10  &  12 \\
-SVD\_IPC\_Regime1 & 5 &14  \\
-SVD\_IPC\_Regime2 & 10   &12  \\
-SVD\_IPC\_Regime3 & 2   &9  \\
-SVD\_IPC\_Regime4 &  3  & 11 \\
-SVD\_IPC\_Regime5 &  23  & 10 \\
-SVD\_IPC\_Regime6 &  30  & 12  \\
-\hline
-\end{tabular}
-\end{center}
-\label{default}
-\end{table}%
-
-\begin{table}[htdp]
-\caption{Average nRMSE over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-                  & Horizon & nRMSE LMA           & nRMSE na\"{i}ve          & $l=5$                   & $l=6$                   \\
-\hline
-gcc               & 4,542   & $0.1407 \pm 0.0063$ & $0.1487 \pm 0.0066$ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-col\_major        & 14,709  & $0.0252 \pm 0.0061$ & $0.1975 \pm 0.0436$ & $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-%SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.8765 \pm 0.0044$ & $0.8508 \pm 0.0035$ \\
-SVD\_IPC\_Regime1 & 2,158   & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ & $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-SVD\_IPC\_Regime2 & 6,902   & $0.1716 \pm 0.0043$ & $0.1762 \pm 0.0012$ & $0.8760 \pm 0.0052$ & $0.8464 \pm 0.0044$ \\
-SVD\_IPC\_Regime3 & 947     & $0.0507 \pm 0.0011$ & $0.5413 \pm 0.0005$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-SVD\_IPC\_Regime4 & 2,929   & $0.1288 \pm 0.0471$ & $0.2308 \pm 0.0867$ & $0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-SVD\_IPC\_Regime5 & 3,255   & $0.0235 \pm 0.0022$ & $0.1306 \pm 0.0003$ & $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-SVD\_IPC\_Regime6 & 5,804   & $0.0196 \pm 0.0022$ & $0.0508 \pm 0.0003$ & $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{default}
-\end{table}%
-
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdipcfull}
-                \caption{SVD IPC without coloring}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdipcregimescolored}
-                \caption{SVD IPC with coloring}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{SVD Full Time Series }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdPrediction}
-                \caption{SVD LMA Prediction}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdPredictionFull}
-                \caption{SVD LMA Prediction with Full Time Series}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{SVD LMA Prediction Figures }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colfullts}
-                \caption{col IPC  Full Time Series}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colshortts}
-                \caption{col IPC  Short Time Series for structure illustration}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{col  Time Series }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colPredShortTS}
-                \caption{col IPC  Full Time Series}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colPredfullTS}
-                \caption{col IPC  Short Time Series for structure illustration}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{col\_major Predictions }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/gccfullts}\caption{403.gcc full time series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/svd_mutuals}\caption{SVD mutuals for full and regimes}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime1}\caption{SVD IPC Regime 1 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime2}\caption{SVD IPC Regime 2 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{figs/regime3}\caption{SVD IPC Regime 3 Time Series}
-%\label{default}
-%\end{center}
-%\end{figure}
-
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{figs/regime4}\caption{SVD IPC Regime 4 Time Series}
-%\label{default}
-%\end{center}
-%\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime5}\caption{SVD IPC Regime 5 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime6}\caption{SVD IPC Regime 6 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.95\textwidth]{figs/SVD_wwpe}
-  \caption{The weighted permutation entropy of one run of SVD. The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    $5,000 \times 100,000$ instructions and the word length is $4$.}
-  \label{fig:wwpe}
-\end{figure}
-
-\section*{Acknowledgment}
-This work was partially supported by NSF grant \#CMMI-1245947 and ARO
-grant \#W911NF-12-1-0288.
-
-\bibliographystyle{unsrt}
-\bibliography{bibliofile}
+\section{Unused Tables}
 

### 2014-02-04 pecomplexity.tex
Cleaning

@@
 %
-\begin{figure}[htbp]
-   \centering
-   \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
-   % where an .eps filename suffix will be assumed under latex,
-   % and a .pdf suffix will be assumed for pdflatex
-   \caption{A small snippet of the L2 cache miss rate of {\tt
-       col\_major}, a three-line C program that repeatedly initializes
-     a matrix in column-major order, running on an Intel Core
-     Duo\textsuperscript{\textregistered}-based machine.  Even this
-     simple program exhibits chaotic performance dynamics.}
-   \label{fig:cache}
- \end{figure}
+% \begin{figure}[htbp]
+%    \centering
+%    \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
+%    % where an .eps filename suffix will be assumed under latex,
+%    % and a .pdf suffix will be assumed for pdflatex
+%    \caption{A small snippet of the L2 cache miss rate of {\tt
+%        col\_major}, a three-line C program that repeatedly initializes
+%      a matrix in column-major order, running on an Intel Core
+%      Duo\textsuperscript{\textregistered}-based machine.  Even this
+%      simple program exhibits chaotic performance dynamics.}
+%    \label{fig:cache}
+%  \end{figure}
 
@@
 procedure, is shown in Figure~\ref{fig:embedding}.
-\begin{figure}
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/3dcolcache}
-    \caption{A 3D projection of a delay-coordinate embedding of the trace
-from Figure~\ref{fig:cache} with a delay ($\tau$) of 100,000 instructions.
-}
-\label{fig:embedding}
-\end{figure}
+
+[[Add a plot of an embedding...]]
+
 %% can cut for space if need be:
@@
 The weighted permutation entropy for the {\tt SVD} program is given in
-Fig.~\cite{fig:wwpe}. To generate this image a window of 5,000 values slid over
+Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
 the time series. Within each of those windows, the statistics over words of
@@
 
+ \section{Results} % (fold)
+ \label{sec:results}
+
+First paragraph: WPE is a good measure of predictability.  Figure:
+best athlete MASE vs. its WPE.
+
 The relationship between prediction accuracy and the weighted permutation
-entropy (WPE) is much as we conjectured: performance traces with high WPE ---
-those whose temporal complexity is high, in the sense that the entropy
-generation per value is approximately equal to the total entropy per value thus
-little information is being propagated forward in time --- are indeed harder to
-predict using the simple deterministic forecast model described in the previous
-section. See Fig.~\cite{fig:pred_vs_ent} for the distribution. The effects of
-changing the word length $n$ are also interesting: using a longer word length
-generally lowers the WPE --- a natural consequence of finite-length data --- but
-the falloff is less rapid in some traces than in others, suggesting that those
-values are closer to the theoretical asymptote that exists for perfect data. The
-persistent WPE values of 0.5--0.6 for the {\tt col\_major} trace are consistent
-with dynamical chaos, further corroborating the results of~\cite{mytkowicz09}.
+entropy (WPE) is much as we conjectured: performance traces with high WPE are
+indeed harder to predict using the forecasting models described in the prior
+section. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
+the exception of one outlier whose behavior we can explain. Furthermore, we find
+that the relationship between the two is roughly linear. The single outlier,
+{\tt SVD} regime 1, does not fit the trend due to a weakness in the WPE: namely
+that in the absence of any large features, as all the other regimes have, the
+WPE effectively falls back to the standard PE for which the noisy behavior
+drives toward 1.0.
+
+The structure being picked up on by WPE may be linear or nonlinear.
+When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
+So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
 
@@
   \includegraphics[width=1.0\textwidth]{figs/prediction_vs_entropy}
-  \caption{The improvement of LMA over na\"{i}ve vs weighted permutation entropy.
-  For each of these, the word length used is $5$.}
+  \caption{The best MASE among all runs and prediction methods vs weighted
+  permutation entropy. For each of these, the word length used is $6$. The
+  dashed line is a least-squares linear fit of all the points except for {\tt
+  SVD$_1$} which we have excluded for reasons explained in the text.}
   \label{fig:pred_vs_ent}
 \end{figure}
+
+Second paragraph: full results.  Image: MASE vs. WPE for both LMA \&
+ARIMA.  Points to make: (1) clusters are distributed differently (2)
+clusters are shaped differently---tight or not (3) clusters move
+differently between LMA and ARIMA (4) finally, the diagonal line is
+important.  If you're below it, you could do better.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1.0\textwidth]{figs/prediction_vs_entropy}
+  \caption{The best MASE among all runs and prediction methods vs weighted
+  permutation entropy. For each of these, the word length used is $6$. The
+  dashed line is a least-squares linear fit of all the points except for {\tt
+  SVD$_1$} which we have excluded for reasons explained in the text.}
+  \label{fig:pred_vs_ent}
+\end{figure}
+
+In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
+compare the performance of the LMA and ARIMA prediction methods (respectively)
+to the value of the weighted permutation entropy for all runs of each program
+under consideration. The LMA MASE values are largely similar to those of the
+best predictions, primarily because LMA often performed superior to ARIMA and
+the naive method. On the other hand, the ARIMA MASE values are largely
+uncorrelated with WPE values. This however does not contradict our hypothesis,
+it just means that ARIMA is ill-suited for the time series under investigation.
+To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since larger
+MASE values constitute worse predictions, ARIMA performed worse than LMA in all
+traces except {\tt SVD$_1$}.
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
+    \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
+    the word length used is $6$.}
+    \label{fig:lma_pred_vs_ent}
+  \end{subfigure}%
+  \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
+    \caption{The MASE of ARIMA vs weighted permutation entropy. For each of
+    these, the word length used is $6$.}
+    \label{fig:arima_pred_vs_ent}
+  \end{subfigure}
+  \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
+    \caption{The MASE values for LMA against ARIMA. The dashed line is the
+    identity, delineating the traces for which either LMA or ARIMA performed
+    better. All traces except those from {\tt SVD$_1$} lie above the line,
+    indicating that LMA is better suited prediction method for the computer
+    traces considered.}
+    \label{fig:lma_vs_arima}
+  \end{subfigure}
+\end{figure}
+
+ % section results (end)
 
@@
 
-The results presented here suggest that permutation entropy---a
-ordinal calculation of forward information transfer in a time
-series---is an effective metric for predictability of computer
-performance traces. Experimentally, traces with a persistent PE
-$\gtrapprox 0.97$ have a natural level of complexity that may
-overshadow the inherent determinism in the system dynamics, whereas
-traces with PE $\lessapprox 0.7$ seem to be highly predictable (viz.,
-at least an order of magnitude improvement in nRMSPE).
-
-If information is the limit, then gathering and using more information
-is an obvious next step.  There is an equally obvious tension here
-between data length and prediction speed: a forecast that requires
-half a second to compute is not useful for the purposes of real-time
-control of a computer system with a MHz clock rate.  Another
-alternative is to sample several system variables simultaneously and
-build multivariate delay-coordinate embeddings.  Existing approaches
-to that are computationally prohibitive
+The results presented here suggest that permutation entropy---a ordinal
+calculation of forward information transfer in a time series---is an effective
+metric for predictability of computer performance traces. Experimentally, traces
+with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
+may overshadow the inherent determinism in the system dynamics, whereas traces
+with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
+of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
+0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
+corroborating the results of~\cite{mytkowicz09}.
+
+If information is the limit, then gathering and using more information is an
+obvious next step.  There is an equally obvious tension here between data length
+and prediction speed: a forecast that requires half a second to compute is not
+useful for the purposes of real-time control of a computer system with a MHz
+clock rate.  Another alternative is to sample several system variables
+simultaneously and build multivariate delay-coordinate embeddings.  Existing
+approaches to that are computationally prohibitive
 \cite{cao-multivariate-embedding}.  We are working on alternative
@@
 
-%****BEYOND THIS POINT IS A DIFFERNT PAPER*** JOSHUA WILL CLEAN THIS OUT, JUST HERE FOR REFERNCE****
-%\section{Introduction: What is Permutation Entropy}
-
-%%\begin{it}
-%%Here I want to talk about the definition and algorithm for computing it.
-Maybe briefly talk about the asymptotic %properties.
-(i.e., same as metric topological in limit.
-Converges to dominate leap if not normalized etc.
-)
-%%\end{it}
-%
-%When analyzing a time series there are several standard characterizations of complexity, including (but not limited to) entropy, Lyapunov exponents and fractal dimensions.
-Rigorous definitions of all these quantities exist if the observation is a \emph{noise-free} trajectory of a \emph{single} dynamical system.
-To an experimental-time-series analyst these are luxuries that are rarely (if ever) realized.
-Many times experimental observation have noise involved such as sensor error, or very commonly in the computer world finite-precision error.
-%
-%Unfortunately the standard algorithms for calculating many of these complexity measures are highly sensitive to noise and give spurious results in the presence of noisy observations.
-Even worse, if a time-series observation measures multiple dynamical systems or dynamical systems that undergo bifurcations\footnote{A drastic change in the behavior of the dynamical system, generally associated with parameter drift.} then all of the standard algorithms for estimating complexity, (e.g., correlation dimension) go out the window.
-Bandt and Pompe first introduced \emph{permutation entropy} as a ``natural complexity measure for time series" in 2002 \cite{bandt2002per}.
-%
-%
-%As explained in \cite{bandt2002per}, permutation entropy is a simple complexity measure that is easily calculated for any time series, be it stochastic, periodic, quasi-periodic, chaotic, noisy or any combination of these.
-Most importantly for analysis of computer performance, permutation entropy as defined in \cite{bandt2002per} yields meaningful results even if observational or dynamical noise is present \cite{bandt2002per}.
-%
-%
-%
-%
-%Analyzing entropies as  a complexity measure is not a new concept, however the way that the entropy is estimated in \cite{bandt2002per} is novel and particularly useful when little is known about the underlying process.
-As discussed in \cite{bandt2002per}, the standard entropy one considers is Shannon entropy.
-%
-%
-%As a quick review of Shannon entropy, we follow the treatment in \cite{amigo2012permutation}.
-Given a finite alphabet $\mathcal{S} = \{s_1,....,s_{|\mathcal{S}|}\}$, define an information source as a discrete-time, stationary stochastic process $X = \{X_n\}_{n \in \mathbb{N}_0}$ where $X_n$ are random variables on a common probability space taking on values from $\mathcal{S}$.
-A realization of $X$ is a one-sided sequence $x_0^{\infty} = (x_n)_{n \in \mathbb{N}_0}$ called a \emph{message}.
-The elements $x_n$ of $\mathcal{S}$ are the \emph{symbols}(a.k.a letters) of the language.
-A finite segment of the message, say $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$ is called a \emph{word} (of length m).
-We then define $p(x_0^{m-1})$ to be the probability of the word $x_0^{m-1}$ to appear some where in the message.
-The shannon entropy of the data source $X$ is defined as: \begin{equation}\label{eq:shannonentropy}
-%h(X) = - \lim_{m\to\infty}\frac{1}{m}\sum p(x_0^{m-1})\log p(x_0^{m-1})
-%\end{equation}
-%
-%For the purposes of this paper once can view entropy as a measure of complexity and predictability in a time series.
-A high entropy almost completely unpredictable and a small entropy is completely predictable.
-This spectrum is outlined very nicely in \cite{amigo2012permutation}: ``Periodic or quasiperiodic sequences have vanishing or negligible complexity.
-At the opposite end, independent and identically distributed random sequences (white noise) have asymptotically divergent permutation entropies, owing to the fact that the number of allowed (or ``admissibleÓ) ordinal patterns grows superexponentially with length.
-Between both ends lie the kind of sequences we are interested in."
-%
-%This illustrates why utilizing entropy as a measure for complexity is, in theory, a great candidate.
-There are several fundamental drawbacks with this approach when analyzing real-world time series.
-The two most fundamental problems are as follows:
-%
-%\begin{enumerate}
-%\item Elements of an experimental time series $x_t \in \mathbb{R}$ which implies $x_t \notin \mathcal{S}$, for a finite alphabet $\mathcal{S}$.
-%\item Time series are always finite in length so the limit in \ref{eq:shannonentropy} never exists in practice.
-%\end{enumerate}
-%
-%
-%The common practice to address item 1 is to perform \emph{symbolization} on the time series you want to analyze.
-Symbolization is the process of defining a dynamic preserving map $\phi: \mathbb{R}^m \to \mathcal{S}$, i.e., for each $x_t$ (or block $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$) we must assign a symbol from $\mathcal{S}$ in such a way that the time series has the same fundamental properties.
-There are several approaches to symbolization, the standard approach is partitioning.
-Assume, $x_t$ has the capability of taking on infinitely many values, defining the space $X$ to be the space where these $x_t$ are drawn.
-We then define a partition on this space $X = P_1 \cup...\cup P_m$, and define $\mathcal{S} = \{1,\dots,m\}$.
-Then $\phi(x_t) = i$ if $x_t \in P_i$.
-We can then calculate the Shannon entropy of the transformed time series, which in the limit of $m$ converges to the Kolmogorov-Sinai entropy \cite{bandt2002per}.
-This approach to symbolization is computationally infeasable as we need an infinite number of partitions of an infinite time series.
-For \emph{extremely} rare cases, there exists a \emph{generating} partition where this limit is not necessary.
-However, finding such a partition even for a simple two-dimensional map like Henon is non-trivial \cite{bandt2002per}.
-Finding a generating partition for an arbitrary time series is for all intensive purposes impossible.
-%
-%[[[Maybe also talk about Mischaikow's approach to this with isolating neighborhoods and homology...but might be off topic. ]]]
-%
-%
-%.....
-%
-%
-%
-%Instead of this approach to symbolization, Bandt et al.
-take the approach that ``the symbol sequence must come naturally for the time series, without any further model assumptions" \cite{bandt2002per}.
-To do this they perform symbolization through ordinal analysis.
-Ordinal analysis of a time series is the process of mapping the ordered successive elements of a time series to a permutation of the same size based on their original time-based ordering.[[Need a better way of wording this]].
-This is also referred to in the literature as the permutation complexity \cite{amigo2012permutation}  For example, if $(x_1,x_2,x_3) = (9,1,7)$ then $\phi$ maps $(x_1,x_2,x_3)$ to the permutation $231$ since $x_2 < x_3<x_1$.
-A great explanation of why this is an interesting symbolization is found in \cite{amigo2012permutation} :
-%
-%\begin{quote}The study of permutation complexity, which we call ordinal analysis, can be envisioned as a new kind of symbolic dynamics whose basic blocks are ordinal patterns.
-Interesting enough, it turns out that under some mild mathematical assumptions, not all ordinal patterns can be materialized by the orbits of a given one- or multi-dimensional deterministic dynamics, not even if this dynamic is chaoticÑ contrarily to what happens with the symbol patterns.
-As a result, the existence of ÒforbiddenÓ (i.e., not occurring) ordinal patterns is always a persistent dynamical feature, in opposition to properties such as proximity and correlation which die out with time in a chaotic dynamic.
-Moreover, if an ordinal pattern is forbidden, its absence pervades all longer patterns in form of more missing ordinal patterns, called outgrowth forbidden patterns.
-Admissible ordinal patterns grow exponentially with length, while forbidden patterns do superexponentially.
-Since random (unconstrained) dynamics has no forbidden patterns with probability 1, their existence can be used as a fingerprint of deterministic orbit generation.
-%\end{quote} Essentially by analyzing the permutation entropy of a time series we are gaining information as to the complexity of the system without any other information available, whether the time series is generated by a random process or the trajectory of a deterministic dynamical system.
-Now that we have justified why the study of permutation entropy is interesting let us define it.
-For this we follow \cite{bandt2002per}.
-%
-%
-%\begin{mydef}[Permutation Entropy]
-%Given a time series $\{x_t\}_{t = 1,\dots,T}$.
-Define $\mathcal{S}_n$ as all $n!$ permutations $\pi$ of order $n$.
-For each $\pi \in \mathcal{S}_n$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$: \begin{equation}\label{eq:permprob}
-%p(\pi) = \frac{|\{t|t\le T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}|}{T-n+1}\end{equation}
-%Where $|\cdot|$ is set cardinality.
-The \emph{permutation entropy} of order $n\ge 2$ is defined as $$H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi)\log_2p(\pi)$$ Notice that $0\le H(n) \le \log_2(n!)$,\cite{bandt2002per}.
-With this in mind, it is common in the literature to present permutation entropy normalized, i.e., $\frac{H(n)}{\log_2(n!)}$.
-With this convention, \emph{low} entropy is close to 0 and \emph{high} entropy is entropy close to 1.
-%\end{mydef}
-%
-%Notes: According to \cite{bandt2002per} equation (\ref{eq:permprob}) estimates the frequency of $\pi$ as good as possible for a finite series of values.
-To determine $p(\pi)$ exactly, we have to assume an infinite time series and take the limit as $T\to \infty$ in (\ref{eq:permprob}).
-Bandt et al., also point out that this limit exists under a very weak stationarity condition: for $k\le n$, the probability for $x_t<x_{t+k}$ should not depend on $t$.
-%
-%It should also be noted that this is discrete-metric (or shannon) permutation entropy, to be distinguished from topological permutation entropy (a.k.a.
-permutation capacity).
-Topological permutation entropy is defined in \cite{amigo2012permutation} as: $$h(\{x_t\}_{t = 1,\dots,\infty}) =  \lim_{L\to\infty}\frac{1}{L}logN(L)$$ where $N(L)$ is the number of distinct ordinal patterns that occur of order $L$.
-This is a measure of how many distinct permutations of length $L$ are realized as opposed to counting the frequency at which each occurs as is the case in metric permutation entropy \cite{amigo2012permutation}.
-It can be shown that metric and topological entropy both converge to the true value with probability 1 in the limit.
-(citation for this).
-When we say permutation entropy in this paper we are referring to that defined in \ref{eq:permprob}.
-%
-%\section{Detecting changes in a time series}
-%
-%Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}.
-To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}.
-Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block.
-(A window shift of 1.)} (or nonoverlapping) blocks.
-Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.
-According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
-%
-%In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy.
-This is identical to the lag used in delay coordinate embedding and often the same language is used.
-Cao et al.
-describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space.
-When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$.
-So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$.
-Cao et al.
-argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1.
-Cao et al.
-then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter.
-The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
-%
-%\subsection{Example: Concatenated Time Series}\label{sec:concatenate}
-%As an example of this detection, consider Figure \ref{fig:ts-banded} (a).
-This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise.
-Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points.
-The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population.
-Each Logistic map trajectory begins at $x=0.6$.
-The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
-%
-%The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$).
-Visually the bands of noise and the chaotic trajectory are indistinguishable.
-If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise.
-If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
-%
-%In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves).
-The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively).
-Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}.
-On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands.
-This is in fact the case when comparing to the way the time series was constructed.
-%
-%The main advantage we can see to the use of lags other than 1 is for analysis of periodicity.
-This is also the major advantage listed in \cite{cao2004det}.
-Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero.
-This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}.
-If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
-%
-%%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
-%
-%\begin{figure}
-%  \centering
-%  \subfigure{\includegraphics[width=\columnwidth]{figs/ts-banded-alone}}
-%    {(a)}
-%
-%  \subfigure{\includegraphics[width=\columnwidth]{figs/ts-banded-pe}}
-%  {(b)}
-%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}.
-(b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves).
-Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
-%\end{figure}
-%
-%
-%
-%
-%
-%
-%and what is it good for with forward pointer to the other sections
-%
-%
-%
-%\subsection{Example: The transient Logistic map}
-%In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed.
-As we showed, permutation was a great proxy to detect this change.
-Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time.
-To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}.
-Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$.
-The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map.
-The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
-%
-% As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well.
-An interesting feature of block calculations are the spikes that occur at bifurcation points.
-As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes.
-This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy.
-The dips you see in the second half correspond to bands of periodicity which occur between the chaos.
-It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range.
-This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
-%
-% This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\columnwidth]{figs/tstranslog-perm}
-%  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
-%  \label{fig:translogistic}
-%\end{figure}
-%%\begin{enumerate}
-%%\item do with henon bifurcation. what happens near the broken chaotic window
-%
-%%\end{enumerate}
-%
-%
-%\section{Possible Applications to Computer Systems Performance}
-%
-%
-%\begin{enumerate}
-%\item In this section I want to discuss the symbolization of performance traces and why this is advantageous over generating partitions and similar approaches.
-%\item No model is assumed, some traces do appear to be stochastic some appear to be deterministic this provides ``proof" thereof. albeit not mathematical proof but CS proof for sure.
-%\item We have essentially infinite data. So limits like $m!$ where $m$ is word length isn't really an issue for us.
-%\end{enumerate}
-%
-%
-%\subsection{Predictability and model change detection}
-%
-%Talk about how some explanatory models may be spinning their wheels because no transferable information.
-That is if you are introducing pure entropy at every time step really no use in using past values and statistically the mean is the most accurate.
-%
-%\begin{enumerate}
-%\item Comp Performance goes under constant bifurcations.
-This allows autonomous detection of when a model is no longer valid.
-(need to be careful here with use of valid as to not be confused with model validation.
-%\item Maybe can tell us if we should use a linear-stochastic, nonlinear-deterministic model or just guess the mean
-%\end{enumerate}
-%
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\columnwidth]{figs/ts-rowcol-pe}
-%  \caption{A time series of cache misses during the execution of the row-col microkernel.
-The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=5,6$ (respectively).
-}\label{fig:rowcol}
-%\end{figure}
-%
-%
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics[width=\columnwidth]{figs/ts-gcc-pe}
-%  \caption{A time series of instructions per cycle for the gcc spec benchmark.
-The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=3,4$ (respectively).
-The lack of structure in such a signal illustrates that the use of deterministic models is most likely not the most advantageous.}\label{fig:gcc}
-%\end{figure}
-%
-%
-%
-%\section{Pro and Cons}
-%\subsection{Advantages of Permutation Entropy}
-%Permutation entropy allows for symbolization of a time series without any further assumptions being made about the underlying system.
-In fact, virtually nothing needs to be known before applying this techniques.
-Other methods of symbolization require expert knowledge of the underlying system, such as generating partitions.
-These techniques have been applied to small data sets, this is truly an advantage especially when dealing with a lot of biological systems whose large datasets can be in the tens of thousands.
-The algorithm for computing permutation entropy is very straight forward and easy to explain.
-The actual code takes longer than I would like to run, but this seems to be an issue with my implementation and not the algorithm, as others state they have exceptionally fast algorithms.
-Although none of these algorithms are ever provided in the literature.
-The major advantages of permutation entropy is that it can quickly give us a rough estimate of where drastic changes in the time series which may have implications in model selection techniques.
-Permutation entropy also can give you a quick idea about if a system has deterministic structure.
-This has direct implications on what prediction strategies to implement.
-%%\begin{enumerate}
-%%\item Allows for symbolization without a model. Really emphasize we can obtain a symbolization without generating partitions which is the traditional approach here.
-%%\item Small amounts of data not really a problem.
-%%\item Maybe talk about how much data is needed per word length for results that mean anything.
-%%\item Fairly easy to calculate, although need a more efficient method.
-%%\item Gives a rough estimation of where drastic changes have occurred in a TS
-%%\item Allows you to see if there is actually something predictable in the signal.
-%%\end{enumerate}
-%
-%\subsection{Disadvantages of PE}
-%There are three disadvantages with permutation entropy we can immediately see that need further thought to be overcome.
-The first is that it is sometimes hard to tell when a change in the dynamical system is not drastic enough to have a huge change on the entropy.
-For example, consider a dynamical system switching from a chaotic to a (different) chaotic regime.
-The permutation entropy may be very similar while the model is slightly different.
-Permutation entropy seems to be ill-suited for fine grain detection of this nature.
-Our current implementation is too slow for on-the-fly analysis.
-We do not believe this is an issue with permutation entropy itself but with our implementation.
-The literature all claims that fast algorithms exist.
-%
-%The biggest disadvantage for us in using permutation entropy is selection of parameters.
-To calculate permutation entropy, we have to choose an embedding dimension, a lag, and a block size.
-In all the literature we have found thus far, parameter selection is made based on persistence or matching the permutation entropy to known values.
-The problem comes when analyzing real time series these parameters seem to have a real impact on the results.
-Finding (or creating) a rigorous method for parameter selection is paramount in the use of this method for the analysis of real world time series.
-%%\begin{enumerate}
-%%\item Paramaters!
-%%\item Currently numerically expensive although the literature disagrees. Think this might just be my implementation. Need to read the numerics section of the PE book.
-%%\item Sometimes hard to tell a regime shift has occurred if the transition is chaotic to chaotic. or low period to low period.
-%%\end{enumerate}
-
 \section{New Figures and Tables}
+
+Possible new error measure
+
+$$q_t = \frac{e_t}{\frac{1}{n-1}\sum^n_{i=2}|Y_i-Y_{i-1}|}$$
+
+$$MASE = mean(|q_t|)$$
+
+Helpful to remember that $F_i = Y_{i-1}$ for random walk prediction
+
+``When $MASE<1$, the proposed method gives, on average, smaller errors than the one-step errors from the na\"ive method.
+If multi-step forecasts are being computed, it is possible to scale by the in- sample MAE[[Mean Absolute Error(mean($|e_t|$)]] computed from multi-step na\"iõve forecasts."
+
+``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series.
+So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method.
+Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one.
+"
+
+We use this error measure with $n$ at the start of the predicted chunk.
 
@@
 \begin{table}[htdp]
-\caption{Average nRMSE over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
+\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
 \begin{center}
@@
 \hline
-                  & Horizon & nRMSE LMA           & nRMSE na\"{i}ve          & $l=5$                   & $l=6$                   \\
-\hline
-gcc               & 4,542   & $0.1407 \pm 0.0063$ & $0.1487 \pm 0.0066$ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-col\_major        & 14,709  & $0.0252 \pm 0.0061$ & $0.1975 \pm 0.0436$ & $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-%SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.8765 \pm 0.0044$ & $0.8508 \pm 0.0035$ \\
-SVD\_IPC\_Regime1 & 2,158   & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ & $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-SVD\_IPC\_Regime2 & 6,902   & $0.1716 \pm 0.0043$ & $0.1762 \pm 0.0012$ & $0.8760 \pm 0.0052$ & $0.8464 \pm 0.0044$ \\
-SVD\_IPC\_Regime3 & 947     & $0.0507 \pm 0.0011$ & $0.5413 \pm 0.0005$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-SVD\_IPC\_Regime4 & 2,929   & $0.1288 \pm 0.0471$ & $0.2308 \pm 0.0867$ & $0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-SVD\_IPC\_Regime5 & 3,255   & $0.0235 \pm 0.0022$ & $0.1306 \pm 0.0003$ & $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-SVD\_IPC\_Regime6 & 5,804   & $0.0196 \pm 0.0022$ & $0.0508 \pm 0.0003$ & $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
+                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
+\hline
+gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+col\_major           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
+
+SVD\_IPC\_Regime1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
+SVD\_IPC\_Regime2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
+SVD\_IPC\_Regime3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
+SVD\_IPC\_Regime4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
+SVD\_IPC\_Regime5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
+SVD\_IPC\_Regime6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
 \hline
@@
 
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdipcfull}
-                \caption{SVD IPC without coloring}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdipcregimescolored}
-                \caption{SVD IPC with coloring}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{SVD Full Time Series }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdPrediction}
-                \caption{SVD LMA Prediction}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdPredictionFull}
-                \caption{SVD LMA Prediction with Full Time Series}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{SVD LMA Prediction Figures }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colfullts}
-                \caption{col IPC  Full Time Series}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colshortts}
-                \caption{col IPC  Short Time Series for structure illustration}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{col  Time Series }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}
-        \centering
-        \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colPredShortTS}
-                \caption{col IPC  Full Time Series}
-                \label{fig:gull}
-        \end{subfigure}%
-        \newline
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colPredfullTS}
-                \caption{col IPC  Short Time Series for structure illustration}
-                \label{fig:svdFullColored}
-        \end{subfigure}
-
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
-          %(or a blank line to force the subfigure onto a new line)
-         \caption{col\_major Predictions }\label{fig:svdFull}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/gccfullts}\caption{403.gcc full time series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/svd_mutuals}\caption{SVD mutuals for full and regimes}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime1}\caption{SVD IPC Regime 1 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime2}\caption{SVD IPC Regime 2 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{figs/regime3}\caption{SVD IPC Regime 3 Time Series}
-%\label{default}
-%\end{center}
-%\end{figure}
-
-%\begin{figure}[htbp]
-%\begin{center}
-%\includegraphics[width=\textwidth]{figs/regime4}\caption{SVD IPC Regime 4 Time Series}
-%\label{default}
-%\end{center}
-%\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime5}\caption{SVD IPC Regime 5 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figs/regime6}\caption{SVD IPC Regime 6 Time Series}
-\label{default}
-\end{center}
-\end{figure}
-
 \section*{Acknowledgment}

### 2014-02-04 unusedFigsTables.tex
Cleaning

@@
 \usepackage{epstopdf}
+%\usepackage{subfigure}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsthm}
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
+\newtheorem*{mydef}{Definition}
                                     % Activate to display a given date or no date
@@
 
+ \begin{figure}
+   \centering
+     \includegraphics[width=0.5\textwidth]{unused-figs/3dcolcache}
+     \caption{A 3D projection of a delay-coordinate embedding of the trace
+ from Figure~\ref{fig:cache} with a delay ($\tau$) of 100,000 instructions.
+ }
+ \label{fig:embedding}
+ \end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdipcfull}
+                \caption{SVD IPC without coloring}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdipcregimescolored}
+                \caption{SVD IPC with coloring}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{SVD Full Time Series }\label{fig:svdFull}
+\end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdregime2prediction}
+                \caption{SVD Regime 2 LMA Prediction}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdregime2predictionFULL}
+                \caption{SVD Regime 2 LMA Prediction with Full Time Series}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{SVD Regime 2 LMA Prediction Figures }\label{fig:svdregime2prediciton}
+\end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdPrediction}
+                \caption{SVD Regime 5 LMA Prediction}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/svdPredictionFull}
+                \caption{SVD Regime 5 LMA Prediction with Full Time Series}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{SVD Regime 5 LMA Prediction Figures }\label{fig:svdregime5prediciton}
+\end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colfullts}
+                \caption{col IPC  Full Time Series}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colshortts}
+                \caption{col IPC  Short Time Series for structure illustration}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{col  Time Series }\label{fig:svdFull}
+\end{figure}
+
+\begin{figure}
+        \centering
+        \begin{subfigure}{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colPredShortTS}
+                \caption{col IPC  Full Time Series}
+                \label{fig:gull}
+        \end{subfigure}%
+        \newline
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{\textwidth}
+                \includegraphics[width=\textwidth]{figs/colPredfullTS}
+                \caption{col IPC  Short Time Series for structure illustration}
+                \label{fig:svdFullColored}
+        \end{subfigure}
+
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc.
+          %(or a blank line to force the subfigure onto a new line)
+         \caption{col\_major Predictions }\label{fig:svdFull}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/gccfullts}\caption{403.gcc full time series}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{unused-figs/svd-mutuals}\caption{SVD mutuals for full and regimes}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime1}\caption{SVD IPC Regime 1 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime2}\caption{SVD IPC Regime 2 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{figs/regime3}\caption{SVD IPC Regime 3 Time Series}
+%\label{default}
+%\end{center}
+%\end{figure}
+
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{figs/regime4}\caption{SVD IPC Regime 4 Time Series}
+%\label{default}
+%\end{center}
+%\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime5}\caption{SVD IPC Regime 5 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figs/regime6}\caption{SVD IPC Regime 6 Time Series}
+\label{default}
+\end{center}
+\end{figure}
+
 \section{Unused Tables}
 
+\begin{table}[htdp]
+\caption{Average nRMSE over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+                   & nRMSE LMA           & nRMSE na\"{i}ve   & $l=5$                   & $l=6$                   \\
+\hline
+gcc                  & $0.1407 \pm 0.0063$ & $0.1487 \pm 0.0066$ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+col\_major          & $0.0252 \pm 0.0061$ & $0.1975 \pm 0.0436$ & $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
+
+SVD\_IPC\_Regime1    & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ & $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
+SVD\_IPC\_Regime2    & $0.1716 \pm 0.0043$ & $0.1762 \pm 0.0012$ & $0.8760 \pm 0.0052$ & $0.8464 \pm 0.0044$ \\
+SVD\_IPC\_Regime3      & $0.0507 \pm 0.0011$ & $0.5413 \pm 0.0005$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
+SVD\_IPC\_Regime4    & $0.1288 \pm 0.0471$ & $0.2308 \pm 0.0867$ & $0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
+SVD\_IPC\_Regime5    & $0.0235 \pm 0.0022$ & $0.1306 \pm 0.0003$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
+SVD\_IPC\_Regime6    & $0.0196 \pm 0.0022$ & $0.0508 \pm 0.0003$ & $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
+
+\begin{table}[htdp]
+\caption{Average nRMSE over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|c|}
+\hline
+                  & Horizon & nRMSE LMA           & nRMSE na\"{i}ve           &LMA MASE (ARIMA) & $l=5$                   & $l=6$                   \\
+\hline
+gcc               & 4,542   & $0.1407 \pm 0.0063$ & $0.1487 \pm 0.0066$ &1.5416(0.929)& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+col\_major        & 14,709  & $0.0252 \pm 0.0061$ & $0.1975 \pm 0.0436$ &0.0519(0.07UF/0.5740/0.9936)& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
+
+SVD\_Full         & 22,072  & $0.0323 \pm 0.0019$ & $0.1784 \pm 0.0040$ & $0.8765 \pm 0.0044$ & $0.8508 \pm 0.0035$ \\
+SVD\_IPC\_Regime1 & 2,158   & $0.1680 \pm 0.0317$ & $0.3517 \pm 0.5223$ &0.7954(0.724UF)& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
+SVD\_IPC\_Regime2 & 6,902   & $0.1716 \pm 0.0043$ & $0.1762 \pm 0.0012$ &1.9990(1.138)             & $0.8760 \pm 0.0052$ & $0.8464 \pm 0.0044$ \\
+SVD\_IPC\_Regime3 & 947     & $0.0507 \pm 0.0011$ & $0.5413 \pm 0.0005$ &1.0089 (1.165215)              & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
+SVD\_IPC\_Regime4 & 2,929   & $0.1288 \pm 0.0471$ & $0.2308 \pm 0.0867$ &0.8976 (0.840548)           & $0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
+SVD\_IPC\_Regime5 & 3,255   & $0.0235 \pm 0.0022$ & $0.1306 \pm 0.0003$ &0.7487 (1.260/2.3446)& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
+SVD\_IPC\_Regime6 & 5,804   & $0.0196 \pm 0.0022$ & $0.0508 \pm 0.0003$ &0.7848 (0.7831) & $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
+
+\begin{table}[htdp]
+\caption{Average MASE over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
+\hline
+gcc                  & $1.9856 \pm $ & $1.7525 \pm $ &1.7904 $\pm$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+col\_major           & $ 0.0587\pm $ & $0.9936 \pm $ & $0.5713\pm$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
+
+SVD\_IPC\_Regime1     & $ \pm $ & $ \pm $ & $\pm$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
+SVD\_IPC\_Regime2     & $ 2.3753\pm $ & $0.7268 \pm $ & $3.1064\pm$& $0.8760 \pm 0.0052$ & $0.8464 \pm 0.0044$ \\
+SVD\_IPC\_Regime3       & $ \pm $ & $ \pm $ & $\pm$& $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
+SVD\_IPC\_Regime4     & $ \pm $ & $ \pm $ & $\pm$&$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
+SVD\_IPC\_Regime5     & $4.9079 \pm $ & $2.3366 \pm $ & $21.0372\pm$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
+SVD\_IPC\_Regime6     & $ \pm $ & $ \pm $ & $\pm$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
+
+\section{Unused Text}
+%****BEYOND THIS POINT IS A DIFFERNT PAPER*** JOSHUA WILL CLEAN THIS OUT, JUST HERE FOR REFERNCE****
+\section{Introduction: What is Permutation Entropy}
+
+%\begin{it}
+%Here I want to talk about the definition and algorithm for computing it.
+Maybe briefly talk about the asymptotic %properties.
+(i.e., same as metric topological in limit.
+Converges to dominate leap if not normalized etc.
+)
+%\end{it}
+
+When analyzing a time series there are several standard characterizations of complexity, including (but not limited to) entropy, Lyapunov exponents and fractal dimensions.
+Rigorous definitions of all these quantities exist if the observation is a \emph{noise-free} trajectory of a \emph{single} dynamical system.
+To an experimental-time-series analyst these are luxuries that are rarely (if ever) realized.
+Many times experimental observation have noise involved such as sensor error, or very commonly in the computer world finite-precision error.
+
+Unfortunately the standard algorithms for calculating many of these complexity measures are highly sensitive to noise and give spurious results in the presence of noisy observations.
+Even worse, if a time-series observation measures multiple dynamical systems or dynamical systems that undergo bifurcations\footnote{A drastic change in the behavior of the dynamical system, generally associated with parameter drift.} then all of the standard algorithms for estimating complexity, (e.g., correlation dimension) go out the window.
+Bandt and Pompe first introduced \emph{permutation entropy} as a ``natural complexity measure for time series" in 2002 \cite{bandt2002per}.
+
+As explained in \cite{bandt2002per}, permutation entropy is a simple complexity measure that is easily calculated for any time series, be it stochastic, periodic, quasi-periodic, chaotic, noisy or any combination of these.
+Most importantly for analysis of computer performance, permutation entropy as defined in \cite{bandt2002per} yields meaningful results even if observational or dynamical noise is present \cite{bandt2002per}.
+
+Analyzing entropies as  a complexity measure is not a new concept, however the way that the entropy is estimated in \cite{bandt2002per} is novel and particularly useful when little is known about the underlying process.
+As discussed in \cite{bandt2002per}, the standard entropy one considers is Shannon entropy.
+
+As a quick review of Shannon entropy, we follow the treatment in \cite{amigo2012permutation}.
+Given a finite alphabet $\mathcal{S} = \{s_1,....,s_{|\mathcal{S}|}\}$, define an information source as a discrete-time, stationary stochastic process $X = \{X_n\}_{n \in \mathbb{N}_0}$ where $X_n$ are random variables on a common probability space taking on values from $\mathcal{S}$.
+A realization of $X$ is a one-sided sequence $x_0^{\infty} = (x_n)_{n \in \mathbb{N}_0}$ called a \emph{message}.
+The elements $x_n$ of $\mathcal{S}$ are the \emph{symbols}(a.k.a letters) of the language.
+A finite segment of the message, say $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$ is called a \emph{word} (of length m).
+We then define $p(x_0^{m-1})$ to be the probability of the word $x_0^{m-1}$ to appear some where in the message.
+The shannon entropy of the data source $X$ is defined as: \begin{equation}\label{eq:shannonentropy}
+h(X) = - \lim_{m\to\infty}\frac{1}{m}\sum p(x_0^{m-1})\log p(x_0^{m-1})
+\end{equation}
+
+For the purposes of this paper once can view entropy as a measure of complexity and predictability in a time series.
+A high entropy almost completely unpredictable and a small entropy is completely predictable.
+This spectrum is outlined very nicely in \cite{amigo2012permutation}: ``Periodic or quasiperiodic sequences have vanishing or negligible complexity.
+At the opposite end, independent and identically distributed random sequences (white noise) have asymptotically divergent permutation entropies, owing to the fact that the number of allowed (or ``admissibleÌ?) ordinal patterns grows superexponentially with length.
+Between both ends lie the kind of sequences we are interested in."
+
+This illustrates why utilizing entropy as a measure for complexity is, in theory, a great candidate.
+There are several fundamental drawbacks with this approach when analyzing real-world time series.
+The two most fundamental problems are as follows:
+
+\begin{enumerate}
+\item Elements of an experimental time series $x_t \in \mathbb{R}$ which implies $x_t \notin \mathcal{S}$, for a finite alphabet $\mathcal{S}$.
+\item Time series are always finite in length so the limit in \ref{eq:shannonentropy} never exists in practice.
+\end{enumerate}
+
+The common practice to address item 1 is to perform \emph{symbolization} on the time series you want to analyze.
+Symbolization is the process of defining a dynamic preserving map $\phi: \mathbb{R}^m \to \mathcal{S}$, i.e., for each $x_t$ (or block $x_k^{k+m-1} = x_kx_{k+1} \dots x_{k+m-1}$) we must assign a symbol from $\mathcal{S}$ in such a way that the time series has the same fundamental properties.
+There are several approaches to symbolization, the standard approach is partitioning.
+Assume, $x_t$ has the capability of taking on infinitely many values, defining the space $X$ to be the space where these $x_t$ are drawn.
+We then define a partition on this space $X = P_1 \cup...\cup P_m$, and define $\mathcal{S} = \{1,\dots,m\}$.
+Then $\phi(x_t) = i$ if $x_t \in P_i$.
+We can then calculate the Shannon entropy of the transformed time series, which in the limit of $m$ converges to the Kolmogorov-Sinai entropy \cite{bandt2002per}.
+This approach to symbolization is computationally infeasable as we need an infinite number of partitions of an infinite time series.
+For \emph{extremely} rare cases, there exists a \emph{generating} partition where this limit is not necessary.
+However, finding such a partition even for a simple two-dimensional map like Henon is non-trivial \cite{bandt2002per}.
+Finding a generating partition for an arbitrary time series is for all intensive purposes impossible.
+
+[[[Maybe also talk about Mischaikow's approach to this with isolating neighborhoods and homology...but might be off topic. ]]]
+
+.....
+
+Instead of this approach to symbolization, Bandt et al.
+take the approach that ``the symbol sequence must come naturally for the time series, without any further model assumptions" \cite{bandt2002per}.
+To do this they perform symbolization through ordinal analysis.
+Ordinal analysis of a time series is the process of mapping the ordered successive elements of a time series to a permutation of the same size based on their original time-based ordering.[[Need a better way of wording this]].
+This is also referred to in the literature as the permutation complexity \cite{amigo2012permutation}  For example, if $(x_1,x_2,x_3) = (9,1,7)$ then $\phi$ maps $(x_1,x_2,x_3)$ to the permutation $231$ since $x_2 < x_3<x_1$.
+A great explanation of why this is an interesting symbolization is found in \cite{amigo2012permutation} :
+
+\begin{quote}The study of permutation complexity, which we call ordinal analysis, can be envisioned as a new kind of symbolic dynamics whose basic blocks are ordinal patterns.
+Interesting enough, it turns out that under some mild mathematical assumptions, not all ordinal patterns can be materialized by the orbits of a given one- or multi-dimensional deterministic dynamics, not even if this dynamic is chaoticÌ?
+contrarily to what happens with the symbol patterns.
+As a result, the existence of Ì?forbiddenÌ?
+(i.e., not occurring) ordinal patterns is always a persistent dynamical feature, in opposition to properties such as proximity and correlation which die out with time in a chaotic dynamic.
+Moreover, if an ordinal pattern is forbidden, its absence pervades all longer patterns in form of more missing ordinal patterns, called outgrowth forbidden patterns.
+Admissible ordinal patterns grow exponentially with length, while forbidden patterns do superexponentially.
+Since random (unconstrained) dynamics has no forbidden patterns with probability 1, their existence can be used as a fingerprint of deterministic orbit generation.
+\end{quote} Essentially by analyzing the permutation entropy of a time series we are gaining information as to the complexity of the system without any other information available, whether the time series is generated by a random process or the trajectory of a deterministic dynamical system.
+Now that we have justified why the study of permutation entropy is interesting let us define it.
+For this we follow \cite{bandt2002per}.
+
+\begin{mydef}[Permutation Entropy]
+Given a time series $\{x_t\}_{t = 1,\dots,T}$.
+Define $\mathcal{S}_n$ as all $n!$ permutations $\pi$ of order $n$.
+For each $\pi \in \mathcal{S}_n$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$: \begin{equation}\label{eq:permprob}
+p(\pi) = \frac{|\{t|t\le T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}|}{T-n+1}\end{equation}
+Where $|\cdot|$ is set cardinality.
+The \emph{permutation entropy} of order $n\ge 2$ is defined as $$H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi)\log_2p(\pi)$$ Notice that $0\le H(n) \le \log_2(n!)$,\cite{bandt2002per}.
+With this in mind, it is common in the literature to present permutation entropy normalized, i.e., $\frac{H(n)}{\log_2(n!)}$.
+With this convention, \emph{low} entropy is close to 0 and \emph{high} entropy is entropy close to 1.
+\end{mydef}
+
+Notes: According to \cite{bandt2002per} equation (\ref{eq:permprob}) estimates the frequency of $\pi$ as good as possible for a finite series of values.
+To determine $p(\pi)$ exactly, we have to assume an infinite time series and take the limit as $T\to \infty$ in (\ref{eq:permprob}).
+Bandt et al., also point out that this limit exists under a very weak stationarity condition: for $k\le n$, the probability for $x_t<x_{t+k}$ should not depend on $t$.
+
+It should also be noted that this is discrete-metric (or shannon) permutation entropy, to be distinguished from topological permutation entropy (a.k.a.
+permutation capacity).
+Topological permutation entropy is defined in \cite{amigo2012permutation} as: $$h(\{x_t\}_{t = 1,\dots,\infty}) =  \lim_{L\to\infty}\frac{1}{L}logN(L)$$ where $N(L)$ is the number of distinct ordinal patterns that occur of order $L$.
+This is a measure of how many distinct permutations of length $L$ are realized as opposed to counting the frequency at which each occurs as is the case in metric permutation entropy \cite{amigo2012permutation}.
+It can be shown that metric and topological entropy both converge to the true value with probability 1 in the limit.
+(citation for this).
+When we say permutation entropy in this paper we are referring to that defined in \ref{eq:permprob}.
+
+\section{Detecting changes in a time series}
+
+Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}.
+To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}.
+Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block.
+(A window shift of 1.)} (or nonoverlapping) blocks.
+Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.
+According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
+
+In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy.
+This is identical to the lag used in delay coordinate embedding and often the same language is used.
+Cao et al.
+describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space.
+When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$.
+So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$.
+Cao et al.
+argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1.
+Cao et al.
+then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter.
+The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
+
+\subsection{Example: Concatenated Time Series}\label{sec:concatenate}
+As an example of this detection, consider Figure \ref{fig:ts-banded} (a).
+This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise.
+Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points.
+The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population.
+Each Logistic map trajectory begins at $x=0.6$.
+The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
+
+The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$).
+Visually the bands of noise and the chaotic trajectory are indistinguishable.
+If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise.
+If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
+
+In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves).
+The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively).
+Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}.
+On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands.
+This is in fact the case when comparing to the way the time series was constructed.
+
+The main advantage we can see to the use of lags other than 1 is for analysis of periodicity.
+This is also the major advantage listed in \cite{cao2004det}.
+Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero.
+This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}.
+If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
+
+%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
+
+%\begin{figure}
+%  \centering
+%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-alone}}
+%    {(a)}
+%
+%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-pe}}
+%  {(b)}
+%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}.
+(b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves).
+Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
+%\end{figure}
+
+and what is it good for with forward pointer to the other sections
+
+\subsection{Example: The transient Logistic map}
+In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed.
+As we showed, permutation was a great proxy to detect this change.
+Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time.
+To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}.
+Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$.
+The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map.
+The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
+
+ As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well.
+An interesting feature of block calculations are the spikes that occur at bifurcation points.
+As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes.
+This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy.
+The dips you see in the second half correspond to bands of periodicity which occur between the chaos.
+It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range.
+This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
+
+ This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\columnwidth]{unused-figs/tstranslog-perm}
+  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
+  \label{fig:translogistic}
+\end{figure}
+%\begin{enumerate}
+%\item do with henon bifurcation. what happens near the broken chaotic window
+
+%\end{enumerate}
+
+\section{Possible Applications to Computer Systems Performance}
+
+\begin{enumerate}
+\item In this section I want to discuss the symbolization of performance traces and why this is advantageous over generating partitions and similar approaches.
+\item No model is assumed, some traces do appear to be stochastic some appear to be deterministic this provides ``proof" thereof. albeit not mathematical proof but CS proof for sure.
+\item We have essentially infinite data. So limits like $m!$ where $m$ is word length isn't really an issue for us.
+\end{enumerate}
+
+\subsection{Predictability and model change detection}
+
+Talk about how some explanatory models may be spinning their wheels because no transferable information.
+That is if you are introducing pure entropy at every time step really no use in using past values and statistically the mean is the most accurate.
+
+\begin{enumerate}
+\item Comp Performance goes under constant bifurcations.
+This allows autonomous detection of when a model is no longer valid.
+(need to be careful here with use of valid as to not be confused with model validation.
+\item Maybe can tell us if we should use a linear-stochastic, nonlinear-deterministic model or just guess the mean
+\end{enumerate}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\columnwidth]{unused-figs/ts-rowcol-pe}
+  \caption{A time series of cache misses during the execution of the row-col microkernel.
+The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=5,6$ (respectively).
+}\label{fig:rowcol}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\columnwidth]{unused-figs/ts-gcc-pe}
+  \caption{A time series of instructions per cycle for the gcc spec benchmark.
+The red and blue curves are permutation entropy calculated using $\tau =1$ and $m=3,4$ (respectively).
+The lack of structure in such a signal illustrates that the use of deterministic models is most likely not the most advantageous.}\label{fig:gcc}
+\end{figure}
+
+\section{Pro and Cons}
+\subsection{Advantages of Permutation Entropy}
+Permutation entropy allows for symbolization of a time series without any further assumptions being made about the underlying system.
+In fact, virtually nothing needs to be known before applying this techniques.
+Other methods of symbolization require expert knowledge of the underlying system, such as generating partitions.
+These techniques have been applied to small data sets, this is truly an advantage especially when dealing with a lot of biological systems whose large datasets can be in the tens of thousands.
+The algorithm for computing permutation entropy is very straight forward and easy to explain.
+The actual code takes longer than I would like to run, but this seems to be an issue with my implementation and not the algorithm, as others state they have exceptionally fast algorithms.
+Although none of these algorithms are ever provided in the literature.
+The major advantages of permutation entropy is that it can quickly give us a rough estimate of where drastic changes in the time series which may have implications in model selection techniques.
+Permutation entropy also can give you a quick idea about if a system has deterministic structure.
+This has direct implications on what prediction strategies to implement.
+%\begin{enumerate}
+%\item Allows for symbolization without a model. Really emphasize we can obtain a symbolization without generating partitions which is the traditional approach here.
+%\item Small amounts of data not really a problem.
+%\item Maybe talk about how much data is needed per word length for results that mean anything.
+%\item Fairly easy to calculate, although need a more efficient method.
+%\item Gives a rough estimation of where drastic changes have occurred in a TS
+%\item Allows you to see if there is actually something predictable in the signal.
+%\end{enumerate}
+
+\subsection{Disadvantages of PE}
+There are three disadvantages with permutation entropy we can immediately see that need further thought to be overcome.
+The first is that it is sometimes hard to tell when a change in the dynamical system is not drastic enough to have a huge change on the entropy.
+For example, consider a dynamical system switching from a chaotic to a (different) chaotic regime.
+The permutation entropy may be very similar while the model is slightly different.
+Permutation entropy seems to be ill-suited for fine grain detection of this nature.
+Our current implementation is too slow for on-the-fly analysis.
+We do not believe this is an issue with permutation entropy itself but with our implementation.
+The literature all claims that fast algorithms exist.
+
+The biggest disadvantage for us in using permutation entropy is selection of parameters.
+To calculate permutation entropy, we have to choose an embedding dimension, a lag, and a block size.
+In all the literature we have found thus far, parameter selection is made based on persistence or matching the permutation entropy to known values.
+The problem comes when analyzing real time series these parameters seem to have a real impact on the results.
+Finding (or creating) a rigorous method for parameter selection is paramount in the use of this method for the analysis of real world time series.
+%\begin{enumerate}
+%\item Paramaters!
+%\item Currently numerically expensive although the literature disagrees. Think this might just be my implementation. Need to read the numerics section of the PE book.
+%\item Sometimes hard to tell a regime shift has occurred if the transition is chaotic to chaotic. or low period to low period.
+%\end{enumerate}
+
 \end{document}

### 2014-02-07 pecomplexity.tex
working on the results/bib/and figures

@@
 \usepackage{amsthm}
+\newcommand{\gcc}{{\tt 403.gcc}~}
+\newcommand{\svd}{{\tt dgesdd}~}
+\newcommand{\row}{{\tt row\_major}~}
+\newcommand{\col}{{\tt col\_major}~}
+
 \usepackage{authblk}
@@
 
-  \begin{table}[htbp]
-   % increase table row spacing, adjust to taste
-   \renewcommand{\arraystretch}{1.3}
-   \caption{Normalized root mean squared error (nRMSE) of 4000-point
-     predictions of memory \& processor performance from different
-     programs.}
-   \label{tab:PredError}
-   \centering
-   % Some packages, such as MDW tools, offer better commands for making tables
-   % than the plain LaTeX2e tabular which is used here.
-%   \begin{tabular}{|c|c|c|c|c|c|}
-%     \hline
-%      & cache misses NRMSE & IPC NRMSE \\
-%     \hline
-%     \tt{row\_major}  & 0.0102  & 0.0095 \\
-%          \hline
-%     \tt{col\_major}  &0.0055& 0.0071 \\
-%     \hline
-%     \tt{403.gcc}  & 0.0865& 0.1805 \\
-%          \hline
-%     \tt{482.sphinx3} & 0.1142& 0.2946 \\
-%      \hline
-%   \end{tabular}
-% \end{table}
-   \begin{tabular}{|c|c|c|c|c|c|}
-     \hline
-      & cache miss rate & instrs per cycle \\
-     \hline
-     \tt{row\_major}  & 0.0324  & 0.0778 \\
-          \hline
-     \tt{col\_major}  &0.0080& 0.0161 \\
-     \hline
-     \tt{403.gcc}  & 0.1416& 0.2033 \\
-          \hline
-     \tt{482.sphinx3} & 0.2032& 0.3670 \\
-      \hline
-   \end{tabular}
- \end{table}
-
 The results in Table~\ref{tab:PredError} show a clear distinction
@@
 regimes.
+[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series differnet levels of complexity and predictability exist.
+Maybe point at some of the predicting predictability papers.]]
 
@@
   \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-  \caption{The weighted permutation entropy of one run of SVD. The gray bands
+  \caption{[[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure]]The weighted permutation entropy of one run of SVD. The gray bands
     are regions where the window overlaps regimes. The window size used is
@@
 
-% Table~\ref{tab:pe} shows the permutation entropy results for the
-% examples considered in this paper, with the nRMSPE prediction
-% accuracies from the previous section included alongside for easy
-% comparison.
-%
-% \begin{table}[htbp]
-%   % increase table row spacing, adjust to taste
-%   \renewcommand{\arraystretch}{1.3}
-%   \caption{TCM NRMSE and PE NORMS}
-%   \label{tab:table_example}
-%   \centering
-   % Some packages, such as MDW tools, offer better commands for making tables
-   % than the plain LaTeX2e tabular which is used here.
-%   \begin{tabular}{|c|c|c|c|c|c|}
-%     \hline
-%     TCM & NRMSE & $m=3$  &$m=4$&$m=5$&$m=6$\\
-%     \hline
-%     \tt{row\_major} & 0.0102 & 0.7900  &0.6751&0.5458&0.4491\\
-%     \hline
-%     \tt{col\_major} & 0.0055 & 0.6411  &0.5029&0.4515&0.3955 \\
-%     \hline
-%     \tt{403.gcc} & 0.0865 & 0.9954  &0.9916&0.9880&0.9835\\
-%     \hline
-%     \tt{482.sphinx3} & 0.1142 & 0.9958  &0.9913&0.9866&0.9802 \\
-%      \hline
-%   \end{tabular}
-% \end{table}
-%
-%%  \begin{table}[htbp]
-%%   % increase table row spacing, adjust to taste
-%%   %\renewcommand{\arraystretch}{1.3}
-%%   \caption{Cache-miss rate performance traces: prediction error
-%%     (nRMSPE) and permutation entropy}
-%%   \label{tab:TCMpe}
-%%   \centering
-%%   % Some packages, such as MDW tools, offer better commands for making tables
-%%   % than the plain LaTeX2e tabular which is used here.
-%%   \begin{tabular}{|c|c|c|c|c|c|}
-%%     \hline
-%%      & nRMSE   &$n=4$&$n=5$&$n=6$\\
-%%     \hline
-%%     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
-%%     \hline
-%%     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
-%%     \hline
-%%     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
-%%     \hline
-%%     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
-%%      \hline
-%%   \end{tabular}
-%% \end{table}
-%
-% \begin{table}[htbp]
-%   % increase table row spacing, adjust to taste
-%   \renewcommand{\arraystretch}{1.3}
-%   \caption{IPC NRMSE and PE NORMS}
-%   \label{tab:table_example}
-%   \centering
-%   % Some packages, such as MDW tools, offer better commands for making tables
-%   % than the plain LaTeX2e tabular which is used here.
-%   \begin{tabular}{|c|c|c|c|c|c|}
-%     \hline
-%     ipc & NRMSE & $n=3$  &$n=4$&$n=5$&$n=6$\\
-%     \hline
-%      \tt{row\_major} & 0.0095 & 0.9912  &0.9723&0.9354&0.8876\\
-%     \hline
-%      \tt{col\_major} & 0.0071 & 0.9428  &0.8356&0.7601&0.6880\\
-%     \hline
-%     \tt{403.gcc} & 0.1805 & 0.9918  &0.9862&0.9814&0.9764\\
-%     \hline
-%     \tt{482.sphinx3} & 0.2946 & 0.9978  &0.9951&0.9914&0.9849 \\
-%      \hline
-%   \end{tabular}
-% \end{table}
-%
-%%  \begin{table}[htbp]
-%%   % increase table row spacing, adjust to taste
-%%   %\renewcommand{\arraystretch}{1.1}
-%%   \caption{Processor load performance traces: prediction error
-%%     (nRMSPE) and permutation entropy}
-%%   \label{tab:IPCpe}
-%%   \centering
-%%   % Some packages, such as MDW tools, offer better commands for making tables
-%%   % than the plain LaTeX2e tabular which is used here.
-%%   \begin{tabular}{|c|c|c|c|c|}
-%%     \hline
-%%      & nRMSPE   &$n=4$&$n=5$&$n=6$\\
-%%     \hline
-%%      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
-%%     \hline
-%%      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
-%%     \hline
-%%     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
-%%     \hline
-%%     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
-%%      \hline
-%%   \end{tabular}
-%% \end{table}
-%
- %  \begin{table}[htbp]
- %   \caption{Prediction error (in nRMSPE) and permutation entropy (for
- %     different wordlengths $n$}
- %   \label{tab:pe}
- %   \centering
- %   \begin{tabular}{|c|c|c|c|c|c|}
- %     \hline
- %      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
- %     \hline
- %     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
- %     \hline
- %     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
- %     \hline
- %     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
- %     \hline
- %     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
- %      \hline
- %      \hline
- %      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
- %     \hline
- %      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
- %     \hline
- %      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
- %     \hline
- %     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
- %     \hline
- %     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
- %      \hline
- %   \end{tabular}
- % \end{table}
-
  \section{Results} % (fold)
@@
 
-First paragraph: WPE is a good measure of predictability.  Figure:
+%\item Introduce MASE and explain why random walk isn't *that* naive
+
+%We calculated a figure of merit for each prediction method as follows.
+
+%As a figure of merit we calculate the Mean Absolute Squared Error (MASE) for each forecast.
+
+In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% refereed to as the ``learning/training" signal, $\{X_{i,obs}\}_{i=1}^{i=n}$ and the last 10\% known as the ``test/correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
+The learning signal is used to build the initial models (e.g., LMA,ARIMA,na\"ive) as described in Section \ref{sec:compModel}.
+The test signal is used both to test the forecasts and to refit the models after forecasting.
+In particular, using the models we perform $k$ 1-step predictions $\{p_j\}_{j=n+1}^{k+n+1}$, after each prediction we append the training signal with the the next point in the correct signal.
+We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.
+
+In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
+
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
+
+The scaling term here:
+
+$$s=\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|$$ 
+is the average in-sample forecast error for a random walk prediction $(p_j=c_{j-1})$.
+This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
+A $MASE<1$ means that the prediction method gives, on average, smaller errors than the 1-step errors from the random walk forecast strategy.
+Analogously, a $MASE>1$ means that the prediction method did worse, on average than the 1-step errors from the random walk forecast strategy.
+In Table \ref{tab:error} we provide the distribution of MASEs for 15 runs of each of the 8 signals across the 3 prediction forecast methods.
+For comparison Table \ref{tab:error} also has weighted permutation
+
+%``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series.
+So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method.
+Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one.
+"
+
+\begin{table}[htdp]
+\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
+\hline
+\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
+
+\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
+\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
+\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
+\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
+\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
+\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
+\hline
+\end{tabular}
+\end{center}
+\label{tab:error}
+\end{table}%
+
+%This is then used as a scaling term in the following way:
+
+%$$qt = \frac{e_t}{\frac{1}{n-1}\}sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$
+
+%$$MASE = \frac{\frac{\sum_{j=n+1}^{k+n+1} |c_j-p_j| }{k}}{\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|}$$
+
+%and is comprised of the first 90\% of the time series, the learning signal is what we build the initial forecast models with.
+The second piece is called the test  the  back the last 10\% of each signal.
+Built the initial forecast model for each signal.
+We then perform an in-sample random walk forecast at every point in the training
+
+%Possible new error measure
+
+%$$MASE = mean(|q_t|)$$
+
+%Helpful to remember that $F_i = Y_{i-1}$ for random walk prediction
+
+%``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series.
+So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method.
+Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one.
+"
+
+%\item Paragraph on how to interpret MASE
+\begin{enumerate}
+
+\item First paragraph: WPE is a good measure of predictability.  Figure:
 best athlete MASE vs. its WPE.
 
-The relationship between prediction accuracy and the weighted permutation
-entropy (WPE) is much as we conjectured: performance traces with high WPE are
-indeed harder to predict using the forecasting models described in the prior
-section. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
-the exception of one outlier whose behavior we can explain. Furthermore, we find
-that the relationship between the two is roughly linear. The single outlier,
-{\tt SVD} regime 1, does not fit the trend due to a weakness in the WPE: namely
-that in the absence of any large features, as all the other regimes have, the
-WPE effectively falls back to the standard PE for which the noisy behavior
-drives toward 1.0.
-
-The structure being picked up on by WPE may be linear or nonlinear.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
-So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
-
+As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
+indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
+the exception of one outlier whose behavior we can explain. In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over methods on that signal) for each of the 8 signals.  
 \begin{figure}[htbp]
   \centering
-  \includegraphics[width=1.0\textwidth]{figs/prediction_vs_entropy}
+  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
   \caption{The best MASE among all runs and prediction methods vs weighted
@@
 \end{figure}
+We find that the relationship between the two is roughly linear. The single outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE: namely
+that in the absence of any large features, as all the other regimes have, the
+WPE effectively falls back to the standard PE for which the noisy behavior
+drives toward 1.0.
 
-Second paragraph: full results.  Image: MASE vs. WPE for both LMA \&
-ARIMA.  Points to make: (1) clusters are distributed differently (2)
-clusters are shaped differently---tight or not (3) clusters move
-differently between LMA and ARIMA (4) finally, the diagonal line is
-important.  If you're below it, you could do better.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=1.0\textwidth]{figs/prediction_vs_entropy}
-  \caption{The best MASE among all runs and prediction methods vs weighted
-  permutation entropy. For each of these, the word length used is $6$. The
-  dashed line is a least-squares linear fit of all the points except for {\tt
-  SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_ent}
-\end{figure}
-
-In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
-compare the performance of the LMA and ARIMA prediction methods (respectively)
-to the value of the weighted permutation entropy for all runs of each program
-under consideration. The LMA MASE values are largely similar to those of the
-best predictions, primarily because LMA often performed superior to ARIMA and
-the naive method. On the other hand, the ARIMA MASE values are largely
-uncorrelated with WPE values. This however does not contradict our hypothesis,
-it just means that ARIMA is ill-suited for the time series under investigation.
-To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since larger
-MASE values constitute worse predictions, ARIMA performed worse than LMA in all
-traces except {\tt SVD$_1$}.
-
+This primary finding: ``Signals that are more complex are harder to predict." At first glance this finding may seem incredibly trivial, of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core result here as we will elucidate now with Figure \ref{fig:lma_vs_arima}.
 \begin{figure}[htbp]
@@
   \end{subfigure}
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
+  %\begin{subfigure}{0.5\textwidth}
+  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
     \caption{The MASE values for LMA against ARIMA. The dashed line is the
@@
     indicating that LMA is better suited prediction method for the computer
-    traces considered.}
-    \label{fig:lma_vs_arima}
-  \end{subfigure}
+   traces considered.}\label{fig:lma_vs_arima}
+
+  %\end{subfigure}
 \end{figure}
+
+\item Talk about structure analysis. Just because there is forward infromation transfer, does not mean that linear predictors can get at this. 
+\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by side plot. 
+\item  full results.  Image: MASE vs. WPE for both LMA \&
+ARIMA.  Points to make: 
+\begin{enumerate}
+\item clusters are distributed differently
+\item clusters are shaped differently---tight or not 
+\item clusters move
+differently between LMA and ARIMA 
+\item finally, the diagonal line is
+important.  If you're below it, you could do better.
+
+\end{enumerate}
+\end{enumerate}
+
+In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
+compare the performance of the LMA and ARIMA prediction methods (respectively)
+to the value of the weighted permutation entropy for all runs of each program
+under consideration. The LMA MASE values are largely similar to those of the
+best predictions, primarily because LMA often performed superior to ARIMA and
+the naive method. On the other hand, the ARIMA MASE values are largely
+uncorrelated with WPE values.
+The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work.
+More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as is proposed as a good strategy in \cite{ARIMA-autofit} and say you got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MASE>1$).
+How do we know if the prediction strategy is not adequte for the prediction task or if the signal is simply too complex to predict.
+If a signal is too complex and little forward information transfer is present we may not be able to do better than the random walk, in which case we should not kill ourselves over finding a more complicated prediction strategy.
+However, if we measure the complexity to be low, $WPE<0.85$ we can most likely do much better than the random walk and should search for more adequate prediction strategies.
+For example, consider \col (the blue +s in Figure \ref{fig:lma_vs_arima}).
+If we use the out-of-the-box ARIMA from \cite{ARIMA-autofit} we get MASE ranging from 0.5 all the way to 1.0.
+That is we may think that the best we can do is twice as good as random-walk but sometimes we only do as well as random walk.
+In which case the random walk prediction strategy is probably your best bet as it is so simple and produces very similar error.
+However, if we calculate WPE we see that \col has a WPE of $0.5131\pm 0.0034$.
+This incredibly low entropy implies a great deal of deterministic structure that we can utilize for prediction, even though this did not appear to be the case with ARIMA.
+In fact using LMA on this signal we get an average MASE of $0.05\pm0.0018$, an error that is on-average 20 times better than the random walk forecast.
+Alternatively, we can look at the other end of the spectrum at \gcc.
+This signal has a WPE of $0.9430\pm0.0013$ and each prediction method we applied was significantly out performed by a simple random walk (the best strategy in the case of structurless time series.
+Since \gcc has such little structure according to WPE it is safe to assume that he best strategy will be random walk and that contining to search for a more accurate prediction strategy will most likely be a fruitless process.
+
+This however does not contradict our hypothesis,
+it just means that ARIMA is ill-suited for the time series under investigation.
+To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since larger
+MASE values constitute worse predictions, ARIMA performed worse than LMA in all
+traces except {\tt SVD$_1$}.
+
+The structure being picked up on by WPE may be linear or nonlinear.
+When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure tha-1t LMA can handle but ARIMA cannot.
+So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
 
@@
 
-%[[Joshua: It may be interesting to look at the PE as a function of
-%prediction horizon. It is known to me that certain start and end
-%locations in the prediction strategy result in different prediction
-%accuracy. Is it the case that when the algorithm tanks it is also the
-%case that PE of the **TEST**(the 10\% we cut off )is high? This may
-%be a great way to illustrate the concept if it works... OR maybe do
-%complexity of forward path in embedding space?]]  some notes about
-%this looking at row Cache 1000 very good RMSPE 21.4167 m = 3 4 5 6
-%1.4192 2.1286 2.5942 2.9230 row Cache 2000 very good RMSPE 19.0990 m
-%= 3 4 5 6 1.4145 2.1305 2.5825 2.9095 so even better row Cache 2010
-%RMSPE 19.9716 m = 3 4 5 6 1.4149 2.1311 2.5830 2.9096 row Cache 1990
-%RMSPE 19.1465 So prediction Horizons ranked in order of best RMSPE
-%RMSPE L 3 4 5 6 1st 19.0990 2000 1.4145 2.1305 2.5825 2.9095 2nd
-%19.1465 1990 1.4151 2.1293 2.5803 2.9067 3rd 19.9716 2010 1.4149
-%2.1311 2.5830 2.9096 3rd 21.4167 1000 1.4192 2.1286 2.5942 2.9230
-%ipcgcc m = 3 pe = 1.7770 penorm = 0.9918 m= 4 pe = 3.1341 penorm =
-%0.9862 m = 5 pe = 4.6985 penorm = 0.9814 m=6 pe = 6.4242 penorm =
-%0.9764 m=7 pe = 8.2450 penorm = 0.9671 >> [pe penorm] =
-%permCalc(gcc100TCM,5,1) pe = 4.7299 penorm = 0.9880 >> [pe penorm] =
-%permCalc(gcc100TCM,6,1) pe = 6.4707 penorm = 0.9835
-
 \section{New Figures and Tables}
-
-Possible new error measure
-
-$$q_t = \frac{e_t}{\frac{1}{n-1}\sum^n_{i=2}|Y_i-Y_{i-1}|}$$
-
-$$MASE = mean(|q_t|)$$
-
-Helpful to remember that $F_i = Y_{i-1}$ for random walk prediction
-
-``When $MASE<1$, the proposed method gives, on average, smaller errors than the one-step errors from the na\"ive method.
-If multi-step forecasts are being computed, it is possible to scale by the in- sample MAE[[Mean Absolute Error(mean($|e_t|$)]] computed from multi-step na\"iõve forecasts."
-
-``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series.
-So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method.
-Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one.
-"
-
-We use this error measure with $n$ at the start of the predicted chunk.
-
-\begin{table}[htdp]
-\caption{ Embedding Parameters for reference if needed. }
-\begin{center}
-\begin{tabular}{|c|c|c|}
-\hline
-&$\tau$ & $m$ \\
-\hline
-gcc & 10& 13\\
-col\_major  &2&12\\
-SVD\_Full &  10  &  12 \\
-SVD\_IPC\_Regime1 & 5 &14  \\
-SVD\_IPC\_Regime2 & 10   &12  \\
-SVD\_IPC\_Regime3 & 2   &9  \\
-SVD\_IPC\_Regime4 &  3  & 11 \\
-SVD\_IPC\_Regime5 &  23  & 10 \\
-SVD\_IPC\_Regime6 &  30  & 12  \\
-\hline
-\end{tabular}
-\end{center}
-\label{default}
-\end{table}%
-
-\begin{table}[htdp]
-\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
-\hline
-gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-col\_major           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-SVD\_IPC\_Regime1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-SVD\_IPC\_Regime2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
-SVD\_IPC\_Regime3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-SVD\_IPC\_Regime4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-SVD\_IPC\_Regime5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-SVD\_IPC\_Regime6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{default}
-\end{table}%
 

### 2014-02-07 unusedFigsTables.tex
working on the results/bib/and figures

@@
 \end{table}%
+
+\begin{table}[htdp]
+\caption{ Embedding Parameters for reference if needed. }
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+&$\tau$ & $m$ \\
+\hline
+gcc & 10& 13\\
+col\_major  &2&12\\
+SVD\_Full &  10  &  12 \\
+SVD\_IPC\_Regime1 & 5 &14  \\
+SVD\_IPC\_Regime2 & 10   &12  \\
+SVD\_IPC\_Regime3 & 2   &9  \\
+SVD\_IPC\_Regime4 &  3  & 11 \\
+SVD\_IPC\_Regime5 &  23  & 10 \\
+SVD\_IPC\_Regime6 &  30  & 12  \\
+\hline
+\end{tabular}
+\end{center}
+\label{default}
+\end{table}%
+
+ Table~\ref{tab:pe} shows the permutation entropy results for the
+ examples considered in this paper, with the nRMSPE prediction
+ accuracies from the previous section included alongside for easy
+ comparison.
+
+ \begin{table}[htbp]
+   % increase table row spacing, adjust to taste
+   \renewcommand{\arraystretch}{1.3}
+   \caption{TCM NRMSE and PE NORMS}
+   \label{tab:table_example}
+   \centering
+   % Some packages, such as MDW tools, offer better commands for making tables
+   % than the plain LaTeX2e tabular which is used here.
+   \begin{tabular}{|c|c|c|c|c|c|}
+     \hline
+     TCM & NRMSE & $m=3$  &$m=4$&$m=5$&$m=6$\\
+     \hline
+     \tt{row\_major} & 0.0102 & 0.7900  &0.6751&0.5458&0.4491\\
+     \hline
+     \tt{col\_major} & 0.0055 & 0.6411  &0.5029&0.4515&0.3955 \\
+     \hline
+     \tt{403.gcc} & 0.0865 & 0.9954  &0.9916&0.9880&0.9835\\
+     \hline
+     \tt{482.sphinx3} & 0.1142 & 0.9958  &0.9913&0.9866&0.9802 \\
+      \hline
+   \end{tabular}
+ \end{table}
+
+%  \begin{table}[htbp]
+%   % increase table row spacing, adjust to taste
+%   %\renewcommand{\arraystretch}{1.3}
+%   \caption{Cache-miss rate performance traces: prediction error
+%     (nRMSPE) and permutation entropy}
+%   \label{tab:TCMpe}
+%   \centering
+%   % Some packages, such as MDW tools, offer better commands for making tables
+%   % than the plain LaTeX2e tabular which is used here.
+%   \begin{tabular}{|c|c|c|c|c|c|}
+%     \hline
+%      & nRMSE   &$n=4$&$n=5$&$n=6$\\
+%     \hline
+%     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
+%     \hline
+%     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
+%     \hline
+%     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
+%     \hline
+%     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
+%      \hline
+%   \end{tabular}
+% \end{table}
+
+ \begin{table}[htbp]
+   % increase table row spacing, adjust to taste
+   \renewcommand{\arraystretch}{1.3}
+   \caption{IPC NRMSE and PE NORMS}
+   \label{tab:table_example}
+   \centering
+   % Some packages, such as MDW tools, offer better commands for making tables
+   % than the plain LaTeX2e tabular which is used here.
+   \begin{tabular}{|c|c|c|c|c|c|}
+     \hline
+     ipc & NRMSE & $n=3$  &$n=4$&$n=5$&$n=6$\\
+     \hline
+      \tt{row\_major} & 0.0095 & 0.9912  &0.9723&0.9354&0.8876\\
+     \hline
+      \tt{col\_major} & 0.0071 & 0.9428  &0.8356&0.7601&0.6880\\
+     \hline
+     \tt{403.gcc} & 0.1805 & 0.9918  &0.9862&0.9814&0.9764\\
+     \hline
+     \tt{482.sphinx3} & 0.2946 & 0.9978  &0.9951&0.9914&0.9849 \\
+      \hline
+   \end{tabular}
+ \end{table}
+
+%  \begin{table}[htbp]
+%   % increase table row spacing, adjust to taste
+%   %\renewcommand{\arraystretch}{1.1}
+%   \caption{Processor load performance traces: prediction error
+%     (nRMSPE) and permutation entropy}
+%   \label{tab:IPCpe}
+%   \centering
+%   % Some packages, such as MDW tools, offer better commands for making tables
+%   % than the plain LaTeX2e tabular which is used here.
+%   \begin{tabular}{|c|c|c|c|c|}
+%     \hline
+%      & nRMSPE   &$n=4$&$n=5$&$n=6$\\
+%     \hline
+%      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
+%     \hline
+%      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
+%     \hline
+%     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
+%     \hline
+%     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
+%      \hline
+%   \end{tabular}
+% \end{table}
+
+ %  \begin{table}[htbp]
+ %   \caption{Prediction error (in nRMSPE) and permutation entropy (for
+ %     different wordlengths $n$}
+ %   \label{tab:pe}
+ %   \centering
+ %   \begin{tabular}{|c|c|c|c|c|c|}
+ %     \hline
+ %      {\bf cache misses} & error   &$n=4$&$n=5$&$n=6$\\
+ %     \hline
+ %     \tt{row\_major}  & 0.0324  &0.6751&0.5458&0.4491\\
+ %     \hline
+ %     \tt{col\_major}  & 0.0080  &0.5029&0.4515&0.3955 \\
+ %     \hline
+ %     \tt{403.gcc}  & 0.1416  &0.9916&0.9880&0.9835\\
+ %     \hline
+ %     \tt{482.sphinx3}  & 0.2032  &0.9913&0.9866&0.9802 \\
+ %      \hline
+ %      \hline
+ %      {\bf insts per cyc} & error   &$n=4$&$n=5$&$n=6$\\
+ %     \hline
+ %      \tt{row\_major} & 0.0778  &0.9723&0.9354&0.8876\\
+ %     \hline
+ %      \tt{col\_major} & 0.0161   &0.8356&0.7601&0.6880\\
+ %     \hline
+ %     \tt{403.gcc} & 0.2033   &0.9862&0.9814&0.9764\\
+ %     \hline
+ %     \tt{482.sphinx3} & 0.3670 & 0.9951&0.9914&0.9849 \\
+ %      \hline
+ %   \end{tabular}
+ % \end{table}
+
+  \begin{table}[htbp]
+   % increase table row spacing, adjust to taste
+   \renewcommand{\arraystretch}{1.3}
+   \caption{Normalized root mean squared error (nRMSE) of 4000-point
+     predictions of memory \& processor performance from different
+     programs.}
+   \label{tab:PredError}
+   \centering
+   % Some packages, such as MDW tools, offer better commands for making tables
+   % than the plain LaTeX2e tabular which is used here.
+%   \begin{tabular}{|c|c|c|c|c|c|}
+%     \hline
+%      & cache misses NRMSE & IPC NRMSE \\
+%     \hline
+%     \tt{row\_major}  & 0.0102  & 0.0095 \\
+%          \hline
+%     \tt{col\_major}  &0.0055& 0.0071 \\
+%     \hline
+%     \tt{403.gcc}  & 0.0865& 0.1805 \\
+%          \hline
+%     \tt{482.sphinx3} & 0.1142& 0.2946 \\
+%      \hline
+%   \end{tabular}
+% \end{table}
+   \begin{tabular}{|c|c|c|c|c|c|}
+     \hline
+      & cache miss rate & instrs per cycle \\
+     \hline
+     \tt{row\_major}  & 0.0324  & 0.0778 \\
+          \hline
+     \tt{col\_major}  &0.0080& 0.0161 \\
+     \hline
+     \tt{403.gcc}  & 0.1416& 0.2033 \\
+          \hline
+     \tt{482.sphinx3} & 0.2032& 0.3670 \\
+      \hline
+   \end{tabular}
+ \end{table}
 
@@
 
+%[[Joshua: It may be interesting to look at the PE as a function of
+%prediction horizon. It is known to me that certain start and end
+%locations in the prediction strategy result in different prediction
+%accuracy. Is it the case that when the algorithm tanks it is also the
+%case that PE of the **TEST**(the 10\% we cut off )is high? This may
+%be a great way to illustrate the concept if it works... OR maybe do
+%complexity of forward path in embedding space?]]  some notes about
+%this looking at row Cache 1000 very good RMSPE 21.4167 m = 3 4 5 6
+%1.4192 2.1286 2.5942 2.9230 row Cache 2000 very good RMSPE 19.0990 m
+%= 3 4 5 6 1.4145 2.1305 2.5825 2.9095 so even better row Cache 2010
+%RMSPE 19.9716 m = 3 4 5 6 1.4149 2.1311 2.5830 2.9096 row Cache 1990
+%RMSPE 19.1465 So prediction Horizons ranked in order of best RMSPE
+%RMSPE L 3 4 5 6 1st 19.0990 2000 1.4145 2.1305 2.5825 2.9095 2nd
+%19.1465 1990 1.4151 2.1293 2.5803 2.9067 3rd 19.9716 2010 1.4149
+%2.1311 2.5830 2.9096 3rd 21.4167 1000 1.4192 2.1286 2.5942 2.9230
+%ipcgcc m = 3 pe = 1.7770 penorm = 0.9918 m= 4 pe = 3.1341 penorm =
+%0.9862 m = 5 pe = 4.6985 penorm = 0.9814 m=6 pe = 6.4242 penorm =
+%0.9764 m=7 pe = 8.2450 penorm = 0.9671 >> [pe penorm] =
+%permCalc(gcc100TCM,5,1) pe = 4.7299 penorm = 0.9880 >> [pe penorm] =
+%permCalc(gcc100TCM,6,1) pe = 6.4707 penorm = 0.9835
+
 \end{document}

### 2014-02-10 pecomplexity.tex
some editing of the results section

@@
 
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% refereed to as the ``learning/training" signal, $\{X_{i,obs}\}_{i=1}^{i=n}$ and the last 10\% known as the ``test/correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to build the initial models (e.g., LMA,ARIMA,na\"ive) as described in Section \ref{sec:compModel}.
-The test signal is used both to test the forecasts and to refit the models after forecasting.
-In particular, using the models we perform $k$ 1-step predictions $\{p_j\}_{j=n+1}^{k+n+1}$, after each prediction we append the training signal with the the next point in the correct signal.
-We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.
+In order to analyze correctness of each prediction we split each time series
+into two pieces: the first 90\% refereed to as the ``learning/training" signal,
+$\{X_{i,obs}\}_{i=1}^{i=n}$ and the last 10\% known as the ``test/correct"
+signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to build the
+initial models (e.g., LMA, ARIMA, na\"ive) as described in Section
+\ref{sec:compModel}. The test signal is used both to test the forecasts and to
+refit the models after forecasting. In particular, using the models we perform
+$k$ 1-step predictions $\{p_j\}_{j=n+1}^{k+n+1}$, after each prediction we
+append the training signal with the the next point in the correct signal. We
+would like to note that this rebuilding occurs due to a problem with ARIMA
+models converging to a mean prediction if too long of a prediction horizon is
+used, this is not a handicap of either LMA or na\"ive.
 
-In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
+In order to compare the resulting forecasts we calculate the Mean Absolute
+Squared Error (MASE)\cite{MASE} between the true and predicted signals:
 
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
+\begin{align*}
+  \textrm{MASE} = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}
+\end{align*}
 
@@
 
-$$s=\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_j=c_{j-1})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
-A $MASE<1$ means that the prediction method gives, on average, smaller errors than the 1-step errors from the random walk forecast strategy.
-Analogously, a $MASE>1$ means that the prediction method did worse, on average than the 1-step errors from the random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution of MASEs for 15 runs of each of the 8 signals across the 3 prediction forecast methods.
-For comparison Table \ref{tab:error} also has weighted permutation
+$$s=\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|$$ is the average in-sample
+forecast error for a random walk prediction $(p_j=c_{j-1})$. This error method
+was introduced in \cite{MASE} as a ``generally applicable measurement of
+forecast accuracy without the problems seen in the other measurements." The
+major advantage of MASE is that it allows fair comparison across methods,
+prediction horizons and varying signal scales. A $MASE<1$ means that the
+prediction method gives, on average, smaller errors than the 1-step errors from
+the random walk forecast strategy. Analogously, a $MASE>1$ means that the
+prediction method did worse, on average than the 1-step errors from the random
+walk forecast strategy. In Table \ref{tab:error} we provide the distribution of
+MASEs for 15 runs of each of the 8 signals across the 3 prediction forecast
+methods. For comparison Table \ref{tab:error} also has weighted permutation
 
@@
 \begin{table}[htdp]
-\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
+\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average WPE at word length 5 and 6 for each signal. }
 \begin{center}
@@
 
-As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
-indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
-the exception of one outlier whose behavior we can explain. In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over methods on that signal) for each of the 8 signals.  
+As can be seen in Table \ref{tab:error} the relationship between prediction
+accuracy and the weighted permutation entropy (WPE) is much as we conjectured:
+performance traces with high WPE are indeed harder to predict using the
+forecasting models described in Section \ref{sec:compModel}.
+Figure~\ref{fig:pred_vs_wpe} demonstrates this primary finding, with the
+exception of one outlier whose behavior we can explain. In
+Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE
+over methods on that signal) for each of the 8 signals.
 \begin{figure}[htbp]
@@
   SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_ent}
+  \label{fig:pred_vs_wpe}
 \end{figure}
-We find that the relationship between the two is roughly linear. The single outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE: namely
-that in the absence of any large features, as all the other regimes have, the
-WPE effectively falls back to the standard PE for which the noisy behavior
+We find that the relationship between the two is roughly linear. The single
+outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
+namely that in the absence of any large features, as all the other regimes have,
+the WPE effectively falls back to the standard PE which the noisy behavior
 drives toward 1.0.
 
-This primary finding: ``Signals that are more complex are harder to predict." At first glance this finding may seem incredibly trivial, of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core result here as we will elucidate now with Figure \ref{fig:lma_vs_arima}.
+This primary finding: ``Signals that are more complex are harder to predict." At
+first glance this finding may seem incredibly trivial, of course a complex
+signal is harder to predict than a trivial one, but this is \emph{not} the core
+result here as we will elucidate now with Figure \ref{fig:predictions_vs_wpe}.
 \begin{figure}[htbp]
@@
     the word length used is $6$.}
-    \label{fig:lma_pred_vs_ent}
+    \label{fig:lma_pred_vs_wpe}
   \end{subfigure}%
@@
     these, the word length used is $6$.}
-    \label{fig:arima_pred_vs_ent}
+    \label{fig:arima_pred_vs_wpe}
   \end{subfigure}
-  %\begin{subfigure}{0.5\textwidth}
+  % \begin{subfigure}{0.5\textwidth}
   %   \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-    \caption{The MASE values for LMA against ARIMA. The dashed line is the
-    identity, delineating the traces for which either LMA or ARIMA performed
-    better. All traces except those from {\tt SVD$_1$} lie above the line,
-    indicating that LMA is better suited prediction method for the computer
-   traces considered.}\label{fig:lma_vs_arima}
-
-  %\end{subfigure}
+  %   \caption{The MASE values for LMA against ARIMA. The dashed line is the
+  %   identity, delineating the traces for which either LMA or ARIMA performed
+  %   better. All traces except those from {\tt SVD$_1$} lie above the line,
+  %   indicating that LMA is better suited prediction method for the computer
+  %  traces considered.}\label{fig:lma_vs_arima}
+  % \end{subfigure}
+  \label{fig:predictions_vs_wpe}
 \end{figure}
 
-\item Talk about structure analysis. Just because there is forward infromation transfer, does not mean that linear predictors can get at this. 
-\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by side plot. 
-\item  full results.  Image: MASE vs. WPE for both LMA \&
-ARIMA.  Points to make: 
+\item Talk about structure analysis. Just because there is forward information
+transfer, does not mean that linear predictors can get at this.
+\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
+side plot.
+\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
 \begin{enumerate}
@@
 \item clusters are shaped differently---tight or not
-\item clusters move
-differently between LMA and ARIMA 
-\item finally, the diagonal line is
-important.  If you're below it, you could do better.
+\item clusters move differently between LMA and ARIMA
+\item finally, the diagonal line is important. If you're below it, you could do
+better.
 
@@
 
-In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
+In Figures~\ref{fig:lma_pred_vs_wpe} and \ref{fig:arima_pred_vs_wpe} we directly
 compare the performance of the LMA and ARIMA prediction methods (respectively)
@@
 the naive method. On the other hand, the ARIMA MASE values are largely
-uncorrelated with WPE values.
-The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as is proposed as a good strategy in \cite{ARIMA-autofit} and say you got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MASE>1$).
-How do we know if the prediction strategy is not adequte for the prediction task or if the signal is simply too complex to predict.
-If a signal is too complex and little forward information transfer is present we may not be able to do better than the random walk, in which case we should not kill ourselves over finding a more complicated prediction strategy.
-However, if we measure the complexity to be low, $WPE<0.85$ we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-For example, consider \col (the blue +s in Figure \ref{fig:lma_vs_arima}).
-If we use the out-of-the-box ARIMA from \cite{ARIMA-autofit} we get MASE ranging from 0.5 all the way to 1.0.
-That is we may think that the best we can do is twice as good as random-walk but sometimes we only do as well as random walk.
-In which case the random walk prediction strategy is probably your best bet as it is so simple and produces very similar error.
-However, if we calculate WPE we see that \col has a WPE of $0.5131\pm 0.0034$.
-This incredibly low entropy implies a great deal of deterministic structure that we can utilize for prediction, even though this did not appear to be the case with ARIMA.
-In fact using LMA on this signal we get an average MASE of $0.05\pm0.0018$, an error that is on-average 20 times better than the random walk forecast.
-Alternatively, we can look at the other end of the spectrum at \gcc.
-This signal has a WPE of $0.9430\pm0.0013$ and each prediction method we applied was significantly out performed by a simple random walk (the best strategy in the case of structurless time series.
-Since \gcc has such little structure according to WPE it is safe to assume that he best strategy will be random walk and that contining to search for a more accurate prediction strategy will most likely be a fruitless process.
+uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in
+large part one of the major findings of this work. More specifically, say we
+tried to predict an arbitrary noisy real-valued time series with an ``out-of-
+the-box" prediction strategy like ARIMA as proposed in \cite{ARIMA-autofit}, and
+say you got inconsistent and bad forecasts, (i.e., perform worse than the
+na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
+strategy is not adequate for the prediction task, or if the signal is simply too
+complex to predict. If a signal is too complex and little forward information
+transfer is present we may simply not be possible to do better than the random
+walk, in which case we should not worry ourselves over finding a more
+complicated prediction strategy. However if we measure the complexity to be low,
+$\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}), we can likely do better
+than the random walk and should search for more adequate prediction strategies.
+
+By way of example, consider \col (the blue +s in
+Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA method
+from \cite{ARIMA-autofit} we get MASE values ranging from $\sim0.5$ to
+$\sim1.0$. From that information we may assume that the best predictions that
+can be done are twice as good as random-walk, and in some cases predictions will
+only do as well as random-walk. In the latter such cases, the random-walk
+prediction strategy is probably the best bet as it is so simple and produces
+very similar error. However, if we calculate the WPE we see that \col has a WPE
+of $0.5131 \pm 0.0034$. This incredibly low entropy implies a great deal of
+deterministic structure which can be utilize for prediction, even though this
+did not appear to be the case with ARIMA. In fact using LMA on this signal we
+get an average MASE value of $0.05\pm0.0018$, an error that is on average 20
+times better than the random walk forecast, and 10 times better than the best
+ARIMA forcast, with little variance. Alternatively, we can look at the other end
+of the spectrum: \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each
+prediction method we applied was significantly out performed by a simple random
+walk (the best strategy in the case of structureless time series). Since \gcc
+has such little structure according to WPE it is safe to assume that he best
+strategy will be random walk and that continuing to search for a more accurate
+prediction strategy will likely be a fruitless process.
 
@@
 
-The structure being picked up on by WPE may be linear or nonlinear.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure tha-1t LMA can handle but ARIMA cannot.
-So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
+The WPE is sensitive to both linear and nonlinear structure. When you have a low
+WPE and a high ARIMA it could be that the structure WPE is picking up is simply
+nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA
+prediction look really bad there is plenty of structure present as suggested by
+WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it
+into account and does bad.
 

### 2014-02-10 pecomplexity.tex
add figures to introduction

@@
 {\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:cache}, and chaos places fundamental limits on
+Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
 predictability.
@@
 %  \end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[t]{0.475\textwidth}
+    \includegraphics[width=0.95\textwidth]{figs/colshortts}
+    \caption{A short segment of the instructions executed per CPU clock cycle
+    (IPC) during the execution of \col. Each point is the IPC during a 100,000
+    cycle period. There is a great deal of periodicity in this time series.}
+    \label{fig:col-ts}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.475\textwidth}
+    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
+    \caption{The full IPC time series during the execution of \gcc. This time
+    series has very little structure.}
+    \label{fig:gcc-ts}
+  \end{subfigure}
+  \label{fig:sample-ts}
+\end{figure}
 

### 2014-02-10 pecomplexity.tex
add better captions

@@
   \centering
-  \begin{subfigure}{0.5\textwidth}
+  \begin{subfigure}[t]{0.5\textwidth}
     \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
     \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
-    the word length used is $6$.}
+    the word length used is $6$. Prediction error vs WPE values largely fall
+    near the fit from Fig.~\ref{fig:pred_vs_wpe} indicating that LMA is a near-
+    optimal prediction method for this data.}
     \label{fig:lma_pred_vs_wpe}
   \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
+  \begin{subfigure}[t]{0.5\textwidth}
     \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
     \caption{The MASE of ARIMA vs weighted permutation entropy. For each of
-    these, the word length used is $6$.}
+    these, the word length used is $6$. Prediction error vs WPE values for many
+    of the traces fall far to the right of the fit from
+    Fig.~\ref{fig:pred_vs_wpe} indicating that ARIMA is a poor choice to predict
+    this data.}
     \label{fig:arima_pred_vs_wpe}

### 2014-02-10 pecomplexity.tex
more content

@@
 \usepackage{amsthm}
+
+\usepackage{pifont} % check and x marks
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+
 \newcommand{\gcc}{{\tt 403.gcc}~}
@@
 
+\begin{enumerate}
+\item \cmark~Introduce MASE.
+\item \cmark~WPE is a good measure of predictability.  Figure: best athlete MASE vs.
+its WPE.
+\item \cmark~Talk about structure analysis. Just because there is forward information
+transfer, does not mean that linear predictors can get at this.
+\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
+side plot.
+\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
+\begin{enumerate}
+\item \cmark~clusters are distributed differently
+\item \cmark~clusters are shaped differently---tight or not
+\item \cmark~clusters move differently between LMA and ARIMA
+\item finally, the diagonal line is important. If you're below it, you could do
+better.
+\end{enumerate}
+\end{enumerate}
+
 %\item Introduce MASE and explain why random walk isn't *that* naive
@@
 major advantage of MASE is that it allows fair comparison across methods,
-prediction horizons and varying signal scales. A $MASE<1$ means that the
-prediction method gives, on average, smaller errors than the 1-step errors from
-the random walk forecast strategy. Analogously, a $MASE>1$ means that the
-prediction method did worse, on average than the 1-step errors from the random
-walk forecast strategy. In Table \ref{tab:error} we provide the distribution of
-MASEs for 15 runs of each of the 8 signals across the 3 prediction forecast
-methods. For comparison Table \ref{tab:error} also has weighted permutation
+prediction horizons and varying signal scales. A $\textrm{MASE}<1$ means that
+the prediction method gives, on average, smaller errors than the 1-step errors
+from the random walk forecast strategy. Analogously, a $\textrm{MASE}>1$ means
+that the prediction method did worse, on average, than the 1-step errors from the
+random walk forecast strategy. In Table \ref{tab:error} we provide the
+distribution of MASEs for 15 runs of each of the 8 signals across the 3
+prediction forecast methods. For comparison Table \ref{tab:error} also has
+weighted permutation
 
@@
 "
-
-%\item Paragraph on how to interpret MASE
-\begin{enumerate}
-
-\item First paragraph: WPE is a good measure of predictability.  Figure:
-best athlete MASE vs. its WPE.
 
@@
 
-\item Talk about structure analysis. Just because there is forward information
-transfer, does not mean that linear predictors can get at this.
-\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
-side plot.
-\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
-\begin{enumerate}
-\item clusters are distributed differently
-\item clusters are shaped differently---tight or not
-\item clusters move differently between LMA and ARIMA
-\item finally, the diagonal line is important. If you're below it, you could do
-better.
-
-\end{enumerate}
-\end{enumerate}
-
 In Figures~\ref{fig:lma_pred_vs_wpe} and \ref{fig:arima_pred_vs_wpe} we directly
@@
 
-This however does not contradict our hypothesis,
-it just means that ARIMA is ill-suited for the time series under investigation.
-To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since larger
-MASE values constitute worse predictions, ARIMA performed worse than LMA in all
-traces except {\tt SVD$_1$}.
-
 The WPE is sensitive to both linear and nonlinear structure. When you have a low
@@
 into account and does bad.
+
+Other noteworthy features of the LMA and ARIMA results are the cluster locations
+and distributions. The WPE values of each run any particular program tend to
+have little variance, leading to the clusters in
+Figures~\ref{fig:arima_pred_vs_wpe}, \ref{fig:lma_pred_vs_wpe} to be fairly
+constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
+low too, resulting is small, tight clusters. The ARIMA MASE values of the \col
+traces, however, have a large variance resulting in the spread seen in
+Fig~\ref{fig:arima_pred_vs_wpe}. Not only are the MASE values of that cluster
+bad, in that other predictors vastly outperform it, but they are inconsistent.
+Furthermore, since LMA can predict nonlinear behavior while ARIMA can not, we
+see that the clusters in Fig.~\ref{fig:arima_pred_vs_wpe} are mostly further to
+the right than those in Fig.~\ref{fig:lma_pred_vs_wpe}.
+
+This however does not contradict our hypothesis,
+it just means that ARIMA is ill-suited for the time series under investigation.
+To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since larger
+MASE values constitute worse predictions, ARIMA performed worse than LMA in all
+traces except {\tt SVD$_1$}.
 

### 2014-02-10 pecomplexity.tex
New Figures Outlines and results

--- 2014-02-10 pecomplexity.tex
@@
 \newtheorem*{mydef}{Definition}
-\title{Determinism, Complexity, and Predictability in Computer Performance}
+\title{Determinism, Complexity, and Predictability in Computer Performance\\
+Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
+\\On Quantifying Predictability through Structural Complexity
+\\On Quantifying Predictability with Structural Complexity
+}
 
@@
 {\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
+Figure~\ref{fig:ipc}, and chaos places fundamental limits on
 predictability.
 %
-% \begin{figure}[htbp]
-%    \centering
-%    \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
-%    % where an .eps filename suffix will be assumed under latex,
-%    % and a .pdf suffix will be assumed for pdflatex
-%    \caption{A small snippet of the L2 cache miss rate of {\tt
-%        col\_major}, a three-line C program that repeatedly initializes
-%      a matrix in column-major order, running on an Intel Core
-%      Duo\textsuperscript{\textregistered}-based machine.  Even this
-%      simple program exhibits chaotic performance dynamics.}
-%    \label{fig:cache}
-%  \end{figure}
-
  \begin{figure}[htbp]
     \centering
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/colshortts}
-    \caption{A short segment of the instructions executed per CPU clock cycle
-    (IPC) during the execution of \col. Each point is the IPC during a 100,000
-    cycle period. There is a great deal of periodicity in this time series.}
-    \label{fig:col-ts}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
-    \caption{The full IPC time series during the execution of \gcc. This time
-    series has very little structure.}
-    \label{fig:gcc-ts}
-  \end{subfigure}
-  \label{fig:sample-ts}
+    \includegraphics[width=\textwidth]{figs/colshortts}
+    % where an .eps filename suffix will be assumed under latex,
+    % and a .pdf suffix will be assumed for pdflatex
+    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
+        \col}, a three-line C program that repeatedly initializes
+      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
+      simple program exhibits chaotic performance dynamics.}
+   \label{fig:ipc}
   \end{figure}
@@
 \cite{josh-ida2011}.  In the case of simple microkernels like the one
-that produced the trace in Figure~\ref{fig:cache}, this deterministic
+that produced the trace in Figure~\ref{fig:ipc}, this deterministic
 modeling and forecast strategy works very well.  In more-complicated
@@
 from these systems---which results from the inherent dimension,
-nonlinearity, and nonstationarity of the dynamics, as well as from
+non-linearity, and non-stationarity of the dynamics, as well as from
 measurement issues like noise, aggregation, and finite data
@@
 
+To Add to this section:
+
+\begin{enumerate}
+\item Description of ARIMA
+\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
+
+\item Maybe add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the begining of the results.
+
+\end{enumerate}
+
 % took out for space
@@
 method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
-plot of the data from Figure~\ref{fig:cache}, embedded following this
+plot of the data from Figure~\ref{fig:ipc}, embedded following this
 procedure, is shown in Figure~\ref{fig:embedding}.
 
-[[Add a plot of an embedding...]]
+ \begin{figure}
+   \centering
+     \includegraphics[width=\textwidth]{figs/colipc3d.png}
+     \caption{A 3D projection of a delay-coordinate embedding of the trace
+ from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
+ }
+ \label{fig:embedding}
+ \end{figure}
 
@@
 space of the given point, then taking that neighbor's path as the
-forecast---works quite well on the trace in Figure~\ref{fig:cache}, as
+forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
 shown in Figure~\ref{fig:cachePredTS}.
@@
     \caption{A forecast of the last 4,000 points of the signal in
-      Figure~\ref{fig:cache} using an LMA-based strategy on the
+      Figure~\ref{fig:ipc} using an LMA-based strategy on the
       embedding in Figure~\ref{fig:embedding}.  Red circles and blue
@@
 regimes.
-[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series differnet levels of complexity and predictability exist.
+[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
 Maybe point at some of the predicting predictability papers.]]
 
+%\begin{figure}[htbp]
+%  \centering
+%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
+%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
+%    are regions where the window overlaps regimes. The window size used is
+%    $5,000 \times 100,000$ instructions and the word length is $4$.}
+%  \label{fig:wwpe}
+%\end{figure}
+
+%%FIGURE INTENT
+%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
+%%%%%%%%%%%%%%%%%
 \begin{figure}[htbp]
   \centering
-  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-  \caption{[[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure]]The weighted permutation entropy of one run of SVD. The gray bands
+         \caption{
+[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly.
+Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD.
+The gray bands
     are regions where the window overlaps regimes. The window size used is
-    $5,000 \times 100,000$ instructions and the word length is $4$.}
-  \label{fig:wwpe}
+    $5,000 \times 100,000$ instructions and the word length is $4$.}\label{fig:wwpe}
+  \begin{subfigure}{0.3\textwidth}
+    \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
+    \caption{The instructions per cycle of \svd.
+Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
+From left to right each change in color represents a change in regime for 6 regimes in total.
+}
+    \label{fig:svd_ts}
+  \end{subfigure}%
+  %\\
+  \begin{subfigure}{0.3\textwidth}
+    \includegraphics[width=0.3\textwidth]{figs/SVD_wwpe}
+    \caption{The MASE of ARIMA vs weighted permutation entropy. }
+    \label{fig:svd_wwpe}
+  \end{subfigure}
+  %\begin{subfigure}{0.5\textwidth}
+  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
+   % \caption{The MASE values for LMA against ARIMA. The dashed line is the
+   % identity, delineating the traces for which either LMA or ARIMA %performed
+%    better. All traces except those from {\tt SVD$_1$} lie above the line,
+%    indicating that LMA is better suited prediction method for the %computer
+%   traces considered.}\label{fig:lma_vs_arima}
+  %\end{subfigure}
 \end{figure}
@@
 %\item Introduce MASE and explain why random walk isn't *that* naive
-
-%We calculated a figure of merit for each prediction method as follows.
+%\item Paragraph on how to interpret MASE
 
@@
 
-In order to analyze correctness of each prediction we split each time series
-into two pieces: the first 90\% refereed to as the ``learning/training" signal,
-$\{X_{i,obs}\}_{i=1}^{i=n}$ and the last 10\% known as the ``test/correct"
-signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to build the
-initial models (e.g., LMA, ARIMA, na\"ive) as described in Section
-\ref{sec:compModel}. The test signal is used both to test the forecasts and to
-refit the models after forecasting. In particular, using the models we perform
-$k$ 1-step predictions $\{p_j\}_{j=n+1}^{k+n+1}$, after each prediction we
-append the training signal with the the next point in the correct signal. We
-would like to note that this rebuilding occurs due to a problem with ARIMA
-models converging to a mean prediction if too long of a prediction horizon is
-used, this is not a handicap of either LMA or na\"ive.
+In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
+The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
+The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
+In particular, we perform $k$ 1-step predictions, after each 1-step prediction we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
+This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.}
 
-In order to compare the resulting forecasts we calculate the Mean Absolute
-Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-
-\begin{align*}
-  \textrm{MASE} = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}
-\end{align*}
-
-The scaling term here:
-
-$$s=\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|$$ is the average in-sample
-forecast error for a random walk prediction $(p_j=c_{j-1})$. This error method
-was introduced in \cite{MASE} as a ``generally applicable measurement of
-forecast accuracy without the problems seen in the other measurements." The
-major advantage of MASE is that it allows fair comparison across methods,
-prediction horizons and varying signal scales. A $MASE<1$ means that the
-prediction method gives, on average, smaller errors than the 1-step errors from
-the random walk forecast strategy. Analogously, a $MASE>1$ means that the
-prediction method did worse, on average than the 1-step errors from the random
-walk forecast strategy. In Table \ref{tab:error} we provide the distribution of
-MASEs for 15 runs of each of the 8 signals across the 3 prediction forecast
-methods. For comparison Table \ref{tab:error} also has weighted permutation
-
-%``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series.
-So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method.
-Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one.
-"
+As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
+%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
+The scaling term for MASE:
+$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
+is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
+This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
+When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
+Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
+In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
+we give mean $\pm$ std.
+dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
+For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for work lengths of $l=5$ and $l=6$.
 
 \begin{table}[htdp]
-\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average WPE at word length 5 and 6 for each signal. }
+\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures]] }
 \begin{center}
@@
 
-%This is then used as a scaling term in the following way:
+%\item First paragraph: WPE is a good measure of predictability.  Figure:
+%best athlete MASE vs. its WPE.
 
-%$$qt = \frac{e_t}{\frac{1}{n-1}\}sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$
-
-%$$MASE = \frac{\frac{\sum_{j=n+1}^{k+n+1} |c_j-p_j| }{k}}{\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|}$$
-
-%and is comprised of the first 90\% of the time series, the learning signal is what we build the initial forecast models with.
-The second piece is called the test  the  back the last 10\% of each signal.
-Built the initial forecast model for each signal.
-We then perform an in-sample random walk forecast at every point in the training
-
-%Possible new error measure
-
-%$$MASE = mean(|q_t|)$$
-
-%Helpful to remember that $F_i = Y_{i-1}$ for random walk prediction
-
-%``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series.
-So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method.
-Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one.
-"
-
-%\item Paragraph on how to interpret MASE
-\begin{enumerate}
-
-\item First paragraph: WPE is a good measure of predictability.  Figure:
-best athlete MASE vs. its WPE.
-
-As can be seen in Table \ref{tab:error} the relationship between prediction
-accuracy and the weighted permutation entropy (WPE) is much as we conjectured:
-performance traces with high WPE are indeed harder to predict using the
-forecasting models described in Section \ref{sec:compModel}.
-Figure~\ref{fig:pred_vs_wpe} demonstrates this primary finding, with the
-exception of one outlier whose behavior we can explain. In
-Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE
-over methods on that signal) for each of the 8 signals.
+As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
+indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
+the exception of one outlier whose behavior we will explain.
+In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over all 3 methods over all runs of that program) for each of the 8 signals.
 \begin{figure}[htbp]
@@
   SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_wpe}
+  \label{fig:pred_vs_ent}
 \end{figure}
-We find that the relationship between the two is roughly linear. The single
-outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
-namely that in the absence of any large features, as all the other regimes have,
-the WPE effectively falls back to the standard PE which the noisy behavior
+We find that the relationship between the two is roughly linear. The single outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE: namely
+that in the absence of any large features, as all the other regimes have, the
+WPE effectively falls back to the standard PE for which the noisy behavior
 drives toward 1.0.
 
-This primary finding: ``Signals that are more complex are harder to predict." At
-first glance this finding may seem incredibly trivial, of course a complex
-signal is harder to predict than a trivial one, but this is \emph{not} the core
-result here as we will elucidate now with Figure \ref{fig:predictions_vs_wpe}.
+It may be tempting to paraphrase this primary finding as: ``Signals that are more complex are harder to predict.", but this truly misses the impact of this work.
+A more correct statement would be ``Time series which exhibit structure-less complexity are harder to predict for any model which uses structure as it's information processor.
+" [[I don't like this at all]]
+
+\begin{itemize}
+\item Complexity need not be hard to predict (can point at the simple predictions paper)
+\item random walk for example is best predicted by guess what just happend
+\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not. 
+\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
+\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
+
+\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
+And a constant, linear and nonlinear prediction strategy all fail.
+We should be able to conclude that guessing random values is the best we can do as is shown by MASE
+
+\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction
+
+\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
+That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
+This gives a practitioner leverage on when to give up and when to keep working.
+
+\end{itemize}
+
+Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}.
+[[Joshua: Make sure we make it clear WPE not only tells you when there is strucure to use but also what kinds of complexity are usable.
+e.g., white noise and deterministic chaos are both complex but one is predictable.
+The kind of complexity is important.
+Talk about chaos and random walk being complex but from a predictive standpoint that complexity spectrum is important **AND** quantifiable by WPE ]]
 \begin{figure}[htbp]
   \centering
-  \begin{subfigure}{0.5\textwidth}
+       \caption{
+For each of
+    these, the word length used is $6$.       
+       The MASE values for LMA against ARIMA. The dashed line is the
+    identity, delineating the traces for which either LMA or ARIMA performed
+   better. All traces except those from {\tt SVD$_1$} lie above the line,
+  indicating that LMA is better suited prediction method for the %computer
+   traces considered.}\label{fig:lma_vs_arima}  
+  \begin{subfigure}{\textwidth}
     \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
-    the word length used is $6$.}
-    \label{fig:lma_pred_vs_wpe}
+    \caption{The MASE of LMA vs weighted permutation entropy. }
+    \label{fig:lma_pred_vs_ent}
   \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
+  \\
+  \begin{subfigure}{\textwidth}
     \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. For each of
-    these, the word length used is $6$.}
-    \label{fig:arima_pred_vs_wpe}
+    \caption{The MASE of ARIMA vs weighted permutation entropy. }
+    \label{fig:arima_pred_vs_ent}
   \end{subfigure}
-  % \begin{subfigure}{0.5\textwidth}
+  %\begin{subfigure}{0.5\textwidth}
   %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
    % \caption{The MASE values for LMA against ARIMA. The dashed line is the
-  %   identity, delineating the traces for which either LMA or ARIMA performed
+   % identity, delineating the traces for which either LMA or ARIMA %performed
 %    better. All traces except those from {\tt SVD$_1$} lie above the line,
-  %   indicating that LMA is better suited prediction method for the computer
+%    indicating that LMA is better suited prediction method for the %computer
 %   traces considered.}\label{fig:lma_vs_arima}
-  % \end{subfigure}
-  \label{fig:predictions_vs_wpe}
+
+  %\end{subfigure}
 \end{figure} 
-
-\item Talk about structure analysis. Just because there is forward information
-transfer, does not mean that linear predictors can get at this.
-\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
-side plot.
-\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
-\begin{enumerate}
-\item clusters are distributed differently
-\item clusters are shaped differently---tight or not
-\item clusters move differently between LMA and ARIMA
-\item finally, the diagonal line is important. If you're below it, you could do
-better.
-
-\end{enumerate}
-\end{enumerate}
-
-In Figures~\ref{fig:lma_pred_vs_wpe} and \ref{fig:arima_pred_vs_wpe} we directly
+In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
 compare the performance of the LMA and ARIMA prediction methods (respectively)
@@
 under consideration. The LMA MASE values are largely similar to those of the
-best predictions, primarily because LMA often performed superior to ARIMA and
-the naive method. On the other hand, the ARIMA MASE values are largely
-uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in
-large part one of the major findings of this work. More specifically, say we
-tried to predict an arbitrary noisy real-valued time series with an ``out-of-
-the-box" prediction strategy like ARIMA as proposed in \cite{ARIMA-autofit}, and
-say you got inconsistent and bad forecasts, (i.e., perform worse than the
-na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
-strategy is not adequate for the prediction task, or if the signal is simply too
-complex to predict. If a signal is too complex and little forward information
-transfer is present we may simply not be possible to do better than the random
-walk, in which case we should not worry ourselves over finding a more
-complicated prediction strategy. However if we measure the complexity to be low,
-$\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}), we can likely do better
-than the random walk and should search for more adequate prediction strategies.
+best predictions, primarily because LMA often performed superior to ARIMA (and
+the na\"ive method). So it is not surprising that the MASE from LMA and the corresponding MASE are highly correlated. On the other hand, the ARIMA MASE values are largely
+uncorrelated with WPE values.
 
-By way of example, consider \col (the blue +s in
-Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA method
-from \cite{ARIMA-autofit} we get MASE values ranging from $\sim0.5$ to
-$\sim1.0$. From that information we may assume that the best predictions that
-can be done are twice as good as random-walk, and in some cases predictions will
-only do as well as random-walk. In the latter such cases, the random-walk
-prediction strategy is probably the best bet as it is so simple and produces
-very similar error. However, if we calculate the WPE we see that \col has a WPE
-of $0.5131 \pm 0.0034$. This incredibly low entropy implies a great deal of
-deterministic structure which can be utilize for prediction, even though this
-did not appear to be the case with ARIMA. In fact using LMA on this signal we
-get an average MASE value of $0.05\pm0.0018$, an error that is on average 20
-times better than the random walk forecast, and 10 times better than the best
-ARIMA forcast, with little variance. Alternatively, we can look at the other end
-of the spectrum: \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each
-prediction method we applied was significantly out performed by a simple random
-walk (the best strategy in the case of structureless time series). Since \gcc
-has such little structure according to WPE it is safe to assume that he best
-strategy will be random walk and that continuing to search for a more accurate
-prediction strategy will likely be a fruitless process.
+The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work.
+More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as is proposed as a good strategy in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MASE>1$).
+How can we know if the prediction strategy is not adequate for the prediction task or if the signal is simply too complex and unstructured to predict.
+If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not kill ourselves over finding a more complicated prediction strategy.
+However, if we measure the complexity to be low, $WPE<0.85$ we can most likely do much better than the random walk and should search for more adequate prediction strategies.
+For example, consider \col (the blue +s in Figure \ref{fig:lma_vs_arima}).
+If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from 0.5 all the way to 1.0.
+That is we may think that the best we can do is twice as good as random-walk but sometimes we only do as well as random walk.
+In which case the random walk prediction strategy is probably your best bet as it is so simple and produces very similar error.
+However, if we calculate WPE we see that \col has a WPE of $0.5131\pm 0.0034$.
+This incredibly low entropy implies a great deal of deterministic structure that we can utilize for prediction, even though this did not appear to be the case with ARIMA.
+In fact using LMA on this signal we get an average MASE of $0.05\pm0.0018$, an error that is on-average 20 times better than the random walk forecast.
+Alternatively, we can look at the other end of the spectrum at \gcc.
+This signal has a WPE of $0.9430\pm0.0013$ and each prediction method we applied performed significantly worse than a simple random walk forecast (the best strategy in the case of a structure-less time series.
+Since \gcc has such little structure according to WPE it is safe to assume that he best strategy will be random walk and that continuing to search for a more accurate prediction strategy will most likely be a fruitless process.
+Many of the other signals made this jump across the $MASE=1$ line for signals with complexity less than $WPE$ of 0.85.
+Signals higher than this did not seem to make the jump.
 
-This however does not contradict our hypothesis,
-it just means that ARIMA is ill-suited for the time series under investigation.
-To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since larger
-MASE values constitute worse predictions, ARIMA performed worse than LMA in all
-traces except {\tt SVD$_1$}.
+This seems to support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
+This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
 
-The WPE is sensitive to both linear and nonlinear structure. When you have a low
-WPE and a high ARIMA it could be that the structure WPE is picking up is simply
-nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA
-prediction look really bad there is plenty of structure present as suggested by
-WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it
-into account and does bad.
+%The structure being picked up on by WPE may be linear or nonlinear.
+When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
+So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
+
+%\item Talk about structure analysis. Just because there is forward information transfer, does not mean that linear predictors can get at this.
+
+%\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by side plot.
+
+Figure \ref{fig:lma_vs_arima} also elucidates another fact: usable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
+In particular, information may be transferring from past to future through the present but because of the mechanism the underlying processe uses to process that information  (e.g., linear or nonlinear) particular prediction strategies may be blind to or not be able to efficiently utilize this information.
+For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
+If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present as is evident in Figure \ref{fig:lma_vs_arima} (b).
+In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos can interpret and utilize this type of structure just fine.
+We believe that many of the shifts in accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that linear predictors can get at this, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
+
+%[[Joshua: I am not sure the following is actually important]]
+%Upon closer examination of Figure \ref{fig:lma_vs_arima} many interesting features of the prediction error versus entropy landscape come to the forefront:
+
+%\item  full results.  Image: MASE vs. WPE for both LMA \&
+%ARIMA.  Points to make: 
+%\begin{itemize}
+%\item clusters are distributed differently
+%\item clusters are shaped differently---tight or not 
+%\item clusters move differently between LMA and ARIMA 
+%\item finally, the diagonal line is
+%important.  If you're below it, you could do better.
+
+%\end{itemize}
+
+%This however does not contradict our hypothesis,
+%it just means that ARIMA is ill-suited for the time series under %investigation.
+%To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since %larger
+%MASE values constitute worse predictions, ARIMA performed worse than LMA in %all
+%traces except {\tt SVD$_1$}.
 

### 2014-02-10 pecomplexity.tex
fixing merge errors

@@
 {\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:ipc}, and chaos places fundamental limits on
+Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
 predictability.
@@
    \label{fig:ipc}
+  \end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[t]{0.475\textwidth}
+    \includegraphics[width=0.95\textwidth]{figs/colshortts}
+    \caption{A short segment of the instructions executed per CPU clock cycle
+    (IPC) during the execution of \col. Each point is the IPC during a 100,000
+    cycle period. There is a great deal of periodicity in this time series.}
+    \label{fig:col-ts}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.475\textwidth}
+    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
+    \caption{The full IPC time series during the execution of \gcc. This time
+    series has very little structure.}
+    \label{fig:gcc-ts}
+  \end{subfigure}
+  \label{fig:sample-ts}
 \end{figure}
@@
 dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for work lengths of $l=5$ and $l=6$.
+For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.
 
@@
 In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over all 3 methods over all runs of that program) for each of the 8 signals.
+
 \begin{figure}[htbp]
@@
   SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_ent}
+  \label{fig:pred_vs_wpe}
 \end{figure}
-We find that the relationship between the two is roughly linear. The single outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE: namely
-that in the absence of any large features, as all the other regimes have, the
-WPE effectively falls back to the standard PE for which the noisy behavior
+We find that the relationship between the two is roughly linear. The single
+outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
+namely that in the absence of any large features, as all the other regimes have,
+the WPE effectively falls back to the standard PE which the noisy behavior
 drives toward 1.0.
@@
 best predictions, primarily because LMA often performed superior to ARIMA (and
-the na\"ive method). So it is not surprising that the MASE from LMA and the corresponding MASE are highly correlated. On the other hand, the ARIMA MASE values are largely
+the na\"ive method). On the other hand, the ARIMA MASE values are largely
 uncorrelated with WPE values.
+The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work.
+More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
+How do we determine if the prediction
+strategy is not adequate for the prediction task, or if the signal is simply too
+complex to predict.
+If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy.
+However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.
 
-The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as is proposed as a good strategy in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MASE>1$).
-How can we know if the prediction strategy is not adequate for the prediction task or if the signal is simply too complex and unstructured to predict.
-If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not kill ourselves over finding a more complicated prediction strategy.
-However, if we measure the complexity to be low, $WPE<0.85$ we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-For example, consider \col (the blue +s in Figure \ref{fig:lma_vs_arima}).
-If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from 0.5 all the way to 1.0.
-That is we may think that the best we can do is twice as good as random-walk but sometimes we only do as well as random walk.
-In which case the random walk prediction strategy is probably your best bet as it is so simple and produces very similar error.
-However, if we calculate WPE we see that \col has a WPE of $0.5131\pm 0.0034$.
-This incredibly low entropy implies a great deal of deterministic structure that we can utilize for prediction, even though this did not appear to be the case with ARIMA.
-In fact using LMA on this signal we get an average MASE of $0.05\pm0.0018$, an error that is on-average 20 times better than the random walk forecast.
-Alternatively, we can look at the other end of the spectrum at \gcc.
-This signal has a WPE of $0.9430\pm0.0013$ and each prediction method we applied performed significantly worse than a simple random walk forecast (the best strategy in the case of a structure-less time series.
-Since \gcc has such little structure according to WPE it is safe to assume that he best strategy will be random walk and that continuing to search for a more accurate prediction strategy will most likely be a fruitless process.
-Many of the other signals made this jump across the $MASE=1$ line for signals with complexity less than $WPE$ of 0.85.
-Signals higher than this did not seem to make the jump.
+By way of example, consider \col (the blue +s in
+Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from $\sim0.5$ to
+$\sim1.0$.  From that information we may assume that the best predictions that
+can be done are twice as good as random-walk, and in some cases predictions will
+only do as well as random-walk. In the latter such cases, the random-walk
+prediction strategy is probably the best bet as it is so simple and produces
+very similar error. However, if we calculate the WPE we see that \col has a WPE
+of $0.5131 \pm 0.0034$. This incredibly low entropy implies a great deal of
+deterministic structure which can be utilize for prediction, even though this
+did not appear to be the case with ARIMA. In fact using LMA on this signal we
+get an average MASE value of $0.05\pm0.0018$, an error that is on average 20
+times better than the random walk forecast, and 10 times better than the best
+ARIMA forcast, with little variance. Alternatively, we can look at the other end
+of the spectrum: \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each
+prediction method we applied was significantly out performed by a simple random
+walk (the best strategy in the case of structureless time series). Since \gcc
+has such little structure according to WPE it is safe to assume that he best
+strategy will be random walk and that continuing to search for a more accurate
+prediction strategy will likely be a fruitless process.
 
@@
 This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
+
+This seems to support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
+This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
+
+%%Now switch form the existence of structure as signaled by WPE to the type of structure
+
+[[Joshua: Merge the following 2 paragraphs.]]
+
+Figure \ref{fig:lma_vs_arima} also elucidates another fact: usable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
+In particular, information may be transferring from past to future through the present but because of the mechanism the underlying processe uses to process that information  (e.g., linear or nonlinear) particular prediction strategies may be blind to or not be able to efficiently utilize this information.
+For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
+If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present as is evident in Figure \ref{fig:lma_vs_arima} (b).
+In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos can interpret and utilize this type of structure just fine.
+We believe that many of the shifts in accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that linear predictors can get at this, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
+
+The WPE is sensitive to both linear and nonlinear structure. When you have a low
+WPE and a high ARIMA it could be that the structure WPE is picking up is simply
+nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA
+prediction look really bad there is plenty of structure present as suggested by
+WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it
+into account and does bad.
+
+Other noteworthy features of the LMA and ARIMA results are the cluster locations
+and distributions. The WPE values of each run any particular program tend to
+have little variance, leading to the clusters in
+Figures~\ref{fig:arima_pred_vs_wpe}, \ref{fig:lma_pred_vs_wpe} to be fairly
+constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
+low too, resulting is small, tight clusters. The ARIMA MASE values of the \col
+traces, however, have a large variance resulting in the spread seen in
+Fig~\ref{fig:arima_pred_vs_wpe}. Not only are the MASE values of that cluster
+bad, in that other predictors vastly outperform it, but they are inconsistent.
+Furthermore, since LMA can predict nonlinear behavior while ARIMA can not, we
+see that the clusters in Fig.~\ref{fig:arima_pred_vs_wpe} are mostly further to
+the right than those in Fig.~\ref{fig:lma_pred_vs_wpe}.
 
@@
 %\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by side plot.
-
-Figure \ref{fig:lma_vs_arima} also elucidates another fact: usable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferring from past to future through the present but because of the mechanism the underlying processe uses to process that information  (e.g., linear or nonlinear) particular prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present as is evident in Figure \ref{fig:lma_vs_arima} (b).
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that linear predictors can get at this, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
 

### 2014-02-10 pecomplexity.tex
moving things

@@
 
+OUTLINE:
 \begin{enumerate}
@@
 transfer, does not mean that linear predictors can get at this.
-\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
+\subitem \cmark~For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
 side plot.
@@
 \end{enumerate}
+\end{enumerate}
+
+In this section we will validate the two key findings of this work:
+
+[[Rephrase these]]
+\begin{enumerate}
+
+\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
+
+\item The way structure/information/complexity is processed internally by a given process plays a role in predictability.
+
 \end{enumerate}
@@
 
-This seems to support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-
 %%Now switch form the existence of structure as signaled by WPE to the type of structure

### 2014-02-10 pecomplexity.tex
adding outines

--- 2014-02-10 pecomplexity.tex
@@
 %\end{it}
+Things to add to introduction
+\begin{enumerate}
+\item Differnt kinds of complexity exist in time series and this makes choosing prediction models difficult 
+\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
+\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is often impossible.
+
+\item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc
+
+\item Introduce the two main contributions of the paper which are outlined at the begining of the results section
+
+\end{enumerate}
 
@@
 \end{enumerate}
+\end{enumerate}
+
+In this section we will validate the two key findings of this work:
+
+[[Rephrase these]]
+\begin{enumerate}
+
+\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)
+
+\subitem WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
+
+%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
+
+\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
+\subitem We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
+The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
+
+\subitem This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
+
 \end{enumerate}

### 2014-02-10 pecomplexity.tex
adding experimental methods and intro outlines

@@
 \begin{enumerate}
-\item Differnt kinds of complexity exist in time series and this makes choosing prediction models difficult 
-\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
-\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is often impossible.
+\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
+\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.
 
 \item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc
+
+\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
+\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
+\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex.
+Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not.
+Maybe even having 4 predictions.
+top being ARIMA of the above signals and bottom being LMA of the above signals.
+Show that one improved and one did not.
+Is it that we used the wrong method to predict or is it that we simply can't predict the signal better than a random walk due to high levels of internal signal complexity.
 
@@
 
-To Add to this section:
+Section outline:
 
 \begin{enumerate}
-\item Description of ARIMA
+\item Experimental methods (how we collect the time series and what the times series are
+\item Description of DCE and parameter estimation  
+\item Description of auto ARIMA 
+\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
 \item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
-
-\item Maybe add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the begining of the results.
+\item Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
 

### 2014-02-13 pecomplexity.tex
editing

@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 %\begin{it}
@@
 
+\begin{figure}[htbp]
+  \centering
+
+  \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
+    \caption{\gcc ARIMA }
+    \label{fig:gccARIMA}
+  \end{subfigure}%
+    \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
+    \caption{\gcc LMA}
+    \label{fig:gccLMA}
+  \end{subfigure}
+  \\
+  \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
+    \caption{\col ARIMA}
+    \label{fig:colARIMA}
+  \end{subfigure}
+    \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/colLMAForecast}
+    \caption{\col LMA}
+    \label{fig:colLMA}
+  \end{subfigure}
+  %\begin{subfigure}{0.5\textwidth}
+  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
+   \caption{
+For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
+On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
+}\label{fig:gcc_vs_col}
+  %\end{subfigure}
+\end{figure}
+
 The computer systems community has applied a variety of prediction
@@
 
-\section{Modeling Computer Performance}\label{sec:compModel}
+\section{Experimental Methods and Modelling}\label{sec:compModel}
 
@@
 % took out for space
-% \subsection{Reconstructing hidden dynamics}
+ \subsection{Reconstructing hidden dynamics}
 
@@
 % took out for space
-% \subsection{LMA: Using dynamics in forecasting}
+ \subsection{LMA: Using dynamics in forecasting}
 
@@
 %% entropy to explore that conjecture.]]
+
+\subsection{Comparing Prediction Accuracy}
+In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
+The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
+The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
+In particular, we perform $k$ 1-step predictions, after each 1-step prediction we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
+This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.}
+
+As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
+%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
+The scaling term for MASE:
+$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
+is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
+This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
+When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
+Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
+In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
+we give mean $\pm$ std.
+dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
+For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.
+
+\begin{table}[htdp]
+\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
+\hline
+\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
+
+\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
+\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
+\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
+\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
+\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
+\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
+\hline
+\end{tabular}
+\end{center}
+\label{tab:error}
+\end{table}%
 
@@
   %\\
-  \begin{subfigure}{0.3\textwidth}
-    \includegraphics[width=0.3\textwidth]{figs/SVD_wwpe}
+  \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
     \caption{The MASE of ARIMA vs weighted permutation entropy. }
@@
 
- \section{Results} % (fold)
+ \section{Validation of Primary Findings} % (fold)
  \label{sec:results}
@@
 
-In this section we will validate the two key findings of this work:
+As we stated in Section \ref{sec:intro} we intend to show two primary contributions of this work:
 
-[[Rephrase these]]
+[[Joshua: Probably need to rephrase these]]
 \begin{enumerate}
@@
 \item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
+
 \subitem We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
@@
 
-%\item Introduce MASE and explain why random walk isn't *that* naive
-%\item Paragraph on how to interpret MASE
+ In this section we will validate these two key findings.
 
-%As a figure of merit we calculate the Mean Absolute Squared Error (MASE) for each forecast.
+%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
+%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
+%%%%%%%%%%%%%%%%%%%%%%%%%
 
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ 1-step predictions, after each 1-step prediction we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.}
-
-As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
-When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.
-
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures]] }
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
+%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
+%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
+%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2014-02-14 pecomplexity.tex
editing results more

@@
 Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\On Quantifying Predictability through Structural Complexity
-\\On Quantifying Predictability with Structural Complexity
+\\ Quantifying Predictability through Structural Complexity
+\\ Quantifying Predictability using Structural Complexity
 }
@@
 
-As we stated in Section \ref{sec:intro} we intend to show two primary contributions of this work:
+BRAINSTORMING 
+\begin{itemize}
+\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
+\item random walk for example is best predicted by guess what just happend[[move to introduction]]
+\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
+\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
+[[talk about again here but justify in information theory section]]
+%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
 
-[[Joshua: Probably need to rephrase these]]
+\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
+And a constant, linear and nonlinear prediction strategy all fail.
+We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
+
+\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
+
+\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
+That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
+This gives a practitioner leverage on when to give up and when to keep working.
+[[use in this section as bridge to next section]]
+
+\end{itemize}
+
+In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:
+
 \begin{enumerate}
 
-\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)
+\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.
 
-\subitem WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
+\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
 
@@
 
-\subitem We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
+\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
 The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
 
-\subitem This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
+\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
 
 \end{enumerate}
-
- In this section we will validate these two key findings.
 
@@
 
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
 %\item First paragraph: WPE is a good measure of predictability.  Figure:
@@
 
-As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
-indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
-the exception of one outlier whose behavior we will explain.
-In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over all 3 methods over all runs of that program) for each of the 8 signals.
+%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
+[[talk about again here but justify in information theory section]]
 
+As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is very non-trivial.
+As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
+Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.
+
+In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation.
 \begin{figure}[htbp]
@@
 \end{figure}
+What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}.
+We will not further validate this claim through some examples.
+
+To further validate this finding we present an in-depth analysis of three examples (\col, \svd$_5$ and \gcc ) which nicely cover the range of complexity from low to high respectively.
+In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
+
+\begin{figure}[htbp]
+  \centering
+
+  \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
+    \caption{The MASE of LMA vs weighted permutation entropy. }
+    \label{fig:lma_pred_vs_ent}
+  \end{subfigure}%
+  \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
+    \caption{The MASE of ARIMA vs weighted permutation entropy. }
+    \label{fig:arima_pred_vs_ent}
+  \end{subfigure}
+\caption{
+For each of these, the word length used is $6$.
+The MASE values for LMA against ARIMA.
+The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
+All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}
+\end{figure}
+
+%\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
+And a constant, linear and nonlinear prediction strategy all fail.
+We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
+
+%\item \also do \svd$_5$
+
+%\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
+
 We find that the relationship between the two is roughly linear. The single
@@
 
-\begin{itemize}
-\item Complexity need not be hard to predict (can point at the simple predictions paper)
-\item random walk for example is best predicted by guess what just happend
-\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not. 
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
-
-\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE
-
-\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction
-
-\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-
-\end{itemize}
-
 Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}.
-[[Joshua: Make sure we make it clear WPE not only tells you when there is strucure to use but also what kinds of complexity are usable.
+[[Joshua: This is not really true ...
+at least by these definitions because it doesn't really explain "hard to predict" ...Make sure we make it clear WPE not only tells you when there is strucure to use but also what kinds of complexity are usable.
 e.g., white noise and deterministic chaos are both complex but one is predictable.
@@
 Talk about chaos and random walk being complex but from a predictive standpoint that complexity spectrum is important **AND** quantifiable by WPE ]]
-\begin{figure}[htbp]
-  \centering
-       \caption{
-For each of
-    these, the word length used is $6$.       
-       The MASE values for LMA against ARIMA. The dashed line is the
-    identity, delineating the traces for which either LMA or ARIMA performed
-   better. All traces except those from {\tt SVD$_1$} lie above the line,
-  indicating that LMA is better suited prediction method for the %computer
-   traces considered.}\label{fig:lma_vs_arima}  
-  \begin{subfigure}{\textwidth}
-    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. }
-    \label{fig:lma_pred_vs_ent}
-  \end{subfigure}%
-  \\
-  \begin{subfigure}{\textwidth}
-    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:arima_pred_vs_ent}
-  \end{subfigure}
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   % \caption{The MASE values for LMA against ARIMA. The dashed line is the
-   % identity, delineating the traces for which either LMA or ARIMA %performed
-%    better. All traces except those from {\tt SVD$_1$} lie above the line,
-%    indicating that LMA is better suited prediction method for the %computer
-%   traces considered.}\label{fig:lma_vs_arima}
 
-  %\end{subfigure}
-\end{figure} 
 In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
@@
 
-%%Now switch form the existence of structure as signaled by WPE to the type of structure
+%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
+%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
+%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2014-02-14 pecomplexity.tex
editing

@@
 
+We will start with the lowest complexity program: \col.
+Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{toddchaos}.
+If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
+This would suggest that there is some predictive structure present but not that much.
+However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive strucutre or equivalently the low complexity.
+This is in fact the case.
+\col is a complex signal, i.e., chaotic, but it is complex in a structured way.
+Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
+The existence of predictve structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
+
+Maybe even more conniving than a low complexity signal like \col is a moderately complex signal like \svd$_5$.
+This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the exsitence of a great deal of predictive structure.
+However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
+The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
+This may be discouragin to someone using an out of the box predictor that the signal is too complex to predict.
+However by \emph{knowing} that there exists predictable strucutre in this signal we can try to use a more advanced technique, in particular LMA.
+If we do this we get forecasts which score 1.5 times better than the random walk.
+Showing that this usable structure (althought hidden to the linear and constant method) was quantifiable by WPE.
+
+Finally, we consider the most complex program: \gcc.
+
 %\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.

### 2014-02-15 pecomplexity.tex
editing results

@@
 \newcommand{\svd}{{\tt dgesdd}~}
+\newcommand{\svdone}{{\tt dgesdd}$_1$~}
+\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
+\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
+\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
+\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
+\newcommand{\svdsix}{{\tt dgesdd$_6$}~}
+
 \newcommand{\row}{{\tt row\_major}~}
@@
 
-\section{Measuring Complexity}\label{sec:meaComplex}
+\section{Measuring Structural Complexity}\label{sec:meaComplex}
 
@@
 
-Maybe even more conniving than a low complexity signal like \col is a moderately complex signal like \svd$_5$.
-This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the exsitence of a great deal of predictive structure.
+Maybe even more conniving than a low complexity signal like \col is a moderately complex signal like \svdfive.
+This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
 However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
 The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
-This may be discouragin to someone using an out of the box predictor that the signal is too complex to predict.
-However by \emph{knowing} that there exists predictable strucutre in this signal we can try to use a more advanced technique, in particular LMA.
+This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict.
+However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA.
 If we do this we get forecasts which score 1.5 times better than the random walk.
-Showing that this usable structure (althought hidden to the linear and constant method) was quantifiable by WPE.
+Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.
 
 Finally, we consider the most complex program: \gcc.
+The complexity of \gcc is $0.94 \pm 0.001$.
+This implies that almost no predictive structure exists in this structure.
+Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future.
+This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex.
+In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step.
+This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.
+
+There is a single outlier in this study which is \svdone.
+This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively).
+The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts.
+We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method.
+MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
+Random walk is then off by the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
+A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure.
+With \svdone
 

### 2014-02-15 pecomplexity.tex
adding explanation of outlier

@@
 
-As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is very non-trivial.
+As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task.
 As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
@@
 MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
-Random walk is then off by the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
-A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure.
-With \svdone
+In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
+A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.
 
-%\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
+To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail.
+This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
+We provide a small snippet because the full time series is so dense the visual loses the structure.
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=\textwidth]{figs/svdonets2}
+\caption{ For each of these, the word length used is $6$.
+The MASE values for LMA against ARIMA.
+The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
+All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}
+\end{figure} 
+As can be seen in Figure \ref{fig:svdone-ts},\svdone has a highly unstructured small band which is about 0.05 instructions per cycle wide.
+Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
+The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}.
+These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle).
+As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
+At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the original signal, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the signals true width.
+This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
+The combination of a band of noise with a high number of random drops that this signal exhibits is a weakness of this method which cause this one outlier.
 
-%\item \also do \svd$_5$
+This seems to support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
+This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
 
-%\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
+%We find that the relationship between the two is roughly %linear. The single
+%outlier, \svd regime 1, does not fit the trend due to a %weakness in the WPE:
+%namely that in the absence of any large features, as all the %other regimes have,
+%the WPE effectively falls back to the standard PE which the %noisy behavior
+%drives toward 1.0.
 
-We find that the relationship between the two is roughly linear. The single
-outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
-namely that in the absence of any large features, as all the other regimes have,
-the WPE effectively falls back to the standard PE which the noisy behavior
-drives toward 1.0.
-
-It may be tempting to paraphrase this primary finding as: ``Signals that are more complex are harder to predict.", but this truly misses the impact of this work.
+%It may be tempting to paraphrase this primary finding as: ``Signals that are more complex are harder to predict.", but this truly misses the impact of this work.
 A more correct statement would be ``Time series which exhibit structure-less complexity are harder to predict for any model which uses structure as it's information processor.
@@
 
-Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}.
+%Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}.
 [[Joshua: This is not really true ...
@@
 prediction strategy will likely be a fruitless process.
-
-This seems to support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
 

### 2014-02-15 pecomplexity.tex
cleaning

@@
 
-Maybe even more conniving than a low complexity signal like \col is a moderately complex signal like \svdfive.
+Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive.
 This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
@@
 This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
-We provide a small snippet because the full time series is so dense the visual loses the structure.
+We provide a small snippet as opposed to the full time series simply for visual clarity.
 \begin{figure}[htbp]
@@
 \end{figure} 
-As can be seen in Figure \ref{fig:svdone-ts},\svdone has a highly unstructured small band which is about 0.05 instructions per cycle wide.
+As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide.
 Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
@@
 As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
-At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the original signal, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the signals true width.
+At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal.
 This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
-The combination of a band of noise with a high number of random drops that this signal exhibits is a weakness of this method which cause this one outlier.
+The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.
 
-This seems to support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
+The results we just presented support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
 This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
+These findings also bring up an interesting point that this work seems to suggest.
+The way that structure, information and complexity is processed by a generating process plays a crucial role in the success of a given prediction scheme.
 
-%We find that the relationship between the two is roughly %linear. The single
-%outlier, \svd regime 1, does not fit the trend due to a %weakness in the WPE:
-%namely that in the absence of any large features, as all the %other regimes have,
-%the WPE effectively falls back to the standard PE which the %noisy behavior
-%drives toward 1.0.
-
-%It may be tempting to paraphrase this primary finding as: ``Signals that are more complex are harder to predict.", but this truly misses the impact of this work.
-A more correct statement would be ``Time series which exhibit structure-less complexity are harder to predict for any model which uses structure as it's information processor.
-" [[I don't like this at all]]
-
-%Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}.
-[[Joshua: This is not really true ...
-at least by these definitions because it doesn't really explain "hard to predict" ...Make sure we make it clear WPE not only tells you when there is strucure to use but also what kinds of complexity are usable.
-e.g., white noise and deterministic chaos are both complex but one is predictable.
-The kind of complexity is important.
-Talk about chaos and random walk being complex but from a predictive standpoint that complexity spectrum is important **AND** quantifiable by WPE ]]
+%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
+%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
+%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-By way of example, consider \col (the blue +s in
-Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from $\sim0.5$ to
-$\sim1.0$.  From that information we may assume that the best predictions that
-can be done are twice as good as random-walk, and in some cases predictions will
-only do as well as random-walk. In the latter such cases, the random-walk
-prediction strategy is probably the best bet as it is so simple and produces
-very similar error. However, if we calculate the WPE we see that \col has a WPE
-of $0.5131 \pm 0.0034$. This incredibly low entropy implies a great deal of
-deterministic structure which can be utilize for prediction, even though this
-did not appear to be the case with ARIMA. In fact using LMA on this signal we
-get an average MASE value of $0.05\pm0.0018$, an error that is on average 20
-times better than the random walk forecast, and 10 times better than the best
-ARIMA forcast, with little variance. Alternatively, we can look at the other end
-of the spectrum: \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each
-prediction method we applied was significantly out performed by a simple random
-walk (the best strategy in the case of structureless time series). Since \gcc
-has such little structure according to WPE it is safe to assume that he best
-strategy will be random walk and that continuing to search for a more accurate
-prediction strategy will likely be a fruitless process.
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-%%%%%%%%%%%%%%%%%%%%%%%%%
+%By way of example, consider \col (the blue +s in
+%Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from $%\sim0.5$ to
+%$\sim1.0$.  From that information we may assume that the best %predictions that
+%can be done are twice as good as random-walk, and in some %cases predictions will
+%only do as well as random-walk. In the latter such cases, the %random-walk
+%prediction strategy is probably the best bet as it is so %simple and produces
+%very similar error. However, if we calculate the WPE we see %that \col has a WPE
+%of $0.5131 \pm 0.0034$. This incredibly low entropy implies a %great deal of
+%deterministic structure which can be utilize for prediction, %even though this
+%did not appear to be the case with ARIMA. In fact using LMA %on this signal we
+%get an average MASE value of $0.05\pm0.0018$, an error that %is on average 20
+%times better than the random walk forecast, and 10 times %better than the best
+%ARIMA forcast, with little variance. Alternatively, we can %look at the other end
+%of the spectrum: \gcc. This signal has a WPE of %$0.9430\pm0.0013$ and each%
+%prediction method we applied was significantly out performed %by a simple random
+%walk (the best strategy in the case of structureless time %series). Since \gcc
+%has such little structure according to WPE it is safe to %assume that he best
+%strategy will be random walk and that continuing to search %for a more accurate
+%prediction strategy will likely be a fruitless process.
 

### 2014-02-15 pecomplexity.tex
finishing results and cleaning up

@@
 
-To further validate this finding we present an in-depth analysis of three examples (\col, \svd$_5$ and \gcc ) which nicely cover the range of complexity from low to high respectively.
+To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively.
 In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
@@
 We will start with the lowest complexity program: \col.
-Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{toddchaos}.
+Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}.
 If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
 This would suggest that there is some predictive structure present but not that much.
-However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive strucutre or equivalently the low complexity.
+However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity.
 This is in fact the case.
@@
 Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
-The existence of predictve structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
+The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
 
@@
 
-The results we just presented support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
+[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
 This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-These findings also bring up an interesting point that this work seems to suggest.
-The way that structure, information and complexity is processed by a generating process plays a crucial role in the success of a given prediction scheme.
+The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.
 
@@
 
-In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
-compare the performance of the LMA and ARIMA prediction methods (respectively)
-to the value of the weighted permutation entropy for all runs of each program
-under consideration. The LMA MASE values are largely similar to those of the
-best predictions, primarily because LMA often performed superior to ARIMA (and
-the na\"ive method). On the other hand, the ARIMA MASE values are largely
-uncorrelated with WPE values.
-The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work.
+Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
+In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information.
+For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
+If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}.
+In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine.
+We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
+
+%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration.
+The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method).
+On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values.
+As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE.
+The WPE is sensitive to both linear and nonlinear structure.
+When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
+So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
+
+%is in large part one of the major findings of this work.
 More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
-How do we determine if the prediction
-strategy is not adequate for the prediction task, or if the signal is simply too
-complex to predict.
+How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict.
 If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy.
@@
 
-%By way of example, consider \col (the blue +s in
-%Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA from \cite{autoArima} we get MASE ranging from $%\sim0.5$ to
-%$\sim1.0$.  From that information we may assume that the best %predictions that
-%can be done are twice as good as random-walk, and in some %cases predictions will
-%only do as well as random-walk. In the latter such cases, the %random-walk
-%prediction strategy is probably the best bet as it is so %simple and produces
-%very similar error. However, if we calculate the WPE we see %that \col has a WPE
-%of $0.5131 \pm 0.0034$. This incredibly low entropy implies a %great deal of
-%deterministic structure which can be utilize for prediction, %even though this
-%did not appear to be the case with ARIMA. In fact using LMA %on this signal we
-%get an average MASE value of $0.05\pm0.0018$, an error that %is on average 20
-%times better than the random walk forecast, and 10 times %better than the best
-%ARIMA forcast, with little variance. Alternatively, we can %look at the other end
-%of the spectrum: \gcc. This signal has a WPE of %$0.9430\pm0.0013$ and each%
-%prediction method we applied was significantly out performed %by a simple random
-%walk (the best strategy in the case of structureless time %series). Since \gcc
-%has such little structure according to WPE it is safe to %assume that he best
-%strategy will be random walk and that continuing to search %for a more accurate
-%prediction strategy will likely be a fruitless process.
-
-[[Joshua: Merge the following 2 paragraphs.]]
-
-Figure \ref{fig:lma_vs_arima} also elucidates another fact: usable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferring from past to future through the present but because of the mechanism the underlying processe uses to process that information  (e.g., linear or nonlinear) particular prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present as is evident in Figure \ref{fig:lma_vs_arima} (b).
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that linear predictors can get at this, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
-
-The WPE is sensitive to both linear and nonlinear structure. When you have a low
-WPE and a high ARIMA it could be that the structure WPE is picking up is simply
-nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA
-prediction look really bad there is plenty of structure present as suggested by
-WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it
-into account and does bad.
-
 Other noteworthy features of the LMA and ARIMA results are the cluster locations
-and distributions. The WPE values of each run any particular program tend to
+and distributions. The WPE values of each run of any particular program tend to
 have little variance, leading to the clusters in
-Figures~\ref{fig:arima_pred_vs_wpe}, \ref{fig:lma_pred_vs_wpe} to be fairly
+Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
 constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
-low too, resulting is small, tight clusters. The ARIMA MASE values of the \col
+low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
 traces, however, have a large variance resulting in the spread seen in
-Fig~\ref{fig:arima_pred_vs_wpe}. Not only are the MASE values of that cluster
+Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
 bad, in that other predictors vastly outperform it, but they are inconsistent.
-Furthermore, since LMA can predict nonlinear behavior while ARIMA can not, we
-see that the clusters in Fig.~\ref{fig:arima_pred_vs_wpe} are mostly further to
-the right than those in Fig.~\ref{fig:lma_pred_vs_wpe}.
-
-%The structure being picked up on by WPE may be linear or nonlinear.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
-So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
-
-%\item Talk about structure analysis. Just because there is forward information transfer, does not mean that linear predictors can get at this.
-
-%\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by side plot.
-
-%[[Joshua: I am not sure the following is actually important]]
-%Upon closer examination of Figure \ref{fig:lma_vs_arima} many interesting features of the prediction error versus entropy landscape come to the forefront:
-
-%\item  full results.  Image: MASE vs. WPE for both LMA \&
-%ARIMA.  Points to make: 
-%\begin{itemize}
-%\item clusters are distributed differently
-%\item clusters are shaped differently---tight or not 
-%\item clusters move differently between LMA and ARIMA 
-%\item finally, the diagonal line is
-%important.  If you're below it, you could do better.
-
-%\end{itemize}
-
-%This however does not contradict our hypothesis,
-%it just means that ARIMA is ill-suited for the time series under %investigation.
-%To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since %larger
-%MASE values constitute worse predictions, ARIMA performed worse than LMA in %all
-%traces except {\tt SVD$_1$}.
-
- % section results (end)
-
-% In Table~\ref{tab:TCMpe} \& \ref{tab:IPCpe} the behavior of the PEs as
-% they limit to their asymptotic values is observed. A feature of note
-% is that some of the signals drop off and some do not.  For example at
-% $m=4$ {\tt col\_major} is at 0.8356, a fairly high entropy, however as
-% $m$ is increased to 6 the PE plummets to 0.6880: a value consistent
-% with a chaotic time series. Notice this also corresponds to a very low
-% nRMSE. In contrast, consider the permutation entropy of the processor
-% load of {\tt 482.sphinx3}: with $m=4$ it is 0.9951 and only drops to
-% 0.9849 at $m=6$. This is consistent with the signal also having the
-% highest nRMSE.
+Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
+see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
+the right than those in Figure~\ref{fig:lma_pred_vs_ent}.
 
 \section{ Conclusions \& Future Work}\label{sec:conc}
+
+[[Joshua: These need to be rewritten to conclude and address this paper...]]
 
@@
 
-\section{New Figures and Tables}
-
 \section*{Acknowledgment}

### 2014-02-15 pecomplexity.tex
cleaning and revising

@@
 \section{Introduction}\label{sec:intro}
+
+BRAINSTORMING 
+\begin{itemize}
+\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
+\item random walk for example is best predicted by guess what just happend[[move to introduction]]
+\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
+\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
+[[talk about again here but justify in information theory section]]
+%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
+
+\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
+And a constant, linear and nonlinear prediction strategy all fail.
+We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
+
+\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
+
+\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
+That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
+This gives a practitioner leverage on when to give up and when to keep working.
+[[use in this section as bridge to next section]]
+
+\end{itemize}
+
 %\begin{it}
@@
 The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ 1-step predictions, after each 1-step prediction we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.}
+In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
+This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
 
@@
 \section{Measuring Structural Complexity}\label{sec:meaComplex}
+
+THINGS TO DO
+\begin{enumerate}
+\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
+\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
+\item justify and explain the regime split in \svd 
+\end{enumerate}
 
@@
   \centering
+
+  %\begin{subfigure}{0.3\textwidth}
+  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
+  %  \caption{The instructions per cycle of \svd.
+Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
+From left to right each change in color represents a change in regime for 6 regimes in total.
+}
+ %   \label{fig:svd_ts}
+ % \end{subfigure}%
+  %\\
+ % \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
+    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
   \caption{
@@
     are regions where the window overlaps regimes. The window size used is
-    $5,000 \times 100,000$ instructions and the word length is $4$.}\label{fig:wwpe}
-  \begin{subfigure}{0.3\textwidth}
-    \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
-    \caption{The instructions per cycle of \svd.
-Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
-From left to right each change in color represents a change in regime for 6 regimes in total.
-}
-    \label{fig:svd_ts}
-  \end{subfigure}%
-  %\\
-  \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:svd_wwpe}
-  \end{subfigure}
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   % \caption{The MASE values for LMA against ARIMA. The dashed line is the
-   % identity, delineating the traces for which either LMA or ARIMA %performed
-%    better. All traces except those from {\tt SVD$_1$} lie above the line,
-%    indicating that LMA is better suited prediction method for the %computer
-%   traces considered.}\label{fig:lma_vs_arima}
-  %\end{subfigure}
+    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
+Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE.
+From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
 \end{figure}
@@
 
-OUTLINE:
-\begin{enumerate}
-\item \cmark~Introduce MASE.
-\item \cmark~WPE is a good measure of predictability.  Figure: best athlete MASE vs.
-its WPE.
-\item \cmark~Talk about structure analysis. Just because there is forward information
-transfer, does not mean that linear predictors can get at this.
-\subitem \cmark~For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
-side plot.
-\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
-\begin{enumerate}
-\item \cmark~clusters are distributed differently
-\item \cmark~clusters are shaped differently---tight or not
-\item \cmark~clusters move differently between LMA and ARIMA
-\item finally, the diagonal line is important. If you're below it, you could do
-better.
-\end{enumerate}
-\end{enumerate}
+%OUTLINE:
+%\begin{enumerate}
+%\item \cmark~Introduce MASE.
+%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
+%its WPE.
+%\item \cmark~Talk about structure analysis. Just because %there is forward information
+%transfer, does not mean that linear predictors can get at %this.
+%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
+%side plot.
+%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
+%\begin{enumerate}
+%\item \cmark~clusters are distributed differently
+%\item \cmark~clusters are shaped differently---tight or not
+%\item \cmark~clusters move differently between LMA and ARIMA
+%\item finally, the diagonal line is important. If you're %below it, you could do
+%better.
+%\end{enumerate}
+%\end{enumerate}
 
-BRAINSTORMING 
-\begin{itemize}
-\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
-\item random walk for example is best predicted by guess what just happend[[move to introduction]]
-\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
+%BRAINSTORMING 
+%\begin{itemize}
+%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
+%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
 [[talk about again here but justify in information theory section]]
@@
 
-\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
+%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
 And a constant, linear and nonlinear prediction strategy all fail.
@@
 
-\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
+%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
 
-\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
+%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
 That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
@@
 
-\end{itemize}
+%\end{itemize}
 
@@
 
-\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
+%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
 
@@
 
-\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
+%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
 The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
 
-\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
+%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
 

### 2014-02-16 pecomplexity.tex
new edits

@@
 \begin{abstract}
-  Computers are deterministic dynamical systems \cite{mytkowicz09}.
+[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
   Among other things, that implies that one should be able to use
@@
   processor loads of some simple programs are easy to predict, for
-  example, but those of more-complex programs like {\tt gcc} are not.
+  example, but those of more-complex programs like \gcc are not.
 %%%%%%%%%%%%%%%%%%%%%%
@@
 \item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]]
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
-
-\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
-
-\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
 
@@
 \item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
+
 \item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
+\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
+[[talk about again here but justify in information theory section]]
 \item justify and explain the regime split in \svd 
@@
 
-[[Joshua: These need to be rewritten to conclude and address this paper...]]
+\begin{enumerate}
+\item  This need to be rewritten to conclude and address this paper...
+\end{enumerate}
 

### 2014-02-17 unusedFigsTables.tex
Cleaning up some figures and fixing gitignore

@@
         \centering
-     \includegraphics[width=0.5\textwidth]{unused-figs/3dcolcache}
-     \caption{A 3D projection of a delay-coordinate embedding of the trace
- from Figure~\ref{fig:cache} with a delay ($\tau$) of 100,000 instructions.
- }
- \label{fig:embedding}
- \end{figure}
-
-\begin{figure}
-        \centering
         \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdipcfull}
+                \includegraphics[width=\textwidth]{unused-figs/svdipcfull}
                 \caption{SVD IPC without coloring}
@@
         \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdipcregimescolored}
+                \includegraphics[width=\textwidth]{unused-figs/svdipcregimescolored}
                 \caption{SVD IPC with coloring}
@@
         \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdregime2prediction}
+                \includegraphics[width=\textwidth]{unused-figs/svdregime2prediction}
                 \caption{SVD Regime 2 LMA Prediction}
@@
         \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdregime2predictionFULL}
+                \includegraphics[width=\textwidth]{unused-figs/svdregime2predictionFULL}
                 \caption{SVD Regime 2 LMA Prediction with Full Time Series}
@@
         \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdPrediction}
+                \includegraphics[width=\textwidth]{unused-figs/svdPrediction}
                 \caption{SVD Regime 5 LMA Prediction}
@@
         \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/svdPredictionFull}
+                \includegraphics[width=\textwidth]{unused-figs/svdPredictionFull}
                 \caption{SVD Regime 5 LMA Prediction with Full Time Series}
@@
         \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colfullts}
+                \includegraphics[width=\textwidth]{unused-figs/colfullts}
                 \caption{col IPC  Full Time Series}
@@
         \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colshortts}
+                \includegraphics[width=\textwidth]{unused-figs/colshortts}
                 \caption{col IPC  Short Time Series for structure illustration}
@@
         \begin{subfigure}{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colPredShortTS}
+                \includegraphics[width=\textwidth]{unused-figs/colPredShortTS}
                 \caption{col IPC  Full Time Series}
@@
         \begin{subfigure}[b]{\textwidth}
-                \includegraphics[width=\textwidth]{figs/colPredfullTS}
+                \includegraphics[width=\textwidth]{unused-figs/colPredfullTS}
                 \caption{col IPC  Short Time Series for structure illustration}
@@
 
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=\textwidth]{unused-figs/%colPredictionVariation.pdf}
+%\end{center}
+%\end{figure}
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics[width=\textwidth]{figs/gccfullts}\caption{403.gcc full time series}
+\includegraphics[width=\textwidth]{unused-figs/gccfullts}\caption{403.gcc full time series}
 \label{default}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figs/regime1}\caption{SVD IPC Regime 1 Time Series}
+\includegraphics[width=\textwidth]{unused-figs/regime1}\caption{SVD IPC Regime 1 Time Series}
 \label{default}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figs/regime2}\caption{SVD IPC Regime 2 Time Series}
+\includegraphics[width=\textwidth]{unused-figs/regime2}\caption{SVD IPC Regime 2 Time Series}
 \label{default}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figs/regime5}\caption{SVD IPC Regime 5 Time Series}
+\includegraphics[width=\textwidth]{unused-figs/regime5}\caption{SVD IPC Regime 5 Time Series}
 \label{default}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figs/regime6}\caption{SVD IPC Regime 6 Time Series}
+\includegraphics[width=\textwidth]{unused-figs/regime6}\caption{SVD IPC Regime 6 Time Series}
 \label{default}

### 2014-02-17 pecomplexity.tex
adding meeting notes

@@
   %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{
+   \caption{[[move this to after the 3 forecast have been described and add the same plot for mean prediction]] 
 For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
@@
 
-\section{Experimental Methods and Modelling}\label{sec:compModel}
+\section{Experimental Methods }
 
+This should be HPM PAPI, which programs we model
+
+Include traces of full \col and \gcc
+
+\section{MODELING}
+\label{sec:compModel}
+
+\subsection{LMA}
+\subsection{ARIMA}
+\subsection{Naive: Random Walk and mean}
+\subsection{Prediction Accuracy}
+MASE
 Section outline:
@@
 \begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
+\caption{[[Move this table to results, and cherry pick a few examples of \gcc and \col to put in the text and chop of $l=5$]] MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
 \begin{center}

### 2014-02-18 pecomplexity.tex
Reorganize after meeting and new figures

--- 2014-02-16 pecomplexity.tex
@@
 
-\begin{figure}[htbp]
-  \centering
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/colshortts}
-    \caption{A short segment of the instructions executed per CPU clock cycle
-    (IPC) during the execution of \col. Each point is the IPC during a 100,000
-    cycle period. There is a great deal of periodicity in this time series.}
-    \label{fig:col-ts}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
-    \caption{The full IPC time series during the execution of \gcc. This time
-    series has very little structure.}
-    \label{fig:gcc-ts}
-  \end{subfigure}
-  \label{fig:sample-ts}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
-    \caption{\gcc ARIMA }
-    \label{fig:gccARIMA}
-  \end{subfigure}%
-    \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
-    \caption{\gcc LMA}
-    \label{fig:gccLMA}
-  \end{subfigure}
-  \\
-  \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
-    \caption{\col ARIMA}
-    \label{fig:colARIMA}
-  \end{subfigure}
-    \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colLMAForecast}
-    \caption{\col LMA}
-    \label{fig:colLMA}
-  \end{subfigure}
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{
-For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
-On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
-}\label{fig:gcc_vs_col}
-  %\end{subfigure}
-\end{figure}
-
 The computer systems community has applied a variety of prediction
@@
 
-\section{Experimental Methods and Modelling}\label{sec:compModel}
+\section{Experimental Methods }
 
+\subsection{Time Series Collection}
+ Experimental methods (how we collect the time series and what the times series are
+This should be HPM PAPI, which programs we model
+
+\subsection{The Programs}
+Include traces of full \col \svd and \gcc
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[t]{0.475\textwidth}
+    \includegraphics[width=0.95\textwidth]{figs/colFullTS}
+    \caption{\col Time Series}
+    \label{fig:col-ts}
+  \end{subfigure}%
+  \begin{subfigure}[t]{0.475\textwidth}
+    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
+    \caption{\gcc Time Series}
+    \label{fig:gcc-ts}
+  \end{subfigure}
+  \caption{In (a) the instructions executed per CPU clock cycle
+    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
+    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
+\end{figure}
+
+\section{Modeling}
 Section outline:
@@
 \begin{enumerate}
-\item Experimental methods (how we collect the time series and what the times series are
+
 \item Description of DCE and parameter estimation  
@@
 \item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
-\item Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
+\item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
 
@@
 
-% took out for space
- \subsection{Reconstructing hidden dynamics}
+\label{sec:compModel}
+
+\subsection{Lorenz Method of Analogues (LMA)}
+TODO:
+\begin{enumerate}
+\item update examples
+\item remove section on RMSE
+\item update figures
+\end{enumerate}
+
+ \subsubsection{Reconstructing hidden dynamics}
 
@@
      \includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{A 3D projection of a delay-coordinate embedding of the trace
+     \caption{[[Joshua:Perhaps add a similar graphic of \gcc to show contrast]]A 3D projection of a delay-coordinate embedding of the trace
  from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
@@
 The coordinates of each point on this plot are differently delayed
-elements of the \verb|col_major| L2 cache miss rate time series
+elements of the \col L2 cache miss rate time series
 $y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
@@
 % took out for space
- \subsection{LMA: Using dynamics in forecasting}
+ \subsubsection{LMA: Using dynamics in forecasting}
 
@@
 %
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
-    \caption{A forecast of the last 4,000 points of the signal in
-      Figure~\ref{fig:ipc} using an LMA-based strategy on the
-      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
-      $\times$s are the true and predicted values, respectively;
-      vertical bars show where these values differ. }
-\label{fig:cachePredTS}
-\end{figure}
+%\begin{figure}[htbp]
+%  \centering
+%    \includegraphics[width=0.5\textwidth]{figs/%colCachePredTS}
+%    \caption{A forecast of the last 4,000 points of the %signal in
+%      Figure~\ref{fig:ipc} using an LMA-based strategy on the
+%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
+%      $\times$s are the true and predicted values, %respectively;
+%      vertical bars show where these values differ. }
+%\label{fig:cachePredTS}
+%\end{figure}
 %
@@
 %
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
-     \caption{An LMA-based forecast of the last 4,000 points of a
-       processor-load performance trace from the {\tt 482.sphinx3}
-       benchmark.  Red circles and blue $\times$s are the true and
-       predicted values, respectively; vertical bars show where these
-       values differ.}
-\label{fig:predsphinx}
-\end{figure}
+%\begin{figure}[htbp]
+%  \centering
+%    \includegraphics[width=0.5\textwidth]{figs/%sphinxPredicTS}
+%     \caption{An LMA-based forecast of the last 4,000 points %of a
+%       processor-load performance trace from the {\tt %482.sphinx3}
+%       benchmark.  Red circles and blue $\times$s are the %true and
+%       predicted values, respectively; vertical bars show %where these
+%       values differ.}
+%\label{fig:predsphinx}
+%\end{figure}
 
-Table~\ref{tab:PredError} presents detailed results about the
-prediction accuracy of this algorithm on four different examples: the
-{\tt col\_major} and {\tt 482.sphinx3} programs in
-Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as
-another simple microkernel that initializes the same matrix as {\tt
-  col\_major}, but in row-major order, and another complex program
-({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.  Both
-microkernels were run on the Intel Core
-Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks
-were run on the Intel i7\textsuperscript{\textregistered} machine.  We
-calculated a figure of merit for each prediction as follows.  We held
-back the last $k$ elements\footnote{Several different prediction
-  horizons were analyzed in our experiment; the results reported in
-  this paper are for $k$=4000} of the $N$ points in each measured time
-series, built the forecast model by embedding the first $N-k$ points,
-used that embedding and the LMA method to predict the next $k$ points,
-then computed the Root Mean Squared Error (RMSE) between the true and
-predicted signals:
-$$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$
-%
-To compare the success of predictions across signals with different
-units, we normalized RMSE as follows:
-$$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
+%Table~\ref{tab:PredError} presents detailed results about the prediction accuracy of this algorithm on four different examples: the {\tt col\_major} and {\tt 482.sphinx3} programs in Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as another simple microkernel that initializes the same matrix as {\tt col\_major}, but in row-major order, and another complex program ({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.
+Both microkernels were run on the Intel Core Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks were run on the Intel i7\textsuperscript{\textregistered} machine.
+We calculated a figure of merit for each prediction as follows.
+We held back the last $k$ elements\footnote{Several different prediction   horizons were analyzed in our experiment; the results reported in this paper are for $k$=4000} of the $N$ points in each measured time series, built the forecast model by embedding the first $N-k$ points, used that embedding and the LMA method to predict the next $k$ points, then computed the Root Mean Squared Error (RMSE) between the true and predicted signals: $$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$ To compare the success of predictions across signals with different units, we normalized RMSE as follows: $$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
 %
@@
 
-The results in Table~\ref{tab:PredError} show a clear distinction
-between the two microkernels, whose future behavior can be
-predicted effectively using this simple deterministic modeling
-strategy, and the more-complex SPEC benchmarks, for which this
-prediction strategy does not work nearly as well.
+%The results in Table~\ref{tab:PredError} show a clear distinction between the two microkernels, whose future behavior can be predicted effectively using this simple deterministic modeling strategy, and the more-complex SPEC benchmarks, for which this prediction strategy does not work nearly as well.
 % Removed for space
@@
 
-\subsection{Comparing Prediction Accuracy}
+%\subsection{Comparing Prediction Accuracy}
+
+\subsection{Autoregressive-integrated-moving average(ARIMA)}
+\subsection{Naive: Random Walk and na\"ive}
+\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE)}
+
+\begin{figure}[htbp]
+  \centering
+   \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/gccMeanForecast}
+    \caption{\gcc na\"ive }
+    \label{fig:gccMEAN}
+  \end{subfigure}%      
+     \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/colMeanForecast}
+    \caption{\col na\"ive }
+    \label{fig:gccMEAN}
+  \end{subfigure}%    
+  \\
+  \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
+    \caption{\gcc ARIMA }
+    \label{fig:gccARIMA}
+  \end{subfigure}%
+  \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
+    \caption{\col ARIMA}
+    \label{fig:colARIMA}
+  \end{subfigure}
+  \\
+      \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
+    \caption{\gcc LMA}
+    \label{fig:gccLMA}
+  \end{subfigure}  
+    \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/colLMAForecast}
+    \caption{\col LMA}
+    \label{fig:colLMA}
+  \end{subfigure}
+
+  %\begin{subfigure}{0.5\textwidth}
+  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
+   \caption{
+For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
+On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
+}\label{fig:gcc_vs_col}
+  %\end{subfigure}
+\end{figure} 
 In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
@@
 dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.
+[[and cherry pick a few examples of \gcc and \col to put in the text For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=6$.]]
 
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
+% took out for space
 
@@
 In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
+We also list all of these values in Table \ref{tab:error}
+
+\begin{table}[htdp]
+\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 6 for each signal.}
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
+\hline
+\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$&  $0.5131 \pm 0.0034$ \\
+
+\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$&  $0.9572 \pm 0.0156$ \\
+\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &   $0.8464 \pm0.0044$ \\
+\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ &  $0.7157 \pm 0.0056$ \\
+\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$ & $0.8246 \pm 0.0077$ \\
+\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$&  $0.6776 \pm 0.0068$ \\
+\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$&  $0.7475 \pm 0.0106$ \\
+\hline
+\end{tabular}
+\end{center}
+\label{tab:error}
+\end{table}%
 

### 2014-02-19 concl.tex
Split Sections to help with conflicts

--- 2014-02-16 pecomplexity.tex
@@
-\documentclass{article}
-\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-%\geometry{landscape}                % Activate for for rotated page geometry
-%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{epstopdf}
-\usepackage{amsthm}
-
-\usepackage{pifont} % check and x marks
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
-
-\newcommand{\gcc}{{\tt 403.gcc}~}
-\newcommand{\svd}{{\tt dgesdd}~}
-\newcommand{\svdone}{{\tt dgesdd}$_1$~}
-\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
-\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
-\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
-\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
-\newcommand{\svdsix}{{\tt dgesdd$_6$}~}
-
-\newcommand{\row}{{\tt row\_major}~}
-\newcommand{\col}{{\tt col\_major}~}
-
-\usepackage{authblk}
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
-\newtheorem*{mydef}{Definition}
-\title{Determinism, Complexity, and Predictability in Computer Performance\\
-Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity
-\\ Quantifying Predictability using Structural Complexity
-}
-
-\author[1]{Joshua Garland \thanks{joshua.garland@colorado.edu}}
-\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
-\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-\affil[1]{Department of Computer Science\\
-  University of Colorado at Boulder\\
-  Colorado, USA
-}
-\affil[2]{Santa Fe Institute\\
-  New Mexico, USA
-}
-
-%Version 1
-%\author{
-%  \IEEEauthorblockN{Joshua Garland}
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: joshua.garland@colorado.edu}
-%  \and
-%  \IEEEauthorblockN{Ryan G.~James}
-%  \IEEEauthorblockA{Complexity Sciences Center \& Dept. of Physics\\
-%    University of California at Davis\\
-%    California, USA\\
-%    Email: rgjames@ucdavis.edu}
-%    \and
-%      \IEEEauthorblockN{Elizabeth Bradley}
-%         \IEEEauthorblockA{Santa Fe Institute \\
-%    New Mexico, USA\\
-%    }
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: lizb@colorado.edu}
-%  }
-
- %\author{
- %  \IEEEauthorblockN{
- %    Joshua Garland\IEEEauthorrefmark{1},
- %    Ryan G.~James\IEEEauthorrefmark{1} and
- %    Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{1}Dept. of Computer Science
- %    University of Colorado, Boulder, Colorado 80309-0430 USA\\
- %    Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
- %    Email: lizb@colorado.edu}
- %  % \IEEEauthorblockA{
- %  %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, %California 95616 USA\\
-%   %   Email: rgjames@ucdavis.edu}
-% }
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
-  Among other things, that implies that one should be able to use
-  deterministic forecast rules to predict aspects of their behavior.
-  That statement is sometimes---but not always---true. The memory and
-  processor loads of some simple programs are easy to predict, for
-  example, but those of more-complex programs like \gcc are not.
-%%%%%%%%%%%%%%%%%%%%%%
-%% I had to change all the \verb|blah| entries because they caused
-%% latex to barf if they were in figure captions.  Odd bug.
-%%%%%%%%%%%%%%%%%%%%%%
-  The goal of this paper is to determine why that is the case. We
-  conjecture that, in practice, complexity can effectively overwhelm
-  the predictive power of deterministic forecast models. To explore
-  that, we build models of a number of performance traces from
-  different programs running on different Intel-based computers. We
-  then calculate the \emph{permutation entropy}---a temporal entropy
-  metric that uses ordinal analysis---of those traces and correlate
-  those values against the prediction success.
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-BRAINSTORMING 
-\begin{itemize}
-\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
-\item random walk for example is best predicted by guess what just happend[[move to introduction]]
-\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]]
-
-\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-\end{itemize}
-
-%\begin{it}
-%Paragraph on computer performance, including citations to Todd paper
-%and summary of the results that indicate that they're deterministic
-%nonlinear dynamical systems.  Given that, we should be able to
-%predict.  What benefits would accrue if we could do so: power mgmt,
-%end world hunger [[this is my primary goal everyday :)]], etc.
-%\end{it}
-Things to add to introduction
-\begin{enumerate}
-\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
-\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.
-
-\item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc
-
-\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
-\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
-\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex.
-Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not.
-Maybe even having 4 predictions.
-top being ARIMA of the above signals and bottom being LMA of the above signals.
-Show that one improved and one did not.
-Is it that we used the wrong method to predict or is it that we simply can't predict the signal better than a random walk due to high levels of internal signal complexity.
-
-\item Introduce the two main contributions of the paper which are outlined at the begining of the results section
-
-\end{enumerate}
-
-Computers are among the most complex engineered artifacts in current
-use.  Modern microprocessor chips contain multiple processing units
-and multi-layer memories, for instance, and they use complicated
-hardware/software strategies to move data and threads of computation
-across those resources.  These features---along with all the others
-that go into the design of these chips---make the patterns of their
-processor loads and memory accesses highly complex and hard to
-predict.  Accurate forecasts of these quantities, if one could
-construct them, could be used to improve computer design.  If one
-could predict that a particular computational thread would be bogged
-down for the next 0.6 seconds waiting for data from main memory, for
-instance, one could save power by putting that thread on hold for that
-time period (e.g., by migrating it to a processing unit whose clock
-speed is scaled back).  Computer performance traces are, however, very
-complex.  Even a simple ``microkernel,'' like a three-line loop that
-repeatedly initializes a matrix in column-major order, can produce
-{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
-predictability.
-%
- \begin{figure}[htbp]
-    \centering
-    \includegraphics[width=\textwidth]{figs/colshortts}
-    % where an .eps filename suffix will be assumed under latex,
-    % and a .pdf suffix will be assumed for pdflatex
-    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
-        \col}, a three-line C program that repeatedly initializes
-      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
-      simple program exhibits chaotic performance dynamics.}
-   \label{fig:ipc}
-  \end{figure}
-
-\begin{figure}[htbp]
-  \centering
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/colshortts}
-    \caption{A short segment of the instructions executed per CPU clock cycle
-    (IPC) during the execution of \col. Each point is the IPC during a 100,000
-    cycle period. There is a great deal of periodicity in this time series.}
-    \label{fig:col-ts}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
-    \caption{The full IPC time series during the execution of \gcc. This time
-    series has very little structure.}
-    \label{fig:gcc-ts}
-  \end{subfigure}
-  \label{fig:sample-ts}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
-    \caption{\gcc ARIMA }
-    \label{fig:gccARIMA}
-  \end{subfigure}%
-    \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
-    \caption{\gcc LMA}
-    \label{fig:gccLMA}
-  \end{subfigure}
-  \\
-  \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
-    \caption{\col ARIMA}
-    \label{fig:colARIMA}
-  \end{subfigure}
-    \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colLMAForecast}
-    \caption{\col LMA}
-    \label{fig:colLMA}
-  \end{subfigure}
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{
-For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
-On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
-}\label{fig:gcc_vs_col}
-  %\end{subfigure}
-\end{figure}
-
-The computer systems community has applied a variety of prediction
-strategies to traces like this, most of which employ regression.  An
-appealing alternative builds on the recently established fact that
-computers can be effectively modeled as deterministic nonlinear
-dynamical systems \cite{mytkowicz09}.  This result implies the
-existence of a deterministic forecast rule for those dynamics.  In
-particular, one can use \emph{delay-coordinate embedding} to
-reconstruct the underlying dynamics of computer performance, then use
-the resulting model to forecast the future values of computer
-performance metrics such as memory or processor loads
-\cite{josh-ida2011}.  In the case of simple microkernels like the one
-that produced the trace in Figure~\ref{fig:ipc}, this deterministic
-modeling and forecast strategy works very well.  In more-complicated
-programs, however, such as speech recognition software or compilers,
-this forecast strategy---as well as the traditional methods---break
-down quickly.
-
-This paper is a first step in understanding when, why, and how
-deterministic forecast strategies fail when they are applied to
-deterministic systems.  We focus here on the specific example of
-computer performance.  We conjecture that the complexity of traces
-from these systems---which results from the inherent dimension,
-non-linearity, and non-stationarity of the dynamics, as well as from
-measurement issues like noise, aggregation, and finite data
-length---can make those deterministic signals \emph{effectively}
-unpredictable.  We argue that \emph{permutation entropy}
-\cite{bandt2002per}, a method for measuring the entropy of a
-real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study four
-examples---two simple microkernels and two complex programs from the
-SPEC benchmark suite---running on different Intel-based machines.  For
-each program, we calculate the permutation entropy of the processor
-load (instructions per cycle) and memory-use efficiency (cache-miss
-rates), then compare that to the prediction accuracy attainable for
-that trace using a simple deterministic model.
-
-% paragraph to appease the theoretician in me
-It is worth taking a moment to consider the theoretical possibility of
-this task. We are not attempting to predict the state of the CPU at an
-arbitrary point in the future --- this, at least with perfect
-accuracy, would be tantamount to solving the halting problem. What we
-are attempting is to predict aspects or functions of the running of
-the CPU: instructions executed per second, cache misses per 100,000
-instructions, and similar statistics. Prediction of these quantities
-at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem.
-
-% The rest of the paper is organized as follows.
-% Section~\ref{sec:compModel} describes the experimental setup, as well
-% as the nonlinear modeling and forecast strategies.  In
-% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
-% its value for a number of different computer performance traces, and
-% compare the results to the prediction accuracy.  In
-% Section~\ref{sec:conc}, we discuss these results and their
-% implications in regard to our conjecture, and consider future areas of
-% research.
-
-\section{Experimental Methods and Modelling}\label{sec:compModel}
-
-Section outline:
-
-\begin{enumerate}
-\item Experimental methods (how we collect the time series and what the times series are
-\item Description of DCE and parameter estimation  
-\item Description of auto ARIMA 
-\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
-\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
-\item Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
-
-\end{enumerate}
-
-% took out for space
- \subsection{Reconstructing hidden dynamics}
-
-Delay-coordinate embedding allows one to reconstruct a system's full
-state-space dynamics from a \emph{single} scalar time-series
-measurement---provided that some conditions hold regarding that data.
-Specifically, if the underlying dynamics and the measurement
-function---the mapping from the unknown state vector $\vec{X}$ to the
-scalar value $x$ that one is measuring---are both smooth and generic,
-Takens~\cite{takens} formally proves that the delay-coordinate map
-\[
-F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
-\]
-from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
-where $t$ is time, is a diffeomorphism on $M$---in other words, that
-the reconstructed dynamics and the true (hidden) dynamics have the
-same topology.
-
-This is an extremely powerful result: among other things, it means
-that one can build a formal model of the full system dynamics without
-measuring (or even knowing) every one of its state variables.  This is
-the foundation of the modeling approach that is used in this paper.
-The first step in the process is to estimate values for the two free
-parameters in the delay-coordinate map: the delay $\tau$ and the
-dimension $m$.  We follow standard procedures for this, choosing the
-first minimum in the average mutual information as an estimate of
-$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
-plot of the data from Figure~\ref{fig:ipc}, embedded following this
-procedure, is shown in Figure~\ref{fig:embedding}.
-
- \begin{figure}
-   \centering
-     \includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{A 3D projection of a delay-coordinate embedding of the trace
- from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
- }
- \label{fig:embedding}
- \end{figure}
-
-%% can cut for space if need be:
-The coordinates of each point on this plot are differently delayed
-elements of the \verb|col_major| L2 cache miss rate time series
-$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
-$y(t+2\tau)$ on the third, and so on.
-%% ...down to here.
-Structure in these kinds of plots---clearly visible in
-Figure~\ref{fig:embedding}---is an indication of
-determinism\footnote{A deeper analysis of
-  Figure~\ref{fig:embedding}---as alluded to on the previous
-  page---supports that diagnosis, confirming the presence of a chaotic
-  attractor in these cache-miss dynamics, with largest Lyapunov
-  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
-  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
-structure can also be used to build a forecast model.
-
-% took out for space
- \subsection{LMA: Using dynamics in forecasting}
-
-Given a nonlinear model of a deterministic dynamical system in the
-form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
-one can build deterministic forecast algorithms by capturing and
-exploiting the geometry of the embedding.  Many techniques have been
-developed by the dynamical systems community for this purpose
-(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
-is the ``Lorenz method of analogues'' (LMA), which is essentially
-nearest-neighbor prediction in the embedded state
-space~\cite{lorenz-analogues}.  Even this simple algorithm---which
-builds predictions by finding the nearest neighbor in the embedded
-space of the given point, then taking that neighbor's path as the
-forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
-shown in Figure~\ref{fig:cachePredTS}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
-    \caption{A forecast of the last 4,000 points of the signal in
-      Figure~\ref{fig:ipc} using an LMA-based strategy on the
-      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
-      $\times$s are the true and predicted values, respectively;
-      vertical bars show where these values differ. }
-\label{fig:cachePredTS}
-\end{figure}
-%
-On the other hand, if we use the same approach to forecast the
-processor load\footnote{Instructions per cycle, or IPC} of the {\tt
-  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
-on an Intel i7\textsuperscript{\textregistered}-based machine, the
-prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
-     \caption{An LMA-based forecast of the last 4,000 points of a
-       processor-load performance trace from the {\tt 482.sphinx3}
-       benchmark.  Red circles and blue $\times$s are the true and
-       predicted values, respectively; vertical bars show where these
-       values differ.}
-\label{fig:predsphinx}
-\end{figure}
-
-Table~\ref{tab:PredError} presents detailed results about the
-prediction accuracy of this algorithm on four different examples: the
-{\tt col\_major} and {\tt 482.sphinx3} programs in
-Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as
-another simple microkernel that initializes the same matrix as {\tt
-  col\_major}, but in row-major order, and another complex program
-({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.  Both
-microkernels were run on the Intel Core
-Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks
-were run on the Intel i7\textsuperscript{\textregistered} machine.  We
-calculated a figure of merit for each prediction as follows.  We held
-back the last $k$ elements\footnote{Several different prediction
-  horizons were analyzed in our experiment; the results reported in
-  this paper are for $k$=4000} of the $N$ points in each measured time
-series, built the forecast model by embedding the first $N-k$ points,
-used that embedding and the LMA method to predict the next $k$ points,
-then computed the Root Mean Squared Error (RMSE) between the true and
-predicted signals:
-$$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$
-%
-To compare the success of predictions across signals with different
-units, we normalized RMSE as follows:
-$$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
-%
-% The smaller the nRMSE, obviously, the more accurate the prediction.
-
-The results in Table~\ref{tab:PredError} show a clear distinction
-between the two microkernels, whose future behavior can be
-predicted effectively using this simple deterministic modeling
-strategy, and the more-complex SPEC benchmarks, for which this
-prediction strategy does not work nearly as well.
-% Removed for space
-% For both processor load (IPC) and memory usage (cache-miss rate),
-% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
-% for \verb|col_major| or \verb|row_major|.
-%
-This begs the question: If these traces all come from deterministic
-systems---computers---then why are they not equally predictable?  Our
-conjecture is that the sheer complexity of the dynamics of the SPEC
-benchmarks running on the Intel i7\textsuperscript{\textregistered}
-machine make them effectively impossible to predict.
-
-%% [[If space: Add a segue sentence: next section uses permutation
-%% entropy to explore that conjecture.]]
-
-\subsection{Comparing Prediction Accuracy}
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
-
-As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
-When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.
-
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
-
-\section{Measuring Structural Complexity}\label{sec:meaComplex}
-
-THINGS TO DO
-\begin{enumerate}
-\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
-
-\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-\item justify and explain the regime split in \svd 
-\end{enumerate}
-
-For the purposes of this paper, one can view entropy as a measure of complexity
-and predictability in a time series.  A high-entropy time series is almost
-completely unpredictable---and conversely.  This can be made more rigorous:
-Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
-Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
-$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
-states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
-difficult prediction becomes.
-
-Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
-categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
-\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
-effectively real-valued.  To get around this, one must discretize the data---
-typically achieved by binning.  Unfortunately, this is rarely a good solution to
-the problem, as the binning of the values introduces an additional dynamic on
-top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
-dynamics studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in the face
-of noise and require further understanding of the underlying system, which
-defeats the purpose of measuring the entropy in the first place.
-
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
-complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
-employs a method of discretizing real-valued time series that follows the
-intrinsic behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the Shannon
-entropy, permutation entropy looks at the statistics of the \emph{orderings} of
-sequences of values using ordinal analysis. Ordinal analysis of a time series is
-the process of mapping successive time-ordered elements of a time series to
-their value-ordered permutation of the same size.  By way of example, if $(x_1,
-x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
-other things, it is generally robust to observational noise and requires no
-knowledge of the underlying mechanisms.
-
-\begin{mydef}[Permutation Entropy]
-
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
-  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
-  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
-  = 1,\dots,T}$:
-  \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
-  \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
-  \ge 2$ is defined as
-  \begin{align*}
-  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
-  \end{align*}
-
-\end{mydef}
-
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
-it is common in the literature to normalize permutation entropy as follows:
-$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
-and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
-many large classes of systems \cite{amigo2012permutation}.
-
-Here we will be utilizing a variation of the permutation entropy, the
-\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
-permutation entropy attempts to correct for observational noise which is larger
-than some trends in the data, but smaller than the larger scale features --- for
-example, a signal that switches between two fixed points with noise about those
-fixed points. The weighted permutation entropy would be dominated by the
-switching rather than by the stochastic fluctuation. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
-\begin{align*}
-  w(x_{t+1:t+n}) = \frac{1}{n}
-                 \sum_{x_i \in \{x_{t+1:t+n}\}}
-                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
-\end{align*}
-where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
-$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
-
-The weighted probability of a permutation is then:
-\begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
-\end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
-probability enhances permutations involved in ``large'' features and demotes
-permutations which are small in amplitude relative to the features of the time
-series. The weighted permutation entropy is then:
-\begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
-\end{align*}
-which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
-results of this paper.
-
-In practice, calculating permutation entropy and weighted permutation entropy
-involves choosing a good value for the word length $n$. The primary
-consideration is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the ordinals are
-gathered: e.g.:
-\begin{align*}
-  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
-\end{align*}
-assuming an average of 100 counts per ordinal is sufficient. In the literature,
-$3 \le n \le 6$ is a standard choice --- generally without any formal
-justification. In theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In practice,
-what one should do is calculate the \emph{persistent} permutation entropy by
-increasing $n$ until the result converges, but data length issues can intrude
-before that convergence is reached.
-
-The weighted permutation entropy for the {\tt SVD} program is given in
-Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
-the time series. Within each of those windows, the statistics over words of
-length 4 are computed and the WPE is calculated. The gray bands denote regions
-where the 5,000 value window overlapped visually-distinct regimes. It can be
-seen that the behaviors of the weighted permutation entropy vary between
-regimes.
-[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
-Maybe point at some of the predicting predictability papers.]]
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
-%    are regions where the window overlaps regimes. The window size used is
-%    $5,000 \times 100,000$ instructions and the word length is $4$.}
-%  \label{fig:wwpe}
-%\end{figure}
-
-%%FIGURE INTENT
-%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
-%%%%%%%%%%%%%%%%%
-\begin{figure}[htbp]
-  \centering
-
-  %\begin{subfigure}{0.3\textwidth}
-  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
-  %  \caption{The instructions per cycle of \svd.
-Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
-From left to right each change in color represents a change in regime for 6 regimes in total.
-}
- %   \label{fig:svd_ts}
- % \end{subfigure}%
-  %\\
- % \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
-    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-  \caption{
-[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly.
-Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD.
-The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
-Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE.
-From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
-\end{figure}
-
- \section{Validation of Primary Findings} % (fold)
- \label{sec:results}
-
-%OUTLINE:
-%\begin{enumerate}
-%\item \cmark~Introduce MASE.
-%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
-%its WPE.
-%\item \cmark~Talk about structure analysis. Just because %there is forward information
-%transfer, does not mean that linear predictors can get at %this.
-%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
-%side plot.
-%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
-%\begin{enumerate}
-%\item \cmark~clusters are distributed differently
-%\item \cmark~clusters are shaped differently---tight or not
-%\item \cmark~clusters move differently between LMA and ARIMA
-%\item finally, the diagonal line is important. If you're %below it, you could do
-%better.
-%\end{enumerate}
-%\end{enumerate}
-
-%BRAINSTORMING 
-%\begin{itemize}
-%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
-%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
-
-%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
-
-%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
-
-%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-%\end{itemize}
-
-In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:
-
-\begin{enumerate}
-
-\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.
-
-%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
-
-%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
-
-\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-
-%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
-The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
-
-%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
-
-\end{enumerate}
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\item First paragraph: WPE is a good measure of predictability.  Figure:
-%best athlete MASE vs. its WPE.
-
-%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-
-As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task.
-As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
-Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.
-
-In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation.
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
-  \caption{The best MASE among all runs and prediction methods vs weighted
-  permutation entropy. For each of these, the word length used is $6$. The
-  dashed line is a least-squares linear fit of all the points except for {\tt
-  SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_wpe}
-\end{figure}
-What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}.
-We will not further validate this claim through some examples.
-
-To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively.
-In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. }
-    \label{fig:lma_pred_vs_ent}
-  \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:arima_pred_vs_ent}
-  \end{subfigure}
-\caption{
-For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}
-\end{figure}
-
-We will start with the lowest complexity program: \col.
-Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}.
-If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
-This would suggest that there is some predictive structure present but not that much.
-However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity.
-This is in fact the case.
-\col is a complex signal, i.e., chaotic, but it is complex in a structured way.
-Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
-The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
-
-Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive.
-This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
-However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
-The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
-This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict.
-However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA.
-If we do this we get forecasts which score 1.5 times better than the random walk.
-Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.
-
-Finally, we consider the most complex program: \gcc.
-The complexity of \gcc is $0.94 \pm 0.001$.
-This implies that almost no predictive structure exists in this structure.
-Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future.
-This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex.
-In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step.
-This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.
-
-There is a single outlier in this study which is \svdone.
-This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively).
-The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts.
-We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method.
-MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
-In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
-A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.
-
-To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail.
-This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
-We provide a small snippet as opposed to the full time series simply for visual clarity.
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=\textwidth]{figs/svdonets2}
-\caption{ For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}
-\end{figure} 
-As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide.
-Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
-The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}.
-These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle).
-As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
-At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal.
-This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
-The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.
-
-[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}.
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
-
-%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration.
-The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method).
-On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values.
-As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE.
-The WPE is sensitive to both linear and nonlinear structure.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
-So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
-
-%is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
-How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict.
-If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy.
-However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-
-Other noteworthy features of the LMA and ARIMA results are the cluster locations
-and distributions. The WPE values of each run of any particular program tend to
-have little variance, leading to the clusters in
-Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
-constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
-low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
-traces, however, have a large variance resulting in the spread seen in
-Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
-bad, in that other predictors vastly outperform it, but they are inconsistent.
-Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
-see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
-the right than those in Figure~\ref{fig:lma_pred_vs_ent}.
-
 \section{ Conclusions \& Future Work}\label{sec:conc}
@@
 %\end{it}
-
-\section*{Acknowledgment}
-This work was partially supported by NSF grant \#CMMI-1245947 and ARO
-grant \#W911NF-12-1-0288.
-
-\bibliographystyle{unsrt}
-\bibliography{bibliofile}
-
-\end{document}

### 2014-02-19 intro.tex
Split Sections to help with conflicts

--- 2014-02-16 pecomplexity.tex
@@
-\documentclass{article}
-\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-%\geometry{landscape}                % Activate for for rotated page geometry
-%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{epstopdf}
-\usepackage{amsthm}
-
-\usepackage{pifont} % check and x marks
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
-
-\newcommand{\gcc}{{\tt 403.gcc}~}
-\newcommand{\svd}{{\tt dgesdd}~}
-\newcommand{\svdone}{{\tt dgesdd}$_1$~}
-\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
-\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
-\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
-\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
-\newcommand{\svdsix}{{\tt dgesdd$_6$}~}
-
-\newcommand{\row}{{\tt row\_major}~}
-\newcommand{\col}{{\tt col\_major}~}
-
-\usepackage{authblk}
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
-\newtheorem*{mydef}{Definition}
-\title{Determinism, Complexity, and Predictability in Computer Performance\\
-Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity
-\\ Quantifying Predictability using Structural Complexity
-}
-
-\author[1]{Joshua Garland \thanks{joshua.garland@colorado.edu}}
-\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
-\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-\affil[1]{Department of Computer Science\\
-  University of Colorado at Boulder\\
-  Colorado, USA
-}
-\affil[2]{Santa Fe Institute\\
-  New Mexico, USA
-}
-
-%Version 1
-%\author{
-%  \IEEEauthorblockN{Joshua Garland}
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: joshua.garland@colorado.edu}
-%  \and
-%  \IEEEauthorblockN{Ryan G.~James}
-%  \IEEEauthorblockA{Complexity Sciences Center \& Dept. of Physics\\
-%    University of California at Davis\\
-%    California, USA\\
-%    Email: rgjames@ucdavis.edu}
-%    \and
-%      \IEEEauthorblockN{Elizabeth Bradley}
-%         \IEEEauthorblockA{Santa Fe Institute \\
-%    New Mexico, USA\\
-%    }
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: lizb@colorado.edu}
-%  }
-
- %\author{
- %  \IEEEauthorblockN{
- %    Joshua Garland\IEEEauthorrefmark{1},
- %    Ryan G.~James\IEEEauthorrefmark{1} and
- %    Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{1}Dept. of Computer Science
- %    University of Colorado, Boulder, Colorado 80309-0430 USA\\
- %    Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
- %    Email: lizb@colorado.edu}
- %  % \IEEEauthorblockA{
- %  %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, %California 95616 USA\\
-%   %   Email: rgjames@ucdavis.edu}
-% }
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
-  Among other things, that implies that one should be able to use
-  deterministic forecast rules to predict aspects of their behavior.
-  That statement is sometimes---but not always---true. The memory and
-  processor loads of some simple programs are easy to predict, for
-  example, but those of more-complex programs like \gcc are not.
-%%%%%%%%%%%%%%%%%%%%%%
-%% I had to change all the \verb|blah| entries because they caused
-%% latex to barf if they were in figure captions.  Odd bug.
-%%%%%%%%%%%%%%%%%%%%%%
-  The goal of this paper is to determine why that is the case. We
-  conjecture that, in practice, complexity can effectively overwhelm
-  the predictive power of deterministic forecast models. To explore
-  that, we build models of a number of performance traces from
-  different programs running on different Intel-based computers. We
-  then calculate the \emph{permutation entropy}---a temporal entropy
-  metric that uses ordinal analysis---of those traces and correlate
-  those values against the prediction success.
-\end{abstract}
-
 \section{Introduction}\label{sec:intro}
@@
 % research.
-
-\section{Experimental Methods and Modelling}\label{sec:compModel}
-
-Section outline:
-
-\begin{enumerate}
-\item Experimental methods (how we collect the time series and what the times series are
-\item Description of DCE and parameter estimation  
-\item Description of auto ARIMA 
-\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
-\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
-\item Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
-
-\end{enumerate}
-
-% took out for space
- \subsection{Reconstructing hidden dynamics}
-
-Delay-coordinate embedding allows one to reconstruct a system's full
-state-space dynamics from a \emph{single} scalar time-series
-measurement---provided that some conditions hold regarding that data.
-Specifically, if the underlying dynamics and the measurement
-function---the mapping from the unknown state vector $\vec{X}$ to the
-scalar value $x$ that one is measuring---are both smooth and generic,
-Takens~\cite{takens} formally proves that the delay-coordinate map
-\[
-F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
-\]
-from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
-where $t$ is time, is a diffeomorphism on $M$---in other words, that
-the reconstructed dynamics and the true (hidden) dynamics have the
-same topology.
-
-This is an extremely powerful result: among other things, it means
-that one can build a formal model of the full system dynamics without
-measuring (or even knowing) every one of its state variables.  This is
-the foundation of the modeling approach that is used in this paper.
-The first step in the process is to estimate values for the two free
-parameters in the delay-coordinate map: the delay $\tau$ and the
-dimension $m$.  We follow standard procedures for this, choosing the
-first minimum in the average mutual information as an estimate of
-$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
-plot of the data from Figure~\ref{fig:ipc}, embedded following this
-procedure, is shown in Figure~\ref{fig:embedding}.
-
- \begin{figure}
-   \centering
-     \includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{A 3D projection of a delay-coordinate embedding of the trace
- from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
- }
- \label{fig:embedding}
- \end{figure}
-
-%% can cut for space if need be:
-The coordinates of each point on this plot are differently delayed
-elements of the \verb|col_major| L2 cache miss rate time series
-$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
-$y(t+2\tau)$ on the third, and so on.
-%% ...down to here.
-Structure in these kinds of plots---clearly visible in
-Figure~\ref{fig:embedding}---is an indication of
-determinism\footnote{A deeper analysis of
-  Figure~\ref{fig:embedding}---as alluded to on the previous
-  page---supports that diagnosis, confirming the presence of a chaotic
-  attractor in these cache-miss dynamics, with largest Lyapunov
-  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
-  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
-structure can also be used to build a forecast model.
-
-% took out for space
- \subsection{LMA: Using dynamics in forecasting}
-
-Given a nonlinear model of a deterministic dynamical system in the
-form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
-one can build deterministic forecast algorithms by capturing and
-exploiting the geometry of the embedding.  Many techniques have been
-developed by the dynamical systems community for this purpose
-(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
-is the ``Lorenz method of analogues'' (LMA), which is essentially
-nearest-neighbor prediction in the embedded state
-space~\cite{lorenz-analogues}.  Even this simple algorithm---which
-builds predictions by finding the nearest neighbor in the embedded
-space of the given point, then taking that neighbor's path as the
-forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
-shown in Figure~\ref{fig:cachePredTS}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
-    \caption{A forecast of the last 4,000 points of the signal in
-      Figure~\ref{fig:ipc} using an LMA-based strategy on the
-      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
-      $\times$s are the true and predicted values, respectively;
-      vertical bars show where these values differ. }
-\label{fig:cachePredTS}
-\end{figure}
-%
-On the other hand, if we use the same approach to forecast the
-processor load\footnote{Instructions per cycle, or IPC} of the {\tt
-  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
-on an Intel i7\textsuperscript{\textregistered}-based machine, the
-prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
-     \caption{An LMA-based forecast of the last 4,000 points of a
-       processor-load performance trace from the {\tt 482.sphinx3}
-       benchmark.  Red circles and blue $\times$s are the true and
-       predicted values, respectively; vertical bars show where these
-       values differ.}
-\label{fig:predsphinx}
-\end{figure}
-
-Table~\ref{tab:PredError} presents detailed results about the
-prediction accuracy of this algorithm on four different examples: the
-{\tt col\_major} and {\tt 482.sphinx3} programs in
-Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as
-another simple microkernel that initializes the same matrix as {\tt
-  col\_major}, but in row-major order, and another complex program
-({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.  Both
-microkernels were run on the Intel Core
-Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks
-were run on the Intel i7\textsuperscript{\textregistered} machine.  We
-calculated a figure of merit for each prediction as follows.  We held
-back the last $k$ elements\footnote{Several different prediction
-  horizons were analyzed in our experiment; the results reported in
-  this paper are for $k$=4000} of the $N$ points in each measured time
-series, built the forecast model by embedding the first $N-k$ points,
-used that embedding and the LMA method to predict the next $k$ points,
-then computed the Root Mean Squared Error (RMSE) between the true and
-predicted signals:
-$$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$
-%
-To compare the success of predictions across signals with different
-units, we normalized RMSE as follows:
-$$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
-%
-% The smaller the nRMSE, obviously, the more accurate the prediction.
-
-The results in Table~\ref{tab:PredError} show a clear distinction
-between the two microkernels, whose future behavior can be
-predicted effectively using this simple deterministic modeling
-strategy, and the more-complex SPEC benchmarks, for which this
-prediction strategy does not work nearly as well.
-% Removed for space
-% For both processor load (IPC) and memory usage (cache-miss rate),
-% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
-% for \verb|col_major| or \verb|row_major|.
-%
-This begs the question: If these traces all come from deterministic
-systems---computers---then why are they not equally predictable?  Our
-conjecture is that the sheer complexity of the dynamics of the SPEC
-benchmarks running on the Intel i7\textsuperscript{\textregistered}
-machine make them effectively impossible to predict.
-
-%% [[If space: Add a segue sentence: next section uses permutation
-%% entropy to explore that conjecture.]]
-
-\subsection{Comparing Prediction Accuracy}
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
-
-As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
-When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.
-
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
-
-\section{Measuring Structural Complexity}\label{sec:meaComplex}
-
-THINGS TO DO
-\begin{enumerate}
-\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
-
-\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-\item justify and explain the regime split in \svd 
-\end{enumerate}
-
-For the purposes of this paper, one can view entropy as a measure of complexity
-and predictability in a time series.  A high-entropy time series is almost
-completely unpredictable---and conversely.  This can be made more rigorous:
-Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
-Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
-$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
-states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
-difficult prediction becomes.
-
-Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
-categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
-\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
-effectively real-valued.  To get around this, one must discretize the data---
-typically achieved by binning.  Unfortunately, this is rarely a good solution to
-the problem, as the binning of the values introduces an additional dynamic on
-top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
-dynamics studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in the face
-of noise and require further understanding of the underlying system, which
-defeats the purpose of measuring the entropy in the first place.
-
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
-complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
-employs a method of discretizing real-valued time series that follows the
-intrinsic behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the Shannon
-entropy, permutation entropy looks at the statistics of the \emph{orderings} of
-sequences of values using ordinal analysis. Ordinal analysis of a time series is
-the process of mapping successive time-ordered elements of a time series to
-their value-ordered permutation of the same size.  By way of example, if $(x_1,
-x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
-other things, it is generally robust to observational noise and requires no
-knowledge of the underlying mechanisms.
-
-\begin{mydef}[Permutation Entropy]
-
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
-  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
-  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
-  = 1,\dots,T}$:
-  \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
-  \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
-  \ge 2$ is defined as
-  \begin{align*}
-  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
-  \end{align*}
-
-\end{mydef}
-
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
-it is common in the literature to normalize permutation entropy as follows:
-$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
-and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
-many large classes of systems \cite{amigo2012permutation}.
-
-Here we will be utilizing a variation of the permutation entropy, the
-\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
-permutation entropy attempts to correct for observational noise which is larger
-than some trends in the data, but smaller than the larger scale features --- for
-example, a signal that switches between two fixed points with noise about those
-fixed points. The weighted permutation entropy would be dominated by the
-switching rather than by the stochastic fluctuation. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
-\begin{align*}
-  w(x_{t+1:t+n}) = \frac{1}{n}
-                 \sum_{x_i \in \{x_{t+1:t+n}\}}
-                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
-\end{align*}
-where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
-$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
-
-The weighted probability of a permutation is then:
-\begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
-\end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
-probability enhances permutations involved in ``large'' features and demotes
-permutations which are small in amplitude relative to the features of the time
-series. The weighted permutation entropy is then:
-\begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
-\end{align*}
-which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
-results of this paper.
-
-In practice, calculating permutation entropy and weighted permutation entropy
-involves choosing a good value for the word length $n$. The primary
-consideration is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the ordinals are
-gathered: e.g.:
-\begin{align*}
-  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
-\end{align*}
-assuming an average of 100 counts per ordinal is sufficient. In the literature,
-$3 \le n \le 6$ is a standard choice --- generally without any formal
-justification. In theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In practice,
-what one should do is calculate the \emph{persistent} permutation entropy by
-increasing $n$ until the result converges, but data length issues can intrude
-before that convergence is reached.
-
-The weighted permutation entropy for the {\tt SVD} program is given in
-Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
-the time series. Within each of those windows, the statistics over words of
-length 4 are computed and the WPE is calculated. The gray bands denote regions
-where the 5,000 value window overlapped visually-distinct regimes. It can be
-seen that the behaviors of the weighted permutation entropy vary between
-regimes.
-[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
-Maybe point at some of the predicting predictability papers.]]
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
-%    are regions where the window overlaps regimes. The window size used is
-%    $5,000 \times 100,000$ instructions and the word length is $4$.}
-%  \label{fig:wwpe}
-%\end{figure}
-
-%%FIGURE INTENT
-%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
-%%%%%%%%%%%%%%%%%
-\begin{figure}[htbp]
-  \centering
-
-  %\begin{subfigure}{0.3\textwidth}
-  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
-  %  \caption{The instructions per cycle of \svd.
-Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
-From left to right each change in color represents a change in regime for 6 regimes in total.
-}
- %   \label{fig:svd_ts}
- % \end{subfigure}%
-  %\\
- % \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
-    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-  \caption{
-[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly.
-Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD.
-The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
-Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE.
-From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
-\end{figure}
-
- \section{Validation of Primary Findings} % (fold)
- \label{sec:results}
-
-%OUTLINE:
-%\begin{enumerate}
-%\item \cmark~Introduce MASE.
-%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
-%its WPE.
-%\item \cmark~Talk about structure analysis. Just because %there is forward information
-%transfer, does not mean that linear predictors can get at %this.
-%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
-%side plot.
-%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
-%\begin{enumerate}
-%\item \cmark~clusters are distributed differently
-%\item \cmark~clusters are shaped differently---tight or not
-%\item \cmark~clusters move differently between LMA and ARIMA
-%\item finally, the diagonal line is important. If you're %below it, you could do
-%better.
-%\end{enumerate}
-%\end{enumerate}
-
-%BRAINSTORMING 
-%\begin{itemize}
-%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
-%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
-
-%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
-
-%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
-
-%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-%\end{itemize}
-
-In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:
-
-\begin{enumerate}
-
-\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.
-
-%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
-
-%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
-
-\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-
-%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
-The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
-
-%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
-
-\end{enumerate}
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\item First paragraph: WPE is a good measure of predictability.  Figure:
-%best athlete MASE vs. its WPE.
-
-%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-
-As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task.
-As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
-Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.
-
-In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation.
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
-  \caption{The best MASE among all runs and prediction methods vs weighted
-  permutation entropy. For each of these, the word length used is $6$. The
-  dashed line is a least-squares linear fit of all the points except for {\tt
-  SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_wpe}
-\end{figure}
-What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}.
-We will not further validate this claim through some examples.
-
-To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively.
-In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. }
-    \label{fig:lma_pred_vs_ent}
-  \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:arima_pred_vs_ent}
-  \end{subfigure}
-\caption{
-For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}
-\end{figure}
-
-We will start with the lowest complexity program: \col.
-Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}.
-If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
-This would suggest that there is some predictive structure present but not that much.
-However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity.
-This is in fact the case.
-\col is a complex signal, i.e., chaotic, but it is complex in a structured way.
-Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
-The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
-
-Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive.
-This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
-However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
-The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
-This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict.
-However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA.
-If we do this we get forecasts which score 1.5 times better than the random walk.
-Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.
-
-Finally, we consider the most complex program: \gcc.
-The complexity of \gcc is $0.94 \pm 0.001$.
-This implies that almost no predictive structure exists in this structure.
-Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future.
-This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex.
-In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step.
-This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.
-
-There is a single outlier in this study which is \svdone.
-This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively).
-The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts.
-We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method.
-MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
-In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
-A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.
-
-To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail.
-This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
-We provide a small snippet as opposed to the full time series simply for visual clarity.
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=\textwidth]{figs/svdonets2}
-\caption{ For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}
-\end{figure} 
-As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide.
-Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
-The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}.
-These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle).
-As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
-At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal.
-This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
-The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.
-
-[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}.
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
-
-%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration.
-The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method).
-On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values.
-As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE.
-The WPE is sensitive to both linear and nonlinear structure.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
-So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
-
-%is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
-How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict.
-If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy.
-However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-
-Other noteworthy features of the LMA and ARIMA results are the cluster locations
-and distributions. The WPE values of each run of any particular program tend to
-have little variance, leading to the clusters in
-Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
-constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
-low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
-traces, however, have a large variance resulting in the spread seen in
-Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
-bad, in that other predictors vastly outperform it, but they are inconsistent.
-Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
-see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
-the right than those in Figure~\ref{fig:lma_pred_vs_ent}.
-
-\section{ Conclusions \& Future Work}\label{sec:conc}
-
-\begin{enumerate}
-\item  This need to be rewritten to conclude and address this paper...
-\end{enumerate}
-
-The results presented here suggest that permutation entropy---a ordinal
-calculation of forward information transfer in a time series---is an effective
-metric for predictability of computer performance traces. Experimentally, traces
-with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
-may overshadow the inherent determinism in the system dynamics, whereas traces
-with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
-of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
-0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
-corroborating the results of~\cite{mytkowicz09}.
-
-If information is the limit, then gathering and using more information is an
-obvious next step.  There is an equally obvious tension here between data length
-and prediction speed: a forecast that requires half a second to compute is not
-useful for the purposes of real-time control of a computer system with a MHz
-clock rate.  Another alternative is to sample several system variables
-simultaneously and build multivariate delay-coordinate embeddings.  Existing
-approaches to that are computationally prohibitive
-\cite{cao-multivariate-embedding}.  We are working on alternative
-methods that sidestep that complexity.
-
-%%This use of permutaiotn entropy is a useful application of information
-%%theory to computer performance modeling because it gives you a simple
-%%and fast way to decide whether building a model and trying to forecast
-%%the future is worthwhile or if guessing the mean is just as effective.
-%%End with a sentence tying back to power management and world peace.
-
-%\begin{it}
-%Summarize the results.
-
-%Paragraphs on the issues that come up, including one about the "amount
-%of info" one: if one could sample more variables, for instance, one
-%might be able to do a better job of predicting more-complex traces.
-%Segue to some handwaving about multivariable LMA models; tie this back
-%to the "computers are NLD systems" stuff in the intro.  This is a real
-%challenge; current approaches to this modelling problem have the major
-%issue of taking way too long to build.  And that's a big issue if
-%you're trying not just to classify, but to predict.  In a system that
-%runs at MHz speeds, a prediction that takes milliseconds to compute is
-%not useful.
-%\end{it}
-
-\section*{Acknowledgment}
-This work was partially supported by NSF grant \#CMMI-1245947 and ARO
-grant \#W911NF-12-1-0288.
-
-\bibliographystyle{unsrt}
-\bibliography{bibliofile}
-
-\end{document}

### 2014-02-19 methods.tex
Split Sections to help with conflicts

--- 2014-02-18 pecomplexity.tex
@@
-\documentclass{article}
-\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-%\geometry{landscape}                % Activate for for rotated page geometry
-%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{epstopdf}
-\usepackage{amsthm}
-
-\usepackage{pifont} % check and x marks
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
-
-\newcommand{\gcc}{{\tt 403.gcc}~}
-\newcommand{\svd}{{\tt dgesdd}~}
-\newcommand{\svdone}{{\tt dgesdd}$_1$~}
-\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
-\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
-\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
-\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
-\newcommand{\svdsix}{{\tt dgesdd$_6$}~}
-
-\newcommand{\row}{{\tt row\_major}~}
-\newcommand{\col}{{\tt col\_major}~}
-
-\usepackage{authblk}
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
-\newtheorem*{mydef}{Definition}
-\title{Determinism, Complexity, and Predictability in Computer Performance\\
-Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity
-\\ Quantifying Predictability using Structural Complexity
-}
-
-\author[1]{Joshua Garland \thanks{joshua.garland@colorado.edu}}
-\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
-\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-\affil[1]{Department of Computer Science\\
-  University of Colorado at Boulder\\
-  Colorado, USA
-}
-\affil[2]{Santa Fe Institute\\
-  New Mexico, USA
-}
-
-%Version 1
-%\author{
-%  \IEEEauthorblockN{Joshua Garland}
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: joshua.garland@colorado.edu}
-%  \and
-%  \IEEEauthorblockN{Ryan G.~James}
-%  \IEEEauthorblockA{Complexity Sciences Center \& Dept. of Physics\\
-%    University of California at Davis\\
-%    California, USA\\
-%    Email: rgjames@ucdavis.edu}
-%    \and
-%      \IEEEauthorblockN{Elizabeth Bradley}
-%         \IEEEauthorblockA{Santa Fe Institute \\
-%    New Mexico, USA\\
-%    }
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: lizb@colorado.edu}
-%  }
-
- %\author{
- %  \IEEEauthorblockN{
- %    Joshua Garland\IEEEauthorrefmark{1},
- %    Ryan G.~James\IEEEauthorrefmark{1} and
- %    Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{1}Dept. of Computer Science
- %    University of Colorado, Boulder, Colorado 80309-0430 USA\\
- %    Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
- %    Email: lizb@colorado.edu}
- %  % \IEEEauthorblockA{
- %  %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, %California 95616 USA\\
-%   %   Email: rgjames@ucdavis.edu}
-% }
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
-  Among other things, that implies that one should be able to use
-  deterministic forecast rules to predict aspects of their behavior.
-  That statement is sometimes---but not always---true. The memory and
-  processor loads of some simple programs are easy to predict, for
-  example, but those of more-complex programs like \gcc are not.
-%%%%%%%%%%%%%%%%%%%%%%
-%% I had to change all the \verb|blah| entries because they caused
-%% latex to barf if they were in figure captions.  Odd bug.
-%%%%%%%%%%%%%%%%%%%%%%
-  The goal of this paper is to determine why that is the case. We
-  conjecture that, in practice, complexity can effectively overwhelm
-  the predictive power of deterministic forecast models. To explore
-  that, we build models of a number of performance traces from
-  different programs running on different Intel-based computers. We
-  then calculate the \emph{permutation entropy}---a temporal entropy
-  metric that uses ordinal analysis---of those traces and correlate
-  those values against the prediction success.
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-BRAINSTORMING 
-\begin{itemize}
-\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
-\item random walk for example is best predicted by guess what just happend[[move to introduction]]
-\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]]
-
-\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-\end{itemize}
-
-%\begin{it}
-%Paragraph on computer performance, including citations to Todd paper
-%and summary of the results that indicate that they're deterministic
-%nonlinear dynamical systems.  Given that, we should be able to
-%predict.  What benefits would accrue if we could do so: power mgmt,
-%end world hunger [[this is my primary goal everyday :)]], etc.
-%\end{it}
-Things to add to introduction
-\begin{enumerate}
-\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
-\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.
-
-\item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc
-
-\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
-\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
-\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex.
-Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not.
-Maybe even having 4 predictions.
-top being ARIMA of the above signals and bottom being LMA of the above signals.
-Show that one improved and one did not.
-Is it that we used the wrong method to predict or is it that we simply can't predict the signal better than a random walk due to high levels of internal signal complexity.
-
-\item Introduce the two main contributions of the paper which are outlined at the begining of the results section
-
-\end{enumerate}
-
-Computers are among the most complex engineered artifacts in current
-use.  Modern microprocessor chips contain multiple processing units
-and multi-layer memories, for instance, and they use complicated
-hardware/software strategies to move data and threads of computation
-across those resources.  These features---along with all the others
-that go into the design of these chips---make the patterns of their
-processor loads and memory accesses highly complex and hard to
-predict.  Accurate forecasts of these quantities, if one could
-construct them, could be used to improve computer design.  If one
-could predict that a particular computational thread would be bogged
-down for the next 0.6 seconds waiting for data from main memory, for
-instance, one could save power by putting that thread on hold for that
-time period (e.g., by migrating it to a processing unit whose clock
-speed is scaled back).  Computer performance traces are, however, very
-complex.  Even a simple ``microkernel,'' like a three-line loop that
-repeatedly initializes a matrix in column-major order, can produce
-{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
-predictability.
-%
- \begin{figure}[htbp]
-    \centering
-    \includegraphics[width=\textwidth]{figs/colshortts}
-    % where an .eps filename suffix will be assumed under latex,
-    % and a .pdf suffix will be assumed for pdflatex
-    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
-        \col}, a three-line C program that repeatedly initializes
-      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
-      simple program exhibits chaotic performance dynamics.}
-   \label{fig:ipc}
-  \end{figure}
-
-The computer systems community has applied a variety of prediction
-strategies to traces like this, most of which employ regression.  An
-appealing alternative builds on the recently established fact that
-computers can be effectively modeled as deterministic nonlinear
-dynamical systems \cite{mytkowicz09}.  This result implies the
-existence of a deterministic forecast rule for those dynamics.  In
-particular, one can use \emph{delay-coordinate embedding} to
-reconstruct the underlying dynamics of computer performance, then use
-the resulting model to forecast the future values of computer
-performance metrics such as memory or processor loads
-\cite{josh-ida2011}.  In the case of simple microkernels like the one
-that produced the trace in Figure~\ref{fig:ipc}, this deterministic
-modeling and forecast strategy works very well.  In more-complicated
-programs, however, such as speech recognition software or compilers,
-this forecast strategy---as well as the traditional methods---break
-down quickly.
-
-This paper is a first step in understanding when, why, and how
-deterministic forecast strategies fail when they are applied to
-deterministic systems.  We focus here on the specific example of
-computer performance.  We conjecture that the complexity of traces
-from these systems---which results from the inherent dimension,
-non-linearity, and non-stationarity of the dynamics, as well as from
-measurement issues like noise, aggregation, and finite data
-length---can make those deterministic signals \emph{effectively}
-unpredictable.  We argue that \emph{permutation entropy}
-\cite{bandt2002per}, a method for measuring the entropy of a
-real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study four
-examples---two simple microkernels and two complex programs from the
-SPEC benchmark suite---running on different Intel-based machines.  For
-each program, we calculate the permutation entropy of the processor
-load (instructions per cycle) and memory-use efficiency (cache-miss
-rates), then compare that to the prediction accuracy attainable for
-that trace using a simple deterministic model.
-
-% paragraph to appease the theoretician in me
-It is worth taking a moment to consider the theoretical possibility of
-this task. We are not attempting to predict the state of the CPU at an
-arbitrary point in the future --- this, at least with perfect
-accuracy, would be tantamount to solving the halting problem. What we
-are attempting is to predict aspects or functions of the running of
-the CPU: instructions executed per second, cache misses per 100,000
-instructions, and similar statistics. Prediction of these quantities
-at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem.
-
-% The rest of the paper is organized as follows.
-% Section~\ref{sec:compModel} describes the experimental setup, as well
-% as the nonlinear modeling and forecast strategies.  In
-% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
-% its value for a number of different computer performance traces, and
-% compare the results to the prediction accuracy.  In
-% Section~\ref{sec:conc}, we discuss these results and their
-% implications in regard to our conjecture, and consider future areas of
-% research.
-
 \section{Experimental Methods }
@@
 \end{figure}
-
-\section{Modeling}
-Section outline:
-
-\begin{enumerate}
-
-\item Description of DCE and parameter estimation  
-\item Description of auto ARIMA 
-\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
-\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
-\item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
-
-\end{enumerate}
-
-\label{sec:compModel}
-
-\subsection{Lorenz Method of Analogues (LMA)}
-TODO:
-\begin{enumerate}
-\item update examples
-\item remove section on RMSE
-\item update figures
-\end{enumerate}
-
- \subsubsection{Reconstructing hidden dynamics}
-
-Delay-coordinate embedding allows one to reconstruct a system's full
-state-space dynamics from a \emph{single} scalar time-series
-measurement---provided that some conditions hold regarding that data.
-Specifically, if the underlying dynamics and the measurement
-function---the mapping from the unknown state vector $\vec{X}$ to the
-scalar value $x$ that one is measuring---are both smooth and generic,
-Takens~\cite{takens} formally proves that the delay-coordinate map
-\[
-F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
-\]
-from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
-where $t$ is time, is a diffeomorphism on $M$---in other words, that
-the reconstructed dynamics and the true (hidden) dynamics have the
-same topology.
-
-This is an extremely powerful result: among other things, it means
-that one can build a formal model of the full system dynamics without
-measuring (or even knowing) every one of its state variables.  This is
-the foundation of the modeling approach that is used in this paper.
-The first step in the process is to estimate values for the two free
-parameters in the delay-coordinate map: the delay $\tau$ and the
-dimension $m$.  We follow standard procedures for this, choosing the
-first minimum in the average mutual information as an estimate of
-$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
-plot of the data from Figure~\ref{fig:ipc}, embedded following this
-procedure, is shown in Figure~\ref{fig:embedding}.
-
- \begin{figure}
-   \centering
-     \includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{[[Joshua:Perhaps add a similar graphic of \gcc to show contrast]]A 3D projection of a delay-coordinate embedding of the trace
- from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
- }
- \label{fig:embedding}
- \end{figure}
-
-%% can cut for space if need be:
-The coordinates of each point on this plot are differently delayed
-elements of the \col L2 cache miss rate time series
-$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
-$y(t+2\tau)$ on the third, and so on.
-%% ...down to here.
-Structure in these kinds of plots---clearly visible in
-Figure~\ref{fig:embedding}---is an indication of
-determinism\footnote{A deeper analysis of
-  Figure~\ref{fig:embedding}---as alluded to on the previous
-  page---supports that diagnosis, confirming the presence of a chaotic
-  attractor in these cache-miss dynamics, with largest Lyapunov
-  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
-  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
-structure can also be used to build a forecast model.
-
-% took out for space
- \subsubsection{LMA: Using dynamics in forecasting}
-
-Given a nonlinear model of a deterministic dynamical system in the
-form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
-one can build deterministic forecast algorithms by capturing and
-exploiting the geometry of the embedding.  Many techniques have been
-developed by the dynamical systems community for this purpose
-(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
-is the ``Lorenz method of analogues'' (LMA), which is essentially
-nearest-neighbor prediction in the embedded state
-space~\cite{lorenz-analogues}.  Even this simple algorithm---which
-builds predictions by finding the nearest neighbor in the embedded
-space of the given point, then taking that neighbor's path as the
-forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
-shown in Figure~\ref{fig:cachePredTS}.
-%
-%\begin{figure}[htbp]
-%  \centering
-%    \includegraphics[width=0.5\textwidth]{figs/%colCachePredTS}
-%    \caption{A forecast of the last 4,000 points of the %signal in
-%      Figure~\ref{fig:ipc} using an LMA-based strategy on the
-%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
-%      $\times$s are the true and predicted values, %respectively;
-%      vertical bars show where these values differ. }
-%\label{fig:cachePredTS}
-%\end{figure}
-%
-On the other hand, if we use the same approach to forecast the
-processor load\footnote{Instructions per cycle, or IPC} of the {\tt
-  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
-on an Intel i7\textsuperscript{\textregistered}-based machine, the
-prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
-%
-%\begin{figure}[htbp]
-%  \centering
-%    \includegraphics[width=0.5\textwidth]{figs/%sphinxPredicTS}
-%     \caption{An LMA-based forecast of the last 4,000 points %of a
-%       processor-load performance trace from the {\tt %482.sphinx3}
-%       benchmark.  Red circles and blue $\times$s are the %true and
-%       predicted values, respectively; vertical bars show %where these
-%       values differ.}
-%\label{fig:predsphinx}
-%\end{figure}
-
-%Table~\ref{tab:PredError} presents detailed results about the prediction accuracy of this algorithm on four different examples: the {\tt col\_major} and {\tt 482.sphinx3} programs in Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as another simple microkernel that initializes the same matrix as {\tt col\_major}, but in row-major order, and another complex program ({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.
-Both microkernels were run on the Intel Core Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks were run on the Intel i7\textsuperscript{\textregistered} machine.
-We calculated a figure of merit for each prediction as follows.
-We held back the last $k$ elements\footnote{Several different prediction   horizons were analyzed in our experiment; the results reported in this paper are for $k$=4000} of the $N$ points in each measured time series, built the forecast model by embedding the first $N-k$ points, used that embedding and the LMA method to predict the next $k$ points, then computed the Root Mean Squared Error (RMSE) between the true and predicted signals: $$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$ To compare the success of predictions across signals with different units, we normalized RMSE as follows: $$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
-%
-% The smaller the nRMSE, obviously, the more accurate the prediction.
-
-%The results in Table~\ref{tab:PredError} show a clear distinction between the two microkernels, whose future behavior can be predicted effectively using this simple deterministic modeling strategy, and the more-complex SPEC benchmarks, for which this prediction strategy does not work nearly as well.
-% Removed for space
-% For both processor load (IPC) and memory usage (cache-miss rate),
-% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
-% for \verb|col_major| or \verb|row_major|.
-%
-This begs the question: If these traces all come from deterministic
-systems---computers---then why are they not equally predictable?  Our
-conjecture is that the sheer complexity of the dynamics of the SPEC
-benchmarks running on the Intel i7\textsuperscript{\textregistered}
-machine make them effectively impossible to predict.
-
-%% [[If space: Add a segue sentence: next section uses permutation
-%% entropy to explore that conjecture.]]
-
-%\subsection{Comparing Prediction Accuracy}
-
-\subsection{Autoregressive-integrated-moving average(ARIMA)}
-\subsection{Naive: Random Walk and na\"ive}
-\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE)}
-
-\begin{figure}[htbp]
-  \centering
-   \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccMeanForecast}
-    \caption{\gcc na\"ive }
-    \label{fig:gccMEAN}
-  \end{subfigure}%      
-     \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colMeanForecast}
-    \caption{\col na\"ive }
-    \label{fig:gccMEAN}
-  \end{subfigure}%    
-  \\
-  \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
-    \caption{\gcc ARIMA }
-    \label{fig:gccARIMA}
-  \end{subfigure}%
-  \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
-    \caption{\col ARIMA}
-    \label{fig:colARIMA}
-  \end{subfigure}
-  \\
-      \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
-    \caption{\gcc LMA}
-    \label{fig:gccLMA}
-  \end{subfigure}  
-    \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colLMAForecast}
-    \caption{\col LMA}
-    \label{fig:colLMA}
-  \end{subfigure}
-
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{
-For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
-On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
-}\label{fig:gcc_vs_col}
-  %\end{subfigure}
-\end{figure} 
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
-
-As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
-When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-[[and cherry pick a few examples of \gcc and \col to put in the text For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=6$.]]
-
-% took out for space
-
-\section{Measuring Structural Complexity}\label{sec:meaComplex}
-
-THINGS TO DO
-\begin{enumerate}
-\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
-
-\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-\item justify and explain the regime split in \svd 
-\end{enumerate}
-
-For the purposes of this paper, one can view entropy as a measure of complexity
-and predictability in a time series.  A high-entropy time series is almost
-completely unpredictable---and conversely.  This can be made more rigorous:
-Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
-Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
-$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
-states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
-difficult prediction becomes.
-
-Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
-categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
-\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
-effectively real-valued.  To get around this, one must discretize the data---
-typically achieved by binning.  Unfortunately, this is rarely a good solution to
-the problem, as the binning of the values introduces an additional dynamic on
-top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
-dynamics studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in the face
-of noise and require further understanding of the underlying system, which
-defeats the purpose of measuring the entropy in the first place.
-
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
-complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
-employs a method of discretizing real-valued time series that follows the
-intrinsic behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the Shannon
-entropy, permutation entropy looks at the statistics of the \emph{orderings} of
-sequences of values using ordinal analysis. Ordinal analysis of a time series is
-the process of mapping successive time-ordered elements of a time series to
-their value-ordered permutation of the same size.  By way of example, if $(x_1,
-x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
-other things, it is generally robust to observational noise and requires no
-knowledge of the underlying mechanisms.
-
-\begin{mydef}[Permutation Entropy]
-
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
-  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
-  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
-  = 1,\dots,T}$:
-  \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
-  \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
-  \ge 2$ is defined as
-  \begin{align*}
-  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
-  \end{align*}
-
-\end{mydef}
-
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
-it is common in the literature to normalize permutation entropy as follows:
-$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
-and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
-many large classes of systems \cite{amigo2012permutation}.
-
-Here we will be utilizing a variation of the permutation entropy, the
-\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
-permutation entropy attempts to correct for observational noise which is larger
-than some trends in the data, but smaller than the larger scale features --- for
-example, a signal that switches between two fixed points with noise about those
-fixed points. The weighted permutation entropy would be dominated by the
-switching rather than by the stochastic fluctuation. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
-\begin{align*}
-  w(x_{t+1:t+n}) = \frac{1}{n}
-                 \sum_{x_i \in \{x_{t+1:t+n}\}}
-                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
-\end{align*}
-where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
-$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
-
-The weighted probability of a permutation is then:
-\begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
-\end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
-probability enhances permutations involved in ``large'' features and demotes
-permutations which are small in amplitude relative to the features of the time
-series. The weighted permutation entropy is then:
-\begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
-\end{align*}
-which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
-results of this paper.
-
-In practice, calculating permutation entropy and weighted permutation entropy
-involves choosing a good value for the word length $n$. The primary
-consideration is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the ordinals are
-gathered: e.g.:
-\begin{align*}
-  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
-\end{align*}
-assuming an average of 100 counts per ordinal is sufficient. In the literature,
-$3 \le n \le 6$ is a standard choice --- generally without any formal
-justification. In theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In practice,
-what one should do is calculate the \emph{persistent} permutation entropy by
-increasing $n$ until the result converges, but data length issues can intrude
-before that convergence is reached.
-
-The weighted permutation entropy for the {\tt SVD} program is given in
-Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
-the time series. Within each of those windows, the statistics over words of
-length 4 are computed and the WPE is calculated. The gray bands denote regions
-where the 5,000 value window overlapped visually-distinct regimes. It can be
-seen that the behaviors of the weighted permutation entropy vary between
-regimes.
-[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
-Maybe point at some of the predicting predictability papers.]]
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
-%    are regions where the window overlaps regimes. The window size used is
-%    $5,000 \times 100,000$ instructions and the word length is $4$.}
-%  \label{fig:wwpe}
-%\end{figure}
-
-%%FIGURE INTENT
-%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
-%%%%%%%%%%%%%%%%%
-\begin{figure}[htbp]
-  \centering
-
-  %\begin{subfigure}{0.3\textwidth}
-  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
-  %  \caption{The instructions per cycle of \svd.
-Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
-From left to right each change in color represents a change in regime for 6 regimes in total.
-}
- %   \label{fig:svd_ts}
- % \end{subfigure}%
-  %\\
- % \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
-    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-  \caption{
-[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly.
-Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD.
-The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
-Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE.
-From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
-\end{figure}
-
- \section{Validation of Primary Findings} % (fold)
- \label{sec:results}
-
-%OUTLINE:
-%\begin{enumerate}
-%\item \cmark~Introduce MASE.
-%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
-%its WPE.
-%\item \cmark~Talk about structure analysis. Just because %there is forward information
-%transfer, does not mean that linear predictors can get at %this.
-%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
-%side plot.
-%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
-%\begin{enumerate}
-%\item \cmark~clusters are distributed differently
-%\item \cmark~clusters are shaped differently---tight or not
-%\item \cmark~clusters move differently between LMA and ARIMA
-%\item finally, the diagonal line is important. If you're %below it, you could do
-%better.
-%\end{enumerate}
-%\end{enumerate}
-
-%BRAINSTORMING 
-%\begin{itemize}
-%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
-%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
-
-%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
-
-%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
-
-%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-%\end{itemize}
-
-In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:
-
-\begin{enumerate}
-
-\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.
-
-%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
-
-%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
-
-\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-
-%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
-The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
-
-%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
-
-\end{enumerate}
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\item First paragraph: WPE is a good measure of predictability.  Figure:
-%best athlete MASE vs. its WPE.
-
-%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-
-As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task.
-As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
-Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.
-
-In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation.
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
-  \caption{The best MASE among all runs and prediction methods vs weighted
-  permutation entropy. For each of these, the word length used is $6$. The
-  dashed line is a least-squares linear fit of all the points except for {\tt
-  SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_wpe}
-\end{figure}
-What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}.
-We will not further validate this claim through some examples.
-
-To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively.
-In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
-We also list all of these values in Table \ref{tab:error}
-
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 6 for each signal.}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$&  $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$&  $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &   $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ &  $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$&  $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$&  $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. }
-    \label{fig:lma_pred_vs_ent}
-  \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:arima_pred_vs_ent}
-  \end{subfigure}
-\caption{
-For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}
-\end{figure}
-
-We will start with the lowest complexity program: \col.
-Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}.
-If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
-This would suggest that there is some predictive structure present but not that much.
-However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity.
-This is in fact the case.
-\col is a complex signal, i.e., chaotic, but it is complex in a structured way.
-Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
-The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
-
-Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive.
-This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
-However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
-The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
-This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict.
-However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA.
-If we do this we get forecasts which score 1.5 times better than the random walk.
-Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.
-
-Finally, we consider the most complex program: \gcc.
-The complexity of \gcc is $0.94 \pm 0.001$.
-This implies that almost no predictive structure exists in this structure.
-Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future.
-This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex.
-In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step.
-This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.
-
-There is a single outlier in this study which is \svdone.
-This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively).
-The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts.
-We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method.
-MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
-In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
-A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.
-
-To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail.
-This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
-We provide a small snippet as opposed to the full time series simply for visual clarity.
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=\textwidth]{figs/svdonets2}
-\caption{ For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}
-\end{figure} 
-As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide.
-Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
-The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}.
-These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle).
-As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
-At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal.
-This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
-The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.
-
-[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}.
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
-
-%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration.
-The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method).
-On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values.
-As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE.
-The WPE is sensitive to both linear and nonlinear structure.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
-So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
-
-%is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
-How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict.
-If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy.
-However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-
-Other noteworthy features of the LMA and ARIMA results are the cluster locations
-and distributions. The WPE values of each run of any particular program tend to
-have little variance, leading to the clusters in
-Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
-constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
-low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
-traces, however, have a large variance resulting in the spread seen in
-Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
-bad, in that other predictors vastly outperform it, but they are inconsistent.
-Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
-see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
-the right than those in Figure~\ref{fig:lma_pred_vs_ent}.
-
-\section{ Conclusions \& Future Work}\label{sec:conc}
-
-\begin{enumerate}
-\item  This need to be rewritten to conclude and address this paper...
-\end{enumerate}
-
-The results presented here suggest that permutation entropy---a ordinal
-calculation of forward information transfer in a time series---is an effective
-metric for predictability of computer performance traces. Experimentally, traces
-with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
-may overshadow the inherent determinism in the system dynamics, whereas traces
-with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
-of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
-0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
-corroborating the results of~\cite{mytkowicz09}.
-
-If information is the limit, then gathering and using more information is an
-obvious next step.  There is an equally obvious tension here between data length
-and prediction speed: a forecast that requires half a second to compute is not
-useful for the purposes of real-time control of a computer system with a MHz
-clock rate.  Another alternative is to sample several system variables
-simultaneously and build multivariate delay-coordinate embeddings.  Existing
-approaches to that are computationally prohibitive
-\cite{cao-multivariate-embedding}.  We are working on alternative
-methods that sidestep that complexity.
-
-%%This use of permutaiotn entropy is a useful application of information
-%%theory to computer performance modeling because it gives you a simple
-%%and fast way to decide whether building a model and trying to forecast
-%%the future is worthwhile or if guessing the mean is just as effective.
-%%End with a sentence tying back to power management and world peace.
-
-%\begin{it}
-%Summarize the results.
-
-%Paragraphs on the issues that come up, including one about the "amount
-%of info" one: if one could sample more variables, for instance, one
-%might be able to do a better job of predicting more-complex traces.
-%Segue to some handwaving about multivariable LMA models; tie this back
-%to the "computers are NLD systems" stuff in the intro.  This is a real
-%challenge; current approaches to this modelling problem have the major
-%issue of taking way too long to build.  And that's a big issue if
-%you're trying not just to classify, but to predict.  In a system that
-%runs at MHz speeds, a prediction that takes milliseconds to compute is
-%not useful.
-%\end{it}
-
-\section*{Acknowledgment}
-This work was partially supported by NSF grant \#CMMI-1245947 and ARO
-grant \#W911NF-12-1-0288.
-
-\bibliographystyle{unsrt}
-\bibliography{bibliofile}
-
-\end{document}

### 2014-02-19 model.tex
Split Sections to help with conflicts

--- 2014-02-18 pecomplexity.tex
@@
-\documentclass{article}
-\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-%\geometry{landscape}                % Activate for for rotated page geometry
-%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{epstopdf}
-\usepackage{amsthm}
-
-\usepackage{pifont} % check and x marks
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
-
-\newcommand{\gcc}{{\tt 403.gcc}~}
-\newcommand{\svd}{{\tt dgesdd}~}
-\newcommand{\svdone}{{\tt dgesdd}$_1$~}
-\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
-\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
-\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
-\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
-\newcommand{\svdsix}{{\tt dgesdd$_6$}~}
-
-\newcommand{\row}{{\tt row\_major}~}
-\newcommand{\col}{{\tt col\_major}~}
-
-\usepackage{authblk}
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
-\newtheorem*{mydef}{Definition}
-\title{Determinism, Complexity, and Predictability in Computer Performance\\
-Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity
-\\ Quantifying Predictability using Structural Complexity
-}
-
-\author[1]{Joshua Garland \thanks{joshua.garland@colorado.edu}}
-\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
-\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-\affil[1]{Department of Computer Science\\
-  University of Colorado at Boulder\\
-  Colorado, USA
-}
-\affil[2]{Santa Fe Institute\\
-  New Mexico, USA
-}
-
-%Version 1
-%\author{
-%  \IEEEauthorblockN{Joshua Garland}
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: joshua.garland@colorado.edu}
-%  \and
-%  \IEEEauthorblockN{Ryan G.~James}
-%  \IEEEauthorblockA{Complexity Sciences Center \& Dept. of Physics\\
-%    University of California at Davis\\
-%    California, USA\\
-%    Email: rgjames@ucdavis.edu}
-%    \and
-%      \IEEEauthorblockN{Elizabeth Bradley}
-%         \IEEEauthorblockA{Santa Fe Institute \\
-%    New Mexico, USA\\
-%    }
-%  \IEEEauthorblockA{Dept. of Computer Science\\
-%    University of Colorado at Boulder\\
-%    Colorado, USA\\
-%    Email: lizb@colorado.edu}
-%  }
-
- %\author{
- %  \IEEEauthorblockN{
- %    Joshua Garland\IEEEauthorrefmark{1},
- %    Ryan G.~James\IEEEauthorrefmark{1} and
- %    Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{1}Dept. of Computer Science
- %    University of Colorado, Boulder, Colorado 80309-0430 USA\\
- %    Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
- %  \IEEEauthorblockA{
- %    \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
- %    Email: lizb@colorado.edu}
- %  % \IEEEauthorblockA{
- %  %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, %California 95616 USA\\
-%   %   Email: rgjames@ucdavis.edu}
-% }
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
-  Among other things, that implies that one should be able to use
-  deterministic forecast rules to predict aspects of their behavior.
-  That statement is sometimes---but not always---true. The memory and
-  processor loads of some simple programs are easy to predict, for
-  example, but those of more-complex programs like \gcc are not.
-%%%%%%%%%%%%%%%%%%%%%%
-%% I had to change all the \verb|blah| entries because they caused
-%% latex to barf if they were in figure captions.  Odd bug.
-%%%%%%%%%%%%%%%%%%%%%%
-  The goal of this paper is to determine why that is the case. We
-  conjecture that, in practice, complexity can effectively overwhelm
-  the predictive power of deterministic forecast models. To explore
-  that, we build models of a number of performance traces from
-  different programs running on different Intel-based computers. We
-  then calculate the \emph{permutation entropy}---a temporal entropy
-  metric that uses ordinal analysis---of those traces and correlate
-  those values against the prediction success.
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-BRAINSTORMING 
-\begin{itemize}
-\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
-\item random walk for example is best predicted by guess what just happend[[move to introduction]]
-\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]]
-
-\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-\end{itemize}
-
-%\begin{it}
-%Paragraph on computer performance, including citations to Todd paper
-%and summary of the results that indicate that they're deterministic
-%nonlinear dynamical systems.  Given that, we should be able to
-%predict.  What benefits would accrue if we could do so: power mgmt,
-%end world hunger [[this is my primary goal everyday :)]], etc.
-%\end{it}
-Things to add to introduction
-\begin{enumerate}
-\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
-\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.
-
-\item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc
-
-\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
-\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
-\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex.
-Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not.
-Maybe even having 4 predictions.
-top being ARIMA of the above signals and bottom being LMA of the above signals.
-Show that one improved and one did not.
-Is it that we used the wrong method to predict or is it that we simply can't predict the signal better than a random walk due to high levels of internal signal complexity.
-
-\item Introduce the two main contributions of the paper which are outlined at the begining of the results section
-
-\end{enumerate}
-
-Computers are among the most complex engineered artifacts in current
-use.  Modern microprocessor chips contain multiple processing units
-and multi-layer memories, for instance, and they use complicated
-hardware/software strategies to move data and threads of computation
-across those resources.  These features---along with all the others
-that go into the design of these chips---make the patterns of their
-processor loads and memory accesses highly complex and hard to
-predict.  Accurate forecasts of these quantities, if one could
-construct them, could be used to improve computer design.  If one
-could predict that a particular computational thread would be bogged
-down for the next 0.6 seconds waiting for data from main memory, for
-instance, one could save power by putting that thread on hold for that
-time period (e.g., by migrating it to a processing unit whose clock
-speed is scaled back).  Computer performance traces are, however, very
-complex.  Even a simple ``microkernel,'' like a three-line loop that
-repeatedly initializes a matrix in column-major order, can produce
-{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
-predictability.
-%
- \begin{figure}[htbp]
-    \centering
-    \includegraphics[width=\textwidth]{figs/colshortts}
-    % where an .eps filename suffix will be assumed under latex,
-    % and a .pdf suffix will be assumed for pdflatex
-    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
-        \col}, a three-line C program that repeatedly initializes
-      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
-      simple program exhibits chaotic performance dynamics.}
-   \label{fig:ipc}
-  \end{figure}
-
-The computer systems community has applied a variety of prediction
-strategies to traces like this, most of which employ regression.  An
-appealing alternative builds on the recently established fact that
-computers can be effectively modeled as deterministic nonlinear
-dynamical systems \cite{mytkowicz09}.  This result implies the
-existence of a deterministic forecast rule for those dynamics.  In
-particular, one can use \emph{delay-coordinate embedding} to
-reconstruct the underlying dynamics of computer performance, then use
-the resulting model to forecast the future values of computer
-performance metrics such as memory or processor loads
-\cite{josh-ida2011}.  In the case of simple microkernels like the one
-that produced the trace in Figure~\ref{fig:ipc}, this deterministic
-modeling and forecast strategy works very well.  In more-complicated
-programs, however, such as speech recognition software or compilers,
-this forecast strategy---as well as the traditional methods---break
-down quickly.
-
-This paper is a first step in understanding when, why, and how
-deterministic forecast strategies fail when they are applied to
-deterministic systems.  We focus here on the specific example of
-computer performance.  We conjecture that the complexity of traces
-from these systems---which results from the inherent dimension,
-non-linearity, and non-stationarity of the dynamics, as well as from
-measurement issues like noise, aggregation, and finite data
-length---can make those deterministic signals \emph{effectively}
-unpredictable.  We argue that \emph{permutation entropy}
-\cite{bandt2002per}, a method for measuring the entropy of a
-real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study four
-examples---two simple microkernels and two complex programs from the
-SPEC benchmark suite---running on different Intel-based machines.  For
-each program, we calculate the permutation entropy of the processor
-load (instructions per cycle) and memory-use efficiency (cache-miss
-rates), then compare that to the prediction accuracy attainable for
-that trace using a simple deterministic model.
-
-% paragraph to appease the theoretician in me
-It is worth taking a moment to consider the theoretical possibility of
-this task. We are not attempting to predict the state of the CPU at an
-arbitrary point in the future --- this, at least with perfect
-accuracy, would be tantamount to solving the halting problem. What we
-are attempting is to predict aspects or functions of the running of
-the CPU: instructions executed per second, cache misses per 100,000
-instructions, and similar statistics. Prediction of these quantities
-at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem.
-
-% The rest of the paper is organized as follows.
-% Section~\ref{sec:compModel} describes the experimental setup, as well
-% as the nonlinear modeling and forecast strategies.  In
-% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
-% its value for a number of different computer performance traces, and
-% compare the results to the prediction accuracy.  In
-% Section~\ref{sec:conc}, we discuss these results and their
-% implications in regard to our conjecture, and consider future areas of
-% research.
-
-\section{Experimental Methods }
-
-\subsection{Time Series Collection}
- Experimental methods (how we collect the time series and what the times series are
-This should be HPM PAPI, which programs we model
-
-\subsection{The Programs}
-Include traces of full \col \svd and \gcc
-\begin{figure}[htbp]
-  \centering
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/colFullTS}
-    \caption{\col Time Series}
-    \label{fig:col-ts}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
-    \caption{\gcc Time Series}
-    \label{fig:gcc-ts}
-  \end{subfigure}
-  \caption{In (a) the instructions executed per CPU clock cycle
-    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
-    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
-\end{figure}
-
 \section{Modeling}
@@
 % took out for space
-
-\section{Measuring Structural Complexity}\label{sec:meaComplex}
-
-THINGS TO DO
-\begin{enumerate}
-\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
-
-\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-\item justify and explain the regime split in \svd 
-\end{enumerate}
-
-For the purposes of this paper, one can view entropy as a measure of complexity
-and predictability in a time series.  A high-entropy time series is almost
-completely unpredictable---and conversely.  This can be made more rigorous:
-Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
-Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
-$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
-states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
-difficult prediction becomes.
-
-Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
-categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
-\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
-effectively real-valued.  To get around this, one must discretize the data---
-typically achieved by binning.  Unfortunately, this is rarely a good solution to
-the problem, as the binning of the values introduces an additional dynamic on
-top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
-dynamics studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in the face
-of noise and require further understanding of the underlying system, which
-defeats the purpose of measuring the entropy in the first place.
-
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
-complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
-employs a method of discretizing real-valued time series that follows the
-intrinsic behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the Shannon
-entropy, permutation entropy looks at the statistics of the \emph{orderings} of
-sequences of values using ordinal analysis. Ordinal analysis of a time series is
-the process of mapping successive time-ordered elements of a time series to
-their value-ordered permutation of the same size.  By way of example, if $(x_1,
-x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
-other things, it is generally robust to observational noise and requires no
-knowledge of the underlying mechanisms.
-
-\begin{mydef}[Permutation Entropy]
-
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
-  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
-  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
-  = 1,\dots,T}$:
-  \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
-  \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
-  \ge 2$ is defined as
-  \begin{align*}
-  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
-  \end{align*}
-
-\end{mydef}
-
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
-it is common in the literature to normalize permutation entropy as follows:
-$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
-and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
-many large classes of systems \cite{amigo2012permutation}.
-
-Here we will be utilizing a variation of the permutation entropy, the
-\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
-permutation entropy attempts to correct for observational noise which is larger
-than some trends in the data, but smaller than the larger scale features --- for
-example, a signal that switches between two fixed points with noise about those
-fixed points. The weighted permutation entropy would be dominated by the
-switching rather than by the stochastic fluctuation. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
-\begin{align*}
-  w(x_{t+1:t+n}) = \frac{1}{n}
-                 \sum_{x_i \in \{x_{t+1:t+n}\}}
-                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
-\end{align*}
-where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
-$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
-
-The weighted probability of a permutation is then:
-\begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
-\end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
-probability enhances permutations involved in ``large'' features and demotes
-permutations which are small in amplitude relative to the features of the time
-series. The weighted permutation entropy is then:
-\begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
-\end{align*}
-which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
-results of this paper.
-
-In practice, calculating permutation entropy and weighted permutation entropy
-involves choosing a good value for the word length $n$. The primary
-consideration is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the ordinals are
-gathered: e.g.:
-\begin{align*}
-  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
-\end{align*}
-assuming an average of 100 counts per ordinal is sufficient. In the literature,
-$3 \le n \le 6$ is a standard choice --- generally without any formal
-justification. In theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In practice,
-what one should do is calculate the \emph{persistent} permutation entropy by
-increasing $n$ until the result converges, but data length issues can intrude
-before that convergence is reached.
-
-The weighted permutation entropy for the {\tt SVD} program is given in
-Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
-the time series. Within each of those windows, the statistics over words of
-length 4 are computed and the WPE is calculated. The gray bands denote regions
-where the 5,000 value window overlapped visually-distinct regimes. It can be
-seen that the behaviors of the weighted permutation entropy vary between
-regimes.
-[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
-Maybe point at some of the predicting predictability papers.]]
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
-%    are regions where the window overlaps regimes. The window size used is
-%    $5,000 \times 100,000$ instructions and the word length is $4$.}
-%  \label{fig:wwpe}
-%\end{figure}
-
-%%FIGURE INTENT
-%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
-%%%%%%%%%%%%%%%%%
-\begin{figure}[htbp]
-  \centering
-
-  %\begin{subfigure}{0.3\textwidth}
-  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
-  %  \caption{The instructions per cycle of \svd.
-Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
-From left to right each change in color represents a change in regime for 6 regimes in total.
-}
- %   \label{fig:svd_ts}
- % \end{subfigure}%
-  %\\
- % \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
-    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-  \caption{
-[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly.
-Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD.
-The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
-Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE.
-From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
-\end{figure}
-
- \section{Validation of Primary Findings} % (fold)
- \label{sec:results}
-
-%OUTLINE:
-%\begin{enumerate}
-%\item \cmark~Introduce MASE.
-%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
-%its WPE.
-%\item \cmark~Talk about structure analysis. Just because %there is forward information
-%transfer, does not mean that linear predictors can get at %this.
-%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
-%side plot.
-%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
-%\begin{enumerate}
-%\item \cmark~clusters are distributed differently
-%\item \cmark~clusters are shaped differently---tight or not
-%\item \cmark~clusters move differently between LMA and ARIMA
-%\item finally, the diagonal line is important. If you're %below it, you could do
-%better.
-%\end{enumerate}
-%\end{enumerate}
-
-%BRAINSTORMING 
-%\begin{itemize}
-%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
-%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
-
-%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
-
-%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
-
-%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-%\end{itemize}
-
-In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:
-
-\begin{enumerate}
-
-\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.
-
-%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
-
-%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
-
-\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-
-%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
-The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
-
-%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
-
-\end{enumerate}
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\item First paragraph: WPE is a good measure of predictability.  Figure:
-%best athlete MASE vs. its WPE.
-
-%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-
-As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task.
-As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
-Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.
-
-In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation.
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
-  \caption{The best MASE among all runs and prediction methods vs weighted
-  permutation entropy. For each of these, the word length used is $6$. The
-  dashed line is a least-squares linear fit of all the points except for {\tt
-  SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_wpe}
-\end{figure}
-What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}.
-We will not further validate this claim through some examples.
-
-To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively.
-In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
-We also list all of these values in Table \ref{tab:error}
-
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 6 for each signal.}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$&  $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$&  $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &   $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ &  $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$&  $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$&  $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. }
-    \label{fig:lma_pred_vs_ent}
-  \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:arima_pred_vs_ent}
-  \end{subfigure}
-\caption{
-For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}
-\end{figure}
-
-We will start with the lowest complexity program: \col.
-Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}.
-If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
-This would suggest that there is some predictive structure present but not that much.
-However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity.
-This is in fact the case.
-\col is a complex signal, i.e., chaotic, but it is complex in a structured way.
-Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
-The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
-
-Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive.
-This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
-However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
-The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
-This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict.
-However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA.
-If we do this we get forecasts which score 1.5 times better than the random walk.
-Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.
-
-Finally, we consider the most complex program: \gcc.
-The complexity of \gcc is $0.94 \pm 0.001$.
-This implies that almost no predictive structure exists in this structure.
-Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future.
-This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex.
-In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step.
-This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.
-
-There is a single outlier in this study which is \svdone.
-This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively).
-The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts.
-We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method.
-MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
-In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
-A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.
-
-To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail.
-This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
-We provide a small snippet as opposed to the full time series simply for visual clarity.
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=\textwidth]{figs/svdonets2}
-\caption{ For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}
-\end{figure} 
-As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide.
-Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
-The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}.
-These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle).
-As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
-At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal.
-This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
-The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.
-
-[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}.
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
-
-%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration.
-The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method).
-On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values.
-As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE.
-The WPE is sensitive to both linear and nonlinear structure.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
-So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
-
-%is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
-How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict.
-If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy.
-However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-
-Other noteworthy features of the LMA and ARIMA results are the cluster locations
-and distributions. The WPE values of each run of any particular program tend to
-have little variance, leading to the clusters in
-Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
-constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
-low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
-traces, however, have a large variance resulting in the spread seen in
-Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
-bad, in that other predictors vastly outperform it, but they are inconsistent.
-Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
-see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
-the right than those in Figure~\ref{fig:lma_pred_vs_ent}.
-
-\section{ Conclusions \& Future Work}\label{sec:conc}
-
-\begin{enumerate}
-\item  This need to be rewritten to conclude and address this paper...
-\end{enumerate}
-
-The results presented here suggest that permutation entropy---a ordinal
-calculation of forward information transfer in a time series---is an effective
-metric for predictability of computer performance traces. Experimentally, traces
-with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
-may overshadow the inherent determinism in the system dynamics, whereas traces
-with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
-of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
-0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
-corroborating the results of~\cite{mytkowicz09}.
-
-If information is the limit, then gathering and using more information is an
-obvious next step.  There is an equally obvious tension here between data length
-and prediction speed: a forecast that requires half a second to compute is not
-useful for the purposes of real-time control of a computer system with a MHz
-clock rate.  Another alternative is to sample several system variables
-simultaneously and build multivariate delay-coordinate embeddings.  Existing
-approaches to that are computationally prohibitive
-\cite{cao-multivariate-embedding}.  We are working on alternative
-methods that sidestep that complexity.
-
-%%This use of permutaiotn entropy is a useful application of information
-%%theory to computer performance modeling because it gives you a simple
-%%and fast way to decide whether building a model and trying to forecast
-%%the future is worthwhile or if guessing the mean is just as effective.
-%%End with a sentence tying back to power management and world peace.
-
-%\begin{it}
-%Summarize the results.
-
-%Paragraphs on the issues that come up, including one about the "amount
-%of info" one: if one could sample more variables, for instance, one
-%might be able to do a better job of predicting more-complex traces.
-%Segue to some handwaving about multivariable LMA models; tie this back
-%to the "computers are NLD systems" stuff in the intro.  This is a real
-%challenge; current approaches to this modelling problem have the major
-%issue of taking way too long to build.  And that's a big issue if
-%you're trying not just to classify, but to predict.  In a system that
-%runs at MHz speeds, a prediction that takes milliseconds to compute is
-%not useful.
-%\end{it}
-
-\section*{Acknowledgment}
-This work was partially supported by NSF grant \#CMMI-1245947 and ARO
-grant \#W911NF-12-1-0288.
-
-\bibliographystyle{unsrt}
-\bibliography{bibliofile}
-
-\end{document}

### 2014-02-19 pecomplexity.tex
Split Sections to help with conflicts

--- 2014-02-16 pecomplexity.tex
@@
 
-\section{Introduction}\label{sec:intro}
+\input{intro.tex}
 
-BRAINSTORMING 
-\begin{itemize}
-\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
-\item random walk for example is best predicted by guess what just happend[[move to introduction]]
-\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]]
+\input{methods.tex}
 
-\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
+\input{model.tex}
 
-\end{itemize}
+\input{wpe.tex}
 
-%\begin{it}
-%Paragraph on computer performance, including citations to Todd paper
-%and summary of the results that indicate that they're deterministic
-%nonlinear dynamical systems.  Given that, we should be able to
-%predict.  What benefits would accrue if we could do so: power mgmt,
-%end world hunger [[this is my primary goal everyday :)]], etc.
-%\end{it}
-Things to add to introduction
-\begin{enumerate}
-\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
-\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.
+\input{results.tex}
 
-\item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc
-
-\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
-\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
-\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex.
-Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not.
-Maybe even having 4 predictions.
-top being ARIMA of the above signals and bottom being LMA of the above signals.
-Show that one improved and one did not.
-Is it that we used the wrong method to predict or is it that we simply can't predict the signal better than a random walk due to high levels of internal signal complexity.
-
-\item Introduce the two main contributions of the paper which are outlined at the begining of the results section
-
-\end{enumerate}
-
-Computers are among the most complex engineered artifacts in current
-use.  Modern microprocessor chips contain multiple processing units
-and multi-layer memories, for instance, and they use complicated
-hardware/software strategies to move data and threads of computation
-across those resources.  These features---along with all the others
-that go into the design of these chips---make the patterns of their
-processor loads and memory accesses highly complex and hard to
-predict.  Accurate forecasts of these quantities, if one could
-construct them, could be used to improve computer design.  If one
-could predict that a particular computational thread would be bogged
-down for the next 0.6 seconds waiting for data from main memory, for
-instance, one could save power by putting that thread on hold for that
-time period (e.g., by migrating it to a processing unit whose clock
-speed is scaled back).  Computer performance traces are, however, very
-complex.  Even a simple ``microkernel,'' like a three-line loop that
-repeatedly initializes a matrix in column-major order, can produce
-{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
-predictability.
-%
- \begin{figure}[htbp]
-    \centering
-    \includegraphics[width=\textwidth]{figs/colshortts}
-    % where an .eps filename suffix will be assumed under latex,
-    % and a .pdf suffix will be assumed for pdflatex
-    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
-        \col}, a three-line C program that repeatedly initializes
-      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
-      simple program exhibits chaotic performance dynamics.}
-   \label{fig:ipc}
-  \end{figure}
-
-\begin{figure}[htbp]
-  \centering
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/colshortts}
-    \caption{A short segment of the instructions executed per CPU clock cycle
-    (IPC) during the execution of \col. Each point is the IPC during a 100,000
-    cycle period. There is a great deal of periodicity in this time series.}
-    \label{fig:col-ts}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
-    \caption{The full IPC time series during the execution of \gcc. This time
-    series has very little structure.}
-    \label{fig:gcc-ts}
-  \end{subfigure}
-  \label{fig:sample-ts}
-\end{figure}
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
-    \caption{\gcc ARIMA }
-    \label{fig:gccARIMA}
-  \end{subfigure}%
-    \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
-    \caption{\gcc LMA}
-    \label{fig:gccLMA}
-  \end{subfigure}
-  \\
-  \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
-    \caption{\col ARIMA}
-    \label{fig:colARIMA}
-  \end{subfigure}
-    \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colLMAForecast}
-    \caption{\col LMA}
-    \label{fig:colLMA}
-  \end{subfigure}
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{
-For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
-On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
-}\label{fig:gcc_vs_col}
-  %\end{subfigure}
-\end{figure}
-
-The computer systems community has applied a variety of prediction
-strategies to traces like this, most of which employ regression.  An
-appealing alternative builds on the recently established fact that
-computers can be effectively modeled as deterministic nonlinear
-dynamical systems \cite{mytkowicz09}.  This result implies the
-existence of a deterministic forecast rule for those dynamics.  In
-particular, one can use \emph{delay-coordinate embedding} to
-reconstruct the underlying dynamics of computer performance, then use
-the resulting model to forecast the future values of computer
-performance metrics such as memory or processor loads
-\cite{josh-ida2011}.  In the case of simple microkernels like the one
-that produced the trace in Figure~\ref{fig:ipc}, this deterministic
-modeling and forecast strategy works very well.  In more-complicated
-programs, however, such as speech recognition software or compilers,
-this forecast strategy---as well as the traditional methods---break
-down quickly.
-
-This paper is a first step in understanding when, why, and how
-deterministic forecast strategies fail when they are applied to
-deterministic systems.  We focus here on the specific example of
-computer performance.  We conjecture that the complexity of traces
-from these systems---which results from the inherent dimension,
-non-linearity, and non-stationarity of the dynamics, as well as from
-measurement issues like noise, aggregation, and finite data
-length---can make those deterministic signals \emph{effectively}
-unpredictable.  We argue that \emph{permutation entropy}
-\cite{bandt2002per}, a method for measuring the entropy of a
-real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study four
-examples---two simple microkernels and two complex programs from the
-SPEC benchmark suite---running on different Intel-based machines.  For
-each program, we calculate the permutation entropy of the processor
-load (instructions per cycle) and memory-use efficiency (cache-miss
-rates), then compare that to the prediction accuracy attainable for
-that trace using a simple deterministic model.
-
-% paragraph to appease the theoretician in me
-It is worth taking a moment to consider the theoretical possibility of
-this task. We are not attempting to predict the state of the CPU at an
-arbitrary point in the future --- this, at least with perfect
-accuracy, would be tantamount to solving the halting problem. What we
-are attempting is to predict aspects or functions of the running of
-the CPU: instructions executed per second, cache misses per 100,000
-instructions, and similar statistics. Prediction of these quantities
-at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem.
-
-% The rest of the paper is organized as follows.
-% Section~\ref{sec:compModel} describes the experimental setup, as well
-% as the nonlinear modeling and forecast strategies.  In
-% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
-% its value for a number of different computer performance traces, and
-% compare the results to the prediction accuracy.  In
-% Section~\ref{sec:conc}, we discuss these results and their
-% implications in regard to our conjecture, and consider future areas of
-% research.
-
-\section{Experimental Methods and Modelling}\label{sec:compModel}
-
-Section outline:
-
-\begin{enumerate}
-\item Experimental methods (how we collect the time series and what the times series are
-\item Description of DCE and parameter estimation  
-\item Description of auto ARIMA 
-\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
-\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
-\item Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
-
-\end{enumerate}
-
-% took out for space
- \subsection{Reconstructing hidden dynamics}
-
-Delay-coordinate embedding allows one to reconstruct a system's full
-state-space dynamics from a \emph{single} scalar time-series
-measurement---provided that some conditions hold regarding that data.
-Specifically, if the underlying dynamics and the measurement
-function---the mapping from the unknown state vector $\vec{X}$ to the
-scalar value $x$ that one is measuring---are both smooth and generic,
-Takens~\cite{takens} formally proves that the delay-coordinate map
-\[
-F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
-\]
-from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
-where $t$ is time, is a diffeomorphism on $M$---in other words, that
-the reconstructed dynamics and the true (hidden) dynamics have the
-same topology.
-
-This is an extremely powerful result: among other things, it means
-that one can build a formal model of the full system dynamics without
-measuring (or even knowing) every one of its state variables.  This is
-the foundation of the modeling approach that is used in this paper.
-The first step in the process is to estimate values for the two free
-parameters in the delay-coordinate map: the delay $\tau$ and the
-dimension $m$.  We follow standard procedures for this, choosing the
-first minimum in the average mutual information as an estimate of
-$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
-plot of the data from Figure~\ref{fig:ipc}, embedded following this
-procedure, is shown in Figure~\ref{fig:embedding}.
-
- \begin{figure}
-   \centering
-     \includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{A 3D projection of a delay-coordinate embedding of the trace
- from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
- }
- \label{fig:embedding}
- \end{figure}
-
-%% can cut for space if need be:
-The coordinates of each point on this plot are differently delayed
-elements of the \verb|col_major| L2 cache miss rate time series
-$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
-$y(t+2\tau)$ on the third, and so on.
-%% ...down to here.
-Structure in these kinds of plots---clearly visible in
-Figure~\ref{fig:embedding}---is an indication of
-determinism\footnote{A deeper analysis of
-  Figure~\ref{fig:embedding}---as alluded to on the previous
-  page---supports that diagnosis, confirming the presence of a chaotic
-  attractor in these cache-miss dynamics, with largest Lyapunov
-  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
-  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
-structure can also be used to build a forecast model.
-
-% took out for space
- \subsection{LMA: Using dynamics in forecasting}
-
-Given a nonlinear model of a deterministic dynamical system in the
-form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
-one can build deterministic forecast algorithms by capturing and
-exploiting the geometry of the embedding.  Many techniques have been
-developed by the dynamical systems community for this purpose
-(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
-is the ``Lorenz method of analogues'' (LMA), which is essentially
-nearest-neighbor prediction in the embedded state
-space~\cite{lorenz-analogues}.  Even this simple algorithm---which
-builds predictions by finding the nearest neighbor in the embedded
-space of the given point, then taking that neighbor's path as the
-forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
-shown in Figure~\ref{fig:cachePredTS}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
-    \caption{A forecast of the last 4,000 points of the signal in
-      Figure~\ref{fig:ipc} using an LMA-based strategy on the
-      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
-      $\times$s are the true and predicted values, respectively;
-      vertical bars show where these values differ. }
-\label{fig:cachePredTS}
-\end{figure}
-%
-On the other hand, if we use the same approach to forecast the
-processor load\footnote{Instructions per cycle, or IPC} of the {\tt
-  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
-on an Intel i7\textsuperscript{\textregistered}-based machine, the
-prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
-%
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
-     \caption{An LMA-based forecast of the last 4,000 points of a
-       processor-load performance trace from the {\tt 482.sphinx3}
-       benchmark.  Red circles and blue $\times$s are the true and
-       predicted values, respectively; vertical bars show where these
-       values differ.}
-\label{fig:predsphinx}
-\end{figure}
-
-Table~\ref{tab:PredError} presents detailed results about the
-prediction accuracy of this algorithm on four different examples: the
-{\tt col\_major} and {\tt 482.sphinx3} programs in
-Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as
-another simple microkernel that initializes the same matrix as {\tt
-  col\_major}, but in row-major order, and another complex program
-({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.  Both
-microkernels were run on the Intel Core
-Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks
-were run on the Intel i7\textsuperscript{\textregistered} machine.  We
-calculated a figure of merit for each prediction as follows.  We held
-back the last $k$ elements\footnote{Several different prediction
-  horizons were analyzed in our experiment; the results reported in
-  this paper are for $k$=4000} of the $N$ points in each measured time
-series, built the forecast model by embedding the first $N-k$ points,
-used that embedding and the LMA method to predict the next $k$ points,
-then computed the Root Mean Squared Error (RMSE) between the true and
-predicted signals:
-$$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$
-%
-To compare the success of predictions across signals with different
-units, we normalized RMSE as follows:
-$$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
-%
-% The smaller the nRMSE, obviously, the more accurate the prediction.
-
-The results in Table~\ref{tab:PredError} show a clear distinction
-between the two microkernels, whose future behavior can be
-predicted effectively using this simple deterministic modeling
-strategy, and the more-complex SPEC benchmarks, for which this
-prediction strategy does not work nearly as well.
-% Removed for space
-% For both processor load (IPC) and memory usage (cache-miss rate),
-% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
-% for \verb|col_major| or \verb|row_major|.
-%
-This begs the question: If these traces all come from deterministic
-systems---computers---then why are they not equally predictable?  Our
-conjecture is that the sheer complexity of the dynamics of the SPEC
-benchmarks running on the Intel i7\textsuperscript{\textregistered}
-machine make them effectively impossible to predict.
-
-%% [[If space: Add a segue sentence: next section uses permutation
-%% entropy to explore that conjecture.]]
-
-\subsection{Comparing Prediction Accuracy}
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
-
-As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
-When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.
-
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
-
-\section{Measuring Structural Complexity}\label{sec:meaComplex}
-
-THINGS TO DO
-\begin{enumerate}
-\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
-
-\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-\item justify and explain the regime split in \svd 
-\end{enumerate}
-
-For the purposes of this paper, one can view entropy as a measure of complexity
-and predictability in a time series.  A high-entropy time series is almost
-completely unpredictable---and conversely.  This can be made more rigorous:
-Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
-Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
-$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
-states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
-difficult prediction becomes.
-
-Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
-categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
-\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
-effectively real-valued.  To get around this, one must discretize the data---
-typically achieved by binning.  Unfortunately, this is rarely a good solution to
-the problem, as the binning of the values introduces an additional dynamic on
-top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
-dynamics studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in the face
-of noise and require further understanding of the underlying system, which
-defeats the purpose of measuring the entropy in the first place.
-
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
-complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
-employs a method of discretizing real-valued time series that follows the
-intrinsic behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the Shannon
-entropy, permutation entropy looks at the statistics of the \emph{orderings} of
-sequences of values using ordinal analysis. Ordinal analysis of a time series is
-the process of mapping successive time-ordered elements of a time series to
-their value-ordered permutation of the same size.  By way of example, if $(x_1,
-x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
-other things, it is generally robust to observational noise and requires no
-knowledge of the underlying mechanisms.
-
-\begin{mydef}[Permutation Entropy]
-
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
-  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
-  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
-  = 1,\dots,T}$:
-  \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
-  \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
-  \ge 2$ is defined as
-  \begin{align*}
-  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
-  \end{align*}
-
-\end{mydef}
-
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
-it is common in the literature to normalize permutation entropy as follows:
-$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
-and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
-many large classes of systems \cite{amigo2012permutation}.
-
-Here we will be utilizing a variation of the permutation entropy, the
-\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
-permutation entropy attempts to correct for observational noise which is larger
-than some trends in the data, but smaller than the larger scale features --- for
-example, a signal that switches between two fixed points with noise about those
-fixed points. The weighted permutation entropy would be dominated by the
-switching rather than by the stochastic fluctuation. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
-\begin{align*}
-  w(x_{t+1:t+n}) = \frac{1}{n}
-                 \sum_{x_i \in \{x_{t+1:t+n}\}}
-                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
-\end{align*}
-where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
-$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
-
-The weighted probability of a permutation is then:
-\begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
-\end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
-probability enhances permutations involved in ``large'' features and demotes
-permutations which are small in amplitude relative to the features of the time
-series. The weighted permutation entropy is then:
-\begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
-\end{align*}
-which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
-results of this paper.
-
-In practice, calculating permutation entropy and weighted permutation entropy
-involves choosing a good value for the word length $n$. The primary
-consideration is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the ordinals are
-gathered: e.g.:
-\begin{align*}
-  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
-\end{align*}
-assuming an average of 100 counts per ordinal is sufficient. In the literature,
-$3 \le n \le 6$ is a standard choice --- generally without any formal
-justification. In theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In practice,
-what one should do is calculate the \emph{persistent} permutation entropy by
-increasing $n$ until the result converges, but data length issues can intrude
-before that convergence is reached.
-
-The weighted permutation entropy for the {\tt SVD} program is given in
-Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
-the time series. Within each of those windows, the statistics over words of
-length 4 are computed and the WPE is calculated. The gray bands denote regions
-where the 5,000 value window overlapped visually-distinct regimes. It can be
-seen that the behaviors of the weighted permutation entropy vary between
-regimes.
-[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
-Maybe point at some of the predicting predictability papers.]]
-
-%\begin{figure}[htbp]
-%  \centering
-%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
-%    are regions where the window overlaps regimes. The window size used is
-%    $5,000 \times 100,000$ instructions and the word length is $4$.}
-%  \label{fig:wwpe}
-%\end{figure}
-
-%%FIGURE INTENT
-%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
-%%%%%%%%%%%%%%%%%
-\begin{figure}[htbp]
-  \centering
-
-  %\begin{subfigure}{0.3\textwidth}
-  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
-  %  \caption{The instructions per cycle of \svd.
-Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
-From left to right each change in color represents a change in regime for 6 regimes in total.
-}
- %   \label{fig:svd_ts}
- % \end{subfigure}%
-  %\\
- % \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
-    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-  \caption{
-[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly.
-Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD.
-The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
-Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE.
-From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
-\end{figure}
-
- \section{Validation of Primary Findings} % (fold)
- \label{sec:results}
-
-%OUTLINE:
-%\begin{enumerate}
-%\item \cmark~Introduce MASE.
-%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
-%its WPE.
-%\item \cmark~Talk about structure analysis. Just because %there is forward information
-%transfer, does not mean that linear predictors can get at %this.
-%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
-%side plot.
-%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
-%\begin{enumerate}
-%\item \cmark~clusters are distributed differently
-%\item \cmark~clusters are shaped differently---tight or not
-%\item \cmark~clusters move differently between LMA and ARIMA
-%\item finally, the diagonal line is important. If you're %below it, you could do
-%better.
-%\end{enumerate}
-%\end{enumerate}
-
-%BRAINSTORMING 
-%\begin{itemize}
-%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
-%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
-
-%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured.
-And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
-
-%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
-
-%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
-That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
-This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
-
-%\end{itemize}
-
-In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:
-
-\begin{enumerate}
-
-\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.
-
-%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
-
-%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
-
-\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-
-%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
-The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
-
-%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
-
-\end{enumerate}
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\item First paragraph: WPE is a good measure of predictability.  Figure:
-%best athlete MASE vs. its WPE.
-
-%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-
-As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task.
-As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
-Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.
-
-In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation.
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
-  \caption{The best MASE among all runs and prediction methods vs weighted
-  permutation entropy. For each of these, the word length used is $6$. The
-  dashed line is a least-squares linear fit of all the points except for {\tt
-  SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_wpe}
-\end{figure}
-What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}.
-We will not further validate this claim through some examples.
-
-To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively.
-In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. }
-    \label{fig:lma_pred_vs_ent}
-  \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:arima_pred_vs_ent}
-  \end{subfigure}
-\caption{
-For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}
-\end{figure}
-
-We will start with the lowest complexity program: \col.
-Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}.
-If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
-This would suggest that there is some predictive structure present but not that much.
-However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity.
-This is in fact the case.
-\col is a complex signal, i.e., chaotic, but it is complex in a structured way.
-Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
-The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
-
-Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive.
-This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
-However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
-The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
-This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict.
-However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA.
-If we do this we get forecasts which score 1.5 times better than the random walk.
-Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.
-
-Finally, we consider the most complex program: \gcc.
-The complexity of \gcc is $0.94 \pm 0.001$.
-This implies that almost no predictive structure exists in this structure.
-Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future.
-This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex.
-In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step.
-This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.
-
-There is a single outlier in this study which is \svdone.
-This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively).
-The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts.
-We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method.
-MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
-In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
-A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.
-
-To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail.
-This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
-We provide a small snippet as opposed to the full time series simply for visual clarity.
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=\textwidth]{figs/svdonets2}
-\caption{ For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}
-\end{figure} 
-As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide.
-Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
-The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}.
-These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle).
-As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
-At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal.
-This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
-The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.
-
-[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.
-
-%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
-%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
-%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}.
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
-
-%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration.
-The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method).
-On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values.
-As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE.
-The WPE is sensitive to both linear and nonlinear structure.
-When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot.
-So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.
-
-%is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
-How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict.
-If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy.
-However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-
-Other noteworthy features of the LMA and ARIMA results are the cluster locations
-and distributions. The WPE values of each run of any particular program tend to
-have little variance, leading to the clusters in
-Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
-constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
-low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
-traces, however, have a large variance resulting in the spread seen in
-Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
-bad, in that other predictors vastly outperform it, but they are inconsistent.
-Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
-see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
-the right than those in Figure~\ref{fig:lma_pred_vs_ent}.
-
-\section{ Conclusions \& Future Work}\label{sec:conc}
-
-\begin{enumerate}
-\item  This need to be rewritten to conclude and address this paper...
-\end{enumerate}
-
-The results presented here suggest that permutation entropy---a ordinal
-calculation of forward information transfer in a time series---is an effective
-metric for predictability of computer performance traces. Experimentally, traces
-with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
-may overshadow the inherent determinism in the system dynamics, whereas traces
-with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
-of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
-0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
-corroborating the results of~\cite{mytkowicz09}.
-
-If information is the limit, then gathering and using more information is an
-obvious next step.  There is an equally obvious tension here between data length
-and prediction speed: a forecast that requires half a second to compute is not
-useful for the purposes of real-time control of a computer system with a MHz
-clock rate.  Another alternative is to sample several system variables
-simultaneously and build multivariate delay-coordinate embeddings.  Existing
-approaches to that are computationally prohibitive
-\cite{cao-multivariate-embedding}.  We are working on alternative
-methods that sidestep that complexity.
-
-%%This use of permutaiotn entropy is a useful application of information
-%%theory to computer performance modeling because it gives you a simple
-%%and fast way to decide whether building a model and trying to forecast
-%%the future is worthwhile or if guessing the mean is just as effective.
-%%End with a sentence tying back to power management and world peace.
-
-%\begin{it}
-%Summarize the results.
-
-%Paragraphs on the issues that come up, including one about the "amount
-%of info" one: if one could sample more variables, for instance, one
-%might be able to do a better job of predicting more-complex traces.
-%Segue to some handwaving about multivariable LMA models; tie this back
-%to the "computers are NLD systems" stuff in the intro.  This is a real
-%challenge; current approaches to this modelling problem have the major
-%issue of taking way too long to build.  And that's a big issue if
-%you're trying not just to classify, but to predict.  In a system that
-%runs at MHz speeds, a prediction that takes milliseconds to compute is
-%not useful.
-%\end{it}
+\input{concl.tex}
 

### 2014-02-19 results.tex
Split Sections to help with conflicts

 \section{Validation of Primary Findings} % (fold)
 \label{sec:results}

%OUTLINE:
%\begin{enumerate}
%\item \cmark~Introduce MASE.
%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
%its WPE.
%\item \cmark~Talk about structure analysis. Just because %there is forward information
%transfer, does not mean that linear predictors can get at %this.
%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
%side plot.
%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
%\begin{enumerate}
%\item \cmark~clusters are distributed differently
%\item \cmark~clusters are shaped differently---tight or not
%\item \cmark~clusters move differently between LMA and ARIMA
%\item finally, the diagonal line is important. If you're %below it, you could do
%better.
%\end{enumerate}
%\end{enumerate}

%BRAINSTORMING 
%\begin{itemize}
%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]

%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]

%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working. [[use in this section as bridge to next section]]

%\end{itemize}

In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:

\begin{enumerate}

\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.

%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"

%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability

\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.

%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure). The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.

%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.

\end{enumerate}

%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
%%%%%%%%%%%%%%%%%%%%%%%%%

%\item First paragraph: WPE is a good measure of predictability.  Figure:
%best athlete MASE vs. its WPE.

%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]

As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task. As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do. Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.

In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_wpe}
\end{figure}
What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}. We will not further validate this claim through some examples.

To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively. In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs. We also list all of these values in Table \ref{tab:error}

\begin{table}[htdp]
\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 6 for each signal.}
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
\hline
\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\

\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$&  $0.5131 \pm 0.0034$ \\

\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$&  $0.9572 \pm 0.0156$ \\
\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &   $0.8464 \pm0.0044$ \\
\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ &  $0.7157 \pm 0.0056$ \\
\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$ & $0.8246 \pm 0.0077$ \\
\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$&  $0.6776 \pm 0.0068$ \\
\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$&  $0.7475 \pm 0.0106$ \\
\hline
\end{tabular}
\end{center}
\label{tab:error}
\end{table}%

\begin{figure}[htbp]
  \centering

  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. }
    \label{fig:lma_pred_vs_ent}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. }
    \label{fig:arima_pred_vs_ent}
  \end{subfigure}
\caption{
For each of these, the word length used is $6$. The MASE values for LMA against ARIMA. The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better. All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}  
\end{figure}

We will start with the lowest complexity program: \col. Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}. If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction. This would suggest that there is some predictive structure present but not that much. However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity. This is in fact the case. \col is a complex signal, i.e., chaotic, but it is complex in a structured way. Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast. The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.

Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive. This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure. However, this increase in complexity is enough to really throw off the constant and linear prediction strategies. The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast. This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict. However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA. If we do this we get forecasts which score 1.5 times better than the random walk. Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.

Finally, we consider the most complex program: \gcc. The complexity of \gcc is $0.94 \pm 0.001$. This implies that almost no predictive structure exists in this structure. Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future. This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex. In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step. This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.

There is a single outlier in this study which is \svdone. This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively). The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts. We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method. MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click. In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast. A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.

To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail. This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal. We provide a small snippet as opposed to the full time series simply for visual clarity.
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{figs/svdonets2}
\caption{ For each of these, the word length used is $6$. The MASE values for LMA against ARIMA. The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better. All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}  
\end{figure} 
As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide. Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal. The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}. These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle). As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting. At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal. This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low. The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.

[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE. This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk. The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.

%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
%%%%%%%%%%%%%%%%%%%%%%%%%

Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it. In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information. For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}. If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}. In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine. We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.

%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration. The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method). On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values. As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE. The WPE is sensitive to both linear and nonlinear structure. When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot. So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.

%is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.

Other noteworthy features of the LMA and ARIMA results are the cluster locations
and distributions. The WPE values of each run of any particular program tend to
have little variance, leading to the clusters in
Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
traces, however, have a large variance resulting in the spread seen in
Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
bad, in that other predictors vastly outperform it, but they are inconsistent.
Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
the right than those in Figure~\ref{fig:lma_pred_vs_ent}.

### 2014-02-19 wpe.tex
Split Sections to help with conflicts

\section{Measuring Structural Complexity}\label{sec:meaComplex}

THINGS TO DO
\begin{enumerate}
\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.

\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
\item justify and explain the regime split in \svd 
\end{enumerate}

For the purposes of this paper, one can view entropy as a measure of complexity
and predictability in a time series.  A high-entropy time series is almost
completely unpredictable---and conversely.  This can be made more rigorous:
Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
difficult prediction becomes.

Utilizing entropy as a measure of temporal complexity is by no means a new idea
\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
effectively real-valued.  To get around this, one must discretize the data---
typically achieved by binning.  Unfortunately, this is rarely a good solution to
the problem, as the binning of the values introduces an additional dynamic on
top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
dynamics studies how to discretize a time series in such a way that the
intrinsic behavior is not perverted, but these methods are fragile in the face
of noise and require further understanding of the underlying system, which
defeats the purpose of measuring the entropy in the first place.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
employs a method of discretizing real-valued time series that follows the
intrinsic behavior of the system under examination.  Rather than looking at the
statistics of sequences of values, as is done when computing the Shannon
entropy, permutation entropy looks at the statistics of the \emph{orderings} of
sequences of values using ordinal analysis. Ordinal analysis of a time series is
the process of mapping successive time-ordered elements of a time series to
their value-ordered permutation of the same size.  By way of example, if $(x_1,
x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
other things, it is generally robust to observational noise and requires no
knowledge of the underlying mechanisms.

\begin{mydef}[Permutation Entropy]

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
  = 1,\dots,T}$:
  \begin{align*}
    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
  \end{align*}
  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
  \ge 2$ is defined as
  \begin{align*}
  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
  \end{align*}

\end{mydef}

Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
it is common in the literature to normalize permutation entropy as follows:
$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
and ``high'' entropy is close to 1. Finally, it should be noted that the
permutation entropy has been shown to be identical to the Shannon entropy for
many large classes of systems \cite{amigo2012permutation}.

Here we will be utilizing a variation of the permutation entropy, the
\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
permutation entropy attempts to correct for observational noise which is larger
than some trends in the data, but smaller than the larger scale features --- for
example, a signal that switches between two fixed points with noise about those
fixed points. The weighted permutation entropy would be dominated by the
switching rather than by the stochastic fluctuation. To accomplish this, the
\emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{t+1:t+n}) = \frac{1}{n}
                 \sum_{x_i \in \{x_{t+1:t+n}\}}
                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
\end{align*}
where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.

The weighted probability of a permutation is then:
\begin{align*}
  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
probability enhances permutations involved in ``large'' features and demotes
permutations which are small in amplitude relative to the features of the time
series. The weighted permutation entropy is then:
\begin{align*}
  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
results of this paper.

In practice, calculating permutation entropy and weighted permutation entropy
involves choosing a good value for the word length $n$. The primary
consideration is that the value be large enough that forbidden ordinals are
discovered, yet small enough that reasonable statistics over the ordinals are
gathered: e.g.:
\begin{align*}
  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
\end{align*}
assuming an average of 100 counts per ordinal is sufficient. In the literature,
$3 \le n \le 6$ is a standard choice --- generally without any formal
justification. In theory, the permutation entropy should reach an asymptote with
increasing $n$, but that requires an arbitrarily long time series. In practice,
what one should do is calculate the \emph{persistent} permutation entropy by
increasing $n$ until the result converges, but data length issues can intrude
before that convergence is reached.

The weighted permutation entropy for the {\tt SVD} program is given in
Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes. [[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist. Maybe point at some of the predicting predictability papers.]]

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
%    are regions where the window overlaps regimes. The window size used is
%    $5,000 \times 100,000$ instructions and the word length is $4$.}
%  \label{fig:wwpe}
%\end{figure}

%%FIGURE INTENT
%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
  \centering

  %\begin{subfigure}{0.3\textwidth}
  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
  %  \caption{The instructions per cycle of \svd. Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}. From left to right each change in color represents a change in regime for 6 regimes in total. }
 %   \label{fig:svd_ts}
 % \end{subfigure}%
  %\\
 % \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
  \caption{
[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly. Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD. The gray bands
    are regions where the window overlaps regimes. The window size used is
    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot. Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE. From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
\end{figure}

### 2014-02-19 intro.tex
titles and notes

@@
 \section{Introduction}\label{sec:intro}
+
+Notes on wording
+
+ If we define entropy as complexity in the paper (as we do currently) we could say that every signal complicated or not has a balance between temporal/predictive structure and complexity (entropy).
+So for example, a chaotic signal and a random walk signal are both "complicated" they exhibit interesting behavior (they aren't monotonic, periodic, constant etc.) but one of them is far more complex and less predictable than the other , e.g.,  a random walk or gcc for example is complicated *and* highly complex (very high entropy) where as col is also complicated but not as complex as it exhibits a great deal of predictive structure (low entropy).
+
+An interesting contribution here is that examining a complicated signal (i.e., something that has interesting behavior, not periodic, monotonic, constant etc) of noisy real-valued entries, it is hard to determine a priori (without WPE) where in the spectrum of temporal structure and complexity the signal lies.
+Is it completely complex and have little to no predictive structure e.g., gcc or is it balanced in complexity and structure like a chaotic signal e.g., col.
+
+One potential issue here is that apparently information theorists say that completely random signals are void of complexity.
+So Saying that a random walk signal (i.e., a signal with high entropy) is complex may cause some issues.
+HOWEVER, if we define complexity as entropy I don't think this is an issue.
+
+Ryan does this seem reasonable?
+Is this an accurate synopsis of our discussion.
+Please feel free to add to it.
+Liz do you think making this distinction between "Complicated" and "Complex" would be useful for the readers understanding and dissemination of this topic?
+
+This would suggest changing the title to something a long the lines of "Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability "
 

### 2014-02-26 intro.tex
adding related works and edits to intro

@@
 \end{enumerate}
+\noindent ***NOTES END HERE****
+
+\noindent ****SECTION START****
+
+Complicated time series are ubiquitous in modern scientific research. 
+%Characterizing and forecasting the future behavior of these observed systems is a vastly researched area with many open questions. 
+Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}\footnote{An approx of (blah ) entropy which we make more rigorous in blah}.
+On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model.
+In particular, there is a an underlying process that generates and transmits information from the past to the future in a perfectly predictable fashion.
+Some examples of this are constant or low-period signals.
+On the opposite end of this spectrum are signals which are ``fully complex", i.e., the underlying generating process creates information so rapidly that no information at all is transmitted from the past to the future.
+Some examples, of this are white noise or random walk processes.
+With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.
+In the middle of this spectrum are where complicated \emph{and} forecast-able signals exist, e.g., deterministic chaotic trajectories, high period signals with some level of noise on top.
+With time-series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model we could predict the future of the observed system with high accuracy.
+Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series, let alone an arbitrary time series.
+[[Rephrase, below here but this is the idea...]]
+ So given a complicated noisy real-valued time series from an observed system, is it possible to reliably quantify where on this spectrum of the time series exists.
+In particular, can we characterize when a time series is predictable without making further assumptions about the system and only knowing possibly noisy real-valued measurements?
+
+One such question is: when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) how can we infer from the time series itself if the signal is too complex to forecast?
+That is, when is a complicated time series too complex to forecast?
+It is well known that every complicated real
+
+[[Introduce limitations and direction]]
+
+For the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which we appear complicated, i.e., they exhibit interesting behavior, e.g., not periodic, strictly monotonic, or constant.
+We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) and we attempt to measure empirically the complexity level of the time series.
+For the purposes of this paper we define complexity as a particular approximation of Shannon \emph{entropy} [[Ryan is this right or is it Renyi?]], which we define rigorously in Section \ref{sec:meaComplex}.
+We then attempt to correlate the empirically estimated complexity with the forecast success of several different forecasting schema.
+
+[[Introduce and make it clear why we choose comp. perf. as test bed.]]
 

### 2014-02-26 related.tex
adding related works and edits to intro

\section{Related Work}

Attempting to model and characterize predictability is a very old problem which arguably began with Yule in 1927 when he invented AR and many attempts to quantify predicatbility have followed including ....

Entropy though Generating partitions 
works **if** you have generating parition but not if you don't

Redundancy (details in SFI forecasting), relies heavily on either estimating the generating partiion which is hard to do *and* estimating the positive lyap spectrum which is hard to do for noisy systems and impossible for systems that are not deterministic.

DVS plots, gives pros and cons in (SFI prediction book)

predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd

Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.

But this method is differnt becasue it uses no assumption about the underlying model, does not require generating partitions, is applicable to noisy real-valued time series.

Maybe talk about ``
wpe has been applied to predicting irregularities in brain wave data but no on one has examined it's correlation with predictive structure. "
but kind of puts the cart before the horse

### 2014-02-26 intro.tex
Adding to and editing intro

--- 2014-02-19 intro.tex
@@
 \end{enumerate}
+\noindent ***NOTES END HERE****
 
-Computers are among the most complex engineered artifacts in current
+\noindent ****SECTION START****
+
+Complicated time series are ubiquitous in modern scientific research. 
+%Characterizing and forecasting the future behavior of these observed systems is a vastly researched area with many open questions. 
+Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}\footnote{An approx of (blah ) entropy which we make more rigorous in blah}.
+On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model.
+In particular, there is a an underlying process that generates and transmits information from the past to the future in a perfectly predictable fashion.
+Some examples of this are constant or low-period signals.
+On the opposite end of this spectrum are signals which are ``fully complex", i.e., the underlying generating process creates information so rapidly that no information at all is transmitted from the past to the future.
+Some examples, of this are white noise or random walk processes.
+With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.
+In the middle of this spectrum are where complicated \emph{and} forecast-able signals exist, e.g., deterministic chaotic trajectories, high period signals with some level of noise on top.
+With time series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model the future of the observed system could be forecast with high accuracy.
+
+Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{sfi-prediction}, let alone an arbitrary time series.
+This naturally leads to an important and hard question: given a complicated real-valued time series, is it possible to reliably quantify where on this spectrum the time series lies?
+i.e., when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) how can we infer from the time series itself if the signal is too \emph{complex} to forecast?
+
+To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant.
+We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
+For the remainder of this paper we define complexity as a particular approximation of Shannon \emph{entropy} [[Ryan is this right or is it Renyi?]], which we define and explain rigorously in Section \ref{sec:meaComplex}.
+To validate these claims we model the time series using a variety of forecasting models which we introduce in Section \ref{sec:model}.
+We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method.
+Which results in two primary findings:
+
+\begin{enumerate}
+
+\item The complexity of a noisy real-valued time series is quantifiable.
+
+\item The way information is processed internally by a given process plays a crucial role in the success of different forecasting schema.
+
+\end{enumerate}
+
+%[[Introduce and make it clear why we choose comp. perf. as test bed.]]
+
+The ideal test bed for this purpose are computer performance traces. Computers are among the most complex engineered artifacts in current
 use.  Modern microprocessor chips contain multiple processing units
@@
 that go into the design of these chips---make the patterns of their
-processor loads and memory accesses highly complex and hard to
-predict.  Accurate forecasts of these quantities, if one could
-construct them, could be used to improve computer design.  If one
-could predict that a particular computational thread would be bogged
-down for the next 0.6 seconds waiting for data from main memory, for
-instance, one could save power by putting that thread on hold for that
-time period (e.g., by migrating it to a processing unit whose clock
-speed is scaled back).  Computer performance traces are, however, very
-complex.  Even a simple ``microkernel,'' like a three-line loop that
-repeatedly initializes a matrix in column-major order, can produce
-{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
-Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
-predictability.
+processor loads highly complicated. Accurate forecasts of these quantities, if one could
+construct them, could be used to improve computer design. 
+% If one
+%could predict that a particular computational thread %would be bogged
+%down for the next 0.6 seconds waiting for data from main memory, for
+%instance, one could save power by putting that thread %on hold for that
+%time period (e.g., by migrating it to a processing %unit whose clock
+%speed is scaled back). 
+ Computer performance traces are, however, very
+complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel,'' like a three-line loop that
+repeatedly initializes a matrix in column-major order, can produce everything from periodic to
+{\sl chaotic} performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
+Figure~\ref{fig:ipc}.%, and chaos places fundamental limits on predictability.
 %
@@
   \end{figure}
+With a single system producing complicated time series that vary so greatly in complexity, it would be invaluable to known a priori if a trace contained enough predictive structure to forecast before time and effort was spent on trying to model and forecast it.
 
@@
 modeling and forecast strategy works very well.  In more-complicated
-programs, however, such as speech recognition software or compilers,
+programs, however, such as numerical software or compilers,
 this forecast strategy---as well as the traditional methods---break
-down quickly.
+down some of the time, but work fine others.
 
@@
 deterministic systems.  We focus here on the specific example of
-computer performance.  We conjecture that the complexity of traces
+computer performance but believe the results apply to a much broader class of time series.  We conjecture that the complexity of traces
 from these systems---which results from the inherent dimension,
@@
 real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study four
-examples---two simple microkernels and two complex programs from the
-SPEC benchmark suite---running on different Intel-based machines.  For
+effective way to explore that conjecture.  We study three
+examples---a simple microkernel and two complex programs: one from the
+SPEC 2006CPU benchmark suite, and one from LAPACK---running on an Intel i7-based machine.  For
 each program, we calculate the permutation entropy of the processor
-load (instructions per cycle) and memory-use efficiency (cache-miss
-rates), then compare that to the prediction accuracy attainable for
-that trace using a simple deterministic model.
+load (instructions per cycle), then compare that to the prediction accuracy attainable for
+that trace using a series of deterministic models.
 
@@
 
-% The rest of the paper is organized as follows.
-% Section~\ref{sec:compModel} describes the experimental setup, as well
-% as the nonlinear modeling and forecast strategies.  In
-% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
-% its value for a number of different computer performance traces, and
-% compare the results to the prediction accuracy.  In
-% Section~\ref{sec:conc}, we discuss these results and their
-% implications in regard to our conjecture, and consider future areas of
-% research.
+ The rest of the paper is organized as follows.
+ Section~\ref{sec:compModel} describes the experimental setup, as well
+ as the nonlinear modeling and forecast strategies.  In
+ Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
+ its value for a number of different computer performance traces, and
+ compare the results to the prediction accuracy.  In
+ Section~\ref{sec:conc}, we discuss these results and their
+ implications in regard to our conjecture, and consider future areas of
+ research.

### 2014-02-26 pecomplexity.tex
titles and notes
adding related works and edits to intro
major edits on the results section
Also some diddling in other sections to get cross refs to work, and to
put the table in an appendix

@@
 \usepackage{subcaption}
+\usepackage{color}
 \usepackage{amssymb}
@@
 Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity
+\\ {\color{green}Quantifying Predictability through Structural Complexity}
 \\ Quantifying Predictability using Structural Complexity 
+\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
+\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
 }
@@
 
+\section{Appendix}
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
+\hline
+\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
+
+\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$&  $0.5131 \pm 0.0034$ \\
+
+\svdone     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$&  $0.9572 \pm 0.0156$ \\
+\svdtwo     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &   $0.8464 \pm0.0044$ \\
+\svdthree       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ &  $0.7157 \pm 0.0056$ \\
+\svdfour     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$ & $0.8246 \pm 0.0077$ \\
+\svdfive     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$&  $0.6776 \pm 0.0068$ \\
+\svdsix     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$&  $0.7475 \pm 0.0106$ \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Detailed MASE scores}
+\label{tab:error}
+\end{table}%
+
 \section*{Acknowledgment}

### 2014-02-26 intro.tex

@@
 
-Notes on wording
+%Notes on wording
 
- If we define entropy as complexity in the paper (as we do currently) we could say that every signal complicated or not has a balance between temporal/predictive structure and complexity (entropy).
+% If we define entropy as complexity in the paper (as we do currently) we could say that every signal complicated or not has a balance between temporal/predictive structure and complexity (entropy).
 So for example, a chaotic signal and a random walk signal are both "complicated" they exhibit interesting behavior (they aren't monotonic, periodic, constant etc.) but one of them is far more complex and less predictable than the other , e.g.,  a random walk or gcc for example is complicated *and* highly complex (very high entropy) where as col is also complicated but not as complex as it exhibits a great deal of predictive structure (low entropy).
 
-An interesting contribution here is that examining a complicated signal (i.e., something that has interesting behavior, not periodic, monotonic, constant etc) of noisy real-valued entries, it is hard to determine a priori (without WPE) where in the spectrum of temporal structure and complexity the signal lies.
+%An interesting contribution here is that examining a complicated signal (i.e., something that has interesting behavior, not periodic, monotonic, constant etc) of noisy real-valued entries, it is hard to determine a priori (without WPE) where in the spectrum of temporal structure and complexity the signal lies.
 Is it completely complex and have little to no predictive structure e.g., gcc or is it balanced in complexity and structure like a chaotic signal e.g., col.
 
-One potential issue here is that apparently information theorists say that completely random signals are void of complexity.
+%One potential issue here is that apparently information %theorists say that completely random signals are void of complexity.
 So Saying that a random walk signal (i.e., a signal with high entropy) is complex may cause some issues.
@@
 
-Ryan does this seem reasonable?
+%Ryan does this seem reasonable?
 Is this an accurate synopsis of our discussion.
@@
 
-This would suggest changing the title to something a long the lines of "Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability "
+%This would suggest changing the title to something a long the lines of "Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability "
 
@@
 This naturally leads to an important and hard question: given a complicated real-valued time series, is it possible to reliably quantify where on this spectrum the time series lies?
-i.e., when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) how can we infer from the time series itself if the signal is too \emph{complex} to forecast?
+That is, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) how can we infer from a noisy real-valued time series if the signal is too \emph{complex} to forecast?
 
@@
 We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
-For the remainder of this paper we define complexity as a particular approximation of Shannon \emph{entropy} [[Ryan is this right or is it Renyi?]], which we define and explain rigorously in Section \ref{sec:meaComplex}.
+We argue that \emph{permutation entropy}
+\cite{bandt2002per}, a method for measuring the entropy of a
+real-valued-finite-length noisy time series through ordinal analysis, is an effective way to explore that conjecture.
+For the purposes of this paper we define complexity as \emph{permutation entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}.
 To validate these claims we model the time series using a variety of forecasting models which we introduce in Section \ref{sec:model}.
@@
 %[[Introduce and make it clear why we choose comp. perf. as test bed.]]
-
-The ideal test bed for this purpose are computer performance traces. Computers are among the most complex engineered artifacts in current
+[[Move this up]]
+To explore this conjecture we require a test bed for generating a broad array of time series to analyze which covers the full spectrum of complexity.
+We argue that the ideal test bed for this purpose is computer performance.
+Computers are among the most complex engineered artifacts in current
 use.  Modern microprocessor chips contain multiple processing units

### 2014-02-26 results.tex

@@
- \section{Validation of Primary Findings} % (fold)
+ \section{Predictability, Complexity, and Permutation Entropy} % (fold)
  \label{sec:results}
+% I changed the title of this section because there are lots of
+% results in the modeling section too.  The results HERE are about the
+% relationship between predictability and WPE
 
@@
 
-In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:
+In this section, we offer an empirical validation of the two key
+conjectures introduced in Section \ref{sec:intro}, namely:
 
@@
 
-\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.
+\item that the weighted permutation entropy (WPE) of a noisy
+  real-valued time series is correlated with prediction
+  accuracy---i.e., that the predictable structure in a time-series
+  data set can be quantified by its WPE.
 
-%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"
+%\subitem [[Simply reminders not to be included]]WPE can quantify when
+%a noisy real-valued time series is predictable.[[I am unsatisfied
+%with predictable in this sentence, need a better word to say "better
+%than random walk" or ``able to be forecast effectively" or ``has the
+%structural capacity to transmit information in a way that the time
+%series can be effectively forecast"
 
-%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability
+%\item The existence of complex structure in a noisy real-valued time
+%series is quantifiable and this type of complexity is directly
+%correlated with predictability
 
-\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
+\item that the way information is processed internally by a system is
+  correlated with the effectiveness of a given predictor on
+  time-series data from that system
 
-%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure).
-The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.
+%\subitem [[Simply reminders not to be included]]We will have shown
+%that the existence whether linear or nonlinear is picked up on with
+%WPE but this point gets at whether the prediction model can use the
+%structure or not (linear can't use nonlinear structure). The right to
+%left shifts in \col and some of the \svd regimes and the lack of
+%shift in \gcc illustrate this nicely.
 
-%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
+%\subitem [[Simply reminders not to be included]]This is the linear vs
+%nonliear vs random we see with the right to left shifts with lower
+%complexity time series.
 
@@
 
-%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
+%\item Quantifying structured and unstructured complexity is
+%nontrivial in the case of real-valued noisy time series but WPE does
+%this. [[talk about again here but justify in information theory
+%section]]
 
-As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task.
-As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do.
-Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.
-
-In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation.
+The experiments below involve the three different prediction methods
+described in Section~\ref{sec:compModel}---na\"ive, ARIMA, and
+LMA---applied to time-series data from eight different systems: {\tt
+  col\_major}, {\tt 403.gcc}, and the six different regimes of the
+{\tt dgesdd} signal in Figure~\ref{fig:wwpe}.  The objective of these
+experiments was to explore how prediction accuracy is related to WPE,
+and how that relationship depends on the generating process and the
+prediction method.  Working from the first 90\% of each signal, we
+generated a prediction of the last 10\% of that signal using each of
+the methods, as described in Section~\ref{sec:accuracy}, then
+calculated the MASE value of those predictions.  We also calculated
+the WPE of each time series, as described in
+Section~\ref{sec:meaComplex}, using a wordlength of six.  In order to
+assess the run-to-run variability of these results, we repeated all of
+these calculations on 15 separate trials: i.e., 15 different runs of
+each program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best}
+of these 45 predictions for each system: i.e., the lowest error over
+all 15 trials and all three methods for each program.  The WPE is
+plotted against the corresponding MASE value in order to bring out the
+correlation between these two quantities.
 \begin{figure}[htbp]
   \centering
-  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
-  \caption{The best MASE among all runs and prediction methods vs weighted
-  permutation entropy. For each of these, the word length used is $6$. The
-  dashed line is a least-squares linear fit of all the points except for {\tt
-  SVD$_1$} which we have excluded for reasons explained in the text.}
-  \label{fig:pred_vs_wpe}
-\end{figure}
-What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}.
-We will not further validate this claim through some examples.
-
-To further validate this finding we present an in-depth analysis of three examples (\col, \svdfive and \gcc ) which nicely cover the range of complexity from low to high respectively.
-In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.
-We also list all of these values in Table \ref{tab:error}
-
-\begin{table}[htdp]
-\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 6 for each signal.}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|}
-\hline
-                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
-\hline
-\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\
-
-\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$&  $0.5131 \pm 0.0034$ \\
-
-\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$&  $0.9572 \pm 0.0156$ \\
-\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &   $0.8464 \pm0.0044$ \\
-\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ &  $0.7157 \pm 0.0056$ \\
-\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$ & $0.8246 \pm 0.0077$ \\
-\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$&  $0.6776 \pm 0.0068$ \\
-\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$&  $0.7475 \pm 0.0106$ \\
-\hline
-\end{tabular}
-\end{center}
-\label{tab:error}
-\end{table}%
-
-\begin{figure}[htbp]
-  \centering
-
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
-    \caption{The MASE of LMA vs weighted permutation entropy. }
-    \label{fig:lma_pred_vs_ent}
-  \end{subfigure}%
-  \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \caption{The MASE of ARIMA vs weighted permutation entropy. }
-    \label{fig:arima_pred_vs_ent}
-  \end{subfigure}
-\caption{
-For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}
+  \includegraphics[width=0.6\textwidth]{figs/prediction_vs_entropy}
+  \caption{Weighted permutation entropy vs. mean absolute scaled error
+    (MASE) of the best prediction of each system.
+% 
+% (across all trials and all prediction methods) 
+% 
+The dashed line is a least-squares linear fit of all the points except
+{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
+text.  [[Ryan, is that correct?]]}
+  \label{fig:wpe_vs_mase_best}
 \end{figure}
 
-We will start with the lowest complexity program: \col.
-Conceptually this program is very simple but due to computer design choices this program can result in highly complex even chaotic time series traces, see \cite{mytkowicz09}.
-If we forecast this signal using an out-of-the-box linear forecaster (ARIMA) or an out-of-the-box constant predictor we receive MASE ($0.59 \pm 0.21$ and $0.57 \pm$ 0.001 respectively) which suggest that on average we can only predict this signal twice as well as the simple random walk prediction.
-This would suggest that there is some predictive structure present but not that much.
-However, if we quantify the predictive structure present in this signal using WPE we see that this signal is highly structured with a WPE of ($0.51 \pm 0.003$.) The existence of structured which we quantified with WPE implies that an ideal predictor capable of using this signals information should be able to predict this signal much better than the random walk due to the high amount of predictive structure or equivalently the low complexity.
-This is in fact the case.
-\col is a complex signal, i.e., chaotic, but it is complex in a structured way.
-Using a prediction technique that can process nonlinear information yields forecasts with a MASE of $0.05 \pm 0.001$, this is a prediction 20 times more accurate than a random walk forecast.
-The existence of predictive structure as quantified by WPE could be precisely the information needed to continue searching for an ideal predictor like LMA.
+The best-case prediction error for each of these eight systems is
+roughly proportional to the weighted permutation entropy, which is
+consistent with our first conjecture.  The dashed line in the figure,
+a linear least-squares fit to these best-case errors---with the
+exception of {\tt dgesdd$_5$}, for reasons described below---captures
+this rough proportionality.  This is not a formal result.  The three
+methods used here were chosen to span the space of standard prediction
+strategies, but they do not, of course, cover that space exhaustively.
+Our goal here is an empirical assessment of predictability and
+complexity, not formal results about a ``best'' predictor for a given
+time series.  There may be other methods that produce lower MASE
+values than those in Figure~\ref{fig:wpe_vs_mase_best}, but the
+sparseness of the points in the upper-left and lower-right quadrants
+of this plot strongly suggests that the underlying predictability of a
+time series is inversely correlated with its WPE.  The rest of this
+section describes our results in more detail, including the measures
+taken to assure meaningful comparisons across methods, trials, and
+programs, and elaborates on the meaning of the dashed line in the
+figure.
 
-Maybe even more convincing than a low complexity signal like \col is a moderately complex signal like \svdfive.
-This signal omits a WPE of $0.677 \pm 0.006$, this is a significantly higher complexity than \col but still implies the existence of a great deal of predictive structure.
-However, this increase in complexity is enough to really throw off the constant and linear prediction strategies.
-The MASE scores for ARIMA and na\"ive imply 2.5 and 20 times worse prediction than using a simple random walk forecast.
-This may be discouraging to someone using an out-of-the-box predictor that the signal is too complex to predict.
-However by \emph{knowing} that there exists predictable  in this signal we can try to use a more advanced technique, in particular LMA.
-If we do this we get forecasts which score 1.5 times better than the random walk.
-Showing that this usable structure (although hidden to the linear and constant method) was quantifiable by WPE.
+Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
+experiments.
+\begin{figure}
+  \centering
+    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
+    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
+    \includegraphics[width=0.6\textwidth]{figs/LMA_prediction_vs_entropy}
+\caption{WPE vs. MASE for all trials, methods, and systems.  \svdone,
+  \svdthree, and \svdfive are omitted from the top plot for scale
+  reasons; their MASE scores are $2.6763\pm4.3282$, $31.3857\pm
+  0.2820$, and $20.8703 \pm 0.1915$ respectively.  {\color{red}Replace
+    the top plot with a plot of the naive results---but running from 0
+    to 3.1 on the horizontal axis and excluding SVD 1, 3, and 5}.  The
+  dashed line is the same as in the previous figure [[Ryan, is this
+      correct?]].  Numerical values, including means and standard
+  deviations of the errors, can be found in Table~\ref{tab:error} in
+  the appendix.
+% 
+%  [[All traces except those from \svd$_1$ lie above the line,
+%      indicating that LMA is better suited prediction method for the
+%      computer traces considered.]]  [[This comes {\bf long} before the
+%      associated discussion in the text, and it's not what this figure
+%      is trying to show here.]]
+% 
+}
+    \label{fig:wpe_vs_mase_all}
+\end{figure} 
+There are 15 points in each cluster, one for each trial\footnote{A
+  trial involves collecting a time series from the corresponding
+  system, predicting the last 10\% of that time series, and
+  calculating the MASE score.}.  (The points in
+Figure~\ref{fig:wpe_vs_mase_best} are the leftmost points in any of
+the three clusters for the corresponding system.)  Most of the
+clusters are quite tight.  Some are flat and wide because the WPE
+values of each run of any particular program tend to have little
+variance; for most traces, the LMA and ARIMA variance is low as well,
+resulting in small, tight clusters.  The {\sl spatial distribution} of
+the clusters is quite different in the three plots: LMA is tight
+around the dashed line, whereas naive and ARIMA are [[need to fill in
+    words here when get the third image]].  This is a geometric
+validation of the second conjecture in this paper, as described in the
+following paragraphs.
 
-Finally, we consider the most complex program: \gcc.
-The complexity of \gcc is $0.94 \pm 0.001$.
-This implies that almost no predictive structure exists in this structure.
-Recall that a WPE $\approx 1$ are similar in complexity to white noise or a random walk process as they do not transmit any information from past to future.
-This would mean that the WPE of $>0.95$ would suggest little to no predictive structure exists and that the signal is simply too complex.
-In this case it should be safe to assume that a random walk forecast would be (on-average) superior to forecast methods which required structure or information to intelligently predict the next time step.
-This is in fact what we see, with each prediction scheme we used we see a MASE which implies (on-average) 1.5-1.8 times worse predictions that performing a simple random walk forecast.
+Though \col is a very simple program, its dynamics are actually quite
+complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
+right-hand column of Figure~\ref{fig:gcc_vs_col} that the naive and
+ARIMA prediction methods perform poorly on this signal (MASE values
+$0.57 \pm$ 0.001 and $0.59 \pm 0.21$, respectively).  That is, the
+errors produced by these methods are a little over half as large as
+the errors produced by the random-walk method, which simply uses the
+current value as the prediction.  However, the WPE value for the \col
+trials is $0.51 \pm 0.003$: i.e., right in the middle of the
+complexity spectrum described in Section~\ref{sec:intro}, implying
+that roughly half of the signal is predictive structure.  This
+disparity---WPE values that suggest a high rate of forward information
+transfer in the signal, but predictions with low MASE
+scores---manifests geometrically in the top two images in
+Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
+the right of the dashed line.  This is important; it says that the
+methods aren't leveraging that information very well.  This is not
+surprising, really.  The dynamics of \col may be complicated, but they
+are not unstructured.  This signal is nonlinear, deterministic, and
+chaotic.  Indeed, if you use a prediction technique that uses a
+nonlinear model (LMA) you get a MASE of $0.05 \pm 0.001$---i.e., 20
+times more accurate than a random-walk forecast.  Note the position of
+the \col cluster in the bottom image in
+Figure~\ref{fig:wpe_vs_mase_all}: much closer to the dashed line.
+From a practitioner's standpoint, what this means is that being below
+that line tells you that there really IS usable structure in the time
+series, that you could do better, and that you should use a better
+method.  Note, too, that the \col points are clustered in LMA but
+spread out horizontally in ARIMA, far to the right of the dashed line
+(i.e., much worse than the other methods).  We need to come up with a
+hypothesis about this spread and skew, even if it's totally tentative.
 
-There is a single outlier in this study which is \svdone.
-This signal has a WPE of $0.95 \pm 0.02$ and both LMA and ARIMA do slightly better than random walk ($0.82 \pm 0.08$ and $0.7141 \pm 0.07$ respectively).
-The na\"ive method did significantly worse than random walk and was highly inconsistent with an average MASE of 2.67 but a standard deviation of 4.33 due to the massive right tail this distribution exhibited in reaction to wildly inconsistent forecasts.
-We do not believe this outlier is really a problem but it does bring about a potential weaknesses of this method.
-MASE is scaled by a random walk forecast which can be very inconsistent in the case of drastic signal changes which occur on a regular basis, e.g., a signal which oscillates from the top to the bottom of the signals scale at every time click.
-In this case a random walk forecast would have error with magnitude of the scale of the signal at every forecast, since guessing the last step will be exactly out of phase with the new forecast.
-A signal like this would normally cause a very low WPE much like \col and a very high Random Walk error resulting in low MASE for methods that can take into account this structure like ARIMA or LMA.
+The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
+  col\_major}.  This indicates that the rate of forward information
+transfer of the underlying process is lower, but that time-series data
+from this system still contain a significant amount of structure that
+can, in theory, be leveraged to predict the future course of the time
+series.  However, the effectiveness of any prediction strategy will
+depend on how well it leverages the available information.
+% 
+% (This is the basis for the second conjecture above.)  
+% 
+This issue becomes apparent in the case of \svdfive, as is clear from
+the positions of the corresponding clusters in
+Figure~\ref{fig:wpe_vs_mase_all}.  The naive and ARIMA prediction
+strategies in the top two plots produce MASE scores of $2.3700 \pm
+0.0505 $ and $20.8703 \pm 0.1915$, respectively: that is, 2.37 and
+20.87 times worse than a simple random walk forecast of the same
+signals.  [[This is big because the amount of variance in this signal
+    makes it a bad idea to just guess the mean.  The same thing is
+    true of \svdthree.]] [[Joshua: can you say that better---in a way
+    that is true of both of these, and distinguishes them from all of
+    the signals that have lower values?]] Again, the positions of the
+ARIMA and LMA clusters---significantly below and to the right of the
+dashed line---should suggest to a practitioner that there is more
+structure in this signal than the method is able to leverage, so s/he
+can do better and should try a different prediction method.  In this
+case, the LMA method produces a MASE score of $ 0.7177\pm 0.0483 $ and
+a cluster of results on the WPE-MASE plot that is much closer to the
+dashed line.  This validates our second conjecture: the naive and
+ARIMA prediction methods, which use constant and linear models,
+respectively, cannot capture or reproduce the way in which the
+\svdfive system processes information, but the nonlinear LMA method
+can.
 
-To understand why this causes \svdone to be an outlier we examine Figure \ref{fig:svdone-ts} in more detail.
-This is a small snippet of the time series of \svdone to illustrate the strucutre of this signal.
-We provide a small snippet as opposed to the full time series simply for visual clarity.
+The WPE of \gcc is much higher: $0.94 \pm 0.001$.  This system
+transmits very little information forward in time and provides almost
+no structure for prediction methods to work with.  What we see: naive
+does the best, then LMA, then ARIMA.  Hypothesis: that that 5\% of the
+structure is nonlinear, so LMA can use it better than ARIMA can.
+Naive does better than LMA because it's filtering out the noise.
+
+\svdone behaves very differently than the other seven systems in this
+study: though its WPE is very high ($0.95 \pm 0.02$), two of the three
+prediction methods (LMA and ARIMA) do quite well: $0.82 \pm 0.08$ and
+$0.7141 \pm 0.07$ MASE scores, respectively.  Accordingly, the
+clusters of \svdone points in the bottom two images in
+Figure~\ref{fig:wpe_vs_mase_all} are well above the trend followed by
+the other seven clusters.  Also, the MASE scores of the predictions
+produced by the na\"ive method are highly inconsistent: $2.67 \pm
+4.33$.  These effects, we believe, are artifacts of the way MASE is
+calculated.  Recall that MASE scores are scaled relative to a random
+walk forecast, which uses the current value as the prediction.  This
+strategy works very badly on signals with frequent, large, rapid
+transitions.  Consider a signal that oscillates from one end of its
+range to the other at every step.  A signal like this will have a low
+WPE, much like \col.  However, a random-walk forecast of this signal
+will be 180 degrees out of phase with the true continuation.  Since
+random-walk error appears in the denominator of the MASE score, this
+effect can shift points leftwards on a WPE vs. MASE plot, and that is
+exactly why the \svdone clusters for ARIMA and LMA are above the
+dashed line in Figure~\ref{fig:wpe_vs_mase_all}.  The \svdone time
+series, which is shown in Figure~\ref{fig:svdone-ts},
 \begin{figure}[htbp]
@@
     \includegraphics[width=\textwidth]{figs/svdonets2}
-\caption{ For each of these, the word length used is $6$.
-The MASE values for LMA against ARIMA.
-The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better.
-All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:svdone-ts}
+\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
 \end{figure} 
-As can be seen in Figure \ref{fig:svdone-ts}, \svdone has a highly unstructured small band on top, which is approximately 0.05 instructions per cycle wide.
-Approximately 80\% of the signal exists within this band of noise, this causes the WPE to be driven up as many words are being omitted by this band of the signal.
-The remaining 20\% of the signal are drops in instructions per cycle away from this band, ranging in size from 0.1 to 0.6 instructions per cycle\footnote{Some of the large drops are not shown for the sake of clarity of the smaller scale}.
-These drastic drops seem to happen every 5-7 points but the drop locations are inconsistent as well as the drops magnitude (ranging from 0.1-0.6 instructions per cycle).
-As discussed in Section \ref{sec:RWMethod} drops of this fashion are the bane of random walk forecasting.
-At every one of these drops a random walk predictor would produce two forecasts with an error 2-12 times the magnitude of the smaller band, i.e., about 40\% of the time the random walk forecast would predict with error 2-12 times the magnitude of the scale exhibited by $\approx$80\% of the signal.
-This inconsistency in random walk would cause cause random walk error to be high and as a result MASE scores to be artificially low.
-The combination of a band of noise with a high number of random drops is a weakness of this method which is present in \svdone and as a result makes this signal an outlier.
-
-[[Joshua: I am not sure if this paragraph is strong enough or even worth having]]These results support that time series with low to moderate complexity ($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive random walk \emph{and} that complexity can be qualitatively measured for a real-valued noisy time series using WPE.
-This will allow practitioners to stop spinning their wheels in the case of signals who are simply better predicted with a simple strategy like random walk.
-The analysis of these results illuminates an interesting point: The way structure, information and complexity are processed by a generating process plays a crucial role in the success of a given prediction scheme.
+is not quite to the level of the worst-case signal described above,
+but it poses a serious challenge to random-walk prediction.  It is
+dominated by a white-noise regime (between 1.86 and 1.88 on the
+vertical scale in the Figure), punctuated by short excursions above
+1.9.  In the former regime, which makes up more than 80\% of the
+signal, there are frequent dips to 1.82 and occasional larger dips
+below 1.8.  These single-point dips are the bane of random walk
+forecasting.  In this particular case, roughly 40\% of the forecasted
+points are off by the width of the associated dip, which skews the
+associated MASE scores.  Signals like this are also problematic for
+the naive prediction strategy, since the outliers have significant
+pull on the mean, which compounds the effect of the skew in the
+scaling factor and creates a large spread in the \svdone MASE values.
+For all of these reasons, we view \svdone as an outlier and thus
+exclude it from the fit calculation of the dashed line in
+Figure~\ref{fig:wpe_vs_mase_best}.
 
@@
 
-Comparing Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} also elucidates another key finding: usable and quantifiable predictive structure can be present in a time series without a prediction scheme being able to utilize it.
-In particular, information may be transferred from past to future through the present but because of the mechanism the underlying process uses to process that information  (e.g., linear or nonlinear) certain prediction strategies may be blind to or not be able to efficiently utilize this information.
-For example, consider \col, programs like this have been shown to exhibit deterministic chaos \cite{mytkowicz09}.
-If this were the case with \col, an out-of-the-box linear method like ARIMA would simply be ill-equipped to model and utilize the kind of structure present, as is evident in Figure \ref{fig:arima_pred_vs_ent}.
-In contrast, a nonlinear predictor like LMA which is built to handle deterministic chaos, can interpret and utilize this type of structure just fine.
-We believe that many of the shifts in forecast accuracy for low-to-moderate WPE programs between ARIMA and LMA is precisely happening for this reason: Just because there is forward information transfer, does not mean that an arbitrary predictor can interpret or utilize it, but luckily WPE can tell us when this structure is present as shown in Figure \ref{fig:lma_vs_arima}.
+Comparing Figures~\ref{fig:lma_pred_vs_ent} and
+\ref{fig:arima_pred_vs_ent} also elucidates another key finding:
+usable and quantifiable predictive structure can be present in a time
+series without a prediction scheme being able to utilize it. In
+particular, information may be transferred from past to future through
+the present but because of the mechanism the underlying process uses
+to process that information (e.g., linear or nonlinear) certain
+prediction strategies may be blind to or not be able to efficiently
+utilize this information. For example, consider \col, programs like
+this have been shown to exhibit deterministic chaos
+\cite{mytkowicz09}. If this were the case with \col, an out-of-the-box
+linear method like ARIMA would simply be ill-equipped to model and
+utilize the kind of structure present, as is evident in Figure
+\ref{fig:arima_pred_vs_ent}. In contrast, a nonlinear predictor like
+LMA which is built to handle deterministic chaos, can interpret and
+utilize this type of structure just fine. We believe that many of the
+shifts in forecast accuracy for low-to-moderate WPE programs between
+ARIMA and LMA is precisely happening for this reason: Just because
+there is forward information transfer, does not mean that an arbitrary
+predictor can interpret or utilize it, but luckily WPE can tell us
+when this structure is present as shown in Figure
+\ref{fig:lma_vs_arima}.
 
@@
 
-Other noteworthy features of the LMA and ARIMA results are the cluster locations
-and distributions. The WPE values of each run of any particular program tend to
-have little variance, leading to the clusters in
-Figures~\ref{fig:arima_pred_vs_ent} and \ref{fig:lma_pred_vs_ent} to be fairly
-constrained in the $y$ direction. For most traces, the LMA and ARIMA variance is
-low as well, resulting is small, tight clusters. The ARIMA MASE values of the \col
-traces, however, have a large variance resulting in the spread seen in
-Figure~\ref{fig:arima_pred_vs_ent}. Not only are the MASE values of that cluster
-bad, in that other predictors vastly outperform it, but they are inconsistent.
-Furthermore, since LMA can predict nonlinear behavior while ARIMA cannot, we
-see that the clusters in Figure ~\ref{fig:arima_pred_vs_ent} are mostly further to
-the right than those in Figure~\ref{fig:lma_pred_vs_ent}.
+The big picture: Furthermore, since LMA can predict nonlinear behavior
+while ARIMA cannot, we see that the clusters in Figure
+~\ref{fig:arima_pred_vs_ent} are mostly further to the right than
+those in Figure~\ref{fig:lma_pred_vs_ent}.

### 2014-02-27 pecomplexity.tex
Fixing pecomplexity.tex
sig figs and rearrange table

--- 2014-02-19 pecomplexity.tex
@@
 Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity
+\\ Quantifying Predictability through Structural Complexity \\
+{\color{green}Quantifying Predictability through Structural Complexity} 
 \\ Quantifying Predictability using Structural Complexity 
+\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
+\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
 }
@@
 \input{intro.tex}
-
+\input{related.tex}
 \input{methods.tex}
@@
 
+\section{Appendix}
+
+\begin{table}[h]
+  \begin{center}
+  \begin{tabular}{|c|c|c|c|c|}
+  \hline
+            & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
+ \hline
+
+ \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\
+
+\gcc           & $ 1.530\pm 0.021$ & $1.837 \pm0.016 $ & $0.951 \pm 0.001$ & $0.943 \pm 0.001$ \\
+
+\svdone     & $ 0.827\pm 0.076$ & $ 0.714\pm 0.075 $ & $2.676\pm4.328$&  $0.957 \pm 0.016$ \\
+
+ \svdtwo    & $1.279 \pm0.020 $ & $2.163 \pm0.027 $ &  $3.054\pm0.040$ &   $0.846 \pm0.004$ \\
+
+ \svdthree     & $0.619 \pm0.021 $ & $0.713 \pm 0.010 $ & $31.386\pm 0.282$ &  $0.716 \pm 0.006$ \\
+
+ \svdfour     & $ 0.779\pm0.036 $ & $0.979 \pm0.032 $ & $2.661\pm0.074$ & $0.825 \pm 0.008$ \\
+
+ \svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
+
+ \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
+  \hline
+  \end{tabular}
+  \end{center}
+ \label{default}
+ \caption{Detailed MASE scores}
+ \label{tab:error}
+  \end{table}%
+
 \section*{Acknowledgment}

### 2014-02-27 results.tex
finished the results section, mostly
moved a small chunk to concl
defined a newcommand
added usepackage{color}

@@
 
-\item that the way information is processed internally by a system is
-  correlated with the effectiveness of a given predictor on
+\item that the way information is generated and processed by a system
+  is correlated with the effectiveness of a given predictor on
   time-series data from that system
@@
 
-The experiments below involve the three different prediction methods
-described in Section~\ref{sec:compModel}---na\"ive, ARIMA, and
-LMA---applied to time-series data from eight different systems: {\tt
+The experiments below involve three different prediction methods
+applied to time-series data from eight different systems: {\tt
   col\_major}, {\tt 403.gcc}, and the six different regimes of the
@@
 prediction method.  Working from the first 90\% of each signal, we
-generated a prediction of the last 10\% of that signal using each of
-the methods, as described in Section~\ref{sec:accuracy}, then
-calculated the MASE value of those predictions.  We also calculated
-the WPE of each time series, as described in
-Section~\ref{sec:meaComplex}, using a wordlength of six.  In order to
-assess the run-to-run variability of these results, we repeated all of
-these calculations on 15 separate trials: i.e., 15 different runs of
-each program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best}
-of these 45 predictions for each system: i.e., the lowest error over
-all 15 trials and all three methods for each program.  The WPE is
-plotted against the corresponding MASE value in order to bring out the
-correlation between these two quantities.
+generated a prediction of the last 10\% using the \naive, ARIMA, and
+LMA prediction methods, as described in Section~\ref{sec:accuracy},
+then calculated the MASE value of those predictions.  We also
+calculated the WPE of each time series using a wordlength of six, as
+described in Section~\ref{sec:meaComplex}.  In order to assess the
+run-to-run variability of these results, we repeated all of these
+calculations on 15 separate trials: i.e., 15 different runs of each
+program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
+these 45 predictions for each system: i.e., the lowest error over all
+15 trials and all three methods for each of the eight programs.  The
+WPE is plotted against the corresponding MASE value here in order to
+bring out the correlation between these two quantities.
 \begin{figure}[htbp]
@@
 consistent with our first conjecture.  The dashed line in the figure,
-a linear least-squares fit to these best-case errors---with the
-exception of {\tt dgesdd$_5$}, for reasons described below---captures
+a linear least-squares fit to these best-case errors\footnote{with the
+  exception of {\tt dgesdd$_1$}, for reasons described below} captures
 this rough proportionality.  This is not a formal result.  The three
 methods used here were chosen to span the space of standard prediction
-strategies, but they do not, of course, cover that space exhaustively.
-Our goal here is an empirical assessment of predictability and
-complexity, not formal results about a ``best'' predictor for a given
-time series.  There may be other methods that produce lower MASE
-values than those in Figure~\ref{fig:wpe_vs_mase_best}, but the
-sparseness of the points in the upper-left and lower-right quadrants
-of this plot strongly suggests that the underlying predictability of a
-time series is inversely correlated with its WPE.  The rest of this
-section describes our results in more detail, including the measures
-taken to assure meaningful comparisons across methods, trials, and
-programs, and elaborates on the meaning of the dashed line in the
-figure.
+strategies, but they do not cover that space exhaustively.  Our goal
+here is an empirical assessment of predictability and complexity, not
+formal results about a ``best'' predictor for a given time series.
+There may be other methods that produce lower MASE values than those
+in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
+in the upper-left and lower-right quadrants of this plot strongly
+suggests that the underlying predictability of a time series is
+inversely correlated with its WPE.  The rest of this section describes
+our results in more detail, including the measures taken to assure
+meaningful comparisons across methods, trials, and programs, and
+elaborates on the meaning of the dashed line in the figure.
 
@@
   \svdthree, and \svdfive are omitted from the top plot for scale
-  reasons; their MASE scores are $2.6763\pm4.3282$, $31.3857\pm
-  0.2820$, and $20.8703 \pm 0.1915$ respectively.  {\color{red}Replace
+  reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm
+  0.282$, and $20.870 \pm 0.192$ respectively.  {\color{red}Replace
     the top plot with a plot of the naive results---but running from 0
@@
 \end{figure} 
-There are 15 points in each cluster, one for each trial\footnote{A
-  trial involves collecting a time series from the corresponding
-  system, predicting the last 10\% of that time series, and
-  calculating the MASE score.}.  (The points in
-Figure~\ref{fig:wpe_vs_mase_best} are the leftmost points in any of
-the three clusters for the corresponding system.)  Most of the
-clusters are quite tight.  Some are flat and wide because the WPE
-values of each run of any particular program tend to have little
-variance; for most traces, the LMA and ARIMA variance is low as well,
-resulting in small, tight clusters.  The {\sl spatial distribution} of
-the clusters is quite different in the three plots: LMA is tight
-around the dashed line, whereas naive and ARIMA are [[need to fill in
-    words here when get the third image]].  This is a geometric
-validation of the second conjecture in this paper, as described in the
-following paragraphs.
+There are 15 points in each cluster, one for each trial.  (The points
+in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
+for the corresponding system in any of these three plots.)  The WPE
+values do not vary very much across trials.  For most traces, the
+variance in MASE scores is low as well, resulting in small, tight
+clusters.  In some cases---ARIMA predictions of {\tt col\_major}, for
+instance---the MASE variance is larger, which spreads out the clusters
+horizontally.  The distribution of the cluster {\sl positions} is
+quite different in the three plots: clusters of predictions generated
+with the nonlinear LMA method are closer to the dashed line, whereas
+the naive and ARIMA prediction clusters are more spread out [[need to
+    revisit this wording when get the third image]].  This is a
+geometric validation of the second conjecture in this paper, as
+described in the following paragraphs.
 
@@
 complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
-right-hand column of Figure~\ref{fig:gcc_vs_col} that the naive and
-ARIMA prediction methods perform poorly on this signal (MASE values
-$0.57 \pm$ 0.001 and $0.59 \pm 0.21$, respectively).  That is, the
-errors produced by these methods are a little over half as large as
-the errors produced by the random-walk method, which simply uses the
-current value as the prediction.  However, the WPE value for the \col
-trials is $0.51 \pm 0.003$: i.e., right in the middle of the
-complexity spectrum described in Section~\ref{sec:intro}, implying
-that roughly half of the signal is predictive structure.  This
-disparity---WPE values that suggest a high rate of forward information
-transfer in the signal, but predictions with low MASE
-scores---manifests geometrically in the top two images in
+graphical representations in the right-hand column of
+Figure~\ref{fig:gcc_vs_col} that the \naive and ARIMA prediction
+methods do not perform as well on this signal as the nonlinear LMA
+method.  The MASE values of the \naive and ARIMA predictions are $0.57
+\pm$ 0.001 and $0.59 \pm 0.21$, respectively, across all 15 trials.
+That is, the errors produced by these methods are a little over half
+as large as the errors produced by the random-walk method---a
+primitive strategy that simply uses the current value as the
+prediction.  However, the WPE value for the \col trials is $0.51 \pm
+0.003$, which is in the center of the complexity spectrum described in
+Section~\ref{sec:intro}.  [[This suggests that roughly half of the
+    signal is predictive structure.]]  [[I'm still uncomfortable with
+    that phrasing; let's revisit it after we have the terminology and
+    definitions hammered out.]]
+
+This disparity---WPE values that suggest a high rate of forward
+information transfer in the signal, but predictions with comparatively
+poor MASE scores---is obvious in the geometry of the top two images in
 Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
-the right of the dashed line.  This is important; it says that the
-methods aren't leveraging that information very well.  This is not
-surprising, really.  The dynamics of \col may be complicated, but they
-are not unstructured.  This signal is nonlinear, deterministic, and
-chaotic.  Indeed, if you use a prediction technique that uses a
-nonlinear model (LMA) you get a MASE of $0.05 \pm 0.001$---i.e., 20
-times more accurate than a random-walk forecast.  Note the position of
-the \col cluster in the bottom image in
-Figure~\ref{fig:wpe_vs_mase_all}: much closer to the dashed line.
-From a practitioner's standpoint, what this means is that being below
-that line tells you that there really IS usable structure in the time
-series, that you could do better, and that you should use a better
-method.  Note, too, that the \col points are clustered in LMA but
-spread out horizontally in ARIMA, far to the right of the dashed line
-(i.e., much worse than the other methods).  We need to come up with a
-hypothesis about this spread and skew, even if it's totally tentative.
+the right of the dashed line.  {\sl This indicates that these methods
+  are not leveraging the available information in the signal.}  The
+dynamics of \col may be complicated, but they are not unstructured.
+This signal is nonlinear and deterministic, and if one uses a
+prediction technique that is based a nonlinear model (LMA)---rather
+than a method that simply predicts the running mean (\naive) or one
+that uses a linear model (ARIMA)---the MASE score is much better:
+$0.050 \pm 0.001$.  This prediction is 20 times more accurate than a
+random-walk forecast, which is more in line with the level of
+predictive structure that the low WPE value suggests is present in the
+signal.  The dashed line is a heuristic that captures this argument.
+It can allow practitioners to recognize when a prediction method is
+not well matched to the task at hand: that is, when the time series
+has more predictive structure than the method is able to use.  When
+the predictions are well below that line (viz., the top images in
+Figure~\ref{fig:wpe_vs_mase_all}), one should try a different method.
+
+Note that the \col points are clustered in LMA but spread out
+horizontally in ARIMA, far to the right of the dashed line (i.e., much
+worse than the other methods).  [[We need to come up with a hypothesis
+    about this spread and skew, even if it's totally tentative.  We
+    may want to merge this discussion into one of the other \col
+    paragraphs---or perhaps put it with the \svdone discussion later
+    on?]]
 
@@
 can, in theory, be leveraged to predict the future course of the time
-series.  However, the effectiveness of any prediction strategy will
-depend on how well it leverages the available information.
+series.  As mentioned above, though, in the discussion of {\tt
+  col\_major}, the effectiveness of any prediction strategy will
+depend on how well it leverages the available information; methods
+that use linear models, for instance, may fail when applied to
+nonlinear processes.
 % 
@@
 % 
-This issue becomes apparent in the case of \svdfive, as is clear from
-the positions of the corresponding clusters in
-Figure~\ref{fig:wpe_vs_mase_all}.  The naive and ARIMA prediction
-strategies in the top two plots produce MASE scores of $2.3700 \pm
-0.0505 $ and $20.8703 \pm 0.1915$, respectively: that is, 2.37 and
-20.87 times worse than a simple random walk forecast of the same
-signals.  [[This is big because the amount of variance in this signal
-    makes it a bad idea to just guess the mean.  The same thing is
-    true of \svdthree.]] [[Joshua: can you say that better---in a way
-    that is true of both of these, and distinguishes them from all of
-    the signals that have lower values?]] Again, the positions of the
-ARIMA and LMA clusters---significantly below and to the right of the
-dashed line---should suggest to a practitioner that there is more
-structure in this signal than the method is able to leverage, so s/he
-can do better and should try a different prediction method.  In this
-case, the LMA method produces a MASE score of $ 0.7177\pm 0.0483 $ and
-a cluster of results on the WPE-MASE plot that is much closer to the
-dashed line.  This validates our second conjecture: the naive and
-ARIMA prediction methods, which use constant and linear models,
-respectively, cannot capture or reproduce the way in which the
-\svdfive system processes information, but the nonlinear LMA method
-can.
+The match between model and task is an issue in the \svdfive
+experiments as well.  The MASE scores of the \naive and ARIMA
+predictions for this system were $20.870 \pm 0.192$ and $2.370 \pm
+0.051$, respectively: that is, 20.87 and 2.37 times worse than a
+simple random walk forecast of the same signals\footnote{The \naive
+  MASE score is large because the amount of variance in this signal is
+  high, which makes guessing the mean a particularly bad strategy.
+  The same thing is true of the \svdthree signal.}.  Again, the
+positions of these points on a WPE vs. MASE plot---significantly below
+and to the right of the dashed line---should suggest to a practitioner
+that there is more structure in this signal than the corresponding
+method is able to leverage, and that a different method might do
+better.  Indeed, for \svdfive, the LMA method produces a MASE score of
+$ 0.718\pm 0.048 $ and a cluster of results that is much closer to the
+dashed line on the WPE-MASE plot.  Again, this validates our second
+conjecture: the LMA method can capture and reproduce the way in which
+the \svdfive system processes information, but the \naive and ARIMA
+prediction methods cannot.
 
-The WPE of \gcc is much higher: $0.94 \pm 0.001$.  This system
+The WPE of \gcc is much higher: $0.940 \pm 0.001$.  This system
 transmits very little information forward in time and provides almost
-no structure for prediction methods to work with.  What we see: naive
-does the best, then LMA, then ARIMA.  Hypothesis: that that 5\% of the
-structure is nonlinear, so LMA can use it better than ARIMA can.
-Naive does better than LMA because it's filtering out the noise.
+no structure for prediction methods to work with.  [[What we see:
+    \naive does the best, then LMA, then ARIMA.  Hypothesis: that that
+    5\% of the structure is nonlinear, so LMA can use it better than
+    ARIMA can.  \naive does better than LMA because it's filtering out
+    the noise.]]
 
@@
 study: though its WPE is very high ($0.95 \pm 0.02$), two of the three
-prediction methods (LMA and ARIMA) do quite well: $0.82 \pm 0.08$ and
-$0.7141 \pm 0.07$ MASE scores, respectively.  Accordingly, the
-clusters of \svdone points in the bottom two images in
-Figure~\ref{fig:wpe_vs_mase_all} are well above the trend followed by
-the other seven clusters.  Also, the MASE scores of the predictions
-produced by the na\"ive method are highly inconsistent: $2.67 \pm
-4.33$.  These effects, we believe, are artifacts of the way MASE is
-calculated.  Recall that MASE scores are scaled relative to a random
-walk forecast, which uses the current value as the prediction.  This
-strategy works very badly on signals with frequent, large, rapid
+prediction methods do quite well, yielding MASE scores of $0.82 \pm
+0.08$ (LMA) and $0.714 \pm 0.07$ (ARIMA).  This pushes the
+corresponding clusters of points in the bottom two images in
+Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
+other seven clusters.  Also, the MASE scores of the predictions of
+this system produced by the \naive method are highly inconsistent:
+$2.67 \pm 4.33$.  These effects, we believe, are artifacts of the way
+MASE is calculated.  Recall that MASE scores are scaled relative to a
+random-walk forecast, which uses the current value as the prediction.
+This strategy works very badly on signals with frequent, large, rapid
 transitions.  Consider a signal that oscillates from one end of its
 range to the other at every step.  A signal like this will have a low
-WPE, much like \col.  However, a random-walk forecast of this signal
-will be 180 degrees out of phase with the true continuation.  Since
-random-walk error appears in the denominator of the MASE score, this
-effect can shift points leftwards on a WPE vs. MASE plot, and that is
-exactly why the \svdone clusters for ARIMA and LMA are above the
-dashed line in Figure~\ref{fig:wpe_vs_mase_all}.  The \svdone time
-series, which is shown in Figure~\ref{fig:svdone-ts},
+WPE, much like {\tt col\_major}.  However, a random-walk forecast of
+this signal will be 180 degrees out of phase with the true
+continuation.  Since random-walk error appears in the denominator of
+the MASE score, this effect can shift points leftwards on a WPE
+vs. MASE plot, and that is exactly why the \svdone clusters for ARIMA
+and LMA are above the dashed line in Figure~\ref{fig:wpe_vs_mase_all}.
+This time series, which is shown in closeup in
+Figure~\ref{fig:svdone-ts},
 \begin{figure}[htbp]
@@
 is not quite to the level of the worst-case signal described above,
-but it poses a serious challenge to random-walk prediction.  It is
-dominated by a white-noise regime (between 1.86 and 1.88 on the
-vertical scale in the Figure), punctuated by short excursions above
-1.9.  In the former regime, which makes up more than 80\% of the
-signal, there are frequent dips to 1.82 and occasional larger dips
-below 1.8.  These single-point dips are the bane of random walk
-forecasting.  In this particular case, roughly 40\% of the forecasted
-points are off by the width of the associated dip, which skews the
-associated MASE scores.  Signals like this are also problematic for
-the naive prediction strategy, since the outliers have significant
-pull on the mean, which compounds the effect of the skew in the
-scaling factor and creates a large spread in the \svdone MASE values.
-For all of these reasons, we view \svdone as an outlier and thus
-exclude it from the fit calculation of the dashed line in
-Figure~\ref{fig:wpe_vs_mase_best}.
+but it still poses a serious challenge to random-walk prediction.  It
+is dominated by a white-noise regime (between $\approx$1.86 and
+$\approx$1.88 on the vertical scale in the Figure), punctuated by
+short excursions above 1.9.  In the former regime, which makes up more
+than 80\% of the signal, there are frequent dips to 1.82 and
+occasional larger dips below 1.8.  These single-point dips are the
+bane of random-walk forecasting.  In this particular case, roughly
+40\% of the forecasted points are off by the width of the associated
+dip, which skews the associated MASE scores.  Signals like this are
+also problematic for the \naive prediction strategy, since the
+outliers have significant influence on the mean.  This compounds the
+effect of the skew in the scaling factor and creates a large spread in
+the \svdone MASE values.  For all of these reasons, we view \svdone as
+an outlier and exclude it from the fit calculation of the dashed line
+in Figure~\ref{fig:wpe_vs_mase_best}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%
-
-Comparing Figures~\ref{fig:lma_pred_vs_ent} and
-\ref{fig:arima_pred_vs_ent} also elucidates another key finding:
-usable and quantifiable predictive structure can be present in a time
-series without a prediction scheme being able to utilize it. In
-particular, information may be transferred from past to future through
-the present but because of the mechanism the underlying process uses
-to process that information (e.g., linear or nonlinear) certain
-prediction strategies may be blind to or not be able to efficiently
-utilize this information. For example, consider \col, programs like
-this have been shown to exhibit deterministic chaos
-\cite{mytkowicz09}. If this were the case with \col, an out-of-the-box
-linear method like ARIMA would simply be ill-equipped to model and
-utilize the kind of structure present, as is evident in Figure
-\ref{fig:arima_pred_vs_ent}. In contrast, a nonlinear predictor like
-LMA which is built to handle deterministic chaos, can interpret and
-utilize this type of structure just fine. We believe that many of the
-shifts in forecast accuracy for low-to-moderate WPE programs between
-ARIMA and LMA is precisely happening for this reason: Just because
-there is forward information transfer, does not mean that an arbitrary
-predictor can interpret or utilize it, but luckily WPE can tell us
-when this structure is present as shown in Figure
-\ref{fig:lma_vs_arima}.
 
@@
 However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.
-
-The big picture: Furthermore, since LMA can predict nonlinear behavior
-while ARIMA cannot, we see that the clusters in Figure
-~\ref{fig:arima_pred_vs_ent} are mostly further to the right than
-those in Figure~\ref{fig:lma_pred_vs_ent}.

### 2014-02-27 intro.tex

@@
 
-Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{sfi-prediction}, let alone an arbitrary time series.
+Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series.
 This naturally leads to an important and hard question: given a complicated real-valued time series, is it possible to reliably quantify where on this spectrum the time series lies?

### 2014-02-27 concl.tex

@@
 \end{enumerate}
+
+Moved the following material here from the results section, since
+they're more general.  But there's a lot of repeated material that
+will obviously need to get compressed and streamlined.
+
+Just because there is forward information transfer, that does not mean
+that an arbitrary predictor can interpret or utilize it, but WPE can
+tell us when this structure is present.
+
+These results support that time series with low to moderate complexity
+($0\le WPE \le 0.85$) can be predicted more efficiently than a na\"ive
+random walk \emph{and} that complexity can be qualitatively measured
+for a real-valued noisy time series using WPE. This will allow
+practitioners to stop spinning their wheels in the case of signals who
+are simply better predicted with a simple strategy like random
+walk. The analysis of these results illuminates an interesting point:
+The way structure, information and complexity are processed by a
+generating process plays a crucial role in the success of a given
+prediction scheme.
+
+As we discussed in Section \ref{sec:meaComplex} distinguishing
+structured from unstructured complexity in the case of real-valued
+time series is non trivial, i.e., distinguishing when a complex signal
+omits predictable structure and when the signal is effectively random
+is not a trivial task. As described in Section \ref{sec:intro}, for a
+practitioner this can be frustrating because it can be nearly
+impossible to find the source of faulty predictions: Is it simply that
+we need to use a more ideal (possibly more advanced) predictor or is
+it the case that the time series is simply so complex that using a
+simple (yet inconsistent) forecast strategy such as random walk is the
+best we can do. Fortunately, weighted permutation entropy (WPE) allows
+us to make this distinction for noisy real-valued time series.
+
+Needs to be more stuff in the WPE section for this to tie back to as
+well as in the new related work section.  [[Joshua, can you elaborate
+    on this so that Ryan can fill this in?]]
 

### 2014-02-28 wpe.tex
editing

@@
   \centering
-
   %\begin{subfigure}{0.3\textwidth}
@@
  % \begin{subfigure}{\textwidth}
+
+%Liz commented this image out temporarily so that her mac doesn't hang
+%when she scrolls past p12
     \includegraphics[width=\textwidth]{figs/SVD_wwpe}
     %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-  \caption{
-[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly.
-Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD.
-The gray bands
-    are regions where the window overlaps regimes. The window size used is
-    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
-Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE.
-From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
+  \caption{The weighted permutation entropy of one run of
+  SVD. The gray bands are regions where the window overlaps
+  regimes. The window size used is 5,000 $\times$ 100,000 instructions
+  and the word length is $6$. {\color{red} make sure this is actually updated to 6} For reference the instructions per cycle
+  of \svd are plotted as a ghost behind this plot. Each color on the
+  ghosted time series corresonds to the different regimes as selected
+  by rapid shifts in WPE. From left to right each change in color
+  represents a change in regime for 6 regimes in
+  total.}\label{fig:wwpe}
 \end{figure}

### 2014-02-28 model.tex
Adding to and editing intro
trying again
head
some cleanup on the results section, following conversation with Joshua
figure update and clean up
adding sectoin on simple methods and cleaning

@@
-\section{Modeling}
+\section{Modeling}\label{sec:model}
+
 Section outline:
@@
 \begin{enumerate}
-
+\item \cmark section intro
+\item \cmark fix order from simplest to most complex
 \item Description of DCE and parameter estimation  
@@
 \subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
-\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
+\item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
 \item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
@@
 
-\label{sec:compModel}
+In this section we describe the four different forecasting schema used in this study.
+In Section \ref{sec:simple} we outline two primitive strategies which often out perform sophisticated prediction strategies in the case of highly complex signals.
+In Section \ref{sec:arima} we describe Autoregressive integrated moving-average (ARIMA) a commonly used linear forecasting scheme.
+In Section \ref{sec:lma} we describe a simple nonlinear deterministic forecasting scheme, Lorenz Method of Analogous (LMA) for capturing and exploiting nonlinear dynamics present in a time series.
+And finally in Section \ref{sec:accuracy} we discuss the error metric we use to evaluate each models predictive accuracy.
 
-\subsection{Lorenz Method of Analogues (LMA)}
+\subsection{Primitive Strategies to combat complexity: Random Walk and \naive}\label{sec:simple}
+In this section we cover two very simple forecast models: Random Walk and \naive.
+With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement, i.e., $p_i = X_{i-1,obs}$.
+The \naive prediction strategy is to forecast the rolling average up to that point.
+i.e., the prediction $p_i$ is simply the average of all observed points up to that point: $$p_i = \sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}$$.
+
+While both of these methods are simplistic they are not without merit.
+When time series possess very little predictive structure,i.e., they exist on the very top of the complexity spectrum these two methods can sometimes be \emph{on-average} very good choices for a forecasting method, as we will see in our results.
+An example is forecasting currency exchange rates.
+In \cite{rwMeese,rwCCE} the authors show that various econometric techniques and prediction models fail to consistently outperform Random Walk predictions in forecasting various exchange rates.
+This is the case because these signals are volatile, noisy and possess very little information structure but they don't \emph{on-average} vary a great deal.
+That is the day-to-day exchange rates \emph{on-average} don't vary very much so guessing the last known exchange rate is often a very good approximation to the new exchange rate.
+[[Not a huge fan of this wording.]] The Na\"ive forecasting scheme can be a great prediction strategy if the time series has very little forward information structure, adheres to a unimodal distribution and has very little variance.
+In this circumstance, forecasting the mean will on average get you in the ball park of the forecast without taking into account the lack of structure.
+The \naive forecasting scheme can be thought of as a filtering mechanism which compresses a signal onto a single point, effectively filtering out both complexity and structure.
+However, if very little predictive structure is present in a signal then this filtering mechanism effectively mitigates the complexity.
+
+While the simplicity of these methods is advantageous when a signal is mostly complexity, i.e., that is the time series exists on the high side of the complexity spectrum, the methods can be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series.
+Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths.
+Fortunately, they somewhat parallel each other, i.e., the weakness of Random Walk and \naive are covered by the other primitive method.
+The situation in which Random Walk fails is when a time series changes drastically on a regular basis.
+The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step.
+At every time step Random Walk would preciesly predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$.
+In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue and error of 0.5 at every step.
+So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.
+
+The \naive forecast fails when you have a signal which shifts between two or more regimes on a regular basis and the regimes are separated by a significant amount.
+In this case the \naive forecast will predict between the multiple regimes at every time step eventually accruing massive error.
+To see how Random Walk can cover this weakness imagine a time series which switches between more than one regime but stays in each regime for several time steps, this behaviour can be seen in some portions of the \svd trace seen in Figure \ref{fig:ts}.
+Random walk in this case could react to the regime shift within one time step, switch to the new Regime and begin forecasting nicely, where as \naive would just continue to predict between the modes of the multimodal distribution resulting in poor forecasting accuracy all of the time.
+
+So clearly these primitive methods have their downfalls, but it is important to keep in mind that when a time series exists on the top of the complexity spectrum they have little predictive structure to exploit and often a primitive method like Random Walk or \naive will consistently out perform methods which rely on predictive structure.
+This is explored further in Section \ref{sec:accuracy} and \ref{sec:results}
+
+\subsection{Autoregressive-integrated-moving average(ARIMA)}\label{sec:arima}
+
+\subsection{Lorenz Method of Analogues (LMA)}\label{sec:lma}
 TODO:
@@
 
-\subsection{Autoregressive-integrated-moving average(ARIMA)}
-\subsection{Naive: Random Walk and na\"ive}
 \subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE)}
+\label{sec:accuracy}
 
@@
   \centering
+      \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/colMeanForecast}
+    \caption{\col na\"ive }
+    \label{fig:gccMEAN}
+  \end{subfigure}%
    \begin{subfigure}{0.49\textwidth}
@@
   \end{subfigure}%      
+  \\
     \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colMeanForecast}
-    \caption{\col na\"ive }
-    \label{fig:gccMEAN}
-  \end{subfigure}%    
-  \\
+    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
+    \caption{\col ARIMA}
+    \label{fig:colARIMA}
+  \end{subfigure}
   \begin{subfigure}{0.49\textwidth}
@@
   \end{subfigure}%
+  \\
       \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
-    \caption{\col ARIMA}
-    \label{fig:colARIMA}
+    \includegraphics[width=\textwidth]{figs/colLMAForecast}
+    \caption{\col LMA}
+    \label{fig:colLMA}
   \end{subfigure}
-  \\
       \begin{subfigure}{0.49\textwidth}
@@
     \label{fig:gccLMA}
-  \end{subfigure}  
-    \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colLMAForecast}
-    \caption{\col LMA}
-    \label{fig:colLMA}
   \end{subfigure}
@@
 
-As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
+As a figure of merit we calculate the Mean Absolute Scaled Error (MASE)\cite{MASE} between the true and predicted signals: 
 %In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:

### 2014-02-28 model.tex
added the ARIMA section and citaitons

@@
 In this section we cover two very simple forecast models: Random Walk and \naive.
-With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement, i.e., $p_i = X_{i-1,obs}$.
+With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement: $$p_i = X_{i-1,obs}$$.
 The \naive prediction strategy is to forecast the rolling average up to that point.
@@
 
-\subsection{Autoregressive-integrated-moving average(ARIMA)}\label{sec:arima}
+\subsection{Seasonal Autoregressive-integrated-moving average(ARIMA)}\label{sec:arima}
+\begin{enumerate}
+\item introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
+\item define rigorously, all the backshift operator garbage
+
+$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = c + \Theta(B^m)\theta(B)\epsilon_i$$
+
+$$B^mX_i  = X_{i-m}$$
+\item discuss that this method needs linear structure to work correctly
+\item maybe talk about 
+\end{enumerate}
+
+For this \cite{davislinearts}
+
+When there is evidence a generating process exhibits predictive structure an obvious strategy would be to fit a hyperplane to the dataset and then use this hyperplane to forecast new data points.
+This is a very old technique stemming back to Yule in 1927\cite{weigend93} with his invention of autoregressive forecasting, which forecasts the next time step through a weighted average of the past $p$ time-steps, i.e., $$p_i = \phi(B)X_{i,obs}.$$ Here $\phi(B)$ is a polynomial of degree $p$ and $B$ is the backshift operator, i.e., $$B^mX_i  = X_{i-m,obs}.$$
+To improve this method further, especially in the presence of noisy data would be to allow for some wiggle room by adding a moving average of white noise to the model: $\theta(B)\epsilon_i$, where $\theta(B)$ is a polynomial of degree $q$ and $\{\epsilon_i\}\sim WN(0,\sigma^2)$.
+Finally if there is evidence of non-stationarity an initial differencing (integrated) is applied as an attempt to remove that non-stationarity: $(1-B^d)X_i$.
+Combining these three give us a nonseasonal  ARIMA$(p,d,q)$ model:  $$\phi(B)(1-B)^dX_i = \theta(B)\epsilon_i.$$ Where $p$, $d$ and $q$ correspond to the orders of the autoregressive, integrated and moving average models respectively.
+If periodic structure is present a seasonal ARIMA$(p,d,q)(P,D,Q)_m$ is often used and is defined by:$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = \Theta(B^m)\theta(B)\epsilon_i.$$ Here $\Phi$ and $\Theta$ are polynomials of degree P and Q respectively, and $m$ is the seasonal frequency.
+Forecasting is then done in two steps, first the integrated step is performed to remove non-stationarity.
+$$\hat{X_i} = (1-B^m)^D(1-B)^d X_{i,obs}$$.
+Then forecasts are made by evaluating $$\Phi(B^m)\phi(B)\hat{X_i} = \Theta(B^m)\theta(B)\epsilon_i.$$
+
+There is a vast amount of theory and literature on constructing, fitting and forecasting with this class of models.
+However, this is not the focus of this paper and  we refer the reader to \cite{davislinearts} for an in-depth exploration.
+In this paper, we intend to explore how predictive accuracy of several different forecasting schema correlates with signal complexity.
+As such, we use ARIMA$(p,d,q)(P,D,Q)_m$ (or simply ARIMA) models as an out of the box black box linear forecasting scheme.
+To do this we use an automated fitting and forecasting techniques described in \cite{autoARIMA}.
+This technique automatically selects model and differencing orders: $p, d, q, P, D, Q$ and $m$ as well as fitting the corresponding polynomials to the time series.
+Hyndman and Khandakar use unit root tests to select differencing orders, $d$ and $D$.
+In particular, they use a combination of
+KPSS unit-root tests \cite{KPSSunit} and their own extension of Canova-Hansen test \cite{Canova1995}.
+To do model order selection they use the Akaike information criterion (AIC) \cite{akaike1974} conditioned on the maximized likelihood of the  model fitted to the differenced data, $(1-B^m)^D(1-B)^d X_{i,obs}$ \cite{autoARIMA}.
+For more detail on this fitting process see \cite{autoARIMA}.
+
+ARIMA forecasting is a time tested modelling process which uses a global linear fit and adjustments for seasonality, non-stationarity and noise.
+If a generating process fits into one of these categories then this model is a great short term predictor.
+If information is instead being transmitted in a nonlinear way this global linear fit falls apart.
+One other weakness of this method is that if long prediction horizons are used the forecast is guaranteed to converge to the mean after some number of predictions.
+For this reason, we limit all methods to doing multiple 1-step predictions as to not handicap this method.
 
@@
 TODO:
-\begin{enumerate}
-\item update examples
-\item remove section on RMSE
-\item update figures
-\end{enumerate}
 
@@
 \]
-from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
+from a $d$-dimensional smooth compact manifold $M$ to $\mathbb{R}^{2d+1}$,
 where $t$ is time, is a diffeomorphism on $M$---in other words, that

### 2014-03-01 model.tex
updating and adding to the LMA section

@@
 
-\subsection{Seasonal Autoregressive-integrated-moving average(ARIMA)}\label{sec:arima}
+\subsection{Linear Forecasting: Seasonal Autoregressive-integrated-moving average(ARIMA)}\label{sec:arima}
 \begin{enumerate}
-\item introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
-\item define rigorously, all the backshift operator garbage
+\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
+\item\cmark define rigorously, all the backshift operator garbage
 
@@
 $$B^mX_i  = X_{i-m}$$
-\item discuss that this method needs linear structure to work correctly
-\item maybe talk about 
+\item\cmark discuss that this method needs linear structure to work correctly
+\item \cmark maybe talk about converging to zero and limits on prediction horizon
 \end{enumerate}
@@
 
-\subsection{Lorenz Method of Analogues (LMA)}\label{sec:lma}
-TODO:
+\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA)}\label{sec:lma}
+
+When a generating process exhibits deterministic predictive nonlinear structure a common approach is to reconstruct the generating processes dynamics using a technique called delay-coordinate embedding and then use the geometric structure of this reconstruction as a proxy for the information transmission process of the underlying process to predict the future.
+In this section we discuss the theory and implementation of this forecasting scheme.
 
@@
 \[
-F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
+F(\tau,m)(X) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
 \]
@@
 measuring (or even knowing) every one of its state variables.  This is
-the foundation of the modeling approach that is used in this paper.
+the foundation of this modeling approach.
 The first step in the process is to estimate values for the two free
@@
 $\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
+method of \cite{KBA92}.  A
 plot of the data from Figure~\ref{fig:ipc}, embedded following this
@@
      \includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{[[Joshua:Perhaps add a similar graphic of \gcc to show contrast]]A 3D projection of a delay-coordinate embedding of the trace
+     \caption{[[{\color{red}Joshua:Perhaps add a similar graphic of \gcc to show contrast}{\color{green}agreed I will work on this}]]A 3D projection of a delay-coordinate embedding of the trace
  from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
@@
 The coordinates of each point on this plot are differently delayed
-elements of the \col L2 cache miss rate time series
-$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
-$y(t+2\tau)$ on the third, and so on.
-%% ...down to here.
+elements of the \col instructions per cycle time series
+$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
+$X_{i+2\tau,obs}$ on the third, and so on.
 Structure in these kinds of plots---clearly visible in
 Figure~\ref{fig:embedding}---is an indication of
-determinism\footnote{A deeper analysis of
-  Figure~\ref{fig:embedding}---as alluded to on the previous
-  page---supports that diagnosis, confirming the presence of a chaotic
-  attractor in these cache-miss dynamics, with largest Lyapunov
-  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
-  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
-structure can also be used to build a forecast model.
-
-% took out for space
+determinism%\footnote{A deeper analysis of
+  %Figure~\ref{fig:embedding}---as alluded to on the previous
+  %page---supports that diagnosis, confirming the presence of a chaotic
+  %attractor in these cache-miss dynamics, with largest Lyapunov
+  %exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
+  %12-dimensional reconstruction space \cite{mytkowicz09}.}
+  .  That structure can be used to build a forecast model.
  \subsubsection{LMA: Using dynamics in forecasting}
@@
 developed by the dynamical systems community for this purpose
-(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
+(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  Perhaps the most straightforward
 is the ``Lorenz method of analogues'' (LMA), which is essentially
@@
 forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
-shown in Figure~\ref{fig:cachePredTS}.
+shown in Figure~\ref{fig:lmacol}.
 %
-%\begin{figure}[htbp]
-%  \centering
-%    \includegraphics[width=0.5\textwidth]{figs/%colCachePredTS}
-%    \caption{A forecast of the last 4,000 points of the %signal in
-%      Figure~\ref{fig:ipc} using an LMA-based strategy on the
-%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
-%      $\times$s are the true and predicted values, %respectively;
-%      vertical bars show where these values differ. }
-%\label{fig:cachePredTS}
-%\end{figure}
+\begin{figure}[htbp]
+  \centering
+   \includegraphics[width=\textwidth]{figs/colPredShortTS}
+    \caption{A forecast of the last 4,000 points of the signal in
+      Figure~\ref{fig:ipc} using an LMA-based strategy on the
+      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
+      $\times$s are the true and predicted values, respectively;
+      vertical bars show where these values differ.}
+\label{fig:lmacol}
+\end{figure}
 %
 On the other hand, if we use the same approach to forecast the
-processor load\footnote{Instructions per cycle, or IPC} of the {\tt
-  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
-on an Intel i7\textsuperscript{\textregistered}-based machine, the
-prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
-%
-%\begin{figure}[htbp]
-%  \centering
-%    \includegraphics[width=0.5\textwidth]{figs/%sphinxPredicTS}
-%     \caption{An LMA-based forecast of the last 4,000 points %of a
-%       processor-load performance trace from the {\tt %482.sphinx3}
-%       benchmark.  Red circles and blue $\times$s are the %true and
-%       predicted values, respectively; vertical bars show %where these
-%       values differ.}
-%\label{fig:predsphinx}
-%\end{figure}
+processor load (ipc) of the \gcc program from the SPEC cpu2006 benchmark suite, the
+prediction is far less accurate; see Figure~\ref{fig:lmagcc}.
 
-%Table~\ref{tab:PredError} presents detailed results about the prediction accuracy of this algorithm on four different examples: the {\tt col\_major} and {\tt 482.sphinx3} programs in Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as another simple microkernel that initializes the same matrix as {\tt col\_major}, but in row-major order, and another complex program ({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.
-Both microkernels were run on the Intel Core Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks were run on the Intel i7\textsuperscript{\textregistered} machine.
-We calculated a figure of merit for each prediction as follows.
-We held back the last $k$ elements\footnote{Several different prediction   horizons were analyzed in our experiment; the results reported in this paper are for $k$=4000} of the $N$ points in each measured time series, built the forecast model by embedding the first $N-k$ points, used that embedding and the LMA method to predict the next $k$ points, then computed the Root Mean Squared Error (RMSE) between the true and predicted signals: $$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$ To compare the success of predictions across signals with different units, we normalized RMSE as follows: $$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
-%
-% The smaller the nRMSE, obviously, the more accurate the prediction.
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+     \caption{{\color{red} [[actually need to generate this figure]]}An LMA-based forecast of the last 4,000 points of a
+       processor-load performance trace from the \gcc
+       benchmark.  Red circles and blue $\times$s are the true and
+       predicted values, respectively; vertical bars show where these
+       values differ.}
+\label{fig:lmagcc}
+\end{figure}
 
-%The results in Table~\ref{tab:PredError} show a clear distinction between the two microkernels, whose future behavior can be predicted effectively using this simple deterministic modeling strategy, and the more-complex SPEC benchmarks, for which this prediction strategy does not work nearly as well.
-% Removed for space
-% For both processor load (IPC) and memory usage (cache-miss rate),
-% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
-% for \verb|col_major| or \verb|row_major|.
-%
-This begs the question: If these traces all come from deterministic
+This gets at the utility of the contribution of this paper: If these traces both come from deterministic
 systems---computers---then why are they not equally predictable?  Our
-conjecture is that the sheer complexity of the dynamics of the SPEC
-benchmarks running on the Intel i7\textsuperscript{\textregistered}
-machine make them effectively impossible to predict.
+conjecture is that while they come from similar systems \gcc produces processor load traces on the top of the complexity spectrum whereas \col produces processor loads with complexity in the mid to low region of the complexity spectrum.
+If this is the case then \gcc might be much better predicted using a method like Random Walk or \naive which can effectively predict in the presence of complexity.
+This is explored more rigorously in the results Section of this paper (Section \ref{sec:results}.
+So that we don't have to compare the forecast accuarcy by visual comparing the predictions we calculate a figure of merit to compare the predictions.
 

### 2014-03-02 methods.tex
adding to methods and detection

@@
-\section{Experimental Methods }
+\section{Experimental Methods and Time Series Explanations}
 
@@
   \end{subfigure}
-  \caption{In (a) the instructions executed per CPU clock cycle
+    \begin{subfigure}[t]{0.475\textwidth}
+    \includegraphics[width=0.95\textwidth]{figs/SVD1RegimesColored}
+    \caption{\svd Time Series with Regime Coloring}
+    \label{fig:svd-ts-colored}
+  \end{subfigure}
+  \caption{[[Maybe this should be just a picture of each time series and then have another figure with color after explaining heuristically the reigmes.
+In (a) the instructions executed per CPU clock cycle
     (IPC) during the execution of \col. Each point is the average IPC during a 100,000
@@
     \end{figure}
+
+\svd brings up a very interesting point: systems change over time.
+While it is clear in Figure \ref{fig:sample-ts} (a) and (b) that a single system is being observed, it also appears \emph{visually} that the complexity of these two time series is consistent over time, i.e., while they are both complicated they don't appear to increase or decrease in complexity over the range of the program.
+\svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually.
+Figure \ref{fig:svd-ts-colored} is a single trace of \svd from start to finish, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series is visually very different from section to section.
+This is caused by the code of \svd moving between different subroutines.
+We call these changes in \svd dynamics  \emph{\svd regimes} and we have colored each of the six regimes different colors in Figure \ref{fig:svd-ts-colored}.
+The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift.
+In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}.
+The purpose of this paper is not to rigorously explore regime detection but to explore quantifying complexity of a time series.
+As such, making this split simply serves as providing 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore.
+For notational convience we refer to these signals as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd.
+The regimes are labeled from left to right.

### 2014-03-02 unusedFigsTables.tex
adding to methods and detection

@@
 
-\section{Detecting changes in a time series}
-
-Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}.
-To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}.
-Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block.
-(A window shift of 1.)} (or nonoverlapping) blocks.
-Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.
-According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
-
-In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy.
-This is identical to the lag used in delay coordinate embedding and often the same language is used.
-Cao et al.
-describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space.
-When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$.
-So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$.
-Cao et al.
-argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1.
-Cao et al.
-then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter.
-The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
-
-\subsection{Example: Concatenated Time Series}\label{sec:concatenate}
-As an example of this detection, consider Figure \ref{fig:ts-banded} (a).
-This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise.
-Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points.
-The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population.
-Each Logistic map trajectory begins at $x=0.6$.
-The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
-
-The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$).
-Visually the bands of noise and the chaotic trajectory are indistinguishable.
-If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise.
-If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
-
-In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves).
-The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively).
-Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}.
-On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands.
-This is in fact the case when comparing to the way the time series was constructed.
-
-The main advantage we can see to the use of lags other than 1 is for analysis of periodicity.
-This is also the major advantage listed in \cite{cao2004det}.
-Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero.
-This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}.
-If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
-
-%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
-
-%\begin{figure}
-%  \centering
-%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-alone}}
-%    {(a)}
-%
-%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-pe}}
-%  {(b)}
-%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}.
-(b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves).
-Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
-%\end{figure}
-
-and what is it good for with forward pointer to the other sections
-
-\subsection{Example: The transient Logistic map}
-In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed.
-As we showed, permutation was a great proxy to detect this change.
-Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time.
-To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}.
-Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$.
-The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map.
-The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
-
- As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well.
-An interesting feature of block calculations are the spikes that occur at bifurcation points.
-As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes.
-This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy.
-The dips you see in the second half correspond to bands of periodicity which occur between the chaos.
-It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range.
-This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
-
- This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\columnwidth]{unused-figs/tstranslog-perm}
-  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
-  \label{fig:translogistic}
-\end{figure}
-%\begin{enumerate}
-%\item do with henon bifurcation. what happens near the broken chaotic window
-
-%\end{enumerate}
-
 \section{Possible Applications to Computer Systems Performance}

### 2014-03-02 wpe.tex
adding to methods and detection

@@
 
-The weighted permutation entropy for the {\tt SVD} program is given in
+\subsection{Regime Selection:Detecting complexity shift in a time series}\label{sec:wpeRegime}
+
+{\color{red} this is from a previous paper I wrote, if this is used we need to update the plots to use wpe and windowed wpe instead of standard pe, also not sure how much of this is necessary but added as starting point}
+
+Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}.
+To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}.
+Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block.
+(A window shift of 1.)} (or nonoverlapping) blocks.
+Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.
+According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
+
+In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy.
+This is identical to the lag used in delay coordinate embedding and often the same language is used.
+Cao et al.
+describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space.
+When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$.
+So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$.
+Cao et al.
+argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1.
+Cao et al.
+then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter.
+The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
+
+\subsubsection{Example: Concatenated Time Series}\label{sec:concatenate}
+As an example of this detection, consider Figure \ref{fig:ts-banded} (a).
+This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise.
+Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points.
+The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population.
+Each Logistic map trajectory begins at $x=0.6$.
+The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
+
+The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$).
+Visually the bands of noise and the chaotic trajectory are indistinguishable.
+If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise.
+If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
+
+In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves).
+The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively).
+Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}.
+On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands.
+This is in fact the case when comparing to the way the time series was constructed.
+
+The main advantage we can see to the use of lags other than 1 is for analysis of periodicity.
+This is also the major advantage listed in \cite{cao2004det}.
+Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero.
+This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}.
+If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
+
+%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
+
+%\begin{figure}
+%  \centering
+%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-alone}}
+%    {(a)}
+%
+%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-pe}}
+%  {(b)}
+%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}.
+(b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves).
+Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
+%\end{figure}
+
+and what is it good for with forward pointer to the other sections
+
+\subsubsection{Example: The transient Logistic map}
+In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed.
+As we showed, permutation was a great proxy to detect this change.
+Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time.
+To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}.
+Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$.
+The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map.
+The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
+
+ As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well.
+An interesting feature of block calculations are the spikes that occur at bifurcation points.
+As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes.
+This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy.
+The dips you see in the second half correspond to bands of periodicity which occur between the chaos.
+It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range.
+This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
+
+ This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\columnwidth]{unused-figs/tstranslog-perm}
+  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
+  \label{fig:translogistic}
+\end{figure}
+%\begin{enumerate}
+%\item do with henon bifurcation. what happens near the broken chaotic window
+
+%\end{enumerate}
+
+The weighted permutation entropy for the \svd program is given in
 Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over

### 2014-03-02 intro.tex
adding experimental methods/collection section and updating

@@
 
-This paper is a first step in understanding when, why, and how
+{\color{red} ARIMA is not deterministic due to MA so we need to change this, actually need to change this entire paragraph} This paper is a first step in understanding when, why, and how
 deterministic forecast strategies fail when they are applied to
@@
 real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study three
+effective way to explore that conjecture.  {\color{red} Probably remove this and move as it is in the experimental methods or point to that section for more details}We study three
 examples---a simple microkernel and two complex programs: one from the

### 2014-03-02 methods.tex
adding experimental methods/collection section and updating

@@
 
+ \begin{enumerate}
+ \item  Experimental methods: (how we collect the time series and what the times series areThis should be HPM PAPI, which programs we model
+\item programs 
+\subitem description of \col
+\subitem description of \gcc
+\subitem description of \svd 
+\subitem description of svd regimes
+ \end{enumerate}
+
 \subsection{Time Series Collection}
- Experimental methods (how we collect the time series and what the times series are
-This should be HPM PAPI, which programs we model
+
+The time-series data for these experiments was collected on an Intel Core\textsuperscript{\textregistered} i7-2600 running the 2.6.38-8 Linux
+kernel.  This Nehalem chip has eight cores running at 3.40GHz and a cache size
+of 8192 KB.  We studied three example programs---a simple microkernel(\col) and two complex programs: one from the
+SPEC 2006CPU benchmark suite(\gcc), and one from LAPACK(\svd), aggregating and measuring the processor load---instructions per cycle (ipc)---at 100,000-instruction intervals.
+To record these measurements,
+we used the {\tt libpfm4}, via PAPI
+\footnote{Performance Application Programming Interface}\cite{papi-website}, which we have instrumented to stop program execution every 100,00-instructions and read the contents of the CPU's onboard hardware performance monitors (HPMs).
+HPMs are specialty registers built for this purpose.
+Obviously using a system to measure itself can cause interference but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way.
+For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see
+\cite{zach-IDA10,mytkowicz09,todd-phd}.
 
 \subsection{The Programs}
-Include traces of full \col \svd and \gcc
-\begin{figure}[htbp]
-  \centering
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/colFullTS}
-    \caption{\col Time Series}
-    \label{fig:col-ts}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
-    \caption{\gcc Time Series}
-    \label{fig:gcc-ts}
-  \end{subfigure}
-    \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/SVD1RegimesColored}
-    \caption{\svd Time Series with Regime Coloring}
-    \label{fig:svd-ts-colored}
-  \end{subfigure}
-  \caption{[[Maybe this should be just a picture of each time series and then have another figure with color after explaining heuristically the reigmes.
-In (a) the instructions executed per CPU clock cycle
-    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
-    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
-    \end{figure}
+We study three example programs---a simple microkernel(\col) and two complex programs: one from the
+SPEC 2006CPU benchmark suite(\gcc), and one from LAPACK(\svd).
+\subsubsection{\col}
+\col is a simple three-line C program that repeatedly initializes the top half of matrix in column-major order.
+While this is a very simple program it has been shown to exhibit very complicated behavior.
+In fact, in \cite{chaos} it was shown
+\subsubsection{\gcc}
 
+\subsubsection{\svd}
 \svd brings up a very interesting point: systems change over time.
@@
 The regimes are labeled from left to right.
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[t]{\textwidth}
+  \centering
+    \includegraphics[width=0.75\textwidth]{figs/colFullTS}
+    \caption{\col Time Series}
+    \label{fig:col-ts}
+  \end{subfigure}%
+  \\
+
+  \begin{subfigure}[t]{\textwidth}
+  \centering
+    \includegraphics[width=0.75\textwidth]{figs/gccfullts}
+    \caption{\gcc Time Series}
+    \label{fig:gcc-ts}
+  \end{subfigure}
+    \begin{subfigure}[t]{\textwidth}
+    \centering
+    \includegraphics[width=0.75\textwidth]{figs/SVD1RegimesColored}
+    \caption{\svd Time Series with Regime Coloring}
+    \label{fig:svd-ts-colored}
+  \end{subfigure}
+  \caption{[[Maybe this should be just a picture of each time series and then have another figure with color after explaining heuristically the reigmes.
+In (a) the instructions executed per CPU clock cycle
+    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
+    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
+    \end{figure}

### 2014-03-02 model.tex
adding experimental methods/collection section and updating

@@
 \item Description of DCE and parameter estimation  
-\item Description of auto ARIMA 
-\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
+\item\cmark Description of auto ARIMA 
 \item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
@@
 
-While the simplicity of these methods is advantageous when a signal is mostly complexity, i.e., that is the time series exists on the high side of the complexity spectrum, the methods can be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series.
+While the simplicity of these methods is advantageous on average when a time series exists on the high side of the complexity spectrum, their simplicity can also be a disadvantage and cause the methods to be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series.
 Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths.
@@
 The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step.
-At every time step Random Walk would preciesly predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$.
+At every time step Random Walk would precisely predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$.
 In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue and error of 0.5 at every step.
@@
 
-For this \cite{davislinearts}
+%For this \cite{davislinearts}
 
@@
    \centering
+\begin{subfigure}{0.49\textwidth}
     \includegraphics[width=\textwidth]{figs/colipc3d.png}
+    \caption{\col }
+    \label{fig:colEmbedding}
+  \end{subfigure}%   
+  \begin{subfigure}{0.49\textwidth}
+    \includegraphics[width=\textwidth]{figs/colipc3d.png}
+    \caption{\gcc}
+    \label{fig:gccEmbedding}
+  \end{subfigure}%
+
+     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
      \caption{[[{\color{red}Joshua:Perhaps add a similar graphic of \gcc to show contrast}{\color{green}agreed I will work on this}]]A 3D projection of a delay-coordinate embedding of the trace

### 2014-03-02 methods.tex
program descriptions

@@
 To record these measurements,
-we used the {\tt libpfm4}, via PAPI
+we used the {\tt libpfm4}, via PAPI 5.2
 \footnote{Performance Application Programming Interface}\cite{papi-website}, which we have instrumented to stop program execution every 100,00-instructions and read the contents of the CPU's onboard hardware performance monitors (HPMs).
@@
 \subsubsection{\col}
-\col is a simple three-line C program that repeatedly initializes the top half of matrix in column-major order.
-While this is a very simple program it has been shown to exhibit very complicated behavior.
-In fact, in \cite{chaos} it was shown
+\col is a simple three-line C program that repeatedly initializes the upper triangle of a  2048 X 2048 matrix in column-major order.
+While this is a very simple program it has been shown to produce time series with very complicated behavior.
+In fact, in \cite{mytkowicz09} it was shown that time series generated from this simple program can exhibit everything from periodic to deterministic chaos.
+A time series of the processor load of this program is shown in Figure \ref{fig:gcc-ts}.
 \subsubsection{\gcc}
 
+\gcc is a benchmark program which is part of the SPEC CPU2006 benchmark suite \cite{spec}.
+This program is written by Richard Stallman and is based on Version 3.2 of gcc.
+This benchmark essentially runs as a compiler with many of its optimization flags enabled, compiling a series of input files and generating x86-64 assembly code files intended for execution on AMD Opteron processors.
+This program is more typical of programs that are used every day, as opposed to the simple micro-kernel, \col.
+A time series of the processor load of this program is shown in Figure \ref{fig:gcc-ts}.
+
 \subsubsection{\svd}
+
+\svd is a Fortran program which calcualtes teh singular value decomposition (SVD) of a rectangular $M$ by $N$  real-valued matrix which need not contain specail structure such as symmetric, positive definite etc.
+We give this program as input a 750 x 1000 matrix of randomly generated entries\footnote{It should be noted that multiple randomly generated matrices were investigated but no measurable effect was present in the resulting time series.}.
+A time series of the processor load of this program is shown in Figure \ref{fig:svd-ts-colored}.
+
 \svd brings up a very interesting point: systems change over time.

### 2014-03-03 methods.tex
adding citations and finishing methods

--- 2014-03-02 methods.tex
@@
 
+ \begin{enumerate}
+ \item \cmark  Experimental methods: (how we collect the time series and what the times series areThis should be HPM PAPI, which programs we model
+\item \cmark program descriptions and relevant citations
+\subitem \cmark description of \col
+\subitem \cmark description of \gcc
+\subitem \cmark description of \svd 
+\subitem \cmark description of svd regimes
+ \end{enumerate}
+
 \subsection{Time Series Collection}
- Experimental methods (how we collect the time series and what the times series are
-This should be HPM PAPI, which programs we model
+
+The time-series data for these experiments was collected on an Intel Core\textsuperscript{\textregistered} i7-2600 running the 2.6.38-8 Linux
+kernel.  This Nehalem chip has eight cores running at 3.40GHz and a cache size
+of 8192 KB.  We collected performance traces of processor efficiency (IPC) during the execution of three exemplar programs---a simple microkernel(\col) and two complex programs: one from the
+SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd).
+In particular, we measured and aggregated the instructions executed per clock cycle (ipc)---a measure of processor efficiency---at 100,000-instruction intervals.
+To record these measurements,
+we used the {\tt libpfm4} library, via PAPI 5.2
+\footnote{Performance Application Programming Interface}\cite{papi}, combined with a custom measurement infrastructure\cite{todd-phd} to stop program execution every 100,00-instructions and read the contents of the CPU's onboard hardware performance monitors (HPMs).
+HPMs are specialty registers who exist solely to store performance data.
+For statistical validation we collected 15 performance traces from each program resulting in 45 total time series.
+
+Obviously using a system to measure itself can cause interference--i.e, are you measuring the generating process cleanly or are you adding noise to the process---but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way.
+For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see
+\cite{zach-IDA10,mytkowicz09,todd-phd}.
 
 \subsection{The Programs}
-Include traces of full \col \svd and \gcc
+We collect performance traces of processor efficiency (IPC) of three exemplar programs---a simple microkernel(\col) and two complex programs: one from the
+SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd).
+In this section we provide a quick explanation of each program as well as an example performance trace (Seen in Figures \ref{fig:col-ts}-\ref{fig:svd-ts-colored}.
+\subsubsection{\col}
+\col is a simple three-line C program that repeatedly initializes the upper triangle of a  2048 X 2048 matrix in column-major order.
+While this is a very simple program it has been shown to produce time series with very complicated behavior.
+In fact, in \cite{mytkowicz09} it was shown that time series generated from this simple program can exhibit everything from periodicity to deterministic chaos.
+A time series of the processor efficiency (IPC) of this program is shown in Figure \ref{fig:col-ts}.
+
 \begin{figure}[htbp]
   \centering
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/colFullTS}
-    \caption{\col Time Series}
+    \includegraphics[width=0.75\textwidth]{figs/colFullTS}
+    \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \col. Each point is the average IPC during a 100,000 instruction period.}
     \label{fig:col-ts}
-  \end{subfigure}%
-  \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
-    \caption{\gcc Time Series}
-    \label{fig:gcc-ts}
-  \end{subfigure}
-    \begin{subfigure}[t]{0.475\textwidth}
-    \includegraphics[width=0.95\textwidth]{figs/SVD1RegimesColored}
-    \caption{\svd Time Series with Regime Coloring}
-    \label{fig:svd-ts-colored}
-  \end{subfigure}
-  \caption{[[Maybe this should be just a picture of each time series and then have another figure with color after explaining heuristically the reigmes.
-In (a) the instructions executed per CPU clock cycle
-    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
-    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
+
   \end{figure}
 
-\svd brings up a very interesting point: systems change over time.
-While it is clear in Figure \ref{fig:sample-ts} (a) and (b) that a single system is being observed, it also appears \emph{visually} that the complexity of these two time series is consistent over time, i.e., while they are both complicated they don't appear to increase or decrease in complexity over the range of the program.
+\subsubsection{\gcc}
+
+\gcc is a benchmark program included in the SPEC CPU2006 benchmark suite \cite{spec}.
+This program is written by Richard Stallman and is based on Version 3.2 of {\tt gcc}.
+This benchmark essentially runs as a compiler with many of its optimization flags enabled, compiling a series of input files and generating x86-64 assembly code files intended for execution on AMD Opteron processors\cite{spec}.
+This program is more typical of programs that are used every day, as opposed to the simple micro-kernel, \col.
+A time series of the processor efficiency of this program running on the Intel i7-2600 is shown in Figure \ref{fig:gcc-ts}.
+
+  \begin{figure}[t]
+  \centering
+    \includegraphics[width=0.75\textwidth]{figs/gccfullts}
+    \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \gcc. Each point is the average IPC during a 100,000 instruction period.}
+    \label{fig:gcc-ts}
+  \end{figure}
+
+\subsubsection{\svd}
+
+\svd is a Fortran program from the LAPACK suite \cite{lapack} which calculates the singular value decomposition (SVD) of a rectangular $M$ by $N$  real-valued matrix which need not contain special structure such as symmetric, positive definite etc.
+To execute this program we input a 750 x 1000 matrix of randomly generated entries\footnote{Multiple randomly generated matrices were investigated but no measurable effect was present in the resulting time series.}.
+A ``regime"-colored time series of the IPC of this program is shown in Figure \ref{fig:svd-ts-colored}.
+We now describe why this is done.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.75\textwidth]{figs/SVD1RegimesColored}
+    \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \svd.
+Each point is the average IPC during a 100,000 instruction period.
+The colors correspond to the six different regimes detected visually.}
+    \label{fig:svd-ts-colored}
+  \end{figure}
+
+\svd brings up a very interesting point: as we observe a system the generating process can change in complexity over time.
+While it is clear in Figures \ref{fig:col-ts} and \ref{fig:gcc-ts} that a single system is being observed, it also appears \emph{visually} that the complexity of these two time series is consistent over time, i.e., while they are both complicated they don't appear to increase or decrease in complexity over the range of the program.
 \svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually.
-Figure \ref{fig:svd-ts-colored} is a single trace of \svd from start to finish, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series is visually very different from section to section.
-This is caused by the code of \svd moving between different subroutines.
-We call these changes in \svd dynamics  \emph{\svd regimes} and we have colored each of the six regimes different colors in Figure \ref{fig:svd-ts-colored}.
+Like Figures \ref{fig:col-ts} and \ref{fig:gcc-ts}, Figure \ref{fig:svd-ts-colored} is a single time series of the IPC of \svd, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series  visually switches between ``regimes".
+
+This is caused by the code of \svd moving between different subroutines as the SVD algorithm evolves, something not present in either \gcc or \col.
+We call these changes in \svd dynamics \emph{\svd regimes} and we have colored each of the six regimes differently in Figure \ref{fig:svd-ts-colored}.
 The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift.
 In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}.
-The purpose of this paper is not to rigorously explore regime detection but to explore quantifying complexity of a time series.
-As such, making this split simply serves as providing 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore.
-For notational convience we refer to these signals as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd.
+Note, the purpose of this paper is \emph{not} to rigorously explore regime detection but to explore quantifying complexity of a time series.
+As such, making these regime splits serves to provide an additional 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore.
+For notational convenience we refer to these time series as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd.
 The regimes are labeled from left to right.
+
+%\begin{figure}[htbp]
+%  \centering
+%  \begin{subfigure}[t]{\textwidth}
+%  \centering
+%    \includegraphics[width=0.75\textwidth]{figs/%colFullTS}
+%    \caption{\col Time Series}
+%    \label{fig:col-ts}
+%  \end{subfigure}%
+%  \\
+
+%  \begin{subfigure}[t]{\textwidth}
+%  \centering
+%    \includegraphics[width=0.75\textwidth]{figs/%gccfullts}
+%    \caption{\gcc Time Series}
+%    \label{fig:gcc-ts}
+%  \end{subfigure}
+%    \begin{subfigure}[t]{\textwidth}
+%    \centering
+%    \includegraphics[width=0.75\textwidth]{figs/%SVD1RegimesColored}
+%    \caption{\svd Time Series with Regime Coloring}
+%    \label{fig:svd-ts-colored}
+%  \end{subfigure}
+%  \caption{[[Maybe this should be just a picture of %each time series and then have another figure with %color after explaining heuristically the reigmes.
+In %(a) the instructions executed per CPU clock cycle%
+%    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
+%    instruction period. Similarly in (b) is a time %series of the IPC during the execution of \gcc.}%\label{fig:sample-ts}%
+%    \end{figure}

### 2014-03-03 intro.tex
updating intro

--- 2014-02-27 intro.tex
@@
 
-%Notes on wording
+%****NOTES START HERE****
 
-% If we define entropy as complexity in the paper (as we do currently) we could say that every signal complicated or not has a balance between temporal/predictive structure and complexity (entropy).
-So for example, a chaotic signal and a random walk signal are both "complicated" they exhibit interesting behavior (they aren't monotonic, periodic, constant etc.) but one of them is far more complex and less predictable than the other , e.g.,  a random walk or gcc for example is complicated *and* highly complex (very high entropy) where as col is also complicated but not as complex as it exhibits a great deal of predictive structure (low entropy).
+%BRAINSTORMING 
+%\begin{itemize}
+%\item Complexity need not be hard to predict (can %point at the simple predictions paper) [[move to %introduction]]
+%\item random walk for example is best predicted by %guess what just happend[[move to introduction]]
+%\item The kind of complexity present matters, i.e., %that is whether the complexity is structured or not.%[[use here and mention in intro]]
 
-%An interesting contribution here is that examining a complicated signal (i.e., something that has interesting behavior, not periodic, monotonic, constant etc) of noisy real-valued entries, it is hard to determine a priori (without WPE) where in the spectrum of temporal structure and complexity the signal lies.
-Is it completely complex and have little to no predictive structure e.g., gcc or is it balanced in complexity and structure like a chaotic signal e.g., col.
-
-%One potential issue here is that apparently information %theorists say that completely random signals are void of complexity.
-So Saying that a random walk signal (i.e., a signal with high entropy) is complex may cause some issues.
-HOWEVER, if we define complexity as entropy I don't think this is an issue.
-
-%Ryan does this seem reasonable?
-Is this an accurate synopsis of our discussion.
-Please feel free to add to it.
-Liz do you think making this distinction between "Complicated" and "Complex" would be useful for the readers understanding and dissemination of this topic?
-
-%This would suggest changing the title to something a long the lines of "Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability "
-
-BRAINSTORMING 
-\begin{itemize}
-\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
-\item random walk for example is best predicted by guess what just happend[[move to introduction]]
-\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]]
-
-\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method.
+%\item \col brings about the point nicely that some %prediction strategies cannot utilize the processes internal information transfer method.
 That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy.
@@
 %\end{it}
-Things to add to introduction
-\begin{enumerate}
-\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
-\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.
+%Things to add to introduction
+%\begin{enumerate}
 
-\item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc
+%\item \cmark Make an argument that Computer Performance %is a great testing ground as it gives signals that completely cover the spectrum of complexity \col \dots \gcc
 
-\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
-\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
-\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex.
+%\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
+%\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
+%\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex.
 Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not.
@@
 
-\item Introduce the two main contributions of the paper which are outlined at the begining of the results section
+%\item\cmark Introduce the two main contributions of the paper which are outlined at the begining of the results section
 
-\end{enumerate}
-\noindent ***NOTES END HERE****
-
-\noindent ****SECTION START****
+%\end{enumerate}
+%\noindent ***NOTES END HERE****
+%
+%\noindent ****SECTION START****
 
@@
 With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.
-In the middle of this spectrum are where complicated \emph{and} forecast-able signals exist, e.g., deterministic chaotic trajectories, high period signals with some level of noise on top.
+In the middle of this spectrum are where complicated \emph{and} interesting---from a modeling perspective---signals exist, e.g., observations of a deterministic chaotic trajectory, high period signals with some level of noise on top, etc.
 With time series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model the future of the observed system could be forecast with high accuracy.
@@
 This naturally leads to an important and hard question: given a complicated real-valued time series, is it possible to reliably quantify where on this spectrum the time series lies?
-That is, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) how can we infer from a noisy real-valued time series if the signal is too \emph{complex} to forecast?
+i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast using a prediction strategy based on predictive structure (temporal forward information transfer)?
+We define this spectrum as well as what is meant by low, medium and high rigorously in Section \ref{sec:meaComplex}.
 
@@
 We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
-We argue that \emph{permutation entropy}
-\cite{bandt2002per}, a method for measuring the entropy of a
-real-valued-finite-length noisy time series through ordinal analysis, is an effective way to explore that conjecture.
-For the purposes of this paper we define complexity as \emph{permutation entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}.
-To validate these claims we model the time series using a variety of forecasting models which we introduce in Section \ref{sec:model}.
-We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method.
-Which results in two primary findings:
-
-\begin{enumerate}
-
-\item The complexity of a noisy real-valued time series is quantifiable.
-
-\item The way information is processed internally by a given process plays a crucial role in the success of different forecasting schema.
-
-\end{enumerate}
+We argue that an extension of \emph{permutation entropy}
+\cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture.
+In light of this, for the purposes of this paper we define \emph{complexity} as an approximation of Shannon \emph{entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}.
+In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity.
 
 %[[Introduce and make it clear why we choose comp. perf. as test bed.]]
-[[Move this up]]
+
 To explore this conjecture we require a test bed for generating a broad array of time series to analyze which covers the full spectrum of complexity.
-We argue that the ideal test bed for this purpose is computer performance.
+We argue that the ideal test bed for this purpose are computer performances, e.g., the processor efficiency during the execution of a particular program.
 Computers are among the most complex engineered artifacts in current
@@
 that go into the design of these chips---make the patterns of their
-processor loads highly complicated. Accurate forecasts of these quantities, if one could
-construct them, could be used to improve computer design. 
+processor efficiency highly complicated. 
+%Accurate forecasts of these quantities, if one could
+%construct them, could be used to improve computer design. 
 % If one
@@
 %speed is scaled back).
- Computer performance traces are, however, very
-complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel,'' like a three-line loop that
-repeatedly initializes a matrix in column-major order, can produce everything from periodic to
+
+As a result computer performance traces are very
+complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel," like a three-line loop that
+repeatedly initializes the upper triangle of a matrix in column-major order, can produce everything from periodic to
 {\sl chaotic} performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
@@
     % and a .pdf suffix will be assumed for pdflatex
-    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
-        \col}, a three-line C program that repeatedly initializes
-      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
-      simple program exhibits chaotic performance dynamics.}
+    \caption{A small snippet of the instructions per cycle(ipc) of \col, a three-line C program that repeatedly initializes
+      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.}
    \label{fig:ipc}
   \end{figure}
-With a single system producing complicated time series that vary so greatly in complexity, it would be invaluable to known a priori if a trace contained enough predictive structure to forecast before time and effort was spent on trying to model and forecast it.
+%
+More interesting programs, e.g., compilers and numerical solvers, produce time series which exist throughout the spectrum of complexity: from completely predictable to completely unstructured.
+In addition, some performance traces even exhibit interesting regime changes as a program switches between varying subroutines.
+With a single system (computer performance) capable of producing complicated time series that vary so greatly in complexity (periodic to noise), it would be invaluable to known a priori if a trace contained enough predictive structure to build a  forecast model based on structure before time and effort was spent on trying to model and forecast it.
 
-The computer systems community has applied a variety of prediction
-strategies to traces like this, most of which employ regression.  An
-appealing alternative builds on the recently established fact that
-computers can be effectively modeled as deterministic nonlinear
-dynamical systems \cite{mytkowicz09}.  This result implies the
-existence of a deterministic forecast rule for those dynamics.  In
-particular, one can use \emph{delay-coordinate embedding} to
-reconstruct the underlying dynamics of computer performance, then use
-the resulting model to forecast the future values of computer
-performance metrics such as memory or processor loads
-\cite{josh-ida2011}.  In the case of simple microkernels like the one
-that produced the trace in Figure~\ref{fig:ipc}, this deterministic
-modeling and forecast strategy works very well.  In more-complicated
-programs, however, such as numerical software or compilers,
-this forecast strategy---as well as the traditional methods---break
-down some of the time, but work fine others.
+%The computer systems community has applied a variety of prediction
+%strategies to traces like this, most of which employ regression.  An
+%appealing alternative builds on the recently established fact that
+%computers can be effectively modeled as deterministic nonlinear
+%dynamical systems \cite{mytkowicz09}.  This result %implies the
+%existence of a deterministic forecast rule for those %dynamics.  In
+%particular, one can use \emph{delay-coordinate %embedding} to
+%reconstruct the underlying dynamics of computer %performance, then use
+%the resulting model to forecast the future values of computer
+%performance metrics such as memory or processor loads
+%\cite{josh-ida2011}.  In the case of simple microkernels like the one
+%that produced the trace in Figure~\ref{fig:ipc}, this deterministic
+%modeling and forecast strategy works very well.  In more-complicated
+%programs, however, such as numerical software or compilers,
+%this forecast strategy---as well as the traditional methods---break
+%down some of the time, but work fine others.
 
-This paper is a first step in understanding when, why, and how
-deterministic forecast strategies fail when they are applied to
-deterministic systems.  We focus here on the specific example of
+This paper provides insight into when, why, and how forecast strategies fail when they are applied to complicated time series by empirically estimating their inherent complexity.
+We focus here on the specific example of
 computer performance but believe the results apply to a much broader class of time series. We conjecture that the complexity of traces
 from these systems---which results from the inherent dimension,
-non-linearity, and non-stationarity of the dynamics, as well as from
+non-linearity, and non-stationarity, as well as from
 measurement issues like noise, aggregation, and finite data
-length---can make those deterministic signals \emph{effectively}
-unpredictable.  We argue that \emph{permutation entropy}
-\cite{bandt2002per}, a method for measuring the entropy of a
-real-valued-finite-length time series through ordinal analysis, is an
-effective way to explore that conjecture.  We study three
-examples---a simple microkernel and two complex programs: one from the
-SPEC 2006CPU benchmark suite, and one from LAPACK---running on an Intel i7-based machine.  For
-each program, we calculate the permutation entropy of the processor
-load (instructions per cycle), then compare that to the prediction accuracy attainable for
-that trace using a series of deterministic models.
+length---is \emph{empirically quantifiable}.
+
+To validate this conjecture we model 120 different performance traces using a variety of forecasting models which we introduce in Section \ref{sec:model}.
+We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method.
+Which results in two primary findings:
+
+\begin{enumerate}
+
+\item The complexity of a noisy real-valued time series is quantifiable and correlated with prediction accuracy of an ideal predictor.
+
+\item The way information is processed internally by a given process plays a crucial role in the success of different forecasting schema.
+
+\end{enumerate}
+
+%We argue that \emph{permutation entropy}
+%\cite{bandt2002per}, a method for measuring the %entropy of a
+%real-valued-finite-length time series through ordinal %analysis, is an
+%effective way to explore that conjecture.  %{\color{red} Probably remove this and move a%s it is in the experimental methods or point to that section for more details}We study three
+%examples---a simple microkernel and two complex %programs: one from the
+%SPEC 2006CPU benchmark suite, and one from LAPACK---%running on an Intel i7-based machine.  For
+%each program, we calculate the permutation entropy of %the processor
+%load (instructions per cycle), then compare that to %the prediction accuracy attainable for
+%that trace using a series of deterministic models.
 
@@
 
- The rest of the paper is organized as follows.
- Section~\ref{sec:compModel} describes the experimental setup, as well
- as the nonlinear modeling and forecast strategies.  In
- Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
- its value for a number of different computer performance traces, and
- compare the results to the prediction accuracy.  In
+ The rest of the paper is organized as follows: Section~\ref{related} discusses similar methods and how this method is different.
+Section~\ref{sec:methods} explains the experimental setup and explains in depth the programs observed to generate each time series.
+Section~\ref{sec:Model} describes the four forecast models we implemented.
+In
+ Section~\ref{sec:meaComplex}, we review weighted permutation entropy. In Section~\ref{sec:results} we estimate the complexity of each time series, and
+ compare the complexity to the prediction accuracy of several different forecast schema.  In
  Section~\ref{sec:conc}, we discuss these results and their

### 2014-03-04 intro.tex
edits to intro,related and methods

@@
-\section{Introduction}\label{sec:intro}
+\section{Introduction {\color{blue} EDITABLE}}\label{sec:intro}
 
@@
 Complicated time series are ubiquitous in modern scientific research. 
-%Characterizing and forecasting the future behavior of these observed systems is a vastly researched area with many open questions. 
-Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}\footnote{An approx of (blah ) entropy which we make more rigorous in blah}.
+Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}.
 On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model.
@@
 Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series.
-This naturally leads to an important and hard question: given a complicated real-valued time series, is it possible to reliably quantify where on this spectrum the time series lies?
+This naturally leads to an important and hard question: given a complicated real-valued time series, does their exist a forecast model for that time series that can utilize the information being transmitted from past to future (if such information exists)?
+A first step in answering this question is to answer: is it possible to reliably quantify where on the spectrum of complexity the time series lies?
 i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast using a prediction strategy based on predictive structure (temporal forward information transfer)?
-We define this spectrum as well as what is meant by low, medium and high rigorously in Section \ref{sec:meaComplex}.
+We rigorously define this \emph{complexity spectrum} as well as what is meant by \emph{low, medium and high} in Section \ref{sec:meaComplex}.
 
@@
 In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity.
+{\color{blue} Ryan: Insert your sentence here, Something like We are not the first people to attempt to quantify complexity of time series with entropy(citations) However, these methods rely on binning which is bad because of biasing (citations) which we explain further in Section \ref{sec:related}.}
 
@@
 
-To explore this conjecture we require a test bed for generating a broad array of time series to analyze which covers the full spectrum of complexity.
-We argue that the ideal test bed for this purpose are computer performances, e.g., the processor efficiency during the execution of a particular program.
+To explore this conjecture we require a test bed for generating a broad array of time series which cover the full spectrum of complexity.
+We argue that the ideal test bed for this purpose are computer performance traces, e.g., the processor efficiency during the execution of a particular program.
 Computers are among the most complex engineered artifacts in current
@@
 repeatedly initializes the upper triangle of a matrix in column-major order, can produce everything from periodic to
-{\sl chaotic} performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
+chaotic performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
 Figure~\ref{fig:ipc}.%, and chaos places fundamental limits on predictability.
@@
 %
-More interesting programs, e.g., compilers and numerical solvers, produce time series which exist throughout the spectrum of complexity: from completely predictable to completely unstructured.
+More interesting programs, e.g., compilers and numerical solvers, produce time series which exist throughout the spectrum of complexity: from completely predictable to completely unstructured,i.e., low to high complexity[[Just trying to reinforce the wording in the readers mind]].
 In addition, some performance traces even exhibit interesting regime changes as a program switches between varying subroutines.
-With a single system (computer performance) capable of producing complicated time series that vary so greatly in complexity (periodic to noise), it would be invaluable to known a priori if a trace contained enough predictive structure to build a  forecast model based on structure before time and effort was spent on trying to model and forecast it.
+With a single system (computer performance) capable of producing complicated time series that vary so greatly in complexity (periodic to noise), it would be invaluable to known a priori if a trace contained enough predictive structure to build a  forecast model based on structure before time and effort was spent searching for an ideal model.
 
@@
 This paper provides insight into when, why, and how forecast strategies fail when they are applied to complicated time series by empirically estimating their inherent complexity.
-We focus here on the specific example of
+[[Liz, do you think here would be better for the sentence about we are not the first, or is above where I asked Ryan to insert it ok?
+I worry it would muddy the water of this synopsis paragraph.]] We focus here on the specific example of
 computer performance but believe the results apply to a much broader class of time series. We conjecture that the complexity of traces
@@
 length---is \emph{empirically quantifiable}.
-
 To validate this conjecture we model 120 different performance traces using a variety of forecasting models which we introduce in Section \ref{sec:model}.
@@
 are attempting is to predict aspects or functions of the running of
-the CPU: instructions executed per second, cache misses per 100,000
-instructions, and similar statistics. Prediction of these quantities
+the CPU, e.g., processor efficiency and similar statistics. Prediction of these quantities
 at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem.
+not violate the Rice-Shapiro{\color{blue}[[Ryan: Do you have a citation for this theorem, also should we give a brief sentence explaining what this theorem is?]]} theorem.
 
- The rest of the paper is organized as follows: Section~\ref{related} discusses similar methods and how this method is different.
+ The rest of the paper is organized as follows: Section~\ref{sec:related} discusses similar methods and how this method is different.
 Section~\ref{sec:methods} explains the experimental setup and explains in depth the programs observed to generate each time series.
-Section~\ref{sec:Model} describes the four forecast models we implemented.
+Section~\ref{sec:model} describes the four forecast models we implemented.
 In

### 2014-03-04 related.tex
edits to intro,related and methods

@@
-\section{Related Work}
+\section{Related Work}\label{sec:related}
 
@@
 works **if** you have generating parition but not if you don't
+{\color{blue} Ryan: Please insert here the paragraph and citations defaming binning / partitioning and illustrating the difficulty of generating partitions / biasing of this method without generating partitions, along with relevant citations.
+}
 

### 2014-03-04 methods.tex
edits to intro,related and methods
editing

@@
-\section{Experimental Methods and Time Series Explanations}
+\section{Experimental Methods and Time Series Explanations}\label{sec:methods}
 
- \begin{enumerate}
- \item \cmark  Experimental methods: (how we collect the time series and what the times series areThis should be HPM PAPI, which programs we model
-\item \cmark program descriptions and relevant citations
-\subitem \cmark description of \col
-\subitem \cmark description of \gcc
-\subitem \cmark description of \svd 
-\subitem \cmark description of svd regimes
- \end{enumerate}
+% \begin{enumerate}
+% \item \cmark  Experimental methods: (how we collect the %time series and what the times series areThis should be HPM %PAPI, which programs we model
+%\item \cmark program descriptions and relevant citations
+%\subitem \cmark description of \col
+%\subitem \cmark description of \gcc
+%\subitem \cmark description of \svd 
+%\subitem \cmark description of svd regimes
+% \end{enumerate}
 
-\subsection{Time Series Collection}
+\subsection{Time Series Collection {\color{blue} EDITABLE}}
 
@@
 
-Obviously using a system to measure itself can cause interference--i.e, are you measuring the generating process cleanly or are you adding noise to the process---but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way.
+Obviously using a system to measure itself can cause interference--i.e, are you measuring the generating process cleanly or are you adding noise to the process?---but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way.
 For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see
@@
 
-\subsection{The Programs}
-We collect performance traces of processor efficiency (IPC) of three exemplar programs---a simple microkernel(\col) and two complex programs: one from the
+\subsection{The Programs: [[Maybe The generating processes]]{\color{blue} EDITABLE}}
+We collect performance traces of processor efficiency (IPC) of three exemplar programs, [[If we change the title of the section, add this: which act as our generating processes]]---a simple microkernel(\col) and two complex programs: one from the
 SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd).
 In this section we provide a quick explanation of each program as well as an example performance trace (Seen in Figures \ref{fig:col-ts}-\ref{fig:svd-ts-colored}.
-\subsubsection{\col}
+\subsubsection{\col {\color{blue} EDITABLE}}
 \col is a simple three-line C program that repeatedly initializes the upper triangle of a  2048 X 2048 matrix in column-major order.
@@
 
-\subsubsection{\gcc}
+\subsubsection{\gcc {\color{blue} EDITABLE}}
 
@@
 This benchmark essentially runs as a compiler with many of its optimization flags enabled, compiling a series of input files and generating x86-64 assembly code files intended for execution on AMD Opteron processors\cite{spec}.
-This program is more typical of programs that are used every day, as opposed to the simple micro-kernel, \col.
+This program is more typical of programs that are used every day, as opposed to a simple micro-kernel,e.g., \col.
 A time series of the processor efficiency of this program running on the Intel i7-2600 is shown in Figure \ref{fig:gcc-ts}.
@@
 
-\subsubsection{\svd}
+\subsubsection{\svd {\color{blue} EDITABLE}}
 
@@
 To execute this program we input a 750 x 1000 matrix of randomly generated entries\footnote{Multiple randomly generated matrices were investigated but no measurable effect was present in the resulting time series.}.
-A ``regime"-colored time series of the IPC of this program is shown in Figure \ref{fig:svd-ts-colored}.
+A ``regime-colored" time series of the IPC of this program is shown in Figure \ref{fig:svd-ts-colored}.
 We now describe why this is done.
@@
 As such, making these regime splits serves to provide an additional 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore.
-For notational convenience we refer to these time series as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd.
-The regimes are labeled from left to right.
-
-%\begin{figure}[htbp]
-%  \centering
-%  \begin{subfigure}[t]{\textwidth}
-%  \centering
-%    \includegraphics[width=0.75\textwidth]{figs/%colFullTS}
-%    \caption{\col Time Series}
-%    \label{fig:col-ts}
-%  \end{subfigure}%
-%  \\
-
-%  \begin{subfigure}[t]{\textwidth}
-%  \centering
-%    \includegraphics[width=0.75\textwidth]{figs/%gccfullts}
-%    \caption{\gcc Time Series}
-%    \label{fig:gcc-ts}
-%  \end{subfigure}
-%    \begin{subfigure}[t]{\textwidth}
-%    \centering
-%    \includegraphics[width=0.75\textwidth]{figs/%SVD1RegimesColored}
-%    \caption{\svd Time Series with Regime Coloring}
-%    \label{fig:svd-ts-colored}
-%  \end{subfigure}
-%  \caption{[[Maybe this should be just a picture of %each time series and then have another figure with %color after explaining heuristically the reigmes.
-In %(a) the instructions executed per CPU clock cycle%
-%    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
-%    instruction period. Similarly in (b) is a time %series of the IPC during the execution of \gcc.}%\label{fig:sample-ts}%
-%    \end{figure}
+For notational convenience we refer to these time series as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd, ordered from left to right.

### 2014-03-04 model.tex
editing

@@
-\section{Modeling}\label{sec:model}
+\section{Modeling {\color{blue} EDITABLE}}\label{sec:model}
 
@@
 
-\subsection{Primitive Strategies to combat complexity: Random Walk and \naive}\label{sec:simple}
+\subsection{Primitive Strategies to combat complexity: Random Walk and \naive {\color{blue} EDITABLE}}\label{sec:simple}
 In this section we cover two very simple forecast models: Random Walk and \naive.
-With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement: $$p_i = X_{i-1,obs}$$.
-The \naive  prediction strategy is to forecast the rolling average up to that point.
-i.e., the prediction $p_i$ is simply the average of all observed points up to that point: $$p_i = \sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}$$.
+With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement: $$p_i = X_{i-1,obs}.$$ The \naive  prediction strategy is to forecast the rolling average up to that point.
+i.e., the prediction $p_i$ is simply the average of all observed points up to that point: $$p_i = \sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}.$$
 
@@
 In this circumstance, forecasting the mean will on average get you in the ball park of the forecast without taking into account the lack of structure.
-The \naive forecasting scheme can be thought of as a filtering mechanism which compresses a signal onto a single point, effectively filtering out both complexity and structure.
+The \naive~ forecasting scheme can be thought of as a filtering mechanism which compresses a signal onto a single point, effectively filtering out both complexity and structure.
 However, if very little predictive structure is present in a signal then this filtering mechanism effectively mitigates the complexity.
@@
 Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths.
-Fortunately, they somewhat parallel each other, i.e., the weakness of Random Walk and \naive are covered by the other primitive method.
+Fortunately, they somewhat compliment each other, i.e., the weakness of Random Walk and \naive~ are covered by the other primitive method.
+
 The situation in which Random Walk fails is when a time series changes drastically on a regular basis.
@@
 At every time step Random Walk would precisely predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$.
-In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue and error of 0.5 at every step.
+In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue an error of 0.5 at every step.
 So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.
@@
 The \naive forecast fails when you have a signal which shifts between two or more regimes on a regular basis and the regimes are separated by a significant amount.
-In this case the \naive forecast will predict between the multiple regimes at every time step eventually accruing massive error.
-To see how Random Walk can cover this weakness imagine a time series which switches between more than one regime but stays in each regime for several time steps, this behaviour can be seen in some portions of the \svd trace seen in Figure \ref{fig:ts}.
+In this case the \naive forecast will predict between the regimes at every time step eventually accruing massive error.
+To see how Random Walk can cover this weakness imagine a time series which switches between more than one regime but stays in each regime for several time steps, e.g., a regularly sampled noisy square wave, this behavior can be seen in some portions of the \svd trace seen in Figure \ref{fig:svd}.
 Random walk in this case could react to the regime shift within one time step, switch to the new Regime and begin forecasting nicely, where as \naive would just continue to predict between the modes of the multimodal distribution resulting in poor forecasting accuracy all of the time.
 
-So clearly these primitive methods have their downfalls, but it is important to keep in mind that when a time series exists on the top of the complexity spectrum they have little predictive structure to exploit and often a primitive method like Random Walk or \naive will consistently out perform methods which rely on predictive structure.
+So clearly these primitive methods have their downfalls, but it is important to keep in mind that when a time series exists on the top of the complexity spectrum they have little predictive structure to exploit and often a primitive method like Random Walk or \naive~ will consistently out perform methods which rely on predictive structure.
 This is explored further in Section \ref{sec:accuracy} and \ref{sec:results}
 
-\subsection{Linear Forecasting: Seasonal Autoregressive-integrated-moving average(ARIMA)}\label{sec:arima}
-\begin{enumerate}
-\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
-\item\cmark define rigorously, all the backshift operator garbage
+\subsection{Linear Forecasting: Seasonal Autoregressive-integrated-moving average(ARIMA){\color{blue} EDITABLE}}\label{sec:arima}
+%\begin{enumerate}
+%\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
+%\item\cmark define rigorously, all the backshift operator garbage
 
-$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = c + \Theta(B^m)\theta(B)\epsilon_i$$
+%$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = c + \Theta(B^m)\theta(B)\epsilon_i$$
 
-$$B^mX_i  = X_{i-m}$$
-\item\cmark discuss that this method needs linear structure to work correctly
-\item \cmark maybe talk about converging to zero and limits on prediction horizon
-\end{enumerate}
+%$$B^mX_i  = X_{i-m}$$
+%\item\cmark discuss that this method needs linear structure to work correctly
+%\item \cmark maybe talk about converging to zero and limits on prediction horizon
+%\end{enumerate}
 
@@
 When there is evidence a generating process exhibits predictive structure an obvious strategy would be to fit a hyperplane to the dataset and then use this hyperplane to forecast new data points.
-This is a very old technique stemming back to Yule in 1927\cite{weigend93} with his invention of autoregressive forecasting, which forecasts the next time step through a weighted average of the past $p$ time-steps, i.e., $$p_i = \phi(B)X_{i,obs}.$$ Here $\phi(B)$ is a polynomial of degree $p$ and $B$ is the backshift operator, i.e., $$B^mX_i  = X_{i-m,obs}.$$
-To improve this method further, especially in the presence of noisy data would be to allow for some wiggle room by adding a moving average of white noise to the model: $\theta(B)\epsilon_i$, where $\theta(B)$ is a polynomial of degree $q$ and $\{\epsilon_i\}\sim WN(0,\sigma^2)$.
+This is a very old technique stemming back to Yule in 1927\cite{weigend93} with his invention of autoregressive forecasting, which forecasts the next time step through a weighted average of the past $p$ time-steps, i.e., $$p_i = \phi(B)X_{i,obs}.$$ Here $\phi(z)$ is a polynomial of degree $p$ and $B$ is the backshift operator, i.e., $$B^mX_i  = X_{i-m,obs}.$$
+To improve this method further--especially in the presence of noisy data---one could allow for some wiggle room by adding a moving average of white noise to the model: $\theta(B)\epsilon_i$, where $\theta(z)$ is a polynomial of degree $q$ and $\{\epsilon_i\}\sim WN(0,\sigma^2)$.
 Finally if there is evidence of non-stationarity an initial differencing (integrated) is applied as an attempt to remove that non-stationarity: $(1-B^d)X_i$.
 Combining these three give us a nonseasonal  ARIMA$(p,d,q)$ model:  $$\phi(B)(1-B)^dX_i = \theta(B)\epsilon_i.$$ Where $p$, $d$ and $q$ correspond to the orders of the autoregressive, integrated and moving average models respectively.
-If periodic structure is present a seasonal ARIMA$(p,d,q)(P,D,Q)_m$ is often used and is defined by:$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = \Theta(B^m)\theta(B)\epsilon_i.$$ Here $\Phi$ and $\Theta$ are polynomials of degree P and Q respectively, and $m$ is the seasonal frequency.
+If periodic structure is present a seasonal ARIMA$(p,d,q)(P,D,Q)_m$ is often used and is defined by:$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = \Theta(B^m)\theta(B)\epsilon_i.$$ Here $\Phi(z)$ and $\Theta(z)$ are polynomials of degree P and Q respectively, and $m$ is the seasonal frequency.
 Forecasting is then done in two steps, first the integrated step is performed to remove non-stationarity.
-$$\hat{X_i} = (1-B^m)^D(1-B)^d X_{i,obs}$$.
-Then forecasts are made by evaluating $$\Phi(B^m)\phi(B)\hat{X_i} = \Theta(B^m)\theta(B)\epsilon_i.$$
+$$\hat{X_i} = (1-B^m)^D(1-B)^d X_{i,obs}.$$ Then forecasts are made by evaluating $$\Phi(B^m)\phi(B)\hat{X_i} = \Theta(B^m)\theta(B)\epsilon_i.$$
 
@@
 ARIMA forecasting is a time tested modelling process which uses a global linear fit and adjustments for seasonality, non-stationarity and noise.
-If a generating process fits into one of these categories then this model is a great short term predictor.
+If a generating process fits into this category then this model is a great short term predictor.
 If information is instead being transmitted in a nonlinear way this global linear fit falls apart.
-One other weakness of this method is that if long prediction horizons are used the forecast is guaranteed to converge to the mean after some number of predictions.
+One other weakness of this method is that if long prediction horizons are needed the forecast is guaranteed to converge to the mean after some number of predictions (dependent on model order).
 For this reason, we limit all methods to doing multiple 1-step predictions as to not handicap this method.
 
-\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA)}\label{sec:lma}
+\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{red} NOT EDITABLE}}\label{sec:lma}
 
@@
 
- \subsubsection{Reconstructing hidden dynamics}
+ \subsubsection{Reconstructing hidden dynamics {\color{red} NOT EDITABLE}}
 
@@
   .  That structure can be used to build a forecast model.
- \subsubsection{LMA: Using dynamics in forecasting}
+ \subsubsection{LMA: Using dynamics in forecasting {\color{red} NOT EDITABLE}}
 
@@
 
-\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE)}
+\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{red} NOT EDITABLE}}
 \label{sec:accuracy}

### 2014-03-04 results.tex
editing

@@
- \section{Predictability, Complexity, and Permutation Entropy} % (fold)
+ \section{Predictability, Complexity, and Permutation Entropy {\color{blue}  EDITABLE}} % (fold)
  \label{sec:results}
@@
   high, which makes guessing the mean a particularly bad strategy.
-  The same thing is true of the \svdthree signal.}.  Again, the
+  The same thing is true of the \svdthree signal.[[Joshua, it isn't the variance it is the strength in the bimodal distribution and only predicting in one mode of the distribution...]]}.  Again, the
 positions of these points on a WPE vs. MASE plot---significantly below

### 2014-03-04 wpe.tex
adding experimental methods/collection section and updating
editing

@@
-\section{Measuring Structural Complexity}\label{sec:meaComplex}
+\section{Measuring Structural Complexity {\color{red} NOT EDITABLE}}\label{sec:meaComplex}
 
@@
 [[talk about again here but justify in information theory section]]
-\item justify and explain the regime split in \svd 
+\item justify and explain the regime split in \svd {\color{red} I(joshua) moved this justification in the experimental methods section, it just needs to be explored here by applying windowed wpe and explaining that method}
 \end{enumerate}

### 2014-03-04 intro.tex
add sentence to intro

@@
 In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity.
-{\color{blue} Ryan: Insert your sentence here, Something like We are not the first people to attempt to quantify complexity of time series with entropy(citations) However, these methods rely on binning which is bad because of biasing (citations) which we explain further in Section \ref{sec:related}.}
+The permutation entropy is ideal in a situation like this as it converges on the true entropy value, unlike other partitioning techniques used in the past which either require specific knowledge of the generating process or produce biased values of the entropy~\cite{bollt2001}.
 

### 2014-03-04 intro.tex
clarify and add citation about rice's theorem

@@
 at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro{\color{blue}[[Ryan: Do you have a citation for this theorem, also should we give a brief sentence explaining what this theorem is?]]} theorem.
+not violate the Rice-Shapiro theorem~\cite{hopcroft} which essentially states that any non-trivial property of a program is uncomputable.
 

### 2014-03-05 model.tex
adding some figures and editing

@@
 
-\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{red} NOT EDITABLE}}\label{sec:lma}
+\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{blue} EDITABLE}}\label{sec:lma}
 
-When a generating process exhibits deterministic predictive nonlinear structure a common approach is to reconstruct the generating processes dynamics using a technique called delay-coordinate embedding and then use the geometric structure of this reconstruction as a proxy for the information transmission process of the underlying process to predict the future.
+When a generating process exhibits deterministic predictive nonlinear structure a common approach is to reconstruct the generating processes dynamics using a technique called delay-coordinate embedding and then use the geometric structure of this reconstruction as a proxy for the information transmission occurring in the underlying process to predict the future.
 In this section we discuss the theory and implementation of this forecasting scheme.
 
- \subsubsection{Reconstructing hidden dynamics {\color{red} NOT EDITABLE}}
+ \subsubsection{Reconstructing hidden dynamics {\color{blue} EDITABLE}}
 
-Delay-coordinate embedding allows one to reconstruct a system's full
+Delay-coordinate embedding allows one to reconstruct a dynamical system's full
 state-space dynamics from a \emph{single} scalar time-series
@@
 function---the mapping from the unknown state vector $\vec{X}$ to the
-scalar value $x$ that one is measuring---are both smooth and generic,
+scalar value $X_{i,obs}$ that one is measuring---are both smooth and generic,
 Takens~\cite{takens} formally proves that the delay-coordinate map
@@
 $\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92}.  A
+method of \cite{KBA92} to estimate $m$.  A
 plot of the data from Figure~\ref{fig:ipc}, embedded following this
@@
    \centering
-\begin{subfigure}{0.49\textwidth}
+\begin{subfigure}{0.32\textwidth}
     \includegraphics[width=\textwidth]{figs/colipc3d.png}
@@
   \end{subfigure}%   
-  \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colipc3d.png}
+  \begin{subfigure}{0.32\textwidth}
+    \includegraphics[width=\textwidth]{figs/svd53dipc.png}
+    \caption{\svdfive}
+    \label{fig:svdfiveEmbedding}
+  \end{subfigure}%   
+  \begin{subfigure}{0.32\textwidth}
+    \includegraphics[width=\textwidth]{figs/gcc3dipc.png}
     \caption{\gcc}
@@
 
+[[Update this paragraph if we include the second plot with \gcc]]
 %% can cut for space if need be:

### 2014-03-05 pecomplexity.tex
finished the results section, mostly
moved a small chunk to concl
defined a newcommand
added usepackage{color}
adding sectoin on simple methods and cleaning
adding some figures and editing

@@
 \newcommand{\svdsix}{{\tt dgesdd$_6$}~}
+\newcommand{\naive}{na\"ive}
+
+\parskip=3pt
 
@@
 \begin{abstract}
-[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
+[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus, should be done after we have a first draft of the paper]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
   Among other things, that implies that one should be able to use

### 2014-03-05 related.tex
editing

--- 2014-02-26 related.tex
@@
-\section{Related Work}
+\section{Related Work {\color{red} NOT EDITABLE}}\label{sec:related}
 
@@
 
-Entropy though Generating partitions 
-works **if** you have generating parition but not if you don't
+The ideal method of computing the entropy of a system is via \emph{generating partitions}~\cite{lind95}, a technique which partitions observations into symbols in a way which preserves the underlying dynamics.
+Unfortunately, outside of one-dimensional unimodal maps and a few select higher dimensional maps, generating partitions are highly non-trivial to compute and often have fractal boundaries~\cite{eisele1999}.
+Furthermore, the underlying dynamic needs to be known to construct the generating partition.
+Unfortunately, using any partition (such as those produced by simply binning the data) other than the generation partition induces spurious complexity to the resulting symbolic sequence, thus misrepresenting the entropy of the underlying system~\cite{bollt2001}.
 

### 2014-03-05 model.tex
editing

@@
 \[
-F(\tau,m)(X) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
+F(\tau,m)(\vec{X}) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
 \]
@@
 method of \cite{KBA92} to estimate $m$.  A
-plot of the data from Figure~\ref{fig:ipc}, embedded following this
-procedure, is shown in Figure~\ref{fig:embedding}.
+plot of the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following this
+procedure, are shown in Figure~\ref{fig:embedding}.
 
@@
     \begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/svd53dipc.png}
+    \includegraphics[width=\textwidth]{figs/gcc3dipc.png}
+    \caption{\gcc}
+    \label{fig:gccEmbedding}
+  \end{subfigure} 
+  \begin{subfigure}{0.32\textwidth}
+    \includegraphics[width=\textwidth]{figs/svd53dipc2.png}
     \caption{\svdfive}
@@
   \end{subfigure}%   
-  \begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/gcc3dipc.png}
-    \caption{\gcc}
-    \label{fig:gccEmbedding}
-  \end{subfigure}%
+%
 
      %\includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{[[{\color{red}Joshua:Perhaps add a similar graphic of \gcc to show contrast}{\color{green}agreed I will work on this}]]A 3D projection of a delay-coordinate embedding of the trace
- from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
+     \caption{A 3D projection of the delay-coordinate embeddings of the traces
+ from: (a) \col seen in Figure~\ref{fig:col-ts} (b)\gcc seen in Figure~\ref{fig:gcc-ts} and (c) \svdfive seen as the fifth regime in Figure~\ref{fig:svd-ts-colored}.
  }
@@
 
-[[Update this paragraph if we include the second plot with \gcc]]
 %% can cut for space if need be:
-The coordinates of each point on this plot are differently delayed
-elements of the \col instructions per cycle time series
+The coordinates of each point in these plots are differently delayed
+elements of the instructions per cycle time series
 $X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
@@
 Structure in these kinds of plots---clearly visible in
-Figure~\ref{fig:embedding}---is an indication of
-determinism%\footnote{A deeper analysis of
-  %Figure~\ref{fig:embedding}---as alluded to on the previous
-  %page---supports that diagnosis, confirming the presence of a chaotic
-  %attractor in these cache-miss dynamics, with largest Lyapunov
-  %exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
-  %12-dimensional reconstruction space \cite{mytkowicz09}.}
-  .  That structure can be used to build a forecast model. 
- \subsubsection{LMA: Using dynamics in forecasting {\color{red} NOT EDITABLE}}
+Figure~\ref{fig:embedding} (a) is an indication of
+structure in the information transmission process, that can be leveraged for prediction.
+Where the water becomes muddy is with embedding like those in Figure~\ref{fig:embedding} (b) \& (c).
+Figure~\ref{fig:embedding} (b) shows little to no structure, this embedding is essentially a noise ball, with an embedding like this we would expect the forecast accuracy that uses this structure to be lacking.
+However, with Figure~\ref{fig:embedding} (c) more structure is present but the structure is surrounded in noise.
+Is the structure enough to aid in accurately forecasting the signal?
+This is a difficult question to answer and one that comes back to the main contributions of this paper.
+It may be that \gcc has structure but structure that is not amenable by this method, similarly \svdfive clearly has structure but it is obfuscated by noise.
+By analyzing qualitatively the amount of complexity in these signals we can answer questions like this.
+If enough predictive structure is present in these embeddings as seen in Figure~\ref{fig:embedding} (a) and possibly in (c), that structure can be used to build a forecast model which we discuss in the next section.
+ \subsubsection{LMA: Using dynamics in forecasting {\color{blue} EDITABLE}}
 
@@
 shown in Figure~\ref{fig:lmacol}.
+Visually distinguishing if Figure~\ref{fig:lmasvdfive} is a good prediction is hard.
+To more fairly compare these forecasts as well as the forecasts made by the other three prediction strategies we calculate a figure of merit, which we introduce in Section \ref{sec:accuracy}.
 %
-\begin{figure}[htbp]
+\begin{figure}
    \centering
+\begin{subfigure}{0.32\textwidth}
     \includegraphics[width=\textwidth]{figs/colPredShortTS}
-    \caption{A forecast of the last 4,000 points of the signal in
-      Figure~\ref{fig:ipc} using an LMA-based strategy on the
-      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
-      $\times$s are the true and predicted values, respectively;
-      vertical bars show where these values differ.}
+    \caption{\col }
     \label{fig:lmacol}
+  \end{subfigure}%  
+    \begin{subfigure}{0.32\textwidth}
+    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+    \caption{\gcc}
+    \label{fig:lmagcc}
+  \end{subfigure} 
+  \begin{subfigure}{0.32\textwidth}
+    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+    \caption{\svdfive}
+    \label{fig:lmasvdfive}
+  \end{subfigure}% 
+  \caption{{\color{red} [[actually need to generate this figure]]}An LMA-based forecast of
+       processor-efficiency performance trace from the\col ,\gcc and \svdfive generating processes.  Red circles and blue $\times$s are the true and
+       predicted values, respectively; vertical bars show where these
+       values differ.}\label{fig:lmapredictions}
  \end{figure}
 %
+%\begin{figure}[htbp]
+%  \centering
+%   \includegraphics[width=\textwidth]{figs/colPredShortTS}
+%    \caption{A forecast of the last 4,000 points of the %signal in
+%      Figure~\ref{fig:ipc} using an LMA-based strategy on %the
+%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
+%      $\times$s are the true and predicted values, %respectively;
+%      vertical bars show where these values differ.}
+%\label{fig:lmacol}
+%\end{figure}
+%
+%\begin{figure}[htbp]
+%  \centering
+%    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+%     \caption{{\color{red} [[actually need to generate this %figure]]}An LMA-based forecast of the last 4,000 points of a
+%       processor-load performance trace from the \gcc
+%       benchmark.  Red circles and blue $\times$s are the %true and
+%       predicted values, respectively; vertical bars show %where these
+%       values differ.}
+%\label{fig:lmagcc}
+%\end{figure}
+%
 On the other hand, if we use the same approach to forecast the
-processor load (ipc) of the \gcc program from the SPEC cpu2006 benchmark suite, the
+processor efficiency (IPC) of the \gcc time series, the
 prediction is far less accurate; see Figure~\ref{fig:lmagcc}.
 
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=\textwidth]{figs/colPredShortTS}
-     \caption{{\color{red} [[actually need to generate this figure]]}An LMA-based forecast of the last 4,000 points of a
-       processor-load performance trace from the \gcc
-       benchmark.  Red circles and blue $\times$s are the true and
-       predicted values, respectively; vertical bars show where these
-       values differ.}
-\label{fig:lmagcc}
-\end{figure}
+This gets at the utility of the contribution of this paper: These time series all come from the same system---computers---but they are not equally predictable (at this point by LMA).
+For a practitioner is it the case that
 
-This gets at the utility of the contribution of this paper: If these traces both come from deterministic
-systems---computers---then why are they not equally predictable?  Our
+  Our
 conjecture is that while they come from similar systems \gcc produces processor load traces on the top of the complexity spectrum whereas \col produces processor loads with complexity in the mid to low region of the complexity spectrum.
@@
   \end{subfigure}  
-
   %\begin{subfigure}{0.5\textwidth}
@@
    \caption{
-For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
+{\color{red}[[Joshua: I should probably add \svdfive as it is a persistent example making this figure 9x9.}For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
 On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.

### 2014-03-06 intro.tex
add comment about redundancy to intro

@@
 A first step in answering this question is to answer: is it possible to reliably quantify where on the spectrum of complexity the time series lies?
-i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast using a prediction strategy based on predictive structure (temporal forward information transfer)?
-We rigorously define this \emph{complexity spectrum} as well as what is meant by \emph{low, medium and high} in Section \ref{sec:meaComplex}.
+i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast?
+Since the information in an observation can be partitioned into two pieces --- redundancy and entropy generation\cite{crutchfield2003} --- our goal is to analyze how the amount of predictive structure present in a signal, as measured by the redundancy, relates to the performance of prediction strategies, as well as what is meant by \emph{low, medium and high} complexity in Section \ref{sec:meaComplex}.
 

### 2014-03-06 intro.tex
some wording changes

@@
 i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast?
-Since the information in an observation can be partitioned into two pieces --- redundancy and entropy generation\cite{crutchfield2003} --- our goal is to analyze how the amount of predictive structure present in a signal, as measured by the redundancy, relates to the performance of prediction strategies, as well as what is meant by \emph{low, medium and high} complexity in Section \ref{sec:meaComplex}.
+Since the information in an observation can be partitioned into two pieces --- redundancy and entropy generation\cite{crutchfield2003} --- our goal is to analyze how the amount of predictive structure present in a signal, as measured by the redundancy, relates to the performance of prediction strategies.
+The formation of this spectrum or partition as well as what is meant by \emph{low, medium and high} complexity are visited in more detail in Section \ref{sec:meaComplex}.
 
@@
 \cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture.
-In light of this, for the purposes of this paper we define \emph{complexity} as an approximation of Shannon \emph{entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}.
+In light of this, for the purposes of this paper we define \emph{complexity} as a particular approximation of Kolmogorov-Sinai \emph{entropy}[[Probably need a citaiton for KS entropy]], which we define and explain rigorously in Section \ref{sec:meaComplex}.
 In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity.

### 2014-03-06 wpe.tex
some wording changes

@@
 and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
+permutation entropy has been shown to be identical to the Shannon entropy for {\color{red}[[Ryan is this the case or is it the KS entropy?]]}
 many large classes of systems \cite{amigo2012permutation}.

### 2014-03-06 pecomplexity.tex
putting it in Phys Rev E Format

@@
-\documentclass{article}
+\documentclass[%
+ %reprint,
+%superscriptaddress,
+%groupedaddress,
+%unsortedaddress,
+%runinaddress,
+%frontmatterverbose, 
+%preprint,
+%showpacs,preprintnumbers,
+%nofootinbib,
+%nobibnotes,
+%bibnotes,
+ amsmath,amssymb,
+% aps,
+%pra,
+prb,
+%rmp,
+%prstab,
+%prstper,
+%floatfix,
+]{revtex4-1}
+%\documentclass{article}
 \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
@@
 \usepackage{subcaption}
-\usepackage{color}
-\usepackage{amssymb}
-\usepackage{amsmath}
+\usepackage{dcolumn}% Align table columns on decimal point
+\usepackage{bm}% bold math
+%\usepackage{amssymb}
+%\usepackage{amsmath}
 \usepackage{epstopdf}
@@
 \usepackage{pifont} % check and x marks
+\usepackage{color}
+
 \newcommand{\cmark}{\ding{51}}
@@
 \newcommand{\naive}{na\"ive}
-
-\parskip=3pt
-
 \newcommand{\row}{{\tt row\_major}~}
 \newcommand{\col}{{\tt col\_major}~}
+%\newcommand{\color{red}}{{\color[rgb]{1,0,0}}}
 
-\usepackage{authblk}
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
+%\parskip=3pt
+
+%\usepackage{authblk}
+%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
 
 \newtheorem*{mydef}{Definition}
-\title{Determinism, Complexity, and Predictability in Computer Performance\\
-Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity \\
-{\color{green}Quantifying Predictability through Structural Complexity} 
-\\ Quantifying Predictability using Structural Complexity 
-\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
-\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
-}
 
-\author[1]{Joshua Garland \thanks{joshua.garland@colorado.edu}}
-\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
-\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-\affil[1]{Department of Computer Science\\
-  University of Colorado at Boulder\\
-  Colorado, USA
-}
-\affil[2]{Santa Fe Institute\\
-  New Mexico, USA
-}
+%\author[1]{Joshua Garland %\thanks{joshua.garland@colorado.edu}}
+%\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
+%\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
+%\affil[1]{Department of Computer Science\\
+%  University of Colorado at Boulder\\
+%  Colorado, USA
+%}
+%\affil[2]{Santa Fe Institute\\
+%  New Mexico, USA
+%}
 
@@
 
-\maketitle
+\preprint{APS/123-QED}
+
+%\title{Manuscript Title}% Force line breaks with \\
+%\thanks{A footnote to the article title}%
+
+\title{Determinism, Complexity, and Predictability in Computer Performance\\
+Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
+\\ Quantifying Predictability through Structural Complexity \\
+{\color{green}Quantifying Predictability through Structural Complexity} 
+\\ Quantifying Predictability using Structural Complexity 
+\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
+\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
+}
+
+\author{Ann Author}
+ \altaffiliation[Also at ]{Physics Department, XYZ University.}%Lines break automatically or can be forced with \\
+\author{Second Author}%
+ \email{Second.Author@institution.edu}
+\affiliation{%
+ Authors' institution and/or address\\
+ This line break forced with \textbackslash\textbackslash
+}%
+
+\collaboration{MUSO Collaboration}%\noaffiliation
+
+\author{Charlie Author}
+ \homepage{http://www.Second.institution.edu/~Charlie.Author}
+\affiliation{
+ Second institution and/or address\\
+ This line break forced% with \\
+}%
+\affiliation{
+ Third institution, the second for Charlie Author
+}%
+\author{Delta Author}
+\affiliation{%
+ Authors' institution and/or address\\
+ This line break forced with \textbackslash\textbackslash
+}%
+
+\collaboration{CLEO Collaboration}%\noaffiliation
+
+\date{\today}% It is always \today, today,
+             %  but any date may be explicitly specified
 
@@
   those values against the prediction success.
+\begin{description}
+\item[Usage]
+Secondary publications and information retrieval purposes.
+\item[PACS numbers]
+May be entered using the \verb+\pacs{#1}+ command.
+\item[Structure]
+You may use the \texttt{description} environment to structure your abstract;
+use the optional argument of the \verb+\item+ command to give the category of each item. 
+\end{description}
 \end{abstract}
+
+\pacs{Valid PACS appear here}% PACS, the Physics and Astronomy
+                             % Classification Scheme.
+%\keywords{Suggested keywords}%Use showkeys class option if keyword
+                              %display desired
+\maketitle
 

### 2014-03-06 pecomplexity.tex
more PRE formatting

@@
 \documentclass[%
- %reprint,
+ reprint,
 %superscriptaddress,
@@
 %pra,
-prb,
+%prb,
 %rmp,
@@
 \newtheorem*{mydef}{Definition}
-
-%\author[1]{Joshua Garland %\thanks{joshua.garland@colorado.edu}}
-%\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
-%\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
-%\affil[1]{Department of Computer Science\\
-%  University of Colorado at Boulder\\
-%  Colorado, USA
-%}
-%\affil[2]{Santa Fe Institute\\
-%  New Mexico, USA
-%}
 
@@
 
-\author{Ann Author}
- \altaffiliation[Also at ]{Physics Department, XYZ University.}%Lines break automatically or can be forced with \\
-\author{Second Author}%
- \email{Second.Author@institution.edu}
+%\author[1]{Joshua Garland %\thanks{joshua.garland@colorado.edu}}
+%\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
+%\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
+%\affil[1]{Department of Computer Science\\
+%  University of Colorado at Boulder\\
+%  Colorado, USA
+%}
+%\affil[2]{Santa Fe Institute\\
+%  New Mexico, USA
+%}
+
+\author{Joshua Garland}
+% \altaffiliation[Also at ]{}%Lines break automatically or can be forced with \\
+%\affiliation{%
+% Department of Computer Science, University of Colorado at %Boulder. Colorado, USA
+%}%
+\email{joshua.garland@colorado.edu}
+\author{Ryan James}%
+ \email{ryan.james@colorado.edu}
 \affiliation{%
- Authors' institution and/or address\\
- This line break forced with \textbackslash\textbackslash
+ Department of Computer Science, University of Colorado at Boulder. Colorado, USA
+ %This line break forced with \textbackslash\textbackslash
 }%
-
-\collaboration{MUSO Collaboration}%\noaffiliation
-
-\author{Charlie Author}
- \homepage{http://www.Second.institution.edu/~Charlie.Author}
+%\collaboration{MUSO Collaboration}%\noaffiliation
+\author{Elizabeth Bradley}\email{lizb@colorado.edu}
 \affiliation{
- Second institution and/or address\\
- This line break forced% with \\
+ Department of Computer Science, University of Colorado at Boulder. Colorado, USA
 }%
 \affiliation{
- Third institution, the second for Charlie Author
+ Santa Fe Institute, New Mexico, USA\\
+ %This line break forced% with \\
 }%
-\author{Delta Author}
-\affiliation{%
- Authors' institution and/or address\\
- This line break forced with \textbackslash\textbackslash
-}%
-
-\collaboration{CLEO Collaboration}%\noaffiliation
 

### 2014-03-06 pecomplexity.tex
fixing bibfor revtex
more pre formatting

@@
 %showpacs,preprintnumbers,
-%nofootinbib,
-%nobibnotes,
+nofootinbib,
+nobibnotes,
 %bibnotes,
@@
 
-\title{Determinism, Complexity, and Predictability in Computer Performance\\
-Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
-\\ Quantifying Predictability through Structural Complexity \\
+\title{
+%Determinism, Complexity, and Predictability in Computer Performance\\
+%Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
+%\\ 
 {\color{green}Quantifying Predictability through Structural Complexity} 
-\\ Quantifying Predictability using Structural Complexity 
+%\\ 
+%Quantifying Predictability using Structural Complexity 
 \\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
@@
 
-\author{Joshua Garland}
-% \altaffiliation[Also at ]{}%Lines break automatically or can be forced with \\
-%\affiliation{%
-% Department of Computer Science, University of Colorado at %Boulder. Colorado, USA
-%}%
-\email{joshua.garland@colorado.edu}
+\author{Joshua Garland}\email{joshua.garland@colorado.edu}
 \author{Ryan James}%
@@
 \affiliation{%
- Department of Computer Science, University of Colorado at Boulder. Colorado, USA
- %This line break forced with \textbackslash\textbackslash
-}%
+ Department of Computer Science, University of Colorado at Boulder. Colorado, USA}%
 %\collaboration{MUSO Collaboration}%\noaffiliation
@@
 \begin{abstract}
-[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus, should be done after we have a first draft of the paper]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
+{\color{green}From the site:It should be about 5\% of the length of the article, but less than 500 words.}[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus, should be done after we have a first draft of the paper]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
   Among other things, that implies that one should be able to use
@@
 \item[PACS numbers]
-May be entered using the \verb+\pacs{#1}+ command.
+May be entered using the \verb+\pacs{#1}+ \pacs{05.45.Tp} command.
+Time series analysis (nonlinear dynamics) 05.45.Tp
+Information theory, 89.70.-a
+	channel capacity in, 89.70.Kn
+	communication complexity in, 89.70.Hj
+	computational complexity in, 89.70.Eg
+	entropy in, 89.70.Cf
+	general biological information, 87.10.Vg
+	in neuroscience, 87.19.lo
+
+Nonlinear dynamics, 05.45.-a
+
+	89.75.Kd	Patterns (complex systems)
+
+Computational Physics	
+95.75.Wx	Time series analysis, time variability
+95.75.Pq	Mathematical procedures and computer techniques
+07.05.-t	Computers in experimental physics
+
 \item[Structure]
@@
 
-\pacs{Valid PACS appear here}% PACS, the Physics and Astronomy
+\pacs{Valid PACS appear here 95.75.Wx	}% PACS, the Physics and Astronomy
                              % Classification Scheme.
@@
 \input{concl.tex}
+
+\section*{Acknowledgment}
+This work was partially supported by NSF grant \#CMMI-1245947 and ARO
+grant \#W911NF-12-1-0288.
 
@@
 
-\section*{Acknowledgment}
-This work was partially supported by NSF grant \#CMMI-1245947 and ARO
-grant \#W911NF-12-1-0288.
-
-\bibliographystyle{unsrt}
+\bibliographystyle{aipauth4-1}
 \bibliography{bibliofile}

### 2014-03-06 results.tex
putting it in Phys Rev E Format
more pre formatting

@@
- \section{Predictability, Complexity, and Permutation Entropy {\color{blue}  EDITABLE}} % (fold)
+ \section{Predictability, Complexity, and Permutation Entropy %{\color{blue}  EDITABLE}
+ } % (fold)
  \label{sec:results}
@@
   \centering
-  \includegraphics[width=0.6\textwidth]{figs/prediction_vs_entropy}
+  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
   \caption{Weighted permutation entropy vs. mean absolute scaled error
@@
   \centering
-    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
-    \includegraphics[width=0.6\textwidth]{figs/LMA_prediction_vs_entropy}
+    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
+    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
+    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
 \caption{WPE vs. MASE for all trials, methods, and systems.  \svdone,
@@
   \centering
-    \includegraphics[width=\textwidth]{figs/svdonets2}
+    \includegraphics[width=\columnwidth]{figs/svdonets2}
 \caption{A small portion of the \svdone time series}\label{fig:svdone-ts}

### 2014-03-07 model.tex
putting it in Phys Rev E Format
more pre formatting
trying to get this section into shape

@@
-\section{Modeling {\color{blue} EDITABLE}}\label{sec:model}
+\section{Modeling }\label{sec:model}
+{\color{blue} EDITABLE}
 
@@
    \centering
-\begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/colipc3d.png}
+\begin{subfigure}{\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
     \caption{\col }
     \label{fig:colEmbedding}
-  \end{subfigure}%  
-    \begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/gcc3dipc.png}
+  \end{subfigure}%  \\
+
+    \begin{subfigure}{\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gcc3dipc.png}
     \caption{\gcc}
@@
   \end{subfigure} 
-  \begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/svd53dipc2.png}
+  \\
+  \begin{subfigure}{\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/svd53dipc2.png}
     \caption{\svdfive}
@@
 Where the water becomes muddy is with embedding like those in Figure~\ref{fig:embedding} (b) \& (c).
-Figure~\ref{fig:embedding} (b) shows little to no structure, this embedding is essentially a noise ball, with an embedding like this we would expect the forecast accuracy that uses this structure to be lacking.
+Figure~\ref{fig:embedding} (b) shows little to no structure, this embedding is essentially a noise cloud; with an embedding like this we would expect the forecast accuracy that uses this structure to be lacking.
 However, with Figure~\ref{fig:embedding} (c) more structure is present but the structure is surrounded in noise.
@@
 This is a difficult question to answer and one that comes back to the main contributions of this paper.
-It may be that \gcc has structure but structure that is not amenable by this method, similarly \svdfive clearly has structure but it is obfuscated by noise.
-By analyzing qualitatively the amount of complexity in these signals we can answer questions like this.
-If enough predictive structure is present in these embeddings as seen in Figure~\ref{fig:embedding} (a) and possibly in (c), that structure can be used to build a forecast model which we discuss in the next section.
+It may be that \gcc has some redundancy but that structure is not amenable to this method, similarly \svdfive clearly has structure but it is obfuscated by noise.
+By quantifying the balance between normalized redundancy (predictive structure) and entropy for these real valued time series, we can begin to answer these questions.
+If enough (redundancy) predictive structure is present in these embeddings as seen in Figure~\ref{fig:embedding} (a) and possibly in (c), that structure can be used to build a forecast model which we discuss in the next section.
  \subsubsection{LMA: Using dynamics in forecasting {\color{blue} EDITABLE}}
@@
 Given a nonlinear model of a deterministic dynamical system in the
-form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
+form of a delay-coordinate embedding like those in Figure~\ref{fig:embedding},
 one can build deterministic forecast algorithms by capturing and
@@
 space~\cite{lorenz-analogues}.  Even this simple algorithm---which
-builds predictions by finding the nearest neighbor in the embedded
-space of the given point, then taking that neighbor's path as the
-forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
+builds predictions by finding the nearest neighbor in the embedded space of the given point, then taking that neighbor's path as the
+forecast provides quite accurate forecasts when a process generates and transmits information in this way.
+In particular, when an observation in a time series has redundancy in the form of nonlinear structure this forecast scheme should be a great choice.
+However, since this method is based on near-neighbor prediction in a reconstructed state-space, if a system does not exhibit deterministic structure this method can do quite poorly.
+{\color{red}[[Joshua:Not a fan of this wording]]}
+%% [[If space: Add a segue sentence: next section uses permutation
+%% entropy to explore that conjecture.]]
+
+%\subsection{Comparing Prediction Accuracy}
+
+\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{red} NOT EDITABLE}}
+\label{sec:accuracy}
+
+\begin{figure}[htbp]
+  \centering
+      \begin{subfigure}{0.48\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
+    \caption{\col na\"ive }
+    \label{fig:gccMEAN}
+  \end{subfigure}%
+   \begin{subfigure}{0.48\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
+    \caption{\gcc na\"ive }
+    \label{fig:gccMEAN}
+  \end{subfigure}%      
+  \\
+    \begin{subfigure}{0.48\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
+    \caption{\col ARIMA}
+    \label{fig:colARIMA}
+  \end{subfigure}
+  \begin{subfigure}{0.48\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
+    \caption{\gcc ARIMA }
+    \label{fig:gccARIMA}
+  \end{subfigure}%
+  \\
+
+      \begin{subfigure}{0.48\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
+    \caption{\col LMA}
+    \label{fig:colLMA}
+  \end{subfigure}
+      \begin{subfigure}{0.48\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
+    \caption{\gcc LMA}
+    \label{fig:gccLMA}
+  \end{subfigure}  
+  %\begin{subfigure}{0.5\textwidth}
+  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
+   \caption{
+{\color{red}[[Joshua: I should probably add \svdfive as it is a persistent example making this figure 9x9.}For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
+On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
+}\label{fig:gcc_vs_col}
+  %\end{subfigure}
+\end{figure} 
+In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
+The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:model}.
+The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
+In particular, we perform $k$ one-step predictions, after each one-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
+This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
+
+As a figure of merit we calculate the Mean Absolute Scaled Error (MASE)\cite{MASE} between the true and predicted signals: 
+%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
+The scaling term for MASE:
+$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
+is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
+This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." We chose MASE because it allows for fair comparison across varying methods, prediction horizons and signal scales.
+When a forecast results in a $MASE<1$ the prediction method produced, on average, smaller errors than the one-step errors from the in-sample random walk forecast strategy.
+Analogously, $MASE>1$ means that the prediction method did worse, on average than the one-step errors for the in-sample random walk forecast strategy.
+In Table \ref{tab:error} we provide the distribution {\color{red}[[Joshua: Ryan, Is this the right word?
+we give mean $\pm$ std.
+dev but some have very skewed right tails]]}  of MASE errors for each of the eight signals and three prediction strategies, these are averaged over 15 runs of each type (signal + method).
+[[and cherry pick a few examples of \gcc and \col to put in the text For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=6$.]]
+
+---works quite well on the trace in Figure~\ref{fig:ipc}, as
 shown in Figure~\ref{fig:lmacol}. 
-Visually distinguishing if Figure~\ref{fig:lmasvdfive} is a good prediction is hard.
-To more fairly compare these forecasts as well as the forecasts made by the other three prediction strategies we calculate a figure of merit, which we introduce in Section \ref{sec:accuracy}.
 %
@@
    \centering
-\begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+\begin{subfigure}{\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colPredShortTS}
     \caption{\col }
@@
   \end{subfigure}%
-    \begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+
+    \begin{subfigure}{\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colPredShortTS}
     \caption{\gcc}
@@
   \end{subfigure} 
-  \begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+  \begin{subfigure}{\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colPredShortTS}
     \caption{\svdfive}
@@
   \end{subfigure}% 
-  \caption{{\color{red} [[actually need to generate this figure]]}An LMA-based forecast of
+  \caption{{\color{red} [[If Liz thinks we should include these I actually need to generate this figure]]}An LMA-based forecast of
        processor-efficiency performance trace from the\col ,\gcc and \svdfive generating processes.  Red circles and blue $\times$s are the true and
@@
 prediction is far less accurate; see Figure~\ref{fig:lmagcc}.
+Figure~\ref{fig:lmasvdfive} is clearly a better prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.
 
@@
 For a practitioner is it the case that
-
   Our
@@
 
-%% [[If space: Add a segue sentence: next section uses permutation
-%% entropy to explore that conjecture.]]
-
-%\subsection{Comparing Prediction Accuracy}
-
-\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{red} NOT EDITABLE}}
-\label{sec:accuracy}
-
-\begin{figure}[htbp]
-  \centering
-      \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colMeanForecast}
-    \caption{\col na\"ive }
-    \label{fig:gccMEAN}
-  \end{subfigure}%
-   \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccMeanForecast}
-    \caption{\gcc na\"ive }
-    \label{fig:gccMEAN}
-  \end{subfigure}%      
-  \\
-    \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
-    \caption{\col ARIMA}
-    \label{fig:colARIMA}
-  \end{subfigure}
-  \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
-    \caption{\gcc ARIMA }
-    \label{fig:gccARIMA}
-  \end{subfigure}%
-  \\
-      \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/colLMAForecast}
-    \caption{\col LMA}
-    \label{fig:colLMA}
-  \end{subfigure}
-      \begin{subfigure}{0.49\textwidth}
-    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
-    \caption{\gcc LMA}
-    \label{fig:gccLMA}
-  \end{subfigure}  
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{
-{\color{red}[[Joshua: I should probably add \svdfive as it is a persistent example making this figure 9x9.}For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
-On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
-}\label{fig:gcc_vs_col}
-  %\end{subfigure}
-\end{figure} 
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
-
-As a figure of merit we calculate the Mean Absolute Scaled Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales.
-When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method).
-[[and cherry pick a few examples of \gcc and \col to put in the text For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=6$.]]
-
 % took out for space

### 2014-03-07 model.tex
fixing and adding forecast figures

@@
 builds predictions by finding the nearest neighbor in the embedded space of the given point, then taking that neighbor's path as the
-forecast provides quite accurate forecasts when a process generates and transmits information in this way.
-In particular, when an observation in a time series has redundancy in the form of nonlinear structure this forecast scheme should be a great choice.
+forecast, provides accurate forecasts when the generating process is a deterministic dynamical system.
+In particular, when an observation in a time series has redundancy in the form of nonlinear predictive structure this forecast scheme should be a great choice.
 However, since this method is based on near-neighbor prediction in a reconstructed state-space, if a system does not exhibit deterministic structure this method can do quite poorly.
@@
   \centering
-      \begin{subfigure}{0.48\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
-    \caption{\col na\"ive }
+      \begin{subfigure}{0.33\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colMeanForecast.png}
+    \caption{\col\\ na\"ive }
     \label{fig:gccMEAN}
   \end{subfigure}%
-   \begin{subfigure}{0.48\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
-    \caption{\gcc na\"ive }
+   \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gccMeanForecast.png}
+    \caption{\gcc\\ na\"ive }
+    \label{fig:gccMEAN}
+  \end{subfigure}% 
+     \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast.png}
+    \caption{\svdfive\\ na\"ive }
     \label{fig:gccMEAN}
@@
   \\
-    \begin{subfigure}{0.48\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
+    \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colARIMAForecast.png}
+
     \caption{\col ARIMA}
@@
   \end{subfigure}
-  \begin{subfigure}{0.48\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
-    \caption{\gcc ARIMA }
+  \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast.png}
+    \caption{\gcc \\ ARIMA }
+    \label{fig:gccARIMA}
+  \end{subfigure}%
+  \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast.png}
+    \caption{\svdfive \\ ARIMA }
     \label{fig:gccARIMA}
@@
 
-      \begin{subfigure}{0.48\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
+      \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
     \caption{\col LMA}
@@
   \end{subfigure}
-      \begin{subfigure}{0.48\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
+      \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
     \caption{\gcc LMA}
+    \label{fig:gccLMA}
+  \end{subfigure}  
+        \begin{subfigure}{0.32\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
+    \caption{\svdfive LMA}
     \label{fig:gccLMA}
@@
 %
-\begin{figure}
-   \centering
-\begin{subfigure}{\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colPredShortTS}
-    \caption{\col }
-    \label{fig:lmacol}
-  \end{subfigure}%
-
-    \begin{subfigure}{\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colPredShortTS}
-    \caption{\gcc}
-    \label{fig:lmagcc}
-  \end{subfigure} 
-  \begin{subfigure}{\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colPredShortTS}
-    \caption{\svdfive}
-    \label{fig:lmasvdfive}
-  \end{subfigure}% 
-  \caption{{\color{red} [[If Liz thinks we should include these I actually need to generate this figure]]}An LMA-based forecast of
-       processor-efficiency performance trace from the\col ,\gcc and \svdfive generating processes.  Red circles and blue $\times$s are the true and
-       predicted values, respectively; vertical bars show where these
-       values differ.}\label{fig:lmapredictions}
- \end{figure}
+%\begin{figure}
+%   \centering
+%\begin{subfigure}{\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
+%    \caption{\col }
+%    \label{fig:lmacol}
+%  \end{subfigure}%  
+%  
+%    \begin{subfigure}{\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
+%    \caption{\gcc}
+%    \label{fig:lmagcc}
+%  \end{subfigure} 
+%  \begin{subfigure}{\columnwidth}
+%%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
+%    \caption{\svdfive}
+%    \label{fig:lmasvdfive}
+%  \end{subfigure}% 
+%  \caption{{\color{red} [[If Liz thinks we should %include these I actually need to generate this %figure]]}An LMA-based forecast of
+%       processor-efficiency performance trace from the%\col ,\gcc and \svdfive generating processes.  Red %circles and blue $\times$s are the true and
+%       predicted values, respectively; vertical bars %show where these
+%       values differ.}\label{fig:lmapredictions}
+% \end{figure}
 %

### 2014-03-10 model.tex

@@
 
-\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{red} NOT EDITABLE}}
+\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{blue} EDITABLE}}
 \label{sec:accuracy}
+
+In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
+The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:model}.
+The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
+In particular, we perform $k$ one-step predictions, after each one-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
+This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
+
+As a figure of merit we calculate the Mean Absolute Scaled Error (MASE)\cite{MASE} between the true and predicted signals: 
+%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
+The scaling term for MASE:
+$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
+is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
+This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." We chose MASE because it allows for fair comparison across varying methods, prediction horizons and signal scales.
+When a forecast results in a $MASE<1$ the prediction method produced, on average, smaller errors than the one-step errors from the in-sample random walk forecast strategy.
+Analogously, $MASE>1$ means that the prediction method did worse, on average than the one-step errors for the in-sample random walk forecast strategy.
+
+Table \ref{tab:error} provides the MASE distributions {\color{red}[[Joshua: Ryan, Is this the right word?
+we give mean $\pm$ std.
+dev but some have very skewed right tails]]} for each of the eight signals and three prediction strategies, these are averaged over 15 runs of each signal-method combination.
+For later comparison, Table \ref{tab:error} also has the complexity measure we introduce in Section  \ref{sec:meaComplex}.
+
+To aid in a conceptual understanding of this error metric we provide example forecasts in Figure~\ref{fig:forecast-example}.
+In plots like this we plot $(c_j,p_j)$, visually points close to the line $y=x$, i.e.
+the line $p_j = c_j$ are more accurate.
+Forecasts which stay close to the line are more accurate than forecasts that stray away from it.
 
@@
    \caption{
-{\color{red}[[Joshua: I should probably add \svdfive as it is a persistent example making this figure 9x9.}For each of these, we plot the predicted value $p_n$ against the correct value $c_n$.
-On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction.
-}\label{fig:gcc_vs_col}
+{\color{red}[[Joshua: Liz, I added \svdfive as it has been a persistent example up to this figure.
+I am a little worried about the 9x9 being too small in 1 column.
+We may need to make this figure span two columns, which I think is ok.]]}For each of these, we plot the predicted value $p_j$ against the correct value $c_j$.
+On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_j = c_j$, e.g., (g) is a near perfect prediction where-as (e) is a poor prediction.}\label{fig:forecast-example}
   %\end{subfigure}
 \end{figure}
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:model}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ one-step predictions, after each one-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
 
-As a figure of merit we calculate the Mean Absolute Scaled Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." We chose MASE because it allows for fair comparison across varying methods, prediction horizons and signal scales.
-When a forecast results in a $MASE<1$ the prediction method produced, on average, smaller errors than the one-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the one-step errors for the in-sample random walk forecast strategy.
-In Table \ref{tab:error} we provide the distribution {\color{red}[[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]}  of MASE errors for each of the eight signals and three prediction strategies, these are averaged over 15 runs of each type (signal + method).
-[[and cherry pick a few examples of \gcc and \col to put in the text For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=6$.]]
+It may be helpful for the reader at this point to compare some of the MASE scores in Table \ref{tab:error} with the example forecasts in Figure~\ref{fig:forecast-example}.
+For example, if we compare Figures~ \ref{fig:forecast-example} (d) and (g), \col is almost perfectly predicted by LMA (close to the line $y=x$ and ARIMA is a poor predictor as the points are spread out around $y=x$.
+Comparing to the MASE values in Table~\ref{tab:error} we see that \col with LMA had a MASE of $0.05 \pm 0.002$, whereas \col with ARIMA had a MASE of $0.599 \pm 0.211$.
+So intuitively the better visual prediction received a much better MASE score (the smaller the better).
+One way of interpreting these are that LMA on \col is on average 20 times more effective than random walk and ARIMA on \col is only 2 times more effective than random walk.
 
----works quite well on the trace in Figure~\ref{fig:ipc}, as
-shown in Figure~\ref{fig:lmacol}. 
+The part that becomes tricky and which must be kept in mind is that this error metric is in comparison to random walk.
+If we compare visually for example the LMA prediction of \col and \svdfive (Figures~\ref{fig:forecast-example} (g) \& (i)) visually they look very similar in terms of accuracy; however, \col was forecast 20 times better than random walk where as \svdfive was forecast only 1.5 times better.
+The reason for this lies in the strengths/weaknesses of Random Walk discussed in Section~\ref{sec:simple}.
+The large oscillations in \col cause Random Walk to do poorly, MASE then gives a model that accounts for this oscillation (e.g., LMA) a boost, whereas \svdfive is a great signal for Random Walk so there is less improvement to be made, and less boost is given for example for correctly reacting to regime shift.
+A more in-depth discussion of this is presented in Section~\ref{sec:results}.
+
+%[[and cherry pick a few examples of \gcc and \col to put in the text ]]
+
+%---works quite well on the trace in Figure~%\ref{fig:ipc}, as%
+%shown in Figure~\ref{fig:lmacol}. 
 %
@@
 %
-On the other hand, if we use the same approach to forecast the
-processor efficiency (IPC) of the \gcc time series, the
-prediction is far less accurate; see Figure~\ref{fig:lmagcc}.
-Figure~\ref{fig:lmasvdfive} is clearly a better prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.
+%On the other hand, if we use the same approach to %forecast the
+%processor efficiency (IPC) of the \gcc time series, %the
+%prediction is far less accurate; see Figure~%\ref{fig:lmagcc}.
+%Figure~\ref{fig:lmasvdfive} is clearly a better %prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.
 
-This gets at the utility of the contribution of this paper: These time series all come from the same system---computers---but they are not equally predictable (at this point by LMA).
+%This gets at the utility of the contribution of this %paper: These time series all come from the same system---computers---but they are not equally predictable (at this point by LMA).
 For a practitioner is it the case that
-  Our
-conjecture is that while they come from similar systems \gcc produces processor load traces on the top of the complexity spectrum whereas \col produces processor loads with complexity in the mid to low region of the complexity spectrum.
+%  Our
+%conjecture is that while they come from similar systems \gcc produces processor load traces on the top of the complexity spectrum whereas \col produces processor loads with complexity in the mid to low region of the complexity spectrum.
 If this is the case then \gcc might be much better predicted using a method like Random Walk or \naive which can effectively predict in the presence of complexity.

### 2014-03-10 related.tex
putting it in Phys Rev E Format

@@
-\section{Related Work {\color{red} NOT EDITABLE}}\label{sec:related}
-
+\section{Related Work }\label{sec:related}
+{\color{red}[[NOT EDITABLE]]}
 Attempting to model and characterize predictability is a very old problem which arguably began with Yule in 1927 when he invented AR and many attempts to quantify predicatbility have followed including ....

### 2014-03-10 intro.tex
putting it in Phys Rev E Format
more PRE formatting
more pre formatting
update hopcroft citation

@@
-\section{Introduction {\color{blue} EDITABLE}}\label{sec:intro}
-
+\section{Introduction  }\label{sec:intro}
+{\color{blue}[[EDITABLE]]}
 %****NOTES START HERE****
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/colshortts}
+    \includegraphics[width=\columnwidth]{figs/colshortts}
     % where an .eps filename suffix will be assumed under latex,
@@
 at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem~\cite{hopcroft} which essentially states that any non-trivial property of a program is uncomputable.
+not violate the Rice-Shapiro theorem~\cite{hopcroft2007} which essentially states that any non-trivial property of a program is uncomputable.
 

### 2014-03-10 wpe.tex
putting it in Phys Rev E Format
more PRE formatting
more pre formatting
first chunk of wpe section edit

--- 2014-03-04 wpe.tex
@@
-\section{Measuring Structural Complexity {\color{red} NOT EDITABLE}}\label{sec:meaComplex}
+\section{Measuring Structural Complexity }\label{sec:meaComplex}
+{\color{red} NOT EDITABLE}
 
@@
 
-For the purposes of this paper, one can view entropy as a measure of complexity
-and predictability in a time series.  A high-entropy time series is almost
-completely unpredictable---and conversely.  This can be made more rigorous:
-Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
-Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
-$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
-states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
-e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
-difficult prediction becomes.
+For the purposes of this paper, one can view entropy as a measure of complexity and unpredictability in a time series.
+A high-entropy time series is almost completely unpredictable---and conversely a low entropy signal is often quite predictable.
+This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Shannon entropy rate is equal to the sum of the positive Lyapunov exponents, $\lambda_i$.
+The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.
+The faster the divergence, the more difficult prediction becomes.
 
-Utilizing entropy as a measure of temporal complexity is by no means a new idea
-\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
-categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
-\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
-effectively real-valued.  To get around this, one must discretize the data---
-typically achieved by binning.  Unfortunately, this is rarely a good solution to
-the problem, as the binning of the values introduces an additional dynamic on
-top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
-dynamics studies how to discretize a time series in such a way that the
-intrinsic behavior is not perverted, but these methods are fragile in the face
-of noise and require further understanding of the underlying system, which
-defeats the purpose of measuring the entropy in the first place.
+Further we can consider the information contained in a single observable, say the \emph{present}, of the system, measured by the entropy.
+This information can be partitioned in to two components: the information shared with past observations --- the mutual information between the past and present --- and the information in the present which is not contained in the past --- the conditional entropy.
+The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the \emph{entropy rate}.
+In theory, the more redundancy in a signal the more predictable it is.
+The difference between prediction methods is the types of redundancy (linear, non-linear, etc) they can detect and make use of.
+
+Utilizing entropy as a measure of temporal complexity is by no means a new idea \cite{Shannon1951, mantegna1994linguistic}.
+Its effective usage requires categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite \emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is effectively\footnote{Measurements from real devices are of finite precision, rendering them discrete.
+Unfortunately the space of values taken is so large that for the purposes of entropy calculations it can be taken as a continuum.} real-valued.
+To get around this, one must discretize the data --- typically achieved by binning.
+Unfortunately, this is rarely a good solution to the problem, as the binning of the values introduces an additional dynamic on top of the intrinsic dynamics whose entropy is desired\cite{bollt2001}.
+The field of symbolic dynamics studies how to discretize a time series in such a way that the intrinsic behavior is not perverted, but these methods are fragile in the face of noise and require further understanding of the underlying system, which defeats the purpose of measuring the entropy in the first place.
 
@@
 
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
-  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
-  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
-  = 1,\dots,T}$:
+  Given a time series $\{x_t\}_{t = 1,\dots,T}$.
+Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$.
+For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$:
   \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
+    p(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
   \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
-  \ge 2$ is defined as
+  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $\ell \ge 2$ is defined as
   \begin{align*}
-  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
+    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p(\pi) \log_2 p(\pi)
   \end{align*}
@@
 
-Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
-it is common in the literature to normalize permutation entropy as follows:
-$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
-and ``high'' entropy is close to 1. Finally, it should be noted that the
-permutation entropy has been shown to be identical to the Shannon entropy for
-many large classes of systems \cite{amigo2012permutation}.
+Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
+With this in mind, it is common in the literature to normalize permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.
+With this convention, ``low'' entropy is close to 0 and ``high'' entropy is close to 1.
+Finally, it should be noted that the permutation entropy has been shown to be identical to the Kolmolgorov-Sinai (KS) entropy for many large classes of systems \cite{amigo2012permutation}.
+In turn, and under very loose conditions, the KS entropy is equal to the Shannon entropy rate.
+What this normalization allows us to do, then, is approximate the redundancy of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
 
@@
 %when she scrolls past p12
-    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
+    \includegraphics[width=\columnwidth]{figs/SVD_wwpe}
     %\caption{The MASE of ARIMA vs weighted permutation entropy. }
@@
   regimes. The window size used is 5,000 $\times$ 100,000 instructions
-  and the word length is $6$. {\color{red} make sure this is actually updated to 6} For reference the instructions per cycle
+  and the word length is $6$. For reference the instructions per cycle
   of \svd are plotted as a ghost behind this plot. Each color on the

### 2014-03-10 wpe.tex
more updates

@@
 
-Here we will be utilizing a variation of the permutation entropy, the
-\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
-permutation entropy attempts to correct for observational noise which is larger
-than some trends in the data, but smaller than the larger scale features --- for
-example, a signal that switches between two fixed points with noise about those
-fixed points. The weighted permutation entropy would be dominated by the
-switching rather than by the stochastic fluctuation. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
+Here we will be utilizing a variation of the permutation entropy, the \emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}.
+The weighted permutation entropy attempts to correct for observational noise which is larger than some trends in the data, but smaller than the larger scale features --- for example, a signal that switches between two fixed points with noise about those fixed points.
+The weighted permutation entropy would be dominated by the switching rather than by the stochastic fluctuation.
+To accomplish this, the \emph{weight} of a permutation is taken into account:
 \begin{align*}
-  w(x_{t+1:t+n}) = \frac{1}{n}
-                 \sum_{x_i \in \{x_{t+1:t+n}\}}
-                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
+  w(x_{t+1:t+\ell}) = \frac{1}{\ell} \sum_{x_i \in \{x_{t+1:t+\ell}\}}
+                      \left( x_i - \bar{x}_{t+1:t+\ell} \right)^2
 \end{align*}
-where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
-$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.
+where $x_{t+1:t+\ell}$ is a sequence of values $x_{t+1}, \ldots, x_{t+\ell}$, and $\bar{x}_{t+1:t+\ell}$ is the arithmetic mean of those values.
 
@@
 \begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
+  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
 \end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
-probability enhances permutations involved in ``large'' features and demotes
-permutations which are small in amplitude relative to the features of the time
-series. The weighted permutation entropy is then:
+where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise.
+Effectively, this weighted probability enhances permutations involved in ``large'' features and demotes permutations which are small in amplitude relative to the features of the time series.
+The weighted permutation entropy is then:
 \begin{align*}
-  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
+  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p_w(\pi) \log_2 p_w(\pi),
 \end{align*}
-which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
-results of this paper.
+which can also be normalized by dividing by $\log_2(\ell!)$, and will be in all the results of this paper.
 
-In practice, calculating permutation entropy and weighted permutation entropy
-involves choosing a good value for the word length $n$. The primary
-consideration is that the value be large enough that forbidden ordinals are
-discovered, yet small enough that reasonable statistics over the ordinals are
-gathered: e.g.:
+In practice, calculating permutation entropy and weighted permutation entropy involves choosing a good value for the word length $\ell$.
+The primary consideration is that the value be large enough that forbidden ordinals are discovered, yet small enough that reasonable statistics over the ordinals are gathered: e.g.:
 \begin{align*}
-  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
+  \ell = \operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \},
 \end{align*}
-assuming an average of 100 counts per ordinal is sufficient. In the literature,
-$3 \le n \le 6$ is a standard choice --- generally without any formal
-justification. In theory, the permutation entropy should reach an asymptote with
-increasing $n$, but that requires an arbitrarily long time series. In practice,
-what one should do is calculate the \emph{persistent} permutation entropy by
-increasing $n$ until the result converges, but data length issues can intrude
-before that convergence is reached.
+assuming an average of 100 counts per ordinal is sufficient.
+In the literature, $3 \le \ell \le 6$ is a standard choice --- generally without any formal justification.
+In theory, the permutation entropy should reach an asymptote with increasing $\ell$, but that requires an arbitrarily long time series.
+In practice, what one should do is calculate the \emph{persistent} permutation entropy by increasing $\ell$ until the result converges, but data length issues can intrude before that convergence is reached.
+Here, we found that $\ell = 6$ was a good choice, because with $\ell \geq 7$ many signals had entropies plummet due to finite data effects and so with a word length of 6 we are capturing as much structure as possible without sacraficing the accuracy of ordinal statistics.
 

### 2014-03-10 intro.tex
adding Liz's edits to intro

--- 2014-03-06 intro.tex
@@
 
-Complicated time series are ubiquitous in modern scientific research.
-Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}.
-On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model.
-In particular, there is a an underlying process that generates and transmits information from the past to the future in a perfectly predictable fashion.
-Some examples of this are constant or low-period signals.
-On the opposite end of this spectrum are signals which are ``fully complex", i.e., the underlying generating process creates information so rapidly that no information at all is transmitted from the past to the future.
-Some examples, of this are white noise or random walk processes.
-With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.
-In the middle of this spectrum are where complicated \emph{and} interesting---from a modeling perspective---signals exist, e.g., observations of a deterministic chaotic trajectory, high period signals with some level of noise on top, etc.
-With time series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model the future of the observed system could be forecast with high accuracy.
+Complicated time-series data are ubiquitous in modern scientific
+research.  The complexity of these data spans a wide range.  On the
+low end of this spectrum are time series that exhibit perfect
+predictive structure, i.e, signals whose future values can be
+successfully predicted from past values.  In situations like this,
+there is an underlying process that generates information and/or
+transmits it from the past to the future in a perfectly predictable
+fashion.  Constant or periodic signals, for example, fall in this
+class.  On the opposite end of this spectrum are signals that are what
+we call \emph{fully complex}, where the underlying generating process
+transmits no information at all from the past to the future.  Some
+examples of this are white noise or random-walk processes.  In signals
+like this, knowledge of the past gives no insight into the future,
+\emph{regardless} of what model one chooses to use.  Signals in the
+midrange of this spectrum---e.g., deterministic chaos, high-period
+signals [[Liz question: why does the rank of the periodicity force
+    them off the low end of the spectrum?  Any periodic signal can be
+    decomposed into sine waves, which can be predicted perfectly...]]
+with some additive noise---pose interesting challenges from a modeling
+perspective.  In these signals, enough information is being
+transmitted from the past to the future that an \emph{ideal}
+model---one that captures the generating process---can forecast the
+future behavior of the observed system with high accuracy.  [[Liz
+    question: if there's noise, is that true?]]
 
-Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series.
-This naturally leads to an important and hard question: given a complicated real-valued time series, does their exist a forecast model for that time series that can utilize the information being transmitted from past to future (if such information exists)?
-A first step in answering this question is to answer: is it possible to reliably quantify where on the spectrum of complexity the time series lies?
-i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast?
-Since the information in an observation can be partitioned into two pieces --- redundancy and entropy generation\cite{crutchfield2003} --- our goal is to analyze how the amount of predictive structure present in a signal, as measured by the redundancy, relates to the performance of prediction strategies, as well as what is meant by \emph{low, medium and high} complexity in Section \ref{sec:meaComplex}.
+Unfortunately, as a corollary of the undecidability of the halting
+problem \cite{halting-problem}, there cannot exist \emph{one}
+[[universally]] [[across {\sl all} time series?  this seems pretty
+    obvious...]] ideal forecasting scheme for even completely
+noise-free deterministic time series\cite{weigend-book}---let alone an
+arbitrary real-world time series.  This naturally leads to an
+important and challenging question: given a complicated real-valued
+time series, does there exist a forecast model that can utilize the
+information (if any) that is being transmitted from past to future by
+the underlying generating process?  A first step in answering this
+question is to reliably quantify where on the complexity spectrum a
+given time series falls; a second step is to determine how complexity
+and predictability are related.  These are the goals of this paper.
 
-To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant.
-We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
-We argue that an extension of \emph{permutation entropy}
-\cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture.
-In light of this, for the purposes of this paper we define \emph{complexity} as an approximation of Shannon \emph{entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}.
-In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity.
-The permutation entropy is ideal in a situation like this as it converges on the true entropy value, unlike other partitioning techniques used in the past which either require specific knowledge of the generating process or produce biased values of the entropy~\cite{bollt2001}.
+In particular, we wish to determine---without knowing anything about
+the generating process---whether a time series is too complex to
+predict.  An important practical corollary to this is assessment of
+forecast methods: if the forecast produced by a particular method is
+poor but the time series contains a significant amount of predictive
+structure, the method is inadequate and one should seek another
+method.
+
+We focus on real-valued, scalar, time-series data that appear
+\emph{complicated}: i.e., not periodic, linear, constant, etc.  We
+make no assumptions about the properties of the generating process:
+whether it is linear, nonlinear, deterministic, stochastic,
+stationary, non-stationary, etc.  The information in an observation
+can be partitioned into two pieces: redundancy and entropy
+generation\cite{crutchfield2003}.  Our approach exploits this
+decomposition in order to assess how much predictive structure is
+present in a signal---i.e., where it falls on the complexity spectrum
+mentioned above.  We define \emph{complexity} as a particular
+approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.  That is,
+we view a random-walk time series (which exhibits high entropy) as
+purely complex, whereas a periodic signal, which exhibits low entropy,
+is on the low end of the complexity spectrum.  We argue that an
+extension of \emph{permutation entropy} \cite{bandt2002per}---a method
+for approximating the entropy through ordinal analysis---is an
+effective way to assess the complexity of a given time series.
+Permutation entropy is ideal for our purposes because it is known to
+converge to the true entropy value, unlike other existing techniques,
+which either require specific knowledge of the generating process or
+produce biased values of the entropy~\cite{bollt2001}.
+
+To explore the relationship between complexity, predictive structure,
+and predictability, we generate forecasts of a number of time-series
+datasets using four different prediction methods, then compare the
+accuracy of those predictions to the permutation entropy of the
+associated signals.  This results in two primary findings:
+\begin{enumerate}
+\item The complexity of a noisy real-valued time series is
+  quantifiable by permutation entropy and correlated with prediction
+  accuracy of an ideal predictor [[we don't know that.  we only tried
+      four methods.]]
+\item The way information is generated and processed internally by a
+  system plays a crucial role in the success of different forecasting
+  schema.
+\end{enumerate}
+The forecast methods used here were chosen to span the space of
+standard prediction strategies, but they do not cover that space
+exhaustively.  Our goal here is an empirical assessment of the
+relationship between predictability and complexity, not formal results
+about a ``best'' predictor for a given time series.  From a
+practitioner's standpoint, it would be useful to know {\sl a priori}
+if a trace contains enough predictive structure to make it worth
+spending the time and effort searching for a good forecast method.  It
+would also be useful to know if a given method is inadequate---that
+is, if one could do better and should keep searching.
 
@@
 
-To explore this conjecture we require a test bed for generating a broad array of time series which cover the full spectrum of complexity.
-We argue that the ideal test bed for this purpose are computer performance traces, e.g., the processor efficiency during the execution of a particular program.
-Computers are among the most complex engineered artifacts in current
-use.  Modern microprocessor chips contain multiple processing units
-and multi-layer memories, for instance, and they use complicated
-hardware/software strategies to move data and threads of computation
-across those resources.  These features---along with all the others
-that go into the design of these chips---make the patterns of their
-processor efficiency highly complicated.
-%Accurate forecasts of these quantities, if one could
-%construct them, could be used to improve computer design.
-% If one
-%could predict that a particular computational thread %would be bogged
-%down for the next 0.6 seconds waiting for data from main memory, for
-%instance, one could save power by putting that thread %on hold for that
-%time period (e.g., by migrating it to a processing %unit whose clock
-%speed is scaled back).
-
-As a result computer performance traces are very
-complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel," like a three-line loop that
-repeatedly initializes the upper triangle of a matrix in column-major order, can produce everything from periodic to
-chaotic performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
-Figure~\ref{fig:ipc}.%, and chaos places fundamental limits on predictability.
+For the purposes of this study, we require a broad array of
+time-series datasets that cover the full complexity spectrum.  Rather
+than use synthetic data, we study sensor data from computer
+performance experiments.  Computers are among the most complex
+engineered artifacts in current use.  Modern microprocessor chips
+contain multiple processing units and multi-layer memories, for
+instance, and they use complicated hardware/software strategies to
+move data and threads of computation across those resources.  As a
+result, the processor and memory loads during the execution of even a
+very simple program can evolve in a very complicated fashion.
+Figure~\ref{fig:ipc} shows a performance trace of a three-line loop
+that repeatedly initializes the upper triangle of a matrix in
+column-major order.
 %
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/colshortts}
+    \includegraphics[width=\columnwidth]{figs/colshortts}
     % where an .eps filename suffix will be assumed under latex,
     % and a .pdf suffix will be assumed for pdflatex
-    \caption{A small snippet of the instructions per cycle(ipc) of \col, a three-line C program that repeatedly initializes
-      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.}
+    \caption{A computer performance trace: the processor load of \col,
+      a three-line C program that repeatedly initializes a matrix in
+      column-major order, running on an Intel
+      i7\textsuperscript{\textregistered}-based machine.}
    \label{fig:ipc}
@@
 %
-More interesting programs, e.g., compilers and numerical solvers, produce time series which exist throughout the spectrum of complexity: from completely predictable to completely unstructured,i.e., low to high complexity[[Just trying to reinforce the wording in the readers mind]].
-In addition, some performance traces even exhibit interesting regime changes as a program switches between varying subroutines.
-With a single system (computer performance) capable of producing complicated time series that vary so greatly in complexity (periodic to noise), it would be invaluable to known a priori if a trace contained enough predictive structure to build a  forecast model based on structure before time and effort was spent searching for an ideal model.
+Although this time series appears to be periodic, it is actually
+chaotic \cite{mytkowicz09}, which places fundamental limits on
+predictability.  Radical changes in this behavior can result if the
+program is changed slightly.  The resulting traces span the whole
+range of the complexity spectrum, from completely predictable to
+completely unstructured, making this an ideal testbed for this study.
 
@@
 
-This paper provides insight into when, why, and how forecast strategies fail when they are applied to complicated time series by empirically estimating their inherent complexity.
-[[Liz, do you think here would be better for the sentence about we are not the first, or is above where I asked Ryan to insert it ok?
-I worry it would muddy the water of this synopsis paragraph.]] We focus here on the specific example of
-computer performance but believe the results apply to a much broader class of time series. We conjecture that the complexity of traces
-from these systems---which results from the inherent dimension,
-non-linearity, and non-stationarity, as well as from
-measurement issues like noise, aggregation, and finite data
-length---is \emph{empirically quantifiable}.
-To validate this conjecture we model 120 different performance traces using a variety of forecasting models which we introduce in Section \ref{sec:model}.
-We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method.
-Which results in two primary findings:
-
-\begin{enumerate}
-
-\item The complexity of a noisy real-valued time series is quantifiable and correlated with prediction accuracy of an ideal predictor.
-
-\item The way information is processed internally by a given process plays a crucial role in the success of different forecasting schema.
-
-\end{enumerate}
-
 %We argue that \emph{permutation entropy}
@@
 
+The rest of the paper is organized as follows:
+Section~\ref{sec:related} discusses previous results on [[what]] and
+[[what]] and situates this work in that
+context. Section~\ref{sec:methods} covers the experimental setup and
+methods used to collect each time series. Section~\ref{sec:model}
+describes the prediction models we implemented.  In
+Section~\ref{sec:meaComplex}, we review weighted permutation entropy.
+In Section~\ref{sec:results}, we estimate the complexity of each time
+series and compare that complexity to the accuracy of predictions
+produced by the methods of Section~\ref{sec:model} on that time
+series.  In Section~\ref{sec:conc}, we discuss these results and their
+implications and consider future areas of research.
+
+[[Following belongs in the methods section:]] In addition, some
+performance traces even exhibit interesting regime changes as a
+program switches between varying subroutines.
+
+[[Following belongs in the conclusion.  Putting it here would be
+    appropriate in a CS paper, but not in a physics journal, and not
+    when we've just made the point that this is just a testbed.]] 
 % paragraph to appease the theoretician in me
@@
 are attempting is to predict aspects or functions of the running of
-the CPU, e.g., processor efficiency and similar statistics. Prediction of these quantities
-at some finite time in the future, even with perfect accuracy, does
-not violate the Rice-Shapiro theorem~\cite{hopcroft} which essentially states that any non-trivial property of a program is uncomputable.
+the CPU, e.g., processor efficiency and similar statistics. Prediction
+of these quantities at some finite time in the future, even with
+perfect accuracy, does not violate the Rice-Shapiro
+theorem~\cite{hopcroft} which essentially states that any non-trivial
+property of a program is uncomputable.
 
- The rest of the paper is organized as follows: Section~\ref{sec:related} discusses similar methods and how this method is different.
-Section~\ref{sec:methods} explains the experimental setup and explains in depth the programs observed to generate each time series.
-Section~\ref{sec:model} describes the four forecast models we implemented.
-In
- Section~\ref{sec:meaComplex}, we review weighted permutation entropy. In Section~\ref{sec:results} we estimate the complexity of each time series, and
- compare the complexity to the prediction accuracy of several different forecast schema.  In
- Section~\ref{sec:conc}, we discuss these results and their
- implications in regard to our conjecture, and consider future areas of
- research.
+[[Following is a good start at an abstract. It doesn't need to be
+    reiterated in the introduction.]]  This paper provides insight
+into when, why, and how forecast strategies fail when they are applied
+to complicated time series by empirically estimating their inherent
+complexity. [[Liz, do you think here would be better for the sentence
+    about we are not the first, or is above where I asked Ryan to
+    insert it ok? I worry it would muddy the water of this synopsis
+    paragraph.]] We focus here on the specific example of computer
+performance but believe the results apply to a much broader class of
+time series. We conjecture that the complexity of traces from these
+systems---which results from the inherent dimension, non-linearity,
+and non-stationarity, as well as from measurement issues like noise,
+aggregation, and finite data length---is \emph{empirically
+  quantifiable}.  To validate this conjecture we model 120 different
+performance traces using a variety of forecasting models.  We then
+compare the empirically estimated \emph{complexity} with the accuracy
+of each forecast method.

### 2014-03-10 intro.tex
fix a sentense

@@
 Permutation entropy is ideal for our purposes because it is known to
-converge to the true entropy value, unlike other existing techniques,
-which either require specific knowledge of the generating process or
-produce biased values of the entropy~\cite{bollt2001}.
+converge to the true entropy value. Other existing techniques either require specific knowledge of the generating process orproduce biased values of the entropy~\cite{bollt2001}.
 
@@
 perfect accuracy, does not violate the Rice-Shapiro
-theorem~\cite{hopcroft} which essentially states that any non-trivial
+theorem~\cite{hopcroft2007} which essentially states that any non-trivial
 property of a program is uncomputable.

### 2014-03-10 intro.tex
ideal -> appropriate

@@
   quantifiable by permutation entropy and correlated with prediction
-  accuracy of an ideal predictor [[we don't know that.  we only tried
-      four methods.]]
+  accuracy of an appropriate predictor.
 \item The way information is generated and processed internally by a
   system plays a crucial role in the success of different forecasting
-  schema.
+  schema, and in the choice of which forcasting scheme is appropriate.
 \end{enumerate}

### 2014-03-10 intro.tex
more tweaks

@@
 The rest of the paper is organized as follows:
-Section~\ref{sec:related} discusses previous results on [[what]] and
-[[what]] and situates this work in that
+Section~\ref{sec:related} discusses previous results on generating partitions, local modeling, and error distribution analysis, and situates this work in that
 context. Section~\ref{sec:methods} covers the experimental setup and

### 2014-03-10 intro.tex
words about 2 conjecture

--- 2014-03-10 intro.tex
@@
 Permutation entropy is ideal for our purposes because it is known to
-converge to the true entropy value. Other existing techniques either require specific knowledge of the generating process orproduce biased values of the entropy~\cite{bollt2001}.
+converge to the true entropy value, unlike other existing techniques,
+which either require specific knowledge of the generating process or
+produce biased values of the entropy~\cite{bollt2001}.
 
@@
 \end{enumerate}
+
+The second conjecture isn't just the corollary of halting problem in
+disguise.  It gives insight into why there's not an ideal predictor.
+It can't exist because the info is being transmitted in different
+ways.
+
 The forecast methods used here were chosen to span the space of
@@
 perfect accuracy, does not violate the Rice-Shapiro
-theorem~\cite{hopcroft2007} which essentially states that any non-trivial
+theorem~\cite{hopcroft} which essentially states that any non-trivial
 property of a program is uncomputable.

### 2014-03-10 intro.tex
new abstract

--- 2014-03-10 intro.tex
@@
 
-[[Following is a good start at an abstract. It doesn't need to be
-    reiterated in the introduction.]]  This paper provides insight
-into when, why, and how forecast strategies fail when they are applied
-to complicated time series by empirically estimating their inherent
-complexity. [[Liz, do you think here would be better for the sentence
-    about we are not the first, or is above where I asked Ryan to
-    insert it ok? I worry it would muddy the water of this synopsis
-    paragraph.]] We focus here on the specific example of computer
-performance but believe the results apply to a much broader class of
-time series. We conjecture that the complexity of traces from these
-systems---which results from the inherent dimension, non-linearity,
-and non-stationarity, as well as from measurement issues like noise,
-aggregation, and finite data length---is \emph{empirically
-  quantifiable}.  To validate this conjecture we model 120 different
-performance traces using a variety of forecasting models.  We then
-compare the empirically estimated \emph{complexity} with the accuracy
-of each forecast method.
+%[[Following is a good start at an abstract. It doesn't %need to be
+%    reiterated in the introduction.]]

### 2014-03-10 pecomplexity.tex
update macros to use xspace
new abstract

@@
 \usepackage{amsthm}
-
+\usepackage{xspace}
 \usepackage{pifont} % check and x marks
@@
 
-\newcommand{\gcc}{{\tt 403.gcc}~}
-\newcommand{\svd}{{\tt dgesdd}~}
-\newcommand{\svdone}{{\tt dgesdd}$_1$~}
-\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
-\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
-\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
-\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
-\newcommand{\svdsix}{{\tt dgesdd$_6$}~}
+\newcommand{\gcc}{{\tt 403.gcc}\xspace}
+\newcommand{\svd}{{\tt dgesdd}\xspace}
+\newcommand{\svdone}{{\tt dgesdd}$_1$\xspace}
+\newcommand{\svdtwo}{{\tt dgesdd$_2$}\xspace}
+\newcommand{\svdthree}{{\tt dgesdd$_3$}\xspace}
+\newcommand{\svdfour}{{\tt dgesdd$_4$}\xspace}
+\newcommand{\svdfive}{{\tt dgesdd$_5$}\xspace}
+\newcommand{\svdsix}{{\tt dgesdd$_6$}\xspace}
 \newcommand{\naive}{na\"ive}
-\newcommand{\row}{{\tt row\_major}~}
-\newcommand{\col}{{\tt col\_major}~}
+\newcommand{\row}{{\tt row\_major}\xspace}
+\newcommand{\col}{{\tt col\_major}\xspace}
 %\newcommand{\color{red}}{{\color[rgb]{1,0,0}}}
@@
 \begin{abstract}
-{\color{green}From the site:It should be about 5\% of the length of the article, but less than 500 words.}[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus, should be done after we have a first draft of the paper]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
-  Among other things, that implies that one should be able to use
-  deterministic forecast rules to predict aspects of their behavior.
-  That statement is sometimes---but not always---true. The memory and
-  processor loads of some simple programs are easy to predict, for
-  example, but those of more-complex programs like \gcc are not.
-%%%%%%%%%%%%%%%%%%%%%%
-%% I had to change all the \verb|blah| entries because they caused
-%% latex to barf if they were in figure captions.  Odd bug.
-%%%%%%%%%%%%%%%%%%%%%%
-  The goal of this paper is to determine why that is the case. We
-  conjecture that, in practice, complexity can effectively overwhelm
-  the predictive power of deterministic forecast models. To explore
-  that, we build models of a number of performance traces from
-  different programs running on different Intel-based computers. We
-  then calculate the \emph{permutation entropy}---a temporal entropy
-  metric that uses ordinal analysis---of those traces and correlate
-  those values against the prediction success.
+{\color{green}From the site:It should be about 5\% of the length of the article, but less than 500 words.} This paper provides insight
+into when, why, and how forecast strategies fail when they are applied
+to complicated time series by empirically estimating their inherent
+complexity. We focus here on the specific example of computer
+performance but believe the results apply to a much broader class of
+time series. We conjecture that the complexity of traces from these
+systems---which results from the inherent dimension, non-linearity,
+and non-stationarity, as well as from measurement issues like noise,
+aggregation, and finite data length---is \emph{empirically
+  quantifiable}.  To validate this conjecture we model 120 different
+performance traces using a variety of forecasting models.  We then
+compare the empirically estimated \emph{complexity} with the accuracy
+of each forecast method.
 \begin{description}

### 2014-03-10 wpe.tex
removing windows

@@
 
-\subsection{Regime Selection:Detecting complexity shift in a time series}\label{sec:wpeRegime}
-
-{\color{red} this is from a previous paper I wrote, if this is used we need to update the plots to use wpe and windowed wpe instead of standard pe, also not sure how much of this is necessary but added as starting point}
-
-Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}.
+%\subsection{Regime Selection:Detecting complexity shift in a time series}\label{sec:wpeRegime}
+%
+%{\color{red} this is from a previous paper I wrote, if this is used we need to update the plots to use wpe and windowed wpe instead of standard pe, also not sure how much of this is necessary but added as starting point}
+%
+%Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}.
 To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}.
@@
 According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
-
-In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy.
+%
+%In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy.
 This is identical to the lag used in delay coordinate embedding and often the same language is used.
@@
 The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
-
-\subsubsection{Example: Concatenated Time Series}\label{sec:concatenate}
-As an example of this detection, consider Figure \ref{fig:ts-banded} (a).
+%
+%\subsubsection{Example: Concatenated Time Series}\label{sec:concatenate}
+%As an example of this detection, consider Figure \ref{fig:ts-banded} (a).
 This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise.
@@
 The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
-
-The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$).
+%
+%The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$).
 Visually the bands of noise and the chaotic trajectory are indistinguishable.
@@
 If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
-
-In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves).
+%
+%In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves).
 The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively).
@@
 This is in fact the case when comparing to the way the time series was constructed.
-
-The main advantage we can see to the use of lags other than 1 is for analysis of periodicity.
+%
+%The main advantage we can see to the use of lags other than 1 is for analysis of periodicity.
 This is also the major advantage listed in \cite{cao2004det}.
@@
 If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
-
-%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
-
-%\begin{figure}
-%  \centering
-%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-alone}}
-%    {(a)}
 %
-%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-pe}}
-%  {(b)}
-%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}.
+%%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
+%
+%%\begin{figure}
+%%  \centering
+%%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-alone}}
+%%    {(a)}
+%%
+%%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-pe}}
+%%  {(b)}
+%%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}.
 (b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves).
 Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
-%\end{figure}
-
-and what is it good for with forward pointer to the other sections
-
-\subsubsection{Example: The transient Logistic map}
-In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed.
+%%\end{figure}
+%
+%
+%
+%
+%
+%
+%and what is it good for with forward pointer to the other sections
+%
+%
+%
+%\subsubsection{Example: The transient Logistic map}
+%In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed.
 As we showed, permutation was a great proxy to detect this change.
@@
 The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
-
- As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well.
+%
+% As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well.
 An interesting feature of block calculations are the spikes that occur at bifurcation points.
@@
 This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
-
- This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\columnwidth]{unused-figs/tstranslog-perm}
-  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
-  \label{fig:translogistic}
-\end{figure}
-%\begin{enumerate}
-%\item do with henon bifurcation. what happens near the broken chaotic window
-
-%\end{enumerate}
-
-The weighted permutation entropy for the \svd program is given in
-Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
-the time series. Within each of those windows, the statistics over words of
-length 4 are computed and the WPE is calculated. The gray bands denote regions
-where the 5,000 value window overlapped visually-distinct regimes. It can be
-seen that the behaviors of the weighted permutation entropy vary between
-regimes.
+%
+% This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
+%
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics[width=\columnwidth]{unused-figs/tstranslog-perm}
+%  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
+%  \label{fig:translogistic}
+%\end{figure}
+%%\begin{enumerate}
+%%\item do with henon bifurcation. what happens near the broken chaotic window
+%
+%%\end{enumerate}
+%
+%
+%The weighted permutation entropy for the \svd program is given in
+%Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
+%the time series. Within each of those windows, the statistics over words of
+%length 4 are computed and the WPE is calculated. The gray bands denote regions
+%where the 5,000 value window overlapped visually-distinct regimes. It can be
+%seen that the behaviors of the weighted permutation entropy vary between
+%regimes.
 [[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
 Maybe point at some of the predicting predictability papers.]]
-
+%
+%%\begin{figure}[htbp]
+%%  \centering
+%%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
+%%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
+%%    are regions where the window overlaps regimes. The window size used is
+%%    $5,000 \times 100,000$ instructions and the word length is $4$.}
+%%  \label{fig:wwpe}
+%%\end{figure}
+%
+%
+%%%FIGURE INTENT
+%%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
+%%%%%%%%%%%%%%%%%%
 %\begin{figure}[htbp]
 %  \centering
-%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
-%    are regions where the window overlaps regimes. The window size used is
-%    $5,000 \times 100,000$ instructions and the word length is $4$.}
-%  \label{fig:wwpe}
-%\end{figure}
-
-%%FIGURE INTENT
-%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
-%%%%%%%%%%%%%%%%%
-\begin{figure}[htbp]
-  \centering
-  %\begin{subfigure}{0.3\textwidth}
-  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
-  %  \caption{The instructions per cycle of \svd.
+%  %\begin{subfigure}{0.3\textwidth}
+%  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
+%  %  \caption{The instructions per cycle of \svd.
 Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}.
@@
 }
- %   \label{fig:svd_ts}
- % \end{subfigure}%
-  %\\
- % \begin{subfigure}{\textwidth}
-
-%Liz commented this image out temporarily so that her mac doesn't hang
-%when she scrolls past p12
-    \includegraphics[width=\columnwidth]{figs/SVD_wwpe}
-    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-  \caption{The weighted permutation entropy of one run of
-  SVD. The gray bands are regions where the window overlaps
-  regimes. The window size used is 5,000 $\times$ 100,000 instructions
-  and the word length is $6$. For reference the instructions per cycle
-  of \svd are plotted as a ghost behind this plot. Each color on the
-  ghosted time series corresonds to the different regimes as selected
-  by rapid shifts in WPE. From left to right each change in color
-  represents a change in regime for 6 regimes in
-  total.}\label{fig:wwpe}
-\end{figure}
+% %   \label{fig:svd_ts}
+% % \end{subfigure}%
+%  %\\
+% % \begin{subfigure}{\textwidth}
+%
+%%Liz commented this image out temporarily so that her mac doesn't hang
+%%when she scrolls past p12
+%    \includegraphics[width=\columnwidth]{figs/SVD_wwpe}
+%    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
+%  \caption{The weighted permutation entropy of one run of
+%  SVD. The gray bands are regions where the window overlaps
+%  regimes. The window size used is 5,000 $\times$ 100,000 instructions
+%  and the word length is $6$. For reference the instructions per cycle
+%  of \svd are plotted as a ghost behind this plot. Each color on the
+%  ghosted time series corresonds to the different regimes as selected
+%  by rapid shifts in WPE. From left to right each change in color
+%  represents a change in regime for 6 regimes in
+%  total.}\label{fig:wwpe}
+%\end{figure}
+%

### 2014-03-10 intro.tex
Liz is now done with intro

--- 2014-03-10 intro.tex
@@
 we call \emph{fully complex}, where the underlying generating process
-transmits no information at all from the past to the future.  Some
-examples of this are white noise or random-walk processes.  In signals
-like this, knowledge of the past gives no insight into the future,
+transmits no information at all from the past to the future.  White
+noise and random-walk processes fall in this class.  In fully complex
+signals, knowledge of the past gives no insight into the future,
 \emph{regardless} of what model one chooses to use.  Signals in the
-midrange of this spectrum---e.g., deterministic chaos, high-period
-signals [[Liz question: why does the rank of the periodicity force
-    them off the low end of the spectrum?  Any periodic signal can be
-    decomposed into sine waves, which can be predicted perfectly...]]
-with some additive noise---pose interesting challenges from a modeling
-perspective.  In these signals, enough information is being
-transmitted from the past to the future that an \emph{ideal}
-model---one that captures the generating process---can forecast the
-future behavior of the observed system with high accuracy.  [[Liz
-    question: if there's noise, is that true?]]
+midrange of this spectrum---e.g., deterministic chaos---pose
+interesting challenges from a modeling perspective.  In these signals,
+enough information is being transmitted from the past to the future
+that an \emph{ideal} model---one that captures the generating
+process---can forecast the future behavior of the observed system with
+high accuracy.
 
-Unfortunately, as a corollary of the undecidability of the halting
-problem \cite{halting-problem}, there cannot exist \emph{one}
-[[universally]] [[across {\sl all} time series?  this seems pretty
-    obvious...]] ideal forecasting scheme for even completely
-noise-free deterministic time series\cite{weigend-book}---let alone an
-arbitrary real-world time series.  This naturally leads to an
-important and challenging question: given a complicated real-valued
+As a corollary of the undecidability of the halting problem
+\cite{halting-problem}, however, no single forecasting schema is ideal
+for all noise-free deterministic signals\cite{weigend-book}---let
+alone all real-world time series data sets.  This leads naturally to
+an important and challenging question: given a complicated real-valued
 time series, does there exist a forecast model that can utilize the
@@
 the generating process---whether a time series is too complex to
-predict.  An important practical corollary to this is assessment of
-forecast methods: if the forecast produced by a particular method is
-poor but the time series contains a significant amount of predictive
-structure, the method is inadequate and one should seek another
-method.
+predict.  An important practical corollary to this is a strategy for
+assessing appropriateness of forecast methods.  If the forecast
+produced by a particular method is poor but the time series contains a
+significant amount of predictive structure, that method is inadequate
+to the task and one should seek another method.  We focus on
+real-valued, scalar, time-series data that appear complicated: i.e.,
+not periodic, linear, constant, etc.  We make no assumptions about the
+properties of the generating process: whether it is linear, nonlinear,
+deterministic, stochastic, stationary, non-stationary, etc.
 
-We focus on real-valued, scalar, time-series data that appear
-\emph{complicated}: i.e., not periodic, linear, constant, etc.  We
-make no assumptions about the properties of the generating process:
-whether it is linear, nonlinear, deterministic, stochastic,
-stationary, non-stationary, etc.  The information in an observation
-can be partitioned into two pieces: redundancy and entropy
-generation\cite{crutchfield2003}.  Our approach exploits this
-decomposition in order to assess how much predictive structure is
-present in a signal---i.e., where it falls on the complexity spectrum
-mentioned above.  We define \emph{complexity} as a particular
-approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.  That is,
-we view a random-walk time series (which exhibits high entropy) as
-purely complex, whereas a periodic signal, which exhibits low entropy,
-is on the low end of the complexity spectrum.  We argue that an
-extension of \emph{permutation entropy} \cite{bandt2002per}---a method
-for approximating the entropy through ordinal analysis---is an
-effective way to assess the complexity of a given time series.
-Permutation entropy is ideal for our purposes because it is known to
-converge to the true entropy value. Other existing techniques either require specific knowledge of the generating process orproduce biased values of the entropy~\cite{bollt2001}.
+The information in an observation can be partitioned into two pieces:
+redundancy and entropy generation\cite{crutchfield2003}.  Our approach
+exploits this decomposition in order to assess how much predictive
+structure is present in a signal---i.e., where it falls on the
+complexity spectrum mentioned above.  We define \emph{complexity} as a
+particular approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.
+That is, we view a random-walk time series (which exhibits high
+entropy) as purely complex, whereas a periodic signal, which exhibits
+low entropy, is on the low end of the complexity spectrum.  We argue
+that an extension of \emph{permutation entropy}
+\cite{bandt2002per}---a method for approximating the entropy through
+ordinal analysis---is an effective way to assess the complexity of a
+given time series.  Permutation entropy is ideal for our purposes
+because it is known to converge to the true entropy value. Other
+existing techniques either require specific knowledge of the
+generating process or produce biased values of the
+entropy~\cite{bollt2001}.
 
@@
   system plays a crucial role in the success of different forecasting
-  schema, and in the choice of which forcasting scheme is appropriate.
+  schema---and in the choice of which forecasting scheme is appropriate.
 \end{enumerate}
@@
 chaotic \cite{mytkowicz09}, which places fundamental limits on
-predictability.  Radical changes in this behavior can result if the
-program is changed slightly.  The resulting traces span the whole
-range of the complexity spectrum, from completely predictable to
-completely unstructured, making this an ideal testbed for this study.
+predictability.  Changes in the program can radically change this
+behavior; indeed, many performance traces exhibit interesting regime
+changes as a program moves through the different phases of its
+operation.  The resulting traces span the whole range of the
+complexity spectrum, from completely predictable to completely
+unstructured, making this an ideal testbed for this
+study\footnote{Predicting the \emph{state} of a computer, of course,
+  would amount to solving the halting problem.  What we are doing here
+  is predicting its \emph{performance}, which does not violate the
+  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
+% footnote to appease the theoretician in Josh
 
@@
 The rest of the paper is organized as follows:
-Section~\ref{sec:related} discusses previous results on generating partitions, local modeling, and error distribution analysis, and situates this work in that
-context. Section~\ref{sec:methods} covers the experimental setup and
-methods used to collect each time series. Section~\ref{sec:model}
-describes the prediction models we implemented.  In
-Section~\ref{sec:meaComplex}, we review weighted permutation entropy.
-In Section~\ref{sec:results}, we estimate the complexity of each time
-series and compare that complexity to the accuracy of predictions
-produced by the methods of Section~\ref{sec:model} on that time
-series.  In Section~\ref{sec:conc}, we discuss these results and their
+Section~\ref{sec:related} discusses previous results on generating
+partitions, local modeling, and error distribution analysis, and
+situates this work in that context. Section~\ref{sec:methods} covers
+the experimental setup and methods used to collect each time
+series. Section~\ref{sec:model} describes the prediction models we
+implemented.  In Section~\ref{sec:meaComplex}, we review weighted
+permutation entropy.  In Section~\ref{sec:results}, we estimate the
+complexity of each time series and compare that complexity to the
+accuracy of predictions produced by the methods of
+Section~\ref{sec:model} on that time series.  In
+Section~\ref{sec:conc}, we discuss these results and their
 implications and consider future areas of research.
-
-[[Following belongs in the methods section:]] In addition, some
-performance traces even exhibit interesting regime changes as a
-program switches between varying subroutines.
-
-[[Following belongs in the conclusion.  Putting it here would be
-    appropriate in a CS paper, but not in a physics journal, and not
-    when we've just made the point that this is just a testbed.]] 
-% paragraph to appease the theoretician in me
-It is worth taking a moment to consider the theoretical possibility of
-this task. We are not attempting to predict the state of the CPU at an
-arbitrary point in the future --- this, at least with perfect
-accuracy, would be tantamount to solving the halting problem. What we
-are attempting is to predict aspects or functions of the running of
-the CPU, e.g., processor efficiency and similar statistics. Prediction
-of these quantities at some finite time in the future, even with
-perfect accuracy, does not violate the Rice-Shapiro
-theorem~\cite{hopcroft2007} which essentially states that any non-trivial
-property of a program is uncomputable.
 

### 2014-03-10 intro.tex
A bit more wordsmithing on intro.tex

@@
 \section{Introduction  }\label{sec:intro}
-{\color{blue}[[EDITABLE]]}
 %****NOTES START HERE****
@@
 predictive structure, i.e, signals whose future values can be
-successfully predicted from past values.  In situations like this,
-there is an underlying process that generates information and/or
+successfully predicted from past values.  This can be viewed as the
+output of an underlying process that generates information and/or
 transmits it from the past to the future in a perfectly predictable
@@
 implications and consider future areas of research.
-
-[[Following is a good start at an abstract. It doesn't need to be
-    reiterated in the introduction.]]  This paper provides insight
-into when, why, and how forecast strategies fail when they are applied
-to complicated time series by empirically estimating their inherent
-complexity. [[Liz, do you think here would be better for the sentence
-    about we are not the first, or is above where I asked Ryan to
-    insert it ok? I worry it would muddy the water of this synopsis
-    paragraph.]] We focus here on the specific example of computer
-performance but believe the results apply to a much broader class of
-time series. We conjecture that the complexity of traces from these
-systems---which results from the inherent dimension, non-linearity,
-and non-stationarity, as well as from measurement issues like noise,
-aggregation, and finite data length---is \emph{empirically
-  quantifiable}.  To validate this conjecture we model 120 different
-performance traces using a variety of forecasting models.  We then
-compare the empirically estimated \emph{complexity} with the accuracy
-of each forecast method.

### 2014-03-10 methods.tex
editing
putting it in Phys Rev E Format
more pre formatting
removing windows
adding window footnote

@@
 \section{Experimental Methods and Time Series Explanations}\label{sec:methods}
-
+{\color{blue} EDITABLE}
 % \begin{enumerate}
@@
   \centering
-    \includegraphics[width=0.75\textwidth]{figs/colFullTS}
+    \includegraphics[width=\columnwidth]{figs/colFullTS}
     \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \col. Each point is the average IPC during a 100,000 instruction period.}
@@
   \centering
-    \includegraphics[width=0.75\textwidth]{figs/gccfullts}
+    \includegraphics[width=\columnwidth]{figs/gccfullts}
     \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \gcc. Each point is the average IPC during a 100,000 instruction period.}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{figs/SVD1RegimesColored}
+    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
     \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \svd.
@@
 \svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually.
-Like Figures \ref{fig:col-ts} and \ref{fig:gcc-ts}, Figure \ref{fig:svd-ts-colored} is a single time series of the IPC of \svd, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series  visually switches between ``regimes".
+Like Figures \ref{fig:col-ts} and \ref{fig:gcc-ts}, Figure \ref{fig:svd-ts-colored} is a single time series of the IPC of \svd, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series  \emph{visually} switches between ``regimes".\footnote{It should be noted that a windowed version of the metric we introduce in Section~\ref{sec:meaComplex} has been used to detect siezures in brainwave data \cite{cao2004det}, which can be viewed as regime detection.
+When we used this method on \svd the regimes we chose visually had different entropies from regime to regime but but the regime boundaries seem fractal in nature.
+This is something we plan to investigate in future work.
+For the purposes of this paper the regime detection was done visually.}
 
@@
 The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift.
-In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}.
 Note, the purpose of this paper is \emph{not} to rigorously explore regime detection but to explore quantifying complexity of a time series.
@@
 For notational convenience we refer to these time series as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd, ordered from left to right.
+
+For regimes \cite{cao2004det}

### 2014-03-11 related.tex
editing related

--- 2014-03-05 related.tex
@@
-\section{Related Work {\color{red} NOT EDITABLE}}\label{sec:related}
+\section{Related Work }\label{sec:related}
+{\color{red}[[NOT EDITABLE]]}
+Attempting to model time series for prediction is an  old problem which arguably began with Yule in 1927 when he invented autoregression \cite{Yule27}.
+Since then hundreds if not thousands of prediction strategies have been developed for a wide variety of prediction tasks.
+The purpose of this paper is \emph{not} to model a time series for prediction but to understand the predictability of a real-valued-scalar time series model free and then to associate the model accuracy with the time series inherent complexity.
 
-Attempting to model and characterize predictability is a very old problem which arguably began with Yule in 1927 when he invented AR and many attempts to quantify predicatbility have followed including ....
+Quantifying predictability or sometimes called predicting predictability is also not a new problem and most solutions fall into two categories: model based and model free where information content is estimated through binning.
 
-The ideal method of computing the entropy of a system is via \emph{generating partitions}~\cite{lind95}, a technique which partitions observations into symbols in a way which preserves the underlying dynamics.
-Unfortunately, outside of one-dimensional unimodal maps and a few select higher dimensional maps, generating partitions are highly non-trivial to compute and often have fractal boundaries~\cite{eisele1999}.
-Furthermore, the underlying dynamic needs to be known to construct the generating partition.
-Unfortunately, using any partition (such as those produced by simply binning the data) other than the generation partition induces spurious complexity to the resulting symbolic sequence, thus misrepresenting the entropy of the underlying system~\cite{bollt2001}.
-
-Redundancy (details in SFI forecasting), relies heavily on either estimating the generating partiion which is hard to do *and* estimating the positive lyap spectrum which is hard to do for noisy systems and impossible for systems that are not deterministic.
+The first class of methods is model oriented.
+These methods often include one of two things: Building local  models around points in a time series or analyzing statistics on the distribution of model errors.
+A few examples of the first case are \cite{Smith199250,dvsplots}, in \cite{Smith199250} for example they build local error distributions around "landmarks"(look up actual word they use) of a reconstructed trajectory of a dynamical system.
+They then use these distributions to predict error bounds on  predictions which occur within that $\epsilon$-ball.
+DVS-plots (Deterministic-vs-Stochastic plots) are somewhat similar to this
 
@@
 
+ and many attempts to quantify predicatbility have followed including ....
+
+The ideal method of computing the entropy of a system is via \emph{generating partitions}~\cite{lind95}, a technique which partitions observations into symbols in a way which preserves the underlying dynamics.
+Unfortunately, outside of one-dimensional unimodal maps and a few select higher dimensional maps, generating partitions are highly non-trivial to compute and often have fractal boundaries~\cite{eisele1999}.
+Furthermore, the underlying dynamic needs to be known to construct the generating partition.
+Unfortunately, using any partition (such as those produced by simply binning the data) other than the generation partition induces spurious complexity to the resulting symbolic sequence, thus misrepresenting the entropy of the underlying system~\cite{bollt2001}.
+
+Redundancy (details in SFI forecasting), relies heavily on either estimating the generating partiion which is hard to do *and* estimating the positive lyap spectrum which is hard to do for noisy systems and impossible for systems that are not deterministic.
+
 But this method is differnt becasue it uses no assumption about the underlying model, does not require generating partitions, is applicable to noisy real-valued time series.

### 2014-03-11 related.tex
editing related

@@
 These methods often include one of two things: Building local  models around points in a time series or analyzing statistics on the distribution of model errors.
-A few examples of the first case are \cite{Smith199250,dvsplots}, in \cite{Smith199250} for example they build local error distributions around "landmarks"(look up actual word they use) of a reconstructed trajectory of a dynamical system.
+A few examples of the first case are \cite{Smith199250,Casdagli92dvsplots}, in \cite{Smith199250} for example they build local error distributions around "landmarks"(look up actual word they use) of a reconstructed trajectory of a dynamical system.
 They then use these distributions to predict error bounds on  predictions which occur within that $\epsilon$-ball.
-DVS-plots (Deterministic-vs-Stochastic plots) are somewhat similar to this
+Casdagli introduced Deterministic-vs-Stochastic plots (DVS-plots) in \cite{Casdagli92dvsplots}
 

### 2014-03-11 methods.tex
finishing first draft of related works section

@@
 \svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually.
-Like Figures \ref{fig:col-ts} and \ref{fig:gcc-ts}, Figure \ref{fig:svd-ts-colored} is a single time series of the IPC of \svd, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series  \emph{visually} switches between ``regimes".\footnote{It should be noted that a windowed version of the metric we introduce in Section~\ref{sec:meaComplex} has been used to detect siezures in brainwave data \cite{cao2004det}, which can be viewed as regime detection.
+Like Figures \ref{fig:col-ts} and \ref{fig:gcc-ts}, Figure \ref{fig:svd-ts-colored} is a single time series of the IPC of \svd, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series  \emph{visually} switches between ``regimes".\footnote{It should be noted that a windowed version of the complexity metric we discuss in Section~\ref{sec:meaComplex} has been used to detect siezures in brainwave data \cite{cao2004det}, which can be viewed as regime detection.
 When we used this method on \svd the regimes we chose visually had different entropies from regime to regime but but the regime boundaries seem fractal in nature.

### 2014-03-11 related.tex
finishing first draft of related works section

--- 2014-03-10 related.tex
@@
 {\color{red}[[NOT EDITABLE]]}
-Attempting to model and characterize predictability is a very old problem which arguably began with Yule in 1927 when he invented AR and many attempts to quantify predicatbility have followed including ....
+Attempting to model time series for prediction is an  old problem which arguably began with Yule in 1927 when he invented autoregression \cite{Yule27}.
+Since then hundreds if not thousands of prediction strategies have been developed for a wide variety of prediction tasks.
+The purpose of this paper is \emph{not} to model a time series for prediction but to understand the predictability of a real-valued-scalar time series model free and then to associate the model accuracy with the time series inherent complexity.
 
-The ideal method of computing the entropy of a system is via \emph{generating partitions}~\cite{lind95}, a technique which partitions observations into symbols in a way which preserves the underlying dynamics.
+Quantifying predictability or sometimes called predicting predictability is also not a new problem and most solutions fall into two categories: model-based error analysis and model-free information analysis.
+
+%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
+
+%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
+
+The first class of methods is model oriented and focuses on either local or global in- or out-of-sample errors of a particular forecast schema.
+The local side of this tries to build error distributions for different areas of a time series reconstruction by doing local ensemble in-sample forecasting and then using the constructed distribution of in-sample errors as a means to predict out-of-sample forecast errors in that area of the time series reconstruction.
+For example, in \cite{Smith199250} in-sample forecasts are done for ensembles around ``data centers" (a subset of the time series) to predict the local predictability of a time series reconstruction.
+This does not however tell you if redundancy exists in a time series but only tells you the local predictive capacity in relation to this particular model (an extension of the model described in Section~\ref{sec:lma}).
+This paper does brings up a very interesting point however that different regions of a time series may exhibit varying levels of local predictive structure, we expand on this finding later in this paper with a time series called ``\svd".
+
+%Distrubtion of error.
+For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process.
+But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process.
+For example, nonlinear structure which is ignored by a linear predictor.
+
+When analyzing post-facto global fits, practitioners often analyze the global error distribution of a forecasting schema.
+For example, it is common with autoregressive models to verify the error is normally distributed.
+If it is not then this is an indicator that more linear structure is present that can be used.
+The problem with this approach is that if the error is normally distributed it can not be said that the practitioner is using the ideal forecasting schema, e.g., if nonlinear structure is present that linear modeling is oblivious to.
+
+A practice known as deterministic vs.
+stochastic modeling\cite{Casdagli92dvsplots, weigend-book} bridges the gap between local and global error analysis.
+The basic idea is to construct a series of local linear fits (single point) to global linear fits (fitting all points known at that time), increasing the number of points in the fit slowly.
+Then analyze how the average out-of-sample error changes as a function of number of points in the linear fits.
+The shape of such a graph indicates how much determinism vs.
+stochasticity is present in a time series, see  \cite{Casdagli92dvsplots, weigend-book} for more information.
+
+ Our approach is completely different than this first class as our approach is model free, each of the methods in this class are based off very particular assumptions about the underlying generating system and knowledge of how error will result if those assumptions hold or break.
+Our approach instead empirically approximates the time series inherent complexity (Kolmogorov-Sinai entropy).
+Quantifying predictability through entropy/redundancy is not a new idea, but past methods (e.g.,\cite{Shannon1951, mantegna1994linguistic}) have hinged on knowledge of the  \emph{generating partitions}.
+
+It is the case that the ideal method of computing the entropy of a system is via \emph{generating partitions}~\cite{lind95}, a technique which partitions observations into symbols in a way which preserves the underlying dynamics.
 Unfortunately, outside of one-dimensional unimodal maps and a few select higher dimensional maps, generating partitions are highly non-trivial to compute and often have fractal boundaries~\cite{eisele1999}.
@@
 
-Redundancy (details in SFI forecasting), relies heavily on either estimating the generating partiion which is hard to do *and* estimating the positive lyap spectrum which is hard to do for noisy systems and impossible for systems that are not deterministic.
+{\color{red}[[Ryan to write]]
 
-DVS plots, gives pros and cons in (SFI prediction book)
+A paragraph on redundancy and its connection with predictability  how it is approximated and how these estimates are bad
 
-predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
+e.g., Redundancy estimates rely heavily on either estimating the generating partition which is hard to do *and* estimating the full positive Lyapunov spectrum which is hard to do for noisy systems and impossible for systems that are not deterministic.
 
-Distrubtion of error.
-For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process.
-But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process.
-For example, nonlinear structure which is ignored by a linear predictor.
+Also please edit / look at the paragrpah proceeding and following this one
+}
 
-But this method is differnt becasue it uses no assumption about the underlying model, does not require generating partitions, is applicable to noisy real-valued time series.
+Our approach expands on \cite{bandt2002per} to approximate normalized redundancy of a time series as a proxy for predictability.
+This method is different from the current literature because it does not rely on generating partitions, and as such does not suffer from the previously mentioned biases and no knowledge of the underlying generating process is necessary, e.g, linear, nonlinear, the Lyapunov spectrum etc.
+In contrast our approach uses no assumption about the underlying generating process, does not require generating partitions, and is applicable to noisy real-valued time series from all classes of systems.
 
+{\color{red}
 Maybe talk about ``
@@
 but kind of puts the cart before the horse
+}

### 2014-03-12 intro.tex
a few edits to intro.tex

@@
 for all noise-free deterministic signals\cite{weigend-book}---let
-alone all real-world time series data sets.  This leads naturally to
+alone all real-world time-series data sets.  This leads naturally to
 an important and challenging question: given a complicated real-valued
@@
 produced by a particular method is poor but the time series contains a
-significant amount of predictive structure, that method is inadequate
-to the task and one should seek another method.  We focus on
-real-valued, scalar, time-series data that appear complicated: i.e.,
-not periodic, linear, constant, etc.  We make no assumptions about the
-properties of the generating process: whether it is linear, nonlinear,
-deterministic, stochastic, stationary, non-stationary, etc.
+significant amount of predictive structure, one can reasonably conclude that that method is inadequate
+to the task and that one should seek another method.
 
@@
 given time series.  Permutation entropy is ideal for our purposes
-because it is known to converge to the true entropy value. Other
+because it works with real-valued data and is known to converge to the true entropy value. Other
 existing techniques either require specific knowledge of the
@@
 
+We focus on real-valued, scalar, time-series data that appear
+complicated: i.e., not periodic, linear, constant, etc.  We make no
+assumptions about the properties of the generating process: whether it
+is linear, nonlinear, deterministic, stochastic, stationary,
+non-stationary, etc.
 To explore the relationship between complexity, predictive structure,
-and predictability, we generate forecasts of a number of time-series
+and predictability, we generate forecasts of a variety of time-series
 datasets using four different prediction methods, then compare the
@@
 \begin{enumerate}
-\item The complexity of a noisy real-valued time series is
-  quantifiable by permutation entropy and correlated with prediction
+\item The complexity of a noisy real-valued time series is {\sl (i)}
+  quantifiable by permutation entropy and {\sl (ii)} correlated with prediction
   accuracy of an appropriate predictor.
@@
   system plays a crucial role in the success of different forecasting
-  schema---and in the choice of which forecasting scheme is appropriate.
+  schema---and in the choice of which one is appropriate for a given 
+time series.
 \end{enumerate}
-The forecast methods used here were chosen to span the space of
-standard prediction strategies, but they do not cover that space
+The forecast methods used in this study were chosen to span the space of
+standard prediction strategies, but they do not, of course, cover that space
 exhaustively.  Our goal here is an empirical assessment of the
@@
 than use synthetic data, we study sensor data from computer
-performance experiments.  Computers are among the most complex
-engineered artifacts in current use.  Modern microprocessor chips
-contain multiple processing units and multi-layer memories, for
-instance, and they use complicated hardware/software strategies to
-move data and threads of computation across those resources.  As a
-result, the processor and memory loads during the execution of even a
-very simple program can evolve in a very complicated fashion.
-Figure~\ref{fig:ipc} shows a performance trace of a three-line loop
-that repeatedly initializes the upper triangle of a matrix in
-column-major order.
+performance experiments.  While this is not a common laboratory
+experiment, it is a highly appropriate choice here.  Computers are
+among the most complex engineered artifacts in current use.  Modern
+microprocessor chips contain multiple processing units and multi-layer
+memories, for instance, and they use complicated hardware/software
+strategies to move data and threads of computation across those
+resources.  As a result, the processor and memory loads during the
+execution of even a very simple program can evolve in a very
+complicated fashion.  Figure~\ref{fig:ipc} shows a performance trace
+of a three-line loop that repeatedly initializes the upper triangle of
+a matrix in column-major order.
 %
@@
   would amount to solving the halting problem.  What we are doing here
-  is predicting its \emph{performance}, which does not violate the
+  is predicting computer \emph{performance}, which does not violate the
   Rice-Shapiro theorem~\cite{hopcroft2007}.}.
@@
 
-The rest of the paper is organized as follows:
+The rest of the paper is organized as follows.
 Section~\ref{sec:related} discusses previous results on generating
@@
 the experimental setup and methods used to collect each time
-series. Section~\ref{sec:model} describes the prediction models we
-implemented.  In Section~\ref{sec:meaComplex}, we review weighted
-permutation entropy.  In Section~\ref{sec:results}, we estimate the
+series. Section~\ref{sec:model} describes the prediction models used
+in this study.  In Section~\ref{sec:meaComplex}, we review the
+permutation entropy technique.  In Section~\ref{sec:results}, we estimate the
 complexity of each time series and compare that complexity to the
 accuracy of predictions produced by the methods of
-Section~\ref{sec:model} on that time series.  In
+Section~\ref{sec:model}, operating on that time series.  In
 Section~\ref{sec:conc}, we discuss these results and their

### 2014-03-12 intro.tex
changed fig 1 name; rewrote section 3.1

@@
 execution of even a very simple program can evolve in a very
-complicated fashion.  Figure~\ref{fig:ipc} shows a performance trace
+complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance trace
 of a three-line loop that repeatedly initializes the upper triangle of
@@
       i7\textsuperscript{\textregistered}-based machine.}
-   \label{fig:ipc}
+   \label{fig:col-ipc}
   \end{figure}

### 2014-03-12 methods.tex
changed fig 1 name; rewrote section 3.1

@@
 
-The time-series data for these experiments was collected on an Intel Core\textsuperscript{\textregistered} i7-2600 running the 2.6.38-8 Linux
-kernel.  This Nehalem chip has eight cores running at 3.40GHz and a cache size
-of 8192 KB.  We collected performance traces of processor efficiency (IPC) during the execution of three exemplar programs---a simple microkernel(\col) and two complex programs: one from the
-SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd).
-In particular, we measured and aggregated the instructions executed per clock cycle (ipc)---a measure of processor efficiency---at 100,000-instruction intervals.
-To record these measurements,
-we used the {\tt libpfm4} library, via PAPI 5.2
-\footnote{Performance Application Programming Interface}\cite{papi}, combined with a custom measurement infrastructure\cite{todd-phd} to stop program execution every 100,00-instructions and read the contents of the CPU's onboard hardware performance monitors (HPMs).
-HPMs are specialty registers who exist solely to store performance data.
-For statistical validation we collected 15 performance traces from each program resulting in 45 total time series.
+The time-series data for these experiments was collected on an Intel
+Core\textsuperscript{\textregistered} i7-2600-based machine running
+the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
+eight processing units---''cores''---a clock rate of 3.40GHz, and a
+cache size of 8192 KB.  We collected performance traces of processor
+efficiency during the execution of three different programs---the
+simple loop (\col) whose performance is depicted in
+Figure~\ref{fig:col-ipc} and two more-complex programs: one from the
+SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK linear
+algebra package (\svd).  To record these measurements, we used the
+{\tt libpfm4} library, via PAPI
+% 
+% \footnote{Performance Application Programming Interface}
+% 
+5.2~\cite{papi}, to stop program execution at 100,000-instruction
+intervals and read the contents of the CPU's onboard hardware
+performance monitors\footnote{These specialty registers are built into
+  modern microprocessors for the purpose of on-board monitoring and
+  storage of performance data.}, which we had programmed to count how
+many instructions were executed in each clock cycle (IPC).  For an
+in-depth description of this custom-measurement infrastructure,
+including discussion of the implications of the measurement interval,
+please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
+validation, we collected 15 performance traces from each of the three
+programs.
 
-Obviously using a system to measure itself can cause interference--i.e, are you measuring the generating process cleanly or are you adding noise to the process?---but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way.
-For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see
-\cite{zach-IDA10,mytkowicz09,todd-phd}.
+% Obviously using a system to measure itself can cause
+% interference--i.e, are you measuring the generating process cleanly or
+% are you adding noise to the process?---but due diligence was put into
+% monitoring the generating process of these time series without
+% interfering with it in a significant way.
 

### 2014-03-13 results.tex
update predictions plot
update caption a bit

@@
   \centering
-    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
-    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
-    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
-\caption{WPE vs. MASE for all trials, methods, and systems.  \svdone,
-  \svdthree, and \svdfive are omitted from the top plot for scale
-  reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm
-  0.282$, and $20.870 \pm 0.192$ respectively.  {\color{red}Replace
-    the top plot with a plot of the naive results---but running from 0
-    to 3.1 on the horizontal axis and excluding SVD 1, 3, and 5}.  The
-  dashed line is the same as in the previous figure [[Ryan, is this
-      correct?]].  Numerical values, including means and standard
-  deviations of the errors, can be found in Table~\ref{tab:error} in
-  the appendix.
+%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
+%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
+%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
+  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy}
+\caption{WPE vs.
+MASE for all trials, methods, and systems.
+\svdone, \svdthree, and \svdfive are omitted from the top plot for scale reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm 0.282$, and $20.870 \pm 0.192$ respectively.
+The dashed line is the same as in the previous figure.
+Numerical values, including means and standard deviations of the errors, can be found in Table~\ref{tab:error} in the appendix.
 % 

### 2014-03-13 methods.tex
Liz's rewrite of section 3

@@
-\section{Experimental Methods and Time Series Explanations}\label{sec:methods}
-{\color{blue} EDITABLE}
+\section{Experimental Methods}\label{sec:methods}
 % \begin{enumerate}
@@
 
-\subsection{Time Series Collection {\color{blue} EDITABLE}}
+% \subsection{Time Series Collection}
 
@@
 the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
-eight processing units---''cores''---a clock rate of 3.40GHz, and a
-cache size of 8192 KB.  We collected performance traces of processor
-efficiency during the execution of three different programs---the
-simple loop (\col) whose performance is depicted in
-Figure~\ref{fig:col-ipc} and two more-complex programs: one from the
-SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK linear
-algebra package (\svd).  To record these measurements, we used the
-{\tt libpfm4} library, via PAPI
+eight processing units, a clock rate of 3.40GHz, and a cache size of
+8192 KB.  We gathered performance traces during the execution of three
+different programs---the simple \col loop whose performance is
+depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
+programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
+from the LAPACK linear algebra package (\svd).  To record these
+measurements, we used the {\tt libpfm4} library, via PAPI
 % 
@@
 in-depth description of this custom-measurement infrastructure,
-including discussion of the implications of the measurement interval,
+including discussion of the implications of the sampling interval,
 please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
 validation, we collected 15 performance traces from each of the three
-programs.
+programs.  These traces, and the processes that generated them, are
+described in more depth in the rest of this section.
 
@@
 
-\subsection{The Programs: [[Maybe The generating processes]]{\color{blue} EDITABLE}}
-We collect performance traces of processor efficiency (IPC) of three exemplar programs, [[If we change the title of the section, add this: which act as our generating processes]]---a simple microkernel(\col) and two complex programs: one from the
-SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd).
-In this section we provide a quick explanation of each program as well as an example performance trace (Seen in Figures \ref{fig:col-ts}-\ref{fig:svd-ts-colored}.
-\subsubsection{\col {\color{blue} EDITABLE}}
-\col is a simple three-line C program that repeatedly initializes the upper triangle of a  2048 X 2048 matrix in column-major order.
-While this is a very simple program it has been shown to produce time series with very complicated behavior.
-In fact, in \cite{mytkowicz09} it was shown that time series generated from this simple program can exhibit everything from periodicity to deterministic chaos.
-A time series of the processor efficiency (IPC) of this program is shown in Figure \ref{fig:col-ts}.
+% \subsection{The Programs: [[Maybe The generating processes]]{\color{blue} EDITABLE}}
+% We collect performance traces of processor efficiency (IPC) of three
+% exemplar programs, [[If we change the title of the section, add this:
+%     which act as our generating processes]]---a simple
+% microkernel(\col) and two complex programs: one from the SPEC 2006CPU
+% benchmark suite (\gcc), and one from LAPACK (\svd). In this section we
+% provide a quick explanation of each program as well as an example
+% performance trace (Seen in Figures
+% \ref{fig:col-ts}-\ref{fig:svd-ts-colored}.
 
+\col is a simple three-line C program that repeatedly initializes the
+upper triangle of a 2048 $\times$ 2048 matrix in column-major order.
+As mentioned in the previous section, this simple program exhibits
+surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
+of the processor utilization during the execution this program is
+shown in Figure~\ref{fig:col-ts}.
 \begin{figure}[htbp]
@@
     \includegraphics[width=\columnwidth]{figs/colFullTS}
-    \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \col. Each point is the average IPC during a 100,000 instruction period.}
+    \caption{The instructions executed per CPU clock cycle (IPC)
+      during the execution of \col.  Each point is the average IPC
+      during a 100,000 instruction period. Figure~\ref{fig:col-ipc} is
+      a closeup of the first part of this signal.  }
     \label{fig:col-ts}
@@
 
-\subsubsection{\gcc {\color{blue} EDITABLE}}
-
-\gcc is a benchmark program included in the SPEC CPU2006 benchmark suite \cite{spec}.
-This program is written by Richard Stallman and is based on Version 3.2 of {\tt gcc}.
-This benchmark essentially runs as a compiler with many of its optimization flags enabled, compiling a series of input files and generating x86-64 assembly code files intended for execution on AMD Opteron processors\cite{spec}.
-This program is more typical of programs that are used every day, as opposed to a simple micro-kernel,e.g., \col.
-A time series of the processor efficiency of this program running on the Intel i7-2600 is shown in Figure \ref{fig:gcc-ts}.
-
+The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
+complicated programs that are used in the computer science community
+to assess and compare the performance of different computers.  \gcc is
+a member of that suite.
+% 
+% This program is written by Richard Stallman and is based on Version
+% 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with
+% many of its optimization flags enabled, compiling a series of input
+% files and generating x86-64 assembly code files intended for execution
+% on AMD Opteron processors\cite{spec}. 
+% 
+It is a \emph{compiler}: a program that translates code written in the
+C language into a lower-level format that can be executed by the
+processor chip.  Its behavior is far more complicated than that of
+\col, as is clear from Figure~\ref{fig:gcc-ts}.
   \begin{figure}[t]
@@
     \includegraphics[width=\columnwidth]{figs/gccfullts}
-    \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \gcc. Each point is the average IPC during a 100,000 instruction period.}
+    \caption{The instructions executed per CPU clock cycle (IPC)
+      during the execution of \gcc. Each point is the average IPC
+      during a 100,000 instruction period.}
     \label{fig:gcc-ts}
   \end{figure}
+Unlike \col, where the processor utilization is quite structured,
+\gcc's performance appears almost random.
 
-\subsubsection{\svd {\color{blue} EDITABLE}}
-
-\svd is a Fortran program from the LAPACK suite \cite{lapack} which calculates the singular value decomposition (SVD) of a rectangular $M$ by $N$  real-valued matrix which need not contain special structure such as symmetric, positive definite etc.
-To execute this program we input a 750 x 1000 matrix of randomly generated entries\footnote{Multiple randomly generated matrices were investigated but no measurable effect was present in the resulting time series.}.
-A ``regime-colored" time series of the IPC of this program is shown in Figure \ref{fig:svd-ts-colored}.
-We now describe why this is done.
-
+\svd is a Fortran program from the LAPACK linear algebra package
+\cite{lapack}.  It calculates the singular value decomposition of an
+rectangular $M$ by $N$ real-valued matrix.  For our experiments, we
+choose $M=750$ and $N=1000$ and generated the matrix entries randomly.
+% 
+% \footnote{Multiple randomly generated matrices were
+%   investigated but no measurable effect was present in the resulting
+%   time series.}. 
+% 
+The behavior of this program as it computes the singular values of
+this matrix is very interesting, as shown in Figure
+\ref{fig:svd-ts-colored}.  As the code moves though its different
+phases---diagonalizing the matrix, computing its transpose,
+multiplying, etc.---the processor utilization changes quite radically,
+as shown in Figure~\ref{fig:svd-ts-colored}.
 \begin{figure}[t]
@@
     \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
-    \caption{The instructions executed per CPU clock cycle (IPC) during the execution of \svd.
-Each point is the average IPC during a 100,000 instruction period.
-The colors correspond to the six different regimes detected visually.}
+    \caption{The instructions executed per CPU clock cycle (IPC)
+      during the execution of \svd. Each point is the average IPC
+      during a 100,000 instruction period.  The colors refer to different
+``regimes'' of the signal, as discussed in the text.}
     \label{fig:svd-ts-colored}
   \end{figure}
+For the first $\sim$27,000 instructions, roughly 1.8 instructions are
+executed per cycle, on the average, by the eight processing units on
+this chip.  After that, the IPC moves through a number of different
+oscillatory regimes, which we have color-coded in the figure in order
+to make textual cross-references easy to track and understand.
 
-\svd brings up a very interesting point: as we observe a system the generating process can change in complexity over time.
-While it is clear in Figures \ref{fig:col-ts} and \ref{fig:gcc-ts} that a single system is being observed, it also appears \emph{visually} that the complexity of these two time series is consistent over time, i.e., while they are both complicated they don't appear to increase or decrease in complexity over the range of the program.
-\svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually.
-Like Figures \ref{fig:col-ts} and \ref{fig:gcc-ts}, Figure \ref{fig:svd-ts-colored} is a single time series of the IPC of \svd, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series  \emph{visually} switches between ``regimes".\footnote{It should be noted that a windowed version of the complexity metric we discuss in Section~\ref{sec:meaComplex} has been used to detect siezures in brainwave data \cite{cao2004det}, which can be viewed as regime detection.
-When we used this method on \svd the regimes we chose visually had different entropies from regime to regime but but the regime boundaries seem fractal in nature.
-This is something we plan to investigate in future work.
-For the purposes of this paper the regime detection was done visually.}
+This wide range of behaviors provides a unique advantage, for the
+purposes of this paper, in that a number of different generating
+processes---with a wide range of complexities---are at work at
+different points in a single time series.  The complexity of \col and
+\gcc in Figures~\ref{fig:col-ts} and \ref{fig:gcc-ts} appears to be
+far more consistent over time---probably the result of a single
+generating process.  \svd, in contrast, has multiple regimes, each the
+result of a different generating process.  To take advantage of this,
+we split the signal into six different segments, thereby obtaining an
+array of examples for the analyses in the following sections.  For
+notational convenience, we refer to these 90 time series\footnote{15
+  runs, each with six regimes} as {\tt dgesdd$_i$} with $i \in
+\{1\dots6\}$ where $i$ corresponds to one of the six color-coded
+segments of \svd, ordered from left to right.  These segments, which
+were determined visually, are shown in different colors in
+Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
+of course, particularly since the regimes exhibit some fractal
+structure\footnote{It should be noted that a version of the complexity
+  metric discussed in Section~\ref{sec:meaComplex} has been used to
+  detect seizures in brainwave data \cite{cao2004det}, which can be
+  viewed as regime detection.  The purpose of this paper is \emph{not}
+  to rigorously explore regime detection but to explore quantifying
+  complexity of a time series.}.  Thus, it may the case that more than
+one generating process may be at work in each of our segments.  This
+is a factor in the discussion of Section~\ref{sec:results}.
 
-This is caused by the code of \svd moving between different subroutines as the SVD algorithm evolves, something not present in either \gcc or \col.
-We call these changes in \svd dynamics \emph{\svd regimes} and we have colored each of the six regimes differently in Figure \ref{fig:svd-ts-colored}.
-The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift.
-Note, the purpose of this paper is \emph{not} to rigorously explore regime detection but to explore quantifying complexity of a time series.
-As such, making these regime splits serves to provide an additional 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore.
-For notational convenience we refer to these time series as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd, ordered from left to right.
-
-For regimes \cite{cao2004det}
+[[Orphan citation?]] For regimes \cite{cao2004det}

### 2014-03-14 related.tex
a few quick edits

@@
 Since then hundreds if not thousands of prediction strategies have been developed for a wide variety of prediction tasks.
-The purpose of this paper is \emph{not} to model a time series for prediction but to understand the predictability of a real-valued-scalar time series model free and then to associate the model accuracy with the time series inherent complexity.
+The purpose of this paper is \emph{not} to model a time series for prediction but to understand the predictability of a real-valued-scalar time series model free and then to associate the prediction accuracy with the time series inherent complexity.
 
-Quantifying predictability or sometimes called predicting predictability is also not a new problem and most solutions fall into two categories: model-based error analysis and model-free information analysis.
+Quantifying predictability, or what is sometimes called predicting predictability, is also not a new problem and most solutions fall into two categories: model-based error analysis and model-free information analysis.
 
@@
 
- Our approach is completely different than this first class as our approach is model free, each of the methods in this class are based off very particular assumptions about the underlying generating system and knowledge of how error will result if those assumptions hold or break.
+ Our approach is completely different than this first class as it is model free.
+Each of the methods in the first class are based off very particular assumptions about the underlying generating system and knowledge of how error will result if those assumptions hold or break.
 Our approach instead empirically approximates the time series inherent complexity (Kolmogorov-Sinai entropy).
-Quantifying predictability through entropy/redundancy is not a new idea, but past methods (e.g.,\cite{Shannon1951, mantegna1994linguistic}) have hinged on knowledge of the  \emph{generating partitions}.
+Quantifying predictability through entropy/redundancy is not a new idea, but past methods (e.g.,\cite{Shannon1951, mantegna1994linguistic}) have hinged on knowledge of the \emph{generating partition} of the generating system.
 

### 2014-03-14 related.tex
boom, paragraph

@@
 
-{\color{red}[[Ryan to write]]
-
-A paragraph on redundancy and its connection with predictability  how it is approximated and how these estimates are bad
-
-e.g., Redundancy estimates rely heavily on either estimating the generating partition which is hard to do *and* estimating the full positive Lyapunov spectrum which is hard to do for noisy systems and impossible for systems that are not deterministic.
-
-Also please edit / look at the paragrpah proceeding and following this one
-}
+It seems desirable to quantify how much information prior observations of a system lend to future values, the \emph{mututal information} between, say, the past $n$ observations at the current one.
+This quantity is known as the \emph{redundancy}.
+For example, this would be the amount of information that delay coordinate embedding captures.
+For IID random processes,  the redundancy is zero, since all observations are independent of each other.
+On the other hand, any deterministic system --- including chaotic ones --- can be perfectly predicted if observed for long enough~\cite{{weigend-book}.
+Unfortunately it is quite difficult to estimate the redundancy in practice for an arbitrary, real-valued time series.
+Doing so requires knowing either the Kolmolgorov-Sinai entropy or the Lyapunov spectrum.
+As discussed above, computing the entropy is quite difficult.
+Similarly, computing the Lyapunov exponent from an arbitrary time series is also fraught with difficulties, and is effectivly impossible if there is significant noise or the generating system is stochastic.
 

### 2014-03-14 related.tex
mark related stuff as editable

@@
 \section{Related Work }\label{sec:related}
-{\color{red}[[NOT EDITABLE]]}
+{\color{blue}[[EDITABLE]]}
 Attempting to model time series for prediction is an old problem which arguably began with Yule in 1927 when he invented autoregression \cite{Yule27}.
@@
 
+The permutation entropy has been used in the past to detect
+
 {\color{red}

### 2014-03-15 related.tex
add a comma

@@
 Our approach expands on \cite{bandt2002per} to approximate normalized redundancy of a time series as a proxy for predictability.
-This method is different from the current literature because it does not rely on generating partitions, and as such does not suffer from the previously mentioned biases and no knowledge of the underlying generating process is necessary, e.g, linear, nonlinear, the Lyapunov spectrum etc.
+This method is different from the current literature because it does not rely on generating partitions, and as such does not suffer from the previously mentioned biases and no knowledge of the underlying generating process is necessary, e.g, linear, nonlinear, the Lyapunov spectrum, etc.
 In contrast our approach uses no assumption about the underlying generating process, does not require generating partitions, and is applicable to noisy real-valued time series from all classes of systems.
-
-The permutation entropy has been used in the past to detect
 
@@
 but kind of puts the cart before the horse
+
+[[ryan]] i tried writing a paragraph about this around 7 times, and couldn't get anything that sounded reasonable.
 }

### 2014-03-15 methods.tex
note and fixing citation orphan

@@
   \end{figure}
-For the first $\sim$27,000 instructions, roughly 1.8 instructions are
+For the first $\sim$27,000 instructions{\color{red}[[Joshua,I am not sure about this, I think it is more instructions than that.
+27,000 instructions are aggregated  in the first point of the time series (every 100,000 instructions is 1 point)...Liz how did you calculate this, I think it should be 21,000*100,000]]}, roughly 1.8 instructions are
 executed per cycle, on the average, by the eight processing units on
@@
 
-[[Orphan citation?]] For regimes \cite{cao2004det}
+%[[Orphan citation?]] For regimes \cite{cao2004det}

### 2014-03-15 model.tex
rewrite of the first subsection of section 4

@@
 \section{Modeling }\label{sec:model}
-{\color{blue} EDITABLE}
+% 
+% {\color{blue} EDITABLE}
+% 
+% Section outline:
+% 
+% \begin{enumerate}
+% \item \cmark section intro
+% \item \cmark fix order from simplest to most complex
+% \item Description of DCE and parameter estimation  
+% \item\cmark Description of auto ARIMA 
+% \item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
+% \item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results. 
+% 
+% \end{enumerate}
 
-Section outline:
+In this section, we describe the four different forecasting schema
+used in this study, as well as the error metric used to evaluate their
+predictive accuracy:
+\begin{itemize}
+\item The \emph{random-walk} method, which uses the previous value in
+  the observed signal as the forecast,
 
-\begin{enumerate}
-\item \cmark section intro
-\item \cmark fix order from simplest to most complex
-\item Description of DCE and parameter estimation  
-\item\cmark Description of auto ARIMA 
-\item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
-\item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
+\item The \emph{naive} method, which uses the mean of the
+  observed signal as the forecast,
 
-\end{enumerate}
+\item The \emph{ARIMA} method, a common linear forecast strategy, and
 
-In this section we describe the four different forecasting schema used in this study.
-In Section \ref{sec:simple} we outline two primitive strategies which often out perform sophisticated prediction strategies in the case of highly complex signals.
-In Section \ref{sec:arima} we describe Autoregressive integrated moving-average (ARIMA) a commonly used linear forecasting scheme.
-In Section \ref{sec:lma} we describe a simple nonlinear deterministic forecasting scheme, Lorenz Method of Analogous (LMA) for capturing and exploiting nonlinear dynamics present in a time series.
-And finally in Section \ref{sec:accuracy} we discuss the error metric we use to evaluate each models predictive accuracy.
+\item The \emph{LMA} method, which uses a near(est) neighbor forecast
+  strategy on a dynamical reconstruction of the signal.
+\end{itemize}
+ARIMA is based on standard autoregressive techniques.  LMA is designed
+to capture and exploit the deterministic structure of a signal from a
+nonlinear dynamical system.  The first two methods, somewhat
+surprisingly, often outperform sophisticated prediction strategies in
+the case of highly complex signals, as discussed below.
 
-\subsection{Primitive Strategies to combat complexity: Random Walk and \naive {\color{blue} EDITABLE}}\label{sec:simple}
-In this section we cover two very simple forecast models: Random Walk and \naive.
-With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement: $$p_i = X_{i-1,obs}.$$ The \naive  prediction strategy is to forecast the rolling average up to that point.
-i.e., the prediction $p_i$ is simply the average of all observed points up to that point: $$p_i = \sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}.$$
+\subsection{Two Simple Prediction Strategies}
+\label{sec:simple}
 
-While both of these methods are simplistic they are not without merit.
-When time series possess very little predictive structure,i.e., they exist on the very top of the complexity spectrum these two methods can sometimes be \emph{on-average} very good choices for a forecasting method, as we will see in our results.
-An example is forecasting currency exchange rates.
-In \cite{rwMeese,rwCCE} the authors show that various econometric techniques and prediction models fail to consistently outperform Random Walk predictions in forecasting various exchange rates.
-This is the case because these signals are volatile, noisy and possess very little information structure but they don't \emph{on-average} vary a great deal.
-That is the day-to-day exchange rates \emph{on-average} don't vary very much so guessing the last known exchange rate is often a very good approximation to the new exchange rate.
-[[Not a huge fan of this wording.]] The Na\"ive forecasting scheme can be a great prediction strategy if the time series has very little forward information structure, adheres to a unimodal distribution and has very little variance.
-In this circumstance, forecasting the mean will on average get you in the ball park of the forecast without taking into account the lack of structure.
-The \naive~ forecasting scheme can be thought of as a filtering mechanism which compresses a signal onto a single point, effectively filtering out both complexity and structure.
-However, if very little predictive structure is present in a signal then this filtering mechanism effectively mitigates the complexity.
+A random-walk predictor simply uses the last observed measurement as
+the forecast: that is, the value $p(i)$ at time $i$ is predicted to
+be $$p_i = X_{i-1,obs}$$ The \naive ~prediction strategy uses a
+rolling average to generate the forecast: the prediction $p_i$ is
+simply the average of [[all]] [[no window?  if so, then it's not a
+    ``rolling'' average]] observed points up to that point: $$p_i =
+\sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}$$
 
-While the simplicity of these methods is advantageous on average when a time series exists on the high side of the complexity spectrum, their simplicity can also be a disadvantage and cause the methods to be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series.
-Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths.
-Fortunately, they somewhat compliment each other, i.e., the weakness of Random Walk and \naive~ are covered by the other primitive method.
+While both of these methods are simplistic, they are not without
+merit.  For a time series near the high end of the complexity
+spectrum---i.e., one that possesses very little predictive
+structure---these two methods can actually be the best choice.  In
+forecasting currency exchange rates, for instance, sophisticated
+econometric techniques and prediction models fail to consistently
+outperform the random-walk method~\cite{rwMeese,rwCCE}.  These signals
+are volatile, noisy, and possess very little predictive structure, but
+they do not---on the average---vary a great deal, so guessing the last
+known value is not a bad strategy.  If a signal has a unimodal
+distribution with low variance, the \naive ~prediction strategy will
+perform quite well---even if the signal is highly complex---simply
+because the mean is a good approximation of the future behavior.
+Also, a temporal average effects a low-pass filtering operation.  In a
+signal with very little predictive structure, then, the \naive
+~prediction strategy effectively mitigates the complexity.
 
-The situation in which Random Walk fails is when a time series changes drastically on a regular basis.
-The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step.
-At every time step Random Walk would precisely predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$.
-In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue an error of 0.5 at every step.
-So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.
+Both of these methods have significant weaknesses, however, because
+they do not model the temporal patterns of the time series or even the
+distribution of its values.  Thus, they cannot track changes in that
+structure, which causes them to fail in a number of important
+situations.
 
-The \naive forecast fails when you have a signal which shifts between two or more regimes on a regular basis and the regimes are separated by a significant amount.
-In this case the \naive forecast will predict between the regimes at every time step eventually accruing massive error.
-To see how Random Walk can cover this weakness imagine a time series which switches between more than one regime but stays in each regime for several time steps, e.g., a regularly sampled noisy square wave, this behavior can be seen in some portions of the \svd trace seen in Figure \ref{fig:svd}.
-Random walk in this case could react to the regime shift within one time step, switch to the new Regime and begin forecasting nicely, where as \naive would just continue to predict between the modes of the multimodal distribution resulting in poor forecasting accuracy all of the time.
+Random-walk strategies are a particularly bad choice for time series
+that change drastically at every time step.  In the worst case---a
+large-amplitude square wave whose period is equivalent to twice the
+sample time---a random-walk prediction would be exactly 180 degrees
+out of phase with the true continuation.  The \naive ~method would be
+better in this situation, since it would always predict the mean.  It
+would, however, fail when a signal has a number of long-lived regimes
+that have significantly different means.  In this situation, the
+inertia of the \naive ~method's rolling mean is a liability and the
+random-walk method's nimbleness is an advantage, allowing it to
+react quickly to regime shifts.
 
-So clearly these primitive methods have their downfalls, but it is important to keep in mind that when a time series exists on the top of the complexity spectrum they have little predictive structure to exploit and often a primitive method like Random Walk or \naive~ will consistently out perform methods which rely on predictive structure.
-This is explored further in Section \ref{sec:accuracy} and \ref{sec:results}
+Of course, a method that could capture and exploit periodic patterns
+would be far more effective in these examples.  The LMA method
+introduced in Section~\ref{sec:lma} is designed to do exactly that.
+[[Couldn't one think about ARIMA as moving in that direction, too?]]
+Note, though, that periodic patterns appear only in signals that are
+at the low end of the complexity spectrum.  If a signal contains
+little predictive structure, forecast strategies like LMA [[and
+    ARIMA?]] have nothing to exploit and thus will be beaten out by
+the two simple strategies described in this section.  This is explored
+further in Sections~\ref{sec:accuracy} and~\ref{sec:results}
 

### 2014-03-16 wpe.tex
fixing wpe
added an intro paragraph to the wpe section

@@
 \section{Measuring Structural Complexity }\label{sec:meaComplex}
-{\color{red} NOT EDITABLE}
+{\color{blue} EDITABLE}
 
-THINGS TO DO
-\begin{enumerate}
-\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
-
-\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
-\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
-[[talk about again here but justify in information theory section]]
-\item justify and explain the regime split in \svd {\color{red} I(joshua) moved this justification in the experimental methods section, it just needs to be explored here by applying windowed wpe and explaining that method}
-\end{enumerate}
+The goal of the section is to describe how we reasonably estimate the entropy of an arbitrary, real-valued time series.
+We will first detail why this is a highly non-trivial task, and then we will weave a path through which this can in fact be accomplished.
+This path will draw upon a myriad of tools and results from a variety of fields including time series analysis, dynamical systems, stochastic processes, and chaos.
 
@@
 
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
-complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
-employs a method of discretizing real-valued time series that follows the
-intrinsic behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the Shannon
-entropy, permutation entropy looks at the statistics of the \emph{orderings} of
-sequences of values using ordinal analysis. Ordinal analysis of a time series is
-the process of mapping successive time-ordered elements of a time series to
-their value-ordered permutation of the same size.  By way of example, if $(x_1,
-x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
-is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
-other things, it is generally robust to observational noise and requires no
-knowledge of the underlying mechanisms.
+Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural complexity measure for time series" \cite{bandt2002per}.
+Permutation entropy employs a method of discretizing real-valued time series that follows the intrinsic behavior of the system under examination.
+Rather than looking at the statistics of sequences of values, as is done when computing the Shannon entropy, permutation entropy looks at the statistics of the \emph{orderings} of sequences of values using ordinal analysis.
+Ordinal analysis of a time series is the process of mapping successive time-ordered elements of a time series to their value-ordered permutation of the same size.
+By way of example, if $(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.
+This method has many features; among other things, it is generally robust to observational noise and requires no knowledge of the underlying mechanisms.
 

### 2014-03-16 wpe.tex
some minor edits to wpe section

@@
 The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.
-The faster the divergence, the more difficult prediction becomes.
+The faster the divergence, the larger the entropy and the more difficult prediction becomes.
 
@@
 This information can be partitioned in to two components: the information shared with past observations --- the mutual information between the past and present --- and the information in the present which is not contained in the past --- the conditional entropy.
-The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the \emph{entropy rate}.
+The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the aforementioned \emph{entropy rate}.
 In theory, the more redundancy in a signal the more predictable it is.

### 2014-03-17 model.tex
rewrite of ARIMA section and some notation changes there and in methods.tex

@@
 used in this study, as well as the error metric used to evaluate their
-predictive accuracy:
+predictive accuracy.  These methods include:
 \begin{itemize}
@@
 
-\item The \emph{LMA} method, which uses a near(est) neighbor forecast
+\item The \emph{LMA} method, which uses a near-neighbor forecast
   strategy on a dynamical reconstruction of the signal.
@@
 nonlinear dynamical system.  The first two methods, somewhat
-surprisingly, often outperform sophisticated prediction strategies in
-the case of highly complex signals, as discussed below.
+surprisingly, often outperform these more-sophisticated prediction
+strategies in the case of highly complex signals, as discussed below.
 
@@
 A random-walk predictor simply uses the last observed measurement as
-the forecast: that is, the value $p(i)$ at time $i$ is predicted to
-be $$p_i = X_{i-1,obs}$$ The \naive ~prediction strategy uses a
-rolling average to generate the forecast: the prediction $p_i$ is
-simply the average of [[all]] [[no window?  if so, then it's not a
-    ``rolling'' average]] observed points up to that point: $$p_i =
-\sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}$$
-
-While both of these methods are simplistic, they are not without
-merit.  For a time series near the high end of the complexity
-spectrum---i.e., one that possesses very little predictive
-structure---these two methods can actually be the best choice.  In
-forecasting currency exchange rates, for instance, sophisticated
-econometric techniques and prediction models fail to consistently
-outperform the random-walk method~\cite{rwMeese,rwCCE}.  These signals
-are volatile, noisy, and possess very little predictive structure, but
-they do not---on the average---vary a great deal, so guessing the last
-known value is not a bad strategy.  If a signal has a unimodal
-distribution with low variance, the \naive ~prediction strategy will
-perform quite well---even if the signal is highly complex---simply
-because the mean is a good approximation of the future behavior.
-Also, a temporal average effects a low-pass filtering operation.  In a
-signal with very little predictive structure, then, the \naive
-~prediction strategy effectively mitigates the complexity.
+the forecast: that is, the value $px_i$ at time $i$ is predicted to
+be $$px_i = x_{i-1}$$ The \naive ~prediction strategy uses an average
+to generate the forecast: the prediction $px_i$ is simply the running
+average of observervations up to that point: $$px_i =
+\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
+simplistic, they are not without merit.  For a time series near the
+high end of the complexity spectrum---i.e., one that possesses very
+little predictive structure---these two methods can actually be the
+best choice.  In forecasting currency exchange rates, for instance,
+sophisticated econometric techniques and prediction models fail to
+consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
+These signals are volatile, noisy, and possess very little predictive
+structure, but they do not---on the average---vary a great deal, so
+the random-walk method's strategy of simply guessing the last known
+value is not a bad choice.  If a signal has a unimodal distribution
+with low variance, the \naive ~prediction strategy will perform quite
+well---even if the signal is highly complex---simply because the mean
+is a good approximation of the future behavior.  Also, a temporal
+average effects a low-pass filtering operation.  In a signal with very
+little predictive structure, then, the \naive ~prediction strategy
+effectively mitigates the complexity.
 
@@
 structure, which causes them to fail in a number of important
-situations.
+situations.  Random-walk strategies are a particularly bad choice for
+time series that change drastically at every time step.  In the worst
+case---a large-amplitude square wave whose period is equivalent to
+twice the sample time---a random-walk prediction would be exactly 180
+degrees out of phase with the true continuation.  The \naive ~method
+would be a better choice in this situation, since it would always
+predict the mean.  It would, however, fail when a signal has a number
+of long-lived regimes that have significantly different means.  In
+this situation, the inertia of the \naive ~method's accumulating mean
+is a liability and the agility of the random-walk method is an
+advantage, since it can respond quickly to regime shifts.
 
-Random-walk strategies are a particularly bad choice for time series
-that change drastically at every time step.  In the worst case---a
-large-amplitude square wave whose period is equivalent to twice the
-sample time---a random-walk prediction would be exactly 180 degrees
-out of phase with the true continuation.  The \naive ~method would be
-better in this situation, since it would always predict the mean.  It
-would, however, fail when a signal has a number of long-lived regimes
-that have significantly different means.  In this situation, the
-inertia of the \naive ~method's rolling mean is a liability and the
-random-walk method's nimbleness is an advantage, allowing it to
-react quickly to regime shifts.
+Of course, methods that could capture and exploit the geometry of the
+data, or its temporal patterns, would be far more effective in the
+situations described in the previous paragraph.  The ARIMA and LMA
+methods introduced in Sections~\ref{sec:lma} and~\ref{sec:arima} are
+designed to do exactly that.  Note, though, that periodic patterns
+appear only in signals that are at the low end of the complexity
+spectrum.  If a signal contains little predictive structure, forecast
+strategies like LMA and ARIMA have [[nothing to exploit and thus will
+    be beaten out by]] [[needs wordsmithing from Liz]] the two simple
+strategies described in this section.  This is explored further in
+Sections~\ref{sec:accuracy} and~\ref{sec:results}.
 
-Of course, a method that could capture and exploit periodic patterns
-would be far more effective in these examples.  The LMA method
-introduced in Section~\ref{sec:lma} is designed to do exactly that.
-[[Couldn't one think about ARIMA as moving in that direction, too?]]
-Note, though, that periodic patterns appear only in signals that are
-at the low end of the complexity spectrum.  If a signal contains
-little predictive structure, forecast strategies like LMA [[and
-    ARIMA?]] have nothing to exploit and thus will be beaten out by
-the two simple strategies described in this section.  This is explored
-further in Sections~\ref{sec:accuracy} and~\ref{sec:results}
-
-\subsection{Linear Forecasting: Seasonal Autoregressive-integrated-moving average(ARIMA){\color{blue} EDITABLE}}\label{sec:arima}
+\subsection{A Regression-Based Prediction Strategy}
+\label{sec:arima}
 %\begin{enumerate}
@@
 
-When there is evidence a generating process exhibits predictive structure an obvious strategy would be to fit a hyperplane to the dataset and then use this hyperplane to forecast new data points.
-This is a very old technique stemming back to Yule in 1927\cite{weigend93} with his invention of autoregressive forecasting, which forecasts the next time step through a weighted average of the past $p$ time-steps, i.e., $$p_i = \phi(B)X_{i,obs}.$$ Here $\phi(z)$ is a polynomial of degree $p$ and $B$ is the backshift operator, i.e., $$B^mX_i  = X_{i-m,obs}.$$
-To improve this method further--especially in the presence of noisy data---one could allow for some wiggle room by adding a moving average of white noise to the model: $\theta(B)\epsilon_i$, where $\theta(z)$ is a polynomial of degree $q$ and $\{\epsilon_i\}\sim WN(0,\sigma^2)$.
-Finally if there is evidence of non-stationarity an initial differencing (integrated) is applied as an attempt to remove that non-stationarity: $(1-B^d)X_i$.
-Combining these three give us a nonseasonal  ARIMA$(p,d,q)$ model:  $$\phi(B)(1-B)^dX_i = \theta(B)\epsilon_i.$$ Where $p$, $d$ and $q$ correspond to the orders of the autoregressive, integrated and moving average models respectively.
-If periodic structure is present a seasonal ARIMA$(p,d,q)(P,D,Q)_m$ is often used and is defined by:$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = \Theta(B^m)\theta(B)\epsilon_i.$$ Here $\Phi(z)$ and $\Theta(z)$ are polynomials of degree P and Q respectively, and $m$ is the seasonal frequency.
-Forecasting is then done in two steps, first the integrated step is performed to remove non-stationarity.
-$$\hat{X_i} = (1-B^m)^D(1-B)^d X_{i,obs}.$$ Then forecasts are made by evaluating $$\Phi(B^m)\phi(B)\hat{X_i} = \Theta(B^m)\theta(B)\epsilon_i.$$
+Perhaps the simplest way to capture and exploit the structure of data
+is to fit a hyperplane to the dataset and then use that it to forecast
+new data points.  The roots of this date back to Yule's 1927 invention
+of the autoregressive schema~\cite{weigend93}, which forecasts the
+next time step through a weighted average of past observations: $$px_i
+= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
+that is fit to the past $p+1$ [[???]] values of $x_i$ via the
+backshift operator $B$ \footnote{I changed the superscript to $k$ to
+  make this definition general---not connected to any of the terms
+  that have specific meanings later in this section.}:
+$$B^k x_i = x_{i-k}$$
+%
+with $k=1...p$ [[presumably?]].  To account for noise in the data, one
+can add a so-called ``moving average'' term to the model:
+$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
+degree $q$ and $\epsilon$ is white noise with the same mean and
+variance as the data.
+% if you leave this in, you'll need to explain a lot about the data
+% and the calculation
+% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
+To remove nonstationarities in the data, one can detrend it using a
+differencing operation: $(1-B^d)x_i$.
 
-There is a vast amount of theory and literature on constructing, fitting and forecasting with this class of models.
-However, this is not the focus of this paper and  we refer the reader to \cite{davislinearts} for an in-depth exploration.
-In this paper, we intend to explore how predictive accuracy of several different forecasting schema correlates with signal complexity.
-As such, we use ARIMA$(p,d,q)(P,D,Q)_m$ (or simply ARIMA) models as an out of the box black box linear forecasting scheme.
-To do this we use an automated fitting and forecasting techniques described in \cite{autoARIMA}.
-This technique automatically selects model and differencing orders: $p, d, q, P, D, Q$ and $m$ as well as fitting the corresponding polynomials to the time series.
-Hyndman and Khandakar use unit root tests to select differencing orders, $d$ and $D$.
-In particular, they use a combination of
-KPSS unit-root tests \cite{KPSSunit} and their own extension of Canova-Hansen test \cite{Canova1995}.
-To do model order selection they use the Akaike information criterion (AIC) \cite{akaike1974} conditioned on the maximized likelihood of the  model fitted to the differenced data, $(1-B^m)^D(1-B)^d X_{i,obs}$ \cite{autoARIMA}.
-For more detail on this fitting process see \cite{autoARIMA}.
+A strategy that incorporates all three of these features is called a
+\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
+ $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
+%
+{\color{red} You had superscripts on some of the $B$s and not on
+  others.  I added them but I may have gotten them wrong.  I also
+  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
+  the first use of that construction.  Please make sure things are
+  consistent, here and throughout this section.}  Here, $p$, $d$ and
+$q$ correspond to the orders of the autoregressive, detrending, and
+moving average terms, respectively.
 
-ARIMA forecasting is a time tested modelling process which uses a global linear fit and adjustments for seasonality, non-stationarity and noise.
-If a generating process fits into this category then this model is a great short term predictor.
-If information is instead being transmitted in a nonlinear way this global linear fit falls apart.
-One other weakness of this method is that if long prediction horizons are needed the forecast is guaranteed to converge to the mean after some number of predictions (dependent on model order).
-For this reason, we limit all methods to doing multiple 1-step predictions as to not handicap this method.
+If periodic structure is present in the data, a \emph{seasonal ARIMA}
+model of order $(p,d,q)(P,D,Q)$ can be a good choice:
+%
+$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
+\Theta(B^Q)\theta(B^q)\epsilon_i$$
+%
+Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
+and $Q$, respectively [[which accomplish what purpose?  why?  how?
+    where did $D$ come from and what does it do?]].  $m$ is the
+seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
+    one should choose $P=Q=m$, since that's what the equation below
+    implicitly does.]]  To use a seasonal ARIMA model, one first uses
+the detrending term to remove nonstationarity: $$\hat{x_i} =
+(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
+evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
+\Theta(B^m)\theta(B^q)\epsilon_i$$
 
-\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{blue} EDITABLE}}\label{sec:lma}
+There is a vast amount of theory and literature regarding the
+construction and use of this class of models; we refer the reader to
+\cite{davislinearts} for an in-depth exploration.  For the purposes of
+this paper, where the goal is to explore the relationship between
+predictability and complexity across a broad array of forecast
+strategies, seasonal ARIMA models are a good exemplar of the class of
+linear forecasting strategies.  In the sections that follow, we use
+the automated fitting and forecasting techniques described
+in~\cite{autoARIMA} to select model and differencing orders and fit
+the corresponding polynomials to the time series.  This procedure uses
+a combination of KPSS unit-root tests \cite{KPSSunit} and a
+customization of the Canova-Hansen test \cite{Canova1995} to select
+the differencing orders $d$ and $D$ and chooses the model order using
+the Akaike information criterion (AIC) \cite{akaike1974}, conditioned
+on the maximized likelihood of the model fitted to the detrended data,
+$(1-B^m)^D(1-B^d) x_{i,obs}$.
 
-When a generating process exhibits deterministic predictive nonlinear structure a common approach is to reconstruct the generating processes dynamics using a technique called delay-coordinate embedding and then use the geometric structure of this reconstruction as a proxy for the information transmission occurring in the underlying process to predict the future.
-In this section we discuss the theory and implementation of this forecasting scheme.
+ARIMA forecasting is a common and time-tested procedure.  Its
+adjustments for seasonality, nonstationarity, and noise make it an
+appropriate choice for short-term predictions of time-series data
+generated by a wide range of processes.  If information is being
+generated and/or transmitted in a nonlinear way, however, a global
+linear fit is inappropriate and ARIMA forecasts will be inaccurate.
+Another weakness of this method is prediction horizon, since an ARIMA
+forecast is guaranteed to converge to the mean after some number of
+predictions, depending on model order.  To avoid this issue, we employ
+it in a stepwise fashion: we build the model, perform a one-step
+prediction, rebuild using the latest observations, and iterate until
+the desired prediction horizon is reached.  (For consistency, we take
+the same approach with the other three models in this study as well.)
+
+\subsection{Nonlinear Forecasting: The Lorenz Method of Analogues}
+\label{sec:lma}
+
+When a generating process exhibits deterministic predictive nonlinear
+structure a common approach is to reconstruct the generating processes
+dynamics using a technique called delay-coordinate embedding and then
+use the geometric structure of this reconstruction as a proxy for the
+information transmission occurring in the underlying process to
+predict the future. In this section we discuss the theory and
+implementation of this forecasting scheme.
 

### 2014-03-17 related.tex

@@
 Since then hundreds if not thousands of prediction strategies have been developed for a wide variety of prediction tasks.
-The purpose of this paper is \emph{not} to model a time series for prediction but to understand the predictability of a real-valued-scalar time series model free and then to associate the prediction accuracy with the time series inherent complexity.
-
-Quantifying predictability, or what is sometimes called predicting predictability, is also not a new problem and most solutions fall into two categories: model-based error analysis and model-free information analysis.
+The purpose of this paper is \emph{not} to find an ideal forecast model for a time series but to quantify empirically the predictive structure present in a real-valued-scalar time series model free and then to associate the prediction accuracy of a set of methods with the time series inherent complexity.
+For this reason we do not cover in-depth the spectrum of previous work in predictive modeling, we do however provide an overview of the models used in this paper in  Section~\ref{sec:model}.
+Quantifying predictability, or what is sometimes called ``predicting predictability", is also not a new problem and most solutions fall into two categories: model-based error analysis and model-free information analysis.
 
@@
 
-The first class of methods is model oriented and focuses on either local or global in- or out-of-sample errors of a particular forecast schema.
-The local side of this tries to build error distributions for different areas of a time series reconstruction by doing local ensemble in-sample forecasting and then using the constructed distribution of in-sample errors as a means to predict out-of-sample forecast errors in that area of the time series reconstruction.
-For example, in \cite{Smith199250} in-sample forecasts are done for ensembles around ``data centers" (a subset of the time series) to predict the local predictability of a time series reconstruction.
-This does not however tell you if redundancy exists in a time series but only tells you the local predictive capacity in relation to this particular model (an extension of the model described in Section~\ref{sec:lma}).
-This paper does brings up a very interesting point however that different regions of a time series may exhibit varying levels of local predictive structure, we expand on this finding later in this paper with a time series called ``\svd".
-
+The first class of methods for quantifying predictive structure is model oriented and focuses on either local or global in- or out-of-sample errors produced by a fixed forecasting schema.
+The local analysis attempts to approximate error distributions for different regions of a time series model by doing local ensemble in-sample forecasting.
+The analysis then uses the approximate in-sample error distribution as a means to estimate out-of-sample forecast errors in that area of the time series model.
+For example, in \cite{Smith199250} in-sample forecasts are done with ensembles around ``data centers" (a subset of the time series) to predict the local predictability of a time series reconstruction using delay-coordinate embedding.
+Using this method, Smtih et al.
+\cite{Smith199250} show  a very interesting corollary: different portions of a time series can exhibit varying levels of local predictive uncertainty.
+We expand on this finding later in this paper with a time series called ``\svd", which exhibits interesting regime changes.
+This local type analysis works quite well but this does not quantify a time series redundancy but only approximates the local predictive uncertainty in relation to a fixed model and tells you nothing of the predictive structure usable by other schema.
 %Distrubtion of error.
@@
 
-When analyzing post-facto global fits, practitioners often analyze the global error distribution of a forecasting schema.
+The global error analysis is generally done post-facto and attempts to quantify if the right model was chosen from a particular class of models by assessing the out-of-sample error distribution.
 For example, it is common with autoregressive models to verify the error is normally distributed.
-If it is not then this is an indicator that more linear structure is present that can be used.
-The problem with this approach is that if the error is normally distributed it can not be said that the practitioner is using the ideal forecasting schema, e.g., if nonlinear structure is present that linear modeling is oblivious to.
+If it is not then this is an indicator that linear structure is present that the current model failed to use.
+The problem with this approach is that if the error is normally distributed it can not be said that the practitioner is using the ideal forecasting schema---e.g., if nonlinear structure is present that linear modeling is oblivious to---but only if this class of model used all the structure it was able to process.
 
@@
 
- Our approach is completely different than this first class as it is model free.
-Each of the methods in the first class are based off very particular assumptions about the underlying generating system and knowledge of how error will result if those assumptions hold or break.
-Our approach instead empirically approximates the time series inherent complexity (Kolmogorov-Sinai entropy).
-Quantifying predictability through entropy/redundancy is not a new idea, but past methods (e.g.,\cite{Shannon1951, mantegna1994linguistic}) have hinged on knowledge of the \emph{generating partition} of the generating system.
+Each of these methods in the first class are based off very particular assumptions about the underlying generating process and knowledge of how error will result if those assumptions hold or fail.
+Our approach instead falls into the second class which is model-free information analysis.
+In particular, we empirically approximates the time series inherent complexity (Kolmogorov-Sinai entropy) and then correlates this complexity with the predictive accuracy of different forecasting schema, as a means to quantify predictability.
+Quantifying predictability through entropy/redundancy is not a new idea, but past methods (e.g.,\cite{Shannon1951, mantegna1994linguistic}) have hinged on knowledge of the \emph{generating partition} of the generating process, a luxury rarely if ever afforded to an analyst.
 
@@
 For IID random processes,  the redundancy is zero, since all observations are independent of each other.
-On the other hand, any deterministic system --- including chaotic ones --- can be perfectly predicted if observed for long enough~\cite{{weigend-book}.
+On the other hand, any deterministic system --- including chaotic ones --- can be perfectly predicted if observed for long enough~\cite{weigend-book}.
 Unfortunately it is quite difficult to estimate the redundancy in practice for an arbitrary, real-valued time series.
-Doing so requires knowing either the Kolmolgorov-Sinai entropy or the Lyapunov spectrum.
+Doing so requires knowing either the Kolmolgorov-Sinai entropy or the \emph{full} positive Lyapunov spectrum.
 As discussed above, computing the entropy is quite difficult.
-Similarly, computing the Lyapunov exponent from an arbitrary time series is also fraught with difficulties, and is effectivly impossible if there is significant noise or the generating system is stochastic.
+Similarly, computing the positive Lyapunov exponent(s) from an arbitrary time series is also fraught with difficulties, and is effectively impossible if there is significant noise or the generating system is stochastic.
 
 Our approach expands on \cite{bandt2002per} to approximate normalized redundancy of a time series as a proxy for predictability.
-This method is different from the current literature because it does not rely on generating partitions, and as such does not suffer from the previously mentioned biases and no knowledge of the underlying generating process is necessary, e.g, linear, nonlinear, the Lyapunov spectrum, etc.
-In contrast our approach uses no assumption about the underlying generating process, does not require generating partitions, and is applicable to noisy real-valued time series from all classes of systems.
+This method is different from the current literature because it does not rely on generating partitions, and as such does not suffer from the previously mentioned biases.
+In addition no knowledge of the underlying generating process is necessary, e.g, linear, nonlinear, the Lyapunov spectrum, etc.
+In fact, our approach uses no assumption about the underlying generating process, does not require generating partitions, and is applicable to noisy real-valued time series from all classes of systems.
 

### 2014-03-17 methods.tex
introduced data notation in methods.tex
rewrite of ARIMA section and some notation changes there and in methods.tex
added a mention to methods.tex that HPMs can measure other metrics besides IPC

@@
 programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
-from the LAPACK linear algebra package (\svd).  To record these
-measurements, we used the {\tt libpfm4} library, via PAPI
+from the LAPACK linear algebra package (\svd).  In these experiments,
+the scalar observation $x_{i}$ at time $i$ was a measurement of how
+well each program was able to exploit the native parellelism of the i7
+chip.  To record these measurements, we used the {\tt libpfm4}
+library, via PAPI
 % 
@@
   modern microprocessors for the purpose of on-board monitoring and
-  storage of performance data.}, which we had programmed to count how
-many instructions were executed in each clock cycle (IPC).  For an
-in-depth description of this custom-measurement infrastructure,
-including discussion of the implications of the sampling interval,
-please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
-validation, we collected 15 performance traces from each of the three
-programs.  These traces, and the processes that generated them, are
-described in more depth in the rest of this section.
+  storage of performance data.  They can be programmed to measure a
+  wide array of processor and memory metrics besides the one used
+  here.}, which we had programmed to count how many instructions were
+executed in each clock cycle (IPC).  For an in-depth description of
+this custom-measurement infrastructure, including discussion of the
+implications of the sampling interval, please see
+\cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical validation,
+we collected 15 performance traces from each of the three programs.
+These traces, and the processes that generated them, are described in
+more depth in the rest of this section.
 

### 2014-03-17 methods.tex
modified methods.tex to get the '~21,000 cycles' stuff right

@@
   \end{figure}
-For the first $\sim$27,000 instructions{\color{red}[[Joshua,I am not sure about this, I think it is more instructions than that.
-27,000 instructions are aggregated  in the first point of the time series (every 100,000 instructions is 1 point)...Liz how did you calculate this, I think it should be 21,000*100,000]]}, roughly 1.8 instructions are
-executed per cycle, on the average, by the eight processing units on
-this chip.  After that, the IPC moves through a number of different
-oscillatory regimes, which we have color-coded in the figure in order
-to make textual cross-references easy to track and understand.
+For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
+cycles), roughly 1.8 instructions are executed per cycle, on the
+average, by the eight processing units on this chip.  After that, the
+IPC moves through a number of different oscillatory regimes, which we
+have color-coded in the figure in order to make textual
+cross-references easy to track and understand.
 

### 2014-03-17 model.tex
rewrote the ARIMA section without a zillion equations

@@
 A random-walk predictor simply uses the last observed measurement as
-the forecast: that is, the value $px_i$ at time $i$ is predicted to
-be $$px_i = x_{i-1}$$ The \naive ~prediction strategy uses an average
-to generate the forecast: the prediction $px_i$ is simply the running
-average of observervations up to that point: $$px_i =
+the forecast: that is, the value $px_i$ at time $i$ is predicted using
+the relation $$px_i = x_{i-1}$$ The \naive ~prediction strategy
+averages the prior observations to generate the forecast: $$px_i =
 \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
@@
 best choice.  In forecasting currency exchange rates, for instance,
-sophisticated econometric techniques and prediction models fail to
+sophisticated econometrics-based prediction models fail to
 consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
@@
 well---even if the signal is highly complex---simply because the mean
-is a good approximation of the future behavior.  Also, a temporal
+is a good approximation of the future behavior.  Moreover, a temporal
 average effects a low-pass filtering operation.  In a signal with very
@@
 
-Both of these methods have significant weaknesses, however, because
-they do not model the temporal patterns of the time series or even the
-distribution of its values.  Thus, they cannot track changes in that
-structure, which causes them to fail in a number of important
+Both of these methods have significant weaknesses, however.  Because
+they do not model the temporal patterns in the data, or even the
+distribution of its values, they cannot track changes in that
+structure.  This causes them to fail in a number of important
 situations.  Random-walk strategies are a particularly bad choice for
@@
 would be a better choice in this situation, since it would always
-predict the mean.  It would, however, fail when a signal has a number
-of long-lived regimes that have significantly different means.  In
-this situation, the inertia of the \naive ~method's accumulating mean
-is a liability and the agility of the random-walk method is an
-advantage, since it can respond quickly to regime shifts.
+predict the mean.  It would, however, perform poorly when a signal has
+a number of long-lived regimes that have significantly different
+means.  In this situation, the inertia of the \naive ~method's
+accumulating mean is a liability and the agility of the random-walk
+method is an advantage, since it can respond quickly to regime shifts.
 
@@
 spectrum.  If a signal contains little predictive structure, forecast
-strategies like LMA and ARIMA have [[nothing to exploit and thus will
-    be beaten out by]] [[needs wordsmithing from Liz]] the two simple
-strategies described in this section.  This is explored further in
-Sections~\ref{sec:accuracy} and~\ref{sec:results}.
+strategies like ARIMA and LMA have nothing to work with and thus will
+often be outperformed by the two simple strategies described in this
+section.  This is explored further in Sections~\ref{sec:accuracy}
+and~\ref{sec:results}.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% here is the old version of this material, for use in other documents...
+%
+%Perhaps the simplest way to capture and exploit the structure of data
+%is to fit a hyperplane to the dataset and then use that it to forecast
+%new data points.  The roots of this date back to Yule's 1927 invention
+%of the autoregressive schema~\cite{weigend93}, which forecasts the
+%next time step through a weighted average of past observations: $$px_i
+%= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
+%that is fit to the past $p+1$ [[???]] values of $x_i$ via the
+%backshift operator $B$ \footnote{I changed the superscript to $k$ to
+%  make this definition general---not connected to any of the terms
+%  that have specific meanings later in this section.}:
+%$$B^k x_i = x_{i-k}$$
+%%
+%with $k=1...p$.  To account for noise in the data, one can add a
+%so-called ``moving average'' term to the model:
+%$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
+%degree $q$ and $\epsilon$ is white noise with the same mean and
+%variance as the data.
+%% if you leave this in, you'll need to explain a lot about the data
+%% and the calculation
+%% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
+%To remove nonstationarities in the data, one can detrend it using a
+%differencing operation: $(1-B^d)x_i$.  
+%
+%A strategy that incorporates all three of these features is called a
+%\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
+% $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
+%%
+%{\color{red} You had superscripts on some of the $B$s and not on
+%  others.  I added them but I may have gotten them wrong.  I also
+%  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
+%  the first use of that construction.  Please make sure things are
+%  consistent, here and throughout this section.}  Here, $p$, $d$ and
+%$q$ correspond to the orders of the autoregressive, detrending, and
+%moving average terms, respectively.
+%
+%If periodic structure is present in the data, a \emph{seasonal ARIMA}
+%model of order $(p,d,q)(P,D,Q)$ can be a good choice:
+%%
+%$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
+%\Theta(B^Q)\theta(B^q)\epsilon_i$$
+%%
+%Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
+%and $Q$, respectively [[which accomplish what purpose?  why?  how?
+%    where did $D$ come from and what does it do?]].  $m$ is the
+%seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
+%    one should choose $P=Q=m$, since that's what the equation below
+%    implicitly does.]]  To use a seasonal ARIMA model, one first uses
+%the detrending term to remove nonstationarity: $$\hat{x_i} =
+%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
+%evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
+%\Theta(B^m)\theta(B^q)\epsilon_i$$
+% ...down to here
+%%%%%%%%%%%%%%%%%%%%%%%%
+
 Perhaps the simplest way to capture and exploit the structure of data
-is to fit a hyperplane to the dataset and then use that it to forecast
-new data points.  The roots of this date back to Yule's 1927 invention
-of the autoregressive schema~\cite{weigend93}, which forecasts the
-next time step through a weighted average of past observations: $$px_i
-= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
-that is fit to the past $p+1$ [[???]] values of $x_i$ via the
-backshift operator $B$ \footnote{I changed the superscript to $k$ to
-  make this definition general---not connected to any of the terms
-  that have specific meanings later in this section.}:
-$$B^k x_i = x_{i-k}$$
-%
-with $k=1...p$ [[presumably?]].  To account for noise in the data, one
-can add a so-called ``moving average'' term to the model:
-$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
-degree $q$ and $\epsilon$ is white noise with the same mean and
-variance as the data.
-% if you leave this in, you'll need to explain a lot about the data
-% and the calculation
-% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
-To remove nonstationarities in the data, one can detrend it using a
-differencing operation: $(1-B^d)x_i$.
-
-A strategy that incorporates all three of these features is called a
-\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
- $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
-%
-{\color{red} You had superscripts on some of the $B$s and not on
-  others.  I added them but I may have gotten them wrong.  I also
-  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
-  the first use of that construction.  Please make sure things are
-  consistent, here and throughout this section.}  Here, $p$, $d$ and
-$q$ correspond to the orders of the autoregressive, detrending, and
-moving average terms, respectively.
-
-If periodic structure is present in the data, a \emph{seasonal ARIMA}
-model of order $(p,d,q)(P,D,Q)$ can be a good choice:
-%
-$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
-\Theta(B^Q)\theta(B^q)\epsilon_i$$
-%
-Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
-and $Q$, respectively [[which accomplish what purpose?  why?  how?
-    where did $D$ come from and what does it do?]].  $m$ is the
-seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
-    one should choose $P=Q=m$, since that's what the equation below
-    implicitly does.]]  To use a seasonal ARIMA model, one first uses
-the detrending term to remove nonstationarity: $$\hat{x_i} =
-(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
-evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
-\Theta(B^m)\theta(B^q)\epsilon_i$$
+is to fit a hyperplane to the dataset and then use that it to make
+predictions.  The roots of this approach date back to Yule's 1927
+invention of the autoregressive schema~\cite{weigend93}, which
+forecasts the next time step through a weighted average of past
+observations: $$px_i = \sum a_i x_i$$ The weighting coefficients $a_i$
+are computed using a polynomial of degree $p$ that is fit to the past
+$p+1$ values of $x_i$.  To account for noise in the data, one can add
+a so-called ``moving average'' term to the model; to remove
+nonstationarities, one can detrend the data using a differencing
+operation.  A strategy that incorporates all three of these features
+is called a \emph{nonseasonal ARIMA model}.  If periodic structure is
+present in the data, a \emph{seasonal ARIMA model}, which adds a
+sampling operation that filters out periodicities, can be a good
+choice.
 
 There is a vast amount of theory and literature regarding the
-construction and use of this class of models; we refer the reader to
+construction and use of models of this type; we refer the reader to
 \cite{davislinearts} for an in-depth exploration.  For the purposes of
@@
 strategies, seasonal ARIMA models are a good exemplar of the class of
-linear forecasting strategies.  In the sections that follow, we use
-the automated fitting and forecasting techniques described
-in~\cite{autoARIMA} to select model and differencing orders and fit
-the corresponding polynomials to the time series.  This procedure uses
-a combination of KPSS unit-root tests \cite{KPSSunit} and a
-customization of the Canova-Hansen test \cite{Canova1995} to select
-the differencing orders $d$ and $D$ and chooses the model order using
-the Akaike information criterion (AIC) \cite{akaike1974}, conditioned
-on the maximized likelihood of the model fitted to the detrended data,
-$(1-B^m)^D(1-B^d) x_{i,obs}$.
+linear forecasting strategies.  Fitting such a model to a dataset
+involves choosing values for the various free parameters in the
+autoregressive, detrending, moving average, and filtering
+terms---e.g., the $p$ in the equation at the beginning of this
+section.  We employ the automated fitting techniques described
+in~\cite{autoARIMA} to accomplish this.  This procedure uses
+sophisticated methods---KPSS unit-root tests~\cite{KPSSunit}, a
+customization of the Canova-Hansen test~\cite{Canova1995}, and the
+Akaike information criterion~\cite{akaike1974}, conditioned on the
+maximized likelihood of the model fitted to the detrended data---to
+select good values for the free parameters of the ARIMA model.
 
@@
 forecast is guaranteed to converge to the mean after some number of
-predictions, depending on model order.  To avoid this issue, we employ
-it in a stepwise fashion: we build the model, perform a one-step
-prediction, rebuild using the latest observations, and iterate until
-the desired prediction horizon is reached.  (For consistency, we take
-the same approach with the other three models in this study as well.)
+predictions, depending on model order.  To avoid this issue, we build
+forecasts in a stepwise fashion: i.e., fit the model to the existing
+data, use it to perform a one-step prediction, rebuild the model using
+the latest observations, and iterate until the desired prediction
+horizon is reached.  (For consistency, we take the same approach with
+the other three models in this study as well, even though that amounts
+to artificially hobbling LMA.)
 

### 2014-03-17 intro.tex
added a pageref label to intro.tex

@@
 The information in an observation can be partitioned into two pieces:
-redundancy and entropy generation\cite{crutchfield2003}.  Our approach
+redundancy and entropy generation\cite{crutchfield2003}.  
+\label{page:redundancy}
+Our approach
 exploits this decomposition in order to assess how much predictive

### 2014-03-17 model.tex
finished rewriting all but the MASE section of model.tex

@@
 the forecast: that is, the value $px_i$ at time $i$ is predicted using
-the relation $$px_i = x_{i-1}$$ The \naive ~prediction strategy
-averages the prior observations to generate the forecast: $$px_i =
-\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
-simplistic, they are not without merit.  For a time series near the
-high end of the complexity spectrum---i.e., one that possesses very
-little predictive structure---these two methods can actually be the
-best choice.  In forecasting currency exchange rates, for instance,
-sophisticated econometrics-based prediction models fail to
-consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
-These signals are volatile, noisy, and possess very little predictive
-structure, but they do not---on the average---vary a great deal, so
-the random-walk method's strategy of simply guessing the last known
-value is not a bad choice.  If a signal has a unimodal distribution
-with low variance, the \naive ~prediction strategy will perform quite
-well---even if the signal is highly complex---simply because the mean
-is a good approximation of the future behavior.  Moreover, a temporal
-average effects a low-pass filtering operation.  In a signal with very
-little predictive structure, then, the \naive ~prediction strategy
+the relation $$px_i = x_{i-1}$$ The prediction strategy that we refer
+to using the term ``\naive'' averages the prior observations to
+generate the forecast: $$px_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
+While both of these methods are simplistic, they are not without
+merit.  For a time series near the high end of the complexity
+spectrum---i.e., one that possesses very little predictive
+structure---these two methods can actually be the best choice.  In
+forecasting currency exchange rates, for instance, sophisticated
+econometrics-based prediction models fail to consistently outperform
+the random-walk method~\cite{rwMeese,rwCCE}.  These signals are
+volatile, noisy, and possess very little predictive structure, but
+they do not---on the average---vary a great deal, so the random-walk
+method's strategy of simply guessing the last known value is not a bad
+choice.  If a signal has a unimodal distribution with low variance,
+the \naive ~prediction strategy will perform quite well---even if the
+signal is highly complex---simply because the mean is a good
+approximation of the future behavior.  Moreover, a temporal average
+effects a low-pass filtering operation.  In a signal with very little
+predictive structure, then, the \naive ~prediction strategy
 effectively mitigates the complexity.
@@
 
-Perhaps the simplest way to capture and exploit the structure of data
-is to fit a hyperplane to the dataset and then use that it to make
-predictions.  The roots of this approach date back to Yule's 1927
+A simple and yet powerful way to capture and exploit the structure of
+data is to fit a hyperplane to the dataset and then use that it to
+make predictions.  The roots of this approach date back to Yule's 1927
 invention of the autoregressive schema~\cite{weigend93}, which
@@
 operation.  A strategy that incorporates all three of these features
-is called a \emph{nonseasonal ARIMA model}.  If periodic structure is
-present in the data, a \emph{seasonal ARIMA model}, which adds a
-sampling operation that filters out periodicities, can be a good
-choice.
+is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
+structure is present in the data, a \emph{seasonal ARIMA model}, which
+adds a sampling operation that filters out periodicities, can be a
+good choice.
 
@@
 horizon is reached.  (For consistency, we take the same approach with
-the other three models in this study as well, even though that amounts
+the other three models in this study as well, even though this amounts
 to artificially hobbling LMA.)
 
-\subsection{Nonlinear Forecasting: The Lorenz Method of Analogues}
+\subsection{A Nonlinear Prediction Strategy}
 \label{sec:lma}
 
-When a generating process exhibits deterministic predictive nonlinear
-structure a common approach is to reconstruct the generating processes
-dynamics using a technique called delay-coordinate embedding and then
-use the geometric structure of this reconstruction as a proxy for the
-information transmission occurring in the underlying process to
-predict the future. In this section we discuss the theory and
-implementation of this forecasting scheme.
+When the temporal progressions in a time series are produced by a
+deterministic nonlinear process, one can use a technique called
+delay-coordinate embedding
+%
+%% Note: claiming that we can reconstruct the dynamics of the
+%% underlying generating process isn't right.  That would be
+%% equivalent to solving the system identification problem!  We're
+%% not reconstructing the underlying system, just its output.
+%
+to model the structure of the information generation and transmission
+occurring in the underlying process, then use that reconstruction to
+generate forecasts.  This section discusses the theory and
+implementation of a prediction strategy that is based on this idea.
 
- \subsubsection{Reconstructing hidden dynamics {\color{blue} EDITABLE}}
-
-Delay-coordinate embedding allows one to reconstruct a dynamical system's full
-state-space dynamics from a \emph{single} scalar time-series
+Delay-coordinate embedding allows one to reconstruct a dynamical
+system's full state-space dynamics from a scalar time-series
 measurement---provided that some conditions hold regarding that data.
@@
 function---the mapping from the unknown state vector $\vec{X}$ to the
-scalar value $X_{i,obs}$ that one is measuring---are both smooth and generic,
+observed value $x_i$---are both smooth and generic,
 Takens~\cite{takens} formally proves that the delay-coordinate map
 \[
-F(\tau,m)(\vec{X}) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
+F(\tau,m)(\vec{X}) = ([x_{i} ~ x_{i+\tau} ~ \dots ~x_{i+m\tau}])
 \]
-from a $d$-dimensional smooth compact manifold $M$ to $\mathbb{R}^{2d+1}$,
-where $t$ is time, is a diffeomorphism on $M$---in other words, that
+from a $d$-dimensional smooth compact manifold $M$ to
+$\mathbb{R}^{2d+1}$ is a diffeomorphism on $M$---in other words, that
 the reconstructed dynamics and the true (hidden) dynamics have the
-same topology.
+same topology.  This is an extremely powerful result: among other
+things, it means that one can model the full system dynamics without
+measuring, or even knowing, every one of its state variables.
 
-This is an extremely powerful result: among other things, it means
-that one can build a formal model of the full system dynamics without
-measuring (or even knowing) every one of its state variables.  This is
-the foundation of this modeling approach.
-The first step in the process is to estimate values for the two free
-parameters in the delay-coordinate map: the delay $\tau$ and the
-dimension $m$.  We follow standard procedures for this, choosing the
-first minimum in the average mutual information as an estimate of
-$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
-method of \cite{KBA92} to estimate $m$.  A
-plot of the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following this
-procedure, are shown in Figure~\ref{fig:embedding}.
-
+The first step in the delay-coordinate embedding process is to
+estimate values for the two free parameters in the map: the delay
+$\tau$ and the dimension $m$.  We follow standard procedures for this,
+choosing the first minimum in the average mutual information as an
+estimate of $\tau$~\cite{fraser-swinney} and using the
+false-near-neighbor method of~\cite{KBA92} to estimate $m$.  Plots of
+the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored},
+embedded following this procedure, are shown in
+Figure~\ref{fig:embedding}.
  \begin{figure}
@@
      %\includegraphics[width=\textwidth]{figs/colipc3d.png}
-     \caption{A 3D projection of the delay-coordinate embeddings of the traces
- from: (a) \col seen in Figure~\ref{fig:col-ts} (b)\gcc seen in Figure~\ref{fig:gcc-ts} and (c) \svdfive seen as the fifth regime in Figure~\ref{fig:svd-ts-colored}.
- }
+     \caption{3D projections of delay-coordinate embeddings of the
+       traces from (a) Figure~\ref{fig:col-ts} (b)
+       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
+       Figure~\ref{fig:svd-ts-colored}.}
  \label{fig:embedding}
  \end{figure}
+%% can cut for space if need be:
+%The coordinates of each point in these plots are differently delayed
+%elements of the instructions per cycle time series $X_{i,obs}$: that
+%is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
+%$X_{i+2\tau,obs}$ on the third, and so on.  
+%
 
-%% can cut for space if need be:
-The coordinates of each point in these plots are differently delayed
-elements of the instructions per cycle time series
-$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
-$X_{i+2\tau,obs}$ on the third, and so on.
-Structure in these kinds of plots---clearly visible in
-Figure~\ref{fig:embedding} (a) is an indication of
-structure in the information transmission process, that can be leveraged for prediction.
-Where the water becomes muddy is with embedding like those in Figure~\ref{fig:embedding} (b) \& (c).
-Figure~\ref{fig:embedding} (b) shows little to no structure, this embedding is essentially a noise cloud; with an embedding like this we would expect the forecast accuracy that uses this structure to be lacking.
-However, with Figure~\ref{fig:embedding} (c) more structure is present but the structure is surrounded in noise.
-Is the structure enough to aid in accurately forecasting the signal?
-This is a difficult question to answer and one that comes back to the main contributions of this paper.
-It may be that \gcc has some redundancy but that structure is not amenable to this method, similarly \svdfive clearly has structure but it is obfuscated by noise.
-By quantifying the balance between normalized redundancy (predictive structure) and entropy for these real valued time series, we can begin to answer these questions.
-If enough (redundancy) predictive structure is present in these embeddings as seen in Figure~\ref{fig:embedding} (a) and possibly in (c), that structure can be used to build a forecast model which we discuss in the next section.
- \subsubsection{LMA: Using dynamics in forecasting {\color{blue} EDITABLE}}
+Geometric structure in these kinds of plots is an indication of
+structure in the underlying information generation/transmission
+process.  The dynamical systems community has developed a number of
+methods that leverage this structure to generate predictions
+(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  One of the
+most straightforward of these is the ``Lorenz method of analogues''
+(LMA), which is essentially nearest-neighbor prediction in the
+embedded\footnote{Lorenz's original formulation used the full system
+  state space; the method was later extended to embedded dynamics by
+  Kantz~\cite{kantz-lma-citation}.} space~\cite{lorenz-analogues}.
 
-Given a nonlinear model of a deterministic dynamical system in the
-form of a delay-coordinate embedding like those in Figure~\ref{fig:embedding},
-one can build deterministic forecast algorithms by capturing and
-exploiting the geometry of the embedding.  Many techniques have been
-developed by the dynamical systems community for this purpose
-(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  Perhaps the most straightforward
-is the ``Lorenz method of analogues'' (LMA), which is essentially
-nearest-neighbor prediction in the embedded state
-space~\cite{lorenz-analogues}.  Even this simple algorithm---which
-builds predictions by finding the nearest neighbor in the embedded space of the given point, then taking that neighbor's path as the
-forecast, provides accurate forecasts when the generating process is a deterministic dynamical system.
-In particular, when an observation in a time series has redundancy in the form of nonlinear predictive structure this forecast scheme should be a great choice.
-However, since this method is based on near-neighbor prediction in a reconstructed state-space, if a system does not exhibit deterministic structure this method can do quite poorly.
-{\color{red}[[Joshua:Not a fan of this wording]]}
-%% [[If space: Add a segue sentence: next section uses permutation
-%% entropy to explore that conjecture.]]
+Even this simple algorithm---which builds predictions by finding the
+nearest neighbor in the embedded space of the given point, then taking
+that neighbor's path as the forecast---provides accurate forecasts
+when the generating process is a deterministic dynamical system.
+Since LMA does not rest on an assumption of linearity, as ARIMA does,
+it can be applied to data produced by both linear and nonlinear
+processes.  If the underlying generating process is not deterministic,
+however, it can perform poorly.  In Figure~\ref{fig:embedding}(b), for
+instance, little structure is visible, so LMA may not work well.  More
+structure appears to be present in~Figure~\ref{fig:embedding}(c), but
+this reconstruction also appears to contain some noise.
 
-%\subsection{Comparing Prediction Accuracy}
+The question as to how much structure is present in a
+reconstruction---and how much of that structure can be captured and
+used by LMA---is apropos of the central question treated in this
+paper.  It may be that \gcc has some redundancy that LMA cannot
+exploit, or that the structure in \svdfive is effectively obfuscated,
+from the standpoint of the LMA method, by noise.  By quantifying the
+balance between normalized redundancy (that is, predictive structure)
+and entropy for these real-valued time series, we can begin to answer
+these questions.
 
-\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{blue} EDITABLE}}
+% If enough predictive structure is present in these embeddings, that
+% structure can be used to build an effective forecast model.
+
+\subsection{Assessing Prediction Accuracy}
 \label{sec:accuracy}
 
-In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:model}.
-The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary.
-In particular, we perform $k$ one-step predictions, after each one-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction.
-This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
+To study the relationship between predictability and complexity, we
+use these four methods to generate predictions of all 360 traces
+described in Section~\ref{sec:methods}, then calculate the error
+between the predictions and the true continuations.  Specifically, we
+split each time series into two pieces: the first 90\%, referred to as
+the ``initial training" signal and denoted $\{x_i\}_{i=1}^{n}$, and
+the last 10\% known as the ``test" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
+The initial training signal is used to build the model, following the
+procedures described in the previous section; that model is used to
+generate a prediction of the value of $x_{n+1}$, which is then
+compared to the true continuation, $c_{n+1}$.  The model is then
+rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process repeats $k$ times,
+out to the end of the observed time series.  This ``one step
+prediction'' process is not technically necessary in the LMA method,
+which can generate arbitrary-length predictions, but the performance
+of the other three methods used here will degrade if the associated
+models are not periodically rebuilt.  In order to make the comparison
+fair, we used an iterative one-step prediction schema {\sl for all
+  four methods}.  This has the slightly confusing effect of causing
+the ``test'' signal to be used both to assess the accuracy of each
+model and for periodic refitting.
 
-As a figure of merit we calculate the Mean Absolute Scaled Error (MASE)\cite{MASE} between the true and predicted signals: 
-%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
-The scaling term for MASE:
-$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$.
-This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." We chose MASE because it allows for fair comparison across varying methods, prediction horizons and signal scales.
-When a forecast results in a $MASE<1$ the prediction method produced, on average, smaller errors than the one-step errors from the in-sample random walk forecast strategy.
-Analogously, $MASE>1$ means that the prediction method did worse, on average than the one-step errors for the in-sample random walk forecast strategy.
+To assess prediction accurary, we calculate the mean absolute scaled
+error (MASE) between the true and predicted signals:
+% 
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
+% 
+This error method was introduced in \cite{MASE} as a ``generally
+applicable measurement of forecast accuracy without the problems seen
+in the other measurements."  Note that the scaling term in the
+denominator
+% 
+% of the MASE value
+% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
+% 
+is the average in-sample forecast error for a random-walk prediction
+$px_i=x_{i-1}$ over a signal of length $k$.  That is, MASE is a
+normalized measure.  A MASE value below one means that the prediction
+error was, on the average, smaller than the error of a one-step
+in-sample random-walk forecast on the same data.  Analogously, a MASE
+value greater than one means that the corresponding prediction method
+did \emph{worse}, on average, than a one-step in-sample random-walk
+method.  We chose this error metric because it allows for fair
+comparison across varying methods, prediction horizons, and signal
+scales.
 

### 2014-03-18 model.tex
finished the model.tex rewrite

@@
 
-In this section, we describe the four different forecasting schema
+In this section, we describe the four different forecasting methods
 used in this study, as well as the error metric used to evaluate their
@@
 to capture and exploit the deterministic structure of a signal from a
-nonlinear dynamical system.  The first two methods, somewhat
-surprisingly, often outperform these more-sophisticated prediction
-strategies in the case of highly complex signals, as discussed below.
+nonlinear dynamical system.  The \naive and random-walk methods,
+somewhat surprisingly, often outperform these more-sophisticated
+prediction strategies in the case of highly complex signals, as
+discussed below.
 
@@
 A random-walk predictor simply uses the last observed measurement as
-the forecast: that is, the value $px_i$ at time $i$ is predicted using
-the relation $$px_i = x_{i-1}$$ The prediction strategy that we refer
-to using the term ``\naive'' averages the prior observations to
+the forecast: that is, the value $px_i$ at time $i$ is calculated
+using the relation $$px_i = x_{i-1}$$ The prediction strategy that we
+refer to using the term ``\naive'' averages the prior observations to
 generate the forecast: $$px_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
@@
 signal is highly complex---simply because the mean is a good
-approximation of the future behavior.  Moreover, a temporal average
-effects a low-pass filtering operation.  In a signal with very little
-predictive structure, then, the \naive ~prediction strategy
-effectively mitigates the complexity.
+approximation of the future behavior.  Moreover, the \naive
+~prediction strategy's temporal average effects a low-pass filtering
+operation, which effectively mitigates the complexity in signals with
+very little predictive structure.
 
@@
 situations.  Random-walk strategies are a particularly bad choice for
-time series that change drastically at every time step.  In the worst
-case---a large-amplitude square wave whose period is equivalent to
-twice the sample time---a random-walk prediction would be exactly 180
-degrees out of phase with the true continuation.  The \naive ~method
-would be a better choice in this situation, since it would always
-predict the mean.  It would, however, perform poorly when a signal has
-a number of long-lived regimes that have significantly different
-means.  In this situation, the inertia of the \naive ~method's
-accumulating mean is a liability and the agility of the random-walk
-method is an advantage, since it can respond quickly to regime shifts.
+time series that change significantly at every time step.  In the
+worst case---a large-amplitude square wave whose period is equivalent
+to twice the sample time---a random-walk prediction would be exactly
+180 degrees out of phase with the true continuation.  The \naive
+~method would be a better choice in this situation, since it would
+always predict the mean.  It would, however, perform poorly when a
+signal has a number of long-lived regimes that have significantly
+different means.  In this situation, the inertia of the \naive
+~method's accumulating mean is a liability and the agility of the
+random-walk method is an advantage, since it can respond quickly to
+regime shifts.
 
@@
 situations described in the previous paragraph.  The ARIMA and LMA
-methods introduced in Sections~\ref{sec:lma} and~\ref{sec:arima} are
+methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
 designed to do exactly that.  Note, though, that periodic patterns
@@
 forecasts the next time step through a weighted average of past
-observations: $$px_i = \sum a_i x_i$$ The weighting coefficients $a_i$
-are computed using a polynomial of degree $p$ that is fit to the past
-$p+1$ values of $x_i$.  To account for noise in the data, one can add
-a so-called ``moving average'' term to the model; to remove
-nonstationarities, one can detrend the data using a differencing
-operation.  A strategy that incorporates all three of these features
-is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
-structure is present in the data, a \emph{seasonal ARIMA model}, which
-adds a sampling operation that filters out periodicities, can be a
-good choice.
+observations: $$px_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
+coefficients $a_i$ are computed using a polynomial of degree $p$ that
+is fit to the past $p+1$ values of $x_i$.  To account for noise in the
+data, one can add a so-called ``moving average'' term to the model; to
+remove nonstationarities, one can detrend the data using a
+differencing operation.  A strategy that incorporates all three of
+these features is called a \emph{nonseasonal ARIMA model}.  If
+evidence of periodic structure is present in the data, a
+\emph{seasonal ARIMA model}, which adds a sampling operation that
+filters out periodicities, can be a good choice.
 
@@
 strategies, seasonal ARIMA models are a good exemplar of the class of
-linear forecasting strategies.  Fitting such a model to a dataset
-involves choosing values for the various free parameters in the
-autoregressive, detrending, moving average, and filtering
-terms---e.g., the $p$ in the equation at the beginning of this
-section.  We employ the automated fitting techniques described
-in~\cite{autoARIMA} to accomplish this.  This procedure uses
-sophisticated methods---KPSS unit-root tests~\cite{KPSSunit}, a
-customization of the Canova-Hansen test~\cite{Canova1995}, and the
-Akaike information criterion~\cite{akaike1974}, conditioned on the
-maximized likelihood of the model fitted to the detrended data---to
-select good values for the free parameters of the ARIMA model.
+linear predictors.  Fitting such a model to a dataset involves
+choosing values for the various free parameters in the autoregressive,
+detrending, moving average, and filtering terms---e.g., the $p$ in the
+equation at the beginning of this section.  We employ the automated
+fitting techniques described in~\cite{autoARIMA} to accomplish this.
+This procedure uses sophisticated methods---KPSS unit-root
+tests~\cite{KPSSunit}, a customization of the Canova-Hansen
+test~\cite{Canova1995}, and the Akaike information
+criterion~\cite{akaike1974}, conditioned on the maximized likelihood
+of the model fitted to the detrended data---to select good values for
+the free parameters of the ARIMA model.
 
@@
 forecast is guaranteed to converge to the mean after some number of
-predictions, depending on model order.  To avoid this issue, we build
-forecasts in a stepwise fashion: i.e., fit the model to the existing
-data, use it to perform a one-step prediction, rebuild the model using
-the latest observations, and iterate until the desired prediction
-horizon is reached.  (For consistency, we take the same approach with
-the other three models in this study as well, even though this amounts
-to artificially hobbling LMA.)
+predictions, depending on model order.  To sidestep this issue, we
+build forecasts in a stepwise fashion: i.e., fit the model to the
+existing data, use it to perform a one-step prediction, rebuild the
+model using the latest observations, and iterate until the desired
+prediction horizon is reached.  (For consistency, we take the same
+approach with the other three models in this study as well, even
+though this amounts to artificially hobbling LMA.)
 
@@
 
-Delay-coordinate embedding allows one to reconstruct a dynamical
-system's full state-space dynamics from a scalar time-series
-measurement---provided that some conditions hold regarding that data.
-Specifically, if the underlying dynamics and the measurement
-function---the mapping from the unknown state vector $\vec{X}$ to the
-observed value $x_i$---are both smooth and generic,
-Takens~\cite{takens} formally proves that the delay-coordinate map
+Delay-coordinate embedding~\cite{packard80,Sauer:1991lr,Takens:1981uq}
+allows one to reconstruct a dynamical system's full state-space
+dynamics from a scalar time-series measurement---provided that some
+conditions hold regarding that data.  Specifically, if the underlying
+dynamics and the measurement function---the mapping from the unknown
+state vector $\vec{X}$ to the observed value $x_i$---are both smooth
+and generic, Takens~\cite{Takens:1981uq} formally proves that the
+delay-coordinate map
 \[
@@
 Geometric structure in these kinds of plots is an indication of
-structure in the underlying information generation/transmission
-process.  The dynamical systems community has developed a number of
-methods that leverage this structure to generate predictions
-(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  One of the
-most straightforward of these is the ``Lorenz method of analogues''
-(LMA), which is essentially nearest-neighbor prediction in the
-embedded\footnote{Lorenz's original formulation used the full system
-  state space; the method was later extended to embedded dynamics by
-  Kantz~\cite{kantz-lma-citation}.} space~\cite{lorenz-analogues}.
+structure in the information generation/transmission process the
+produced the time series.  The dynamical systems community has
+developed a number of methods that leverage this structure to generate
+predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
+One of the most straightforward of these is the ``Lorenz method of
+analogues'' (LMA), which is essentially nearest-neighbor prediction in
+the embedded\footnote{Lorenz's original formulation used the full
+  system state space; the method was later extended to embedded
+  dynamics by Kantz~\cite{kantz-lma-citation}.}
+space~\cite{lorenz-analogues}.  Even this simple algorithm---which
+builds predictions by finding the nearest neighbor in the embedded
+space of the given point, then taking that neighbor's path as the
+prediction---provides accurate forecasts when the generating process
+is a deterministic dynamical system.
 
-Even this simple algorithm---which builds predictions by finding the
-nearest neighbor in the embedded space of the given point, then taking
-that neighbor's path as the forecast---provides accurate forecasts
-when the generating process is a deterministic dynamical system.
-Since LMA does not rest on an assumption of linearity, as ARIMA does,
-it can be applied to data produced by both linear and nonlinear
-processes.  If the underlying generating process is not deterministic,
-however, it can perform poorly.  In Figure~\ref{fig:embedding}(b), for
-instance, little structure is visible, so LMA may not work well.  More
-structure appears to be present in~Figure~\ref{fig:embedding}(c), but
-this reconstruction also appears to contain some noise.
-
-The question as to how much structure is present in a
-reconstruction---and how much of that structure can be captured and
-used by LMA---is apropos of the central question treated in this
-paper.  It may be that \gcc has some redundancy that LMA cannot
-exploit, or that the structure in \svdfive is effectively obfuscated,
-from the standpoint of the LMA method, by noise.  By quantifying the
-balance between normalized redundancy (that is, predictive structure)
-and entropy for these real-valued time series, we can begin to answer
+Since LMA does not rest on an assumption of linearity (as ARIMA does)
+it can handle both linear and nonlinear processes.  If the underlying
+generating process is nondeterministic, however, it can perform
+poorly.  In Figure~\ref{fig:embedding}(b), for instance, little
+structure is visible, so LMA may not work well.  More structure
+appears to be present in~Figure~\ref{fig:embedding}(c), but this
+reconstruction also appears to contain some noise.  The question as to
+how much structure is present in a reconstruction---and how much of
+that structure can be captured and used by LMA---is apropos of the
+central question treated in this paper.  It may be that \gcc has some
+redundancy that LMA cannot exploit, or that the structure in \svdfive
+is effectively obfuscated, from the standpoint of the LMA method, by
+noise.  By quantifying the balance between normalized redundancy (that
+is, predictive structure) and entropy for these real-valued time
+series, as shown in Section~\ref{sec:results}, we can begin to answer
 these questions.
@@
 To study the relationship between predictability and complexity, we
-use these four methods to generate predictions of all 360 traces
-described in Section~\ref{sec:methods}, then calculate the error
-between the predictions and the true continuations.  Specifically, we
-split each time series into two pieces: the first 90\%, referred to as
-the ``initial training" signal and denoted $\{x_i\}_{i=1}^{n}$, and
-the last 10\% known as the ``test" signal $\{c_j\}_{j=n+1}^{k+n+1}$.
-The initial training signal is used to build the model, following the
-procedures described in the previous section; that model is used to
-generate a prediction of the value of $x_{n+1}$, which is then
-compared to the true continuation, $c_{n+1}$.  The model is then
-rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process repeats $k$ times,
-out to the end of the observed time series.  This ``one step
-prediction'' process is not technically necessary in the LMA method,
-which can generate arbitrary-length predictions, but the performance
-of the other three methods used here will degrade if the associated
-models are not periodically rebuilt.  In order to make the comparison
-fair, we used an iterative one-step prediction schema {\sl for all
-  four methods}.  This has the slightly confusing effect of causing
-the ``test'' signal to be used both to assess the accuracy of each
-model and for periodic refitting.
+use the four methods outlined above to generate predictions of all 360
+traces described in Section~\ref{sec:methods}, then calculate the
+error between the predictions and the true continuations.
+Specifically, we split each time series into two pieces: the first
+90\%, referred to as the ``initial training" signal and denoted
+$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
+$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
+build the model, following the procedures described in the previous
+section; that model is used to generate a prediction of the value of
+$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
+The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
+repeats $k$ times, out to the end of the observed time series.  This
+``one step prediction'' process is not technically necessary in the
+LMA method, which can generate arbitrary-length predictions, but the
+performance of the other three methods used here will degrade if the
+associated models are not periodically rebuilt.  In order to make the
+comparison fair, we used an iterative one-step prediction schema {\sl
+  for all four methods}.  This has the slightly confusing effect of
+causing the ``test'' signal to be used both to assess the accuracy of
+each model and for periodic refitting.
 
-To assess prediction accurary, we calculate the mean absolute scaled
-error (MASE) between the true and predicted signals:
-% 
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
-% 
-This error method was introduced in \cite{MASE} as a ``generally
-applicable measurement of forecast accuracy without the problems seen
-in the other measurements."  Note that the scaling term in the
-denominator
-% 
-% of the MASE value
-% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
-% 
-is the average in-sample forecast error for a random-walk prediction
-$px_i=x_{i-1}$ over a signal of length $k$.  That is, MASE is a
-normalized measure.  A MASE value below one means that the prediction
-error was, on the average, smaller than the error of a one-step
-in-sample random-walk forecast on the same data.  Analogously, a MASE
-value greater than one means that the corresponding prediction method
-did \emph{worse}, on average, than a one-step in-sample random-walk
-method.  We chose this error metric because it allows for fair
-comparison across varying methods, prediction horizons, and signal
-scales.
-
-Table \ref{tab:error} provides the MASE distributions {\color{red}[[Joshua: Ryan, Is this the right word?
-we give mean $\pm$ std.
-dev but some have very skewed right tails]]} for each of the eight signals and three prediction strategies, these are averaged over 15 runs of each signal-method combination.
-For later comparison, Table \ref{tab:error} also has the complexity measure we introduce in Section  \ref{sec:meaComplex}.
-
-To aid in a conceptual understanding of this error metric we provide example forecasts in Figure~\ref{fig:forecast-example}.
-In plots like this we plot $(c_j,p_j)$, visually points close to the line $y=x$, i.e.
-the line $p_j = c_j$ are more accurate.
-Forecasts which stay close to the line are more accurate than forecasts that stray away from it.
-
-\begin{figure}[htbp]
+Figure~\ref{fig:forecast-example} shows example forecasts made using
+the \naive, ARIMA, and LMA methods for the \col, \gcc, and \svdfive
+time series.
+\begin{figure*}[htbp]
   \centering
-      \begin{subfigure}{0.33\columnwidth}
+      \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/colMeanForecast.png}
@@
   \end{subfigure}%
-   \begin{subfigure}{0.32\columnwidth}
+   \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/gccMeanForecast.png}
@@
   \end{subfigure}% 
-     \begin{subfigure}{0.32\columnwidth}
+     \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast.png}
@@
   \\
-    \begin{subfigure}{0.32\columnwidth}
+    \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/colARIMAForecast.png}
 
-    \caption{\col ARIMA}
+    \caption{\col \\ ARIMA}
     \label{fig:colARIMA}
   \end{subfigure}
-  \begin{subfigure}{0.32\columnwidth}
+  \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/gccARIMAForecast.png}
@@
   \end{subfigure}%
-  \begin{subfigure}{0.32\columnwidth}
+  \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast.png}
@@
 
-      \begin{subfigure}{0.32\columnwidth}
+      \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
-    \caption{\col LMA}
+    \caption{\col \\ LMA}
     \label{fig:colLMA}
   \end{subfigure}
-      \begin{subfigure}{0.32\columnwidth}
+      \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
-    \caption{\gcc LMA}
+    \caption{\gcc \\ LMA}
     \label{fig:gccLMA}
   \end{subfigure}  
-        \begin{subfigure}{0.32\columnwidth}
+        \begin{subfigure}{0.6\columnwidth}
     \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
-    \caption{\svdfive LMA}
+    \caption{\svdfive \\ LMA}
     \label{fig:gccLMA}
@@
   %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{
-{\color{red}[[Joshua: Liz, I added \svdfive as it has been a persistent example up to this figure.
-I am a little worried about the 9x9 being too small in 1 column.
-We may need to make this figure span two columns, which I think is ok.]]}For each of these, we plot the predicted value $p_j$ against the correct value $c_j$.
-On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_j = c_j$, e.g., (g) is a near perfect prediction where-as (e) is a poor prediction.}\label{fig:forecast-example}
+   \caption{Predicted ($px_j$) versus true values ($c_j$) for
+     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
+     perfect prediction lies exactly on the diagonal.
+%
+{\color{red} Joshua, what would you think of adding a row of
+  random-walk plots at the top?  The letters in the subcaptions---(a)
+  (b) etc.---feel clunky here.  Finally, we should eventually change
+  the label of the vertical axes on these plots to be whatever symbol
+  we choose for the prediction.}
+%
+}
+\label{fig:forecast-example}
   %\end{subfigure}
-\end{figure}
+\end{figure*} 
+In these images, the vertical axis is the prediction $px_j$ and the
+horizontal axis is the true continuation $c_j$.  On such a plot, a
+perfect prediction would lie on the diagonal.  LMA, for instance,
+generates a very accurate prediction of the \col data, while ARIMA
+does not.  [[We should comment on what a noise ball means on such a
+    plot, as well as what a horizontal line means.]]
 
-It may be helpful for the reader at this point to compare some of the MASE scores in Table \ref{tab:error} with the example forecasts in Figure~\ref{fig:forecast-example}.
-For example, if we compare Figures~ \ref{fig:forecast-example} (d) and (g), \col is almost perfectly predicted by LMA (close to the line $y=x$ and ARIMA is a poor predictor as the points are spread out around $y=x$.
-Comparing to the MASE values in Table~\ref{tab:error} we see that \col with LMA had a MASE of $0.05 \pm 0.002$, whereas \col with ARIMA had a MASE of $0.599 \pm 0.211$.
-So intuitively the better visual prediction received a much better MASE score (the smaller the better).
-One way of interpreting these are that LMA on \col is on average 20 times more effective than random walk and ARIMA on \col is only 2 times more effective than random walk.
+As a numerical measure of prediction accurary, we calculate the mean
+absolute scaled error ($MASE$) between the true and predicted signals:
+% 
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-x_j|
+}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
+% 
+This error metric was introduced in \cite{MASE} as a ``generally
+applicable measurement of forecast accuracy without the problems seen
+in the other measurements."  $MASE$ is a normalized measure; the
+scaling term in the denominator
+% 
+% of the MASE value
+% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
+% 
+is the average in-sample forecast error for a random-walk prediction
+over a signal of length $k$.  That is, $MASE<1$ means that the
+prediction error in question was, on the average, smaller than the
+error of a random-walk forecast on the same data.  Analogously,
+$MASE>1$ means that the corresponding prediction method did
+\emph{worse}, on average, than the random-walk method.  We chose this
+error metric because it allows for fair comparison across varying
+methods, prediction horizons, and signal scales.
 
-The part that becomes tricky and which must be kept in mind is that this error metric is in comparison to random walk.
-If we compare visually for example the LMA prediction of \col and \svdfive (Figures~\ref{fig:forecast-example} (g) \& (i)) visually they look very similar in terms of accuracy; however, \col was forecast 20 times better than random walk where as \svdfive was forecast only 1.5 times better.
-The reason for this lies in the strengths/weaknesses of Random Walk discussed in Section~\ref{sec:simple}.
-The large oscillations in \col cause Random Walk to do poorly, MASE then gives a model that accounts for this oscillation (e.g., LMA) a boost, whereas \svdfive is a great signal for Random Walk so there is less improvement to be made, and less boost is given for example for correctly reacting to regime shift.
-A more in-depth discussion of this is presented in Section~\ref{sec:results}.
+$MASE$ scores for all 360 experiments are tabulated in the three
+middle columns in Table~\ref{tab:error}.
+\begin{table*}
+  \begin{center}
+  \begin{tabular}{|c|c|c|c|c|}
+  \hline Signal & MASE LMA & MASE ARIMA &MASE na\"{i}ve & Weighted Permutation \\ 
+  &  &  & &  Entropy \\ \hline
+
+ \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\
+
+\gcc           & $ 1.530\pm 0.021$ & $1.837 \pm0.016 $ & $0.951 \pm 0.001$ & $0.943 \pm 0.001$ \\
+
+\svdone     & $ 0.827\pm 0.076$ & $ 0.714\pm 0.075 $ & $2.676\pm4.328$&  $0.957 \pm 0.016$ \\
+
+ \svdtwo    & $1.279 \pm0.020 $ & $2.163 \pm0.027 $ &  $3.054\pm0.040$ &   $0.846 \pm0.004$ \\
+
+ \svdthree     & $0.619 \pm0.021 $ & $0.713 \pm 0.010 $ & $31.386\pm 0.282$ &  $0.716 \pm 0.006$ \\
+
+ \svdfour     & $ 0.779\pm0.036 $ & $0.979 \pm0.032 $ & $2.661\pm0.074$ & $0.825 \pm 0.008$ \\
+
+ \svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
+
+ \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
+  \hline
+  \end{tabular}
+  \end{center}
+ \label{default}
+ \caption{MASE scores and weighted permutation entropies for all eight
+   processes studied in this paper}
+ \label{tab:error}
+  \end{table*}%
+Comparing these with the geometry of the plots in
+Figure~\ref{fig:forecast-example}, one can see some obvious
+correspondences.  The average LMA $MASE$ score for the \col signals
+was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
+That is, LMA performed roughly 20 times better on \col signals than a
+random-walk predictor, while ARIMA only outperformed random walk by a
+factor of 1.7 [[is this right?  there was a ``2'' here, but 1/0.599 is
+    closer to 1.7]].  This is in accordance with the visual appearance
+of the corresponding images in Figure~\ref{fig:forecast-example}.
+While the correspondence between $MASE$ score and the visual
+appearance of a $px$ versus $c$ plot is clear cut in this case, that
+is not always so.  The plots of LMA predictions of \col and \svdfive
+both lie near the diagonal, for instance, but the corresponding $MASE$
+scores were very different: $0.050$ for \col and $0.718$ for \svdfive
+(resp., 20 and 1.4 times better than random-walk forecasts of the same
+signals).  This is a result of the normalization in the $MASE$ score.
+Recall from Section~\ref{sec:simple} that the random-walk method
+performs especially poorly on signals that oscillate rapidly.  \col
+fits this description, so the random-walk error on this signal---the
+denominator of the three $MASE$ scores in the first row of the
+Table---is pathologically high, which skews those scores down.
+Random-walk prediction is very effective for the\svdfive signal, on
+the other hand, so the denominator is small and the $MASE$ scores are
+skewes upwards.  Normalization is a two-edged sword here.  It
+facilitates comparisons across signals of different types and lengths,
+but its particular pathologies must be taken into account, as
+discussed at more length in Section~\ref{sec:results}.
+
+% Table \ref{tab:error} provides the $MASE$ distributions
+% {\color{red}[[Joshua: Ryan, Is this the right word? we give mean $\pm$
+%       std. dev but some have very skewed right tails]]} for each of
+% the eight signals and three prediction strategies, these are averaged
+% over 15 runs of each signal-method combination. For later comparison,
+% Table \ref{tab:error} also has the complexity measure we introduce in
+% Section \ref{sec:meaComplex}.
 
@@
 So that we don't have to compare the forecast accuarcy by visual comparing the predictions we calculate a figure of merit to compare the predictions.
-
-% took out for space

### 2014-03-22 intro.tex
minor formatting to intro.tex

@@
 
-As a corollary of the undecidability of the halting problem
-\cite{halting-problem}, however, no single forecasting schema is ideal
-for all noise-free deterministic signals\cite{weigend-book}---let
-alone all real-world time-series data sets.  This leads naturally to
-an important and challenging question: given a complicated real-valued
-time series, does there exist a forecast model that can utilize the
-information (if any) that is being transmitted from past to future by
-the underlying generating process?  A first step in answering this
-question is to reliably quantify where on the complexity spectrum a
-given time series falls; a second step is to determine how complexity
-and predictability are related.  These are the goals of this paper.
+As a corollary of the undecidability of the halting
+problem~\cite{halting-problem}, however, no single forecasting schema
+is ideal for all noise-free deterministic
+signals~\cite{weigend-book}---let alone all real-world time-series
+data sets.  This leads naturally to an important and challenging
+question: given a complicated real-valued time series, does there
+exist a forecast model that can utilize the information (if any) that
+is being transmitted from past to future by the underlying generating
+process?  A first step in answering this question is to reliably
+quantify where on the complexity spectrum a given time series falls; a
+second step is to determine how complexity and predictability are
+related.  These are the goals of this paper.
 
@@
 The information in an observation can be partitioned into two pieces:
-redundancy and entropy generation\cite{crutchfield2003}.  
+redundancy and entropy generation~\cite{crutchfield2003}.  
 \label{page:redundancy}
-Our approach
-exploits this decomposition in order to assess how much predictive
-structure is present in a signal---i.e., where it falls on the
-complexity spectrum mentioned above.  We define \emph{complexity} as a
-particular approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.
-That is, we view a random-walk time series (which exhibits high
-entropy) as purely complex, whereas a periodic signal, which exhibits
-low entropy, is on the low end of the complexity spectrum.  We argue
-that an extension of \emph{permutation entropy}
-\cite{bandt2002per}---a method for approximating the entropy through
-ordinal analysis---is an effective way to assess the complexity of a
-given time series.  Permutation entropy is ideal for our purposes
-because it works with real-valued data and is known to converge to the true entropy value. Other
-existing techniques either require specific knowledge of the
-generating process or produce biased values of the
-entropy~\cite{bollt2001}.
+Our approach exploits this decomposition in order to assess how much
+predictive structure is present in a signal---i.e., where it falls on
+the complexity spectrum mentioned above.  We define \emph{complexity}
+as a particular approximation of Kolmogorov-Sinai
+entropy~\cite{KS-entropy}.  That is, we view a random-walk time series
+(which exhibits high entropy) as purely complex, whereas a periodic
+signal, which exhibits low entropy, is on the low end of the
+complexity spectrum.  We argue that an extension of \emph{permutation
+  entropy}~\cite{bandt2002per}---a method for approximating the
+entropy through ordinal analysis---is an effective way to assess the
+complexity of a given time series.  Permutation entropy is ideal for
+our purposes because it works with real-valued data and is known to
+converge to the true entropy value. Other existing techniques either
+require specific knowledge of the generating process or produce biased
+values of the entropy~\cite{bollt2001}.
 
@@
 Although this time series appears to be periodic, it is actually
-chaotic \cite{mytkowicz09}, which places fundamental limits on
+chaotic~\cite{mytkowicz09}, which places fundamental limits on
 predictability.  Changes in the program can radically change this

### 2014-03-22 model.tex
px to p in model.tex

@@
 A random-walk predictor simply uses the last observed measurement as
-the forecast: that is, the value $px_i$ at time $i$ is calculated
-using the relation $$px_i = x_{i-1}$$ The prediction strategy that we
+the forecast: that is, the value $p_i$ at time $i$ is calculated
+using the relation $$p_i = x_{i-1}$$ The prediction strategy that we
 refer to using the term ``\naive'' averages the prior observations to
-generate the forecast: $$px_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
+generate the forecast: $$p_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
 While both of these methods are simplistic, they are not without
@@
 %of the autoregressive schema~\cite{weigend93}, which forecasts the
-%next time step through a weighted average of past observations: $$px_i
+%next time step through a weighted average of past observations: $$p_i
 %= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
@@
 %the detrending term to remove nonstationarity: $$\hat{x_i} =
-%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
+%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $p_i$ by
 %evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
@@
 forecasts the next time step through a weighted average of past
-observations: $$px_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
+observations: $$p_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
 coefficients $a_i$ are computed using a polynomial of degree $p$ that
@@
   %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
-   \caption{Predicted ($px_j$) versus true values ($c_j$) for
+   \caption{Predicted ($p_j$) versus true values ($c_j$) for
      forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
@@
 \end{figure*} 
-In these images, the vertical axis is the prediction $px_j$ and the
+In these images, the vertical axis is the prediction $p_j$ and the
 horizontal axis is the true continuation $c_j$.  On such a plot, a
@@
 While the correspondence between $MASE$ score and the visual
-appearance of a $px$ versus $c$ plot is clear cut in this case, that
+appearance of a $p$ versus $c$ plot is clear cut in this case, that
 is not always so.  The plots of LMA predictions of \col and \svdfive

### 2014-03-22 model.tex
making notation consistent in model.tex

@@
 observations: $$p_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
-coefficients $a_i$ are computed using a polynomial of degree $p$ that
-is fit to the past $p+1$ values of $x_i$.  To account for noise in the
+coefficients $a_i$ are computed using a polynomial of degree $l$ that
+is fit to the past $l+1$ values of $x_i$.  To account for noise in the
 data, one can add a so-called ``moving average'' term to the model; to
@@
 choosing values for the various free parameters in the autoregressive,
-detrending, moving average, and filtering terms---e.g., the $p$ in the
+detrending, moving average, and filtering terms---e.g., the $l$ in the
 equation at the beginning of this section.  We employ the automated
@@
 {\color{red} Joshua, what would you think of adding a row of
-  random-walk plots at the top?  The letters in the subcaptions---(a)
-  (b) etc.---feel clunky here.  Finally, we should eventually change
-  the label of the vertical axes on these plots to be whatever symbol
-  we choose for the prediction.}
+  random-walk plots at the top?  Also, the letters in the
+  subcaptions---(a) (b) etc.---feel clunky here.  What would you think
+  of removing them?}
 %
@@
 generates a very accurate prediction of the \col data, while ARIMA
-does not.  [[We should comment on what a noise ball means on such a
-    plot, as well as what a horizontal line means.]]
+does not.  {\color{red}[[We should comment on what a noise ball means
+      on such a plot, as well as what a horizontal line means.]]}
 
@@
 % 
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-x_j|
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-x_j|
 }{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
 % 
+{\color{red} That equation had $c_j$ instead of $p_j$ in the numerator.
+I changed it.  Let me know if that's wrong.}
 This error metric was introduced in \cite{MASE} as a ``generally
@@
 random-walk predictor, while ARIMA only outperformed random walk by a
-factor of 1.7 [[is this right?  there was a ``2'' here, but 1/0.599 is
-    closer to 1.7]].  This is in accordance with the visual appearance
-of the corresponding images in Figure~\ref{fig:forecast-example}.
-While the correspondence between $MASE$ score and the visual
-appearance of a $p$ versus $c$ plot is clear cut in this case, that
-is not always so.  The plots of LMA predictions of \col and \svdfive
-both lie near the diagonal, for instance, but the corresponding $MASE$
-scores were very different: $0.050$ for \col and $0.718$ for \svdfive
-(resp., 20 and 1.4 times better than random-walk forecasts of the same
-signals).  This is a result of the normalization in the $MASE$ score.
-Recall from Section~\ref{sec:simple} that the random-walk method
-performs especially poorly on signals that oscillate rapidly.  \col
-fits this description, so the random-walk error on this signal---the
-denominator of the three $MASE$ scores in the first row of the
-Table---is pathologically high, which skews those scores down.
-Random-walk prediction is very effective for the\svdfive signal, on
-the other hand, so the denominator is small and the $MASE$ scores are
-skewes upwards.  Normalization is a two-edged sword here.  It
-facilitates comparisons across signals of different types and lengths,
-but its particular pathologies must be taken into account, as
+factor of 1.7 {\color{red}[[is this right?  there was a ``2'' here,
+      but 1/0.599 is closer to 1.7]]}.  This is in accordance with the
+visual appearance of the corresponding images in
+Figure~\ref{fig:forecast-example}.  While the correspondence between
+$MASE$ score and the visual appearance of a $p$ versus $c$ plot is
+clear cut in this case, that is not always so.  The plots of LMA
+predictions of \col and \svdfive both lie near the diagonal, for
+instance, but the corresponding $MASE$ scores were very different:
+$0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times
+better than random-walk forecasts of the same signals).  This is a
+result of the normalization in the $MASE$ score.  Recall from
+Section~\ref{sec:simple} that the random-walk method performs
+especially poorly on signals that oscillate rapidly.  \col fits this
+description, so the random-walk error on this signal---the denominator
+of the three $MASE$ scores in the first row of the Table---is
+pathologically high, which skews those scores down.  Random-walk
+prediction is very effective for the \svdfive signal, on the other
+hand, so the denominator is small and the $MASE$ scores are skewes
+upwards.  Normalization is, as is often the case, a two-edged sword.
+Here, it facilitates comparisons across signals of different types and
+lengths, but its particular pathologies must be taken into account, as
 discussed at more length in Section~\ref{sec:results}.

### 2014-03-24 model.tex
removed duplicated Yule wording from model.tex

@@
 data is to fit a hyperplane to the dataset and then use that it to
-make predictions.  The roots of this approach date back to Yule's 1927
-invention of the autoregressive schema~\cite{weigend93}, which
-forecasts the next time step through a weighted average of past
-observations: $$p_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
-coefficients $a_i$ are computed using a polynomial of degree $l$ that
-is fit to the past $l+1$ values of $x_i$.  To account for noise in the
-data, one can add a so-called ``moving average'' term to the model; to
-remove nonstationarities, one can detrend the data using a
-differencing operation.  A strategy that incorporates all three of
-these features is called a \emph{nonseasonal ARIMA model}.  If
-evidence of periodic structure is present in the data, a
-\emph{seasonal ARIMA model}, which adds a sampling operation that
-filters out periodicities, can be a good choice.
+make predictions.  The roots of this approach date back to the
+original autoregressive schema~\cite{weigend93}, which forecasts the
+next time step through a weighted average of past observations: $$p_i
+= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_i$ are
+computed using a polynomial of degree $l$ that is fit to the past
+$l+1$ values of $x_i$.  To account for noise in the data, one can add
+a so-called ``moving average'' term to the model; to remove
+nonstationarities, one can detrend the data using a differencing
+operation.  A strategy that incorporates all three of these features
+is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
+structure is present in the data, a \emph{seasonal ARIMA model}, which
+adds a sampling operation that filters out periodicities, can be a
+good choice.
 

### 2014-03-24 related.tex
beginning of rewrite of related.tex

@@
 \section{Related Work }\label{sec:related}
-{\color{blue}[[EDITABLE]]}
-Attempting to model time series for prediction is an old problem which arguably began with Yule in 1927 when he invented autoregression \cite{Yule27}.
-Since then hundreds if not thousands of prediction strategies have been developed for a wide variety of prediction tasks.
-The purpose of this paper is \emph{not} to find an ideal forecast model for a time series but to quantify empirically the predictive structure present in a real-valued-scalar time series model free and then to associate the prediction accuracy of a set of methods with the time series inherent complexity.
-For this reason we do not cover in-depth the spectrum of previous work in predictive modeling, we do however provide an overview of the models used in this paper in  Section~\ref{sec:model}.
-Quantifying predictability, or what is sometimes called ``predicting predictability", is also not a new problem and most solutions fall into two categories: model-based error analysis and model-free information analysis.
 
+Modeling time-series data for the purposes of prediction is an old
+problem that dates back at least to Yule's 1927 invention of
+autoregression~\cite{Yule27}.  Since then hundreds, if not thousands,
+of prediction strategies have been developed for a wide variety of
+prediction tasks.  The purpose of this paper is not to add a new
+method to this arsenal, nor to assess or compare the effectiveness of
+these existing methods.  Our goals are broader: (i) to empirically
+quantify the predictive structure that is present in a real-valued
+scalar time series and (ii) to explore how that inherent complexity
+affects the performance of those existing prediction methods.  It
+would, of course, be neither practical nor interesting to report
+results for every existing forecast method; instead, we choose a
+representative set, which are described in Section~\ref{sec:model}.
+
+Quantifying predictability, which is sometimes called ``predicting
+predictability,'' is also not a new problem.  Most solutions fall into
+two categories: model-based error analysis and model-free information
+analysis.
+%
 %Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
+%
+%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
+%
+The first class focuses on errors produced by a fixed forecasting
+schema.  This can proceed locally or globally.  The local version of
+this type of analysis approximates error distributions for different
+regions of a time-series model using local ensemble in-sample
+forecasting.  [[Joshua, we probably want a quick definition of
+    in-sample and out-of-sample.]] These distributions are then used
+to estimate out-of-sample forecast errors [[in that area of]] the time
+series model.  For example, Smith {\sl et al.} make in-sample
+forecasts using ensembles around selected points in the time series in
+order to predict the local predictability of a time series
+reconstructed using delay-coordinate embedding~\cite{Smith199250}.
+This approach can be used to show that different portions of a time
+series can exhibit varying levels of local predictive uncertainty.  We
+expand on this finding later in this paper with a time series that
+exhibits interesting regime shifts.
 
-%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
-
-The first class of methods for quantifying predictive structure is model oriented and focuses on either local or global in- or out-of-sample errors produced by a fixed forecasting schema.
-The local analysis attempts to approximate error distributions for different regions of a time series model by doing local ensemble in-sample forecasting.
-The analysis then uses the approximate in-sample error distribution as a means to estimate out-of-sample forecast errors in that area of the time series model.
-For example, in \cite{Smith199250} in-sample forecasts are done with ensembles around ``data centers" (a subset of the time series) to predict the local predictability of a time series reconstruction using delay-coordinate embedding.
-Using this method, Smtih et al.
-\cite{Smith199250} show  a very interesting corollary: different portions of a time series can exhibit varying levels of local predictive uncertainty.
-We expand on this finding later in this paper with a time series called ``\svd", which exhibits interesting regime changes.
-This local type analysis works quite well but this does not quantify a time series redundancy but only approximates the local predictive uncertainty in relation to a fixed model and tells you nothing of the predictive structure usable by other schema.
+Local model-based error analysis works quite well, but it only
+approximates the \emph{local} predictive uncertainty \emph{in relation
+  to a fixed model}.  It cannot quantify the inherent redundancy of a
+time series~\cite{crutchfield2003} and thus cannot be used to draw
+conclusions about predictive structure that may be usable by other
+forecast methods.
+%
 %Distrubtion of error.
@@
 For example, nonlinear structure which is ignored by a linear predictor.
+%
+Global model-based error analysis moves in this direction.  It uses
+out-of-sample error distributions, computed {\sl post facto} from a
+class of models, to determine which of those models was best.  When
+building an autoregressive model, for example, it is common to verify
+that the model error is normally distributed.  If it is not, that
+indicates that there is linear structure in the time series that the
+model-building process failed to use.  The problem with this approach
+is that if the error is normally distributed it can not be said that
+the practitioner is using the ideal forecasting schema---e.g., if
+nonlinear structure is present that linear modeling is oblivious
+to---but only if this class of model used all the structure it was
+able to process.
 
-The global error analysis is generally done post-facto and attempts to quantify if the right model was chosen from a particular class of models by assessing the out-of-sample error distribution.
-For example, it is common with autoregressive models to verify the error is normally distributed.
-If it is not then this is an indicator that linear structure is present that the current model failed to use.
-The problem with this approach is that if the error is normally distributed it can not be said that the practitioner is using the ideal forecasting schema---e.g., if nonlinear structure is present that linear modeling is oblivious to---but only if this class of model used all the structure it was able to process.
+A practice known as deterministic vs. stochastic
+modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
+between local and global error analysis. The basic idea is to
+construct a series of local linear fits (single point) to global
+linear fits (fitting all points known at that time), increasing the
+number of points in the fit slowly. Then analyze how the average
+out-of-sample error changes as a function of number of points in the
+linear fits. The shape of such a graph indicates how much determinism
+vs. stochasticity is present in a time series, see
+\cite{Casdagli92dvsplots, weigend-book} for more information.
 
-A practice known as deterministic vs.
-stochastic modeling\cite{Casdagli92dvsplots, weigend-book} bridges the gap between local and global error analysis.
-The basic idea is to construct a series of local linear fits (single point) to global linear fits (fitting all points known at that time), increasing the number of points in the fit slowly.
-Then analyze how the average out-of-sample error changes as a function of number of points in the linear fits.
-The shape of such a graph indicates how much determinism vs.
-stochasticity is present in a time series, see  \cite{Casdagli92dvsplots, weigend-book} for more information.
+The model-based error analysis methods described in the previous three
+paragraphs are based on specific assumptions about the underlying
+generating process and knowledge about what will happen to the error
+if those assumptions hold or fail.  Model-free information analysis
+moves away from those restrictions.  Our approach falls into this
+class.  In particular, we empirically measure the inherent complexity
+of a time series using an approximation of the Kolmogorov-Sinai
+entropy, then correlate that complexity with the predictive accuracy
+of forecasts made a number of different schema for that time series.
 
-Each of these methods in the first class are based off very particular assumptions about the underlying generating process and knowledge of how error will result if those assumptions hold or fail.
-Our approach instead falls into the second class which is model-free information analysis.
-In particular, we empirically approximates the time series inherent complexity (Kolmogorov-Sinai entropy) and then correlates this complexity with the predictive accuracy of different forecasting schema, as a means to quantify predictability.
-Quantifying predictability through entropy/redundancy is not a new idea, but past methods (e.g.,\cite{Shannon1951, mantegna1994linguistic}) have hinged on knowledge of the \emph{generating partition} of the generating process, a luxury rarely if ever afforded to an analyst.
+[[I think we need a short paragraph about redundancy in here
+    somewhere.  We keep referring to it as if the reader knows it...]]
 
-It is the case that the ideal method of computing the entropy of a system is via \emph{generating partitions}~\cite{lind95}, a technique which partitions observations into symbols in a way which preserves the underlying dynamics.
-Unfortunately, outside of one-dimensional unimodal maps and a few select higher dimensional maps, generating partitions are highly non-trivial to compute and often have fractal boundaries~\cite{eisele1999}.
-Furthermore, the underlying dynamic needs to be known to construct the generating partition.
-Unfortunately, using any partition (such as those produced by simply binning the data) other than the generation partition induces spurious complexity to the resulting symbolic sequence, thus misrepresenting the entropy of the underlying system~\cite{bollt2001}.
+Using entropy and redundancy to quantify the inherent predictability
+of a time series is not a new idea.  Past methods for this, however,
+(e.g.,~\cite{Shannon1951, mantegna1994linguistic}) have hinged on
+knowledge of the \emph{generating partition} of the generating
+process, which partitions observations into symbols in a way that
+preserves the underlying dynamics~\cite{lind95}.  This is a luxury
+that is rarely, if ever, afforded to an analyst, since one needs to
+know the underlying dynamics in order to construct a generating
+partition.  Even if the dynamics are known, these partitions are
+difficult to compute and often have fractal
+boundaries~\cite{eisele1999}.
+
+Unfortunately, using any partition (such as those produced by simply
+binning the data) other than the generation partition induces spurious
+complexity to the resulting symbolic sequence, thus misrepresenting
+the entropy of the underlying system~\cite{bollt2001}.
 

### 2014-03-24 related.tex
quitting for the night on related.tex

@@
 autoregression~\cite{Yule27}.  Since then hundreds, if not thousands,
-of prediction strategies have been developed for a wide variety of
-prediction tasks.  The purpose of this paper is not to add a new
-method to this arsenal, nor to assess or compare the effectiveness of
-these existing methods.  Our goals are broader: (i) to empirically
-quantify the predictive structure that is present in a real-valued
-scalar time series and (ii) to explore how that inherent complexity
-affects the performance of those existing prediction methods.  It
-would, of course, be neither practical nor interesting to report
-results for every existing forecast method; instead, we choose a
-representative set, which are described in Section~\ref{sec:model}.
+of strategies have been developed for a wide variety of prediction
+tasks.  The purpose of this paper is not to add a new weapon to this
+arsenal, nor to assess or compare the effectiveness of existing
+methods.  Our goals are broader: (i) to empirically quantify the
+predictive structure that is present in a real-valued scalar time
+series and (ii) to explore how that inherent complexity affects the
+performance of those existing prediction methods.  It would, of
+course, be neither practical nor interesting to report results for
+every existing forecast method; instead, we choose a representative
+set, which are described in Section~\ref{sec:model}.
 
@@
 regions of a time-series model using local ensemble in-sample
-forecasting.  [[Joshua, we probably want a quick definition of
-    in-sample and out-of-sample.]] These distributions are then used
-to estimate out-of-sample forecast errors [[in that area of]] the time
-series model.  For example, Smith {\sl et al.} make in-sample
-forecasts using ensembles around selected points in the time series in
-order to predict the local predictability of a time series
-reconstructed using delay-coordinate embedding~\cite{Smith199250}.
-This approach can be used to show that different portions of a time
-series can exhibit varying levels of local predictive uncertainty.  We
-expand on this finding later in this paper with a time series that
-exhibits interesting regime shifts.
+forecasting.  {\color{red}[[Joshua, we probably want a quick
+      definition of in-sample and out-of-sample.]]} These
+distributions are then used to estimate out-of-sample forecast errors
+{\color{red}[[in that area of]] [[means?]]} the time series model.
+For example, Smith {\sl et al.}  make in-sample forecasts using
+ensembles around selected points in the time series in order to
+predict the local predictability of a time series reconstructed using
+delay-coordinate embedding~\cite{Smith199250}.  This approach can be
+used to show that different portions of a time series can exhibit
+varying levels of local predictive uncertainty.  We expand on this
+finding later in this paper with a time series that exhibits
+interesting regime shifts.
 
@@
 out-of-sample error distributions, computed {\sl post facto} from a
-class of models, to determine which of those models was best.  When
-building an autoregressive model, for example, it is common to verify
-that the model error is normally distributed.  If it is not, that
-indicates that there is linear structure in the time series that the
-model-building process failed to use.  The problem with this approach
-is that if the error is normally distributed it can not be said that
-the practitioner is using the ideal forecasting schema---e.g., if
-nonlinear structure is present that linear modeling is oblivious
-to---but only if this class of model used all the structure it was
-able to process.
+class of models, to determine which of those models was best.  After
+building an autoregressive model, for example, it is common to go back
+and verify that the model error is normally distributed.  If it is
+not, that indicates that there is structure in the time series that
+the model-building process was unable to capture and use.
+{\color{red} [[The problem with this approach is that if the error is
+      normally distributed it can not be said that the practitioner is
+      using the ideal forecasting schema---e.g., if nonlinear
+      structure is present that linear modeling is oblivious to---but
+      only if this class of model used all the structure it was able
+      to process.]]  [[having trouble understanding this...]]}
 
@@
 modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
-between local and global error analysis. The basic idea is to
-construct a series of local linear fits (single point) to global
-linear fits (fitting all points known at that time), increasing the
-number of points in the fit slowly. Then analyze how the average
-out-of-sample error changes as a function of number of points in the
-linear fits. The shape of such a graph indicates how much determinism
-vs. stochasticity is present in a time series, see
-\cite{Casdagli92dvsplots, weigend-book} for more information.
+between local and global error analysis {\color{red} [[just in
+      model-based error analysis?  or also in model-free?]]}  The
+basic idea is to construct a series of local linear fits (single
+point) to global linear fits (fitting all points known at that time),
+increasing the number of points in the fit slowly, and then analyze
+how the average out-of-sample error changes as a function of number of
+points in the linear fits. The shape of such a graph indicates how
+much determinism vs. stochasticity is present in a time series.
 
@@
 
-[[I think we need a short paragraph about redundancy in here
-    somewhere.  We keep referring to it as if the reader knows it...]]
+[[I think we need a short paragraph about redundancy in here somewhere
+    --- with a bit more info than was in the intro where we first
+    defined it.  We keep referring to it as if the reader knows
+    it...]]
 

### 2014-03-24 intro.tex
minor wordsmithing on intro.tex

@@
 \end{enumerate}
-The forecast methods used in this study were chosen to span the space of
-standard prediction strategies, but they do not, of course, cover that space
-exhaustively.  Our goal here is an empirical assessment of the
-relationship between predictability and complexity, not formal results
-about a ``best'' predictor for a given time series.  From a
-practitioner's standpoint, it would be useful to know {\sl a priori}
-if a trace contains enough predictive structure to make it worth
-spending the time and effort searching for a good forecast method.  It
-would also be useful to know if a given method is inadequate---that
-is, if one could do better and should keep searching.
+The forecast methods used in this study were chosen to be a
+representative sampling of standard prediction strategies, but they do
+not, of course, cover that space exhaustively.  Our goal here is an
+empirical assessment of the relationship between predictability and
+complexity, not formal results about a ``best'' predictor for a given
+time series.  From a practitioner's standpoint, it would be useful to
+know {\sl a priori} if a trace contains enough predictive structure to
+make it worth spending the time and effort searching for a good
+forecast method.  It would also be useful to know if a given method is
+inadequate---that is, if one could do better and should keep
+searching.
 
@@
 For the purposes of this study, we require a broad array of
-time-series datasets that cover the full complexity spectrum.  Rather
-than use synthetic data, we study sensor data from computer
-performance experiments.  While this is not a common laboratory
-experiment, it is a highly appropriate choice here.  Computers are
-among the most complex engineered artifacts in current use.  Modern
-microprocessor chips contain multiple processing units and multi-layer
-memories, for instance, and they use complicated hardware/software
-strategies to move data and threads of computation across those
-resources.  As a result, the processor and memory loads during the
-execution of even a very simple program can evolve in a very
-complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance trace
-of a three-line loop that repeatedly initializes the upper triangle of
-a matrix in column-major order.
+time-series datasets from across the complexity spectrum.  Rather than
+use synthetic data, we study sensor data from computer performance
+experiments.  While this is not a common laboratory experiment, it is
+a highly appropriate choice here.  Computers are among the most
+complex engineered artifacts in current use.  Modern microprocessor
+chips contain multiple processing units and multi-layer memories, for
+instance, and they use complicated hardware/software strategies to
+move data and threads of computation across those resources.  As a
+result, the processor and memory loads during the execution of even a
+very simple program can evolve in a very complicated fashion.
+Figure~\ref{fig:col-ipc} shows a performance trace of a three-line
+loop that repeatedly initializes the upper triangle of a matrix in
+column-major order.
 %
@@
 chaotic~\cite{mytkowicz09}, which places fundamental limits on
-predictability.  Changes in the program can radically change this
+predictability.  Changes in the code can radically change this
 behavior; indeed, many performance traces exhibit interesting regime
@@
   would amount to solving the halting problem.  What we are doing here
-  is predicting computer \emph{performance}, which does not violate the
-  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
+  is predicting computer \emph{performance}, which does not violate
+  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
 % footnote to appease the theoretician in Josh
@@
 series. Section~\ref{sec:model} describes the prediction models used
-in this study.  In Section~\ref{sec:meaComplex}, we review the
-permutation entropy technique.  In Section~\ref{sec:results}, we estimate the
-complexity of each time series and compare that complexity to the
-accuracy of predictions produced by the methods of
-Section~\ref{sec:model}, operating on that time series.  In
-Section~\ref{sec:conc}, we discuss these results and their
-implications and consider future areas of research.
+in this study.  In Section~\ref{sec:meaComplex}, we review permutation
+entropy, the technique that we use to measure complexity.  In
+Section~\ref{sec:results}, we estimate the complexity of each time
+series and compare that complexity to the accuracy of predictions
+produced by the methods of Section~\ref{sec:model}, operating on that
+time series.  In Section~\ref{sec:conc}, we discuss these results and
+their implications, and consider future areas of research.

### 2014-03-24 related.tex
new draft of related.tex

@@
 
-Modeling time-series data for the purposes of prediction is an old
-problem that dates back at least to Yule's 1927 invention of
-autoregression~\cite{Yule27}.  Since then hundreds, if not thousands,
-of strategies have been developed for a wide variety of prediction
-tasks.  The purpose of this paper is not to add a new weapon to this
-arsenal, nor to assess or compare the effectiveness of existing
-methods.  Our goals are broader: (i) to empirically quantify the
-predictive structure that is present in a real-valued scalar time
-series and (ii) to explore how that inherent complexity affects the
-performance of those existing prediction methods.  It would, of
-course, be neither practical nor interesting to report results for
-every existing forecast method; instead, we choose a representative
-set, which are described in Section~\ref{sec:model}.
+Modeling time-series data for the purposes of prediction dates back at
+least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
+then hundreds, if not thousands, of strategies have been developed for
+a wide variety of prediction tasks.  The purpose of this paper is not
+to add a new weapon to this arsenal, nor to assess or compare the
+effectiveness of existing methods.  Our goals are more general: {\sl
+  (i)} to empirically quantify the predictive structure that is
+present in a real-valued scalar time series and {\sl (ii)} to explore
+how the performance of prediction methods is related to that inherent
+complexity.  It would, of course, be neither practical nor interesting
+to report results for every existing forecast method; instead, we
+choose a representative set, as described in Section~\ref{sec:model}.
 
 Quantifying predictability, which is sometimes called ``predicting
-predictability,'' is also not a new problem.  Most solutions fall into
-two categories: model-based error analysis and model-free information
-analysis.
+predictability,'' is not a new problem.  Most solutions to it fall
+into two categories: model-based error analysis and model-free
+information analysis.
 %
@@
 The first class focuses on errors produced by a fixed forecasting
-schema.  This can proceed locally or globally.  The local version of
-this type of analysis approximates error distributions for different
-regions of a time-series model using local ensemble in-sample
-forecasting.  {\color{red}[[Joshua, we probably want a quick
-      definition of in-sample and out-of-sample.]]} These
-distributions are then used to estimate out-of-sample forecast errors
-{\color{red}[[in that area of]] [[means?]]} the time series model.
-For example, Smith {\sl et al.}  make in-sample forecasts using
-ensembles around selected points in the time series in order to
-predict the local predictability of a time series reconstructed using
-delay-coordinate embedding~\cite{Smith199250}.  This approach can be
-used to show that different portions of a time series can exhibit
-varying levels of local predictive uncertainty.  We expand on this
-finding later in this paper with a time series that exhibits
-interesting regime shifts.
+schema.  This analysis can proceed locally or globally.  The local
+version approximates error distributions for different regions of a
+time-series model using local ensemble in-sample forecasting.
+{\color{red}[[Joshua, we probably want a quick definition of in-sample
+      and out-of-sample.]]} These distributions are then used to
+estimate out-of-sample forecast errors {\color{red}[[in that area of]]
+  [[means?]]} the time series model.  For example, Smith {\sl et al.}
+make in-sample forecasts using ensembles around selected points in
+order to predict the local predictability of that time
+series~\cite{Smith199250}.  This approach can be used to show that
+different portions of a time series can exhibit varying levels of
+local predictive uncertainty.  We expand on this finding later in this
+paper with a time series that exhibits interesting regime shifts.
 
@@
 approximates the \emph{local} predictive uncertainty \emph{in relation
-  to a fixed model}.  It cannot quantify the inherent redundancy of a
-time series~\cite{crutchfield2003} and thus cannot be used to draw
-conclusions about predictive structure that may be usable by other
-forecast methods.
+  to a fixed model}.  It cannot quantify the inherent predictability
+of a time series and thus cannot be used to draw conclusions about
+predictive structure that may be usable by other forecast methods.
 %
@@
 Global model-based error analysis moves in this direction.  It uses
-out-of-sample error distributions, computed {\sl post facto} from a
+out-of-sample error distributions, computed \emph{post facto} from a
 class of models, to determine which of those models was best.  After
-building an autoregressive model, for example, it is common to go back
-and verify that the model error is normally distributed.  If it is
-not, that indicates that there is structure in the time series that
-the model-building process was unable to capture and use.
-{\color{red} [[The problem with this approach is that if the error is
-      normally distributed it can not be said that the practitioner is
-      using the ideal forecasting schema---e.g., if nonlinear
-      structure is present that linear modeling is oblivious to---but
-      only if this class of model used all the structure it was able
-      to process.]]  [[having trouble understanding this...]]}
+building an autoregressive model, for example, it is common to
+calculate the errors and verify that they are normally distributed.
+If they are not, that indicates that there is structure in the time
+series that the model-building process was unable to capture and use.
+The problem with this approach is lack of generality.  Normally
+distributed errors indicate that a model has captured the structure in
+the data insofar as is possible, \emph{given the formulation of that
+  particular model} (viz., the best possible linear fit to a nonlinear
+dataset).  This gives no indication as to whether another modeling
+strategy might do better.  {\color{red} [[I'm not sure I believe that
+      a model's bad fit wouldn't manifest in the error distribution.
+      Is that really true?]]}
 
-A practice known as deterministic vs. stochastic
+A practice known as deterministic/sto\-chas\-tic
 modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
-between local and global error analysis {\color{red} [[just in
-      model-based error analysis?  or also in model-free?]]}  The
-basic idea is to construct a series of local linear fits (single
-point) to global linear fits (fitting all points known at that time),
-increasing the number of points in the fit slowly, and then analyze
+between local and global model-based error analysis {\color{red} [[or
+      also in model-free?]]}  The basic idea is to construct a series
+of local linear fits, beginning with a few points and working up to to
+a global linear fit that includes all known points, and then analyze
 how the average out-of-sample error changes as a function of number of
-points in the linear fits. The shape of such a graph indicates how
-much determinism vs. stochasticity is present in a time series.
+points in the fit. The shape of such a graph indicates the amounts of
+determinism and stochasticity are present in a time series.
 
@@
 moves away from those restrictions.  Our approach falls into this
-class.  In particular, we empirically measure the inherent complexity
-of a time series using an approximation of the Kolmogorov-Sinai
-entropy, then correlate that complexity with the predictive accuracy
-of forecasts made a number of different schema for that time series.
+class.  Our goal is to empirically measure the inherent complexity of
+a time series, then correlate that complexity with the predictive
+accuracy of forecasts made using a number of different methods.
 
-[[I think we need a short paragraph about redundancy in here somewhere
-    --- with a bit more info than was in the intro where we first
-    defined it.  We keep referring to it as if the reader knows
-    it...]]
+We build on the notion of \emph{redundancy} that was introduced on
+page~\pageref{page:redundancy}, which formally quantifies how
+information propagates forward through a time series:
+% how much information prior observations of a system lend to future
+% values
+e.g., the mutual information between the past $n$ observations and the
+current one.  {\color{red} [[How is redundancy different from/more
+      general than MI, then?]]}  For example, this would be the amount
+of information that delay coordinate embedding captures.  {\color{red}
+  [[This will be very hard to understand, coming as it does before the
+      description of that method.  Does it add something essential
+      here?]]}  The redundancy of i.i.d. random processes is zero,
+since all observations are independent of one another.  On the other
+hand, any deterministic system --- including chaotic ones --- can be
+perfectly predicted if observed for long enough~\cite{weigend-book}.
+{\color{red} [[Previous sentence leaves the connection to redundancy
+      implicit.  Need to make it explicit.]]}  
+In practice, it is quite difficult to estimate the redundancy of an
+arbitrary, real-valued time series.  Doing so requires knowing either
+the Kolmolgorov-Sinai entropy or the values of all positive Lyapunov
+exponents of the system.  Both of these calculations are
+difficult---the latter particularly so if the data are very noisy or
+the generating system is stochastic.
 
@@
 (e.g.,~\cite{Shannon1951, mantegna1994linguistic}) have hinged on
-knowledge of the \emph{generating partition} of the generating
-process, which partitions observations into symbols in a way that
-preserves the underlying dynamics~\cite{lind95}.  This is a luxury
-that is rarely, if ever, afforded to an analyst, since one needs to
-know the underlying dynamics in order to construct a generating
-partition.  Even if the dynamics are known, these partitions are
-difficult to compute and often have fractal
+knowledge of the \emph{generating partition} of the underlying
+process, which lets one transform {\color{red}[[what kind of
+      observations?  real-valued? scalar?]]}  observations into
+symbols in a way that preserves the underlying dynamics~\cite{lind95}.
+Different projections---e.g., simply binning the data---can create
+spurious complexity in the resulting symbolic sequence and thus
+misrepresent the entropy of the underlying system~\cite{bollt2001}.
+Generating partitions are luxuries that are rarely, if ever, afforded
+to an analyst, since one needs to know the underlying dynamics in
+order to construct one.  And even if the dynamics are known, these
+partitions are difficult to compute and often have fractal
 boundaries~\cite{eisele1999}.
 
-Unfortunately, using any partition (such as those produced by simply
-binning the data) other than the generation partition induces spurious
-complexity to the resulting symbolic sequence, thus misrepresenting
-the entropy of the underlying system~\cite{bollt2001}.
-
-It seems desirable to quantify how much information prior observations of a system lend to future values, the \emph{mututal information} between, say, the past $n$ observations at the current one.
-This quantity is known as the \emph{redundancy}.
-For example, this would be the amount of information that delay coordinate embedding captures.
-For IID random processes,  the redundancy is zero, since all observations are independent of each other.
-On the other hand, any deterministic system --- including chaotic ones --- can be perfectly predicted if observed for long enough~\cite{weigend-book}.
-Unfortunately it is quite difficult to estimate the redundancy in practice for an arbitrary, real-valued time series.
-Doing so requires knowing either the Kolmolgorov-Sinai entropy or the \emph{full} positive Lyapunov spectrum.
-As discussed above, computing the entropy is quite difficult.
-Similarly, computing the positive Lyapunov exponent(s) from an arbitrary time series is also fraught with difficulties, and is effectively impossible if there is significant noise or the generating system is stochastic.
-
-Our approach expands on \cite{bandt2002per} to approximate normalized redundancy of a time series as a proxy for predictability.
-This method is different from the current literature because it does not rely on generating partitions, and as such does not suffer from the previously mentioned biases.
-In addition no knowledge of the underlying generating process is necessary, e.g, linear, nonlinear, the Lyapunov spectrum, etc.
-In fact, our approach uses no assumption about the underlying generating process, does not require generating partitions, and is applicable to noisy real-valued time series from all classes of systems.
-
-{\color{red}
-Maybe talk about ``
-wpe has been applied to predicting irregularities in brain wave data but no on one has examined it's correlation with predictive structure. "
-but kind of puts the cart before the horse
-
-[[ryan]] i tried writing a paragraph about this around 7 times, and couldn't get anything that sounded reasonable.
-}
+We sidestep these issues by using a variant of the \emph{permutation
+  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate a value
+for the Kolmogorov-Sinai entropy of a real-valued time series---and
+thus the redundancy in that data, which is an effective proxy for
+predictability.  This differs from existing approaches in a number of
+ways.  It does not rely on generating partitions---and thus does not
+introduce bias into the results if one does not know the dynamics or
+cannot compute the partition.  It makes no assumptions about, and
+requires no knowledge of, the underlying generating process: linear,
+nonlinear, the Lyapunov spectrum, etc.  These features make our
+approach applicable to noisy real-valued time series from all classes
+of systems.

### 2014-03-24 model.tex
reduced red font in related

@@
 does not.  {\color{red}[[We should comment on what a noise ball means
-      on such a plot, as well as what a horizontal line means.]]}
+      on such a plot, as well as what a horizontal line means. maybe
+      also tie back to the comment on
+      page~\pageref{page:normal-errors}.]]}
 

### 2014-03-24 related.tex
reduced red font in related

@@
 version approximates error distributions for different regions of a
-time-series model using local ensemble in-sample forecasting.
-{\color{red}[[Joshua, we probably want a quick definition of in-sample
-      and out-of-sample.]]} These distributions are then used to
-estimate out-of-sample forecast errors {\color{red}[[in that area of]]
-  [[means?]]} the time series model.  For example, Smith {\sl et al.}
-make in-sample forecasts using ensembles around selected points in
-order to predict the local predictability of that time
+time-series model using local ensemble in-sample
+forecasting\footnote{[[The terms ``in sample'' and ``out of sample''
+      are used in different ways in the forecasting community.  By in
+      sample, we mean a ``forecast'' that is done on the stuff that
+      you know, for the purposes of assessing models, errors, and
+      prediction horizons.  An out-of-sample prediction is what one
+      thinks of as a prediction: forecasting the future.]]}.  These
+distributions are then used as estimates of out-of-sample forecast
+errors in those regions.  For example, Smith {\sl et al.}  make
+in-sample forecasts using ensembles around selected points in order to
+predict the local predictability of that time
 series~\cite{Smith199250}.  This approach can be used to show that
@@
 series that the model-building process was unable to capture and use.
-The problem with this approach is lack of generality.  Normally
-distributed errors indicate that a model has captured the structure in
-the data insofar as is possible, \emph{given the formulation of that
-  particular model} (viz., the best possible linear fit to a nonlinear
-dataset).  This gives no indication as to whether another modeling
-strategy might do better.  {\color{red} [[I'm not sure I believe that
-      a model's bad fit wouldn't manifest in the error distribution.
-      Is that really true?]]}
+The problem with this approach is lack of generality.  
+\label{page:normal-errors}
+Normally distributed errors indicate that a model has captured the
+structure in the data insofar as is possible, \emph{given the
+  formulation of that particular model} (viz., the best possible
+linear fit to a nonlinear dataset).  This gives no indication as to
+whether another modeling strategy might do better.
 
@@
 modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
-between local and global model-based error analysis {\color{red} [[or
-      also in model-free?]]}  The basic idea is to construct a series
-of local linear fits, beginning with a few points and working up to to
-a global linear fit that includes all known points, and then analyze
-how the average out-of-sample error changes as a function of number of
-points in the fit. The shape of such a graph indicates the amounts of
-determinism and stochasticity are present in a time series.
+between local and global approaches to model-based error analysis.
+The basic idea is to construct a series of local linear fits,
+beginning with a few points and working up to to a global linear fit
+that includes all known points, and then analyze how the average
+out-of-sample error changes as a function of number of points in the
+fit. The shape of such a graph indicates the amounts of determinism
+and stochasticity are present in a time series.
 
@@
 % values
-e.g., the mutual information between the past $n$ observations and the
-current one.  {\color{red} [[How is redundancy different from/more
-      general than MI, then?]]}  For example, this would be the amount
-of information that delay coordinate embedding captures.  {\color{red}
-  [[This will be very hard to understand, coming as it does before the
-      description of that method.  Does it add something essential
-      here?]]}  The redundancy of i.i.d. random processes is zero,
-since all observations are independent of one another.  On the other
-hand, any deterministic system --- including chaotic ones --- can be
-perfectly predicted if observed for long enough~\cite{weigend-book}.
-{\color{red} [[Previous sentence leaves the connection to redundancy
-      implicit.  Need to make it explicit.]]}  
-In practice, it is quite difficult to estimate the redundancy of an
-arbitrary, real-valued time series.  Doing so requires knowing either
-the Kolmolgorov-Sinai entropy or the values of all positive Lyapunov
-exponents of the system.  Both of these calculations are
-difficult---the latter particularly so if the data are very noisy or
-the generating system is stochastic.
+i.e., the mutual information between the past $n$ observations and the
+current one.
+% took this out because DCE hasn't been introduced yet: 
+% 
+% For example,
+% this would be the amount of information that delay coordinate
+% embedding captures.
+The redundancy of i.i.d. random processes is zero, since all
+observations are independent of one another.  On the other hand, any
+deterministic system---including chaotic ones---has maximal redundancy
+and thus can be perfectly predicted if observed for long
+enough~\cite{weigend-book}.  In practice, it is quite difficult to
+estimate the redundancy of an arbitrary, real-valued time series.
+Doing so requires knowing either the Kolmolgorov-Sinai entropy or the
+values of all positive Lyapunov exponents of the system.  Both of
+these calculations are difficult---the latter particularly so if the
+data are very noisy or the generating system is stochastic.
 

### 2014-03-24 model.tex
various small changes

--- 2014-03-24 model.tex
@@
 generates a very accurate prediction of the \col data, while ARIMA
-does not.  {\color{red}[[We should comment on what a noise ball means
-      on such a plot, as well as what a horizontal line means.]]}
+does not.  Horizontal lines result when a constant predictor (e.g.,
+\naive) is used on a non-constant signal.  Point clouds reflect the
+structure of the distribution of the errors---roughly normally
+distributed, for example, in the case of LMA on \gcc.
 
@@
 % 
-$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-x_j|
+$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
 }{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
 % 
-{\color{red} That equation had $c_j$ instead of $p_j$ in the numerator.
-I changed it.  Let me know if that's wrong.}
 This error metric was introduced in \cite{MASE} as a ``generally
@@
 random-walk predictor, while ARIMA only outperformed random walk by a
-factor of 1.7 {\color{red}[[is this right?  there was a ``2'' here,
-      but 1/0.599 is closer to 1.7]]}.  This is in accordance with the
+factor of 1.7.  This is in accordance with the
 visual appearance of the corresponding images in

### 2014-03-24 pecomplexity.tex
various small changes

@@
 %\\ 
-{\color{green}Quantifying Predictability through Structural Complexity} 
+Quantifying Predictability through Structural Complexity
 %\\ 
 %Quantifying Predictability using Structural Complexity 
-\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
-\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
+% \\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
+% \\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
 }
@@
 
-\section{Appendix}
-
-\begin{table}[h]
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}
-  \hline
-            & MASE LMA    & MASE ARIMA &MASE na\"{i}ve   & $l=6$ \\
- \hline
-
- \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\
-
-\gcc           & $ 1.530\pm 0.021$ & $1.837 \pm0.016 $ & $0.951 \pm 0.001$ & $0.943 \pm 0.001$ \\
-
-\svdone     & $ 0.827\pm 0.076$ & $ 0.714\pm 0.075 $ & $2.676\pm4.328$&  $0.957 \pm 0.016$ \\
-
- \svdtwo    & $1.279 \pm0.020 $ & $2.163 \pm0.027 $ &  $3.054\pm0.040$ &   $0.846 \pm0.004$ \\
-
- \svdthree     & $0.619 \pm0.021 $ & $0.713 \pm 0.010 $ & $31.386\pm 0.282$ &  $0.716 \pm 0.006$ \\
-
- \svdfour     & $ 0.779\pm0.036 $ & $0.979 \pm0.032 $ & $2.661\pm0.074$ & $0.825 \pm 0.008$ \\
-
- \svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
-
- \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
-  \hline
-  \end{tabular}
-  \end{center}
- \label{default}
- \caption{Detailed MASE scores}
- \label{tab:error}
-  \end{table}%
-
 \bibliographystyle{aipauth4-1}

### 2014-03-24 related.tex
various small changes

@@
 knowledge of the \emph{generating partition} of the underlying
-process, which lets one transform {\color{red}[[what kind of
-      observations?  real-valued? scalar?]]}  observations into
+process, which lets one transform real-valued observations into
 symbols in a way that preserves the underlying dynamics~\cite{lind95}.

### 2014-03-26 related.tex
wordsmithed in/out sample footnote

@@
 time-series model using local ensemble in-sample
-forecasting\footnote{[[The terms ``in sample'' and ``out of sample''
-      are used in different ways in the forecasting community.  By in
-      sample, we mean a ``forecast'' that is done on the stuff that
-      you know, for the purposes of assessing models, errors, and
-      prediction horizons.  An out-of-sample prediction is what one
-      thinks of as a prediction: forecasting the future.]]}.  These
-distributions are then used as estimates of out-of-sample forecast
-errors in those regions.  For example, Smith {\sl et al.}  make
-in-sample forecasts using ensembles around selected points in order to
-predict the local predictability of that time
+forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
+  used in different ways in the forecasting community.  The meanings
+  that take here are distinguished by the part of the time series that
+  is the focus of the prediction: the observed data for the former and
+  the unknown future for the latter.  In-sample
+  forecasts---comparisons of predictions generated from \emph{part} of
+  the observed time series---are useful for assessing model error and
+  prediction horizons, among other things.}.
+%
+These distributions are then used as estimates of out-of-sample
+forecast errors in those regions.  For example, Smith {\sl et al.}
+make in-sample forecasts using ensembles around selected points in
+order to predict the local predictability of that time
 series~\cite{Smith199250}.  This approach can be used to show that

### 2014-03-26 related.tex
typo in related

@@
 forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
-  used in different ways in the forecasting community.  The meanings
-  that take here are distinguished by the part of the time series that
-  is the focus of the prediction: the observed data for the former and
-  the unknown future for the latter.  In-sample
-  forecasts---comparisons of predictions generated from \emph{part} of
-  the observed time series---are useful for assessing model error and
-  prediction horizons, among other things.}.
+  used in different ways in the forecasting community.  Here, we
+  distinguish those terms by the part of the time series that is the
+  focus of the prediction: the observed data for the former and the
+  unknown future for the latter.  In-sample forecasts---comparisons of
+  predictions generated from \emph{part} of the observed time
+  series---are useful for assessing model error and prediction
+  horizons, among other things.}.
 %

### 2014-03-26 related.tex
typo in related

@@
 % embedding captures.
-The redundancy of i.i.d. random processes is zero, since all
-observations are independent of one another.  On the other hand, any
-deterministic system---including chaotic ones---has maximal redundancy
-and thus can be perfectly predicted if observed for long
-enough~\cite{weigend-book}.  In practice, it is quite difficult to
-estimate the redundancy of an arbitrary, real-valued time series.
-Doing so requires knowing either the Kolmolgorov-Sinai entropy or the
-values of all positive Lyapunov exponents of the system.  Both of
-these calculations are difficult---the latter particularly so if the
-data are very noisy or the generating system is stochastic.
+The redundancy of i.i.d. random processes, for instance, is zero,
+since all observations in such a process are independent of one
+another.  On the other hand, deterministic systems---including chaotic
+ones---have maximal redundancy and thus can be perfectly predicted if
+observed for long enough~\cite{weigend-book}.  In practice, it is
+quite difficult to estimate the redundancy of an arbitrary,
+real-valued time series.  Doing so requires knowing either the
+Kolmolgorov-Sinai entropy or the values of all positive Lyapunov
+exponents of the system.  Both of these calculations are
+difficult---the latter particularly so if the data are very noisy or
+the generating system is stochastic.
 
@@
 symbols in a way that preserves the underlying dynamics~\cite{lind95}.
-Different projections---e.g., simply binning the data---can create
-spurious complexity in the resulting symbolic sequence and thus
-misrepresent the entropy of the underlying system~\cite{bollt2001}.
-Generating partitions are luxuries that are rarely, if ever, afforded
-to an analyst, since one needs to know the underlying dynamics in
-order to construct one.  And even if the dynamics are known, these
-partitions are difficult to compute and often have fractal
-boundaries~\cite{eisele1999}.
+Using a projection that is not a generating partition---e.g., simply
+binning the data---can introduce spurious complexity into the
+resulting symbolic sequence and thus misrepresent the entropy of the
+underlying system~\cite{bollt2001}.  Generating partitions are
+luxuries that are rarely, if ever, afforded to an analyst, since one
+needs to know the underlying dynamics in order to construct one.  And
+even if the dynamics are known, these partitions are difficult to
+compute and often have fractal boundaries~\cite{eisele1999}.
 

### 2014-03-26 methods.tex
minor edits to methods.tex incl regime -> segment
moved seizure detection citation to concl

@@
       during the execution of \svd. Each point is the average IPC
-      during a 100,000 instruction period.  The colors refer to different
-``regimes'' of the signal, as discussed in the text.}
+      during a 100,000 instruction period.  The colors refer to
+      different {\sl segments} of the signal, as discussed in the
+      text.}
     \label{fig:svd-ts-colored}
@@
 of course, particularly since the regimes exhibit some fractal
-structure\footnote{It should be noted that a version of the complexity
-  metric discussed in Section~\ref{sec:meaComplex} has been used to
-  detect seizures in brainwave data \cite{cao2004det}, which can be
-  viewed as regime detection.  The purpose of this paper is \emph{not}
-  to rigorously explore regime detection but to explore quantifying
-  complexity of a time series.}.  Thus, it may the case that more than
-one generating process may be at work in each of our segments.  This
-is a factor in the discussion of Section~\ref{sec:results}.
+structure.  Thus, it may the case that more than one generating
+process may be at work in each of our segments.  This is a factor in
+the discussion of Section~\ref{sec:results}.
 

### 2014-03-26 concl.tex
editing
putting it in Phys Rev E Format
various small changes
see previous commit

@@
-\section{ Conclusions \& Future Work}\label{sec:conc}
-
+\section{ Conclusions \& Future Work }\label{sec:conc}
+{\color{red} NOT EDITABLE}
 \begin{enumerate}
@@
 \end{enumerate}
+
+Paragraph about PE as regime-shift detection.  Here's some text that
+we had in a footnote in the methods section, which no longer talks
+about regimes or regime-shift detection: It should be noted that
+[[non-weighted?  non-windowed?]]  permutation entropy has been used to
+detect seizures in brainwave data \cite{cao2004det}, which can be
+viewed as regime detection.  The purpose of this paper is \emph{not}
+to rigorously explore regime detection but to explore quantifying
+complexity of a time series.
 

### 2014-03-26 model.tex
cleaned up model.tex

@@
 to capture and exploit the deterministic structure of a signal from a
-nonlinear dynamical system.  The \naive and random-walk methods,
+nonlinear dynamical system.  The \naive ~and random-walk methods,
 somewhat surprisingly, often outperform these more-sophisticated
@@
 A random-walk predictor simply uses the last observed measurement as
-the forecast: that is, the value $p_i$ at time $i$ is calculated
-using the relation $$p_i = x_{i-1}$$ The prediction strategy that we
-refer to using the term ``\naive'' averages the prior observations to
-generate the forecast: $$p_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
-While both of these methods are simplistic, they are not without
-merit.  For a time series near the high end of the complexity
-spectrum---i.e., one that possesses very little predictive
-structure---these two methods can actually be the best choice.  In
-forecasting currency exchange rates, for instance, sophisticated
-econometrics-based prediction models fail to consistently outperform
-the random-walk method~\cite{rwMeese,rwCCE}.  These signals are
-volatile, noisy, and possess very little predictive structure, but
-they do not---on the average---vary a great deal, so the random-walk
-method's strategy of simply guessing the last known value is not a bad
-choice.  If a signal has a unimodal distribution with low variance,
-the \naive ~prediction strategy will perform quite well---even if the
-signal is highly complex---simply because the mean is a good
-approximation of the future behavior.  Moreover, the \naive
+the forecast: that is, the predicted value $p_i$ at time $i$ is
+calculated using the relation $$p_i = x_{i-1}$$ The prediction
+strategy that we refer to using the term ``\naive'' averages the prior
+observations to generate the forecast: $$p_i =
+\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
+simplistic, they are not without merit.  For a time series near the
+high end of the complexity spectrum---i.e., one that possesses very
+little predictive structure---these two methods can actually be the
+best choice.  In forecasting currency exchange rates, for instance,
+sophisticated econometrics-based prediction models fail to
+consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
+These signals are volatile, noisy, and possess very little predictive
+structure, but they do not---on the average---vary a great deal, so
+the random-walk method's strategy of simply guessing the last known
+value is not a bad choice.  If a signal has a unimodal distribution
+with low variance, the \naive ~prediction strategy will perform quite
+well---even if the signal is highly complex---simply because the mean
+is a good approximation of the future behavior.  Moreover, the \naive
 ~prediction strategy's temporal average effects a low-pass filtering
@@
 
-Since LMA does not rest on an assumption of linearity (as ARIMA does)
+Since LMA does not rest on an assumption of linearity (as ARIMA does),
 it can handle both linear and nonlinear processes.  If the underlying
@@
 %
-{\color{red} Joshua, what would you think of adding a row of
-  random-walk plots at the top?  Also, the letters in the
-  subcaptions---(a) (b) etc.---feel clunky here.  What would you think
-  of removing them?}
+{\color{red} Joshua will add a row of random-walk plots at the top}
 %

### 2014-03-26 pecomplexity.tex
added time-series to the title
adding RW Figures

@@
 %\\ 
-Quantifying Predictability through Structural Complexity
+{\color{green}Quantifying Predictability through Structural Complexity} 
 %\\ 
 %Quantifying Predictability using Structural Complexity 
-% \\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
-% \\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
+\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
+\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
 }

### 2014-03-27 wpe.tex
new draft of related.tex
various small changes
make WPE section more readable and consistant

@@
 
-For the purposes of this paper, one can view entropy as a measure of complexity and unpredictability in a time series.
-A high-entropy time series is almost completely unpredictable---and conversely a low entropy signal is often quite predictable.
-This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Shannon entropy rate is equal to the sum of the positive Lyapunov exponents, $\lambda_i$.
+For the purposes of this paper one can view the Shannon entropy, and in particular its growth rate  with respect to word length (the \emph{Shannon entropy rate}), as a measure of complexity and unpredictability in a time series.
+A high-entropy rate time series is almost completely unpredictable---and conversely a low entropy rate signal is often quite predictable.
+This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Kolmogorov-Sanai (KS) entropy is equal to the sum of the positive Lyapunov exponents, $\lambda_i$.
 The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.
 The faster the divergence, the larger the entropy and the more difficult prediction becomes.
+Further, the KS entropy is defined as the supremum of the Shannon entropy rates of all partitions.
+The partition which achieves this supremum is known as a \emph{generating partition}.
 
-Further we can consider the information contained in a single observable, say the \emph{present}, of the system, measured by the entropy.
-This information can be partitioned in to two components: the information shared with past observations --- the mutual information between the past and present --- and the information in the present which is not contained in the past --- the conditional entropy.
-The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the aforementioned \emph{entropy rate}.
+From a different point of view, we can consider the information (as measured by the Shannon entropy) contained in a single observable, say the \emph{present}, of the system.
+This information can be partitioned in to two components: the information shared with past observations, the mutual information between the past and present, and the information in the present which is not contained in the past, the conditional entropy of the present given the past.
+The first part is known as the \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the aforementioned \emph{Shannon entropy rate}.
 In theory, the more redundancy in a signal the more predictable it is.
@@
 
-Utilizing entropy as a measure of temporal complexity is by no means a new idea \cite{Shannon1951, mantegna1994linguistic}.
-Its effective usage requires categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite \emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is effectively\footnote{Measurements from real devices are of finite precision, rendering them discrete.
-Unfortunately the space of values taken is so large that for the purposes of entropy calculations it can be taken as a continuum.} real-valued.
+Utilizing Shannon entropy rate as a measure of temporal complexity is not a new idea \cite{Shannon1951, mantegna1994linguistic}, but these approaches required categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite \emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is effectively\footnote{Measurements from real devices are always discrete though typically of very high resolution, for example \texttt{double} precision.
+Though these values are discete, for the purposes of entropy calculations they are effectively a continuum.} real-valued.
 To get around this, one must discretize the data --- typically achieved by binning.
 Unfortunately, this is rarely a good solution to the problem, as the binning of the values introduces an additional dynamic on top of the intrinsic dynamics whose entropy is desired\cite{bollt2001}.
-The field of symbolic dynamics studies how to discretize a time series in such a way that the intrinsic behavior is not perverted, but these methods are fragile in the face of noise and require further understanding of the underlying system, which defeats the purpose of measuring the entropy in the first place.
+The field of symbolic dynamics studies how to discretize a time series in such a way that the intrinsic behavior is not perverted, but these methods are fragile in the face of noise and furthermore require understanding of the underlying system, which defeats the purpose of measuring the entropy to quantify predictability in the first place.
 
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural complexity measure for time series" \cite{bandt2002per}.
+Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural complexity measure for time series'' \cite{bandt2002per}.
 Permutation entropy employs a method of discretizing real-valued time series that follows the intrinsic behavior of the system under examination.
@@
   \begin{align*}
-    p(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
+    P(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
   \end{align*}
-  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $\ell \ge 2$ is defined as
+  Where $P(\pi)$ quantifies the probability of an ordinal and $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $\ell \ge 2$ is defined as
   \begin{align*}
-    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p(\pi) \log_2 p(\pi)
+    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P(\pi) \log_2 P(\pi)
   \end{align*}
@@
 With this in mind, it is common in the literature to normalize permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.
-With this convention, ``low'' entropy is close to 0 and ``high'' entropy is close to 1.
-Finally, it should be noted that the permutation entropy has been shown to be identical to the Kolmolgorov-Sinai (KS) entropy for many large classes of systems \cite{amigo2012permutation}.
-In turn, and under very loose conditions, the KS entropy is equal to the Shannon entropy rate.
+With this convention, ``low'' PE is close to 0 and ``high'' PE is close to 1.
+Finally, it should be noted that the permutation entropy has been shown to be identical to the Kolmolgorov-Sinai entropy for many large classes of systems \cite{amigo2012permutation}, so long as observational noise is sufficiently small.
+As mentioned before, this is equal to the Shannon entropy rate of a generating partition of the system.
 What this normalization allows us to do, then, is approximate the redundancy of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
@@
 \begin{align*}
-  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
+  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
 \end{align*}
@@
 \begin{align*}
-  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p_w(\pi) \log_2 p_w(\pi),
+  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
 \end{align*}
@@
 %
+% {\color{red}TODO: Put this image back in.}
 %%Liz commented this image out temporarily so that her mac doesn't hang
 %%when she scrolls past p12
+%% We decided to skip the figure completely, since we took out
+%% all stuff about PE as regime-shift detection.  
+%% [[should add that as future work, though.]]
 %  \includegraphics[width=\columnwidth]{figs/SVD_wwpe}
-%    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
-%  \caption{The weighted permutation entropy of one run of
-%  SVD. The gray bands are regions where the window overlaps
-%  regimes. The window size used is 5,000 $\times$ 100,000 instructions
-%  and the word length is $6$. For reference the instructions per cycle
-%  of \svd are plotted as a ghost behind this plot. Each color on the
-%  ghosted time series corresonds to the different regimes as selected
-%  by rapid shifts in WPE. From left to right each change in color
-%  represents a change in regime for 6 regimes in
-%  total.}\label{fig:wwpe}
+%  \caption{The weighted permutation entropy of one run of SVD.
+The gray bands are regions where the window overlaps regimes.
+The window size used is 5,000 $\times$ 100,000 instructions and the word length is $6$.
+For reference the instructions per cycle of \svd are plotted as a ghost behind this plot.
+Each color on the ghosted time series corresonds to the different regimes as determined by visual inspection.
+From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
 %\end{figure}

### 2014-03-27 model.tex
adding RW Figures
edited text to fit new figure.  also added a to-do note in red

@@
 Figure~\ref{fig:forecast-example} shows example forecasts made using
-the \naive, ARIMA, and LMA methods for the \col, \gcc, and \svdfive
-time series.
+all four methods for the \col, \gcc, and \svdfive time series.
 \begin{figure*}[htbp]
   \centering
+
+  \begin{subfigure}{0.6\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/colRWForecast.png}
+    \caption{\col\\ random walk }
+    \label{fig:gccMEAN}
+  \end{subfigure}%
+   \begin{subfigure}{0.6\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/gccRWForecast.png}
+    \caption{\gcc\\ random walk }
+    \label{fig:gccMEAN}
+  \end{subfigure}% 
+     \begin{subfigure}{0.6\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast.png}
+    \caption{\svdfive\\ random walk}
+    \label{fig:gccMEAN}
+  \end{subfigure}%      
+  \\  
       \begin{subfigure}{0.6\columnwidth}
@@
 %
-{\color{red} Joshua will add a row of random-walk plots at the top}
 %
@@
 distributed, for example, in the case of LMA on \gcc.
+{\color{red} [[We need to comment on and explain the structure in parts
+      (a) and (g).]]}
 

### 2014-03-27 model.tex
typos in model

@@
 prediction is very effective for the \svdfive signal, on the other
-hand, so the denominator is small and the $MASE$ scores are skewes
+hand, so the denominator is small and the $MASE$ scores are skewed
 upwards.  Normalization is, as is often the case, a two-edged sword.

### 2014-03-28 wpe.tex
add a citation about the relationship between KS and Shannon entropy
liz draft of wpe.tex

@@
 \section{Measuring Structural Complexity }\label{sec:meaComplex}
-{\color{blue} EDITABLE}
 
-The goal of the section is to describe how we reasonably estimate the entropy of an arbitrary, real-valued time series.
-We will first detail why this is a highly non-trivial task, and then we will weave a path through which this can in fact be accomplished.
-This path will draw upon a myriad of tools and results from a variety of fields including time series analysis, dynamical systems, stochastic processes, and chaos.
+The goal of the section is to describe our solution to the challenging
+problem of estimating the entropy of an arbitrary, real-valued time
+series.  This approach draws upon methods and results from a variety
+of fields including time-series analysis, dynamical systems, and
+stochastic processes.
 
-For the purposes of this paper one can view the Shannon entropy, and in particular its growth rate  with respect to word length (the \emph{Shannon entropy rate}), as a measure of complexity and unpredictability in a time series.
-A high-entropy rate time series is almost completely unpredictable---and conversely a low entropy rate signal is often quite predictable.
-This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Kolmogorov-Sanai (KS) entropy is equal to the sum of the positive Lyapunov exponents, $\lambda_i$.
-The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.
-The faster the divergence, the larger the entropy and the more difficult prediction becomes.
-Further, the KS entropy is defined as the supremum of the Shannon entropy rates of all partitions.
-The partition which achieves this supremum is known as a \emph{generating partition}.
+For the purposes of this paper, we view the Shannon entropy---in
+particular its growth rate with respect to word length (the
+\emph{Shannon entropy rate})---as a measure of complexity and
+unpredictability in a time series.  A time series with a high entropy
+rate is almost completely unpredictable; conversely one with low
+entropy rate is often quite predictable. {\color{red} [[does the claim
+      in the previous sentence need a citation, a justification, or an
+      ``in practice,''?]]}  This can be made more rigorous: Pesin's
+relation \cite{pesin77} states that in chaotic dynamical systems, the
+Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
+Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
+quantify the rate at which nearby states of the system diverge with
+time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
+x(0) \right|$.  The faster the divergence, the larger the entropy, and
+the more difficult prediction becomes.  The KS entropy is defined as
+the supremum of the Shannon entropy rates of all
+partitions~\cite{petersen1989}. The partition that achieves this
+supremum is the \emph{generating partition} discussed in
+Section~\ref{sec:related}.
 
-From a different point of view, we can consider the information (as measured by the Shannon entropy) contained in a single observable, say the \emph{present}, of the system.
-This information can be partitioned in to two components: the information shared with past observations, the mutual information between the past and present, and the information in the present which is not contained in the past, the conditional entropy of the present given the past.
-The first part is known as the \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the aforementioned \emph{Shannon entropy rate}.
-In theory, the more redundancy in a signal the more predictable it is.
-The difference between prediction methods is the types of redundancy (linear, non-linear, etc) they can detect and make use of.
+From a different point of view, we can consider the information (as
+measured by the Shannon entropy) contained in a single observable, say
+the \emph{present}, of the system. This information can be partitioned
+into two components: the information shared with past
+observations---e.g., the mutual information between the past and
+present---and the information in the present that is not contained in
+the past (aka ``the conditional entropy of the present given the
+past'').  The first part is known as the \emph{redundancy}, as it is
+information in the present which is also in the past.  The second part
+is the aforementioned \emph{Shannon entropy rate}.  In theory, the
+more redundancy in a signal, the more predictable it is {\color{red}
+  [[citation for that theory?]]}.  Redundancy is critical to
+prediction, in general.  And the specific \emph{form} of the
+redundancy dictates whether a particular prediction method will work
+well or poorly on the corresponding signal.  A linear method cannot
+detect or make use of nonlinear redundancy, for instance.  This issue
+is central to the claims in this paper and the discussion in the
+following section.
 
-Utilizing Shannon entropy rate as a measure of temporal complexity is not a new idea \cite{Shannon1951, mantegna1994linguistic}, but these approaches required categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite \emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is effectively\footnote{Measurements from real devices are always discrete though typically of very high resolution, for example \texttt{double} precision.
-Though these values are discete, for the purposes of entropy calculations they are effectively a continuum.} real-valued.
-To get around this, one must discretize the data --- typically achieved by binning.
-Unfortunately, this is rarely a good solution to the problem, as the binning of the values introduces an additional dynamic on top of the intrinsic dynamics whose entropy is desired\cite{bollt2001}.
-The field of symbolic dynamics studies how to discretize a time series in such a way that the intrinsic behavior is not perverted, but these methods are fragile in the face of noise and furthermore require understanding of the underlying system, which defeats the purpose of measuring the entropy to quantify predictability in the first place.
+Utilizing Shannon entropy rate as a measure of temporal complexity is
+not a new idea \cite{Shannon1951, mantegna1994linguistic}, but
+previous approaches required categorical data: $x_t \in \mathcal{S}$
+for some finite or countably infinite \emph{alphabet} $\mathcal{S}$.
+Data taken from real-world systems is, however,
+effectively\footnote{Measurements from finite-precision sensors are
+  discrete, but data from modern high-resolution sensors are, for the
+  purposes of entropy calculations, effectively continuous.}
+real-valued.  To analyze real-valued data using a method that requires
+categorical values, one must discretize the data---typically by
+binning.  Unfortunately, this is rarely a good solution, as the
+binning of the values introduces spurious dynamics~\cite{bollt2001}.
+The field of symbolic dynamics offers discretization methods that do
+not disturb the intrinsic behavior.  These methods are, however,
+fragile in the face of noise; worse yet, they require knowledge of the
+underlying system.  This is not useful in our context, where we want
+to measure the entropy to quantify predictability.
 
-Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural complexity measure for time series'' \cite{bandt2002per}.
-Permutation entropy employs a method of discretizing real-valued time series that follows the intrinsic behavior of the system under examination.
-Rather than looking at the statistics of sequences of values, as is done when computing the Shannon entropy, permutation entropy looks at the statistics of the \emph{orderings} of sequences of values using ordinal analysis.
-Ordinal analysis of a time series is the process of mapping successive time-ordered elements of a time series to their value-ordered permutation of the same size.
-By way of example, if $(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.
-This method has many features; among other things, it is generally robust to observational noise and requires no knowledge of the underlying mechanisms.
+Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a
+``natural complexity measure for time series''
+\cite{bandt2002per}. Permutation entropy employs a method of
+discretizing real-valued time series that follows the intrinsic
+behavior of the system under examination.  Rather than looking at the
+statistics of sequences of values, as is done when computing the
+Shannon entropy, permutation entropy looks at the statistics of the
+\emph{orderings} of sequences of values using ordinal
+analysis. Ordinal analysis of a time series is the process of mapping
+successive time-ordered elements of a time series to their
+value-ordered permutation of the same size.  By way of example, if
+$(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern},
+$\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  The
+ordinal pattern of the permutation $(x_1, x_2, x_3) = (9, 7, 1)$ is
+$321$.  {\color{red} [[This method has many features; among other
+      things, it is generally robust to observational noise and
+      requires no knowledge of the underlying mechanisms.]] [[This
+      sentence is out of place.]]}
 
@@
   \end{align*}
-  Where $P(\pi)$ quantifies the probability of an ordinal and $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $\ell \ge 2$ is defined as
+  where $P(\pi)$ quantifies the probability of an ordinal and
+  $|\cdot|$ is set cardinality. The \emph{permutation entropy} of
+  order $\ell \ge 2$ is defined as
   \begin{align*}
@@
 Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
-With this in mind, it is common in the literature to normalize permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.
-With this convention, ``low'' PE is close to 0 and ``high'' PE is close to 1.
-Finally, it should be noted that the permutation entropy has been shown to be identical to the Kolmolgorov-Sinai entropy for many large classes of systems \cite{amigo2012permutation}, so long as observational noise is sufficiently small.
-As mentioned before, this is equal to the Shannon entropy rate of a generating partition of the system.
-What this normalization allows us to do, then, is approximate the redundancy of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
+With this in mind, it is common in the literature to normalize
+permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
+this convention, ``low'' PE is close to 0 and ``high'' PE is close to
+1. Finally, it should be noted that the permutation entropy has been
+shown to be identical to the Kolmolgorov-Sinai entropy for many large
+classes of systems \cite{amigo2012permutation}, so long as
+observational noise is sufficiently small. As mentioned before, this
+is equal to the Shannon entropy rate of a generating partition of the
+system.  {\color{red} [[What this normalization allows us to do, then,
+      is approximate the redundancy of a signal by $1 -
+      \frac{H(\ell)}{\log_2(\ell!)}$.]] [[This sentence is out of
+      place.  Shouldn't it be before the sentence that starts with
+      ``Finally...''?]]}
 
-Here we will be utilizing a variation of the permutation entropy, the \emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}.
-The weighted permutation entropy attempts to correct for observational noise which is larger than some trends in the data, but smaller than the larger scale features --- for example, a signal that switches between two fixed points with noise about those fixed points.
-The weighted permutation entropy would be dominated by the switching rather than by the stochastic fluctuation.
-To accomplish this, the \emph{weight} of a permutation is taken into account:
+Here we will be utilizing a variation of the basic permutation entropy
+technique, the \emph{weighted permutation entropy} (WPE), which was
+introduced in~\cite{fadlallah2013}.  The intent behind the weighting
+is to correct for observational noise that is larger than the trends
+in the data, but smaller than the larger scale features.  Consider,
+for example, a signal that switches between two fixed points and
+contains some additive noise.  {\color{red} The weighted permutation
+  entropy would be dominated by the switching rather than by the
+  stochastic fluctuation. To accomplish this, [[I am completely
+      confused by the wording here.  Do you mean that the {\bf
+        original} PE would be dominated?  And that you don't want
+      that?  If not, you need to fix this wording -- i.e., ``would''
+      to ``should'' and an explanation of why one would want to
+      calculate a PE that captures the switching.]]} the \emph{weight}
+of a permutation is taken into account:
 \begin{align*}
@@
 \end{align*}
-where $x_{t+1:t+\ell}$ is a sequence of values $x_{t+1}, \ldots, x_{t+\ell}$, and $\bar{x}_{t+1:t+\ell}$ is the arithmetic mean of those values.
+where $x_{t+1:t+\ell}$ is a sequence of values $x_{t+1}, \ldots,
+x_{t+\ell}$, and $\bar{x}_{t+1:t+\ell}$ is the arithmetic mean of
+those values.
 
@@
 \end{align*}
-where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise.
-Effectively, this weighted probability enhances permutations involved in ``large'' features and demotes permutations which are small in amplitude relative to the features of the time series.
-The weighted permutation entropy is then:
+where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
+this weighted probability enhances permutations that are involved in
+``large'' features and de-emphasizes permutations that are small in
+amplitude relative to the features of the time series.  {\color{red}
+  [[The weighted permutation entropy is then:]] [[you used exactly
+      these words above.  what is the textual causality of this
+      ``then'' and that ``then''?]]}
 \begin{align*}
@@
 \end{align*}
-which can also be normalized by dividing by $\log_2(\ell!)$, and will be in all the results of this paper.
+which can also be normalized {\color{red} [[why?]]}  by dividing by
+$\log_2(\ell!)$.  This normalization is used in all the results that
+follow.
 
-In practice, calculating permutation entropy and weighted permutation entropy involves choosing a good value for the word length $\ell$.
-The primary consideration is that the value be large enough that forbidden ordinals are discovered, yet small enough that reasonable statistics over the ordinals are gathered: e.g.:
-\begin{align*}
-  \ell = \operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \},
-\end{align*}
-assuming an average of 100 counts per ordinal is sufficient.
-In the literature, $3 \le \ell \le 6$ is a standard choice --- generally without any formal justification.
-In theory, the permutation entropy should reach an asymptote with increasing $\ell$, but that requires an arbitrarily long time series.
-In practice, what one should do is calculate the \emph{persistent} permutation entropy by increasing $\ell$ until the result converges, but data length issues can intrude before that convergence is reached.
-Here, we found that $\ell = 6$ was a good choice, because with $\ell \geq 7$ many signals had entropies plummet due to finite data effects and so with a word length of 6 we are capturing as much structure as possible without sacraficing the accuracy of ordinal statistics.
+In practice, calculating permutation entropy and weighted permutation
+entropy involves choosing a good value for the word length $\ell$. The
+primary consideration here is that the value be large enough that
+forbidden ordinals are discovered, yet small enough that reasonable
+statistics over the {\color{red} [[existing??]]}  ordinals are
+gathered.  If an average of 100 counts per ordinal is considered to be
+sufficient, for instance, then $\ell =
+\operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \}$.  In the
+literature, $3 \le \ell \le 6$ is a standard choice---generally
+without any formal justification.  In theory, the permutation entropy
+should reach an asymptote with increasing $\ell$, but that can require
+an arbitrarily long time series. In practice, what one should do is
+calculate the \emph{persistent} permutation entropy by increasing
+$\ell$ until the result converges, but data length issues can intrude
+before that convergence is reached.  We used this approach to choose
+$\ell = 6$ for the experiments in the following section.  This value
+represents a good balance between accurate ordinal statistics and
+finite-data effects.
+
+% was a good choice, because with $\ell \geq 7$ many signals had
+% entropies plummet due to finite data effects and so with a word length
+% of 6 we are capturing as much structure as possible without
+% sacraficing the accuracy of ordinal statistics.
 

### 2014-03-28 wpe.tex
added haven paragraph

@@
 
+As a final note: there has been prior work under a similar title to
+ours~\cite{haven}, but there are only superficial similarities between
+the two research projects.  Haven {\sl et al.} utilize the relative
+entropy to study how well small ensembles predict the distribution of
+future states of dynamical (and potentially chaotic) systems.  Our
+work quantifies the predictability of a single observed time series
+using weighted permutation entropy and makes no assumptions about the
+underlying dynamics.  {\color{red}[[Why is WPE better than relative
+      entropy?]]}.
+
 % was a good choice, because with $\ell \geq 7$ many signals had

### 2014-03-28 wpe.tex
more work on the wpe section

--- 2014-03-28 wpe.tex
@@
 \emph{Shannon entropy rate})---as a measure of complexity and
-unpredictability in a time series.  A time series with a high entropy
-rate is almost completely unpredictable; conversely one with low
-entropy rate is often quite predictable. {\color{red} [[does the claim
-      in the previous sentence need a citation, a justification, or an
-      ``in practice,''?]]}  This can be made more rigorous: Pesin's
+unpredictability in a time series.  Time-series consisting of IID 
+random variables, such as white noise, have maximal entropy rates, 
+whereas highly structured time-series, for example periodic, have very
+low or zero entropy rate. A time series with a high entropy rate is 
+almost completely unpredictable; conversely one with low entropy rate
+is often quite predictable. This can be made more rigorous: Pesin's
 relation \cite{pesin77} states that in chaotic dynamical systems, the
@@
 is equal to the Shannon entropy rate of a generating partition of the
-system.  {\color{red} [[What this normalization allows us to do, then,
-      is approximate the redundancy of a signal by $1 -
-      \frac{H(\ell)}{\log_2(\ell!)}$.]] [[This sentence is out of
-      place.  Shouldn't it be before the sentence that starts with
-      ``Finally...''?]]}
+system. This transitive chain of equalities, from permutation entropy 
+to Shannon entropy rate via the KS entropy, then allows us to
+approximate the redundancy --- being the dual of the Shannon entropy 
+rate --- of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
 
@@
 for example, a signal that switches between two fixed points and
-contains some additive noise.  {\color{red} The weighted permutation
-  entropy would be dominated by the switching rather than by the
-  stochastic fluctuation. To accomplish this, [[I am completely
-      confused by the wording here.  Do you mean that the {\bf
-        original} PE would be dominated?  And that you don't want
-      that?  If not, you need to fix this wording -- i.e., ``would''
-      to ``should'' and an explanation of why one would want to
-      calculate a PE that captures the switching.]]} the \emph{weight}
-of a permutation is taken into account:
+contains some additive noise. The PE is dominated by the noise about 
+the fixed points driving it to be $\approx 1$, in some sense hiding
+the fact that the signal is actually quite structed. The weighted 
+permutation entropy on the other hand would be dominated by the 
+switching rather than by the stochastic fluctuations. To accomplish 
+this, the \emph{weight} of a permutation is taken into account:
 \begin{align*}
-  w(x_{t+1:t+\ell}) = \frac{1}{\ell} \sum_{x_i \in \{x_{t+1:t+\ell}\}}
-                      \left( x_i - \bar{x}_{t+1:t+\ell} \right)^2
+  w(x_{t+1}^\ell) = \frac{1}{\ell} \sum_{i = t}^{t+\ell}
+                      \left( x_i - \bar{x}_{t+1}^\ell \right)^2
 \end{align*}
-where $x_{t+1:t+\ell}$ is a sequence of values $x_{t+1}, \ldots,
-x_{t+\ell}$, and $\bar{x}_{t+1:t+\ell}$ is the arithmetic mean of
+where $x_{t+1}^\ell$ is a sequence of values $x_{t+1}, \ldots,
+x_{t+\ell}$, and $\bar{x}_{t+1}^\ell$ is the arithmetic mean of
 those values.
 
-The weighted probability of a permutation is then:
+The weighted probability of a permutation is defined as:
 \begin{align*}
-  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
+  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell) \cdot \delta(\phi(x_{t+1}^\ell), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell)}
 \end{align*}
@@
 ``large'' features and de-emphasizes permutations that are small in
-amplitude relative to the features of the time series.  {\color{red}
-  [[The weighted permutation entropy is then:]] [[you used exactly
-      these words above.  what is the textual causality of this
-      ``then'' and that ``then''?]]}
+amplitude relative to the features of the time series. Using the
+standard form of an entropy, the weighted permutation entropy is then:
 \begin{align*}
@@
 \end{align*}
-which can also be normalized {\color{red} [[why?]]}  by dividing by
-$\log_2(\ell!)$.  This normalization is used in all the results that
-follow.
+which can also be normalized by dividing by $\log_2(\ell!)$, making $0
+\le \textrm{WPE} \le 1$.  This normalization is used in all the
+results that follow.
 
@@
 forbidden ordinals are discovered, yet small enough that reasonable
-statistics over the {\color{red} [[existing??]]}  ordinals are
+statistics over the ordinals are
 gathered.  If an average of 100 counts per ordinal is considered to be
@@
 finite-data effects.
+
+As a final note: there has been prior work under a similar title to
+ours~\cite{haven2005}, but there are only superficial similarities 
+between the two research projects.  Haven {\sl et al.} utilize the 
+relative entropy to quantify how much better a distribution over 
+system states evolved from small ensembles predicts the future of a 
+dynamical system compared to its asymptotic distribution. Our work 
+quantifies the predictability of a single observed time series using 
+weighted permutation entropy and makes no assumptions about the 
+underlying dynamics. These approaches are very different: Haven's 
+compares two distributions, the average behavior versus the behaviors 
+that can be seen given a reasonably known state in the past; ours 
+attempts to quantify how well an appropriate predictor will do for a 
+given time-series without any understanding of how that time-series
+was generated.
 

### 2014-03-28 wpe.tex
more edits to wpe

@@
 information in the present which is also in the past.  The second part
-is the aforementioned \emph{Shannon entropy rate}.  In theory, the
-more redundancy in a signal, the more predictable it is {\color{red}
-  [[citation for that theory?]]}.  Redundancy is critical to
-prediction, in general.  And the specific \emph{form} of the
-redundancy dictates whether a particular prediction method will work
-well or poorly on the corresponding signal.  A linear method cannot
-detect or make use of nonlinear redundancy, for instance.  This issue
-is central to the claims in this paper and the discussion in the
-following section.
+is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
+the more redundancy in a signal, the more predictable it is.  
+Redundancy is critical to prediction, in general.  And the specific 
+\emph{form} of the redundancy dictates whether a particular prediction 
+method will work well or poorly on the corresponding signal.  A linear 
+method cannot detect or make use of nonlinear redundancy, for 
+instance.  This issue is central to the claims in this paper and the 
+discussion in the following section.
 
@@
 ``natural complexity measure for time series''
-\cite{bandt2002per}. Permutation entropy employs a method of
+\cite{bandt2002per}. The permutation entropy employs a method of
 discretizing real-valued time series that follows the intrinsic
-behavior of the system under examination.  Rather than looking at the
-statistics of sequences of values, as is done when computing the
-Shannon entropy, permutation entropy looks at the statistics of the
-\emph{orderings} of sequences of values using ordinal
-analysis. Ordinal analysis of a time series is the process of mapping
-successive time-ordered elements of a time series to their
+behavior of the system under examination, and has many advantages 
+including being robust to observational noise and its application does 
+not require any knowledge of the underlying mechanisms.  Rather than 
+looking at the statistics of sequences of values, as is done when 
+computing the Shannon entropy, permutation entropy looks at the 
+statistics of the \emph{orderings} of sequences of values using 
+ordinal analysis. Ordinal analysis of a time series is the process of 
+mapping successive time-ordered elements of a time series to their
 value-ordered permutation of the same size.  By way of example, if
@@
 ordinal pattern of the permutation $(x_1, x_2, x_3) = (9, 7, 1)$ is
-$321$.  {\color{red} [[This method has many features; among other
-      things, it is generally robust to observational noise and
-      requires no knowledge of the underlying mechanisms.]] [[This
-      sentence is out of place.]]}
+$321$.
 

### 2014-03-28 pecomplexity.tex
title
putting title back
update affiliation format and stuff

--- 2014-03-24 pecomplexity.tex
@@
 \documentclass[%
+pre,
 reprint,
-%superscriptaddress,
+superscriptaddress,
 %groupedaddress,
@@
 %\\ 
-Quantifying Predictability through Structural Complexity
+Quantifying Time-Series Predictability through Structural Complexity
 %\\ 
 %Quantifying Predictability using Structural Complexity 
-% \\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
-% \\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
+%\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
+%\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
 }
@@
 
-\author{Joshua Garland}\email{joshua.garland@colorado.edu}
-\author{Ryan James}%
+\author{Joshua Garland}
+\email{joshua.garland@colorado.edu}
+\affiliation{Department of Computer Science, University of Colorado at Boulder. Colorado, USA}
+
+\author{Ryan James}
 \email{ryan.james@colorado.edu}
-\affiliation{%
- Department of Computer Science, University of Colorado at Boulder. Colorado, USA}%
-%\collaboration{MUSO Collaboration}%\noaffiliation
-\author{Elizabeth Bradley}\email{lizb@colorado.edu}
-\affiliation{
- Department of Computer Science, University of Colorado at Boulder. Colorado, USA
-}%
-\affiliation{
- Santa Fe Institute, New Mexico, USA\\
- %This line break forced% with \\
-}%
+\affiliation{Department of Computer Science, University of Colorado at Boulder. Colorado, USA}
+
+\author{Elizabeth Bradley}
+\email{lizb@colorado.edu}
+\affiliation{Department of Computer Science, University of Colorado at Boulder. Colorado, USA}
+\affiliation{Santa Fe Institute, New Mexico, USA}
 
@@
 of each forecast method.
-\begin{description}
-\item[Usage]
-Secondary publications and information retrieval purposes.
-\item[PACS numbers]
-May be entered using the \verb+\pacs{#1}+ \pacs{05.45.Tp} command.
-Time series analysis (nonlinear dynamics) 05.45.Tp
-Information theory, 89.70.-a
-	channel capacity in, 89.70.Kn
-	communication complexity in, 89.70.Hj
-	computational complexity in, 89.70.Eg
-	entropy in, 89.70.Cf
-	general biological information, 87.10.Vg
-	in neuroscience, 87.19.lo
 
-Nonlinear dynamics, 05.45.-a
+\vspace{0.1in}
+\noindent
+{\bf Keywords}: entropy, permutation entropy, weighted permutation
+entropy, time-series analysis, predictibility
 
-	89.75.Kd	Patterns (complex systems)
+\end{abstract}
+\pacs{
+05.45.Tp    % Time series analysis (nonlinear dynamics)
+89.70.-a    % Information theory
+89.70.Hj    %   communication complexity
+89.70.Cf    %   entropy
+05.45.-a    % Nonlinear dynamics
+89.75.Kd    %   Patterns (complex systems)
+95.75.Wx    % Time series analysis, time variability
+95.75.Pq    % Mathematical procedures and computer techniques
+07.05.-t    % Computers in experimental physics
+}
 
-Computational Physics	
-95.75.Wx	Time series analysis, time variability
-95.75.Pq	Mathematical procedures and computer techniques
-07.05.-t	Computers in experimental physics
-
-\item[Structure]
-You may use the \texttt{description} environment to structure your abstract;
-use the optional argument of the \verb+\item+ command to give the category of each item. 
-\end{description}
-\end{abstract}
-
-\pacs{Valid PACS appear here 95.75.Wx	}% PACS, the Physics and Astronomy
-                             % Classification Scheme.
-%\keywords{Suggested keywords}%Use showkeys class option if keyword
-                              %display desired
 \maketitle

### 2014-03-31 results.tex
edits,figures and variables

@@
   col\_major}, {\tt 403.gcc}, and the six different regimes of the
-{\tt dgesdd} signal in Figure~\ref{fig:wwpe}.  The objective of these
+{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
 experiments was to explore how prediction accuracy is related to WPE,

### 2014-03-31 wpe.tex
edits,figures and variables

@@
 not a new idea \cite{Shannon1951, mantegna1994linguistic}, but
-previous approaches required categorical data: $x_t \in \mathcal{S}$
+previous approaches required categorical data: $x_i \in \mathcal{S}$
 for some finite or countably infinite \emph{alphabet} $\mathcal{S}$.
@@
 
-  Given a time series $\{x_t\}_{t = 1,\dots,T}$.
+  Given a time series $\{x_i\}_{i = 1,\dots,N}$.
 Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$.
-For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$:
+For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_i\}_{i = 1,\dots,N}$:
   \begin{align*}
-    P(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
+    P(\pi) = \frac{\left|\{i|i \leq N-\ell,\phi(x_{i+1},\dots,x_{i+\ell}) = \pi\}\right|}{N-\ell+1}
   \end{align*}
@@
 \begin{align*}
-  w(x_{t+1}^\ell) = \frac{1}{\ell} \sum_{i = t}^{t+\ell}
-                      \left( x_i - \bar{x}_{t+1}^\ell \right)^2
+  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i}^{i+\ell}
+                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
 \end{align*}
-where $x_{t+1}^\ell$ is a sequence of values $x_{t+1}, \ldots,
-x_{t+\ell}$, and $\bar{x}_{t+1}^\ell$ is the arithmetic mean of
+where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
+x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
 those values.
@@
 \begin{align*}
-  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell) \cdot \delta(\phi(x_{t+1}^\ell), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1}^\ell)}
+  P_w(\pi) = \frac{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell) \cdot \delta(\phi(x_{i+1}^\ell), \pi) }{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell)}
 \end{align*}
@@
 sufficient, for instance, then $\ell =
-\operatornamewithlimits{argmax}_n \{ T \gtrapprox 100 n! \}$.  In the
+\operatornamewithlimits{argmax}_{\hat{\ell}} \{ N \gtrapprox 100 \hat{\ell}! \}$.  In the
 literature, $3 \le \ell \le 6$ is a standard choice---generally

### 2014-04-01 related.tex
some edits, and making comments red

@@
 another.  On the other hand, deterministic systems---including chaotic
-ones---have maximal redundancy and thus can be perfectly predicted if
-observed for long enough~\cite{weigend-book}.  In practice, it is
+ones---have high redundancy which is maximal in the infinite limit, 
+and thus can be perfectly predicted if observed for long 
+enough~\cite{weigend-book}.  In practice, it is
 quite difficult to estimate the redundancy of an arbitrary,
 real-valued time series.  Doing so requires knowing either the
-Kolmolgorov-Sinai entropy or the values of all positive Lyapunov
+Kolmogorov-Sinai entropy or the values of all positive Lyapunov
 exponents of the system.  Both of these calculations are

### 2014-04-01 results.tex
some edits, and making comments red

@@
 {\tt dgesdd$_1$}, which we have excluded for reasons explained in the
-text.  [[Ryan, is that correct?]]}
+text.}
   \label{fig:wpe_vs_mase_best}
@@
 with the nonlinear LMA method are closer to the dashed line, whereas
-the naive and ARIMA prediction clusters are more spread out [[need to
-    revisit this wording when get the third image]].  This is a
+the naive and ARIMA prediction clusters are more spread out {\color{red}[[need to
+    revisit this wording when get the third image]]}.  This is a
 geometric validation of the second conjecture in this paper, as
@@
 0.003$, which is in the center of the complexity spectrum described in
-Section~\ref{sec:intro}.  [[This suggests that roughly half of the
+Section~\ref{sec:intro}.  {\color{red}[[This suggests that roughly half of the
     signal is predictive structure.]]  [[I'm still uncomfortable with
     that phrasing; let's revisit it after we have the terminology and
-    definitions hammered out.]]
+    definitions hammered out.]]  }
 
@@
 horizontally in ARIMA, far to the right of the dashed line (i.e., much
-worse than the other methods).  [[We need to come up with a hypothesis
+worse than the other methods).  {\color{red}[[We need to come up with a hypothesis
     about this spread and skew, even if it's totally tentative.  We
@@
     paragraphs---or perhaps put it with the \svdone discussion later
-    on?]]
+    on?]]}
 
@@
   high, which makes guessing the mean a particularly bad strategy.
-  The same thing is true of the \svdthree signal.[[Joshua, it isn't the variance it is the strength in the bimodal distribution and only predicting in one mode of the distribution...]]}.  Again, the
+  The same thing is true of the \svdthree signal.{\color{red}[[Joshua, it isn't the variance it is the strength in the bimodal distribution and only predicting in one mode of the distribution...]]}}.
+Again, the
 positions of these points on a WPE vs. MASE plot---significantly below
@@
 transmits very little information forward in time and provides almost
-no structure for prediction methods to work with.  [[What we see:
+no structure for prediction methods to work with.  {\color{red}[[What we see:
     \naive does the best, then LMA, then ARIMA.  Hypothesis: that that
@@
     ARIMA can.  \naive does better than LMA because it's filtering out
-    the noise.]]
+    the noise.]]}
 

### 2014-04-02 intro.tex
edits and consistency changes

@@
 In particular, we wish to determine---without knowing anything about
-the generating process---whether a time series is too complex to
+the generating process---whether a time series is too complex to successfully
 predict.  An important practical corollary to this is a strategy for
@@
 time series.  From a practitioner's standpoint, it would be useful to
-know {\sl a priori} if a trace contains enough predictive structure to
+know {\sl a priori} if a time series contains enough predictive structure to
 make it worth spending the time and effort searching for a good
@@
 experiments.  While this is not a common laboratory experiment, it is
-a highly appropriate choice here.  Computers are among the most
+a highly appropriate choice here.
+{\color{red} [[Liz, these are the few sentences I was worried about being too similar but we may have only used them in a grant proposal.]]} Computers are among the most
 complex engineered artifacts in current use.  Modern microprocessor

### 2014-04-02 methods.tex
edits and consistency changes

@@
 programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
-from the LAPACK linear algebra package (\svd).  In these experiments,
-the scalar observation $x_{i}$ at time $i$ was a measurement of how
-well each program was able to exploit the native parellelism of the i7
-chip.  To record these measurements, we used the {\tt libpfm4}
+from the LAPACK linear algebra package (\svd).
+In these experiments, the scalar observation $x_i$ was a measurement of the processor performance at time $i$ during the execution of each program.
+To record these measurements, we used the {\tt libpfm4}
 library, via PAPI
@@
 surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
-of the processor utilization during the execution this program is
+of the processor performance during the execution this program is
 shown in Figure~\ref{fig:col-ts}.
@@
 \svd is a Fortran program from the LAPACK linear algebra package
-\cite{lapack}.  It calculates the singular value decomposition of an
+\cite{lapack}.  It calculates the singular value decomposition of a
 rectangular $M$ by $N$ real-valued matrix.  For our experiments, we
@@
 For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
-cycles), roughly 1.8 instructions are executed per cycle, on the
+instructions), roughly 1.8 instructions are executed per cycle, on the
 average, by the eight processing units on this chip.  After that, the
@@
 of course, particularly since the regimes exhibit some fractal
-structure.  Thus, it may the case that more than one generating
+structure.  Thus, it may be the case that more than one generating
 process may be at work in each of our segments.  This is a factor in

### 2014-04-02 model.tex
some edits, and making comments red
edits and consistency changes

@@
 consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
-These signals are volatile, noisy, and possess very little predictive
-structure, but they do not---on the average---vary a great deal, so
+These signals are constantly changing, noisy, and possess very little predictive
+structure, but those variations are not---on the average---very large, so
 the random-walk method's strategy of simply guessing the last known
@@
 methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
-designed to do exactly that.  Note, though, that periodic patterns
-appear only in signals that are at the low end of the complexity
-spectrum.  If a signal contains little predictive structure, forecast
+designed to do exactly that. 
+% Note, though, that periodic patterns
+%appear only in signals that are at the low end of the %complexity
+%spectrum. 
+However, if a signal contains little predictive structure, forecast
 strategies like ARIMA and LMA have nothing to work with and thus will
@@
 next time step through a weighted average of past observations: $$p_i
-= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_i$ are
-computed using a polynomial of degree $l$ that is fit to the past
-$l+1$ values of $x_i$.  To account for noise in the data, one can add
+= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
+computed generally using either an ordinary least squares approach, or with the method of moments using the Yule-Walker equations.  To account for noise in the data, one can add
 a so-called ``moving average'' term to the model; to remove
@@
 generated and/or transmitted in a nonlinear way, however, a global
-linear fit is inappropriate and ARIMA forecasts will be inaccurate.
+linear fit is inappropriate and ARIMA forecasts can be inaccurate.
 Another weakness of this method is prediction horizon, since an ARIMA
@@
 Geometric structure in these kinds of plots is an indication of
-structure in the information generation/transmission process the
+structure in the information generation/transmission process that
 produced the time series.  The dynamical systems community has
@@
   system state space; the method was later extended to embedded
-  dynamics by Kantz~\cite{kantz-lma-citation}.}
+  dynamics by Kantz~\cite{kantz97}.}
 space~\cite{lorenz-analogues}.  Even this simple algorithm---which
@@
 ``one step prediction'' process is not technically necessary in the
-LMA method, which can generate arbitrary-length predictions, but the
-performance of the other three methods used here will degrade if the
+LMA method, which can generate arbitrary-length predictions (although these predictions will degrade over time in the presence of positive Lyapunov exponents), but the
+performance of the other three methods used here will degrade severely if the
 associated models are not periodically rebuilt.  In order to make the
@@
 structure of the distribution of the errors---roughly normally
-distributed, for example, in the case of LMA on \gcc.
-{\color{red} [[We need to comment on and explain the structure in parts
-      (a) and (g).]]}
+distributed, for example, in the case of LMA on \gcc. The triangular 
+shapes that ARIMA and random-walk produce for \col simply reflects 
+three phases of values that is visually present in the time-series.{\color{red}not happy with this explanation, I(joshua) will think about how to word this}
 
@@
 is the average in-sample forecast error for a random-walk prediction
-over a signal of length $k$.  That is, $MASE<1$ means that the
+over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$ means that the
 prediction error in question was, on the average, smaller than the
@@
 Figure~\ref{fig:forecast-example}.  While the correspondence between
-$MASE$ score and the visual appearance of a $p$ versus $c$ plot is
+$MASE$ score and the visual appearance of these kinds of plots is
 clear cut in this case, that is not always so.  The plots of LMA

### 2014-04-02 related.tex
edits and consistency changes

@@
 building an autoregressive model, for example, it is common to
-calculate the errors and verify that they are normally distributed.
+calculate forecast errors and verify that they are normally distributed.
 If they are not, that indicates that there is structure in the time
@@
 
-A practice known as deterministic/sto\-chas\-tic
+A practice known as deterministic vs. stochastic
 modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
@@
 out-of-sample error changes as a function of number of points in the
-fit. The shape of such a graph indicates the amounts of determinism
+fit. The shape of such a ``DVS" graph indicates the amounts of determinism
 and stochasticity are present in a time series.

### 2014-04-02 wpe.tex
edits and consistency changes

@@
 the fixed points driving it to be $\approx 1$, in some sense hiding
-the fact that the signal is actually quite structed. The weighted 
+the fact that the signal is actually quite structured. The weighted 
 permutation entropy on the other hand would be dominated by the 
@@
 relative entropy to quantify how much better a distribution over 
-system states evolved from small ensembles predicts the future of a 
+system states evolved from small ensembles to predict the future of a 
 dynamical system compared to its asymptotic distribution. Our work 

### 2014-04-04 intro.tex
adding a citation
rephrased copied sentences

@@
 as a particular approximation of Kolmogorov-Sinai
-entropy~\cite{KS-entropy}.  That is, we view a random-walk time series
+entropy~\cite{lind95}.  That is, we view a random-walk time series
 (which exhibits high entropy) as purely complex, whereas a periodic
@@
 experiments.  While this is not a common laboratory experiment, it is
-a highly appropriate choice here.
-{\color{red} [[Liz, these are the few sentences I was worried about being too similar but we may have only used them in a grant proposal.]]} Computers are among the most
-complex engineered artifacts in current use.  Modern microprocessor
-chips contain multiple processing units and multi-layer memories, for
-instance, and they use complicated hardware/software strategies to
-move data and threads of computation across those resources.  As a
-result, the processor and memory loads during the execution of even a
-very simple program can evolve in a very complicated fashion.
+a highly appropriate choice here.  Computers are extremely complicated
+systems.  Modern microprocessor chips contain multiple processing
+units and multi-layer memories, for instance, and they use complicated
+hardware/software strategies to maximize performance by moving data
+and threads of computation across those resources.  As a result, the
+processor and memory loads during the execution of even a very simple
+program can evolve in a very complicated fashion.
 Figure~\ref{fig:col-ipc} shows a performance trace of a three-line

### 2014-04-04 methods.tex
small edits to methods.tex

@@
 upper triangle of a 2048 $\times$ 2048 matrix in column-major order.
-As mentioned in the previous section, this simple program exhibits
+As mentioned in Section~\ref{sec:intro}, this simple program exhibits
 surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
@@
     \includegraphics[width=\columnwidth]{figs/colFullTS}
-    \caption{The instructions executed per CPU clock cycle (IPC)
-      during the execution of \col.  Each point is the average IPC
-      during a 100,000 instruction period. Figure~\ref{fig:col-ipc} is
-      a closeup of the first part of this signal.  }
+    \caption{The instructions executed per CPU clock cycle (IPC) as
+      the \col program runs.  Each point is the average IPC in a
+      100,000 instruction period. Figure~\ref{fig:col-ipc} is a
+      closeup of the first part of this signal.  }
     \label{fig:col-ts}
@@
     \includegraphics[width=\columnwidth]{figs/gccfullts}
-    \caption{The instructions executed per CPU clock cycle (IPC)
-      during the execution of \gcc. Each point is the average IPC
-      during a 100,000 instruction period.}
+    \caption{The instructions per CPU clock cycle (IPC) during the
+      execution of \gcc. Each point is the average IPC in a 100,000
+      instruction period.}
     \label{fig:gcc-ts}
@@
 \cite{lapack}.  It calculates the singular value decomposition of a
-rectangular $M$ by $N$ real-valued matrix.  For our experiments, we
-choose $M=750$ and $N=1000$ and generated the matrix entries randomly.
+rectangular $M$ by $N$ matrix with real-valued entries.  For our experiments, we
+choose $M=750$ and $N=1000$ and generate the matrix entries randomly.
 % 
@@
     \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
-    \caption{The instructions executed per CPU clock cycle (IPC)
-      during the execution of \svd. Each point is the average IPC
-      during a 100,000 instruction period.  The colors refer to
-      different {\sl segments} of the signal, as discussed in the
-      text.}
+    \caption{The instructions per CPU clock cycle (IPC) during the
+      execution of \svd. Each point is the average IPC in a 100,000
+      instruction period.  The colors identify the different {\sl
+        segments} of the signal that are discussed in the text.}
     \label{fig:svd-ts-colored}
@@
 purposes of this paper, in that a number of different generating
-processes---with a wide range of complexities---are at work at
-different points in a single time series.  The complexity of \col and
+processes---with a wide range of complexities---are at work in
+different phases of a single time series.  The complexity of \col and
 \gcc in Figures~\ref{fig:col-ts} and \ref{fig:gcc-ts} appears to be
@@
 array of examples for the analyses in the following sections.  For
-notational convenience, we refer to these 90 time series\footnote{15
-  runs, each with six regimes} as {\tt dgesdd$_i$} with $i \in
-\{1\dots6\}$ where $i$ corresponds to one of the six color-coded
-segments of \svd, ordered from left to right.  These segments, which
-were determined visually, are shown in different colors in
+notational convenience, we refer to these 90 time-series data
+sets\footnote{15 runs, each with six regimes} as {\tt dgesdd$_i$},
+with $i \in \{1\dots6\}$ where $i$ corresponds to one of the six
+segments of the signal, ordered from left to right.  These segments,
+which were determined visually, are shown in different colors in
 Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
@@
 structure.  Thus, it may be the case that more than one generating
-process may be at work in each of our segments.  This is a factor in
-the discussion of Section~\ref{sec:results}.
+process is at work in each of our segments.  This is a factor in the
+discussion of Section~\ref{sec:results}.
 

### 2014-04-04 related.tex
minor edits to related.tex

@@
 least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
-then hundreds, if not thousands, of strategies have been developed for
-a wide variety of prediction tasks.  The purpose of this paper is not
-to add a new weapon to this arsenal, nor to assess or compare the
+then, hundreds, if not thousands, of strategies have been developed
+for a wide variety of prediction tasks.  The purpose of this paper is
+not to add a new weapon to this arsenal, nor to assess or compare the
 effectiveness of existing methods.  Our goals are more general: {\sl
@@
 Quantifying predictability, which is sometimes called ``predicting
-predictability,'' is not a new problem.  Most solutions to it fall
-into two categories: model-based error analysis and model-free
-information analysis.
+predictability,'' is not a new problem.  Most of the corresponding
+solutions fall into two categories: model-based error analysis and
+model-free information analysis.
 %
@@
 another.  On the other hand, deterministic systems---including chaotic
-ones---have high redundancy which is maximal in the infinite limit, 
-and thus can be perfectly predicted if observed for long 
+ones---have high redundancy that is maximal in the infinite limit, 
+and thus they can be perfectly predicted if observed for long 
 enough~\cite{weigend-book}.  In practice, it is

### 2014-04-04 model.tex
minor edits on model.tex

@@
 These signals are constantly changing, noisy, and possess very little predictive
-structure, but those variations are not---on the average---very large, so
+structure, but their variations are not---on the average---very large, so
 the random-walk method's strategy of simply guessing the last known
@@
 choosing values for the various free parameters in the autoregressive,
-detrending, moving average, and filtering terms---e.g., the $l$ in the
-equation at the beginning of this section.  We employ the automated
-fitting techniques described in~\cite{autoARIMA} to accomplish this.
-This procedure uses sophisticated methods---KPSS unit-root
-tests~\cite{KPSSunit}, a customization of the Canova-Hansen
+detrending, moving average, and filtering terms.  We employ the
+automated fitting techniques described in~\cite{autoARIMA} to
+accomplish this.  This procedure uses sophisticated methods---KPSS
+unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
 test~\cite{Canova1995}, and the Akaike information
@@
 ``one step prediction'' process is not technically necessary in the
-LMA method, which can generate arbitrary-length predictions (although these predictions will degrade over time in the presence of positive Lyapunov exponents), but the
-performance of the other three methods used here will degrade severely if the
-associated models are not periodically rebuilt.  In order to make the
-comparison fair, we used an iterative one-step prediction schema {\sl
-  for all four methods}.  This has the slightly confusing effect of
-causing the ``test'' signal to be used both to assess the accuracy of
-each model and for periodic refitting.
+LMA method, which can generate arbitrary-length
+predictions\footnote{although the accuracy of these predictions will
+  degrade over prediction length, in the presence of positive Lyapunov
+  exponents}, but the performance of the other three methods used here
+will degrade severely if the associated models are not periodically
+rebuilt.  In order to make the comparison fair, we used an iterative
+one-step prediction schema {\sl for all four methods}.  This has the
+slightly confusing effect of causing the ``test'' signal to be used
+both to assess the accuracy of each model and for periodic refitting.
 
@@
   \begin{tabular}{|c|c|c|c|c|}
-  \hline Signal & MASE LMA & MASE ARIMA &MASE na\"{i}ve & Weighted Permutation \\ 
+  \hline Signal & LMA $MASE$  & ARIMA $MASE$ & na\"{i}ve $MASE$  & Weighted Permutation \\ 
   &  &  & &  Entropy \\ \hline
@@
  \label{default}
- \caption{MASE scores and weighted permutation entropies for all eight
+ \caption{$MASE$< scores and weighted permutation entropies for all eight
    processes studied in this paper}
@@
 random-walk predictor, while ARIMA only outperformed random walk by a
-factor of 1.7.  This is in accordance with the
-visual appearance of the corresponding images in
-Figure~\ref{fig:forecast-example}.  While the correspondence between
-$MASE$ score and the visual appearance of these kinds of plots is
-clear cut in this case, that is not always so.  The plots of LMA
-predictions of \col and \svdfive both lie near the diagonal, for
-instance, but the corresponding $MASE$ scores were very different:
-$0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times
-better than random-walk forecasts of the same signals).  This is a
-result of the normalization in the $MASE$ score.  Recall from
-Section~\ref{sec:simple} that the random-walk method performs
-especially poorly on signals that oscillate rapidly.  \col fits this
-description, so the random-walk error on this signal---the denominator
-of the three $MASE$ scores in the first row of the Table---is
-pathologically high, which skews those scores down.  Random-walk
-prediction is very effective for the \svdfive signal, on the other
-hand, so the denominator is small and the $MASE$ scores are skewed
-upwards.  Normalization is, as is often the case, a two-edged sword.
-Here, it facilitates comparisons across signals of different types and
-lengths, but its particular pathologies must be taken into account, as
-discussed at more length in Section~\ref{sec:results}.
+factor of 1.7.  This is in accordance with the visual appearance of
+the corresponding images in Figure~\ref{fig:forecast-example}.  While
+the correspondence between $MASE$ score and the visual appearance of
+these kinds of plots is clear cut in this case, that is not always so.
+The plots of LMA predictions of \col and \svdfive both lie near the
+diagonal, for instance, but the corresponding $MASE$ scores were very
+different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
+1.4 times better than random-walk forecasts of the same signals).
+This is a result of the normalization in the $MASE$ score calculation.
+Recall from Section~\ref{sec:simple} that the random-walk method
+performs especially poorly on signals that oscillate rapidly.  \col
+fits this description, so the random-walk error on this signal---the
+denominator of the three $MASE$ scores in the first row of the
+Table---is pathologically high, which skews those scores down.
+Random-walk prediction is very effective for the \svdfive signal, on
+the other hand, so the denominator is small and the $MASE$ scores are
+skewed upwards.  Normalization is, as is often the case, a two-edged
+sword.  Here, it facilitates comparisons across signals of different
+types and lengths, but its particular pathologies must be taken into
+account, as discussed at more length in Section~\ref{sec:results}.
 

### 2014-04-04 wpe.tex
minor edits to wpe, with a note at the end

@@
 
-The goal of the section is to describe our solution to the challenging
-problem of estimating the entropy of an arbitrary, real-valued time
-series.  This approach draws upon methods and results from a variety
-of fields including time-series analysis, dynamical systems, and
+Estimating the entropy of an arbitrary, real-valued time series is a
+challenging problem.  Our approach to this problem, which is the topic
+of this section, draws upon methods and results from a variety of
+fields including time-series analysis, dynamical systems, and
 stochastic processes.
@@
 \emph{Shannon entropy rate})---as a measure of complexity and
-unpredictability in a time series.  Time-series consisting of IID 
+unpredictability in a time series.  Time-series consisting of i.i.d.
 random variables, such as white noise, have maximal entropy rates, 
@@
 low or zero entropy rate. A time series with a high entropy rate is 
-almost completely unpredictable; conversely one with low entropy rate
+almost completely unpredictable; conversely, one with low entropy rate
 is often quite predictable. This can be made more rigorous: Pesin's
@@
 past'').  The first part is known as the \emph{redundancy}, as it is
-information in the present which is also in the past.  The second part
+information in the present that is also in the past.  The second part
 is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
@@
 
-Utilizing Shannon entropy rate as a measure of temporal complexity is
-not a new idea \cite{Shannon1951, mantegna1994linguistic}, but
-previous approaches required categorical data: $x_i \in \mathcal{S}$
-for some finite or countably infinite \emph{alphabet} $\mathcal{S}$.
-Data taken from real-world systems is, however,
-effectively\footnote{Measurements from finite-precision sensors are
-  discrete, but data from modern high-resolution sensors are, for the
-  purposes of entropy calculations, effectively continuous.}
-real-valued.  To analyze real-valued data using a method that requires
-categorical values, one must discretize the data---typically by
-binning.  Unfortunately, this is rarely a good solution, as the
-binning of the values introduces spurious dynamics~\cite{bollt2001}.
-The field of symbolic dynamics offers discretization methods that do
-not disturb the intrinsic behavior.  These methods are, however,
-fragile in the face of noise; worse yet, they require knowledge of the
-underlying system.  This is not useful in our context, where we want
-to measure the entropy to quantify predictability.
+Previous approaches to measuring temporal complexity via the Shannon
+entropy rate \cite{Shannon1951, mantegna1994linguistic} required
+categorical data: $x_i \in \mathcal{S}$ for some finite or countably
+infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
+systems is, however, effectively\footnote{Measurements from
+  finite-precision sensors are discrete, but data from modern
+  high-resolution sensors are, for the purposes of entropy
+  calculations, effectively continuous.}  real-valued.  To analyze
+real-valued data using a method that requires categorical values, one
+must discretize the data---typically by binning.  Unfortunately, this
+is rarely a good solution, as the binning of the values introduces
+spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
+offers discretization methods that do not disturb the intrinsic
+behavior.  These methods are, however, fragile in the face of noise;
+worse yet, they require knowledge of the underlying system.  This is
+not useful in our context, where we want to measure the entropy to
+quantify predictability.
 
@@
 to Shannon entropy rate via the KS entropy, then allows us to
-approximate the redundancy --- being the dual of the Shannon entropy 
-rate --- of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
+approximate the redundancy---being the dual of the Shannon entropy
+rate---of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
 
@@
 contains some additive noise. The PE is dominated by the noise about
-the fixed points driving it to be $\approx 1$, in some sense hiding
-the fact that the signal is actually quite structured. The weighted 
-permutation entropy on the other hand would be dominated by the 
-switching rather than by the stochastic fluctuations. To accomplish 
-this, the \emph{weight} of a permutation is taken into account:
+the fixed points driving it to $\approx 1$, which in some sense hides
+the fact that the signal is actually quite structured.  In this
+situation, the weighted permutation entropy measures on switching
+rather than the stochastic fluctuations. To accomplish this, the
+\emph{weight} of a permutation is taken into account:
 \begin{align*}
@@
 amplitude relative to the features of the time series. Using the
-standard form of an entropy, the weighted permutation entropy is then:
+standard form of an entropy, the weighted permutation entropy is:
 \begin{align*}
@@
 entropy involves choosing a good value for the word length $\ell$. The
-primary consideration here is that the value be large enough that
-forbidden ordinals are discovered, yet small enough that reasonable
-statistics over the ordinals are
-gathered.  If an average of 100 counts per ordinal is considered to be
-sufficient, for instance, then $\ell =
-\operatornamewithlimits{argmax}_{\hat{\ell}} \{ N \gtrapprox 100 \hat{\ell}! \}$.  In the
-literature, $3 \le \ell \le 6$ is a standard choice---generally
-without any formal justification.  In theory, the permutation entropy
-should reach an asymptote with increasing $\ell$, but that can require
-an arbitrarily long time series. In practice, what one should do is
-calculate the \emph{persistent} permutation entropy by increasing
-$\ell$ until the result converges, but data length issues can intrude
-before that convergence is reached.  We used this approach to choose
-$\ell = 6$ for the experiments in the following section.  This value
-represents a good balance between accurate ordinal statistics and
-finite-data effects.
+primary consideration in that choice is that the value be large enough
+that forbidden ordinals are discovered, yet small enough that
+reasonable statistics over the ordinals are gathered.  If an average
+of 100 counts per ordinal is considered to be sufficient, for
+instance, then $\ell = \operatornamewithlimits{argmax}_{\hat{\ell}} \{
+N \gtrapprox 100 \hat{\ell}! \}$.  In the literature, $3 \le \ell \le
+6$ is a standard choice---generally without any formal justification.
+In theory, the permutation entropy should reach an asymptote with
+increasing $\ell$, but that can require an arbitrarily long time
+series. In practice, what one should do is calculate the
+\emph{persistent} permutation entropy by increasing $\ell$ until the
+result converges, but data length issues can intrude before that
+convergence is reached.  We used this approach to choose $\ell = 6$
+for the experiments in the following section.  This value represents a
+good balance between accurate ordinal statistics and finite-data
+effects.
 
@@
 weighted permutation entropy and makes no assumptions about the
-underlying dynamics. These approaches are very different: Haven's 
-compares two distributions, the average behavior versus the behaviors 
-that can be seen given a reasonably known state in the past; ours 
-attempts to quantify how well an appropriate predictor will do for a 
-given time-series without any understanding of how that time-series
-was generated.
+underlying dynamics. {\color{red} These approaches are very different:
+  Haven's compares two distributions, the average behavior versus the
+  behaviors that can be seen given a reasonably known state in the
+  past; ours attempts to quantify how well an appropriate predictor
+  will do for a given time-series without any understanding of how
+  that time-series was generated. [[Seems to me like the first part of
+      this paragraph made the point.  Is this second set of stuff
+      needed?]]}
 

### 2014-04-05 model.tex
messing with the table

@@
   \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}
-  \hline Signal & LMA $MASE$  & ARIMA $MASE$ & na\"{i}ve $MASE$  & Weighted Permutation \\ 
-  &  &  & &  Entropy \\ \hline
-
+  \begin{tabular}{|c|c|c|l|c|}
+  \hline Signal & LMA  & ARIMA & na\"{i}ve  & Weighted Permutation \\ 
+  &  $MASE$ & $MASE$  & $MASE$ &  Entropy \\ \hline
  \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\
@@
 \svdthree     & $0.619 \pm0.021 $ & $0.713 \pm 0.010 $ & $31.386\pm 0.282$ &  $0.716 \pm 0.006$ \\
-
  \svdfour     & $ 0.779\pm0.036 $ & $0.979 \pm0.032 $ & $2.661\pm0.074$ & $0.825 \pm 0.008$ \\
@@
 \svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
-
- \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
+ \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
   \hline
@@
  \label{default}
- \caption{$MASE$< scores and weighted permutation entropies for all eight
+ \caption{$MASE$ scores and weighted permutation entropies for all eight
    processes studied in this paper}

### 2014-04-05 results.tex
rewrite of results.tex

@@
 LMA prediction methods, as described in Section~\ref{sec:accuracy},
-then calculated the MASE value of those predictions.  We also
+then calculated the $MASE$ value of those predictions.  We also
 calculated the WPE of each time series using a wordlength of six, as
@@
 calculations on 15 separate trials: i.e., 15 different runs of each
-program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
-these 45 predictions for each system: i.e., the lowest error over all
-15 trials and all three methods for each of the eight programs.  The
-WPE is plotted against the corresponding MASE value here in order to
-bring out the correlation between these two quantities.
+program.
+
+Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of these 45
+predictions for each system: that is, the lowest error over all 15
+trials and all three methods for each of the eight programs.  The WPE
+is plotted against the corresponding $MASE$ value here in order to bring
+out the correlation between these two quantities.
 \begin{figure}[htbp]
@@
   \caption{Weighted permutation entropy vs. mean absolute scaled error
-    (MASE) of the best prediction of each system.
+    ($MASE$) of the best prediction of each system.
 % 
@@
 \end{figure}
-
 The best-case prediction error for each of these eight systems is
@@
 formal results about a ``best'' predictor for a given time series.
-There may be other methods that produce lower MASE values than those
+There may be other methods that produce lower $MASE$ values than those
 in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
@@
 
-Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
+Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. $MASE$ plots for all 360
 experiments.
@@
   \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy}
-\caption{WPE vs.
-MASE for all trials, methods, and systems.
-\svdone, \svdthree, and \svdfive are omitted from the top plot for scale reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm 0.282$, and $20.870 \pm 0.192$ respectively.
-The dashed line is the same as in the previous figure.
-Numerical values, including means and standard deviations of the errors, can be found in Table~\ref{tab:error} in the appendix.
+\caption{WPE vs. $MASE$ for all trials, methods, and systems. \svdone,
+  \svdthree, and \svdfive are omitted from the top plot for scale
+  reasons. {\color{red} All three plots need to have the same
+    horizontal scale!}
+% 
+% ; their $MASE$ scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
+%   $20.87 \pm 0.192$ respectively. 
+% 
+Numerical values, including means and standard deviations of the
+errors, can be found in Table~\ref{tab:error}.  The dashed line is the
+same as in the previous figure.
 % 
@@
 in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
-for the corresponding system in any of these three plots.)  The WPE
-values do not vary very much across trials.  For most traces, the
-variance in MASE scores is low as well, resulting in small, tight
-clusters.  In some cases---ARIMA predictions of {\tt col\_major}, for
-instance---the MASE variance is larger, which spreads out the clusters
+for the corresponding system in any of the three plots in
+Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
+much across trials.  For most traces, the variance in $MASE$ scores is
+low as well, resulting in small, tight clusters.  In some
+cases---ARIMA predictions of {\tt col\_major}, for instance---the
+$MASE$ variance is larger, which spreads out the clusters
 horizontally.  The distribution of the cluster {\sl positions} is
-quite different in the three plots: clusters of predictions generated
-with the nonlinear LMA method are closer to the dashed line, whereas
-the naive and ARIMA prediction clusters are more spread out {\color{red}[[need to
-    revisit this wording when get the third image]]}.  This is a
-geometric validation of the second conjecture in this paper, as
-described in the following paragraphs.
+quite different in the three plots.  The mean $MASE$ scores of
+predictions generated with the nonlinear LMA method are generally
+closer to the dashed line; the ARIMA method clusters are more
+widely spread, the \naive ~clusters even more so.  This geometric
+validation of the second conjecture in this paper is discussed in more
+details in the following paragraphs.
 
@@
 complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
-graphical representations in the right-hand column of
-Figure~\ref{fig:gcc_vs_col} that the \naive and ARIMA prediction
-methods do not perform as well on this signal as the nonlinear LMA
-method.  The MASE values of the \naive and ARIMA predictions are $0.57
-\pm$ 0.001 and $0.59 \pm 0.21$, respectively, across all 15 trials.
-That is, the errors produced by these methods are a little over half
-as large as the errors produced by the random-walk method---a
-primitive strategy that simply uses the current value as the
-prediction.  However, the WPE value for the \col trials is $0.51 \pm
-0.003$, which is in the center of the complexity spectrum described in
-Section~\ref{sec:intro}.  {\color{red}[[This suggests that roughly half of the
-    signal is predictive structure.]]  [[I'm still uncomfortable with
-    that phrasing; let's revisit it after we have the terminology and
-    definitions hammered out.]]  }
+graphical representations in the left-hand column of
+Figure~\ref{fig:forecast-example} that the \naive ~and ARIMA
+prediction methods do not perform as well on this signal as the
+nonlinear LMA method.  The $MASE$ scores of the \naive ~and ARIMA
+predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
+across all 15 trials.  That is, these methods are performing only
+$\approx 1.7$ times better than the random-walk method, a primitive
+strategy that simply uses the current value as the prediction.
+However, the WPE value for the \col trials is $0.513 \pm 0.003$, which
+is in the center of the complexity spectrum described in
+Section~\ref{sec:intro}.
 
@@
 information transfer in the signal, but predictions with comparatively
-poor MASE scores---is obvious in the geometry of the top two images in
-Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
-the right of the dashed line.  {\sl This indicates that these methods
-  are not leveraging the available information in the signal.}  The
-dynamics of \col may be complicated, but they are not unstructured.
-This signal is nonlinear and deterministic, and if one uses a
-prediction technique that is based a nonlinear model (LMA)---rather
-than a method that simply predicts the running mean (\naive) or one
-that uses a linear model (ARIMA)---the MASE score is much better:
-$0.050 \pm 0.001$.  This prediction is 20 times more accurate than a
-random-walk forecast, which is more in line with the level of
-predictive structure that the low WPE value suggests is present in the
-signal.  The dashed line is a heuristic that captures this argument.
-It can allow practitioners to recognize when a prediction method is
-not well matched to the task at hand: that is, when the time series
-has more predictive structure than the method is able to use.  When
-the predictions are well below that line (viz., the top images in
-Figure~\ref{fig:wpe_vs_mase_all}), one should try a different method.
+poor $MASE$ scores---is obvious in the geometry of the top two images
+in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
+to the right of the dashed line.  {\sl This indicates that these
+  methods are not leveraging the available information in the signal.}
+The dynamics of \col may be complicated, but they are not
+unstructured.  This signal is nonlinear and deterministic, and if one
+uses a prediction technique that is based a nonlinear model
+(LMA)---rather than a method that simply predicts the running mean
+(\naive) or one that uses a linear model (ARIMA)---the $MASE$ score is
+much better: $0.050 \pm 0.001$.  This prediction is 20 times more
+accurate than a random-walk forecast, which is more in line with the
+level of predictive structure that the low WPE value suggests is
+present in the signal.
 
-Note that the \col points are clustered in LMA but spread out
-horizontally in ARIMA, far to the right of the dashed line (i.e., much
-worse than the other methods).  {\color{red}[[We need to come up with a hypothesis
-    about this spread and skew, even if it's totally tentative.  We
-    may want to merge this discussion into one of the other \col
-    paragraphs---or perhaps put it with the \svdone discussion later
-    on?]]}
+The dashed line in the Figures is a heuristic that captures the
+argument at the end of the previous paragraph.  It can allow
+practitioners to recognize when a prediction method is not well
+matched to the task at hand: that is, when the time series has more
+predictive structure than the method is able to use.  When the
+predictions are well below that line (as in the top and middle images
+in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
+method.
+
+Note that the \col points (blue {\color{blue}$+$} icons) are clustered
+very tightly in the lower left quadrant of the LMA and \naive ~$MASE$
+vs. WPE plots, but spread out horizontally in the ARIMA plot.
+{\color{red}[[We need to come up with a hypothesis about this spread,
+      even if it's totally tentative.  Was it that the random-walk got
+      really good in a few of the \col trials?  Or that ARIMA got
+      really bad in those trials?  In either case, why did that
+      happen---what was it about those trials that made ARIMA bark
+      and/or random walk do well?]]} The position of the LMA
+cluster---well to the left of the other two, near the dashed
+line---reflects this method's ability to capture and exploit the
+structure that is present in this signal.
 
@@
 can, in theory, be leveraged to predict the future course of the time
-series.  As mentioned above, though, in the discussion of {\tt
-  col\_major}, the effectiveness of any prediction strategy will
-depend on how well it leverages the available information; methods
-that use linear models, for instance, may fail when applied to
-nonlinear processes.
+series.  As mentioned above, though, the effectiveness of any
+prediction strategy depends on how well it leverages the available
+information: methods that use linear models, for instance, may fail
+when applied to nonlinear processes.
 % 
@@
 % 
-The match between model and task is an issue in the \svdfive
-experiments as well.  The MASE scores of the \naive and ARIMA
-predictions for this system were $20.870 \pm 0.192$ and $2.370 \pm
+This match between model and task is an issue in the \svdfive
+experiments as well.  The $MASE$ scores of the \naive ~and ARIMA
+predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
 0.051$, respectively: that is, 20.87 and 2.37 times worse than a
 simple random walk forecast of the same signals\footnote{The \naive
-  MASE score is large because the amount of variance in this signal is
-  high, which makes guessing the mean a particularly bad strategy.
-  The same thing is true of the \svdthree signal.{\color{red}[[Joshua, it isn't the variance it is the strength in the bimodal distribution and only predicting in one mode of the distribution...]]}}.
-Again, the
-positions of these points on a WPE vs. MASE plot---significantly below
-and to the right of the dashed line---should suggest to a practitioner
-that there is more structure in this signal than the corresponding
-method is able to leverage, and that a different method might do
-better.  Indeed, for \svdfive, the LMA method produces a MASE score of
-$ 0.718\pm 0.048 $ and a cluster of results that is much closer to the
-dashed line on the WPE-MASE plot.  Again, this validates our second
-conjecture: the LMA method can capture and reproduce the way in which
-the \svdfive system processes information, but the \naive and ARIMA
-prediction methods cannot.
+  ~$MASE$ score is large because of the bimodal nature of the
+  distribution of the values of the signal, which makes guessing the
+  mean a particularly bad strategy.  The same thing is true of the
+  \svdthree signal.}.  As before, the positions of these points on a
+WPE vs. $MASE$ plot---significantly below and to the right of the
+dashed line---should suggest to a practitioner that there is more
+structure in this signal than the corresponding method is able to
+leverage, and that a different method might do better.  Indeed, for
+\svdfive, the LMA method produces a $MASE$ score of $ 0.718\pm 0.048 $
+and a cluster of results that is much closer to the dashed line on the
+WPE-$MASE$ plot.  Again, this validates our second conjecture: the LMA
+method can capture and reproduce the way in which the \svdfive system
+processes information, but the \naive ~and ARIMA prediction methods
+cannot.
 
-The WPE of \gcc is much higher: $0.940 \pm 0.001$.  This system
+The WPE of \gcc is higher still: $0.943 \pm 0.001$.  This system
 transmits very little information forward in time and provides almost
-no structure for prediction methods to work with.  {\color{red}[[What we see:
-    \naive does the best, then LMA, then ARIMA.  Hypothesis: that that
-    5\% of the structure is nonlinear, so LMA can use it better than
-    ARIMA can.  \naive does better than LMA because it's filtering out
-    the noise.]]}
+no structure for prediction methods to work with.  Here, the \naive
+~method performs best ($MASE=0.951 \pm 0.001$), simply because its
+calculation of the mean filters out the noise in the signal.  Since
+fitting a hyperplane using least squares should have similar effects,
+the fact that LMA outperforms ARIMA ($1.530 \pm 0.021$ vs. $1.837 \pm
+0.016$) is somewhat counterintuitive.  However, the small amount of
+predictive structure that is present in this signal is nonlinear
+(cf.,~\cite{mytkowicz09}), and LMA is designed to capture and exploit
+that kind of structure.
 
 \svdone behaves very differently than the other seven systems in this
-study: though its WPE is very high ($0.95 \pm 0.02$), two of the three
-prediction methods do quite well, yielding MASE scores of $0.82 \pm
-0.08$ (LMA) and $0.714 \pm 0.07$ (ARIMA).  This pushes the
-corresponding clusters of points in the bottom two images in
-Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
-other seven clusters.  Also, the MASE scores of the predictions of
-this system produced by the \naive method are highly inconsistent:
-$2.67 \pm 4.33$.  These effects, we believe, are artifacts of the way
-MASE is calculated.  Recall that MASE scores are scaled relative to a
-random-walk forecast, which uses the current value as the prediction.
-This strategy works very badly on signals with frequent, large, rapid
-transitions.  Consider a signal that oscillates from one end of its
-range to the other at every step.  A signal like this will have a low
-WPE, much like {\tt col\_major}.  However, a random-walk forecast of
-this signal will be 180 degrees out of phase with the true
+study.  Though its weighted permutation entropy is very high ($0.957
+\pm 0.016$), two of the three prediction methods do quite well,
+yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
+0.075$ (ARIMA).  This pushes the corresponding clusters of points in
+the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
+the trend followed by the other seven clusters.  The $MASE$ scores of
+the predictions of this system produced by the \naive ~method,
+however, are highly inconsistent: $2.676 \pm 4.328$.  The majority of
+the blue diamond-shaped points on the top plot in the figure are
+clustered near a $MASE$ score of 0.6, but three of the 15 \svdone
+trials posed challenges to this prediction method, resulting in $MASE$
+scores $\approx 2.5$.
+
+These effects, we believe, are artifacts of the way $MASE$ is
+calculated, as well as the nonstationarity properties of the \svdone
+signal.  Recall that $MASE$ scores are scaled \emph{relative to a
+  random-walk forecast}, which uses the current value as the
+prediction.  This strategy works very badly on signals with frequent,
+large, rapid transitions.  Consider a signal that oscillates from one
+end of its range to the other at every step.  A signal like this will
+have a low WPE, much like {\tt col\_major}.  However, a random-walk
+forecast of this signal will be 180 degrees out of phase with the true
 continuation.  Since random-walk error appears in the denominator of
-the MASE score, this effect can shift points leftwards on a WPE
-vs. MASE plot, and that is exactly why the \svdone clusters for ARIMA
-and LMA are above the dashed line in Figure~\ref{fig:wpe_vs_mase_all}.
-This time series, which is shown in closeup in
-Figure~\ref{fig:svdone-ts},
+the $MASE$ score, this effect can shift points leftwards on a WPE
+vs. $MASE$ plot, and that is exactly why the \svdone clusters for
+ARIMA and LMA are above the dashed line in
+Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
+closeup in Figure~\ref{fig:svdone-ts},
 \begin{figure}[htbp]
@@
 40\% of the forecasted points are off by the width of the associated
-dip, which skews the associated MASE scores.  Signals like this are
-also problematic for the \naive prediction strategy, since the
+dip, which skews the associated $MASE$ scores.  Signals like this are
+also problematic for the \naive ~prediction strategy, since the
 outliers have significant influence on the mean.  This compounds the
 effect of the skew in the scaling factor and creates a large spread in
-the \svdone MASE values.  For all of these reasons, we view \svdone as
-an outlier and exclude it from the fit calculation of the dashed line
-in Figure~\ref{fig:wpe_vs_mase_best}.
+the \svdone $MASE$ values.  {\color{red} We need to think about the
+  three outliers and put in a sentence here about them.}  For all of
+these reasons, we view \svdone as an outlier and exclude it from the
+fit calculation of the dashed line in
+Figure~\ref{fig:wpe_vs_mase_best}.
 
@@
 %is in large part one of the major findings of this work.
-More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$).
+More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MAS>1$).
 How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict.

### 2014-04-11 intro.tex
edit through the model section

@@
 is being transmitted from past to future by the underlying generating
-process?  A first step in answering this question is to reliably
+process \alert{in order to predict it}?  A first step in answering this question is to reliably
 quantify where on the complexity spectrum a given time series falls; a
@@
 
-We focus on real-valued, scalar, time-series data that appear
+We focus on real-valued, scalar, time-series data that appear\alert{s}
 complicated: i.e., not periodic, linear, constant, etc.  We make no
@@
 To explore the relationship between complexity, predictive structure,
-and predictability, we generate forecasts of a variety of time-series
+and predictability, we generate forecasts for a variety of time-series
 datasets using four different prediction methods, then compare the
@@
 forecast method.  It would also be useful to know if a given method is
-inadequate---that is, if one could do better and should keep
-searching.
+inadequate---that is, if an alternative method could do better and so 
+one should continue searching.
 
@@
 experiments.  While this is not a common laboratory experiment, it is
-a highly appropriate choice here.  Computers are extremely complicated
+a highly appropriate choice.  Computers are extremely complicated
 systems.  Modern microprocessor chips contain multiple processing
@@
 complexity spectrum, from completely predictable to completely
-unstructured, making this an ideal testbed for this
+unstructured, making them an ideal testbed for this
 study\footnote{Predicting the \emph{state} of a computer, of course,

### 2014-04-11 pecomplexity.tex
edit through the model section

@@
 \usepackage{graphicx}
+\usepackage{hyperref}
 \usepackage{caption}
@@
 \newcommand{\col}{{\tt col\_major}\xspace}
+\newcommand{\alert}[1]{{\color{red}#1}}
 %\newcommand{\color{red}}{{\color[rgb]{1,0,0}}}
@@
 \begin{abstract}
-{\color{green}From the site:It should be about 5\% of the length of the article, but less than 500 words.} This paper provides insight
+{\color{green}From the site: It should be about 5\% of the length of the article, but less than 500 words.}
+
+This paper provides insight
 into when, why, and how forecast strategies fail when they are applied

### 2014-04-11 related.tex
edit through the model section

@@
 solutions fall into two categories: model-based error analysis and
-model-free information analysis.
+model-free information analysis \alert{citation?}.
 %
@@
 We sidestep these issues by using a variant of the \emph{permutation
-  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate a value
-for the Kolmogorov-Sinai entropy of a real-valued time series---and
+  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate the value
+of the Kolmogorov-Sinai entropy of a real-valued time series---and
 thus the redundancy in that data, which is an effective proxy for

### 2014-04-11 model.tex
more messing w/table
edit model section

@@
 
-\item The \emph{ARIMA} method, a common linear forecast strategy, and
+\item The \emph{ARIMA} (auto-regressive integrated moving average)
+  method, a common linear forecast strategy, and
 
-\item The \emph{LMA} method, which uses a near-neighbor forecast
-  strategy on a dynamical reconstruction of the signal.
+\item The \emph{LMA} (Lorenz method of analogues) method, which uses a
+  near-neighbor forecast strategy on a dynamical reconstruction of the
+  signal.
 \end{itemize}
@@
 linear fit is inappropriate and ARIMA forecasts can be inaccurate.
-Another weakness of this method is prediction horizon, since an ARIMA
+Another weakness of this method is prediction horizon: an ARIMA
 forecast is guaranteed to converge to the mean after some number of
@@
   \begin{center}
-  \begin{tabular}{|c|c|c|l|c|}
+  \begin{tabular}{|c|c|c|c|c|}
   \hline Signal & LMA  & ARIMA & na\"{i}ve  & Weighted Permutation \\ 
@@
 the other hand, so the denominator is small and the $MASE$ scores are
-skewed upwards.  Normalization is, as is often the case, a two-edged
+skewed upwards.  Normalization is, as is often the case, a double-edged
 sword.  Here, it facilitates comparisons across signals of different

### 2014-04-11 results.tex
adding some notes

@@
 The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
-  col\_major}.  This indicates that the rate of forward information
+  col\_major}.  This indicates that the rate{\color{red} is this really a rate} of forward information
 transfer of the underlying process is lower, but that time-series data
@@
 trials posed challenges to this prediction method, resulting in $MASE$
-scores $\approx 2.5$.
+scores $\approx 2.5$. {\color{red} this is becasue of the weird split it does some of the time. Should this be said?}
 
@@
 but it still poses a serious challenge to random-walk prediction.  It
-is dominated by a white-noise regime (between $\approx$1.86 and
+is dominated by a white-noise{\color{red} can we actually say white-noise here} regime (between $\approx$1.86 and
 $\approx$1.88 on the vertical scale in the Figure), punctuated by
@@
 the \svdone $MASE$ values.  {\color{red} We need to think about the
-  three outliers and put in a sentence here about them.}  For all of
+  three outliers and put in a sentence here about them., these are the dip ones}  For all of
 these reasons, we view \svdone as an outlier and exclude it from the

### 2014-04-11 wpe.tex
clean up haven paragraph, use different WPE vs MASE image

@@
 between the two research projects.  Haven {\sl et al.} utilize the
-relative entropy to quantify how much better a distribution over
-system states evolved from small ensembles to predict the future of a
-dynamical system compared to its asymptotic distribution. Our work
+relative entropy to quantify how much better one can predict the exact
+state of a system from a distribution over system states evolved from 
+a small ensemble around a prior known exact state, verses its asymptotic
+distribution, requiring a model to perform the evolution. Our work
 quantifies the predictability of a single observed time series using
 weighted permutation entropy and makes no assumptions about the
-underlying dynamics. {\color{red} These approaches are very different:
-  Haven's compares two distributions, the average behavior versus the
-  behaviors that can be seen given a reasonably known state in the
-  past; ours attempts to quantify how well an appropriate predictor
-  will do for a given time-series without any understanding of how
-  that time-series was generated. [[Seems to me like the first part of
-      this paragraph made the point.  Is this second set of stuff
-      needed?]]}
+underlying dynamics.
 

### 2014-04-11 intro.tex
red text edits

@@
 question: given a complicated real-valued time series, does there
-exist a forecast model that can utilize the information (if any) that
+exist a forecast model that can leverage the information (if any) that
 is being transmitted from past to future by the underlying generating
-process \alert{in order to predict it}?  A first step in answering this question is to reliably
+process?  A first step in answering this question is to reliably
 quantify where on the complexity spectrum a given time series falls; a

### 2014-04-11 methods.tex
edit through the model section
red text edits

@@
 
-\col is a simple three-line C program that repeatedly initializes the
-upper triangle of a 2048 $\times$ 2048 matrix in column-major order.
-As mentioned in Section~\ref{sec:intro}, this simple program exhibits
-surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
-of the processor performance during the execution this program is
-shown in Figure~\ref{fig:col-ts}.
+\col is a simple C program that repeatedly initializes the upper
+triangle of a 2048 $\times$ 2048 matrix in column-major order.
+
+{\color{red}Joshua, can you please insert code here?}
+
+\noindent As mentioned in Section~\ref{sec:intro}, this simple program
+exhibits surprisingly complicated behavior~\cite{mytkowicz09}.  A time
+series of the processor performance during the execution this program
+is shown in Figure~\ref{fig:col-ts}. \alert{would it be useful to
+  include said 3-line loop?}
 \begin{figure}[htbp]

### 2014-04-11 results.tex
explanation of col arima cluster spread

@@
 %    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
-  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy}
+  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
 \caption{WPE vs. $MASE$ for all trials, methods, and systems. \svdone,
   \svdthree, and \svdfive are omitted from the top plot for scale
-  reasons. {\color{red} All three plots need to have the same
-    horizontal scale!}
+  reasons.
 % 
@@
 in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
-method.
+method.  The position of the LMA cluster, for instance---well to the
+left of the other two, near the dashed line---reflects this method's
+ability to capture and exploit the structure that is present in this
+signal.
 
@@
 very tightly in the lower left quadrant of the LMA and \naive ~$MASE$
-vs. WPE plots, but spread out horizontally in the ARIMA plot.
-{\color{red}[[We need to come up with a hypothesis about this spread,
-      even if it's totally tentative.  Was it that the random-walk got
-      really good in a few of the \col trials?  Or that ARIMA got
-      really bad in those trials?  In either case, why did that
-      happen---what was it about those trials that made ARIMA bark
-      and/or random walk do well?]]} The position of the LMA
-cluster---well to the left of the other two, near the dashed
-line---reflects this method's ability to capture and exploit the
-structure that is present in this signal.
+vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
+because of the way that these models are built.  If a KPSS test of the
+time series indicates that it is nonstationary, an integration term is
+added to the model.  According to this test, the \col time series is
+right on the edge of nonstationarity, so the ARIMA procedure sometimes
+omits the integration term, which increases the error, thereby
+spreading out the points.
 
 The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
-  col\_major}.  This indicates that the rate{\color{red} is this really a rate} of forward information
-transfer of the underlying process is lower, but that time-series data
-from this system still contain a significant amount of structure that
-can, in theory, be leveraged to predict the future course of the time
-series.  As mentioned above, though, the effectiveness of any
-prediction strategy depends on how well it leverages the available
-information: methods that use linear models, for instance, may fail
-when applied to nonlinear processes.
+  col\_major}.  This indicates that the rate{\color{red} is this
+  really a rate} of forward information transfer of the underlying
+process is lower, but that time-series data from this system still
+contain a significant amount of structure that can, in theory, be
+leveraged to predict the future course of the time series.  As
+mentioned above, though, the effectiveness of any prediction strategy
+depends on how well it leverages the available information: methods
+that use linear models, for instance, may fail when applied to
+nonlinear processes.
 % 

### 2014-04-11 model.tex
filled in the stuff about geometry of p vs c

@@
     \caption{\col\\ random walk }
-    \label{fig:gccMEAN}
+    \label{fig:colRW}
   \end{subfigure}%
@@
     \caption{\gcc\\ random walk }
-    \label{fig:gccMEAN}
+    \label{fig:gccRW}
   \end{subfigure}% 
@@
     \caption{\svdfive\\ random walk}
-    \label{fig:gccMEAN}
+    \label{fig:svd5RW}
   \end{subfigure}%      
@@
     \caption{\col\\ na\"ive }
-    \label{fig:gccMEAN}
+    \label{fig:colMEAN}
   \end{subfigure}%
@@
     \caption{\svdfive\\ na\"ive }
-    \label{fig:gccMEAN}
+    \label{fig:svd5MEAN}
   \end{subfigure}%      
@@
     \includegraphics[width=\columnwidth]{figs/colARIMAForecast.png}
-
     \caption{\col \\ ARIMA}
@@
     \caption{\svdfive \\ ARIMA }
-    \label{fig:gccARIMA}
+    \label{fig:svd5ARIMA}
   \end{subfigure}%    
@@
 structure of the distribution of the errors---roughly normally
-distributed, for example, in the case of LMA on \gcc. The triangular 
-shapes that ARIMA and random-walk produce for \col simply reflects 
-three phases of values that is visually present in the time-series.{\color{red}not happy with this explanation, I(joshua) will think about how to word this}
+distributed, for example, in the case of LMA on \gcc.  The shapes of
+Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are reminiscent of the
+projected embedding in Figure~\ref{fig:embedding}(a).  For a
+random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
+two-dimensional embedding with $\tau=1$.  For ARIMA, the
+correspondence is not quite as simple, since the $p_j$ values are
+linear combinations of a number of past values of the $c_j$, but the
+effect is largely the same\footnote{The temporal ordering of the
+  points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of a
+  projected embedding of the time series, however.}.
 

### 2014-04-12 results.tex
fixed up results.tex

--- 2014-04-05 results.tex
@@
 %\begin{itemize}
-%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
+%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.
 %\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. 
-[[talk about again here but justify in information theory section]]
 %\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.
@@
 And a constant, linear and nonlinear prediction strategy all fail.
-We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]
+We should be able to conclude that guessing random values is the best we can do as is shown by MASE
 
-%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]
+%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction
 
@@
 This gives a practitioner leverage on when to give up and when to keep working.
-[[use in this section as bridge to next section]]
 
@@
 
-%\subitem [[Simply reminders not to be included]]WPE can quantify when
-%a noisy real-valued time series is predictable.[[I am unsatisfied
+%\subitem Simply reminders not to be included]]WPE can quantify when
+%a noisy real-valued time series is predictable.I am unsatisfied
 %with predictable in this sentence, need a better word to say "better
@@
 
-%\subitem [[Simply reminders not to be included]]We will have shown
+%\subitem Simply reminders not to be included]]We will have shown
 %that the existence whether linear or nonlinear is picked up on with
@@
 
-%\subitem [[Simply reminders not to be included]]This is the linear vs
+%\subitem Simply reminders not to be included]]This is the linear vs
 %nonliear vs random we see with the right to left shifts with lower
@@
 %nontrivial in the case of real-valued noisy time series but WPE does
-%this. [[talk about again here but justify in information theory
+%this. talk about again here but justify in information theory
 %section]]
@@
 same as in the previous figure.
-% 
-%  [[All traces except those from \svd$_1$ lie above the line,
-%      indicating that LMA is better suited prediction method for the
-%      computer traces considered.]]  [[This comes {\bf long} before the
-%      associated discussion in the text, and it's not what this figure
-%      is trying to show here.]]
-% 
 }
@@
 in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
-method.
+method.  Conversely, the position of the LMA cluster---well to the
+left of the other two, near the dashed line---reflects this method's
+ability to capture and exploit the structure that is present in this
+signal.
 
@@
 very tightly in the lower left quadrant of the LMA and \naive ~$MASE$
-vs. WPE plots, but spread out horizontally in the ARIMA plot.
-{\color{red}[[We need to come up with a hypothesis about this spread,
-      even if it's totally tentative.  Was it that the random-walk got
-      really good in a few of the \col trials?  Or that ARIMA got
-      really bad in those trials?  In either case, why did that
-      happen---what was it about those trials that made ARIMA bark
-      and/or random walk do well?]]} The position of the LMA
-cluster---well to the left of the other two, near the dashed
-line---reflects this method's ability to capture and exploit the
-structure that is present in this signal.
+vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
+because of the way that these models are built.  If a KPSS test of the
+time series in question indicates that it is nonstationary, an
+integration term is added to the model.  According to this test, the
+\col time series is right on the edge of nonstationarity, so the ARIMA
+procedure sometimes omits the integration term, which increases the
+error, thereby spreading out the points.
 
@@
 study.  Though its weighted permutation entropy is very high ($0.957
-\pm 0.016$), two of the three prediction methods do quite well,
-yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
-0.075$ (ARIMA).  This pushes the corresponding clusters of points in
-the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
-the trend followed by the other seven clusters.  The $MASE$ scores of
-the predictions of this system produced by the \naive ~method,
-however, are highly inconsistent: $2.676 \pm 4.328$.  The majority of
-the blue diamond-shaped points on the top plot in the figure are
-clustered near a $MASE$ score of 0.6, but three of the 15 \svdone
-trials posed challenges to this prediction method, resulting in $MASE$
-scores $\approx 2.5$.
+\pm 0.016$), two of the three prediction methods do quite well on this
+signal, yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714
+\pm 0.075$ (ARIMA).  This pushes the corresponding clusters of points
+in the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well
+above the trend followed by the other seven signals.  The $MASE$
+scores of the predictions that were produced by the \naive ~method for
+this system, however, are highly inconsistent.  The majority of the
+blue diamond-shaped points on the top plot in
+Figure~\ref{fig:wpe_vs_mase_all} are clustered near a $MASE$ score of
+0.6.  In five of the 15 \svdone trials, however, there were step
+changes in the signal.  The \naive ~method has a very difficult time
+with signals like this, particularly if there are multiple step
+changes.  This raised the $MASE$ scores of these trials, pushing the
+corresponding points to the right, and in turn raising both the mean
+and variance of this set of trials.
 
-These effects, we believe, are artifacts of the way $MASE$ is
-calculated, as well as the nonstationarity properties of the \svdone
-signal.  Recall that $MASE$ scores are scaled \emph{relative to a
-  random-walk forecast}, which uses the current value as the
-prediction.  This strategy works very badly on signals with frequent,
-large, rapid transitions.  Consider a signal that oscillates from one
-end of its range to the other at every step.  A signal like this will
-have a low WPE, much like {\tt col\_major}.  However, a random-walk
-forecast of this signal will be 180 degrees out of phase with the true
-continuation.  Since random-walk error appears in the denominator of
-the $MASE$ score, this effect can shift points leftwards on a WPE
-vs. $MASE$ plot, and that is exactly why the \svdone clusters for
-ARIMA and LMA are above the dashed line in
+The effects described in the previous paragraph are exacerbated by the
+way $MASE$ is calculated.  Recall that $MASE$ scores are scaled
+\emph{relative to a random-walk forecast}, which uses the current
+value as the prediction.  This strategy works very badly on signals
+with frequent, large, rapid transitions.  Consider a signal that
+oscillates from one end of its range to the other at every step.  A
+signal like this will have a low WPE, much like {\tt col\_major}.
+However, a random-walk forecast of this signal will be 180 degrees out
+of phase with the true continuation.  Since random-walk error appears
+in the denominator of the $MASE$ score, this effect can shift points
+leftwards on a WPE vs. $MASE$ plot, and that is exactly why the
+\svdone clusters for ARIMA and LMA are above the dashed line in
 Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
@@
 but it still poses a serious challenge to random-walk prediction.  It
-is dominated by a white-noise regime (between $\approx$1.86 and
+is dominated by a noisy regime (between $\approx$1.86 and
 $\approx$1.88 on the vertical scale in the Figure), punctuated by
@@
 outliers have significant influence on the mean.  This compounds the
-effect of the skew in the scaling factor and creates a large spread in
-the \svdone $MASE$ values.  {\color{red} We need to think about the
-  three outliers and put in a sentence here about them.}  For all of
-these reasons, we view \svdone as an outlier and exclude it from the
-fit calculation of the dashed line in
-Figure~\ref{fig:wpe_vs_mase_best}.
+effect of the skew in the scaling factor and exacerbates the spread in
+the \svdone $MASE$ values.  For all of these reasons, we view \svdone
+as an outlier and exclude it from the fit calculation of the dashed
+line in Figure~\ref{fig:wpe_vs_mase_best}.
 

### 2014-04-12 wpe.tex
typo in wpe

@@
 state of a system from a distribution over system states evolved from 
-a small ensemble around a prior known exact state, verses its asymptotic
+a small ensemble around a prior known exact state, versus its asymptotic
 distribution, requiring a model to perform the evolution. Our work

### 2014-04-12 pecomplexity.tex
adding question
red text edits
rewrote abstract

@@
 
-This paper provides insight
-into when, why, and how forecast strategies fail when they are applied
-to complicated time series by empirically estimating their inherent
-complexity. We focus here on the specific example of computer
-performance but believe the results apply to a much broader class of
-time series. We conjecture that the complexity of traces from these
-systems---which results from the inherent dimension, non-linearity,
-and non-stationarity, as well as from measurement issues like noise,
-aggregation, and finite data length---is \emph{empirically
-  quantifiable}.  To validate this conjecture we model 120 different
-performance traces using a variety of forecasting models.  We then
-compare the empirically estimated \emph{complexity} with the accuracy
-of each forecast method.
+This paper provides insight into when, why, and how forecast
+strategies fail when they are applied to complicated time series.  We
+conjecture that the inherent complexity of a time series---which
+results from the inherent dimension, nonlinearity, and
+non-stationarity of the generating process, as well as from
+measurement issues like noise, aggregation, and finite data
+length---is both \emph{empirically quantifiable} and \emph{directly
+  correlated with predictability}.  To validate this conjecture, we
+study 120 different time-series data sets and a wide variety of
+forecast models.  We estimate the complexity of each data set using a
+variant of permutation entropy, then compare that complexity with the
+accuracy of each forecast method.
 

### 2014-04-12 pecomplexity.tex
added sentence to abstract

@@
 \begin{abstract}
-{\color{green}From the site: It should be about 5\% of the length of the article, but less than 500 words.}
 
@@
 variant of permutation entropy, then compare that complexity with the
-accuracy of each forecast method.
+accuracy of each forecast method.  We use the results to develop a
+heuristic that can help practitioners recognize when a particular
+prediction method is not well matched to the task at hand: that is,
+when the time series has more predictive structure than that method
+can capture and exploit.
 

### 2014-04-12 results.tex
language about mase spread

@@
 changes.  This raised the $MASE$ scores of these trials, pushing the
-corresponding points to the right, and in turn raising both the mean
-and variance of this set of trials.
+corresponding points to the right\footnote{This includes the cluster
+  of three points near $MASE \approx 2.5$, as well as two points that
+  are beyond the domain of the graph, at $MASE$ 11.2 and 14.8.}, and
+in turn raising both the mean and variance of this set of trials.
 

### 2014-04-12 results.tex
added JG reasoning about col arima to results

@@
 vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
-because of the way that these models are built.  If a KPSS test of the
-time series in question indicates that it is nonstationary, an
-integration term is added to the model.  According to this test, the
-\col time series is right on the edge of nonstationarity, so the ARIMA
-procedure sometimes omits the integration term, which increases the
-error, thereby spreading out the points.
+because of the way that these models are built \cite{auto-arima}.  If
+a KPSS test of the time series in question indicates that it is
+nonstationary, an integration term is added to the model.  This test
+gives mixed results in the case of the \col process, flagging five of
+the 15 trials as stationary and ten as nonstationary.  ARIMA models
+without an integration term perform more poorly on this signal, which
+increases the error, thereby spreading out the points\footnote{We
+  tested this hypothesis by forcing the inclusion of an integration
+  term in the five cases where a KPSS test indicated that such a term
+  was not needed.  This action removed the spread, pushing all 15 of
+  the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
+  tight cluster.}.
 

### 2014-04-12 pecomplexity.tex
adding code and packages to compile code
abstract redone (again)

@@
 \usepackage{color}
+\usepackage{listings}
+\lstset{language=C}
 
@@
 strategies fail when they are applied to complicated time series. We
-conjecture that the inherent complexity of a time series---which
-results from the inherent dimension, nonlinearity, and
-non-stationarity of the generating process, as well as from
-measurement issues like noise, aggregation, and finite data
-length---is both \emph{empirically quantifiable} and \emph{directly
-  correlated with predictability}.  To validate this conjecture, we
-study 120 different time-series data sets and a wide variety of
-forecast models.  We estimate the complexity of each data set using a
-variant of permutation entropy, then compare that complexity with the
-accuracy of each forecast method.  We use the results to develop a
-heuristic that can help practitioners recognize when a particular
-prediction method is not well matched to the task at hand: that is,
-when the time series has more predictive structure than that method
-can capture and exploit.
+conjecture that the inherent complexity of a time series—which re-
+sults from the inherent dimension, nonlinearity, and non-stationarity
+of the generating process, as well as from measurement issues like
+noise, aggregation, and finite data length—is both empirically
+quantifiable and directly correlated with predictability. To validate
+this conjecture, we study 120 different time-series data sets and a
+wide variety of forecast models. We estimate the complexity of each
+data set using a variant of permutation entropy, then compare that
+complexity with the accu- racy of each forecast method. We use the
+results to develop a heuristic that can help practitioners recognize
+when a particular prediction method is not well matched to the task at
+hand: that is, when the time series has more predictive structure than
+that method can capture and exploit.
 

### 2014-04-12 pecomplexity.tex
fixing abstract, which got trompled by pull/push

@@
 strategies fail when they are applied to complicated time series. We
-conjecture that the inherent complexity of a time series—which re-
-sults from the inherent dimension, nonlinearity, and non-stationarity
-of the generating process, as well as from measurement issues like
-noise, aggregation, and finite data length—is both empirically
-quantifiable and directly correlated with predictability. To validate
-this conjecture, we study 120 different time-series data sets and a
-wide variety of forecast models. We estimate the complexity of each
-data set using a variant of permutation entropy, then compare that
-complexity with the accu- racy of each forecast method. We use the
-results to develop a heuristic that can help practitioners recognize
-when a particular prediction method is not well matched to the task at
-hand: that is, when the time series has more predictive structure than
-that method can capture and exploit.
+conjecture that the inherent complexity of a time series—--which
+results from the inherent dimension, nonlinearity, and
+non-stationarity of the generating process, as well as from
+measurement issues like noise, aggregation, and finite data
+length---is both empirically quantifiable and directly correlated with
+predictability. To validate this conjecture, we study 120 different
+time-series data sets and a wide variety of forecast models. We
+estimate the complexity of each time series using a variant of
+permutation entropy, then compare that complexity with the accuracy of
+each forecast method.  We use the results to develop a heuristic that
+can help practitioners recognize when a particular prediction method
+is not well matched to the task at hand: that is, when the time series
+has more predictive structure than that method can capture and
+exploit.
 

### 2014-04-12 pecomplexity.tex
added notion of redundancy to the abstract

--- 2014-04-12 pecomplexity.tex
@@
 strategies fail when they are applied to complicated time series. We
-conjecture that the inherent complexity of a time series—which re-
-sults from the inherent dimension, nonlinearity, and non-stationarity
-of the generating process, as well as from measurement issues like
-noise, aggregation, and finite data length—is both empirically
-quantifiable and directly correlated with predictability. To validate
-this conjecture, we study 120 different time-series data sets and a
-wide variety of forecast models. We estimate the complexity of each
-data set using a variant of permutation entropy, then compare that
-complexity with the accu- racy of each forecast method. We use the
-results to develop a heuristic that can help practitioners recognize
-when a particular prediction method is not well matched to the task at
-hand: that is, when the time series has more predictive structure than
-that method can capture and exploit.
+conjecture that the inherent complexity of a time series —-- which
+results from the inherent dimension, nonlinearity, and
+non-stationarity of the generating process, as well as from
+measurement issues like noise, aggregation, and finite data length ---
+is both empirically quantifiable and directly correlated with
+predictability. To validate this conjecture, we study 120 different
+time-series data sets.  We argue that \emph{redundancy} is an
+effective way to measure complexity and predictive structure in a time
+series and that \emph{weighted permutation entropy} is an effective
+way to estimate that complexity.  For each time series, we construct
+predictions using a wide variety of forecast models, then compare the
+accuracy of the predictions with the permutation entropy of that time
+series.  We use the results to develop a heuristic that can help
+practitioners recognize when a particular prediction method is not
+well matched to the task at hand: that is, when the time series has
+more predictive structure than that method can capture and exploit.
 

### 2014-04-12 concl.tex
note to concl
synching everything in case I forgot something

--- 2014-02-19 concl.tex
@@
-\section{ Conclusions \& Future Work}\label{sec:conc}
+\section{ Conclusions \& Future Work }\label{sec:conc}
 
-\begin{enumerate}
-\item  This need to be rewritten to conclude and address this paper...
-\end{enumerate}
+It's obvious that no forecast method can work if a time series has no
+predictive structure, but it's hard to define or measure predictive
+structure.  In this paper, we have argued that redundancy---insert
+short definition here---is a useful definition of predictive structure
+in a time series, and that weighted permutation entropy is an
+effective way to estimate how much of that structure is present in a
+given real-valued, scalar time series.  Even if predictive structure
+exists, a given forecast method may not work---simply because it
+cannot capture and/or exploit the structure that is present (e.g., a
+linear model of a nonlinear process).  This fairly obvious point can
+be a serious practical challenge.  When a model isn't working, the
+practitioner doesn't know whether it's because there's no predictive
+structure (i.e., no model will work) or whether the model simply isn't
+good enough.  Fortunately, weighted permutation entropy allows us to
+make this distinction for noisy real-valued time series.
 
-The results presented here suggest that permutation entropy---a ordinal
-calculation of forward information transfer in a time series---is an effective
-metric for predictability of computer performance traces. Experimentally, traces
-with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
-may overshadow the inherent determinism in the system dynamics, whereas traces
-with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
-of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
-0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
-corroborating the results of~\cite{mytkowicz09}.
+In particular, we wish to determine---without knowing anything about
+the generating process---whether a time series is too complex to
+successfully predict.  We have argued that
+redundancy~\cite{crutchfield2003} is a useful definition of the
+inherent predictability of a time series.  To operationalize that
+definition, we use an approximation of the Kolmogorov-Sinai
+entropy~\cite{lind95}, estimated using a weighted version of the
+permutation entropy of~\cite{bandt2002per}.  This WPE technique---an
+ordinal calculation of forward information transfer in a time
+series---is ideal for our purposes because it works with real-valued
+data and is known to converge to the true entropy value.  Using a
+variety of forecast models and more than a hundred time-series data
+sets from a computer-performance experiment, we have shown that
+prediction accuracy is indeed correlated with weighted permutation
+entropy: the higher the WPE, in general, the higher the prediction
+error.  There are some exceptions: summarize \& discuss here.
 
-If information is the limit, then gathering and using more information is an
-obvious next step.  There is an equally obvious tension here between data length
-and prediction speed: a forecast that requires half a second to compute is not
-useful for the purposes of real-time control of a computer system with a MHz
-clock rate.  Another alternative is to sample several system variables
-simultaneously and build multivariate delay-coordinate embeddings.  Existing
-approaches to that are computationally prohibitive
-\cite{cao-multivariate-embedding}.  We are working on alternative
-methods that sidestep that complexity.
+% Further, the persistent WPE values of 0.5--0.6 for the {\tt
+%   col\_major} trace are consistent with dynamical chaos, further
+% corroborating the results of~\cite{mytkowicz09}.
 
-%%This use of permutaiotn entropy is a useful application of information
-%%theory to computer performance modeling because it gives you a simple
-%%and fast way to decide whether building a model and trying to forecast
-%%the future is worthwhile or if guessing the mean is just as effective.
-%%End with a sentence tying back to power management and world peace.
+An important practical corollary to this empirical correlation of
+predictability and WPE is a strategy for assessing appropriateness of
+forecast methods.  If the forecast produced by a particular method is
+poor but the time series contains a significant amount of predictive
+structure, one can reasonably conclude that that method is inadequate
+to the task and that one should seek another method.  The nonlinear
+LMA method, for instance, performs better in most cases because it is
+more general.  The \naive ~method, which simply predicts the mean,
+works better in noisy signals because it's a filter.  The dashed line
+in Figures blah is a useful empirical guideline for knowing when a
+model is not well-matched to the task at hand: predictions whose MASE
+scores are more than foo times their WPE are an indication that the
+time series has more predictive structure than that method can capture
+and exploit.
 
-%\begin{it}
-%Summarize the results.
+If information is the limit, then gathering and using more information
+is an obvious next step.  There is an equally obvious tension here
+between data length and prediction speed: a forecast that requires
+half a second to compute is not useful for the purposes of real-time
+control of a computer system with a MHz clock rate.  Another
+alternative is to sample several system variables simultaneously and
+build multivariate delay-coordinate embeddings.  Existing approaches
+to that (e.g., \cite{cao-multivariate-embedding,deyle-sugihara2011})
+are computationally prohibitive, however.  We are working on
+alternative methods that sidestep that complexity.
 
@@
 %\end{it}
+
+Nonstationarity is a serious challenge in any time-series modeling
+problem.  One of the most common uses of permutation entropy is in
+regime-shift detection: e.g., to detect seizures in brainwave data
+\cite{cao2004det}.  The SVD signal in Figure~\ref{fig:what} was useful
+here because it contains a number of different regimes.  We segmented
+this signal visually, but one could imagine using permutation entropy
+to do so instead.  This would be a win because it would be automatic,
+which would move towards an adaptive modeling strategy, wherein one
+would toss the old model at the regime boundary, measure WPE, then use
+the result to choose which kind of model to build for the next regime.
+More here from Ryan and Joshua.
+
+Experiments here are from one system, but should work on others.  Give
+some examples.  Close with mention of what that could enable (viz.,
+adaptive models like ones at end of prev paragraph used as the core of
+controllers).
+
+% The way structure, information and complexity are processed by a
+% generating process play a crucial role in the success of a given
+% prediction scheme.

### 2014-04-12 concl.tex
first draft of conclusion

@@
 
-It's obvious that no forecast method can work if a time series has no
-predictive structure, but it's hard to define or measure predictive
-structure.  In this paper, we have argued that redundancy---insert
-short definition here---is a useful definition of predictive structure
-in a time series, and that weighted permutation entropy is an
-effective way to estimate how much of that structure is present in a
-given real-valued, scalar time series.  Even if predictive structure
-exists, a given forecast method may not work---simply because it
-cannot capture and/or exploit the structure that is present (e.g., a
-linear model of a nonlinear process).  This fairly obvious point can
-be a serious practical challenge.  When a model isn't working, the
-practitioner doesn't know whether it's because there's no predictive
-structure (i.e., no model will work) or whether the model simply isn't
-good enough.  Fortunately, weighted permutation entropy allows us to
-make this distinction for noisy real-valued time series.
+No forecast strategy will be effective if a time series has no
+predictive structure.  In practice, this almost-tautological statement
+poses real problems; without knowing anything about the generating
+process, it is difficult to determine how much predictive structure is
+present in a given time series.  And even if predictive structure
+exists, a given forecast method may not work, simply because it cannot
+capture and/or exploit the structure that is present (e.g., a linear
+model of a nonlinear process).  This too can be a serious practical
+challenge.  If a forecast model is not producing good results, a
+practitioner needs to know why: is the reason that the time series
+contains no predictive structure---i.e., that no model will work---or
+is the model that s/he is using simply not good enough?
 
-In particular, we wish to determine---without knowing anything about
-the generating process---whether a time series is too complex to
-successfully predict.  We have argued that
-redundancy~\cite{crutchfield2003} is a useful definition of the
-inherent predictability of a time series.  To operationalize that
-definition, we use an approximation of the Kolmogorov-Sinai
-entropy~\cite{lind95}, estimated using a weighted version of the
-permutation entropy of~\cite{bandt2002per}.  This WPE technique---an
-ordinal calculation of forward information transfer in a time
-series---is ideal for our purposes because it works with real-valued
-data and is known to converge to the true entropy value.  Using a
-variety of forecast models and more than a hundred time-series data
-sets from a computer-performance experiment, we have shown that
+In this paper, we have argued that redundancy~\cite{crutchfield2003}
+is a useful definition of the inherent predictability of a time
+series.  To operationalize that definition, we use an approximation of
+the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
+version of the permutation entropy of~\cite{bandt2002per}.  This WPE
+technique---an ordinal calculation of forward information transfer in
+a time series---is ideal for our purposes because it works with
+real-valued data and is known to converge to the true entropy value.
+Using a variety of forecast models and more than a hundred time-series
+data sets from a computer-performance experiment, we have shown that
 prediction accuracy is indeed correlated with weighted permutation
 entropy: the higher the WPE, in general, the higher the prediction
-error.  There are some exceptions: summarize \& discuss here.
+error.  There are some exceptions: [[Liz to do: summarize \& discuss
+    here.]]
 
@@
 An important practical corollary to this empirical correlation of
-predictability and WPE is a strategy for assessing appropriateness of
-forecast methods.  If the forecast produced by a particular method is
-poor but the time series contains a significant amount of predictive
-structure, one can reasonably conclude that that method is inadequate
-to the task and that one should seek another method.  The nonlinear
-LMA method, for instance, performs better in most cases because it is
-more general.  The \naive ~method, which simply predicts the mean,
-works better in noisy signals because it's a filter.  The dashed line
-in Figures blah is a useful empirical guideline for knowing when a
-model is not well-matched to the task at hand: predictions whose MASE
-scores are more than foo times their WPE are an indication that the
-time series has more predictive structure than that method can capture
-and exploit.
+predictability and WPE is a practical strategy for assessing
+appropriateness of forecast methods.  If the forecast produced by a
+particular method is poor but the time series contains a significant
+amount of predictive structure, one can reasonably conclude that that
+method is inadequate to the task and that one should seek another
+method.  The nonlinear LMA method, for instance, performs better in
+most cases because it is more general.  The \naive ~method, which
+simply predicts the mean, works better in noisy signals because it
+effects a filtering operation.  The dashed line in
+Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all} is a
+useful empirical guideline for knowing when a model is not
+well-matched to the task at hand: a $MASE$ score that is more than
+{\color{red}[[what's the equation for that line, Ryan?]]} times the
+WPE of the corresponding signal is an indication that that time series
+has more predictive structure than that forecast model can capture and
+exploit.
 
-If information is the limit, then gathering and using more information
-is an obvious next step.  There is an equally obvious tension here
-between data length and prediction speed: a forecast that requires
-half a second to compute is not useful for the purposes of real-time
-control of a computer system with a MHz clock rate.  Another
-alternative is to sample several system variables simultaneously and
-build multivariate delay-coordinate embeddings.  Existing approaches
-to that (e.g., \cite{cao-multivariate-embedding,deyle-sugihara2011})
-are computationally prohibitive, however.  We are working on
-alternative methods that sidestep that complexity.
+Given that information is a fundamental limitation in predictability,
+then gathering and using more information is an obvious next step.
+But there is an equally obvious tension here between data length and
+prediction speed: a forecast that requires half a second to compute is
+not useful for the purposes of real-time control of a computer system
+with a MHz clock rate.  Another alternative is to sample several
+system variables simultaneously and build multivariate models.  This
+is a particular challenge in nonlinear LMA-type models, since
+multivariate delay-coordinate embedding (e.g.,
+\cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
+computationally prohibitive.  We are working on alternative methods
+that sidestep that complexity.
 
@@
 Nonstationarity is a serious challenge in any time-series modeling
-problem.  One of the most common uses of permutation entropy is in
-regime-shift detection: e.g., to detect seizures in brainwave data
-\cite{cao2004det}.  The SVD signal in Figure~\ref{fig:what} was useful
-here because it contains a number of different regimes.  We segmented
-this signal visually, but one could imagine using permutation entropy
-to do so instead.  This would be a win because it would be automatic,
-which would move towards an adaptive modeling strategy, wherein one
-would toss the old model at the regime boundary, measure WPE, then use
-the result to choose which kind of model to build for the next regime.
-More here from Ryan and Joshua.
+problem.  Indeed, one of the first uses of permutation entropy was
+regime-shift detection for the purposes of recognizing seizure onset
+in brainwave data~\cite{cao2004det}.  Recall that the signal in
+Figure~\ref{fig:svd_ts_colored} was especially useful for the study in
+this paper because it contained a number of different regimes.  We
+segmented this signal visually, but one could imagine using
+permutation entropy to do so instead.  Automating regime-shift
+detection would be an important step towards a fully adaptive modeling
+strategy, where old models are discarded and new ones are rebuilt
+whenever shifts are detected.  WPE would be particularly powerful in
+this scenario, as its value can suggest what kind of model might work
+well in each new regime.
 
-Experiments here are from one system, but should work on others.  Give
-some examples.  Close with mention of what that could enable (viz.,
-adaptive models like ones at end of prev paragraph used as the core of
-controllers).
+All of the experiments described in this paper involve a particular
+physical system.  We chose that system because its behavior spans a
+broad spectrum of classical time series patterns.  We believe that our
+results will generalize far beyond this particular system; proving
+that claim with examples is another important next step.  [[Liz to do:
+    insert some examples.  Close with mention of what that could
+    enable (viz., adaptive models like ones at end of prev paragraph
+    used as the core of controllers).]]
 

### 2014-04-13 concl.tex
notes to liz

@@
 No forecast strategy will be effective if a time series has no
-predictive structure.  In practice, this almost-tautological statement
+predictive structure.{\color{red} no forecast strategy that uses **structure** will be effective, but naive forecast may do just fine.
+Maybe something like ..."Forecast strategies which depend on predictive structure (e.g., LMA or ARIMA) are rendered ineffective when signal complexity outweighs information redundancy."}  In practice, this almost-tautological statement
 poses real problems; without knowing anything about the generating
@@
 simply predicts the mean, works better in noisy signals because it
-effects a filtering operation.  The dashed line in
+effects a filtering operation.{\color{red} might be helpful to say \naive wins on \svdone with the exception to the 5 outliers and that random walk beats everyone on \gcc, this shows the two simple predictions beat everyone for the two highly complex signals but might be too late in the paper to say this, but might be able to say this to reinforce it}   The dashed line in
 Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all} is a
@@
 in brainwave data~\cite{cao2004det}.  Recall that the signal in
-Figure~\ref{fig:svd_ts_colored} was especially useful for the study in
+Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
 this paper because it contained a number of different regimes.  We

### 2014-04-15 methods.tex
minor edits to methods
adding code and packages to compile code
edited the code listing

@@
 \col is a simple C program that repeatedly initializes the upper
-triangle of a 2048 $\times$ 2048 matrix in column-major order.
-
-{\color{red}Joshua, can you please insert code here?}
-
+triangle of a 2048 $\times$ 2048 matrix in column-major order by
+looping over the following three lines of code:
+\begin{lstlisting}
+% #define SIZE (2048)
+% 
+% int main(int argc, char ** argv)
+% {
+%   int i, j, n;
+%   static int data[SIZE][SIZE];
+%   
+%   for(n=0;n<numInits;n++){  
+  for(i=0;i<SIZE;i++)
+  	for(j=i;j<SIZE;j++)
+  		data[j][i] = 0;
+%   }
+%   return 0;
+}
+\end{lstlisting}
 \noindent As mentioned in Section~\ref{sec:intro}, this simple program
 exhibits surprisingly complicated behavior~\cite{mytkowicz09}.  A time
-series of the processor performance during the execution this program
-is shown in Figure~\ref{fig:col-ts}. \alert{would it be useful to
-  include said 3-line loop?}
+series of the processor performance during the execution of this
+program is shown in Figure~\ref{fig:col-ts}.
 \begin{figure}[htbp]

### 2014-04-15 concl.tex
committing everything

--- 2014-04-12 concl.tex
@@
 
-No forecast strategy will be effective if a time series has no
-predictive structure.  In practice, this almost-tautological statement
-poses real problems; without knowing anything about the generating
-process, it is difficult to determine how much predictive structure is
-present in a given time series.  And even if predictive structure
-exists, a given forecast method may not work, simply because it cannot
-capture and/or exploit the structure that is present (e.g., a linear
-model of a nonlinear process).  This too can be a serious practical
-challenge.  If a forecast model is not producing good results, a
+Forecast strategies that are designed to capture predictive structure
+are ineffective when signal complexity outweighs information
+redundancy.  This poses a number of serious challenges in practice.
+First, without knowing anything about the generating process, it is
+difficult to determine how much predictive structure is present in a
+given time series.  And even if predictive structure exists, a given
+forecast method may not work, simply because it cannot exploit the
+structure that is present (e.g., a linear model of a nonlinear
+process).  If a forecast model is not producing good results, a
 practitioner needs to know why: is the reason that the time series
@@
 simply predicts the mean, works better in noisy signals because it
-effects a filtering operation.  The dashed line in
-Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all} is a
-useful empirical guideline for knowing when a model is not
-well-matched to the task at hand: a $MASE$ score that is more than
-{\color{red}[[what's the equation for that line, Ryan?]]} times the
-WPE of the corresponding signal is an indication that that time series
-has more predictive structure than that forecast model can capture and
-exploit.
+effects a filtering operation.  The simple random-walk strategy is
+better than LMA, ARIMA, or the \naive ~method on the \gcc signal,
+which is extremely complex---i.e., extremely low redundancy.
+{\color{red} I suggest we don't mention \svdone here since it's a
+  pretty detailed thing and conclusions are for high-level summaries.
+  It's a good point, though, and I'll leave it in comments in the
+  file.}
+%  \naive ~wins on \svdone with the exception to the 5 outliers.
+The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
+and~\ref{fig:wpe_vs_mase_all} is a useful empirical guideline for
+knowing when a model is not well-matched to the task at hand: a $MASE$
+score that is more than {\color{red}[[what's the equation for that
+      line, Ryan?]]} times the WPE of the corresponding signal is an
+indication that that time series has more predictive structure than
+that forecast model can capture and exploit.
 
@@
 Nonstationarity is a serious challenge in any time-series modeling
-problem.  Indeed, one of the first uses of permutation entropy was
-regime-shift detection for the purposes of recognizing seizure onset
-in brainwave data~\cite{cao2004det}.  Recall that the signal in
-Figure~\ref{fig:svd_ts_colored} was especially useful for the study in
+problem.  Indeed, one of the first applications of permutation entropy
+was regime-shift detection for the purposes of recognizing seizure
+onset in brainwave data~\cite{cao2004det}.  Recall that the signal in
+Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
 this paper because it contained a number of different regimes.  We

### 2014-04-15 concl.tex
some fill-ins on concl.tex

@@
 entropy: the higher the WPE, in general, the higher the prediction
-error.  There are some exceptions: [[Liz to do: summarize \& discuss
-    here.]]
+error.  There are some exceptions: \svdone is easier to predict than
+its WPE value suggests, for instance, using any of the three forecast
+strategies studied here.  Conversely, forecast errors for the \col and
+\svdfive signals are higher than the corresponding WPEs
+suggest---except for the nonlinear LMA predictor.
 
@@
 method.  The nonlinear LMA method, for instance, performs better in
-most cases because it is more general.  The \naive ~method, which
-simply predicts the mean, works better in noisy signals because it
-effects a filtering operation.  The simple random-walk strategy is
-better than LMA, ARIMA, or the \naive ~method on the \gcc signal,
-which is extremely complex---i.e., extremely low redundancy.
-{\color{red} I suggest we don't mention \svdone here since it's a
-  pretty detailed thing and conclusions are for high-level summaries.
-  It's a good point, though, and I'll leave it in comments in the
-  file.}
+most cases because it is more general.  This becomes particularly
+apparent in the \col and \svdfive examples, where the other two
+methods do not perform well.  The \naive ~method, which simply
+predicts the mean, works better in noisy signals because it effects a
+filtering operation.  The simple random-walk strategy outperforms
+nLMA, ARIMA, and the \naive ~method on the \gcc signal, which is
+extremely complex---i.e., extremely low redundancy.  {\color{red} I
+  suggest we don't mention \svdone here since getting into outliers is
+  a pretty detailed thing and conclusions are for high-level
+  summaries.  It's a good point, though, and I'll leave it in comments
+  in the file.}
 %  \naive ~wins on \svdone with the exception to the 5 outliers.

### 2014-04-15 concl.tex
finished draft of conclusion

--- 2014-04-15 concl.tex
@@
 redundancy.  This poses a number of serious challenges in practice.
-First, without knowing anything about the generating process, it is
-difficult to determine how much predictive structure is present in a
-given time series.  And even if predictive structure exists, a given
-forecast method may not work, simply because it cannot exploit the
-structure that is present (e.g., a linear model of a nonlinear
-process).  If a forecast model is not producing good results, a
-practitioner needs to know why: is the reason that the time series
-contains no predictive structure---i.e., that no model will work---or
-is the model that s/he is using simply not good enough?
+Without knowing anything about the generating process, it is difficult
+to determine how much predictive structure is present in a given time
+series.  And even if predictive structure exists, a given forecast
+method may not work, simply because it cannot exploit the structure
+that is present (e.g., a linear model of a nonlinear process).  If a
+forecast model is not producing good results, a practitioner needs to
+know why: is the reason that the time series contains no predictive
+structure---i.e., that no model will work---or is the model that s/he
+is using simply not good enough?
 
@@
 entropy: the higher the WPE, in general, the higher the prediction
-error.  There are some exceptions: [[Liz to do: summarize \& discuss
-    here.]]
+error.  There are some exceptions: \svdone is easier to predict than
+its WPE value suggests using any of the three forecast strategies
+studied here.  Conversely, forecast errors for the \col and \svdfive
+signals are higher than the corresponding WPEs suggest---except for
+the nonlinear LMA predictor.
 
@@
 method.  The nonlinear LMA method, for instance, performs better in
-most cases because it is more general.  The \naive ~method, which
-simply predicts the mean, works better in noisy signals because it
-effects a filtering operation.  The simple random-walk strategy is
-better than LMA, ARIMA, or the \naive ~method on the \gcc signal,
-which is extremely complex---i.e., extremely low redundancy.
-{\color{red} I suggest we don't mention \svdone here since it's a
-  pretty detailed thing and conclusions are for high-level summaries.
-  It's a good point, though, and I'll leave it in comments in the
-  file.}
+most cases because it is more general.  (This is particularly apparent
+in the \col and \svdfive examples.)
+% , where the other two methods do not perform well.)  
+The \naive ~method, which simply predicts the mean, can work very well
+on noisy signals because it effects a filtering operation.  The simple
+random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
+the \gcc signal, which is extremely complex---i.e., extremely low
+redundancy.
 %  \naive ~wins on \svdone with the exception to the 5 outliers.
 The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
-and~\ref{fig:wpe_vs_mase_all} is a useful empirical guideline for
-knowing when a model is not well-matched to the task at hand: a $MASE$
-score that is more than {\color{red}[[what's the equation for that
-      line, Ryan?]]} times the WPE of the corresponding signal is an
-indication that that time series has more predictive structure than
-that forecast model can capture and exploit.
+and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
+paragraph.  It is a useful empirical guideline for knowing when a
+model is not well-matched to the task at hand: a MASE score that is
+more than {\color{red}[[what's the equation for that line, Ryan?]]}
+times the WPE of the corresponding signal is an indication that that
+time series has more predictive structure than that forecast model can
+capture and exploit.
 
@@
 problem.  Indeed, one of the first applications of permutation entropy
-was regime-shift detection for the purposes of recognizing seizure
-onset in brainwave data~\cite{cao2004det}.  Recall that the signal in
+was to recognize the regime shift in brainwave data that occurs when
+someone has a seizure~\cite{cao2004det}.  Recall that the signal in
 Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
@@
 strategy, where old models are discarded and new ones are rebuilt
-whenever shifts are detected.  WPE would be particularly powerful in
-this scenario, as its value can suggest what kind of model might work
-well in each new regime.
+whenever the time series enters a new regime.  WPE would be
+particularly powerful in this scenario, as its value can not only help
+with regime-shift detection, but also suggest what kind of model might
+work well in each new regime.
 
@@
 physical system.  We chose that system because its behavior spans a
-broad spectrum of classical time series patterns.  We believe that our
-results will generalize far beyond this particular system; proving
-that claim with examples is another important next step.  [[Liz to do:
-    insert some examples.  Close with mention of what that could
-    enable (viz., adaptive models like ones at end of prev paragraph
-    used as the core of controllers).]]
+broad spectrum of classical time series patterns.  Since these
+patterns---dynamical chaos, periodicities, noise, etc.--- are generic,
+we believe that our results will generalize beyond this particular
+system.  Testing that claim with different examples is another
+important next step.  Of particular interest would be the class of
+so-called \emph{hybrid systems}: [[insert def]].  A lathe, for
+instance, that has an intermittent instability.  Traffic at an
+internet router, which has characteristic patterns during normal
+operation that shift radically during an attack.  Modeling and
+predicting the behavior of these kinds of systems effectively is quite
+difficult; doing so adaptively and automatically---in the manner that
+is alluded to at the end of the previous paragraph---would be an even
+more important challenge.
 

### 2014-04-15 model.tex
typo in fig labels
finished draft of conclusion

@@
     \caption{\svdfive \\ LMA}
-    \label{fig:gccLMA}
+    \label{fig:svd5LMA}
   \end{subfigure}
@@
 As a numerical measure of prediction accurary, we calculate the mean
-absolute scaled error ($MASE$) between the true and predicted signals:
+absolute scaled error (MASE) between the true and predicted signals:
 % 
@@
 applicable measurement of forecast accuracy without the problems seen
-in the other measurements."  $MASE$ is a normalized measure; the
+in the other measurements."  MASE is a normalized measure; the
 scaling term in the denominator
@@
 
-$MASE$ scores for all 360 experiments are tabulated in the three
+MASE scores for all 360 experiments are tabulated in the three
 middle columns in Table~\ref{tab:error}.
@@
   \hline Signal & LMA  & ARIMA & na\"{i}ve  & Weighted Permutation \\ 
-  &  $MASE$ & $MASE$  & $MASE$ &  Entropy \\ \hline
+  &  MASE & MASE  & MASE &  Entropy \\ \hline
  \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\
@@
  \label{default}
- \caption{$MASE$ scores and weighted permutation entropies for all eight
+ \caption{MASE scores and weighted permutation entropies for all eight
    processes studied in this paper}
@@
 Figure~\ref{fig:forecast-example}, one can see some obvious
-correspondences.  The average LMA $MASE$ score for the \col signals
+correspondences.  The average LMA MASE score for the \col signals
 was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
@@
 the corresponding images in Figure~\ref{fig:forecast-example}.  While
-the correspondence between $MASE$ score and the visual appearance of
+the correspondence between MASE score and the visual appearance of
 these kinds of plots is clear cut in this case, that is not always so.
 The plots of LMA predictions of \col and \svdfive both lie near the
-diagonal, for instance, but the corresponding $MASE$ scores were very
+diagonal, for instance, but the corresponding MASE scores were very
 different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
 1.4 times better than random-walk forecasts of the same signals).
-This is a result of the normalization in the $MASE$ score calculation.
+This is a result of the normalization in the MASE score calculation.
 Recall from Section~\ref{sec:simple} that the random-walk method
@@
 fits this description, so the random-walk error on this signal---the
-denominator of the three $MASE$ scores in the first row of the
+denominator of the three MASE scores in the first row of the
 Table---is pathologically high, which skews those scores down.
 Random-walk prediction is very effective for the \svdfive signal, on
-the other hand, so the denominator is small and the $MASE$ scores are
+the other hand, so the denominator is small and the MASE scores are
 skewed upwards.  Normalization is, as is often the case, a double-edged
@@
 
-% Table \ref{tab:error} provides the $MASE$ distributions
+% Table \ref{tab:error} provides the MASE distributions
 % {\color{red}[[Joshua: Ryan, Is this the right word? we give mean $\pm$

### 2014-04-15 results.tex
typo in autoarima ref
getting autoarima citations to work
finished draft of conclusion

@@
 LMA prediction methods, as described in Section~\ref{sec:accuracy},
-then calculated the $MASE$ value of those predictions.  We also
+then calculated the MASE value of those predictions.  We also
 calculated the WPE of each time series using a wordlength of six, as
@@
 trials and all three methods for each of the eight programs.  The WPE
-is plotted against the corresponding $MASE$ value here in order to bring
+is plotted against the corresponding MASE value here in order to bring
 out the correlation between these two quantities.
@@
   \caption{Weighted permutation entropy vs. mean absolute scaled error
-    ($MASE$) of the best prediction of each system.
+    (MASE) of the best prediction of each system.
 % 
@@
 formal results about a ``best'' predictor for a given time series.
-There may be other methods that produce lower $MASE$ values than those
+There may be other methods that produce lower MASE values than those
 in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
@@
 
-Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. $MASE$ plots for all 360
+Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
 experiments.
@@
   \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
-\caption{WPE vs. $MASE$ for all trials, methods, and systems. \svdone,
+\caption{WPE vs. MASE for all trials, methods, and systems. \svdone,
   \svdthree, and \svdfive are omitted from the top plot for scale
@@
 % 
-% ; their $MASE$ scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
+% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
 %   $20.87 \pm 0.192$ respectively. 
@@
 Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
-much across trials.  For most traces, the variance in $MASE$ scores is
+much across trials.  For most traces, the variance in MASE scores is
 low as well, resulting in small, tight clusters.  In some
 cases---ARIMA predictions of {\tt col\_major}, for instance---the
-$MASE$ variance is larger, which spreads out the clusters
+MASE variance is larger, which spreads out the clusters
 horizontally.  The distribution of the cluster {\sl positions} is
-quite different in the three plots.  The mean $MASE$ scores of
+quite different in the three plots.  The mean MASE scores of
 predictions generated with the nonlinear LMA method are generally
@@
 prediction methods do not perform as well on this signal as the
-nonlinear LMA method.  The $MASE$ scores of the \naive ~and ARIMA
+nonlinear LMA method.  The MASE scores of the \naive ~and ARIMA
 predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
@@
 information transfer in the signal, but predictions with comparatively
-poor $MASE$ scores---is obvious in the geometry of the top two images
+poor MASE scores---is obvious in the geometry of the top two images
 in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
-to the right of the dashed line.  {\sl This indicates that these
-  methods are not leveraging the available information in the signal.}
+to the right of the dashed line.  \emph{This indicates that these
+  methods are not leveraging the available information in the signal}.
 The dynamics of \col may be complicated, but they are not
@@
 (LMA)---rather than a method that simply predicts the running mean
-(\naive) or one that uses a linear model (ARIMA)---the $MASE$ score is
+(\naive) or one that uses a linear model (ARIMA)---the MASE score is
 much better: $0.050 \pm 0.001$.  This prediction is 20 times more
@@
 Note that the \col points (blue {\color{blue}$+$} icons) are clustered
-very tightly in the lower left quadrant of the LMA and \naive ~$MASE$
+very tightly in the lower left quadrant of the LMA and \naive ~MASE
 vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
-because of the way that these models are built \cite{auto-arima}.  If
+because of the way that these models are built \cite{autoARIMA}.  If
 a KPSS test of the time series in question indicates that it is
@@
 This match between model and task is an issue in the \svdfive
-experiments as well.  The $MASE$ scores of the \naive ~and ARIMA
+experiments as well.  The MASE scores of the \naive ~and ARIMA
 predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
@@
 simple random walk forecast of the same signals\footnote{The \naive
-  ~$MASE$ score is large because of the bimodal nature of the
+  ~MASE score is large because of the bimodal nature of the
   distribution of the values of the signal, which makes guessing the
@@
   \svdthree signal.}.  As before, the positions of these points on a
-WPE vs. $MASE$ plot---significantly below and to the right of the
+WPE vs. MASE plot---significantly below and to the right of the
 dashed line---should suggest to a practitioner that there is more
@@
 leverage, and that a different method might do better.  Indeed, for
-\svdfive, the LMA method produces a $MASE$ score of $ 0.718\pm 0.048 $
+\svdfive, the LMA method produces a MASE score of $ 0.718\pm 0.048 $
 and a cluster of results that is much closer to the dashed line on the
-WPE-$MASE$ plot.  Again, this validates our second conjecture: the LMA
+WPE-MASE plot.  Again, this validates our second conjecture: the LMA
 method can capture and reproduce the way in which the \svdfive system
@@
 \pm 0.016$), two of the three prediction methods do quite well on this
-signal, yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714
-\pm 0.075$ (ARIMA).  This pushes the corresponding clusters of points
-in the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well
-above the trend followed by the other seven signals.  The $MASE$
-scores of the predictions that were produced by the \naive ~method for
-this system, however, are highly inconsistent.  The majority of the
-blue diamond-shaped points on the top plot in
-Figure~\ref{fig:wpe_vs_mase_all} are clustered near a $MASE$ score of
+signal, yielding MASE scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
+0.075$ (ARIMA).  This pushes the corresponding clusters of points in
+the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
+the trend followed by the other seven signals.  The MASE scores of the
+predictions that were produced by the \naive ~method for this system,
+however, are highly inconsistent.  The majority of the blue
+diamond-shaped points on the top plot in
+Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
 0.6.  In five of the 15 \svdone trials, however, there were step
@@
 with signals like this, particularly if there are multiple step
-changes.  This raised the $MASE$ scores of these trials, pushing the
+changes.  This raised the MASE scores of these trials, pushing the
 corresponding points to the right\footnote{This includes the cluster
-  of three points near $MASE \approx 2.5$, as well as two points that
-  are beyond the domain of the graph, at $MASE$ 11.2 and 14.8.}, and
-in turn raising both the mean and variance of this set of trials.
+  of three points near MASE $\approx 2.5$, as well as two points that
+  are beyond the domain of the graph, at MASE 11.2 and 14.8.}, and in
+turn raising both the mean and variance of this set of trials.
 
 The effects described in the previous paragraph are exacerbated by the
-way $MASE$ is calculated.  Recall that $MASE$ scores are scaled
+way MASE is calculated.  Recall that MASE scores are scaled
 \emph{relative to a random-walk forecast}, which uses the current
@@
 of phase with the true continuation.  Since random-walk error appears
-in the denominator of the $MASE$ score, this effect can shift points
-leftwards on a WPE vs. $MASE$ plot, and that is exactly why the
+in the denominator of the MASE score, this effect can shift points
+leftwards on a WPE vs. MASE plot, and that is exactly why the
 \svdone clusters for ARIMA and LMA are above the dashed line in
@@
 40\% of the forecasted points are off by the width of the associated
-dip, which skews the associated $MASE$ scores.  Signals like this are
+dip, which skews the associated MASE scores.  Signals like this are
 also problematic for the \naive ~prediction strategy, since the
@@
 effect of the skew in the scaling factor and exacerbates the spread in
-the \svdone $MASE$ values.  For all of these reasons, we view \svdone
+the \svdone MASE values.  For all of these reasons, we view \svdone
 as an outlier and exclude it from the fit calculation of the dashed

### 2014-04-20 concl.tex
add equation

@@
 model is not well-matched to the task at hand: a MASE score that is
-more than {\color{red}[[what's the equation for that line, Ryan?]]}
+more than $WPE = 0.2794 \cdot MASE + 0.5334$
 times the WPE of the corresponding signal is an indication that that

### 2014-04-21 concl.tex
clean it up a bit

@@
 model is not well-matched to the task at hand: a MASE score that is
-more than $WPE = 0.2794 \cdot MASE + 0.5334$
-times the WPE of the corresponding signal is an indication that that
-time series has more predictive structure than that forecast model can
-capture and exploit.
+more than $3.5788 \cdot \mathrm{WPE} - 1.909$, where WPE is that of
+the corresponding signal, is an indication that that time series has
+more predictive structure than that forecast model can capture and
+exploit.
 

### 2014-04-21 results.tex
some changes to wpe, results sections

@@
 The dynamics of \col may be complicated, but they are not
-unstructured.  This signal is nonlinear and deterministic, and if one
+unstructured.  This signal is nonlinear and \alert{largely} deterministic, and if one
 uses a prediction technique that is based a nonlinear model

### 2014-04-21 wpe.tex
some changes to wpe, results sections

@@
 not useful in our context, where we want to measure the entropy to
-quantify predictability.
+quantify predictability of an arbitrary time-series.
 
@@
 the fact that the signal is actually quite structured.  In this
-situation, the weighted permutation entropy measures on switching
-rather than the stochastic fluctuations. To accomplish this, the
-\emph{weight} of a permutation is taken into account:
+situation, the terms of the weighted permutation entropy are dominated
+by the switching rather than the stochastic fluctuations. To 
+accomplish this, the \emph{weight} of a permutation is taken into
+account:
 \begin{align*}

### 2014-04-22 methods.tex
more editing to code
cleaning up after meeting

@@
 the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
-eight processing units, a clock rate of 3.40GHz, and a cache size of
+eight processing units, a clock rate of 3.40 GHz, and a cache size of
 8192 KB.  We gathered performance traces during the execution of three
@@
 \begin{lstlisting}
+  for(i=0;i<2048;i++)
+  	for(j=i;j<2048;j++)
+  		data[j][i] = 0;
+\end{lstlisting}
+% full listing
 % #define SIZE (2048)
@@
 %   for(n=0;n<numInits;n++){  
-  for(i=0;i<SIZE;i++)
-  	for(j=i;j<SIZE;j++)
-  		data[j][i] = 0;
+%   for(i=0;i<SIZE;i++)
+%   	for(j=i;j<SIZE;j++)
+%   		data[j][i] = 0;
 %   }
 %   return 0;
-}
-\end{lstlisting}
+% }
+%
 \noindent As mentioned in Section~\ref{sec:intro}, this simple program
@@
 % 
-It is a \emph{compiler}: a program that translates code written in the
+It is a \emph{C-compiler}: a program that translates code written in the
 C language into a lower-level format that can be executed by the

### 2014-04-22 model.tex
changed table caption
cleaning up after meeting

@@
 
-\item The \emph{naive} method, which uses the mean of the
+\item The \emph{\naive} method, which uses the mean of the
   observed signal as the forecast,
@@
 A simple and yet powerful way to capture and exploit the structure of
-data is to fit a hyperplane to the dataset and then use that it to
+data is to fit a hyperplane to the dataset and then use that to
 make predictions.  The roots of this approach date back to the
@@
 test~\cite{Canova1995}, and the Akaike information
-criterion~\cite{akaike1974}, conditioned on the maximized likelihood
+criterion~\cite{akaike1974}, conditioned on the maximum likelihood
 of the model fitted to the detrended data---to select good values for
@@
 $\tau$ and the dimension $m$.  We follow standard procedures for this,
-choosing the first minimum in the average mutual information as an
+choosing the first minimum in the time-delayed mutual information as an
 estimate of $\tau$~\cite{fraser-swinney} and using the
-false-near-neighbor method of~\cite{KBA92} to estimate $m$.  Plots of
+false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.  Plots of
 the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored},
@@
 the embedded\footnote{Lorenz's original formulation used the full
-  system state space; the method was later extended to embedded
-  dynamics by Kantz~\cite{kantz97}.}
+  system state space; According to \cite{kantz97} this method was first extended to embedded dynamics by Pikovsky ~\cite{pikovsky86-sov}, but more closely related to the prediction work of Sugihara \& May in \cite{sugihara90}}
 space~\cite{lorenz-analogues}.  Even this simple algorithm---which
@@
 is effectively obfuscated, from the standpoint of the LMA method, by
-noise.  By quantifying the balance between normalized redundancy (that
+noise.  By quantifying the balance between normalized (that
 is, predictive structure) and entropy for these real-valued time
@@
 LMA method, which can generate arbitrary-length
-predictions\footnote{although the accuracy of these predictions will
+predictions---although the accuracy of these predictions will
   degrade over prediction length, in the presence of positive Lyapunov
-  exponents}, but the performance of the other three methods used here
+  exponents, but the performance of the other three methods used here
 will degrade severely if the associated models are not periodically
@@
  \label{default}
- \caption{MASE scores and weighted permutation entropies for all eight
+ \caption{Mean
+absolute scaled error (MASE) scores and weighted permutation entropies for all eight
    processes studied in this paper}

### 2014-04-22 related.tex
red text edits
cleaning up after meeting

@@
 predictability,'' is not a new problem.  Most of the corresponding
-solutions fall into two categories: model-based error analysis and
-model-free information analysis \alert{citation?}.
+solutions fall into two categories that we call model-based error
+analysis and model-free information analysis.
 %
@@
 The basic idea is to construct a series of local linear fits,
-beginning with a few points and working up to to a global linear fit
+beginning with a few points and working up to a global linear fit
 that includes all known points, and then analyze how the average
@@
 fit. The shape of such a ``DVS" graph indicates the amounts of determinism
-and stochasticity are present in a time series.
+and stochasticity present in a time series.\alert{Liz Should we say what this has to do with predictability, realized this paragraph may seem unrelated to a casual reader}
 

### 2014-04-22 results.tex
cleaning up after meeting

--- 2014-04-15 results.tex
@@
 The dynamics of \col may be complicated, but they are not
-unstructured.  This signal is nonlinear and deterministic, and if one
+unstructured.  This signal is nonlinear and deterministic\cite{mytkowicz09}, and if one
 uses a prediction technique that is based a nonlinear model

### 2014-04-22 wpe.tex
fix up haven paragraph

@@
 between the two research projects.  Haven {\sl et al.} utilize the
-relative entropy to quantify how much better one can predict the exact
-state of a system from a distribution over system states evolved from 
-a small ensemble around a prior known exact state, versus its asymptotic
-distribution, requiring a model to perform the evolution. Our work
-quantifies the predictability of a single observed time series using
-weighted permutation entropy and makes no assumptions about the
-underlying dynamics.
+relative entropy to quantify the difference in predictability between 
+two distributions: one using the known dynamical system from a small 
+ensamble in the past, and the other the asymptotic distribution. Our 
+work quantifies the predictability of a single observed time series 
+using weighted permutation entropy and makes no assumptions about the
+generating process.
 

### 2014-04-22 concl.tex
Liz is done rewriting

--- 2014-04-20 concl.tex
@@
 signals are higher than the corresponding WPEs suggest---except for
-the nonlinear LMA predictor.
+the nonlinear LMA predictor.  On the whole, though, signals with
+higher weighted permutation entropy are harder to predict.
 
@@
 and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
-paragraph.  It is a useful empirical guideline for knowing when a
-model is not well-matched to the task at hand: a MASE score that is
-more than $WPE = 0.2794 \cdot MASE + 0.5334$
-times the WPE of the corresponding signal is an indication that that
-time series has more predictive structure than that forecast model can
-capture and exploit.
+paragraph.  It is a preliminary, but potentially useful, empirical
+guideline for knowing when a model is not well-matched to the task at
+hand: a MASE score that is more than $3.58 \cdot \mathrm{WPE} - 1.91$,
+where $\mathrm{WPE}$ is that of the corresponding signal, is an
+indication that that time series has more predictive structure than
+that forecast model can capture and exploit.
 
@@
 not useful for the purposes of real-time control of a computer system
-with a MHz clock rate.  Another alternative is to sample several
-system variables simultaneously and build multivariate models.  This
-is a particular challenge in nonlinear LMA-type models, since
+with a MHz clock rate, for instance.  Another alternative is to sample
+several system variables simultaneously and build multivariate models.
+This is a particular challenge in nonlinear LMA-type models, since
 multivariate delay-coordinate embedding (e.g.,
@@
 broad spectrum of classical time series patterns.  Since these
-patterns---dynamical chaos, periodicities, noise, etc.--- are generic,
-we believe that our results will generalize beyond this particular
+patterns (dynamical chaos, periodicities, noise, etc.) are generic, we
+believe that our results will generalize beyond this particular
 system.  Testing that claim with different examples is another
-important next step.  Of particular interest would be the class of
-so-called \emph{hybrid systems}: [[insert def]].  A lathe, for
-instance, that has an intermittent instability.  Traffic at an
-internet router, which has characteristic patterns during normal
-operation that shift radically during an attack.  Modeling and
-predicting the behavior of these kinds of systems effectively is quite
-difficult; doing so adaptively and automatically---in the manner that
-is alluded to at the end of the previous paragraph---would be an even
-more important challenge.
+important next step---and one that will be critical to validating our
+claims about the utility of the dashed line as a heuristic for
+determining whether a particular forecast method is well-matched to a
+particular task.  Of particular interest would be the class of
+so-called \emph{hybrid systems}, which exhibit discrete transitions
+between different continuous regimes---e.g., a lathe that has an
+intermittent instability or traffic at an internet router, which has
+characteristic patterns during normal operation that shift radically
+during an attack.  Effective modeling and prediction of these kinds of
+systems effectively is quite difficult; doing so adaptively and
+automatically---in the manner that is alluded to at the end of the
+previous paragraph---would be an even more important challenge.
 

### 2014-04-22 intro.tex
added sentence to abstract
Liz is done rewriting

@@
 predictive structure, i.e, signals whose future values can be
-successfully predicted from past values.  This can be viewed as the
-output of an underlying process that generates information and/or
-transmits it from the past to the future in a perfectly predictable
-fashion.  Constant or periodic signals, for example, fall in this
-class.  On the opposite end of this spectrum are signals that are what
-we call \emph{fully complex}, where the underlying generating process
-transmits no information at all from the past to the future.  White
-noise and random-walk processes fall in this class.  In fully complex
-signals, knowledge of the past gives no insight into the future,
-\emph{regardless} of what model one chooses to use.  Signals in the
-midrange of this spectrum---e.g., deterministic chaos---pose
-interesting challenges from a modeling perspective.  In these signals,
-enough information is being transmitted from the past to the future
-that an \emph{ideal} model---one that captures the generating
-process---can forecast the future behavior of the observed system with
-high accuracy.
+successfully predicted from past values.  Signals like this can be
+viewed as the product of an underlying process that generates
+information and/or transmits it from the past to the future in a
+perfectly predictable fashion.  Constant or periodic signals, for
+example, fall in this class.  On the opposite end of this spectrum are
+signals that are what we call \emph{fully complex}, where the
+underlying generating process transmits no information at all from the
+past to the future.  White noise and random-walk processes fall in
+this class.  In fully complex signals, knowledge of the past gives no
+insight into the future, regardless of what model one chooses to use.
+Signals in the midrange of this spectrum---e.g., deterministic
+chaos---pose interesting challenges from a modeling perspective.  In
+these signals, enough information is being transmitted from the past
+to the future that an \emph{ideal} model---one that captures the
+generating process---can forecast the future behavior of the observed
+system with high accuracy.
 
@@
 entropy~\cite{lind95}.  That is, we view a random-walk time series
-(which exhibits high entropy) as purely complex, whereas a periodic
-signal, which exhibits low entropy, is on the low end of the
-complexity spectrum.  We argue that an extension of \emph{permutation
+(which exhibits high entropy) as purely complex, whereas a low-entropy
+periodic signal is on the low end of the complexity spectrum.  We
+argue that an extension of \emph{permutation
   entropy}~\cite{bandt2002per}---a method for approximating the
@@
 
-We focus on real-valued, scalar, time-series data that appear\alert{s}
-complicated: i.e., not periodic, linear, constant, etc.  We make no
-assumptions about the properties of the generating process: whether it
-is linear, nonlinear, deterministic, stochastic, stationary,
-non-stationary, etc.
-To explore the relationship between complexity, predictive structure,
-and predictability, we generate forecasts for a variety of time-series
-datasets using four different prediction methods, then compare the
-accuracy of those predictions to the permutation entropy of the
-associated signals.  This results in two primary findings:
+We focus on real-valued, scalar, time-series data.
+% 
+% that appears complicated: i.e., not periodic, linear, constant, etc.
+% 
+We make no assumptions about the properties of the generating process:
+whether it is linear, nonlinear, deterministic, stochastic,
+stationary, non-stationary, etc.  To explore the relationship between
+complexity, predictive structure, and predictability, we generate
+forecasts for a variety of time-series datasets using four different
+prediction methods, then compare the accuracy of those predictions to
+the permutation entropy of the associated signals.  This results in
+two primary findings:
 \begin{enumerate}
-\item The complexity of a noisy real-valued time series is {\sl (i)}
-  quantifiable by permutation entropy and {\sl (ii)} correlated with prediction
-  accuracy of an appropriate predictor.
+\item The complexity of a noisy real-valued time series is \emph{(i)}
+  quantifiable by permutation entropy and \emph{(ii)} correlated with
+  prediction accuracy of an appropriate predictor.
 \item The way information is generated and processed internally by a
@@
 time series.  From a practitioner's standpoint, it would be useful to
-know {\sl a priori} if a time series contains enough predictive structure to
-make it worth spending the time and effort searching for a good
-forecast method.  It would also be useful to know if a given method is
-inadequate---that is, if an alternative method could do better and so 
-one should continue searching.
+know \emph{ a priori} if a time series contains enough predictive
+structure to make it worth spending the time and effort searching for
+a good forecast method.  It would also be useful to know if a given
+method is inadequate---that is, if an alternative method could do
+better and so one should continue searching.
 
@@
 For the purposes of this study, we require a broad array of
-time-series datasets from across the complexity spectrum.  Rather than
-use synthetic data, we study sensor data from computer performance
-experiments.  While this is not a common laboratory experiment, it is
-a highly appropriate choice.  Computers are extremely complicated
-systems.  Modern microprocessor chips contain multiple processing
-units and multi-layer memories, for instance, and they use complicated
-hardware/software strategies to maximize performance by moving data
-and threads of computation across those resources.  As a result, the
-processor and memory loads during the execution of even a very simple
-program can evolve in a very complicated fashion.
-Figure~\ref{fig:col-ipc} shows a performance trace of a three-line
-loop that repeatedly initializes the upper triangle of a matrix in
-column-major order.
+time-series datasets from across the complexity spectrum.  Since
+forecasting is a real-world problem, we chose to study sensor data
+from a computer-performance experiment.  While this is not a common
+laboratory experiment, it is a highly appropriate choice here.
+Computers are extremely complicated systems.  Modern microprocessor
+chips contain multiple processing units and multi-layer memories, for
+instance, and they use complicated hardware/software strategies to
+maximize performance by moving data and threads of computation across
+those resources.  As a result, the processor and memory loads during
+the execution of even a very simple program can evolve in a very
+complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance
+trace of a short program that repeatedly initializes the upper
+triangle of a matrix in column-major order.
 %
@@
     \caption{A computer performance trace: the processor load of \col,
-      a three-line C program that repeatedly initializes a matrix in
+      a simple C program that repeatedly initializes a matrix in
       column-major order, running on an Intel

### 2014-04-22 methods.tex
Liz is done rewriting

@@
 
-The time-series data for these experiments was collected on an Intel
-Core\textsuperscript{\textregistered} i7-2600-based machine running
-the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
-eight processing units, a clock rate of 3.40 GHz, and a cache size of
-8192 KB.  We gathered performance traces during the execution of three
-different programs---the simple \col loop whose performance is
-depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
+The time-series data sets for these experiments were collected on an
+Intel Core\textsuperscript{\textregistered} i7-2600-based machine
+running the 2.6.38-8 Linux kernel.  This particular microprocessor
+chip has eight processing units, a clock rate of 3.40 GHz, and a cache
+size of 8192 KB.  We gathered performance traces during the execution
+of three different programs---the simple \col loop whose performance
+is depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
 programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
-from the LAPACK linear algebra package (\svd).
-In these experiments, the scalar observation $x_i$ was a measurement of the processor performance at time $i$ during the execution of each program.
-To record these measurements, we used the {\tt libpfm4}
-library, via PAPI
+from the LAPACK linear algebra package (\svd).  In these experiments,
+the scalar observation $x_i$ was a measurement of the processor
+performance at time $i$ during the execution of each program.  To
+record these measurements, we used the {\tt libpfm4} library, via PAPI
 % 
@@
 5.2~\cite{papi}, to stop program execution at 100,000-instruction
-intervals and read the contents of the CPU's onboard hardware
-performance monitors\footnote{These specialty registers are built into
-  modern microprocessors for the purpose of on-board monitoring and
-  storage of performance data.  They can be programmed to measure a
-  wide array of processor and memory metrics besides the one used
-  here.}, which we had programmed to count how many instructions were
-executed in each clock cycle (IPC).  For an in-depth description of
-this custom-measurement infrastructure, including discussion of the
-implications of the sampling interval, please see
-\cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical validation,
-we collected 15 performance traces from each of the three programs.
-These traces, and the processes that generated them, are described in
-more depth in the rest of this section.
+intervals---the unit of time in this paper---and read the contents of
+the CPU's onboard hardware performance monitors\footnote{These
+  specialty registers are built into modern microprocessors for the
+  purpose of monitoring and storage of performance data.  They can be
+  programmed to measure a wide array of processor and memory metrics
+  besides the one used here.}, which we had programmed to count how
+many instructions were executed in each clock cycle (IPC).  For an
+in-depth description of this custom-measurement infrastructure,
+including discussion of the implications of the sampling interval,
+please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
+validation, we collected 15 performance traces from each of the three
+programs.  These traces, and the processes that generated them, are
+described in more depth in the rest of this section.
 
@@
 The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
-complicated programs that are used in the computer science community
+complicated programs that are used in the computer-science community
 to assess and compare the performance of different computers.  \gcc is
@@
 % 
-It is a \emph{C-compiler}: a program that translates code written in the
-C language into a lower-level format that can be executed by the
-processor chip.  Its behavior is far more complicated than that of
-\col, as is clear from Figure~\ref{fig:gcc-ts}.
+It is a \emph{compiler}: a program that translates code written in a
+high-level language (C, in the case of \gcc) into a lower-level format
+that can be executed by the processor chip.  Its behavior is far more
+complicated than that of \col, as is clear from
+Figure~\ref{fig:gcc-ts}.
   \begin{figure}[t]
@@
 phases---diagonalizing the matrix, computing its transpose,
-multiplying, etc.---the processor utilization changes quite radically,
+multiplying, etc.---the processor utilization patterns change quite radically,
 as shown in Figure~\ref{fig:svd-ts-colored}.
@@
       execution of \svd. Each point is the average IPC in a 100,000
-      instruction period.  The colors identify the different {\sl
+      instruction period.  The colors identify the different \emph{
         segments} of the signal that are discussed in the text.}
@@
 have color-coded in the figure in order to make textual
-cross-references easy to track and understand.
+cross-references easy to track.
 

### 2014-04-22 model.tex
Liz is done rewriting

@@
 the forecast: that is, the predicted value $p_i$ at time $i$ is
-calculated using the relation $$p_i = x_{i-1}$$ The prediction
-strategy that we refer to using the term ``\naive'' averages the prior
-observations to generate the forecast: $$p_i =
+calculated using the following relation: $$p_i = x_{i-1}$$ The
+prediction strategy that we refer to using the term ``\naive''
+averages the prior observations to generate the forecast: $$p_i =
 \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
@@
 consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
-These signals are constantly changing, noisy, and possess very little predictive
-structure, but their variations are not---on the average---very large, so
-the random-walk method's strategy of simply guessing the last known
-value is not a bad choice.  If a signal has a unimodal distribution
-with low variance, the \naive ~prediction strategy will perform quite
-well---even if the signal is highly complex---simply because the mean
-is a good approximation of the future behavior.  Moreover, the \naive
-~prediction strategy's temporal average effects a low-pass filtering
-operation, which effectively mitigates the complexity in signals with
-very little predictive structure.
+These signals are constantly changing, noisy, and possess very little
+predictive structure, but their variations are not---on the
+average---very large, so the random-walk method's strategy of simply
+guessing the last known value is not a bad choice.  If a signal has a
+unimodal distribution with low variance, the \naive ~prediction
+strategy will perform quite well---even if the signal is highly
+complex---simply because the mean is a good approximation of the
+future behavior.  Moreover, the \naive ~prediction strategy's temporal
+average effects a low-pass filtering operation, which can  mitigate the
+complexity in signals with very little predictive structure.
 
@@
 ~method would be a better choice in this situation, since it would
-always predict the mean.  It would, however, perform poorly when a
+always split the difference.  It would, however, perform poorly when a
 signal has a number of long-lived regimes that have significantly
@@
 often be outperformed by the two simple strategies described in this
-section.  This is explored further in Sections~\ref{sec:accuracy}
+section.  This effect is explored further in Sections~\ref{sec:accuracy}
 and~\ref{sec:results}.
@@
 A simple and yet powerful way to capture and exploit the structure of
-data is to fit a hyperplane to the dataset and then use that to
-make predictions.  The roots of this approach date back to the
-original autoregressive schema~\cite{weigend93}, which forecasts the
-next time step through a weighted average of past observations: $$p_i
-= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
-computed generally using either an ordinary least squares approach, or with the method of moments using the Yule-Walker equations.  To account for noise in the data, one can add
-a so-called ``moving average'' term to the model; to remove
-nonstationarities, one can detrend the data using a differencing
-operation.  A strategy that incorporates all three of these features
-is called a \emph{nonseasonal ARIMA model}.  If evidence of periodic
-structure is present in the data, a \emph{seasonal ARIMA model}, which
-adds a sampling operation that filters out periodicities, can be a
-good choice.
+data is to fit a hyperplane to the dataset and then use it to make
+predictions.  The roots of this approach date back to the original
+autoregressive schema~\cite{weigend93}, which forecasts the next time
+step through a weighted average of past observations: $$p_i =
+\sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
+generally computed using either an ordinary least squares approach, or
+with the method of moments using the Yule-Walker equations.  To
+account for noise in the data, one can add a so-called ``moving
+average'' term to the model; to remove nonstationarities, one can
+detrend the data using a differencing operation.  A strategy that
+incorporates all three of these features is called a \emph{nonseasonal
+  ARIMA model}.  If evidence of periodic structure is present in the
+data, a \emph{seasonal ARIMA model}, which adds a sampling operation
+that filters out periodicities, can be a good choice.
 
@@
 build forecasts in a stepwise fashion: i.e., fit the model to the
-existing data, use it to perform a one-step prediction, rebuild the
-model using the latest observations, and iterate until the desired
+existing data, use that model to perform a one-step prediction, rebuild it
+using the latest observations, and iterate until the desired
 prediction horizon is reached.  (For consistency, we take the same
 approach with the other three models in this study as well, even
-though this amounts to artificially hobbling LMA.)
+though doing so amounts to artificially hobbling LMA.)
 
@@
 dynamics from a scalar time-series measurement---provided that some
-conditions hold regarding that data.  Specifically, if the underlying
+conditions hold regarding those data.  Specifically, if the underlying
 dynamics and the measurement function---the mapping from the unknown
@@
 same topology.  This is an extremely powerful result: among other
-things, it means that one can model the full system dynamics without
-measuring, or even knowing, every one of its state variables.
+things, it means that one can model the full system dynamics, up to
+diffeomorphism, without measuring---or even knowing---every one of its
+state variables.
 
@@
 $\tau$ and the dimension $m$.  We follow standard procedures for this,
-choosing the first minimum in the time-delayed mutual information as an
-estimate of $\tau$~\cite{fraser-swinney} and using the
-false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.  Plots of
-the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored},
-embedded following this procedure, are shown in
-Figure~\ref{fig:embedding}.
+choosing the first minimum in the time-delayed mutual information as
+an estimate of $\tau$~\cite{fraser-swinney} and using the
+false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.
+Some example plots of data from
+Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following
+this procedure, are shown in Figure~\ref{fig:embedding}.
  \begin{figure}
@@
 predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
-One of the most straightforward of these is the ``Lorenz method of
-analogues'' (LMA), which is essentially nearest-neighbor prediction in
-the embedded\footnote{Lorenz's original formulation used the full
-  system state space; According to \cite{kantz97} this method was first extended to embedded dynamics by Pikovsky ~\cite{pikovsky86-sov}, but more closely related to the prediction work of Sugihara \& May in \cite{sugihara90}}
+One of the most straightforward of these is the \emph{Lorenz method of
+  analogues} (LMA), which is essentially nearest-neighbor prediction
+in the embedded\footnote{Lorenz's original formulation used the full
+  system state space; According to \cite{kantz97} this method was
+  first extended to embedded dynamics by Pikovsky
+  ~\cite{pikovsky86-sov}, but more closely related to the prediction
+  work of Sugihara \& May in \cite{sugihara90}}
 space~\cite{lorenz-analogues}.  Even this simple algorithm---which
@@
 generating process is nondeterministic, however, it can perform
-poorly.  In Figure~\ref{fig:embedding}(b), for instance, little
+poorly.  In Figure~\ref{fig:embedding}b, for instance, little
 structure is visible, so LMA may not work well.  More structure
-appears to be present in~Figure~\ref{fig:embedding}(c), but this
+appears to be present in~Figure~\ref{fig:embedding}c, but this
 reconstruction also appears to contain some noise.  The question as to
@@
 is effectively obfuscated, from the standpoint of the LMA method, by
-noise.  By quantifying the balance between normalized (that
-is, predictive structure) and entropy for these real-valued time
-series, as shown in Section~\ref{sec:results}, we can begin to answer
-these questions.
+noise.  By quantifying the balance between redundancy (that is,
+predictive structure) and entropy for these real-valued time series,
+as shown in Section~\ref{sec:results}, we can begin to answer these
+questions.
 
@@
 ``one step prediction'' process is not technically necessary in the
-LMA method, which can generate arbitrary-length
-predictions---although the accuracy of these predictions will
-  degrade over prediction length, in the presence of positive Lyapunov
-  exponents, but the performance of the other three methods used here
-will degrade severely if the associated models are not periodically
-rebuilt.  In order to make the comparison fair, we used an iterative
-one-step prediction schema {\sl for all four methods}.  This has the
-slightly confusing effect of causing the ``test'' signal to be used
-both to assess the accuracy of each model and for periodic refitting.
+LMA method, whose ability to generate accurate predictions is limited
+only by the positive Lyapunov exponents of the system.  However, the
+performance of the other three methods used here will degrade severely
+if the associated models are not periodically rebuilt.  In order to
+make the comparison fair, we used an iterative one-step prediction
+schema \emph{for all four methods}.  This has the slightly confusing
+effect of causing the ``test'' signal to be used both to assess the
+accuracy of each model and for periodic refitting.
 
@@
    \caption{Predicted ($p_j$) versus true values ($c_j$) for
-     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
-     perfect prediction lies exactly on the diagonal.
+     forecasts of \col, \gcc, and \svdfive.  
+% On this type of plot, a perfect prediction lies exactly on the
+% diagonal.
 %
@@
 structure of the distribution of the errors---roughly normally
-distributed, for example, in the case of LMA on \gcc.  The shapes of
-Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are reminiscent of the
-projected embedding in Figure~\ref{fig:embedding}(a).  For a
-random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
+distributed, for example, in the case of LMA on \gcc.  Note that the
+shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
+reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
+For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
 two-dimensional embedding with $\tau=1$.  For ARIMA, the
@@
 linear combinations of a number of past values of the $c_j$, but the
-effect is largely the same\footnote{The temporal ordering of the
-  points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of a
-  projected embedding of the time series, however.}.
+effect is largely the same\footnote{The temporal \emph{ordering} of
+  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
+  a projected embedding of the time series, however.}.
 
@@
 applicable measurement of forecast accuracy without the problems seen
-in the other measurements."  MASE is a normalized measure; the
-scaling term in the denominator
+in the other measurements."  MASE is a normalized measure: the scaling
+term in the denominator
 % 

### 2014-04-22 pecomplexity.tex
Liz is done rewriting

@@
 conjecture that the inherent complexity of a time series —-- which
-results from the inherent dimension, nonlinearity, and
-non-stationarity of the generating process, as well as from
-measurement issues like noise, aggregation, and finite data length ---
-is both empirically quantifiable and directly correlated with
-predictability. To validate this conjecture, we study 120 different
-time-series data sets.  We argue that \emph{redundancy} is an
-effective way to measure complexity and predictive structure in a time
-series and that \emph{weighted permutation entropy} is an effective
-way to estimate that complexity.  For each time series, we construct
+results from the dimension, nonlinearity, and non-stationarity of the
+generating process, as well as from measurement issues like noise,
+aggregation, and finite data length --- is both empirically
+quantifiable and directly correlated with predictability.  In
+particular, we argue that \emph{redundancy} is an effective way to
+measure complexity and predictive structure in a time series and that
+\emph{weighted permutation entropy} is an effective way to estimate
+that complexity.  To validate these conjectures, we study 120
+different time-series data sets.  For each time series, we construct
 predictions using a wide variety of forecast models, then compare the

### 2014-04-22 related.tex
Liz is done rewriting

@@
 not to add a new weapon to this arsenal, nor to assess or compare the
-effectiveness of existing methods.  Our goals are more general: {\sl
-  (i)} to empirically quantify the predictive structure that is
-present in a real-valued scalar time series and {\sl (ii)} to explore
+effectiveness of existing methods.  Our goals are more general:
+\emph{(i)} to empirically quantify the predictive structure that is
+present in a real-valued scalar time series and \emph{(ii)} to explore
 how the performance of prediction methods is related to that inherent
@@
 These distributions are then used as estimates of out-of-sample
-forecast errors in those regions.  For example, Smith {\sl et al.}
+forecast errors in those regions.  For example, Smith \emph{et al.}
 make in-sample forecasts using ensembles around selected points in
@@
 calculate forecast errors and verify that they are normally distributed.
-If they are not, that indicates that there is structure in the time
+If they are not, that suggests that there is structure in the time
 series that the model-building process was unable to capture and use.
@@
 of the Kolmogorov-Sinai entropy of a real-valued time series---and
-thus the redundancy in that data, which is an effective proxy for
+thus the redundancy in that data, which we show to be an effective proxy for
 predictability.  This differs from existing approaches in a number of

### 2014-04-22 results.tex
Liz is done rewriting

@@
 
-In this section, we offer an empirical validation of the two key
-conjectures introduced in Section \ref{sec:intro}, namely:
+In this section, we offer an empirical validation of the two
+conjectures offered in Section \ref{sec:intro}, namely:
 
@@
 
-\item that the weighted permutation entropy (WPE) of a noisy
+\item That the weighted permutation entropy (WPE) of a noisy
   real-valued time series is correlated with prediction
@@
 
-\item that the way information is generated and processed by a system
+\item That the way information is generated and processed by a system
   is correlated with the effectiveness of a given predictor on
@@
 
-Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of these 45
+Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these 45
 predictions for each system: that is, the lowest error over all 15
@@
 \end{figure}
-The best-case prediction error for each of these eight systems is
-roughly proportional to the weighted permutation entropy, which is
-consistent with our first conjecture.  The dashed line in the figure,
-a linear least-squares fit to these best-case errors\footnote{with the
-  exception of {\tt dgesdd$_1$}, for reasons described below} captures
-this rough proportionality.  This is not a formal result.  The three
-methods used here were chosen to span the space of standard prediction
-strategies, but they do not cover that space exhaustively.  Our goal
-here is an empirical assessment of predictability and complexity, not
-formal results about a ``best'' predictor for a given time series.
-There may be other methods that produce lower MASE values than those
-in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
-in the upper-left and lower-right quadrants of this plot strongly
+As the plot shows, the best-case prediction error for each of these
+eight systems is roughly proportional to the weighted permutation
+entropy, which is consistent with our first conjecture.  The dashed
+line in the figure, a linear least-squares fit to these best-case
+errors\footnote{with the exception of {\tt dgesdd$_1$}, for reasons
+  described below} captures this rough proportionality.  This is not a
+formal result.  The three forecast methods used here were chosen to
+span the space of standard prediction strategies, but they do not
+cover that space exhaustively.  Our goal here is an empirical
+assessment of predictability and complexity, not formal results about
+a ``best'' predictor for a given time series.  There may be other
+methods that produce lower MASE values than those in
+Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points in
+the upper-left and lower-right quadrants of this plot strongly
 suggests that the underlying predictability of a time series is
 inversely correlated with its WPE.  The rest of this section describes
-our results in more detail, including the measures taken to assure
-meaningful comparisons across methods, trials, and programs, and
+our results in more detail---including the measures taken to assure
+meaningful comparisons across methods, trials, and programs---and
 elaborates on the meaning of the dashed line in the figure.
 
-Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
-experiments.
+Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
+set of experiments.
 \begin{figure}
@@
   \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
-\caption{WPE vs. MASE for all trials, methods, and systems. \svdone,
-  \svdthree, and \svdfive are omitted from the top plot for scale
-  reasons.
+\caption{WPE vs. MASE for all trials, methods, and systems---with the
+  exception of \svdone, \svdthree, and \svdfive are omitted from the
+  top plot for scale reasons.
 % 
@@
 low as well, resulting in small, tight clusters.  In some
-cases---ARIMA predictions of {\tt col\_major}, for instance---the
-MASE variance is larger, which spreads out the clusters
-horizontally.  The distribution of the cluster {\sl positions} is
-quite different in the three plots.  The mean MASE scores of
-predictions generated with the nonlinear LMA method are generally
-closer to the dashed line; the ARIMA method clusters are more
-widely spread, the \naive ~clusters even more so.  This geometric
-validation of the second conjecture in this paper is discussed in more
-details in the following paragraphs.
+cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
+variance is larger, which spreads out the clusters horizontally.  The
+distribution of the cluster \emph{positions} is quite different in the
+three plots.  The mean MASE scores of predictions generated with the
+nonlinear LMA method are generally closer to the dashed line; the
+ARIMA method clusters are more widely spread, the \naive ~clusters
+even more so.  This geometric validation of the second conjecture in
+this paper is discussed in more detail in the following paragraphs.
 
@@
 
-The dashed line in the Figures is a heuristic that captures the
+The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
+and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
 argument at the end of the previous paragraph.  It can allow
@@
 
-\svdone behaves very differently than the other seven systems in this
-study.  Though its weighted permutation entropy is very high ($0.957
-\pm 0.016$), two of the three prediction methods do quite well on this
-signal, yielding MASE scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
-0.075$ (ARIMA).  This pushes the corresponding clusters of points in
-the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
-the trend followed by the other seven signals.  The MASE scores of the
-predictions that were produced by the \naive ~method for this system,
-however, are highly inconsistent.  The majority of the blue
-diamond-shaped points on the top plot in
-Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
-0.6.  In five of the 15 \svdone trials, however, there were step
-changes in the signal.  The \naive ~method has a very difficult time
-with signals like this, particularly if there are multiple step
-changes.  This raised the MASE scores of these trials, pushing the
-corresponding points to the right\footnote{This includes the cluster
-  of three points near MASE $\approx 2.5$, as well as two points that
-  are beyond the domain of the graph, at MASE 11.2 and 14.8.}, and in
-turn raising both the mean and variance of this set of trials.
+\svdone---the dark blue segment of
+Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
+other seven systems in this study.  Though its weighted permutation
+entropy is very high ($0.957 \pm 0.016$), two of the three prediction
+methods do quite well on this signal, yielding MASE scores of $0.827
+\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (ARIMA).  This pushes the
+corresponding clusters of points in the bottom two images in
+Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
+other seven signals.  The MASE scores of the predictions that were
+produced by the \naive ~method for this system, however, are highly
+inconsistent.  The majority of the blue diamond-shaped points on the
+top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
+score of 0.6, which is better than LMA and ARIMA.  In five of the 15
+\svdone trials, however, there were step changes in the signal.  The
+\naive ~method has a very difficult time with signals like this,
+particularly if there are multiple step changes.  This raised the MASE
+scores of these trials, pushing the corresponding points to the
+right\footnote{This includes the cluster of three points near MASE
+  $\approx 2.5$, as well as two points that are beyond the domain of
+  the graph, at MASE 11.2 and 14.8.}, and in turn raising both the
+mean and variance of this set of trials.
 
@@
 is dominated by a noisy regime (between $\approx$1.86 and
-$\approx$1.88 on the vertical scale in the Figure), punctuated by
-short excursions above 1.9.  In the former regime, which makes up more
-than 80\% of the signal, there are frequent dips to 1.82 and
-occasional larger dips below 1.8.  These single-point dips are the
+$\approx$1.88 on the vertical scale in Figure~\ref{fig:svdone-ts}),
+punctuated by short excursions above 1.9.  In the former regime, which
+makes up more than 80\% of the signal, there are frequent dips to 1.82
+and occasional larger dips below 1.8.  These single-point dips are the
 bane of random-walk forecasting.  In this particular case, roughly
@@
 effect of the skew in the scaling factor and exacerbates the spread in
-the \svdone MASE values.  For all of these reasons, we view \svdone
-as an outlier and exclude it from the fit calculation of the dashed
-line in Figure~\ref{fig:wpe_vs_mase_best}.
+the \svdone MASE values.  For all of these reasons, we view \svdone as
+an outlier and exclude it from the fit calculation of the dashed line
+in Figure~\ref{fig:wpe_vs_mase_best}.
 

### 2014-04-22 wpe.tex
Liz is done rewriting

@@
 whereas highly structured time-series, for example periodic, have very
-low or zero entropy rate. A time series with a high entropy rate is 
+low (or zero) entropy rates. A time series with a high entropy rate is
 almost completely unpredictable; conversely, one with low entropy rate
@@
 time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
-x(0) \right|$.  The faster the divergence, the larger the entropy, and
-the more difficult prediction becomes.  The KS entropy is defined as
-the supremum of the Shannon entropy rates of all
-partitions~\cite{petersen1989}. The partition that achieves this
-supremum is the \emph{generating partition} discussed in
+x(0) \right|$.  The faster the divergence, the larger the entropy.
+The KS entropy is defined as the supremum of the Shannon entropy rates
+of all partitions~\cite{petersen1989}. The partition that achieves
+this supremum is the \emph{generating partition} discussed in
 Section~\ref{sec:related}.
@@
 method will work well or poorly on the corresponding signal.  A linear 
-method cannot detect or make use of nonlinear redundancy, for 
+method cannot capture or make use of nonlinear redundancy, for 
 instance.  This issue is central to the claims in this paper and the 
@@
 infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
-systems is, however, effectively\footnote{Measurements from
+systems are, however, effectively\footnote{Measurements from
   finite-precision sensors are discrete, but data from modern
@@
 discretizing real-valued time series that follows the intrinsic
-behavior of the system under examination, and has many advantages 
-including being robust to observational noise and its application does 
-not require any knowledge of the underlying mechanisms.  Rather than 
-looking at the statistics of sequences of values, as is done when 
+behavior of the system under examination.  It has many advantages,
+including being robust to observational noise, and its application
+does not require any knowledge of the underlying mechanisms.  Rather
+than looking at the statistics of sequences of values, as is done when
 computing the Shannon entropy, permutation entropy looks at the
@@
 shown to be identical to the Kolmolgorov-Sinai entropy for many large
-classes of systems \cite{amigo2012permutation}, so long as
+classes of systems \cite{amigo2012permutation}, as long as
 observational noise is sufficiently small. As mentioned before, this
@@
 system. This transitive chain of equalities, from permutation entropy
-to Shannon entropy rate via the KS entropy, then allows us to
-approximate the redundancy---being the dual of the Shannon entropy
-rate---of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
+to Shannon entropy rate via the KS entropy, allows us to approximate
+the redundancy---being the dual of the Shannon entropy rate---of a
+signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.
 
@@
 ours~\cite{haven2005}, but there are only superficial similarities
-between the two research projects.  Haven {\sl et al.} utilize the
+between the two research projects.  Haven \emph{et al.} utilize the
 relative entropy to quantify the difference in predictability between
-two distributions: one using the known dynamical system from a small 
-ensamble in the past, and the other the asymptotic distribution. Our 
-work quantifies the predictability of a single observed time series 
-using weighted permutation entropy and makes no assumptions about the
-generating process.
+two distributions: one {\color{red}[[Ryan: a distribution ``using'' a
+      dynamical system is confusing.  Do you mean ``created using''?
+      Or ``measured from''?  And what do you mean by ``from an
+      ensemble''?  ensemble of what?  used how?]]}  using the known
+dynamical system from a small ensemble in the past, and the other the
+asymptotic distribution. Our work quantifies the predictability of a
+single observed time series using weighted permutation entropy and
+makes no assumptions about the generating process.
 

### 2014-04-22 wpe.tex
fix horrible writing. shouldn't write while tired...

@@
 relative entropy to quantify the difference in predictability between
-two distributions: one {\color{red}[[Ryan: a distribution ``using'' a
-      dynamical system is confusing.  Do you mean ``created using''?
-      Or ``measured from''?  And what do you mean by ``from an
-      ensemble''?  ensemble of what?  used how?]]}  using the known
-dynamical system from a small ensemble in the past, and the other the
+two distributions: one evolved from a small ensamble of past
+states using the known dynamical system, and the other the
 asymptotic distribution. Our work quantifies the predictability of a

### 2014-04-25 concl.tex
some minor fixes

@@
 between different continuous regimes---e.g., a lathe that has an
-intermittent instability or traffic at an internet router, which has
-characteristic patterns during normal operation that shift radically
-during an attack.  Effective modeling and prediction of these kinds of
-systems effectively is quite difficult; doing so adaptively and
+intermittent instability or traffic at an internet router, whose
+characteristic normal traffic patterns shift radically during an
+attack.  Effective modeling and prediction of these kinds of systems
+effectively is quite difficult; doing so adaptively and
 automatically---in the manner that is alluded to at the end of the

### 2014-04-25 model.tex
some minor fixes

@@
 in the embedded\footnote{Lorenz's original formulation used the full
-  system state space; According to \cite{kantz97} this method was
-  first extended to embedded dynamics by Pikovsky
-  ~\cite{pikovsky86-sov}, but more closely related to the prediction
-  work of Sugihara \& May in \cite{sugihara90}}
+  system state space; 
+%
+%According to \cite{kantz97}, 
+%
+this method was first extended to embedded dynamics by
+Pikovsky~\cite{pikovsky86-sov}, but is also related to the prediction
+work of Sugihara \& May~\cite{sugihara90}}
 space~\cite{lorenz-analogues}.  Even this simple algorithm---which

### 2014-04-25 related.tex
some minor fixes

@@
 The basic idea is to construct a series of local linear fits,
-beginning with a few points and working up to a global linear fit
-that includes all known points, and then analyze how the average
-out-of-sample error changes as a function of number of points in the
-fit. The shape of such a ``DVS" graph indicates the amounts of determinism
-and stochasticity present in a time series.\alert{Liz Should we say what this has to do with predictability, realized this paragraph may seem unrelated to a casual reader}
+beginning with a few points and working up to a global linear fit that
+includes all known points, and then analyze how the average
+out-of-sample forecast error changes as a function of number of points
+in the fit. The shape of such a ``DVS" graph indicates the amounts of
+determinism and stochasticity present in a time series.
 

### 2014-04-25 wpe.tex
some minor fixes

@@
 relative entropy to quantify the difference in predictability between
-two distributions: one evolved from a small ensamble of past
+two distributions: one evolved from a small ensemble of past
 states using the known dynamical system, and the other the

### 2014-04-25 concl.tex
typo in concl

@@
 attack.  Effective modeling and prediction of these kinds of systems
-effectively is quite difficult; doing so adaptively and
-automatically---in the manner that is alluded to at the end of the
-previous paragraph---would be an even more important challenge.
+is quite difficult; doing so adaptively and automatically---in the
+manner that is alluded to at the end of the previous paragraph---would
+be an even more important challenge.
 

### 2014-04-25 wpe.tex
Liz is done

@@
 states using the known dynamical system, and the other the
-asymptotic distribution. Our work quantifies the predictability of a
+observed distribution. Our work quantifies the predictability of a
 single observed time series using weighted permutation entropy and

### 2014-04-26 model.tex
cleaning up post submission

@@
 
-Delay-coordinate embedding~\cite{packard80,Sauer:1991lr,Takens:1981uq}
+Delay-coordinate embedding~\cite{Sauer:1991lr,Takens:1981uq}
 allows one to reconstruct a dynamical system's full state-space

### 2014-04-27 concl.tex
beating this thing into PRE shape...

@@
 particular task.  Of particular interest would be the class of
-so-called \emph{hybrid systems}, which exhibit discrete transitions
+so-called \emph{hybrid systems}~\cite{hybrid}, which exhibit discrete transitions
 between different continuous regimes---e.g., a lathe that has an

### 2014-04-27 pecomplexity.tex
cleaning up post submission
beating this thing into PRE shape...

@@
 %preprint,
-%showpacs,preprintnumbers,
+showpacs,
+%preprintnumbers,
 nofootinbib,
@@
  amsmath,amssymb,
-% aps,
+ aps,
 %pra,
@@
 \usepackage{hyperref}
-\usepackage{caption}
+\usepackage[sort&compress]{natbib}
+%\usepackage{caption}
 \usepackage{subcaption}
+\usepackage{microtype}
 \usepackage{dcolumn}% Align table columns on decimal point
@@
 89.70.-a    % Information theory
-89.70.Hj    %   communication complexity
-89.70.Cf    %   entropy
-05.45.-a    % Nonlinear dynamics
+%89.70.Hj    %   communication complexity
+%89.70.Cf    %   entropy
+%05.45.-a    % Nonlinear dynamics
 89.75.Kd    %   Patterns (complex systems)
-95.75.Wx    % Time series analysis, time variability
+%95.75.Wx    % Time series analysis, time variability
 95.75.Pq    % Mathematical procedures and computer techniques
-07.05.-t    % Computers in experimental physics
+%07.05.-t    % Computers in experimental physics
 }
@@
 grant \#W911NF-12-1-0288.
+We would also like to acknowledge the Santa Fe Institute whose interdisciplinary halls created a place for this work to be born.
 
-\bibliographystyle{aipauth4-1}
+%\bibliographystyle{aipauth4-1}
+\bibliographystyle{apsrev4-1}
 \bibliography{bibliofile}

### 2014-05-30 pecomplexity.tex
add preprint info

@@
 {\bf Keywords}: entropy, permutation entropy, weighted permutation
-entropy, time-series analysis, predictibility
+entropy, time-series analysis, predictability
+
+\preprint{Santa Fe Institute Working Paper XX-XX-XXX}
+\preprint{arxiv.org:XXXX.XXXX [nlin.CD]}
 

### 2014-05-30 pecomplexity.tex
fix an dash

@@
 strategies fail when they are applied to complicated time series. We
-conjecture that the inherent complexity of a time series —-- which
+conjecture that the inherent complexity of a time series --- which
 results from the dimension, nonlinearity, and non-stationarity of the

### 2014-05-30 response.tex
start response document

\documentclass[12pt]{article}

\usepackage{amsmath}

\title{Response to the Reviewers}
\author{Joshua Garland, Ryan James, and Elizabeth Bradley}
\date{\today}

\begin{document}

\maketitle

\section*{Response to the First Referee}

\emph{In this manuscript the authors present a way to explore the relationship
between predictability and complexity across a broad (but not complete) array of
forecast strategies such as random walk (last value), na\"ive (average value),
regression based (ARIMA) and a nonlinear method based on state space
reconstruction (LMA). They use weighted permutation entropy as a criterion/index
of predictive structure in the time series and illustrate their approach on
eight real-world datasets derived from three different systems (computer
performance traces).}

\emph{Overall the manuscript is well written and clearly understandable. The
analysis seems to be carried out carefully and the conclusions drawn are quite
reasonable. While I don't get the impression that this is a breakthrough paper I
can still recommend the manuscript for publication in PRE if the issues listed
below are addressed thoroughly.}

\noindent\emph{Major issues:}

\emph{I don't get the reason for normalizing the three methods by the random
walk method (for example in Table I). As the authors write themselves this makes
the indicator vulnerable to a bias caused by the high influence of this
(arbitrarily selected) reference method. Why not just use a non-referential
estimator, which then would also allow to compare the inherent predictability of
different systems (as it is now the values obtained for each systems are
strongly influenced by the random walk estimate on that particular system and
thus can not be compared). This will also change the results of the fitting (the
authors explicitly exclude one value because of the effect caused by the
normalization). And it is not that the random walk predictor is the one standard
method that all other methods have to compete against.}

\emph{In many parts the article reads almost like a review. There is a lot of
unnecessary redundancy (the entropy rate is quite low). In particular, the main
point (``The aim of the paper is not ... but ...'') is repeated far too many
times. The paper while easy to read could be shortened considerably without any
loss of information.}

\noindent\emph{Minor issues:}

\emph{Page 1: Why mention the `halting problem'? I don't see a straightforward
connection to the problem at hand. The only connection is `undecidability' but
there are many other examples for that.}

\emph{Page 10: `error between the predictions and the true continuations' -->
`error of the predictions with respect to the true continuations' Figure 6 on
Page 11: It would be more intuitive to turn this into a `table of figures'
with the four methods as row labels on the left and the three systems as column
labels at the top. Also: Caption `for forecast of ... and all four prediction
strategies'.}

\emph{First equation on Page 14: Shouldn't the index $j$ run from $i+1$ to
$i+\ell$ (instead of from $i$ to $i+\ell$)? You normalize by $\ell$ (not
$\ell+1$) values.}

\emph{Page 19: ``There are some exceptions: [...] Conversely, forecast errors
for the \texttt{col\_major} and \texttt{dgesdd$_5$} signals are higher than the
corresponding WPEs suggest --- except for the nonlinear LMA predictor.'' This is
not really an exception. As mentioned by the authors before, in general, the
result only holds under the assumption that a reasonably good model has been
used for the prediction. But for ill-matched datasets and models like in the
cases mentioned above this is just not the case so it is not a surprise that the
forecast errors are higher.}

\section*{Response to the Second Referee}

\emph{The problem of time series prediction is considered in the paper of J.
Garland et al. Namely, the authors suggest to quantify predictability in advance
(before using any concrete prediction scheme) based on the complexity estimate
(weighted permutation entropy, WPE). They hypothesize that WPE is related to the
best possible prediction error in a simple way. For several benchmark time
series (reflecting variations of a computer performance during execution of
different tasks), they obtain estimates of WPE and prediction error (the least
one over four techniques), and build a linear regression using seven or eight
resulting data points. They suppose that the obtained linear function can be
used to ``predict'' the best prediction error from the model-free WPE estimate.
Thus, the authors finally assume (as could be seen from the last section) that
WPE and optimal prediction error for all systems are related via the same
universal linear function with fixed coefficients. This is what I can understand
from the entire paper, though the problem is not so definitely described in the
Introduction.}

\emph{I think that the possibility to estimate the best prediction error in
advance from a relatively simple WPE characteristic would be interesting and
useful, if it could hold true for a wide class of systems. Such a question would
deserve a careful study as that started in the presented work. However, I have a
number of critical remarks and strongly doubt that the conjecture of the authors
can be valid for a reasonably wide class of systems (not even speaking of ``all
systems''). Other aspects of the paper are not so important, in my view, despite
some of them are fresh and interesting (e.g. taking a computer as an object of
prediction). My critical remarks are listed below.}

\emph{1) Introduction is too long, but the main problem is not formulated
definitely enough. I see that the main point is the approximately linear (or at
least one-to-one) relationship between WPE and the best prediction error. This
aim is not claimed explicitly in the Introduction. All other aspects of the
manuscript do not seem important enough to be published in a separate article.}

\emph{2) In particular, two primary findings mentioned on page 2 (Introduction)
are formulated quite vaguely. The first part of the first finding reads ``(i)
complexity of a noisy real-value time series is quantifiable by permutation
entropy''. However, it seems obvious from the previous research, including that
cited by the authors. The authors do not suggest PE or WPE here, they exploit
them as previously known approaches. It was previously known that PE is an
estimate of KS entropy as the authors also state. It was well known that KS
entropy relates to predictability and, in this sense, it is a measure of
complexity (see e.g. the work [G. Boffetta, M. Cencini, M. Falcioni, and A.
Vulpiani, ``Predictability: a way to characterize complexity'' // Physics
Reports 356 (2002) 367–474] which is not cited by the authors). Thus, the first
part of the first finding does not provide any new information. The second part
of the first finding reads ``(ii) complexity of the noisy real-valued time
series is correlated with prediction accuracy of an appropriate predictor''.
However, it seems also obvious and directly follows from the definition of KS-
entropy and its discussion in many papers including [Boffetta et al.] cited
above.}

\emph{3) By the way, the authors call ``fully complex'' the white noise process,
i.e. they equate predictability and complexity. However, they do not even
mention another approach to the notion of complexity which ascribes low
complexity to the white noise [the works e.g. by C.R. Shalizi and J. Crutchfield
carried out in the Santa Fe Institute from where one of the authors is]. It
would be appropriate to mention that ambiguity of the notion of complexity if
the latter plays an important role in many formulations of the authors.}

\emph{4) The second finding reads ``The way information is generated and
processed internally by a system plays a crucial role in the success of
different forecasting schema - and in the choice of which one is appropriate for
a given time series''. This statement also looks trivial. Indeed, an optimal
predictor for a linear stochastic system and low-dimensional nonlinear
deterministic system are quite different (an AR model versus a local or global
nonlinear model) due to different properties of the original systems. Thus,
there is nothing to be proven here. Probably, the authors imply their particular
benchmark signals (reflecting a computer behavior) but they do not claim that.
Thus, both primary findings seem to provide nothing new.}

\emph{5) The benchmark system - a computer as an object of prediction - seems to
be an unexpected and, probably, interesting choice but just as an additional
illustration of a statement which should be first shown for well-controlled and
well-understood paradigmatic systems (like low-dimensional maps or ordinary
differential equations). Indeed, it is not known in advance what error one
should expect for such a complicated object using a certain predictive
technique, what technique is optimal or close to optimal, etc. If the main point
of the research is the relationship between WPE and the best prediction error
(if any), selection of such a complicated object as computer performance
variations only introduces its own difficulties into the problem.}

\emph{6) The description of the choice of the benchmark system in the
Introduction is too long. On the other hand, if the computer signals are the
main interest of the work, then it should be claimed explicitly. Then, the paper
should be rewritten and submitted to a kind of engineering journals where it
might well be of interest.}

\emph{7) As for the choice of the four prediction techniques, some of them seem
superfluous. E.g. naive approach is just a global AR model of the zero order.
Random-walk is similar to the AR model of the first order where the coefficient
is not estimated by the least-squares but set equal to 1 (or this is an ARIMA
model with the first difference and ``zero ARMA part''). Inefficiencies of the
ARIMA model as compared to those choices in the paper can represent an
inappropriate procedure for the order selection rather than indicate
inappropriateness of the entire ARIMA approach. Thus, linear ARIMA and nonlinear
LMA might be sufficient (in combination with mathematical benchmark examples) to
represent close-to-optimal prediction errors.}

\emph{8) Conditions for inefficiency of simple prediction schemes (like naive
and random-walk) discussed in the paper look rather obvious as well and do not
give new information. Is it the purpose of the paper to discuss when and where a
concrete simple prediction scheme fails?}

\emph{9) A better justification of the prediction error metrics could be also in
order if this work is continued. At least, comparison of prediction errors
quantified with different metrics seems necessary. Especially, taking into
account the following major concerns.}

\noindent\emph{General remarks.}

\emph{I) Strict relationship between WPE and prediction error applicable to all
systems is hardly possible. In particular, there is a notion of epsilon-entropy
which tends to the KS entropy if epsilon (the cell size) tends to zero (see e.g.
[Boffetta et al.] cited above). For stochastic systems, the KS entropy is
infinite and epsilon-entropy varies strongly with the scale epsilon. Seemingly,
the permutation entropy estimate in this case will strongly depend on the choice
of the word length L. The authors concentrate on L = 6 and obtain coefficients
of the regression between WPE and prediction error only for that choice and for
their particular signals. However, other systems and other choices of L may
easily lead to other regression coefficients so that deciding about the best
possible prediction error from WPE seems impossible in general. It is good if
such a decision is possible for a certain narrow class of systems. However, the
authors do not even discuss to which class of systems their results apply.}

\emph{II) PE is invariant under an invertible nonlinear change of variables. The
prediction error measured as the authors suggest is not invariant. Thus, such a
simple change of variables can lead, at least, to a change in regression
coefficient between WPE and prediction error. Thus, the authors should also
discuss to which variable for that class of systems their result apply. I think
that the fact that the authors observe a certain regression line at all seems to
be due to small number of different signals they consider (only 7 or 8 different
data points, such a small number of points may lie approximately near a straight
line by chance).}

\emph{To summarize, the paper shows that the authors are well familiar with the
entire topic and highly qualified, they overview some well-known facts (WPE,
various prediction techniques, etc) and ``invent'' an unexpected test system
(computer performance data), they report about their attempts to play around
with all that staff. However, their results do not seem to provide any new
information to a specialist. Probably, this work can be continued to get some
more reliable results about concrete relationship between WPE and prediction
error for certain classes of systems (however, I doubt that it is possible to
derive any simple and widely applicable relationship like the linear function
suggested by the authors). Seemingly, it should then be a new paper which can be
hardly considered as a revised version of the present manuscript.}

\end{document}

### 2014-05-30 response.tex
put a few responses in. things in red have not actually been done yet

@@
 \usepackage{amsmath}
+\usepackage{xcolor}
+
+\newcommand{\alert}[1]{{\color{red}#1}}
 
@@
 
+We thank the reviewer for their kind words and appreciate the time they put in
+to reviewing our work.
+
 \noindent\emph{Major issues:}
@@
 
+While we recognize the drawbacks inherent in using the industry-standard error
+metric MASE, we chose it over other error metrics due to some of its favorable
+qualities; namely that it is scale-free and comparable across different time
+series. In our opinion this made it the superior candidate despite the bias it
+can introduce.
+
 \emph{In many parts the article reads almost like a review. There is a lot of
@@
 loss of information.}
+
+Due to the broad selection of fields that this work draws upon, we thought it
+necessary to provide background for each of them in case the reader is not
+familiar with some subset of them. While this necessitated some redundancy,
+\alert{we have made an effort to trim the fat and make the review portions more
+streamlined.}
 
@@
 
-\emph{Page 10: `error between the predictions and the true continuations' -->
-`error of the predictions with respect to the true continuations' Figure 6 on
-Page 11: It would be more intuitive to turn this into a `table of figures'
-with the four methods as row labels on the left and the three systems as column
-labels at the top. Also: Caption `for forecast of ... and all four prediction
-strategies'.}
+Our intent in mentioning this theoretical result was to side-step the possible
+response of simply ``always use the most sophisticated prediction scheme''. This
+corollary implies that no matter how sophisticated and broad a scheme we
+construct, it will not be the optimal one for some time series. Thus there is a
+real need for determining when a better scheme exists. \alert{We have modified
+the paragraph to make this point more clear.}
+
+\emph{Page 10: `error between the predictions and the true continuations'
+$\rightarrow$ `error of the predictions with respect to the true continuations'
+Figure 6 on Page 11: It would be more intuitive to turn this into a `table of
+figures' with the four methods as row labels on the left and the three systems
+as column labels at the top. Also: Caption `for forecast of ... and all four
+prediction strategies'.}
+
+We have modified the text as the reviewer suggested, and agree that it reads
+more clearly now. \alert{We have also modified the figure as suggested.}
 

### 2014-05-30 model.tex
beating this thing into PRE shape...
put a few responses in. things in red have not actually been done yet

--- 2014-04-25 model.tex
@@
 \begin{subfigure}{\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
+    \includegraphics[width=\columnwidth]{figs/colipc3d}
     \caption{\col }
@@
     \begin{subfigure}{\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gcc3dipc.png}
+    \includegraphics[width=\columnwidth]{figs/gcc3dipc}
     \caption{\gcc}
@@
   \begin{subfigure}{\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svd53dipc2.png}
+    \includegraphics[width=\columnwidth]{figs/svd53dipc2}
     \caption{\svdfive}
@@
 
-     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
+     %\includegraphics[width=\textwidth]{colipc3d}
      \caption{3D projections of delay-coordinate embeddings of the
@@
 traces described in Section~\ref{sec:methods}, then calculate the
-error between the predictions and the true continuations.
+error of the predictions with respect to the true continuations.
 Specifically, we split each time series into two pieces: the first
@@
   \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colRWForecast.png}
+    \includegraphics[width=\columnwidth]{figs/colRWForecast}
     \caption{\col\\ random walk }
@@
    \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccRWForecast.png}
+    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
     \caption{\gcc\\ random walk }
@@
      \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast.png}
+    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
     \caption{\svdfive\\ random walk}
@@
       \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colMeanForecast.png}
+    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
     \caption{\col\\ na\"ive }
@@
    \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccMeanForecast.png}
+    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
     \caption{\gcc\\ na\"ive }
@@
      \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast.png}
+    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
     \caption{\svdfive\\ na\"ive }
@@
     \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colARIMAForecast.png}
+    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
     \caption{\col \\ ARIMA}
@@
   \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast.png}
+    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
     \caption{\gcc \\ ARIMA }
@@
   \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast.png}
+    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
     \caption{\svdfive \\ ARIMA }
@@
       \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
+    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
     \caption{\col \\ LMA}
@@
       \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
+    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
     \caption{\gcc \\ LMA}
@@
         \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
+    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
     \caption{\svdfive \\ LMA}
@@
   %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
+  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
    \caption{Predicted ($p_j$) versus true values ($c_j$) for
-     forecasts of \col, \gcc, and \svdfive.  
+     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
 % On this type of plot, a perfect prediction lies exactly on the
@@
 \begin{table*}
+\caption{Mean
+absolute scaled error (MASE) scores and weighted permutation entropies for all eight
+   processes studied in this paper}
   \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}
-  \hline Signal & LMA  & ARIMA & na\"{i}ve  & Weighted Permutation \\ 
-  &  MASE & MASE  & MASE &  Entropy \\ \hline
+  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
+  \hline\hline Signal & LMA MASE  & ARIMA MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
+  \hline
  \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\
@@
  \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
-  \hline
-  \end{tabular}
+  \hline\hline
+  \end{tabular*}
   \end{center}
  \label{default}
- \caption{Mean
-absolute scaled error (MASE) scores and weighted permutation entropies for all eight
-   processes studied in this paper}
  \label{tab:error}
@@
 %\begin{subfigure}{\columnwidth}
-%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
+%    \includegraphics[width=\columnwidth]{%colPredShortTS}
 %    \caption{\col }
@@
 %    \begin{subfigure}{\columnwidth}
-%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
+%    \includegraphics[width=\columnwidth]{%colPredShortTS}
 %    \caption{\gcc}
@@
 %  \begin{subfigure}{\columnwidth}
-%%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
+%%    \includegraphics[width=\columnwidth]{%colPredShortTS}
 %    \caption{\svdfive}
@@
 %  \centering
-%   \includegraphics[width=\textwidth]{figs/colPredShortTS}
+%   \includegraphics[width=\textwidth]{colPredShortTS}
 %    \caption{A forecast of the last 4,000 points of the %signal in
@@
 %  \centering
-%    \includegraphics[width=\textwidth]{figs/colPredShortTS}
+%    \includegraphics[width=\textwidth]{colPredShortTS}
 %     \caption{{\color{red} [[actually need to generate this %figure]]}An LMA-based forecast of the last 4,000 points of a

### 2014-06-17 response.tex
fix the lower limit of a sum
address some stuff

@@
 
-While we recognize the drawbacks inherent in using the industry-standard error
-metric MASE, we chose it over other error metrics due to some of its favorable
+While we recognize the drawbacks inherent in using the standard error metric
+MASE, we chose it over other error metrics due to some of its favorable
 qualities; namely that it is scale-free and comparable across different time
 series. In our opinion this made it the superior candidate despite the bias it
-can introduce.
+can introduce. For a complete comparison of MASE relative to other error
+metrics, please see the citation contained within our paper.
 
@@
 
+Thank you for catching our mistake, we have corrected the equation.
+
 \emph{Page 19: ``There are some exceptions: [...] Conversely, forecast errors
@@
 
+You are absolutely correct, and we've adjusted the text accordingly.
+
 \section*{Response to the Second Referee}
+
+We thank the reviewer for taking the time to both read our paper and provide us
+with valuable feedback. While we agree with several of the reviewers criticisms,
+and have taken steps to address them, we disagree that doing so changes the
+nature of the paper to the point of requiring a new submission.
 
@@
 
+We have modified the introduction to make our goals more clearly defined.
+
 \emph{I think that the possibility to estimate the best prediction error in
@@
 
+As we stated on page 19 of the manuscript, we believe that due to the broad
+spectrum of behaviors exhibited by the computer performance data these results
+will hold for time series drawn from other physical systems. We are aware that
+this claim may not be true, and believe that to be critical follow up work.
+However, we have included here some preliminary results that support our claim
+by analyzing a classical random walk, three time series derived from the classic
+Lorenz attractor, and the time series used in the SFI prediction competition.
+Though we are not prepared to fully discuss or publish the results regarding
+these systems, we hope that the preliminary findings are enough to justify our
+claim for the reviewer.
+
 \emph{1) Introduction is too long, but the main problem is not formulated
@@
 manuscript do not seem important enough to be published in a separate article.}
+
+We have modified the manuscript to make it more clear that our goal is to derive
+a criterion for best-case prediction error in a model-free way. We have also
+shortened it.
 
@@
 
+While we agree that on the surface these claims are both trivial and reasonably
+well known in theory and for toy models, our specific intent is that these also
+hold and can be practically applied via the WPE in real-world systems with
+potentially both dynamical and observational noise. We also greatly appreciated
+being informed about Boffetta et al's paper, and have included a citation to it
+in our paper.
+
 \emph{3) By the way, the authors call ``fully complex'' the white noise process,
@@
 the latter plays an important role in many formulations of the authors.}
+
+We have added a short paragraph discussing the differences in notions of
+complexity, and clarified our working definition of it as used for this paper.
 
@@
 
+It is indeed true that the way we worded this finding makes it sound trivial.
+Our intent, as we have modified the text to reflect, is that the appropriateness
+of a predictor can be discovered empirically via comparison with the WPE.
+\alert{i'm not entirely sure how to respond to this one correctly.}
+
 \emph{5) The benchmark system - a computer as an object of prediction - seems to
@@
 
+As mentioned earlier in our response, we have included preliminary results
+regarding some simple systems here for the reviewer, and indeed the well-known
+results relating entropy and predictability hold for these systems. Because
+these results are not surprising, but rather simply confirming what is known in
+theory, we did not want to include them in the paper but rather focus on how
+these results still hold in the face of very ``dirty'' real-world data.
+
 \emph{6) The description of the choice of the benchmark system in the
@@
 might well be of interest.}
+
+Since our primary goal with this paper is the relationship between prediction
+error and WPE in noisy, real-world systems, and the practical applicability of
+this relationship, we have reduced the amount of space spent in the introduction
+discussing the particular time series we use.
 
@@
 
+It is certainly true that the naive and random-walk predictors can be seen as
+subclasses of the ARIMA model. We have explicitly included them, however, since
+they are used (sometimes to great success) in practice. Further, we have used
+standard methods \alert{josh -- should we add some names here?} of fitting the
+parameters used in our ARIMA predictions, and we believe any inefficiencies of
+that parameter selection support our findings --- namely, if a particular
+selection of model parameters predicts poorly (compared to what the WPE value
+might suggest), it is an indication that the specific model, including
+parameters, is not optimal for the job. This might mean that a different
+parameter selection scheme is needed, or that the entire model is inappropriate.
+Unfortunately the WPE value alone is insufficient to differentiate these two
+situations.
+
 \emph{8) Conditions for inefficiency of simple prediction schemes (like naive
@@
 
+The purpose of this short paragraph in this paper, which explains the
+efficiencies and inefficiencies of these simple prediction schemes, is present
+for two purposes. First, to explain why we have included these simple schemes
+rather than just the more complicated ones, and second to help in explaining the
+specific results obtained later in the paper.
+
 \emph{9) A better justification of the prediction error metrics could be also in
@@
 account the following major concerns.}
+
+We have strengthened our justification for the use of the MASE metric. The work
+we have cited which introduced MASE extensively compares it to other schemes and
+highlights the differences obtained therein. We further believe that a
+comparison of different prediction error metrics as they relate to WPE is best
+left as a sequel to this work.
 
@@
 
+It is certainly true that the specific relationship between WPE and MASE values
+is dependent on word length, though for any reasonable choice of $\ell$
+(discussion regarding reasonable values in the manuscript) we believe a similar
+relationship to hold. We also believe, due to the spectrum of time series
+behaviors we have included, that the curve is not highly dependent on which
+systems are considered. As stated before, we have hopefully assuaged the
+reviewers concern about this point by including here the WPE and MASE values for
+a variety of other time series, including a random walk, simple chaotic systems,
+as time series used in the SFI prediction contest.
+
 \emph{II) PE is invariant under an invertible nonlinear change of variables. The
@@
 line by chance).}
+
+Both WPE and MASE are sensitive to changes of variables, and from some initial
+testing they are so in complementary ways. This leads a change in variables
+to...  We have also redone our regression line to include all 105 WPE-MASE
+pairs, rather than just a representative from each program.
 
@@
 
+It is our hope that between the additional preliminary results here, the
+modifications to the paper, as well as our clarifying the goals of our paper
+(understandably misunderstood in the draft), that the reviewer now considers the
+manuscript appropriate for publication in PRE.
+
 \end{document}

### 2014-06-17 results.tex
address some stuff

@@
 
-Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these 45
-predictions for each system: that is, the lowest error over all 15
-trials and all three methods for each of the eight programs.  The WPE
-is plotted against the corresponding MASE value here in order to bring
-out the correlation between these two quantities.
+Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these predictions for
+each trial of each system: that is, the lowest error over all three methods for
+each of the eight programs.  The WPE is plotted against the corresponding MASE
+value here in order to bring out the correlation between these two quantities.
 \begin{figure}[htbp]
   \centering
-  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
+  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy}
   \caption{Weighted permutation entropy vs. mean absolute scaled error
-    (MASE) of the best prediction of each system.
+    (MASE) of the best prediction of each trial for each system.
 %
@@
 %
-The dashed line is a least-squares linear fit of all the points except
-{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
+The dashed line is a least-squares linear fit of all the points except those
+from {\tt dgesdd$_1$}, which we have excluded for reasons explained in the
 text.}

### 2014-06-17 response.tex
some tweaks

@@
 We have modified the text as the reviewer suggested, and agree that it reads
-more clearly now. \alert{We have also modified the figure as suggested.}
+more clearly now. We have also modified the figure as suggested.
 
@@
 
-We have modified the introduction to make our goals more clearly defined.
+\alert{We have modified the introduction to make our goals more clearly
+defined.}
 
@@
 this claim may not be true, and believe that to be critical follow up work.
-However, we have included here some preliminary results that support our claim
-by analyzing a classical random walk, three time series derived from the classic
-Lorenz attractor, and the time series used in the SFI prediction competition.
-Though we are not prepared to fully discuss or publish the results regarding
-these systems, we hope that the preliminary findings are enough to justify our
-claim for the reviewer.
+\alert{However, we have included here some preliminary results that support our
+claim by analyzing a classical random walk, three time series derived from the
+classic Lorenz attractor, and the time series used in the SFI prediction
+competition.} Though we are not prepared to fully discuss or publish the results
+regarding these systems, we hope that the preliminary findings are enough to
+justify our claim for the reviewer.
 
@@
 
-We have modified the manuscript to make it more clear that our goal is to derive
-a criterion for best-case prediction error in a model-free way. We have also
-shortened it.
+\alert{We have modified the manuscript to make it more clear that our goal is to
+derive a criterion for best-case prediction error in a model-free way. We have
+also shortened it.}
 
@@
 potentially both dynamical and observational noise. We also greatly appreciated
-being informed about Boffetta et al's paper, and have included a citation to it
-in our paper.
+being informed about Boffetta et al's paper, \alert{and have included a citation
+to it in our paper.}
 
@@
 
-We have added a short paragraph discussing the differences in notions of
-complexity, and clarified our working definition of it as used for this paper.
+\alert{We have added a short paragraph discussing the differences in notions of
+complexity, and clarified our working definition of it as used for this paper.}
 

### 2014-06-17 concl.tex

@@
 
-In this paper, we have argued that redundancy~\cite{crutchfield2003}
-is a useful definition of the inherent predictability of a time
-series.  To operationalize that definition, we use an approximation of
-the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
-version of the permutation entropy of~\cite{bandt2002per}.  This WPE
-technique---an ordinal calculation of forward information transfer in
-a time series---is ideal for our purposes because it works with
-real-valued data and is known to converge to the true entropy value.
-Using a variety of forecast models and more than a hundred time-series
-data sets from a computer-performance experiment, we have shown that
-prediction accuracy is indeed correlated with weighted permutation
-entropy: the higher the WPE, in general, the higher the prediction
-error.  There are some exceptions: \svdone is easier to predict than
-its WPE value suggests using any of the three forecast strategies
-studied here.  Conversely, forecast errors for the \col and \svdfive
-signals are higher than the corresponding WPEs suggest---except for
-the nonlinear LMA predictor.  On the whole, though, signals with
-higher weighted permutation entropy are harder to predict.
+In this paper, we have argued that redundancy~\cite{crutchfield2003} is a useful
+definition of the inherent predictability of a time series.  To operationalize
+that definition, we use an approximation of the Kolmogorov-Sinai
+entropy~\cite{lind95}, estimated using a weighted version of the permutation
+entropy of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation of
+forward information transfer in a time series---is ideal for our purposes
+because it works with real-valued data and is known to converge to the true
+entropy value. Using a variety of forecast models and more than a hundred time-
+series data sets from a computer-performance experiment, we have shown that
+prediction accuracy is indeed correlated with weighted permutation entropy: the
+higher the WPE, in general, the higher the prediction error.  Supporting our
+hypothesis, forecast errors for the \col and \svdfive signals are higher than
+the corresponding WPEs suggest---except for the nonlinear LMA predictor.  There
+are some exceptions: \svdone is easier to predict than its WPE value suggests
+using any of the three forecast strategies studied here.  On the whole, though,
+signals with higher weighted permutation entropy are harder to predict.
 

### 2014-06-17 model.tex

@@
 \end{figure*}
+
+\begin{table*}
+  \renewcommand{\arraystretch}{1.25}
+  \begin{tabular}{lccc}
+    & {\large \col} & {\large \gcc} & {\large \svdfive} \\
+    & & & \\
+    \begin{sideways}\large random walk\end{sideways} &
+    \includegraphics[width=0.6\columnwidth]{figs/colRWForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/gccRWForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/svdfiveRWForecast} \\
+    \begin{sideways}\large na\"ive\end{sideways} &
+    \includegraphics[width=0.6\columnwidth]{figs/colMeanForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/gccMeanForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
+    \begin{sideways}\large ARIMA\end{sideways} &
+    \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/gccARIMAForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/svdfiveARIMAForecast} \\
+    \begin{sideways}\large LMA\end{sideways} &
+    \includegraphics[width=0.6\columnwidth]{figs/colLMAForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/gccLMAForecast} &
+    \includegraphics[width=0.6\columnwidth]{figs/svdfiveLMAForecast}
+  \label{fig:forecast-example}
+  \end{tabular}
+  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts of \col,
+     \gcc, and \svdfive and all four prediction strategies.
+  }
+  \label{fig:forecast-example}
+\end{table*}
+
 In these images, the vertical axis is the prediction $p_j$ and the

### 2014-06-27 response.tex
more updates

@@
 \usepackage{xcolor}
+\usepackage{graphicx}
 
@@
 this claim may not be true, and believe that to be critical follow up work.
-\alert{However, we have included here some preliminary results that support our
-claim by analyzing a classical random walk, three time series derived from the
-classic Lorenz attractor, and the time series used in the SFI prediction
-competition.} Though we are not prepared to fully discuss or publish the results
-regarding these systems, we hope that the preliminary findings are enough to
-justify our claim for the reviewer.
+However, we have included here some preliminary results that support our claim
+by analyzing a random walk, the Henon map, the Lorenz attractor, and a time
+series of voltages from a chaotic laser used in the SFI prediction competition.
+Though we are not prepared to fully discuss or publish the results regarding
+these systems, we hope that the preliminary findings are enough to support our
+claim of generality. The fit in this plot, of the form $y = a \log(b x + c)$, is
+based solely on the computer performance data. Despite this, the four
+supplemental time series fall within a narrow band around the fit. The use of
+this log fit --- now used in the paper instead of the linear one --- is also
+theoretically advantageous because any signal with zero entropy should be
+capable of being predicted perfectly and so should ideally have a MASE value of
+zero.
+
+\begin{center}
+    \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_extra_autolog}
+\end{center}
 
@@
 potentially both dynamical and observational noise. We also greatly appreciated
-being informed about Boffetta et al's paper, \alert{and have included a citation
-to it in our paper.}
+being informed about Boffetta et al's paper, and have included a citation to it
+in our paper.
 
@@
 
-\alert{We have added a short paragraph discussing the differences in notions of
-complexity, and clarified our working definition of it as used for this paper.}
+We have added a short discussion about the differences in notions of complexity,
+and clarified our working definition of it as used for this paper.
 

### 2014-06-27 intro.tex
beating this thing into PRE shape...
address some stuff
more updates

@@
 
-Complicated time-series data are ubiquitous in modern scientific
-research.  The complexity of these data spans a wide range.  On the
-low end of this spectrum are time series that exhibit perfect
-predictive structure, i.e, signals whose future values can be
-successfully predicted from past values.  Signals like this can be
-viewed as the product of an underlying process that generates
-information and/or transmits it from the past to the future in a
-perfectly predictable fashion.  Constant or periodic signals, for
-example, fall in this class.  On the opposite end of this spectrum are
-signals that are what we call \emph{fully complex}, where the
-underlying generating process transmits no information at all from the
-past to the future.  White noise and random-walk processes fall in
-this class.  In fully complex signals, knowledge of the past gives no
-insight into the future, regardless of what model one chooses to use.
-Signals in the midrange of this spectrum---e.g., deterministic
-chaos---pose interesting challenges from a modeling perspective.  In
-these signals, enough information is being transmitted from the past
-to the future that an \emph{ideal} model---one that captures the
-generating process---can forecast the future behavior of the observed
-system with high accuracy.
+Complicated time-series data are ubiquitous in modern scientific research.  The
+complexity of these data spans a wide range.  On the low end of this spectrum
+are time series that exhibit perfect predictive structure, i.e, signals whose
+future values can be successfully predicted from past values.  Signals like this
+can be viewed as the product of an underlying process that generates information
+and/or transmits it from the past to the future in a perfectly predictable
+fashion.  Constant or periodic signals, for example, fall in this class.  On the
+opposite end of this spectrum are signals that are what we call \emph{fully
+complex}, where the underlying generating process transmits no information at
+all from the past to the future.  White noise processes fall in this class.  In
+fully complex signals, knowledge of the past gives no insight into the future,
+regardless of what model one chooses to use. Signals in the midrange of this
+spectrum---e.g., deterministic chaos---pose interesting challenges from a
+modeling perspective.  In these signals, enough information is being transmitted
+from the past to the future that an \emph{ideal} model---one that captures the
+generating process---can forecast the future behavior of the observed system
+with high accuracy. Though this notion of complexity differs from that
+considered by e.g. \cite{Shalizi2008}, which would consider a time series
+without any statistical regularities to be non-complex, it has been extensively
+considered by Boffetta et al~\cite{boffetta02}.
 

### 2014-07-14 intro.tex
first round of LizB edits to revisions and response

@@
 
-Complicated time-series data are ubiquitous in modern scientific research.  The
-complexity of these data spans a wide range.  On the low end of this spectrum
-are time series that exhibit perfect predictive structure, i.e, signals whose
-future values can be successfully predicted from past values.  Signals like this
-can be viewed as the product of an underlying process that generates information
-and/or transmits it from the past to the future in a perfectly predictable
-fashion.  Constant or periodic signals, for example, fall in this class.  On the
-opposite end of this spectrum are signals that are what we call \emph{fully
-complex}, where the underlying generating process transmits no information at
-all from the past to the future.  White noise processes fall in this class.  In
-fully complex signals, knowledge of the past gives no insight into the future,
-regardless of what model one chooses to use. Signals in the midrange of this
-spectrum---e.g., deterministic chaos---pose interesting challenges from a
-modeling perspective.  In these signals, enough information is being transmitted
-from the past to the future that an \emph{ideal} model---one that captures the
-generating process---can forecast the future behavior of the observed system
-with high accuracy. Though this notion of complexity differs from that
-considered by e.g. \cite{Shalizi2008}, which would consider a time series
-without any statistical regularities to be non-complex, it has been extensively
-considered by Boffetta et al~\cite{boffetta02}.
+Complicated time-series data are ubiquitous in modern scientific
+research.  The complexity of these data spans a wide range.  On the
+low end of this spectrum are time series that exhibit perfect
+predictive structure, i.e, signals whose future values can be
+successfully predicted from past values.  Signals like this can be
+viewed as the product of an underlying process that generates
+information and/or transmits it from the past to the future in a
+perfectly predictable fashion.  Constant or periodic signals, for
+example, fall in this class.  On the opposite end of this spectrum are
+signals that are what one can call \emph{fully complex}, where the
+underlying generating process transmits no information at all from the
+past to the future.  White noise processes fall in this class.  In
+fully complex signals, knowledge of the past gives no insight into the
+future, regardless of what model one chooses to use. Signals in the
+midrange of this spectrum---e.g., deterministic chaos---pose
+interesting challenges from a modeling perspective.  In these signals,
+enough information is being transmitted from the past to the future
+that an \emph{ideal} model---one that captures the generating
+process---can forecast the future behavior of the observed system with
+high accuracy.
 
-As a corollary of the undecidability of the halting
-problem~\cite{halting-problem}, however, no single forecasting schema
-is ideal for all noise-free deterministic
-signals~\cite{weigend-book}---let alone all real-world time-series
-data sets.  This leads naturally to an important and challenging
-question: given a complicated real-valued time series, does there
-exist a forecast model that can leverage the information (if any) that
-is being transmitted from past to future by the underlying generating
-process?  A first step in answering this question is to reliably
-quantify where on the complexity spectrum a given time series falls; a
-second step is to determine how complexity and predictability are
-related.  These are the goals of this paper.
+{\color{red} Though this notion of complexity differs from that
+  considered by e.g. \cite{Shalizi2008}, which would consider a time
+  series without any statistical regularities to be non-complex, it
+  has been extensively considered by Boffetta et al~\cite{boffetta02}.
+  I still don't understand how this sentence both addresses the
+  reviewer's pointing us to this paper and flows, logically.  Either
+  way, it is out of place.  It should be in a part of this paper that
+  gives citations, not one that lays out context in a general manner,
+  without citations.}
 
-In particular, we wish to determine---without knowing anything about
-the generating process---whether a time series is too complex to successfully
-predict.  An important practical corollary to this is a strategy for
-assessing appropriateness of forecast methods.  If the forecast
-produced by a particular method is poor but the time series contains a
-significant amount of predictive structure, one can reasonably conclude that that method is inadequate
-to the task and that one should seek another method.
+% As a corollary of the undecidability of the halting
+% problem~\cite{halting-problem}, however, no single forecasting schema
+% is ideal for all noise-free deterministic
+% signals~\cite{weigend-book}---let alone all real-world time-series
+% data sets.
+
+This leads naturally to an important and challenging question: given a
+noisy real-valued empirical time series, does there exist a forecast
+model that can leverage the information (if any) that is being
+transmitted from past to future by the underlying generating process?
+A first step in answering this question is to reliably quantify where
+on the complexity spectrum a given time series falls; a second step is
+to determine how complexity and predictability are related in these
+kinds of data sets.  With these answers in hand, one can develop a
+practical strategy for assessing appropriateness of forecast methods
+for a given time series.  If the forecast produced by a particular
+method is poor, for instance, but the time series contains a
+significant amount of predictive structure, one can reasonably
+conclude that that method is inadequate to the task and that one
+should seek another method.  The goal of this paper is to explore and
+formalize that strategy.
+
+% We have two goals in this paper: to determine---without knowing
+% anything about the generating process---whether an empirical time
+% series is too complex to successfully predict.
 
@@
 %
-We make no assumptions about the properties of the generating process:
-whether it is linear, nonlinear, deterministic, stochastic,
-stationary, non-stationary, etc.  To explore the relationship between
-complexity, predictive structure, and predictability, we generate
-forecasts for a variety of time-series datasets using four different
-prediction methods, then compare the accuracy of those predictions to
-the permutation entropy of the associated signals.  This results in
-two primary findings:
+We do not make any knowledge of the generating process or its
+properties: whether it is linear, nonlinear, deterministic,
+stochastic, stationary, non-stationary, etc.  To explore the
+relationship between complexity, predictive structure, and
+predictability, we generate forecasts for a variety of time-series
+datasets using four different prediction methods, then compare the
+accuracy of those predictions to the permutation entropy of the
+associated signals.  This results in two primary findings:
 \begin{enumerate}
@@
   quantifiable by permutation entropy and \emph{(ii)} correlated with
-  prediction accuracy of an appropriate predictor.
+  the accuracy of an appropriate predictor.
 \item The way information is generated and processed internally by a
@@
   schema---and in the choice of which one is appropriate for a given
-time series.
+  empirical time series.
 \end{enumerate}
+There has, of course, been a great deal of good work on different ways
+to measure the complexity of data, and previous explorations have
+confirmed repeatedly that complexity is a challenge to prediction, but
+this collection of issues has not been properly explored in the
+context of noisy, poorly sampled, real-world data from unknown
+systems.  That is the contribution of this paper.
+
 The forecast methods used in this study were chosen to be a
@@
 For the purposes of this study, we require a broad array of
-time-series datasets from across the complexity spectrum.  Since
-forecasting is a real-world problem, we chose to study sensor data
-from a computer-performance experiment.  While this is not a common
-laboratory experiment, it is a highly appropriate choice here.
-Computers are extremely complicated systems.  Modern microprocessor
-chips contain multiple processing units and multi-layer memories, for
-instance, and they use complicated hardware/software strategies to
-maximize performance by moving data and threads of computation across
-those resources.  As a result, the processor and memory loads during
-the execution of even a very simple program can evolve in a very
-complicated fashion.  Figure~\ref{fig:col-ipc} shows a performance
-trace of a short program that repeatedly initializes the upper
-triangle of a matrix in column-major order.
+time-series datasets from across the complexity spectrum.  We chose to
+study sensor data from a computer-performance experiment.  While this
+is not a common laboratory experiment, it is a highly appropriate
+choice here.  Computers are extremely complicated systems and their
+dynamics is surprisingly rich.
+% Modern microprocessor chips contain multiple processing units and
+% multi-layer memories, for instance, and they use complicated
+% hardware/software strategies to maximize performance by moving data
+% and threads of computation across those resources.  As a result, the
+The processor and memory loads during the execution of even a very
+simple program can exhibit dynamical chaos, for
+instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
+example: a performance trace of a short program that repeatedly
+initializes the upper triangle of a matrix in column-major order.
 %
@@
 %
-Although this time series appears to be periodic, it is actually
-chaotic~\cite{mytkowicz09}, which places fundamental limits on
-predictability.  Changes in the code can radically change this
-behavior; indeed, many performance traces exhibit interesting regime
-changes as a program moves through the different phases of its
-operation.  The resulting traces span the whole range of the
-complexity spectrum, from completely predictable to completely
-unstructured, making them an ideal testbed for this
-study\footnote{Predicting the \emph{state} of a computer, of course,
-  would amount to solving the halting problem.  What we are doing here
-  is predicting computer \emph{performance}, which does not violate
-  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
+A small change in the code can cause the dynamics to bifurcate to a
+periodic regime.  
+% 
+% Indeed, many computer performance traces exhibit
+% interesting regime changes as a program moves through the different
+% phases of its operation.
+By running different programs on the same computer, then, we can
+produce traces that span the whole range of the complexity spectrum,
+from completely predictable to completely unstructured---which makes
+this an ideal testbed for this study\footnote{Predicting the
+  \emph{state} of a computer, of course, would amount to solving the
+  halting problem.  What we are doing here is predicting computer
+  \emph{performance}, which does not violate the Rice-Shapiro
+  theorem~\cite{hopcroft2007}.}.
 % footnote to appease the theoretician in Josh
@@
 series. Section~\ref{sec:model} describes the prediction models used
-in this study.  In Section~\ref{sec:meaComplex}, we review permutation
+in this study.  Section~\ref{sec:meaComplex} reviews permutation
 entropy, the technique that we use to measure complexity.  In
-Section~\ref{sec:results}, we estimate the complexity of each time
-series and compare that complexity to the accuracy of predictions
-produced by the methods of Section~\ref{sec:model}, operating on that
-time series.  In Section~\ref{sec:conc}, we discuss these results and
-their implications, and consider future areas of research.
+Section~\ref{sec:results}, we estimate the complexity of each
+empirical time series and compare that complexity to the accuracy of
+predictions produced by the methods of Section~\ref{sec:model},
+operating on that time series.  In Section~\ref{sec:conc}, we discuss
+these results and their implications, and consider future areas of
+research.

### 2014-07-14 related.tex
first round of LizB edits to revisions and response

@@
 for a wide variety of prediction tasks.  The purpose of this paper is
-not to add a new weapon to this arsenal, nor to assess or compare the
-effectiveness of existing methods.  Our goals are more general:
-\emph{(i)} to empirically quantify the predictive structure that is
-present in a real-valued scalar time series and \emph{(ii)} to explore
-how the performance of prediction methods is related to that inherent
+not to add a new weapon to this arsenal, nor to do any sort of
+theoretical assessment or comparison of existing methods.  Our goals
+are focused more on the \emph{practice} of predication: \emph{(i)} to
+empirically quantify the predictive structure that is present in a
+real-valued scalar time series and \emph{(ii)} to explore how the
+performance of prediction methods is related to that inherent
 complexity.  It would, of course, be neither practical nor interesting
@@
 moves away from those restrictions.  Our approach falls into this
-class.  Our goal is to empirically measure the inherent complexity of
-a time series, then correlate that complexity with the predictive
+class.  Our goal is to measure the inherent complexity of an empirical
+time series, then correlate that complexity with the predictive
 accuracy of forecasts made using a number of different methods.
@@
 We sidestep these issues by using a variant of the \emph{permutation
-  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate the value
-of the Kolmogorov-Sinai entropy of a real-valued time series---and
-thus the redundancy in that data, which we show to be an effective proxy for
-predictability.  This differs from existing approaches in a number of
-ways.  It does not rely on generating partitions---and thus does not
-introduce bias into the results if one does not know the dynamics or
-cannot compute the partition.  It makes no assumptions about, and
+  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate the
+value of the Kolmogorov-Sinai entropy of a real-valued time
+series---and thus the redundancy in that data, which our results
+confirm to be an effective proxy for predictability.  This differs
+from existing approaches in a number of ways.  It does not rely on
+generating partitions---and thus does not introduce bias into the
+results if one does not know the dynamics or cannot compute the
+partition.  Permutation entropy makes no assumptions about, and
 requires no knowledge of, the underlying generating process: linear,

### 2014-07-14 intro.tex
fixing biblio

--- 2014-06-27 intro.tex
@@
 is ideal for all noise-free deterministic
-signals~\cite{weigend-book}---let alone all real-world time-series
+signals~\cite{weigend93}---let alone all real-world time-series
 data sets.  This leads naturally to an important and challenging

### 2014-07-14 related.tex
fixing biblio

--- 2014-04-25 related.tex
@@
 A practice known as deterministic vs. stochastic
-modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
+modeling~\cite{weigend93, Casdagli92dvsplots} bridges the gap
 between local and global approaches to model-based error analysis.
@@
 and thus they can be perfectly predicted if observed for long 
-enough~\cite{weigend-book}.  In practice, it is
+enough~\cite{weigend93}.  In practice, it is
 quite difficult to estimate the redundancy of an arbitrary,

### 2014-07-14 response.tex

--- 2014-06-17 response.tex
@@
 
-\emph{In this manuscript the authors present a way to explore the relationship
-between predictability and complexity across a broad (but not complete) array of
-forecast strategies such as random walk (last value), na\"ive (average value),
-regression based (ARIMA) and a nonlinear method based on state space
-reconstruction (LMA). They use weighted permutation entropy as a criterion/index
-of predictive structure in the time series and illustrate their approach on
-eight real-world datasets derived from three different systems (computer
+\emph{In this manuscript the authors present a way to explore the
+  relationship between predictability and complexity across a broad
+  (but not complete) array of forecast strategies such as random walk
+  (last value), na\"ive (average value), regression based (ARIMA) and
+  a nonlinear method based on state space reconstruction (LMA). They
+  use weighted permutation entropy as a criterion/index of predictive
+  structure in the time series and illustrate their approach on eight
+  real-world datasets derived from three different systems (computer
   performance traces).}
 
-\emph{Overall the manuscript is well written and clearly understandable. The
-analysis seems to be carried out carefully and the conclusions drawn are quite
-reasonable. While I don't get the impression that this is a breakthrough paper I
-can still recommend the manuscript for publication in PRE if the issues listed
-below are addressed thoroughly.}
+\emph{Overall the manuscript is well written and clearly
+  understandable. The analysis seems to be carried out carefully and
+  the conclusions drawn are quite reasonable. While I don't get the
+  impression that this is a breakthrough paper I can still recommend
+  the manuscript for publication in PRE if the issues listed below are
+  addressed thoroughly.}
 
-We thank the reviewer for their kind words and appreciate the time they put in
-to reviewing our work.
+We thank the reviewer for their kind words and appreciate the time
+they put into reviewing our work.
 
@@
 
-\emph{I don't get the reason for normalizing the three methods by the random
-walk method (for example in Table I). As the authors write themselves this makes
-the indicator vulnerable to a bias caused by the high influence of this
-(arbitrarily selected) reference method. Why not just use a non-referential
-estimator, which then would also allow to compare the inherent predictability of
-different systems (as it is now the values obtained for each systems are
-strongly influenced by the random walk estimate on that particular system and
-thus can not be compared). This will also change the results of the fitting (the
-authors explicitly exclude one value because of the effect caused by the
-normalization). And it is not that the random walk predictor is the one standard
-method that all other methods have to compete against.}
+\emph{I don't get the reason for normalizing the three methods by the
+  random walk method (for example in Table I). As the authors write
+  themselves this makes the indicator vulnerable to a bias caused by
+  the high influence of this (arbitrarily selected) reference
+  method. Why not just use a non-referential estimator, which then
+  would also allow to compare the inherent predictability of different
+  systems (as it is now the values obtained for each systems are
+  strongly influenced by the random walk estimate on that particular
+  system and thus can not be compared). This will also change the
+  results of the fitting (the authors explicitly exclude one value
+  because of the effect caused by the normalization). And it is not
+  that the random walk predictor is the one standard method that all
+  other methods have to compete against.}
 
-While we recognize the drawbacks inherent in using the standard error metric
-MASE, we chose it over other error metrics due to some of its favorable
-qualities; namely that it is scale-free and comparable across different time
-series. In our opinion this made it the superior candidate despite the bias it
-can introduce. For a complete comparison of MASE relative to other error
-metrics, please see the citation contained within our paper.
+No error metric is perfect.  For our purposes, we needed one that was
+scale free and comparable across different time series.  While MASE
+has its drawbacks, it was a very good fit to those requirements.  In
+our opinion, this outweighed its disadvantages.  There is a great deal
+of discussion of this cluster of issues in [Hyndman and Koehler, 2006]
+(our reference \alert{[[fill in]])}.  If the reviewer thinks that it
+would strengthen our paper, we can recapitulate some of that material
+in our Section IVD.
 
-\emph{In many parts the article reads almost like a review. There is a lot of
-unnecessary redundancy (the entropy rate is quite low). In particular, the main
-point (``The aim of the paper is not ... but ...'') is repeated far too many
-times. The paper while easy to read could be shortened considerably without any
-loss of information.}
+\emph{In many parts the article reads almost like a review. There is a
+  lot of unnecessary redundancy (the entropy rate is quite low). In
+  particular, the main point (``The aim of the paper is not ... but
+  ...'') is repeated far too many times. The paper while easy to read
+  could be shortened considerably without any loss of information.}
 
-Due to the broad selection of fields that this work draws upon, we thought it
-necessary to provide background for each of them in case the reader is not
-familiar with some subset of them. While this necessitated some redundancy,
-\alert{we have made an effort to trim the fat and make the review portions more
-streamlined.}
+Due to the broad selection of fields that this work draws upon, we
+thought it necessary to provide background for readers who might not
+be familiar with all of them.  While this necessitated some
+redundancy, we have made an effort to trim the fat and make the review
+portions more streamlined.
 
@@
 
-\emph{Page 1: Why mention the `halting problem'? I don't see a straightforward
-connection to the problem at hand. The only connection is `undecidability' but
-there are many other examples for that.}
+\emph{Page 1: Why mention the `halting problem'? I don't see a
+  straightforward connection to the problem at hand. The only
+  connection is `undecidability' but there are many other examples for
+  that.}
 
-Our intent in mentioning this theoretical result was to side-step the possible
-response of simply ``always use the most sophisticated prediction scheme''. This
-corollary implies that no matter how sophisticated and broad a scheme we
-construct, it will not be the optimal one for some time series. Thus there is a
-real need for determining when a better scheme exists. \alert{We have modified
-the paragraph to make this point more clear.}
+This is an important issue to the computer-science community, but we
+have taken the reviewer's advice that it is less so for Physical
+Review E---and hence removed it.
+
+% Our intent in mentioning this theoretical result was to side-step the
+% possible response of simply ``always use the most sophisticated
+% prediction scheme''. This corollary implies that no matter how
+% sophisticated and broad a scheme we construct, it will not be the
+% optimal one for some time series. Thus there is a real need for
+% determining when a better scheme exists. \alert{We have modified the
+%   paragraph to make this point more clear.}
 
@@
 
-Thank you for catching our mistake, we have corrected the equation.
+Thank you for catching our mistake; we have corrected the equation.
 
@@
 
-You are absolutely correct, and we've adjusted the text accordingly.
+You are absolutely correct and we've adjusted the text accordingly.
 
@@
 
-We thank the reviewer for taking the time to both read our paper and provide us
-with valuable feedback. While we agree with several of the reviewers criticisms,
-and have taken steps to address them, we disagree that doing so changes the
-nature of the paper to the point of requiring a new submission.
+We thank the reviewer for taking the time to read our paper and
+provide us with valuable feedback.
 
-\emph{The problem of time series prediction is considered in the paper of J.
-Garland et al. Namely, the authors suggest to quantify predictability in advance
-(before using any concrete prediction scheme) based on the complexity estimate
-(weighted permutation entropy, WPE). They hypothesize that WPE is related to the
-best possible prediction error in a simple way. For several benchmark time
-series (reflecting variations of a computer performance during execution of
-different tasks), they obtain estimates of WPE and prediction error (the least
-one over four techniques), and build a linear regression using seven or eight
-resulting data points. They suppose that the obtained linear function can be
-used to ``predict'' the best prediction error from the model-free WPE estimate.
-Thus, the authors finally assume (as could be seen from the last section) that
-WPE and optimal prediction error for all systems are related via the same
-universal linear function with fixed coefficients. This is what I can understand
-from the entire paper, though the problem is not so definitely described in the
-Introduction.}
+\emph{The problem of time series prediction is considered in the paper
+  of J.  Garland et al. Namely, the authors suggest to quantify
+  predictability in advance (before using any concrete prediction
+  scheme) based on the complexity estimate (weighted permutation
+  entropy, WPE). They hypothesize that WPE is related to the best
+  possible prediction error in a simple way. For several benchmark
+  time series (reflecting variations of a computer performance during
+  execution of different tasks), they obtain estimates of WPE and
+  prediction error (the least one over four techniques), and build a
+  linear regression using seven or eight resulting data points. They
+  suppose that the obtained linear function can be used to ``predict''
+  the best prediction error from the model-free WPE estimate.  Thus,
+  the authors finally assume (as could be seen from the last section)
+  that WPE and optimal prediction error for all systems are related
+  via the same universal linear function with fixed coefficients. This
+  is what I can understand from the entire paper, though the problem
+  is not so definitely described in the Introduction.}
 
-\alert{We have modified the introduction to make our goals more clearly
-defined.}
+We have modified the introduction in order to define our goals more
+clearly.
 
-\emph{I think that the possibility to estimate the best prediction error in
-advance from a relatively simple WPE characteristic would be interesting and
-useful, if it could hold true for a wide class of systems. Such a question would
-deserve a careful study as that started in the presented work. However, I have a
-number of critical remarks and strongly doubt that the conjecture of the authors
-can be valid for a reasonably wide class of systems (not even speaking of ``all
-systems''). Other aspects of the paper are not so important, in my view, despite
-some of them are fresh and interesting (e.g. taking a computer as an object of
-prediction). My critical remarks are listed below.}
+\emph{I think that the possibility to estimate the best prediction
+  error in advance from a relatively simple WPE characteristic would
+  be interesting and useful, if it could hold true for a wide class of
+  systems. Such a question would deserve a careful study as that
+  started in the presented work. However, I have a number of critical
+  remarks and strongly doubt that the conjecture of the authors can be
+  valid for a reasonably wide class of systems (not even speaking of
+  ``all systems''). Other aspects of the paper are not so important,
+  in my view, despite some of them are fresh and interesting
+  (e.g. taking a computer as an object of prediction). My critical
+  remarks are listed below.}
 
-As we stated on page 19 of the manuscript, we believe that due to the broad
-spectrum of behaviors exhibited by the computer performance data these results
-will hold for time series drawn from other physical systems. We are aware that
-this claim may not be true, and believe that to be critical follow up work.
-\alert{However, we have included here some preliminary results that support our
-claim by analyzing a classical random walk, three time series derived from the
-classic Lorenz attractor, and the time series used in the SFI prediction
-competition.} Though we are not prepared to fully discuss or publish the results
-regarding these systems, we hope that the preliminary findings are enough to
-justify our claim for the reviewer.
+Computer dynamics spans the full range of behavior reported in the
+nonlinear dynamics literature.  Standard nonlinear time-series
+analysis of data from these systems has confirmed the existence of
+fixed points, periodic orbits of different periods, chaotic
+trajectories with various Lyapunov exponents, etc.  in their dynamics.
+While it is certainly possible that the nonlinear dynamics of computer
+performance is subtly different from the nonlinear dynamics of other
+systems---in a manner that cannot be detected by traditional nonlinear
+time-series analysis---we believe that our results will hold for time
+series drawn from other physical systems.
 
-\emph{1) Introduction is too long, but the main problem is not formulated
-definitely enough. I see that the main point is the approximately linear (or at
-least one-to-one) relationship between WPE and the best prediction error. This
-aim is not claimed explicitly in the Introduction. All other aspects of the
-manuscript do not seem important enough to be published in a separate article.}
+We are certainly aware that this claim may not be true, however, and
+we are in the process of the follow-up study suggested by the
+reviewer.  We have found that the relationship posited in our paper
+holds for a time series of voltages from a chaotic laser used in the
+SFI prediction competition, for instance.  (It also holds for various
+paradigmatic systems, such as the Henon map and the Lorenz system, as
+discussed in a later section of this response.)  The image below is an
+augmented version of our revised Figure \alert{[[fill in]]} that
+includes all of these examples.
 
-\alert{We have modified the manuscript to make it more clear that our goal is to
-derive a criterion for best-case prediction error in a model-free way. We have
-also shortened it.}
+\begin{center}
+    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extra_autolog}
+\end{center}
 
-\emph{2) In particular, two primary findings mentioned on page 2 (Introduction)
-are formulated quite vaguely. The first part of the first finding reads ``(i)
-complexity of a noisy real-value time series is quantifiable by permutation
-entropy''. However, it seems obvious from the previous research, including that
-cited by the authors. The authors do not suggest PE or WPE here, they exploit
-them as previously known approaches. It was previously known that PE is an
-estimate of KS entropy as the authors also state. It was well known that KS
-entropy relates to predictability and, in this sense, it is a measure of
-complexity (see e.g. the work [G. Boffetta, M. Cencini, M. Falcioni, and A.
-Vulpiani, ``Predictability: a way to characterize complexity'' // Physics
-Reports 356 (2002) 367–474] which is not cited by the authors). Thus, the first
-part of the first finding does not provide any new information. The second part
-of the first finding reads ``(ii) complexity of the noisy real-valued time
-series is correlated with prediction accuracy of an appropriate predictor''.
-However, it seems also obvious and directly follows from the definition of KS-
-entropy and its discussion in many papers including [Boffetta et al.] cited
-above.}
+\noindent The dashed-line fit in this plot (and in Figure
+\alert{[[fill in]])} is based solely on the computer performance data.
+Nonetheless, the points from the other time series fall within a
+narrow band around the fit.  Note that this fit ($y = a \log(b x +
+c)$) is different than in the submitted version.  A log fit is more
+appropriate because any signal with zero entropy should be perfectly
+predictable, and so should ideally have a MASE value of zero.  Note,
+too, that there are a lot more points on this plot than in our
+original version; this is discussed further below.
 
-While we agree that on the surface these claims are both trivial and reasonably
-well known in theory and for toy models, our specific intent is that these also
-hold and can be practically applied via the WPE in real-world systems with
-potentially both dynamical and observational noise. We also greatly appreciated
-being informed about Boffetta et al's paper, \alert{and have included a citation
-to it in our paper.}
+\emph{1) Introduction is too long, but the main problem is not
+  formulated definitely enough. I see that the main point is the
+  approximately linear (or at least one-to-one) relationship between
+  WPE and the best prediction error. This aim is not claimed
+  explicitly in the Introduction. All other aspects of the manuscript
+  do not seem important enough to be published in a separate article.}
 
-\emph{3) By the way, the authors call ``fully complex'' the white noise process,
-i.e. they equate predictability and complexity. However, they do not even
-mention another approach to the notion of complexity which ascribes low
-complexity to the white noise [the works e.g. by C.R. Shalizi and J. Crutchfield
-carried out in the Santa Fe Institute from where one of the authors is]. It
-would be appropriate to mention that ambiguity of the notion of complexity if
-the latter plays an important role in many formulations of the authors.}
+We have modified the manuscript to make it more clear that our goal is
+to derive a practical criterion for best-case prediction error in a
+model-free way, and to test it in the context of empirical data.
 
-\alert{We have added a short paragraph discussing the differences in notions of
-complexity, and clarified our working definition of it as used for this paper.}
+\emph{2) In particular, two primary findings mentioned on page 2
+  (Introduction) are formulated quite vaguely. The first part of the
+  first finding reads ``(i) complexity of a noisy real-value time
+  series is quantifiable by permutation entropy''. However, it seems
+  obvious from the previous research, including that cited by the
+  authors. The authors do not suggest PE or WPE here, they exploit
+  them as previously known approaches. It was previously known that PE
+  is an estimate of KS entropy as the authors also state. It was well
+  known that KS entropy relates to predictability and, in this sense,
+  it is a measure of complexity (see e.g. the work [G. Boffetta,
+    M. Cencini, M. Falcioni, and A.  Vulpiani, ``Predictability: a way
+    to characterize complexity'' // Physics Reports 356 (2002)
+    367–474] which is not cited by the authors). Thus, the first part
+  of the first finding does not provide any new information. The
+  second part of the first finding reads ``(ii) complexity of the
+  noisy real-valued time series is correlated with prediction accuracy
+  of an appropriate predictor''.  However, it seems also obvious and
+  directly follows from the definition of KS- entropy and its
+  discussion in many papers including [Boffetta et al.] cited above.}
 
-\emph{4) The second finding reads ``The way information is generated and
-processed internally by a system plays a crucial role in the success of
-different forecasting schema - and in the choice of which one is appropriate for
-a given time series''. This statement also looks trivial. Indeed, an optimal
-predictor for a linear stochastic system and low-dimensional nonlinear
-deterministic system are quite different (an AR model versus a local or global
-nonlinear model) due to different properties of the original systems. Thus,
-there is nothing to be proven here. Probably, the authors imply their particular
-benchmark signals (reflecting a computer behavior) but they do not claim that.
-Thus, both primary findings seem to provide nothing new.}
+We agree that these claims are both trivial and reasonably well known,
+in theory and for toy models.  Our specific claim, which was not well
+explained in the submitted version, is that those claims also
+hold---and can be practically applied---in real-world time-series data
+that may contain dynamical and/or observational noise.  This has not,
+to our knowledge, received much attention in the literature.  We have
+added a paragraph to the introduction that hopefully makes all of this
+more clear.
 
-It is indeed true that the way we worded this finding makes it sound trivial.
-Our intent, as we have modified the text to reflect, is that the appropriateness
-of a predictor can be discovered empirically via comparison with the WPE.
-\alert{i'm not entirely sure how to respond to this one correctly.}
+\alert{We appreciate being informed about [Boffetta et al.]  paper,
+  and we have included a discussion of it in \alert{Section [[what]].
+    The discussion is in the wrong place and doesn't make logical
+    sense.  Needs discussion.}
 
-\emph{5) The benchmark system - a computer as an object of prediction - seems to
-be an unexpected and, probably, interesting choice but just as an additional
-illustration of a statement which should be first shown for well-controlled and
-well-understood paradigmatic systems (like low-dimensional maps or ordinary
-differential equations). Indeed, it is not known in advance what error one
-should expect for such a complicated object using a certain predictive
-technique, what technique is optimal or close to optimal, etc. If the main point
-of the research is the relationship between WPE and the best prediction error
-(if any), selection of such a complicated object as computer performance
-variations only introduces its own difficulties into the problem.}
+\emph{3) By the way, the authors call ``fully complex'' the white
+  noise process, i.e. they equate predictability and
+  complexity. However, they do not even mention another approach to
+  the notion of complexity which ascribes low complexity to the white
+  noise [the works e.g. by C.R. Shalizi and J. Crutchfield carried out
+    in the Santa Fe Institute from where one of the authors is]. It
+  would be appropriate to mention that ambiguity of the notion of
+  complexity if the latter plays an important role in many
+  formulations of the authors.}
 
-As mentioned earlier in our response, we have included preliminary results
-regarding some simple systems here for the reviewer, and indeed the well-known
-results relating entropy and predictability hold for these systems. Because
-these results are not surprising, but rather simply confirming what is known in
-theory, we did not want to include them in the paper but rather focus on how
-these results still hold in the face of very ``dirty'' real-world data.
+We have added a short discussion about the differences in notions and
+definitions of complexity, and clarified the working definition that
+we use in this paper.
+
+\emph{4) The second finding reads ``The way information is generated
+  and processed internally by a system plays a crucial role in the
+  success of different forecasting schema - and in the choice of which
+  one is appropriate for a given time series''. This statement also
+  looks trivial. Indeed, an optimal predictor for a linear stochastic
+  system and low-dimensional nonlinear deterministic system are quite
+  different (an AR model versus a local or global nonlinear model) due
+  to different properties of the original systems. Thus, there is
+  nothing to be proven here. Probably, the authors imply their
+  particular benchmark signals (reflecting a computer behavior) but
+  they do not claim that.  Thus, both primary findings seem to provide
+  nothing new.}
+
+The original wording of this finding did make it sound trivial.  Our
+intent, as we have modified the text to reflect, is that the
+appropriateness of a predictor can be discovered \emph{empirically}
+via comparison with the WPE.  Previous explorations, in contrast, have
+used clean data and/or systems where the generating process is known.
+The paragraph that we have added to the introduction (immediately
+following these two claims) should hopefully make this more clear.
+\alert{Ryan said: ``i'm not entirely sure how to respond to this one
+  correctly.''  Liz asks: do you mean in the rebuttal or in the text?}
+
+\emph{5) The benchmark system - a computer as an object of prediction
+  - seems to be an unexpected and, probably, interesting choice but
+  just as an additional illustration of a statement which should be
+  first shown for well-controlled and well-understood paradigmatic
+  systems (like low-dimensional maps or ordinary differential
+  equations). Indeed, it is not known in advance what error one should
+  expect for such a complicated object using a certain predictive
+  technique, what technique is optimal or close to optimal, etc. If
+  the main point of the research is the relationship between WPE and
+  the best prediction error (if any), selection of such a complicated
+  object as computer performance variations only introduces its own
+  difficulties into the problem.}
+
+As mentioned in conjunction with the image above, the relationship
+posited in our paper does indeed hold for paradigmatic systems like
+the Henon map and the Lorenz equations.  We had not included these
+results in our paper because they simply confirmed well-known results,
+and because our focus was on confirming those results in the context
+of noisy, real-world data.  \alert{We have added a sentence in the
+  conclusion (or maybe end of discussion) about the fact that the WPE
+  vs. MASE relationship for these synthetic examples follows the
+  empirically determined pattern in our paper.}
 
 \emph{6) The description of the choice of the benchmark system in the
-Introduction is too long. On the other hand, if the computer signals are the
-main interest of the work, then it should be claimed explicitly. Then, the paper
-should be rewritten and submitted to a kind of engineering journals where it
-might well be of interest.}
+  Introduction is too long. On the other hand, if the computer signals
+  are the main interest of the work, then it should be claimed
+  explicitly. Then, the paper should be rewritten and submitted to a
+  kind of engineering journals where it might well be of interest.}
 
-Since our primary goal with this paper is the relationship between prediction
-error and WPE in noisy, real-world systems, and the practical applicability of
-this relationship, we have reduced the amount of space spent in the introduction
-discussing the particular time series we use.
+The primary goal of this paper is to explore the relationship between
+prediction error and WPE in noisy, real-world systems, and the
+practical applicability of this relationship.  Accordingly, we have
+cut down on the level of detail regarding the benchmark system and
+pointed interested readers to our other papers for the engineering
+details.  We did not feel that removing all of this material would
+be effective, however, because it helps the reader understand why
+this is a good testbed for exploring the problem at hand.
 
-\emph{7) As for the choice of the four prediction techniques, some of them seem
-superfluous. E.g. naive approach is just a global AR model of the zero order.
-Random-walk is similar to the AR model of the first order where the coefficient
-is not estimated by the least-squares but set equal to 1 (or this is an ARIMA
-model with the first difference and ``zero ARMA part''). Inefficiencies of the
-ARIMA model as compared to those choices in the paper can represent an
-inappropriate procedure for the order selection rather than indicate
-inappropriateness of the entire ARIMA approach. Thus, linear ARIMA and nonlinear
-LMA might be sufficient (in combination with mathematical benchmark examples) to
-represent close-to-optimal prediction errors.}
+\emph{7) As for the choice of the four prediction techniques, some of
+  them seem superfluous. E.g. naive approach is just a global AR model
+  of the zero order.  Random-walk is similar to the AR model of the
+  first order where the coefficient is not estimated by the
+  least-squares but set equal to 1 (or this is an ARIMA model with the
+  first difference and ``zero ARMA part''). Inefficiencies of the
+  ARIMA model as compared to those choices in the paper can represent
+  an inappropriate procedure for the order selection rather than
+  indicate inappropriateness of the entire ARIMA approach. Thus,
+  linear ARIMA and nonlinear LMA might be sufficient (in combination
+  with mathematical benchmark examples) to represent close-to-optimal
+  prediction errors.}
 
-It is certainly true that the naive and random-walk predictors can be seen as
-subclasses of the ARIMA model. We have explicitly included them, however, since
-they are used (sometimes to great success) in practice. Further, we have used
-standard methods \alert{josh -- should we add some names here?} of fitting the
-parameters used in our ARIMA predictions, and we believe any inefficiencies of
-that parameter selection support our findings --- namely, if a particular
-selection of model parameters predicts poorly (compared to what the WPE value
-might suggest), it is an indication that the specific model, including
-parameters, is not optimal for the job. This might mean that a different
-parameter selection scheme is needed, or that the entire model is inappropriate.
-Unfortunately the WPE value alone is insufficient to differentiate these two
-situations.
+\alert{We need to discuss this.  We also need to discuss whether this
+  response requires changes in the text.It is certainly true that the
+  naive and random-walk predictors can be seen as subclasses of the
+  ARIMA model. We have explicitly included them, however, since they
+  are used (sometimes to great success) in practice. Further, we have
+  used standard methods of fitting the parameters used in our ARIMA
+  predictions, and we believe any inefficiencies of that parameter
+  selection support our findings --- namely, if a particular selection
+  of model parameters predicts poorly (compared to what the WPE value
+  might suggest), it is an indication that the specific model,
+  including the mechanism for selecting parameter values, is not
+  optimal for the job. This might mean that a different parameter
+  selection scheme is needed, or that the entire model is
+  inappropriate.  Unfortunately the WPE value alone is insufficient to
+  differentiate these two situations.  It can identify both, but
+  cannot distinguish.}
 
@@
 
-The purpose of this short paragraph in this paper, which explains the
-efficiencies and inefficiencies of these simple prediction schemes, is present
-for two purposes. First, to explain why we have included these simple schemes
-rather than just the more complicated ones, and second to help in explaining the
-specific results obtained later in the paper.
+This short paragraph in this paper, which explains the efficiencies
+and inefficiencies of these simple prediction schemes, was present for
+two purposes: first, to explain why we have included these simple
+schemes rather than just the more complicated ones; second, to help in
+explaining the specific results described later in the paper.
 
-\emph{9) A better justification of the prediction error metrics could be also in
-order if this work is continued. At least, comparison of prediction errors
-quantified with different metrics seems necessary. Especially, taking into
-account the following major concerns.}
+\emph{9) A better justification of the prediction error metrics could
+  be also in order if this work is continued. At least, comparison of
+  prediction errors quantified with different metrics seems
+  necessary. Especially, taking into account the following major
+  concerns.}
 
-We have strengthened our justification for the use of the MASE metric. The work
-we have cited which introduced MASE extensively compares it to other schemes and
-highlights the differences obtained therein. We further believe that a
-comparison of different prediction error metrics as they relate to WPE is best
-left as a sequel to this work.
+We have extended the justification for our use of the MASE metric.
+\alert{did we?  need to check.}  The paper in which MASE was
+introduced [Hyndman and Koehler, 2006] (our reference \alert{[[fill
+      in]])} includes an extensive evaluation of this metric, as well
+as a comprehensive comparison between it and other metrics.  If the
+reviewer thinks that it would strengthen our paper, we can
+recapitulate more of that material in our Section IVD.<  As part of our
+current work on this topic, we are evaluating how other choices of
+metric affect the shape of our curves, but we feel that a
+comprehensive study of this issue is best left as a sequel to this
+work.
 
@@
 
-\emph{I) Strict relationship between WPE and prediction error applicable to all
-systems is hardly possible. In particular, there is a notion of epsilon-entropy
-which tends to the KS entropy if epsilon (the cell size) tends to zero (see e.g.
-[Boffetta et al.] cited above). For stochastic systems, the KS entropy is
-infinite and epsilon-entropy varies strongly with the scale epsilon. Seemingly,
-the permutation entropy estimate in this case will strongly depend on the choice
-of the word length L. The authors concentrate on L = 6 and obtain coefficients
-of the regression between WPE and prediction error only for that choice and for
-their particular signals. However, other systems and other choices of L may
-easily lead to other regression coefficients so that deciding about the best
-possible prediction error from WPE seems impossible in general. It is good if
-such a decision is possible for a certain narrow class of systems. However, the
-authors do not even discuss to which class of systems their results apply.}
+\emph{I) Strict relationship between WPE and prediction error
+  applicable to all systems is hardly possible. In particular, there
+  is a notion of epsilon-entropy which tends to the KS entropy if
+  epsilon (the cell size) tends to zero (see e.g.  [Boffetta et al.]
+  cited above). For stochastic systems, the KS entropy is infinite and
+  epsilon-entropy varies strongly with the scale epsilon. Seemingly,
+  the permutation entropy estimate in this case will strongly depend
+  on the choice of the word length L. The authors concentrate on L = 6
+  and obtain coefficients of the regression between WPE and prediction
+  error only for that choice and for their particular
+  signals. However, other systems and other choices of L may easily
+  lead to other regression coefficients so that deciding about the
+  best possible prediction error from WPE seems impossible in
+  general. It is good if such a decision is possible for a certain
+  narrow class of systems. However, the authors do not even discuss to
+  which class of systems their results apply.}
 
-It is certainly true that the specific relationship between WPE and MASE values
-is dependent on word length, though for any reasonable choice of $\ell$
-(discussion regarding reasonable values in the manuscript) we believe a similar
-relationship to hold. We also believe, due to the spectrum of time series
-behaviors we have included, that the curve is not highly dependent on which
-systems are considered. As stated before, we have hopefully assuaged the
-reviewers concern about this point by including here the WPE and MASE values for
-a variety of other time series, including a random walk, simple chaotic systems,
-as time series used in the SFI prediction contest.
+\alert{This needs discussion.  It is certainly true that WPE values
+  depend on wordlength.  This is a real issue, particularly because
+  word length choices are rarely justified in the PE literature.
+  Quote from a few papers.  The right thing to do, as we argued in
+  section n, is to compute PE across a range of word lengths and look
+  for a flat spot in the curve.  A systematic approach like this
+  mitigates, but of course does not remove, any variation in the
+  results.  We used that method to choose individual l values for each
+  time series studied here.  We believe that this systematic approach
+  to l choice mitigates the differences in the WPE vs. MASE results
+  between time series; we have not yet confirmed that conjecture but
+  we are in the process of doing so.
 
-\emph{II) PE is invariant under an invertible nonlinear change of variables. The
-prediction error measured as the authors suggest is not invariant. Thus, such a
-simple change of variables can lead, at least, to a change in regression
-coefficient between WPE and prediction error. Thus, the authors should also
-discuss to which variable for that class of systems their result apply. I think
-that the fact that the authors observe a certain regression line at all seems to
-be due to small number of different signals they consider (only 7 or 8 different
-data points, such a small number of points may lie approximately near a straight
-line by chance).}
+Ryan had written: ``the specific relationship between WPE and MASE
+values is dependent on word length, though for any reasonable choice
+of $\ell$, we believe a similar relationship to hold.'' Liz is not
+sure she understands what ``similar'' means or why that's justified.}
 
-Both WPE and MASE are sensitive to changes of variables, and from some initial
-testing they are so in complementary ways. This leads a change in variables
-to...  We have also redone our regression line to include all 105 WPE-MASE
-pairs, rather than just a representative from each program.
+Regarding generality: the goal of this paper was to explore model-free
+heuristics for empirical time-series data, where one does not know
+anything about the underlying system.  In situations like this,
+equivalence classes are difficult to define.  We addressed
+this---insofar as is possible in an experimental context---by choosing
+a range of time-series data sets whose behavior spans a wide range of
+dynamical behaviors.  Recapitulate that henon, lorenz, SFI A all fit
+the curve determined by the computer perf data; cite as justification
+for generality and point to future work.
+
+\emph{II) PE is invariant under an invertible nonlinear change of
+  variables. The prediction error measured as the authors suggest is
+  not invariant. Thus, such a simple change of variables can lead, at
+  least, to a change in regression coefficient between WPE and
+  prediction error. Thus, the authors should also discuss to which
+  variable for that class of systems their result apply. I think that
+  the fact that the authors observe a certain regression line at all
+  seems to be due to small number of different signals they consider
+  (only 7 or 8 different data points, such a small number of points
+  may lie approximately near a straight line by chance).}
+
+Both WPE and MASE are sensitive to changes of variables, and from some
+initial testing they are so in complementary ways. This leads a change
+in variables to...  We have also redone our regression line to include
+all 105 WPE-MASE pairs, rather than just a representative from each
+program.
 

### 2014-07-14 methods.tex

@@
 The time-series data sets for these experiments were collected on an
-Intel Core\textsuperscript{\textregistered} i7-2600-based machine
-running the 2.6.38-8 Linux kernel.  This particular microprocessor
-chip has eight processing units, a clock rate of 3.40 GHz, and a cache
-size of 8192 KB.  We gathered performance traces during the execution
-of three different programs---the simple \col loop whose performance
-is depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
-programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
-from the LAPACK linear algebra package (\svd).  In these experiments,
-the scalar observation $x_i$ was a measurement of the processor
-performance at time $i$ during the execution of each program.  To
-record these measurements, we used the {\tt libpfm4} library, via PAPI
+Intel Core\textsuperscript{\textregistered} i7-2600-based machine.
 % 
-% \footnote{Performance Application Programming Interface}
+% running the 2.6.38-8 Linux kernel.  This particular microprocessor
+% chip has eight processing units, a clock rate of 3.40 GHz, and a cache
+% size of 8192 KB.  
 % 
-5.2~\cite{papi}, to stop program execution at 100,000-instruction
-intervals---the unit of time in this paper---and read the contents of
-the CPU's onboard hardware performance monitors\footnote{These
-  specialty registers are built into modern microprocessors for the
-  purpose of monitoring and storage of performance data.  They can be
-  programmed to measure a wide array of processor and memory metrics
-  besides the one used here.}, which we had programmed to count how
-many instructions were executed in each clock cycle (IPC).  For an
-in-depth description of this custom-measurement infrastructure,
-including discussion of the implications of the sampling interval,
+We gathered performance traces during the execution of three different
+programs---the simple \col loop whose performance is depicted in
+Figure~\ref{fig:col-ipc}, as well as two more-complex programs: one
+from the SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK
+linear algebra package (\svd).  In these experiments, the scalar
+observation $x_i$ was a measurement of the processor performance at
+time $i$ during the execution of each program.  
+% 
+% To record these measurements, we used the {\tt libpfm4} library, via
+% PAPI
+% % 
+% % \footnote{Performance Application Programming Interface}
+% % 
+% 5.2~\cite{papi}, to stop program execution at 100,000-instruction
+% intervals---the unit of time in this paper---and read the contents of
+% the CPU's onboard hardware performance monitors\footnote{These
+%   specialty registers are built into modern microprocessors for the
+%   purpose of monitoring and storage of performance data.  They can be
+%   programmed to measure a wide array of processor and memory metrics
+%   besides the one used here.}, which we had programmed to count how
+% many instructions were executed in each clock cycle (IPC).  
+% 
+For an in-depth description of this experimental setup,
+% including discussion of the implications of the sampling interval,
 please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
@@
 \col is a simple C program that repeatedly initializes the upper
-triangle of a 2048 $\times$ 2048 matrix in column-major order by
-looping over the following three lines of code:
-\begin{lstlisting}
-  for(i=0;i<2048;i++)
-  	for(j=i;j<2048;j++)
-  		data[j][i] = 0;
-\end{lstlisting}
+triangle of a 2048 $\times$ 2048 matrix in column-major order.
+% 
+% by looping over the following three lines of code:
+% \begin{lstlisting}
+%   for(i=0;i<2048;i++)
+%   	for(j=i;j<2048;j++)
+%   		data[j][i] = 0;
+% \end{lstlisting}
 % full listing
@@
 %
-\noindent As mentioned in Section~\ref{sec:intro}, this simple program
-exhibits surprisingly complicated behavior~\cite{mytkowicz09}.  A time
-series of the processor performance during the execution of this
-program is shown in Figure~\ref{fig:col-ts}.
+As mentioned in Section~\ref{sec:intro}, this simple program exhibits
+surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
+of the processor performance during the execution of this program is
+shown in Figure~\ref{fig:col-ts}.
 \begin{figure}[htbp]
@@
       execution of \svd. Each point is the average IPC in a 100,000
-      instruction period.  The colors identify the different \emph{
-        segments} of the signal that are discussed in the text.}
+      instruction period.  The colors identify the different
+      \emph{segments} of the signal that are discussed in the text.}
     \label{fig:svd-ts-colored}

### 2014-07-14 model.tex

@@
 is effectively obfuscated, from the standpoint of the LMA method, by
-noise.  By quantifying the balance between redundancy (that is,
-predictive structure) and entropy for these real-valued time series,
-as shown in Section~\ref{sec:results}, we can begin to answer these
-questions.
+noise.  Without any knowledge of the generating process, answers to
+these questions can only be derived from the data, with all of the
+attendant problems (noise, sampling issues, etc.)  By quantifying the
+balance between redundancy (that is, predictive structure) and entropy
+for these real-valued time series, as shown in
+Section~\ref{sec:results}, we can begin to answer these questions
+in an effective and practical manner.
 
@@
 \end{table*}
-
 In these images, the vertical axis is the prediction $p_j$ and the

### 2014-07-14 pecomplexity.tex
address some stuff
edits

@@
 \usepackage{graphicx}
-\usepackage{hyperref}
+%\usepackage{hyperref}
 \usepackage[sort&compress]{natbib}
@@
 \usepackage{listings}
+\usepackage{rotating}
+\usepackage{array,booktabs}
 \lstset{language=C}
@@
 strategies fail when they are applied to complicated time series. We
-conjecture that the inherent complexity of a time series --- which
-results from the dimension, nonlinearity, and non-stationarity of the
-generating process, as well as from measurement issues like noise,
-aggregation, and finite data length --- is both empirically
-quantifiable and directly correlated with predictability.  In
-particular, we argue that \emph{redundancy} is an effective way to
-measure complexity and predictive structure in a time series and that
+conjecture that the inherent complexity of real-world time-series
+data---which results from the dimension, nonlinearity, and
+non-stationarity of the generating process, as well as from
+measurement issues like noise, aggregation, and finite data
+length---is both \emph{empirically} quantifiable and directly
+correlated with predictability.  In particular, we argue that
+\emph{redundancy} is an effective way to measure complexity and
+predictive structure in an experimental time series and that
 \emph{weighted permutation entropy} is an effective way to estimate
@@
 accuracy of the predictions with the permutation entropy of that time
-series.  We use the results to develop a heuristic that can help
-practitioners recognize when a particular prediction method is not
-well matched to the task at hand: that is, when the time series has
-more predictive structure than that method can capture and exploit.
+series.  We use the results to develop a model-free heuristic that can
+help practitioners recognize when a particular prediction method is
+not well matched to the task at hand: that is, when the time series
+has more predictive structure than that method can capture and
+exploit.
 

### 2014-07-14 response.tex
second round of revisions.  awaiting today's conversation before can finish.

@@
 intent, as we have modified the text to reflect, is that the
-appropriateness of a predictor can be discovered \emph{empirically}
-via comparison with the WPE.  Previous explorations, in contrast, have
-used clean data and/or systems where the generating process is known.
-The paragraph that we have added to the introduction (immediately
-following these two claims) should hopefully make this more clear.
-\alert{Ryan said: ``i'm not entirely sure how to respond to this one
-  correctly.''  Liz asks: do you mean in the rebuttal or in the text?}
+appropriateness of a predictor can be discovered \emph{empirically},
+via comparison with the WPE.  Previous explorations have used clean
+data and/or systems where the generating process is known.  Our work
+makes neither of these assumptions.  The paragraph that we have added
+to the introduction (immediately following these two claims) should
+hopefully make this more clear.  \alert{Ryan said: ``i'm not entirely
+  sure how to respond to this one correctly.''  Liz asks: do you mean
+  in the rebuttal or in the text?}
 
@@
 
-Both WPE and MASE are sensitive to changes of variables, and from some
-initial testing they are so in complementary ways. This leads a change
-in variables to...  We have also redone our regression line to include
-all 105 WPE-MASE pairs, rather than just a representative from each
-program.
+Both WPE and MASE are sensitive to changes of variables; our initial
+testing, however, indicates they are so \emph{in complementary ways},
+so the shape of the WPE vs. MASE curve is [[largely invariant with
+    respect to changes in what]] \alert{Ryan wrote: ``This leads a
+  change in variables to...''  What should go in here?  What variables
+  is he talking about?  Did we actually do that test (varying them)?
+  What should we say about that word ``largely''?}
 
-\emph{To summarize, the paper shows that the authors are well familiar with the
-entire topic and highly qualified, they overview some well-known facts (WPE,
-various prediction techniques, etc) and ``invent'' an unexpected test system
-(computer performance data), they report about their attempts to play around
-with all that staff. However, their results do not seem to provide any new
-information to a specialist. Probably, this work can be continued to get some
-more reliable results about concrete relationship between WPE and prediction
-error for certain classes of systems (however, I doubt that it is possible to
-derive any simple and widely applicable relationship like the linear function
-suggested by the authors). Seemingly, it should then be a new paper which can be
+The fit of the regression line was not a function of the small number
+of points; when we included all 105 WPE-MASE pairs, rather than just a
+representative from each program, the slope of the fit changed by 2\%
+and the y intercept by 1\%.
+
+We have revised the Figure, and the associated text, to include all
+105 points.  We have also moved from a linear fit to a log fit in this
+curve to better match the theoretical relationship between MASE and
+WPE, as mentioned above.  Finally, we now show a ``fuzzy'' version of
+the fitted curve in an effort to emphasize that the fundamental claim
+of the paper is not related to the crispness of the fit.
+
+\emph{To summarize, the paper shows that the authors are well familiar
+  with the entire topic and highly qualified, they overview some
+  well-known facts (WPE, various prediction techniques, etc) and
+  ``invent'' an unexpected test system (computer performance data),
+  they report about their attempts to play around with all that
+  staff. However, their results do not seem to provide any new
+  information to a specialist. Probably, this work can be continued to
+  get some more reliable results about concrete relationship between
+  WPE and prediction error for certain classes of systems (however, I
+  doubt that it is possible to derive any simple and widely applicable
+  relationship like the linear function suggested by the
+  authors). Seemingly, it should then be a new paper which can be
   hardly considered as a revised version of the present manuscript.}
 
-It is our hope that between the additional preliminary results here, the
-modifications to the paper, as well as our clarifying the goals of our paper
-(understandably misunderstood in the draft), that the reviewer now considers the
-manuscript appropriate for publication in PRE.
+It is our hope that between the additional preliminary results here,
+the modifications to the paper, as well as our clarifying the goals of
+our paper (understandably misunderstood in the draft), that the
+reviewer now considers the manuscript appropriate for publication in
+PRE.
 

### 2014-07-14 concl.tex
second round of revisions.  awaiting today's conversation before can finish.

--- 2014-04-27 concl.tex
@@
 Without knowing anything about the generating process, it is difficult
-to determine how much predictive structure is present in a given time
-series.  And even if predictive structure exists, a given forecast
-method may not work, simply because it cannot exploit the structure
-that is present (e.g., a linear model of a nonlinear process).  If a
-forecast model is not producing good results, a practitioner needs to
-know why: is the reason that the time series contains no predictive
-structure---i.e., that no model will work---or is the model that s/he
-is using simply not good enough?
+to determine how much predictive structure is present in a noisy,
+real-world time series.  And even if predictive structure exists, a
+given forecast method may not work, simply because it cannot exploit
+the structure that is present (e.g., a linear model of a nonlinear
+process).  If a forecast model is not producing good results, a
+practitioner needs to know why: is the reason that the data contain no
+predictive structure---i.e., that no model will work---or is the model
+that s/he is using simply not good enough?
 
 In this paper, we have argued that redundancy~\cite{crutchfield2003}
-is a useful definition of the inherent predictability of a time
-series.  To operationalize that definition, we use an approximation of
-the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
-version of the permutation entropy of~\cite{bandt2002per}.  This WPE
-technique---an ordinal calculation of forward information transfer in
-a time series---is ideal for our purposes because it works with
-real-valued data and is known to converge to the true entropy value.
-Using a variety of forecast models and more than a hundred time-series
-data sets from a computer-performance experiment, we have shown that
-prediction accuracy is indeed correlated with weighted permutation
-entropy: the higher the WPE, in general, the higher the prediction
-error.  There are some exceptions: \svdone is easier to predict than
-its WPE value suggests using any of the three forecast strategies
-studied here.  Conversely, forecast errors for the \col and \svdfive
-signals are higher than the corresponding WPEs suggest---except for
-the nonlinear LMA predictor.  On the whole, though, signals with
-higher weighted permutation entropy are harder to predict.
+is a useful definition of the inherent predictability of an empirical
+time series.  To operationalize that definition, we use an
+approximation of the Kolmogorov-Sinai entropy~\cite{lind95}, estimated
+using a weighted version of the permutation entropy
+of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation
+of forward information transfer in a time series---is ideal for our
+purposes because it works with real-valued data and is known to
+converge to the true entropy value. Using a variety of forecast models
+and more than a hundred time-series data sets from a
+computer-performance experiment, we have shown that prediction
+accuracy is indeed correlated with weighted permutation entropy: the
+higher the WPE, in general, the higher the prediction error.  The
+relationship is roughly logarithmic, which makes sense because any
+signal with zero entropy should be perfectly predictable, and so
+should ideally have a MASE value of zero.  Supporting our hypothesis,
+forecast errors for the \col and \svdfive signals are higher than the
+corresponding WPEs suggest---except for the nonlinear LMA predictor.
+There are some exceptions: \svdone is easier to predict than its WPE
+value suggests using any of the three forecast strategies studied
+here.  On the whole, though, signals with higher weighted permutation
+entropy are harder to predict.
 
@@
 %  \naive ~wins on \svdone with the exception to the 5 outliers.
-The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
+The curve in Figures~\ref{fig:wpe_vs_mase_best}
 and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
@@
 guideline for knowing when a model is not well-matched to the task at
-hand: a MASE score that is more than $3.58 \cdot \mathrm{WPE} - 1.91$,
-where $\mathrm{WPE}$ is that of the corresponding signal, is an
-indication that that time series has more predictive structure than
-that forecast model can capture and exploit.
+hand: a MASE score that is significantly higher than [[$3.58 \cdot
+    \mathrm{WPE} - 1.91$]], where $\mathrm{WPE}$ is that of the
+corresponding signal, is an indication that that time series has more
+predictive structure than that forecast model can capture and exploit.
+
+[[This paragraph duplicates material in the last paragraph of this
+    section.  Some of it needs to be here rather than there, judging
+    by the ref's reactions.  But the later paragraph has a nice
+    concluding tone, so need to figure out what goes where.]]
+
+This curve was determined empirically using a finite set of forecast
+methods and data sets---the latter drawn from a single experimental
+system.  Computer dynamics spans the full range of behavior reported
+in the nonlinear dynamics literature.  Standard nonlinear time-series
+analysis of data from these systems has confirmed the existence of
+fixed points, periodic orbits of different periods, chaotic
+trajectories with various Lyapunov exponents, etc. in their dynamics.
+While it is certainly possible that the nonlinear dynamics of computer
+performance is subtly different from the nonlinear dynamics of other
+systems---in a manner that cannot be detected by traditional nonlinear
+time-series analysis---we believe that our results will hold for time
+series drawn from other physical systems.  We are certainly aware that
+this claim may not be true, however, and we are in the process of a
+follow-up study exploring a broader range of applications.
+Preliminary results of this study show that MASE vs. WPE points for
+data from the SFI prediction competition \cite{sfi} fall in the shaded
+region on Figure [[what]]\footnote{Synthetic data from paradigmatic
+  systems like the Henon map \cite{henon} and the Lorenz system
+  \cite{lorenz63} also fit this pattern.}.
 

### 2014-07-14 results.tex
more updates
second round of revisions.  awaiting today's conversation before can finish.

--- 2014-04-22 results.tex
@@
 \item That the weighted permutation entropy (WPE) of a noisy
-  real-valued time series is correlated with prediction
-  accuracy---i.e., that the predictable structure in a time-series
-  data set can be quantified by its WPE.
+  real-valued time series from an unknown system is correlated with
+  prediction accuracy---i.e., that the predictable structure in an
+  empirical time-series data set can be quantified by its WPE.
 
@@
 
-\item That the way information is generated and processed by a system
-  is correlated with the effectiveness of a given predictor on
-  time-series data from that system
+\item That the way information is generated and processed by an
+  unknown system is correlated with the effectiveness of a given
+  predictor on empirical time-series data from that system
 
@@
   col\_major}, {\tt 403.gcc}, and the six different regimes of the
-{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
-experiments was to explore how prediction accuracy is related to WPE,
-and how that relationship depends on the generating process and the
-prediction method.  Working from the first 90\% of each signal, we
-generated a prediction of the last 10\% using the \naive, ARIMA, and
-LMA prediction methods, as described in Section~\ref{sec:accuracy},
-then calculated the MASE value of those predictions.  We also
-calculated the WPE of each time series using a wordlength of six, as
-described in Section~\ref{sec:meaComplex}.  In order to assess the
-run-to-run variability of these results, we repeated all of these
-calculations on 15 separate trials: i.e., 15 different runs of each
-program.
+{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective
+of these experiments was to explore how prediction accuracy is related
+to WPE.
+% 
+% , and how that relationship depends on the generating process
+% and the prediction method.  
+% 
+Working from the first 90\% of each signal, we generated a prediction
+of the last 10\% using the \naive, ARIMA, and LMA prediction methods,
+as described in Section~\ref{sec:accuracy}, then calculated the MASE
+value of those predictions.  We also calculated the WPE of each time
+series using a wordlength chosen via the procedure described in
+Section~\ref{sec:meaComplex}.  In order to assess the run-to-run
+variability of these results, we repeated all of these calculations on
+15 separate trials: i.e., 15 different runs of each program.
 
-Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these 45
-predictions for each system: that is, the lowest error over all 15
-trials and all three methods for each of the eight programs.  The WPE
-is plotted against the corresponding MASE value here in order to bring
-out the correlation between these two quantities.
+% Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these
+% predictions for each trial of each system: that is, the lowest error
+% over all three methods for each of the eight programs.  The WPE is
+% plotted against the corresponding MASE value here in order to bring
+% out the correlation between these two quantities.
+
+Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
+corresponding MASE values [[for all of these trials.]]
 \begin{figure}[htbp]
   \centering
-  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
+  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_autolog}
   \caption{Weighted permutation entropy vs. mean absolute scaled error
-    (MASE) of the best prediction of each system.
+    (MASE) of [[all trials for each system]].
 %
@@
 %
-The dashed line is a least-squares linear fit of all the points except
-{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
-text.}
+The dashed curve is a least-squares log fit of all the points except
+those from {\tt dgesdd$_1$}, which we have excluded for reasons
+explained in the text.  [[Say what the colored region is and why we
+    included it.]]}
   \label{fig:wpe_vs_mase_best}
 \end{figure}
-As the plot shows, the best-case prediction error for each of these
-eight systems is roughly proportional to the weighted permutation
-entropy, which is consistent with our first conjecture.  The dashed
-line in the figure, a linear least-squares fit to these best-case
-errors\footnote{with the exception of {\tt dgesdd$_1$}, for reasons
-  described below} captures this rough proportionality.  This is not a
-formal result.  The three forecast methods used here were chosen to
-span the space of standard prediction strategies, but they do not
-cover that space exhaustively.  Our goal here is an empirical
-assessment of predictability and complexity, not formal results about
-a ``best'' predictor for a given time series.  There may be other
-methods that produce lower MASE values than those in
-Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points in
-the upper-left and lower-right quadrants of this plot strongly
-suggests that the underlying predictability of a time series is
-inversely correlated with its WPE.  The rest of this section describes
-our results in more detail---including the measures taken to assure
-meaningful comparisons across methods, trials, and programs---and
-elaborates on the meaning of the dashed line in the figure.
+As the plot shows, the [[best-case]] prediction error for each of
+these eight systems is roughly proportional to the log of the weighted
+permutation entropy, which is consistent with our first conjecture.
+The dashed line in the figure, a least-squares fit of $y = a \log(b x
++ c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
+  for reasons described below} captures this rough
+proportionality\footnote{The logarithmic fit is appropriate here
+  because any signal with zero entropy should be perfectly
+  predictable, and so should ideally have a MASE value of zero.}.
+This is not a formal result.  The forecast methods and data sets used
+here were chosen to span the space of standard prediction strategies
+and the range of dynamical behaviors, but they do not cover those
+spaces exhaustively.  Our goal here is an\emph{empirical} assessment
+of the relationship between predictability and complexity, not formal
+results about a ``best'' predictor for a given time series.  There may
+be other methods that produce lower MASE values than those in
+Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
+above and below the curve in this plot strongly suggests a pattern of
+(inverse) correlation between the underlying predictability of a time
+series and its WPE.  The rest of this section describes our results in
+more detail---including the measures taken to assure meaningful
+comparisons across methods, trials, and programs---and elaborates on
+the meaning of the dashed curve in the figure.
 
@@
 %    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
-  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
+  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
 \caption{WPE vs. MASE for all trials, methods, and systems---with the
@@
 Numerical values, including means and standard deviations of the
-errors, can be found in Table~\ref{tab:error}.  The dashed line is the
+errors, can be found in Table~\ref{tab:error}.  The dashed curve is the
 same as in the previous figure.
@@
 \end{figure}
-There are 15 points in each cluster, one for each trial.  (The points
-in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
-for the corresponding system in any of the three plots in
-Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
-much across trials.  For most traces, the variance in MASE scores is
-low as well, resulting in small, tight clusters.  In some
+There are 15 points in each cluster, one for each trial.  [[(The
+    points in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of
+    the points for the corresponding system in any of the three plots
+    in Figure~\ref{fig:wpe_vs_mase_all}.)]]  The WPE values do not
+vary very much across trials.  For most traces, the variance in MASE
+scores is low as well, resulting in small, tight clusters.  In some
 cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
@@
 three plots.  The mean MASE scores of predictions generated with the
-nonlinear LMA method are generally closer to the dashed line; the
+nonlinear LMA method are generally closer to the dashed curve; the
 ARIMA method clusters are more widely spread, the \naive ~clusters
@@
 information transfer in the signal, but predictions with comparatively
-poor MASE scores---is obvious in the geometry of the top two images
-in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
-to the right of the dashed line.  \emph{This indicates that these
+poor MASE scores---is obvious in the geometry of the top two images in
+Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
+the right of the dashed curve.  \emph{This indicates that these
   methods are not leveraging the available information in the signal}.
 The dynamics of \col may be complicated, but they are not
-unstructured.  This signal is nonlinear and deterministic\cite{mytkowicz09}, and if one
-uses a prediction technique that is based a nonlinear model
-(LMA)---rather than a method that simply predicts the running mean
-(\naive) or one that uses a linear model (ARIMA)---the MASE score is
-much better: $0.050 \pm 0.001$.  This prediction is 20 times more
-accurate than a random-walk forecast, which is more in line with the
-level of predictive structure that the low WPE value suggests is
-present in the signal.
+unstructured.  This signal is nonlinear and
+deterministic\cite{mytkowicz09}, and if one uses a prediction
+technique that is based a nonlinear model (LMA)---rather than a method
+that simply predicts the running mean (\naive) or one that uses a
+linear model (ARIMA)---the MASE score is much better: $0.050 \pm
+0.001$.  This prediction is 20 times more accurate than a random-walk
+forecast, which is more in line with the level of predictive structure
+that the low WPE value suggests is present in the signal.
 
-The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
+The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
 and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
@@
 predictive structure than the method is able to use.  When the
-predictions are well below that line (as in the top and middle images
+predictions are well below that curve (as in the top and middle images
 in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
 method.  Conversely, the position of the LMA cluster---well to the
-left of the other two, near the dashed line---reflects this method's
+left of the other two, near the dashed curve---reflects this method's
 ability to capture and exploit the structure that is present in this
-signal.
+signal.  [[Again, this curve was determined from a finite set of
+    methods and data.  But we worked hard to make these experiments
+    representative and comprehensive.  Forecast methods from simple to
+    sophisticated; time-series data from a system whose behavior spans
+    the dynamical behavior space.  While we are cautiously opmistic
+    about its generality, more exploration will be required to firm up
+    that conclusion.  Say that Henon, Lorenz and SFI A fall within the
+    colored region too, and that we're working on a broader study.]]
 
@@
 vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
-because of the way that these models are built \cite{autoARIMA}.  If
-a KPSS test of the time series in question indicates that it is
+because of the way that these models are built \cite{autoARIMA}.  If a
+KPSS test of the time series in question indicates that it is
 nonstationary, an integration term is added to the model.  This test
@@
   the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
-  tight cluster.}.
+  tight cluster.}.  \alert{This paragraph probably needs more, in
+  response to the ref's gritch about fitting.}
 
@@
 transfer of the underlying process is lower, but that time-series data
-from this system still contain a significant amount of structure that
-can, in theory, be leveraged to predict the future course of the time
-series.  As mentioned above, though, the effectiveness of any
-prediction strategy depends on how well it leverages the available
-information: methods that use linear models, for instance, may fail
-when applied to nonlinear processes.
+observed from this system still contain a significant amount of
+structure that can, in theory, be leveraged to predict the future
+course of the time series.  As mentioned above, though, the
+effectiveness of any prediction strategy depends on how well it
+leverages the information that is available in the data: methods that
+use linear models, for instance, may fail when applied to nonlinear
+processes.
 %
@@
   \svdthree signal.}.  As before, the positions of these points on a
-WPE vs. MASE plot---significantly below and to the right of the
-dashed line---should suggest to a practitioner that there is more
-structure in this signal than the corresponding method is able to
-leverage, and that a different method might do better.  Indeed, for
-\svdfive, the LMA method produces a MASE score of $ 0.718\pm 0.048 $
-and a cluster of results that is much closer to the dashed line on the
-WPE-MASE plot.  Again, this validates our second conjecture: the LMA
-method can capture and reproduce the way in which the \svdfive system
-processes information, but the \naive ~and ARIMA prediction methods
-cannot.
+WPE vs. MASE plot---significantly below and to the right of the shaded
+region---should suggest to a practitioner that there is more structure
+in this signal than the corresponding method is able to leverage, and
+that a different method might do better.  Indeed, for \svdfive, the
+LMA method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster
+of results that is much closer to the dashed curve on the WPE-MASE
+plot.  Again, this validates our second conjecture: the LMA method can
+capture and reproduce the way in which the \svdfive system processes
+information, but the \naive ~and ARIMA prediction methods cannot.
 
@@
 
-The effects described in the previous paragraph are exacerbated by the
-way MASE is calculated.  Recall that MASE scores are scaled
+The effects described in the previous paragraph are also exacerbated
+by the way MASE is calculated.  Recall that MASE scores are scaled
 \emph{relative to a random-walk forecast}, which uses the current
@@
 in the denominator of the MASE score, this effect can shift points
-leftwards on a WPE vs. MASE plot, and that is exactly why the
-\svdone clusters for ARIMA and LMA are above the dashed line in
+leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
+clusters for ARIMA and LMA are above the dashed curve in
 Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
@@
 the \svdone MASE values.  For all of these reasons, we view \svdone as
-an outlier and exclude it from the fit calculation of the dashed line
+an outlier and exclude it from the fit calculation of the dashed curve
 in Figure~\ref{fig:wpe_vs_mase_best}.
+
+[[Short paragraph about how MASE has its drawbacks but is the best
+    overall choice here because it is scale free and can compare
+    across signals.  Point to the MASE paper for extensive evaluation
+    and comparison wrt other metrics. We are in the process of
+    repeating this study with other metrics, other forecast methods,
+    and time-series data from other systems.  Segue to next section.]]
 

### 2014-07-14 wpe.tex
fix the lower limit of a sum
second round of revisions.  awaiting today's conversation before can finish.

@@
 information in the present that is also in the past.  The second part
-is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
-the more redundancy in a signal, the more predictable it is.  
+is the aforementioned \emph{Shannon entropy rate}.  It seems obvious
+that the more redundancy in a signal, the more predictable it is.
 Redundancy is critical to prediction, in general.  And the specific
@@
 method cannot capture or make use of nonlinear redundancy, for
-instance.  This issue is central to the claims in this paper and the 
-discussion in the following section.
+instance.  This issue---which has been explored extensively using
+clean data and/or systems where the generating process is known, but
+not so much with empirical data from unknown systems---is central to
+the claims in this paper and the discussion in the following section.
 
@@
 is to correct for observational noise that is larger than the trends
-in the data, but smaller than the larger scale features.  Consider,
+in the data, but smaller than the larger-scale features.  Consider,
 for example, a signal that switches between two fixed points and
-contains some additive noise. The PE is dominated by the noise about
-the fixed points driving it to $\approx 1$, which in some sense hides
-the fact that the signal is actually quite structured.  In this
-situation, the terms of the weighted permutation entropy are dominated
-by the switching rather than the stochastic fluctuations. To
-accomplish this, the \emph{weight} of a permutation is taken into
-account:
+contains some additive noise. The PE of such a signal will be
+dominated by the noise about the fixed points, driving it to $\approx
+1$---which in some sense hides the fact that the signal is actually
+quite structured.  In this situation, the terms of the weighted
+permutation entropy are dominated by the switching rather than the
+stochastic fluctuations. To accomplish this, the \emph{weight} of a
+permutation is taken into account:
 \begin{align*}
-  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i}^{i+\ell}
+  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i + 1}^{i+\ell}
                       \left( x_j - \bar{x}_{i+1}^\ell \right)^2

### 2014-07-14 intro.tex

--- 2014-07-14 intro.tex
@@
 
+%<<<<<<< HEAD
+%Complicated time-series data are ubiquitous in modern scientific research.  The
+%complexity of these data spans a wide %range.  On the low end of this spectrum
+%are time series that exhibit perfect %predictive structure, i.e, signals whose
+%future values can be successfully %predicted from past values.  Signals like %this
+%can be viewed as the product of an %underlying process that generates %information
+%and/or transmits it from the past to the %future in a perfectly predictable
+%fashion.  Constant or periodic signals, %for example, fall in this class.  On the
+%opposite end of this spectrum are signals %that are what we call \emph{fully
+%complex}, where the underlying generating %process transmits no information at
+%all from the past to the future.  White %noise processes fall in this class.  In
+%fully complex signals, knowledge of the %past gives no insight into the future,
+%regardless of what model one chooses to %use. Signals in the midrange of this
+%spectrum---e.g., deterministic chaos---%pose interesting challenges from a
+%modeling perspective.  In these signals, %enough information is being transmitted
+%%from the past to the future that an \emph{ideal} model---one that captures the
+%generating process---can forecast the future behavior of the observed system
+%with high accuracy. Though this notion of complexity differs from that
+%considered by e.g. \cite{Shalizi2008}, which would consider a time series
+%without any statistical regularities to be non-complex, it has been extensively
+%considered by Boffetta et al~\cite{boffetta02}.
+%
+%As a corollary of the undecidability of the halting
+%problem~\cite{halting-problem}, however, no single forecasting schema
+%is ideal for all noise-free deterministic
+%signals~\cite{weigend93}---let alone all real-world time-series
+%data sets.  This leads naturally to an important and challenging
+%question: given a complicated real-valued time series, does there
+%exist a forecast model that can leverage the information (if any) that
+%is being transmitted from past to future by the underlying generating
+%process?  A first step in answering this question is to reliably
+%quantify where on the complexity spectrum a given time series falls; a
+%second step is to determine how complexity and predictability are
+%related.  These are the goals of this paper.
+%
+%In particular, we wish to determine---without knowing anything about
+%the generating process---whether a time series is too complex to successfully
+%predict.  An important practical corollary to this is a strategy for
+%assessing appropriateness of forecast methods.  If the forecast
+%produced by a particular method is poor but the time series contains a
+%significant amount of predictive structure, one can reasonably conclude that that method is inadequate
+%to the task and that one should seek another method.
+%=======
 Complicated time-series data are ubiquitous in modern scientific
@@
 % series is too complex to successfully predict.
+%>>>>>>> %1574864f4a0a007a6917eac812688ba02a5de1d5
 

### 2014-07-14 concl.tex
forgot a few minor changes

@@
 signal with zero entropy should be perfectly predictable, and so
-should ideally have a MASE value of zero.  Supporting our hypothesis,
-forecast errors for the \col and \svdfive signals are higher than the
+should ideally have a MASE value of zero.  [[Are there other reasons
+    this shape makes sense?]]  Supporting our hypothesis, forecast
+errors for the \col and \svdfive signals are higher than the
 corresponding WPEs suggest---except for the nonlinear LMA predictor.
@@
 %  \naive ~wins on \svdone with the exception to the 5 outliers.
-The curve in Figures~\ref{fig:wpe_vs_mase_best}
-and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
-paragraph.  It is a preliminary, but potentially useful, empirical
+The dashed curve and shaded region in
+Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
+operationalize the discussion in this paragraph.  These geometric
+features are a preliminary, but potentially useful, empirical
 guideline for knowing when a model is not well-matched to the task at
-hand: a MASE score that is significantly higher than [[$3.58 \cdot
-    \mathrm{WPE} - 1.91$]], where $\mathrm{WPE}$ is that of the
-corresponding signal, is an indication that that time series has more
-predictive structure than that forecast model can capture and exploit.
+hand: [[a MASE score that is significantly higher than $3.58 \cdot
+    \mathrm{WPE} - 1.91$, where $\mathrm{WPE}$ is that of the
+    corresponding signal]] [[need to reword]], is an indication that
+that time series has more predictive structure than that forecast
+model can capture and exploit.
 
@@
 important next step---and one that will be critical to validating our
-claims about the utility of the dashed line as a heuristic for
-determining whether a particular forecast method is well-matched to a
-particular task.  Of particular interest would be the class of
-so-called \emph{hybrid systems}~\cite{hybrid}, which exhibit discrete transitions
-between different continuous regimes---e.g., a lathe that has an
-intermittent instability or traffic at an internet router, whose
-characteristic normal traffic patterns shift radically during an
-attack.  Effective modeling and prediction of these kinds of systems
-is quite difficult; doing so adaptively and automatically---in the
-manner that is alluded to at the end of the previous paragraph---would
-be an even more important challenge.
+claims about the utility of the shaded region around the dashed curve
+in Figure [[what]] as a heuristic for determining whether a particular
+forecast method is well-matched to a particular task.  Of particular
+interest would be the class of so-called \emph{hybrid
+  systems}~\cite{hybrid}, which exhibit discrete transitions between
+different continuous regimes---e.g., a lathe that has an intermittent
+instability or traffic at an internet router, whose characteristic
+normal traffic patterns shift radically during an attack.  Effective
+modeling and prediction of these kinds of systems is quite difficult;
+doing so adaptively and automatically---in the manner that is alluded
+to at the end of the previous paragraph---would be an even more
+important challenge.
 

### 2014-07-14 pecomplexity.tex
forgot a few minor changes

@@
 This work was partially supported by NSF grant \#CMMI-1245947 and ARO
-grant \#W911NF-12-1-0288.
-We would also like to acknowledge the Santa Fe Institute whose interdisciplinary halls created a place for this work to be born.
+grant \#W911NF-12-1-0288.  We would also like to acknowledge the Santa
+Fe Institute whose interdisciplinary halls created a place for this
+work to be born, and the anonymous reviewers whose comments
+strengthened the presentation.
 

### 2014-07-14 results.tex
forgot a few minor changes

@@
 + c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
-  for reasons described below} captures this rough
-proportionality\footnote{The logarithmic fit is appropriate here
-  because any signal with zero entropy should be perfectly
-  predictable, and so should ideally have a MASE value of zero.}.
-This is not a formal result.  The forecast methods and data sets used
-here were chosen to span the space of standard prediction strategies
-and the range of dynamical behaviors, but they do not cover those
-spaces exhaustively.  Our goal here is an\emph{empirical} assessment
-of the relationship between predictability and complexity, not formal
-results about a ``best'' predictor for a given time series.  There may
-be other methods that produce lower MASE values than those in
-Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
+  for reasons described below} captures this relationship between WPE
+and [[best-case]] prediction error\footnote{The logarithmic fit is
+  appropriate here because any signal with zero entropy should be
+  perfectly predictable, and so should ideally have a MASE value of
+  zero.}.  This is not a formal result.  The forecast methods and data
+sets used here were chosen to span the space of standard prediction
+strategies and the range of dynamical behaviors, but they do not cover
+those spaces exhaustively.  Our goal here is an\emph{empirical}
+assessment of the relationship between predictability and complexity,
+not formal results about a ``best'' predictor for a given time series.
+There may be other methods that produce lower MASE values than those
+in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
 above and below the curve in this plot strongly suggests a pattern of
@@
 Numerical values, including means and standard deviations of the
-errors, can be found in Table~\ref{tab:error}.  The dashed curve is the
-same as in the previous figure.
-}
+errors, can be found in Table~\ref{tab:error}.  The dashed curve [[and
+    shaded regions?  check in all captions]] is the same as in the
+previous figure.  }
     \label{fig:wpe_vs_mase_all}
@@
 KPSS test of the time series in question indicates that it is
-nonstationary, an integration term is added to the model.  This test
-gives mixed results in the case of the \col process, flagging five of
-the 15 trials as stationary and ten as nonstationary.  ARIMA models
-without an integration term perform more poorly on this signal, which
-increases the error, thereby spreading out the points\footnote{We
-  tested this hypothesis by forcing the inclusion of an integration
-  term in the five cases where a KPSS test indicated that such a term
-  was not needed.  This action removed the spread, pushing all 15 of
-  the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
-  tight cluster.}.  \alert{This paragraph probably needs more, in
-  response to the ref's gritch about fitting.}
+nonstationary, the ARIMA recipe adds an integration term to the model.
+This test gives mixed results in the case of the \col process,
+flagging five of the 15 trials as stationary and ten as nonstationary.
+ARIMA models without an integration term perform more poorly on this
+signal, which increases the error, thereby spreading out the
+points\footnote{We tested this hypothesis by forcing the inclusion of
+  an integration term in the five cases where a KPSS test indicated
+  that such a term was not needed.  This action removed the spread,
+  pushing all 15 of the \col ARIMA points in
+  Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.}.
+\alert{This paragraph probably needs more, in response to the ref's
+  gritch about fitting.}
 
@@
 observed from this system still contain a significant amount of
-structure that can, in theory, be leveraged to predict the future
-course of the time series.  As mentioned above, though, the
-effectiveness of any prediction strategy depends on how well it
-leverages the information that is available in the data: methods that
-use linear models, for instance, may fail when applied to nonlinear
-processes.
+structure that can, in theory, be used to predict the future course of
+the time series.  As mentioned above, though, the effectiveness of any
+prediction strategy depends on how well it leverages the information
+that is available in the data: methods that use linear models, for
+instance, may fail when applied to nonlinear processes.
 %
@@
 
-[[Short paragraph about how MASE has its drawbacks but is the best
-    overall choice here because it is scale free and can compare
-    across signals.  Point to the MASE paper for extensive evaluation
-    and comparison wrt other metrics. We are in the process of
-    repeating this study with other metrics, other forecast methods,
-    and time-series data from other systems.  Segue to next section.]]
+[[Even though MASE has weaknesses, we concur with \cite{mase} that it
+    is an effective way to compare prediction error across time series
+    of different lengths and different scales.  Point to that paper
+    for extensive evaluation and comparison wrt other metrics. We are
+    in the process of repeating our own study with other metrics,
+    other forecast methods, and time-series data from other systems.
+    Segue to next section.]]
 

### 2014-07-16 response.tex
response update
second round of LizB edits

@@
 discussed in a later section of this response.)  The image below is an
-augmented version of our revised Figure \alert{[[fill in]]} that
-includes all of these examples.
+augmented version of our \alert{revised Figure 6 [[This needs to be
+      regenerated with the new error bars]]} that includes all of
+these examples.
 
@@
 
-\noindent The dashed-line fit in this plot (and in Figure
-\alert{[[fill in]])} is based solely on the computer performance data.
-Nonetheless, the points from the other time series fall within a
-narrow band around the fit.  Note that this fit ($y = a \log(b x +
-c)$) is different than in the submitted version.  A log fit is more
-appropriate because any signal with zero entropy should be perfectly
-predictable, and so should ideally have a MASE value of zero.  Note,
-too, that there are a lot more points on this plot than in our
-original version; this is discussed further below.
+\noindent The dashed curve in this plot (and in Figure 6) was fit
+\emph{solely on the computer performance data}.  Nonetheless, the
+points from the other time series fall within a narrow band around the
+fit.  Note that this fit ($y = a \log(b x + c)$) is different than in
+the submitted version.  A careful examination of the data, prompted by
+the reviewer's questions, suggested that this was a better match to
+the data; theoretically, too, a log fit is more appropriate because
+any signal with zero entropy should be perfectly predictable, and so
+should ideally have a MASE value of zero.  Finally, note that there
+are a lot more points on this plot than in our original version; this
+is discussed further below.
+
+\smallskip
 
@@
 
-We have modified the manuscript to make it more clear that our goal is
-to derive a practical criterion for best-case prediction error in a
-model-free way, and to test it in the context of empirical data.
+We have shortened the introduction and modified the manuscript
+throughout to make it more clear that our goal is to derive a
+practical criterion for best-case prediction error in a model-free
+way, and to test it in the context of empirical data.
+
+\smallskip
 
@@
 
-We agree that these claims are both trivial and reasonably well known,
-in theory and for toy models.  Our specific claim, which was not well
-explained in the submitted version, is that those claims also
-hold---and can be practically applied---in real-world time-series data
-that may contain dynamical and/or observational noise.  This has not,
-to our knowledge, received much attention in the literature.  We have
-added a paragraph to the introduction that hopefully makes all of this
-more clear.
+These claims have indeed been explored extensively in theory and in
+toy models.  Our specific claim, which was not well explained in the
+submitted version, is that those claims also hold---and can be
+practically applied---in real-world time-series data that may contain
+dynamical and/or observational noise.  This has not, to our knowledge,
+received much attention in the literature.  We have added a paragraph
+to the introduction that hopefully makes all of this more clear.
 
-\alert{We appreciate being informed about [Boffetta et al.]  paper,
-  and we have included a discussion of it in \alert{Section [[what]].
-    The discussion is in the wrong place and doesn't make logical
-    sense.  Needs discussion.}
+We appreciate being informed about the Boffetta et al.  paper, and we
+have included a discussion of it at the end of the related-work
+section.
+
+\smallskip
 
@@
 We have added a short discussion about the differences in notions and
-definitions of complexity, and clarified the working definition that
-we use in this paper.
+definitions of complexity to the introduction and related-work
+section, and clarified the working definition that we use in this
+paper.
+
+\smallskip
 
@@
 
-The original wording of this finding did make it sound trivial.  Our
-intent, as we have modified the text to reflect, is that the
+The original wording of this finding did indeed make it sound trivial.
+Our intent, as we have modified the text to reflect, is that the
 appropriateness of a predictor can be discovered \emph{empirically},
-via comparison with the WPE.  Previous explorations have used clean
-data and/or systems where the generating process is known.  Our work
-makes neither of these assumptions.  The paragraph that we have added
-to the introduction (immediately following these two claims) should
-hopefully make this more clear.  \alert{Ryan said: ``i'm not entirely
-  sure how to respond to this one correctly.''  Liz asks: do you mean
-  in the rebuttal or in the text?}
+via comparison of its MASE value with the WPE.  Previous explorations
+of predictability have used clean data and/or systems where the
+generating process is known.  Our work makes neither of these
+assumptions.  \alert{[[Sentence from Josh here about why that's
+      important: given a time series from an unknown system, there's
+      no way a priori to know what model is right.  Point to the col
+      example in the text as an example of this.  But need to check
+      after inserting it to make sure that this point has not been
+      belabored too much already in this letter.]]}  The paragraph
+that we have added to the introduction (immediately following these
+two claims) should hopefully make this more clear.
+
+\smallskip
 
@@
 
-As mentioned in conjunction with the image above, the relationship
-posited in our paper does indeed hold for paradigmatic systems like
-the Henon map and the Lorenz equations.  We had not included these
-results in our paper because they simply confirmed well-known results,
-and because our focus was on confirming those results in the context
-of noisy, real-world data.  \alert{We have added a sentence in the
-  conclusion (or maybe end of discussion) about the fact that the WPE
-  vs. MASE relationship for these synthetic examples follows the
-  empirically determined pattern in our paper.}
+Computers are indeed complicated objects, but we have studied their
+dynamics extensively and established quite firmly that they are
+garden-variety dynamical systems with surprisingly low-dimensional
+dynamics [Mytkowicz et al. CHAOS, 2009].  Standard nonlinear
+time-series analysis techniques on performance data measured from
+running computers produce clean and consistent results, as mentioned
+above: fixed points, bifurcations, chaotic attractors, and so on.
+
+Complicated, unknown systems are where prediction really
+matters---especially systems where a good prediction can make a
+practical difference (e.g., reallocation of resources in a computer to
+suit the dynamically changing needs of a program).  Well-known systems
+are certainly good testcases.  As is also mentioned above, the
+relationship posited in our paper does indeed hold for paradigmatic
+systems like the Henon map and the Lorenz equations.  We had not
+included these examples in our paper because they simply confirmed
+well-known results, and because our focus was on confirming those
+results in the context of noisy, real-world data.  \alert{[[Liz to do,
+      then finish this wording: We have added a sentence in the
+      conclusion (or maybe end of discussion) about the fact that the
+      WPE vs. MASE relationship for these synthetic examples follows
+      the empirically determined pattern in our paper.]]}
+
+\smallskip
 
@@
 
+\smallskip
+
 \emph{7) As for the choice of the four prediction techniques, some of
@@
 
-\alert{We need to discuss this.  We also need to discuss whether this
-  response requires changes in the text.It is certainly true that the
-  naive and random-walk predictors can be seen as subclasses of the
-  ARIMA model. We have explicitly included them, however, since they
-  are used (sometimes to great success) in practice. Further, we have
-  used standard methods of fitting the parameters used in our ARIMA
-  predictions, and we believe any inefficiencies of that parameter
-  selection support our findings --- namely, if a particular selection
-  of model parameters predicts poorly (compared to what the WPE value
-  might suggest), it is an indication that the specific model,
-  including the mechanism for selecting parameter values, is not
-  optimal for the job. This might mean that a different parameter
-  selection scheme is needed, or that the entire model is
-  inappropriate.  Unfortunately the WPE value alone is insufficient to
-  differentiate these two situations.  It can identify both, but
-  cannot distinguish.}
+It is certainly true that the naive and random-walk predictors can be
+seen as subclasses of the ARIMA model family.  We have explicitly
+included them, however, because they are used (sometimes to great
+success) in practice and because that made for a more comprehensive
+exploration.
 
-\emph{8) Conditions for inefficiency of simple prediction schemes (like naive
-and random-walk) discussed in the paper look rather obvious as well and do not
-give new information. Is it the purpose of the paper to discuss when and where a
-concrete simple prediction scheme fails?}
+The ``AutoARIMA'' method that was used to build all of the ARIMA
+models in this paper---another ``off the shelf'' method in the
+modeling and prediction communities---follows the procedure outlined
+in Section \alert{[[blah]]} to choose values for the model parameters.
+Sometimes the choices made by that procedure (e.g., setting a
+particular parameter to zero) produce a naive or random-walk
+predictor.  Our method can help practioners know whether this is a
+good or bad thing.  If a particular model performs poorly (compared to
+what the WPE value might suggest), it may be an indication that the
+chosen parameter values were suboptimal---or even that no member of
+the ARIMA family will work and one should try a nonlinear model.
+
+% Unfortunately the WPE value alone is insufficient to differentiate
+% these two situations.  It can identify both, but cannot distinguish.
+
+We have \alert{Someone needs to do this: [[changed the term used in
+      the paper to AutoARIMA]]} in order to make it clear that this
+refers not to the general class of these models, but to a particular
+member of that class chosen via a particular procedure.
+
+\smallskip
+
+\emph{8) Conditions for inefficiency of simple prediction schemes
+  (like naive and random-walk) discussed in the paper look rather
+  obvious as well and do not give new information. Is it the purpose
+  of the paper to discuss when and where a concrete simple prediction
+  scheme fails?}
 
 This short paragraph in this paper, which explains the efficiencies
-and inefficiencies of these simple prediction schemes, was present for
-two purposes: first, to explain why we have included these simple
-schemes rather than just the more complicated ones; second, to help in
-explaining the specific results described later in the paper.
+and inefficiencies of these schemes, was present for two purposes:
+first, to explain why we have included these simple prediction schemes
+and not just the more-complicated ones; second, to help in explaining
+the specific results described later in the paper.
+
+\smallskip
 
@@
 
-We have extended the justification for our use of the MASE metric.
-\alert{did we?  need to check.}  The paper in which MASE was
-introduced [Hyndman and Koehler, 2006] (our reference \alert{[[fill
-      in]])} includes an extensive evaluation of this metric, as well
-as a comprehensive comparison between it and other metrics.  If the
-reviewer thinks that it would strengthen our paper, we can
-recapitulate more of that material in our Section IVD.<  As part of our
-current work on this topic, we are evaluating how other choices of
-metric affect the shape of our curves, but we feel that a
-comprehensive study of this issue is best left as a sequel to this
-work.
+\alert{Liz to do: We have extended the justification for our use of
+  the MASE metric.}  The paper in which MASE was introduced---[Hyndman
+  and Koehler, 2006], our reference \alert{[[fill in]]}---includes an
+extensive evaluation of this metric, as well as a comprehensive
+comparison to other metrics.  If the reviewer thinks that it would
+strengthen our paper, we can recapitulate more of that material in our
+Section IVD. As part of our current work on this topic, we are
+evaluating how other choices of metric affect the shape of our curves,
+but we feel that a comprehensive study of this issue is best left as a
+sequel to this work.
+
+\smallskip
 
@@
 
-\alert{This needs discussion.  It is certainly true that WPE values
-  depend on wordlength.  This is a real issue, particularly because
-  word length choices are rarely justified in the PE literature.
-  Quote from a few papers.  The right thing to do, as we argued in
-  section n, is to compute PE across a range of word lengths and look
-  for a flat spot in the curve.  A systematic approach like this
-  mitigates, but of course does not remove, any variation in the
-  results.  We used that method to choose individual l values for each
-  time series studied here.  We believe that this systematic approach
-  to l choice mitigates the differences in the WPE vs. MASE results
-  between time series; we have not yet confirmed that conjecture but
-  we are in the process of doing so.
+WPE values depend not only on wordlength, but also on \emph{data}
+length---since one needs more data to get good statistics on the words
+for the PE calculation.  These two interacting knobs are not easy to
+set.  Perhaps as a result of this, wordlength choices are rarely
+justified in the permutation entropy literature.  \alert{[[Can you
+      guys please extract a few representatively bad quotes from some
+      of the PE papers and put them in here?]]}
 
-Ryan had written: ``the specific relationship between WPE and MASE
-values is dependent on word length, though for any reasonable choice
-of $\ell$, we believe a similar relationship to hold.'' Liz is not
-sure she understands what ``similar'' means or why that's justified.}
+The right thing to do, as we argued in Section [[what]], is to compute
+PE across a \emph{range} of wordlengths ($\ell$) and look for a flat
+spot in the curve of PE versus $\ell$.  We used that method to choose
+individual $\ell$ values for each time series studied here.  We
+conjecture that a systematic approach like sill mitigate any variation
+in the results, thereby preserving the shape of the WPE vs. MASE
+curve.  We have not yet confirmed that conjecture, but we are in the
+process of doing so.
 
@@
 a range of time-series data sets whose behavior spans a wide range of
-dynamical behaviors.  Recapitulate that henon, lorenz, SFI A all fit
-the curve determined by the computer perf data; cite as justification
-for generality and point to future work.
+dynamical behaviors.  For all of these examples---as well as for the
+three additional ones constructed for this rebuttal letter (Henon,
+Lorenz, SFI A)---the pattern (Figure 6) was consistent.  This strongly
+suggests, but of course does not prove, that our approach is general.
+As the reviewer suggested, this will require extensive study and
+evaluation in future work.
+
+\smallskip
 
@@
 
-Both WPE and MASE are sensitive to changes of variables; our initial
-testing, however, indicates they are so \emph{in complementary ways},
-so the shape of the WPE vs. MASE curve is [[largely invariant with
-    respect to changes in what]] \alert{Ryan wrote: ``This leads a
-  change in variables to...''  What should go in here?  What variables
-  is he talking about?  Did we actually do that test (varying them)?
-  What should we say about that word ``largely''?}
+\alert{response and modification of text both are pending experiments}
+
+PE is invariant under an invertible nonlinear transformation of the
+values in the time series, but WPE is not.  But MASE varies \emph{in a
+  complementary way}.  Ryan will do $x(t)$ and $5e^{x(t)}$ and show
+that it's on the curve.  This is consistent with our claim of
+generality; this is just another time series.  A xform like this
+effectively changes the generating process, but our model-free result
+handles it.
 
@@
 
-It is our hope that between the additional preliminary results here,
-the modifications to the paper, as well as our clarifying the goals of
-our paper (understandably misunderstood in the draft), that the
-reviewer now considers the manuscript appropriate for publication in
-PRE.
+It is our hope that these clarifications, the attendant modifications
+to the paper, and the additional preliminary results that we included
+in this rebuttal letter have convinced the reviewer that this
+manuscript is appropriate for publication in PRE.
 

### 2014-07-16 concl.tex
second round of LizB edits

@@
 higher the WPE, in general, the higher the prediction error.  The
-relationship is roughly logarithmic, which makes sense because any
-signal with zero entropy should be perfectly predictable, and so
-should ideally have a MASE value of zero.  [[Are there other reasons
-    this shape makes sense?]]  Supporting our hypothesis, forecast
-errors for the \col and \svdfive signals are higher than the
-corresponding WPEs suggest---except for the nonlinear LMA predictor.
-There are some exceptions: \svdone is easier to predict than its WPE
-value suggests using any of the three forecast strategies studied
-here.  On the whole, though, signals with higher weighted permutation
-entropy are harder to predict.
+relationship is roughly logarithmic.  [[This is just an observed result.
+It makes sense because, firstly, any signal with zero entropy should
+be perfectly predictable, and so should ideally have a MASE value of
+zero.  As the entropy rises, the MASE value rise: slowly at first
+because the signal still has lots of info, then more rapidly because
+[[why]], and finally saturating at [[what]] because of [[why]].
+Supporting our hypothesis, forecast errors for the \col and \svdfive
+signals are higher than the corresponding WPEs suggest---except for
+the nonlinear LMA predictor.  There are some exceptions: \svdone is
+easier to predict than its WPE value suggests using any of the three
+forecast strategies studied here.  On the whole, though, signals with
+higher weighted permutation entropy are harder to predict.
 
@@
 on noisy signals because it effects a filtering operation.  The simple
-random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
-the \gcc signal, which is extremely complex---i.e., extremely low
-redundancy.
+random-walk strategy outperforms LMA, AutoARIMA, and the \naive
+~method on the \gcc signal, which is extremely complex---i.e.,
+extremely low redundancy.
 %  \naive ~wins on \svdone with the exception to the 5 outliers.

### 2014-07-16 intro.tex
second round of LizB edits

@@
 %\begin{itemize}
-%\item Complexity need not be hard to predict (can %point at the simple predictions paper) [[move to %introduction]]
-%\item random walk for example is best predicted by %guess what just happend[[move to introduction]]
-%\item The kind of complexity present matters, i.e., %that is whether the complexity is structured or not.%[[use here and mention in intro]]
+%\item Complexity need not be hard to predict (can %point at the simple predictions paper) 
+%\item random walk for example is best predicted by %guess what just happend
+%\item The kind of complexity present matters, i.e., %that is whether the complexity is structured or not.%
 
@@
 %predict.  What benefits would accrue if we could do so: power mgmt,
-%end world hunger [[this is my primary goal everyday :)]], etc.
+%end world hunger [this is my primary goal everyday :)], etc.
 %\end{it}
@@
 
-{\color{red} Though this notion of complexity differs from that
-  considered by e.g. \cite{Shalizi2008}, which would consider a time
-  series without any statistical regularities to be non-complex, it
-  has been extensively considered by Boffetta et al~\cite{boffetta02}.
-  I still don't understand how this sentence both addresses the
-  reviewer's pointing us to this paper and flows, logically.  Either
-  way, it is out of place.  It should be in a part of this paper that
-  gives citations, not one that lays out context in a general manner,
-  without citations.}
-
 % As a corollary of the undecidability of the halting
@@
 (which exhibits high entropy) as purely complex, whereas a low-entropy
-periodic signal is on the low end of the complexity spectrum.  We
-argue that an extension of \emph{permutation
-  entropy}~\cite{bandt2002per}---a method for approximating the
-entropy through ordinal analysis---is an effective way to assess the
-complexity of a given time series.  Permutation entropy is ideal for
-our purposes because it works with real-valued data and is known to
-converge to the true entropy value. Other existing techniques either
-require specific knowledge of the generating process or produce biased
-values of the entropy~\cite{bollt2001}.
+periodic signal is on the low end of the complexity spectrum.
+\alert{This differs from the notion of complexity used by
+  e.g. \cite{Shalizi2008}, which would consider a time series without
+  any statistical regularities to be non-complex.  Liz moved this
+  here.  Is there a better place for it?}  We argue that an extension
+of \emph{permutation entropy}~\cite{bandt2002per}---a method for
+approximating the entropy through ordinal analysis---is an effective
+way to assess the complexity of a given time series.  Permutation
+entropy is ideal for our purposes because it works with real-valued
+data and is known to converge to the true entropy value. Other
+existing techniques either require specific knowledge of the
+generating process or produce biased values of the
+entropy~\cite{bollt2001}.
 
@@
 
-%[[Introduce and make it clear why we choose comp. perf. as test bed.]]
+%[Introduce and make it clear why we choose comp. perf. as test bed.]
 

### 2014-07-16 model.tex
second round of LizB edits

@@
 \item The \emph{ARIMA} (auto-regressive integrated moving average)
-  method, a common linear forecast strategy, and
+  method, a common linear forecast strategy, built using the
+  \emph{AutoARIMA} procedure \cite{arima}, and
 
@@
 \end{itemize}
-ARIMA is based on standard autoregressive techniques.  LMA is designed
-to capture and exploit the deterministic structure of a signal from a
-nonlinear dynamical system.  The \naive ~and random-walk methods,
-somewhat surprisingly, often outperform these more-sophisticated
-prediction strategies in the case of highly complex signals, as
-discussed below.
+ARIMA models are based on standard autoregressive techniques.  LMA is
+designed to capture and exploit the deterministic structure of a
+signal from a nonlinear dynamical system.  The \naive ~and random-walk
+methods, somewhat surprisingly, often outperform these
+more-sophisticated prediction strategies in the case of highly complex
+signals, as discussed below.
 
@@
 data, or its temporal patterns, would be far more effective in the
-situations described in the previous paragraph.  The ARIMA and LMA
+situations described in the previous paragraph.  The AutoARIMA and LMA
 methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
@@
 %= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
-%that is fit to the past $p+1$ [[???]] values of $x_i$ via the
+%that is fit to the past $p+1$ [???] values of $x_i$ via the
 %backshift operator $B$ \footnote{I changed the superscript to $k$ to
@@
 %Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
-%and $Q$, respectively [[which accomplish what purpose?  why?  how?
-%    where did $D$ come from and what does it do?]].  $m$ is the
-%seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
+%and $Q$, respectively [which accomplish what purpose?  why?  how?
+%    where did $D$ come from and what does it do?].  $m$ is the
+%seasonal frequency [shouldn't this be ``period''?].  [Explain why
 %    one should choose $P=Q=m$, since that's what the equation below
-%    implicitly does.]]  To use a seasonal ARIMA model, one first uses
+%    implicitly does.]  To use a seasonal ARIMA model, one first uses
 %the detrending term to remove nonstationarity: $$\hat{x_i} =
@@
 automated fitting techniques described in~\cite{autoARIMA} to
-accomplish this.  This procedure uses sophisticated methods---KPSS
-unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
+accomplish this, producing what we call an \emph{AutoARIMA model}.
+This procedure uses sophisticated methods---KPSS unit-root
+tests~\cite{KPSSunit}, a customization of the Canova-Hansen
 test~\cite{Canova1995}, and the Akaike information
-criterion~\cite{akaike1974}, conditioned on the maximum likelihood
-of the model fitted to the detrended data---to select good values for
-the free parameters of the ARIMA model.
+criterion~\cite{akaike1974}, conditioned on the maximum likelihood of
+the model fitted to the detrended data---to select good values for the
+free parameters of the AutoARIMA model.
 
@@
 predictions, depending on model order.  To sidestep this issue, we
-build forecasts in a stepwise fashion: i.e., fit the model to the
-existing data, use that model to perform a one-step prediction, rebuild it
-using the latest observations, and iterate until the desired
-prediction horizon is reached.  (For consistency, we take the same
-approach with the other three models in this study as well, even
+build forecasts in a stepwise fashion: i.e., fit the AutoARIMA model
+to the existing data, use that model to perform a one-step prediction,
+rebuild it using the latest observations, and iterate until the
+desired prediction horizon is reached.  (For consistency, we take the
+same approach with the other three models in this study as well, even
 though doing so amounts to artificially hobbling LMA.)
@@
     \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
-    \caption{\col \\ ARIMA}
+    \caption{\col \\ AutoARIMA}
     \label{fig:colARIMA}
@@
     \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
-    \caption{\gcc \\ ARIMA }
+    \caption{\gcc \\ AutoARIMA }
     \label{fig:gccARIMA}
@@
     \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
-    \caption{\svdfive \\ ARIMA }
+    \caption{\svdfive \\ AutoARIMA }
     \label{fig:svd5ARIMA}
@@
     \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
-    \begin{sideways}\large ARIMA\end{sideways} &
+    \begin{sideways}\large AutoARIMA\end{sideways} &
     \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
@@
 perfect prediction would lie on the diagonal.  LMA, for instance,
-generates a very accurate prediction of the \col data, while ARIMA
+generates a very accurate prediction of the \col data, while AutoARIMA
 does not.  Horizontal lines result when a constant predictor (e.g.,
@@
 For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
-two-dimensional embedding with $\tau=1$.  For ARIMA, the
+two-dimensional embedding with $\tau=1$.  For AutoARIMA, the
 correspondence is not quite as simple, since the $p_j$ values are
@@
 effect is largely the same\footnote{The temporal \emph{ordering} of
-  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
+  the points on the AutoARIMA $p_j$ vs. $c_j$ plot does not match that of
   a projected embedding of the time series, however.}.
@@
   \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
-  \hline\hline Signal & LMA MASE  & ARIMA MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
+  \hline\hline Signal & LMA MASE  & AutoARIMA MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
   \hline
@@
 Figure~\ref{fig:forecast-example}, one can see some obvious
-correspondences.  The average LMA MASE score for the \col signals
-was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
+correspondences.  The average LMA MASE score for the \col signals was
+$0.050$, for instance, while AutoARIMA scored much worse ($0.599$).
 That is, LMA performed roughly 20 times better on \col signals than a
-random-walk predictor, while ARIMA only outperformed random walk by a
-factor of 1.7.  This is in accordance with the visual appearance of
-the corresponding images in Figure~\ref{fig:forecast-example}.  While
-the correspondence between MASE score and the visual appearance of
-these kinds of plots is clear cut in this case, that is not always so.
-The plots of LMA predictions of \col and \svdfive both lie near the
-diagonal, for instance, but the corresponding MASE scores were very
-different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
-1.4 times better than random-walk forecasts of the same signals).
+random-walk predictor, while AutoARIMA only outperformed random walk
+by a factor of 1.7.  This is in accordance with the visual appearance
+of the corresponding images in Figure~\ref{fig:forecast-example}.
+While the correspondence between MASE score and the visual appearance
+of these kinds of plots is clear cut in this case, that is not always
+so.  The plots of LMA predictions of \col and \svdfive both lie near
+the diagonal, for instance, but the corresponding MASE scores were
+very different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20
+and 1.4 times better than random-walk forecasts of the same signals).
 This is a result of the normalization in the MASE score calculation.
@@
 the other hand, so the denominator is small and the MASE scores are
-skewed upwards.  Normalization is, as is often the case, a double-edged
-sword.  Here, it facilitates comparisons across signals of different
-types and lengths, but its particular pathologies must be taken into
-account, as discussed at more length in Section~\ref{sec:results}.
+skewed upwards.  Normalization is, as is often the case, a
+double-edged sword.  Here, it facilitates comparisons across signals
+of different types and lengths, but its particular pathologies must be
+taken into account, as discussed at more length in
+Section~\ref{sec:results}.
 
 % Table \ref{tab:error} provides the MASE distributions
-% {\color{red}[[Joshua: Ryan, Is this the right word? we give mean $\pm$
-%       std. dev but some have very skewed right tails]]} for each of
+% {\color{red}[Joshua: Ryan, Is this the right word? we give mean $\pm$
+%       std. dev but some have very skewed right tails]} for each of
 % the eight signals and three prediction strategies, these are averaged
@@
 
-%[[and cherry pick a few examples of \gcc and \col to put in the text ]]
+%[and cherry pick a few examples of \gcc and \col to put in the text ]
 
@@
 %  \end{subfigure}%
-%  \caption{{\color{red} [[If Liz thinks we should %include these I actually need to generate this %figure]]}An LMA-based forecast of
+%  \caption{{\color{red} [If Liz thinks we should %include these I actually need to generate this %figure]}An LMA-based forecast of
 %       processor-efficiency performance trace from the%\col ,\gcc and \svdfive generating processes.  Red %circles and blue $\times$s are the true and
@@
 %    \includegraphics[width=\textwidth]{colPredShortTS}
-%     \caption{{\color{red} [[actually need to generate this %figure]]}An LMA-based forecast of the last 4,000 points of a
+%     \caption{{\color{red} [actually need to generate this %figure]}An LMA-based forecast of the last 4,000 points of a
 %       processor-load performance trace from the \gcc

### 2014-07-16 related.tex
second round of LizB edits

@@
 
+The notion of complexity used here differs from that considered by
+e.g. \cite{Shalizi2008}, which would consider a time series without
+any statistical regularities to be non-complex.
+
 Using entropy and redundancy to quantify the inherent predictability
@@
 of systems.
+
+As a final note: there has been prior work under a similar title to
+ours~\cite{haven2005}, but there are only superficial similarities
+between the two research projects.  Haven \emph{et al.} utilize the
+relative entropy to quantify the difference in predictability between
+two distributions: one evolved from a small ensemble of past states
+using the known dynamical system, and the other the observed
+distribution. Our work quantifies the predictability of a single
+observed time series using weighted permutation entropy and makes no
+assumptions about the generating process.  More closely related is the
+work by Boffetta et al.~\cite{boffetta02}.  \alert{[[Sentences from
+      Ryan explaining exactly what the difference is.]]]

### 2014-07-16 results.tex
second round of LizB edits

@@
 Working from the first 90\% of each signal, we generated a prediction
-of the last 10\% using the \naive, ARIMA, and LMA prediction methods,
+of the last 10\% using the \naive, AutoARIMA, and LMA prediction methods,
 as described in Section~\ref{sec:accuracy}, then calculated the MASE
@@
 Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
-corresponding MASE values [[for all of these trials.]]
+corresponding MASE values of the best prediction of each time series.
 \begin{figure}[htbp]
@@
   \caption{Weighted permutation entropy vs. mean absolute scaled error
-    (MASE) of [[all trials for each system]].
+    (MASE) of the best prediction of each time series.
 %
@@
 \end{figure}
-As the plot shows, the [[best-case]] prediction error for each of
+As the plot shows, the best-case prediction error for each trace from
 these eight systems is roughly proportional to the log of the weighted
@@
   for reasons described below} captures this relationship between WPE
-and [[best-case]] prediction error\footnote{The logarithmic fit is
-  appropriate here because any signal with zero entropy should be
-  perfectly predictable, and so should ideally have a MASE value of
-  zero.}.  This is not a formal result.  The forecast methods and data
-sets used here were chosen to span the space of standard prediction
-strategies and the range of dynamical behaviors, but they do not cover
-those spaces exhaustively.  Our goal here is an\emph{empirical}
-assessment of the relationship between predictability and complexity,
-not formal results about a ``best'' predictor for a given time series.
-There may be other methods that produce lower MASE values than those
-in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
+and prediction error\footnote{The logarithmic fit is appropriate here
+  because any signal with zero entropy should be perfectly
+  predictable, and so should ideally have a MASE value of zero.}.
+This is not a formal result.  The forecast methods and data sets used
+here were chosen to span the space of standard prediction strategies
+and the range of dynamical behaviors, but they do not cover those
+spaces exhaustively.  Our goal here is an\emph{empirical} assessment
+of the relationship between predictability and complexity, not formal
+results about a ``best'' predictor for a given time series.  There may
+be other methods that produce lower MASE values than those in
+Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
 above and below the curve in this plot strongly suggests a pattern of
@@
 errors, can be found in Table~\ref{tab:error}.  The dashed curve [[and
-    shaded regions?  check in all captions]] is the same as in the
+    shaded regions?  need to check this throughout the text and in all
+    captions after we finalize the plot]] is the same as in the
 previous figure.  }
@@
 \end{figure}
-There are 15 points in each cluster, one for each trial.  [[(The
-    points in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of
-    the points for the corresponding system in any of the three plots
-    in Figure~\ref{fig:wpe_vs_mase_all}.)]]  The WPE values do not
-vary very much across trials.  For most traces, the variance in MASE
-scores is low as well, resulting in small, tight clusters.  In some
-cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
-variance is larger, which spreads out the clusters horizontally.  The
-distribution of the cluster \emph{positions} is quite different in the
-three plots.  The mean MASE scores of predictions generated with the
-nonlinear LMA method are generally closer to the dashed curve; the
-ARIMA method clusters are more widely spread, the \naive ~clusters
+There are 15 points in each cluster, one for each trial.  (The points
+in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
+for the corresponding trace in any of the three plots in
+Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
+much across trials.  For most traces, the variance in MASE scores is
+low as well, resulting in small, tight clusters.  In some
+cases---AutoARIMA predictions of {\tt col\_major}, for instance---the
+MASE variance is larger, which spreads out the clusters horizontally.
+The distribution of the cluster \emph{positions} is quite different in
+the three plots.  The mean MASE scores of predictions generated with
+the nonlinear LMA method are generally closer to the [[dashed curve]];
+the AutoARIMA method clusters are more widely spread, the \naive ~clusters
 even more so.  This geometric validation of the second conjecture in
@@
 graphical representations in the left-hand column of
-Figure~\ref{fig:forecast-example} that the \naive ~and ARIMA
+Figure~\ref{fig:forecast-example} that the \naive ~and AutoARIMA
 prediction methods do not perform as well on this signal as the
-nonlinear LMA method.  The MASE scores of the \naive ~and ARIMA
+nonlinear LMA method.  The MASE scores of the \naive ~and AutoARIMA
 predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
@@
 Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
-the right of the dashed curve.  \emph{This indicates that these
+the right of the [[dashed curve]].  \emph{This indicates that these
   methods are not leveraging the available information in the signal}.
@@
 that simply predicts the running mean (\naive) or one that uses a
-linear model (ARIMA)---the MASE score is much better: $0.050 \pm
+linear model (AutoARIMA)---the MASE score is much better: $0.050 \pm
 0.001$.  This prediction is 20 times more accurate than a random-walk
@@
 
-The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
+The [[dashed curve]] in Figures~\ref{fig:wpe_vs_mase_best}
 and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
@@
 predictive structure than the method is able to use.  When the
-predictions are well below that curve (as in the top and middle images
-in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
-method.  Conversely, the position of the LMA cluster---well to the
-left of the other two, near the dashed curve---reflects this method's
-ability to capture and exploit the structure that is present in this
-signal.  [[Again, this curve was determined from a finite set of
-    methods and data.  But we worked hard to make these experiments
-    representative and comprehensive.  Forecast methods from simple to
-    sophisticated; time-series data from a system whose behavior spans
-    the dynamical behavior space.  While we are cautiously opmistic
-    about its generality, more exploration will be required to firm up
-    that conclusion.  Say that Henon, Lorenz and SFI A fall within the
-    colored region too, and that we're working on a broader study.]]
+predictions are well below [[that curve]] (as in the top and middle
+images in Figure~\ref{fig:wpe_vs_mase_all}), one should try a
+different method.  Conversely, the position of the LMA cluster---well
+to the left of the other two, near the [[dashed curve]]---reflects
+this method's ability to capture and exploit the structure that is
+present in this signal.  [[Liz to wordsmith: Again, this curve was
+    determined from a finite set of methods and data.  But we worked
+    hard to make these experiments representative and comprehensive:
+    forecast methods from simple to sophisticated; time-series data
+    from a system whose behavior spans the dynamical behavior space.
+    While we are cautiously opmistic about its generality, more
+    exploration will be required to firm up that conclusion.  Say that
+    Henon, Lorenz and SFI A fall within the colored region too, and
+    that we're working on a broader study.]]
 
@@
 very tightly in the lower left quadrant of the LMA and \naive ~MASE
-vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
-because of the way that these models are built \cite{autoARIMA}.  If a
-KPSS test of the time series in question indicates that it is
-nonstationary, the ARIMA recipe adds an integration term to the model.
-This test gives mixed results in the case of the \col process,
-flagging five of the 15 trials as stationary and ten as nonstationary.
-ARIMA models without an integration term perform more poorly on this
-signal, which increases the error, thereby spreading out the
-points\footnote{We tested this hypothesis by forcing the inclusion of
-  an integration term in the five cases where a KPSS test indicated
-  that such a term was not needed.  This action removed the spread,
-  pushing all 15 of the \col ARIMA points in
+vs. WPE plots, but spread out horizontally in the AutoARIMA plot.
+This is because of the way that these models are built
+\cite{autoARIMA}.  If a KPSS test of the time series in question
+indicates that it is nonstationary, the AutoARIMA recipe adds an
+integration term to the model.  This test gives mixed results in the
+case of the \col process, flagging five of the 15 trials as stationary
+and ten as nonstationary.  ARIMA models without an integration term
+perform more poorly on this signal, which increases the error, thereby
+spreading out the points\footnote{We tested this hypothesis by forcing
+  the inclusion of an integration term in the five cases where a KPSS
+  test indicated that such a term was not needed.  This action removed
+  the spread, pushing all 15 of the \col AutoARIMA points in
   Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.}.
 \alert{This paragraph probably needs more, in response to the ref's
-  gritch about fitting.}
+  gritch about fitting.  We certainly want to call his attention to
+  it, either way.}
 
@@
 This match between model and task is an issue in the \svdfive
-experiments as well.  The MASE scores of the \naive ~and ARIMA
+experiments as well.  The MASE scores of the \naive ~and AutoARIMA
 predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
@@
 capture and reproduce the way in which the \svdfive system processes
-information, but the \naive ~and ARIMA prediction methods cannot.
+information, but the \naive ~and AutoARIMA prediction methods cannot.
 
@@
 fitting a hyperplane using least squares should have similar effects,
-the fact that LMA outperforms ARIMA ($1.530 \pm 0.021$ vs. $1.837 \pm
-0.016$) is somewhat counterintuitive.  However, the small amount of
-predictive structure that is present in this signal is nonlinear
+the fact that LMA outperforms AutoARIMA ($1.530 \pm 0.021$ vs. $1.837
+\pm 0.016$) is somewhat counterintuitive.  However, the small amount
+of predictive structure that is present in this signal is nonlinear
 (cf.,~\cite{mytkowicz09}), and LMA is designed to capture and exploit
@@
 methods do quite well on this signal, yielding MASE scores of $0.827
-\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (ARIMA).  This pushes the
+\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (AutoARIMA).  This pushes the
 corresponding clusters of points in the bottom two images in
@@
 top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
-score of 0.6, which is better than LMA and ARIMA.  In five of the 15
-\svdone trials, however, there were step changes in the signal.  The
-\naive ~method has a very difficult time with signals like this,
+score of 0.6, which is better than LMA and AutoARIMA.  In five of the
+15 \svdone trials, however, there were step changes in the signal.
+The \naive ~method has a very difficult time with signals like this,
 particularly if there are multiple step changes.  This raised the MASE
@@
 leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
-clusters for ARIMA and LMA are above the dashed curve in
+clusters for AutoARIMA and LMA are above the dashed curve in
 Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
@@
 
-[[Even though MASE has weaknesses, we concur with \cite{mase} that it
-    is an effective way to compare prediction error across time series
-    of different lengths and different scales.  Point to that paper
-    for extensive evaluation and comparison wrt other metrics. We are
-    in the process of repeating our own study with other metrics,
-    other forecast methods, and time-series data from other systems.
-    Segue to next section.]]
+[[Liz to wordsmith: Even though MASE has weaknesses, we concur with
+    \cite{mase} that it is an effective way to compare prediction
+    error across time series of different lengths and different
+    scales.  Point to that paper for extensive evaluation and
+    comparison wrt other metrics. We are in the process of repeating
+    our own study with other metrics, other forecast methods, and
+    time-series data from other systems.  Segue to next section.]]
 

### 2014-07-16 wpe.tex
second round of LizB edits

@@
 
-As a final note: there has been prior work under a similar title to
-ours~\cite{haven2005}, but there are only superficial similarities
-between the two research projects.  Haven \emph{et al.} utilize the
-relative entropy to quantify the difference in predictability between
-two distributions: one evolved from a small ensemble of past
-states using the known dynamical system, and the other the
-observed distribution. Our work quantifies the predictability of a
-single observed time series using weighted permutation entropy and
-makes no assumptions about the generating process.
-
 % was a good choice, because with $\ell \geq 7$ many signals had
@@
 %regimes.
-[[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
-Maybe point at some of the predicting predictability papers.]]
+[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist.
+Maybe point at some of the predicting predictability papers.]
 %
@@
 %%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
-%%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
+%%  \caption{[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]The weighted permutation entropy of %one run of SVD. The gray bands
 %%    are regions where the window overlaps regimes. The window size used is
@@
 %% all stuff about PE as regime-shift detection.
-%% [[should add that as future work, though.]]
+%% [should add that as future work, though.]
 %  \includegraphics[width=\columnwidth]{figs/SVD_wwpe}

### 2014-07-16 concl.tex
bunch of formatting issues

@@
 Preliminary results of this study show that MASE vs. WPE points for
-data from the SFI prediction competition \cite{sfi} fall in the shaded
+data from the SFI prediction competition \cite{weigend-book} fall in the shaded
 region on Figure [[what]]\footnote{Synthetic data from paradigmatic
-  systems like the Henon map \cite{henon} and the Lorenz system
+  systems like the H\'{e}non map \cite{henon} and the Lorenz system
   \cite{lorenz63} also fit this pattern.}.

### 2014-07-16 model.tex
bunch of formatting issues

@@
 this procedure, are shown in Figure~\ref{fig:embedding}.
- \begin{figure}
-   \centering
-\begin{subfigure}{\columnwidth}
+% \begin{figure}
+%   \centering
+%\begin{subfigure}{\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/colipc3d}
+%    \caption{\col }
+%    \label{fig:colEmbedding}
+%  \end{subfigure}%  \\
+%
+%    \begin{subfigure}{\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/gcc3dipc}
+%    \caption{\gcc}
+%    \label{fig:gccEmbedding}
+%  \end{subfigure}
+%  \\
+%  \begin{subfigure}{\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/svd53dipc2}
+%    \caption{\svdfive}
+%    \label{fig:svdfiveEmbedding}
+%  \end{subfigure}%
+%%
+%
+%
+%     %\includegraphics[width=\textwidth]{colipc3d}
+%     \caption{3D projections of delay-coordinate embeddings of the
+%       traces from (a) Figure~\ref{fig:col-ts} (b)
+%       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
+%       Figure~\ref{fig:svd-ts-colored}.}
+% \label{fig:embedding}
+% \end{figure}
+ \begin{figure}[!ht]
+    \subfloat[\col\label{fig:colEmbedding}]{%
       \includegraphics[width=\columnwidth]{figs/colipc3d}
-    \caption{\col }
-    \label{fig:colEmbedding}
-  \end{subfigure}%  \\
-
-    \begin{subfigure}{\columnwidth}
+    }
+    \hfill
+    \subfloat[\gcc\label{fig:gccEmbedding}]{%
       \includegraphics[width=\columnwidth]{figs/gcc3dipc}
-    \caption{\gcc}
-    \label{fig:gccEmbedding}
-  \end{subfigure}
-  \\
-  \begin{subfigure}{\columnwidth}
+    }
+        \hfill
+    \subfloat[\svdfive \label{fig:svdfiveEmbedding}]{%
       \includegraphics[width=\columnwidth]{figs/svd53dipc2}
-    \caption{\svdfive}
-    \label{fig:svdfiveEmbedding}
-  \end{subfigure}%
-%
-
-     %\includegraphics[width=\textwidth]{colipc3d}
+    }
     \caption{3D projections of delay-coordinate embeddings of the
@@
 all four methods for the \col, \gcc, and \svdfive time series.
-\begin{figure*}[htbp]
-  \centering
+%\begin{figure*}[htbp]
+%  \centering
+%
+%  \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/colRWForecast}
+%    \caption{\col\\ random walk }
+%    \label{fig:colRW}
+%  \end{subfigure}%
+%   \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
+%    \caption{\gcc\\ random walk }
+%    \label{fig:gccRW}
+%  \end{subfigure}%
+%     \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
+%    \caption{\svdfive\\ random walk}
+%    \label{fig:svd5RW}
+%  \end{subfigure}%
+%  \\
+%      \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
+%    \caption{\col\\ na\"ive }
+%    \label{fig:colMEAN}
+%  \end{subfigure}%
+%   \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
+%    \caption{\gcc\\ na\"ive }
+%    \label{fig:gccMEAN}
+%  \end{subfigure}%
+%     \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
+%    \caption{\svdfive\\ na\"ive }
+%    \label{fig:svd5MEAN}
+%  \end{subfigure}%
+%  \\
+%    \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
+%    \caption{\col \\ AutoARIMA}
+%    \label{fig:colARIMA}
+%  \end{subfigure}
+%  \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
+%    \caption{\gcc \\ AutoARIMA }
+%    \label{fig:gccARIMA}
+%  \end{subfigure}%
+%  \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
+%    \caption{\svdfive \\ AutoARIMA }
+%    \label{fig:svd5ARIMA}
+%  \end{subfigure}%
+%  \\
+%
+%      \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
+%    \caption{\col \\ LMA}
+%    \label{fig:colLMA}
+%  \end{subfigure}
+%      \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
+%    \caption{\gcc \\ LMA}
+%    \label{fig:gccLMA}
+%  \end{subfigure}
+%        \begin{subfigure}{0.6\columnwidth}
+%    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
+%    \caption{\svdfive \\ LMA}
+%    \label{fig:svd5LMA}
+%  \end{subfigure}
+%  %\begin{subfigure}{0.5\textwidth}
+%  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
+%   \caption{Predicted ($p_j$) versus true values ($c_j$) for
+%     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
+%% On this type of plot, a perfect prediction lies exactly on the
+%% diagonal.
+%%
+%%
+%}
+%\label{fig:forecast-example}
+%  %\end{subfigure}
 
-  \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colRWForecast}
-    \caption{\col\\ random walk }
-    \label{fig:colRW}
-  \end{subfigure}%
-   \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
-    \caption{\gcc\\ random walk }
-    \label{fig:gccRW}
-  \end{subfigure}%
-     \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
-    \caption{\svdfive\\ random walk}
-    \label{fig:svd5RW}
-  \end{subfigure}%
-  \\
-      \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
-    \caption{\col\\ na\"ive }
-    \label{fig:colMEAN}
-  \end{subfigure}%
-   \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
-    \caption{\gcc\\ na\"ive }
-    \label{fig:gccMEAN}
-  \end{subfigure}%
-     \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
-    \caption{\svdfive\\ na\"ive }
-    \label{fig:svd5MEAN}
-  \end{subfigure}%
-  \\
-    \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
-    \caption{\col \\ AutoARIMA}
-    \label{fig:colARIMA}
-  \end{subfigure}
-  \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
-    \caption{\gcc \\ AutoARIMA }
-    \label{fig:gccARIMA}
-  \end{subfigure}%
-  \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
-    \caption{\svdfive \\ AutoARIMA }
-    \label{fig:svd5ARIMA}
-  \end{subfigure}%
-  \\
+%\end{figure*}
 
-      \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
-    \caption{\col \\ LMA}
-    \label{fig:colLMA}
-  \end{subfigure}
-      \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
-    \caption{\gcc \\ LMA}
-    \label{fig:gccLMA}
-  \end{subfigure}
-        \begin{subfigure}{0.6\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
-    \caption{\svdfive \\ LMA}
-    \label{fig:svd5LMA}
-  \end{subfigure}
-  %\begin{subfigure}{0.5\textwidth}
-  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
-   \caption{Predicted ($p_j$) versus true values ($c_j$) for
-     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
-% On this type of plot, a perfect prediction lies exactly on the
-% diagonal.
-%
-%
-}
-\label{fig:forecast-example}
-  %\end{subfigure}
-\end{figure*}
-
-\begin{table*}
+\newcolumntype{C}{>{\centering\arraybackslash} m{0.6\columnwidth} }
+\begin{figure*}
   \renewcommand{\arraystretch}{1.25}
-  \begin{tabular}{lccc}
+  \begin{tabular}{m{0.1\columnwidth}|CCC}
     & {\large \col} & {\large \gcc} & {\large \svdfive} \\
     & & & \\
-    \begin{sideways}\large random walk\end{sideways} &
+            \hline \\
+    \rotatebox{90}{\large{random walk}} &
     \includegraphics[width=0.6\columnwidth]{figs/colRWForecast} &
@@
     \includegraphics[width=0.6\columnwidth]{figs/svdfiveRWForecast} \\
-    \begin{sideways}\large na\"ive\end{sideways} &
+    \rotatebox{90}{\large \naive} &
     \includegraphics[width=0.6\columnwidth]{figs/colMeanForecast} &
@@
     \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
-    \begin{sideways}\large AutoARIMA\end{sideways} &
+    \rotatebox{90}{\large \arima} &
     \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
@@
     \includegraphics[width=0.6\columnwidth]{figs/svdfiveARIMAForecast} \\
-    \begin{sideways}\large LMA\end{sideways} &
+    \rotatebox{90}{\large LMA} &
     \includegraphics[width=0.6\columnwidth]{figs/colLMAForecast} &
@@
   \label{fig:forecast-example}
-\end{table*}
+\end{figure*}
 In these images, the vertical axis is the prediction $p_j$ and the

### 2014-07-16 concl.tex
fixing autoarima command and formatting bugs

@@
 on noisy signals because it effects a filtering operation.  The simple
-random-walk strategy outperforms LMA, AutoARIMA, and the \naive
+random-walk strategy outperforms LMA, \arima, and the \naive
 ~method on the \gcc signal, which is extremely complex---i.e.,
@@
   systems like the H\'{e}non map \cite{henon} and the Lorenz system
-  \cite{lorenz63} also fit this pattern.}.
+  \cite{lorenz} also fit this pattern.}.
 

### 2014-07-16 model.tex
fixing autoarima command and formatting bugs

@@
   method, a common linear forecast strategy, built using the
-  \emph{AutoARIMA} procedure \cite{arima}, and
+  \emph{\arima} procedure \cite{autoARIMA}, and
 
@@
 data, or its temporal patterns, would be far more effective in the
-situations described in the previous paragraph.  The AutoARIMA and LMA
+situations described in the previous paragraph.  The \arima and LMA
 methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
@@
 automated fitting techniques described in~\cite{autoARIMA} to
-accomplish this, producing what we call an \emph{AutoARIMA model}.
+accomplish this, producing what we call an \emph{\arima model}.
 This procedure uses sophisticated methods---KPSS unit-root
@@
 the model fitted to the detrended data---to select good values for the
-free parameters of the AutoARIMA model.
+free parameters of the \arima model.
 
@@
 predictions, depending on model order.  To sidestep this issue, we
-build forecasts in a stepwise fashion: i.e., fit the AutoARIMA model
+build forecasts in a stepwise fashion: i.e., fit the \arima model
 to the existing data, use that model to perform a one-step prediction,
@@
 %    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
-%    \caption{\col \\ AutoARIMA}
+%    \caption{\col \\ \arima}
 %    \label{fig:colARIMA}
@@
 %    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
-%    \caption{\gcc \\ AutoARIMA }
+%    \caption{\gcc \\ \arima }
 %    \label{fig:gccARIMA}
@@
 %    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
-%    \caption{\svdfive \\ AutoARIMA }
+%    \caption{\svdfive \\ \arima }
 %    \label{fig:svd5ARIMA}
@@
 perfect prediction would lie on the diagonal.  LMA, for instance,
-generates a very accurate prediction of the \col data, while AutoARIMA
+generates a very accurate prediction of the \col data, while \arima
 does not.  Horizontal lines result when a constant predictor (e.g.,
@@
 distributed, for example, in the case of LMA on \gcc.  Note that the
-shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
+shapes of some of the plots in Figure~\ref{fig:forecast-example}, random walk and \arima on \col for example,  are
 reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
 For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
-two-dimensional embedding with $\tau=1$.  For AutoARIMA, the
+two-dimensional embedding with $\tau=1$.  For \arima, the
 correspondence is not quite as simple, since the $p_j$ values are
@@
 effect is largely the same\footnote{The temporal \emph{ordering} of
-  the points on the AutoARIMA $p_j$ vs. $c_j$ plot does not match that of
+  the points on the \arima $p_j$ vs. $c_j$ plot does not match that of
   a projected embedding of the time series, however.}.
@@
   \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
-  \hline\hline Signal & LMA MASE  & AutoARIMA MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
+  \hline\hline Signal & LMA MASE  & \arima MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
   \hline
@@
 correspondences.  The average LMA MASE score for the \col signals was
-$0.050$, for instance, while AutoARIMA scored much worse ($0.599$).
+$0.050$, for instance, while \arima scored much worse ($0.599$).
 That is, LMA performed roughly 20 times better on \col signals than a
-random-walk predictor, while AutoARIMA only outperformed random walk
+random-walk predictor, while \arima only outperformed random walk
 by a factor of 1.7.  This is in accordance with the visual appearance

### 2014-07-16 results.tex
bunch of formatting issues
fixing autoarima command and formatting bugs

@@
 Working from the first 90\% of each signal, we generated a prediction
-of the last 10\% using the \naive, AutoARIMA, and LMA prediction methods,
+of the last 10\% using the \naive, \arima, and LMA prediction methods,
 as described in Section~\ref{sec:accuracy}, then calculated the MASE
@@
 low as well, resulting in small, tight clusters.  In some
-cases---AutoARIMA predictions of {\tt col\_major}, for instance---the
+cases---\arima predictions of {\tt col\_major}, for instance---the
 MASE variance is larger, which spreads out the clusters horizontally.
@@
 the nonlinear LMA method are generally closer to the [[dashed curve]];
-the AutoARIMA method clusters are more widely spread, the \naive ~clusters
+the \arima method clusters are more widely spread, the \naive ~clusters
 even more so.  This geometric validation of the second conjecture in
@@
 graphical representations in the left-hand column of
-Figure~\ref{fig:forecast-example} that the \naive ~and AutoARIMA
+Figure~\ref{fig:forecast-example} that the \naive ~and \arima
 prediction methods do not perform as well on this signal as the
-nonlinear LMA method.  The MASE scores of the \naive ~and AutoARIMA
+nonlinear LMA method.  The MASE scores of the \naive ~and \arima
 predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
@@
 that simply predicts the running mean (\naive) or one that uses a
-linear model (AutoARIMA)---the MASE score is much better: $0.050 \pm
+linear model (\arima)---the MASE score is much better: $0.050 \pm
 0.001$.  This prediction is 20 times more accurate than a random-walk
@@
 very tightly in the lower left quadrant of the LMA and \naive ~MASE
-vs. WPE plots, but spread out horizontally in the AutoARIMA plot.
+vs. WPE plots, but spread out horizontally in the \arima plot.
 This is because of the way that these models are built
 \cite{autoARIMA}.  If a KPSS test of the time series in question
-indicates that it is nonstationary, the AutoARIMA recipe adds an
+indicates that it is nonstationary, the \arima recipe adds an
 integration term to the model.  This test gives mixed results in the
@@
   test indicated that such a term was not needed.  This action removed
-  the spread, pushing all 15 of the \col AutoARIMA points in
+  the spread, pushing all 15 of the \col \arima points in
   Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.}.
@@
 This match between model and task is an issue in the \svdfive
-experiments as well.  The MASE scores of the \naive ~and AutoARIMA
+experiments as well.  The MASE scores of the \naive ~and \arima
 predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
@@
 capture and reproduce the way in which the \svdfive system processes
-information, but the \naive ~and AutoARIMA prediction methods cannot.
+information, but the \naive ~and \arima prediction methods cannot.
 
@@
 fitting a hyperplane using least squares should have similar effects,
-the fact that LMA outperforms AutoARIMA ($1.530 \pm 0.021$ vs. $1.837
+the fact that LMA outperforms \arima ($1.530 \pm 0.021$ vs. $1.837
 \pm 0.016$) is somewhat counterintuitive.  However, the small amount
@@
 methods do quite well on this signal, yielding MASE scores of $0.827
-\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (AutoARIMA).  This pushes the
+\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (\arima).  This pushes the
 corresponding clusters of points in the bottom two images in
@@
 top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
-score of 0.6, which is better than LMA and AutoARIMA.  In five of the
+score of 0.6, which is better than LMA and \arima.  In five of the
 15 \svdone trials, however, there were step changes in the signal.
@@
 leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
-clusters for AutoARIMA and LMA are above the dashed curve in
+clusters for \arima and LMA are above the dashed curve in
 Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
@@
 [[Liz to wordsmith: Even though MASE has weaknesses, we concur with
-    \cite{mase} that it is an effective way to compare prediction
+    \cite{MASE} that it is an effective way to compare prediction
     error across time series of different lengths and different

### 2014-07-16 model.tex
formatting errors

@@
 Another weakness of this method is prediction horizon: an ARIMA
-forecast is guaranteed to converge to the mean after some number of
+forecast is guaranteed to converge to the\alert{ mean, a constant, or a linear trend}, after some number of
 predictions, depending on model order.  To sidestep this issue, we
@@
     \includegraphics[width=0.6\columnwidth]{figs/svdfiveLMAForecast}
-  \label{fig:forecast-example}
   \end{tabular}

### 2014-07-16 related.tex
bunch of formatting issues
add comparison to boffetta

@@
 As a final note: there has been prior work under a similar title to
-ours~\cite{haven2005}, but there are only superficial similarities
-between the two research projects.  Haven \emph{et al.} utilize the
-relative entropy to quantify the difference in predictability between
-two distributions: one evolved from a small ensemble of past states
-using the known dynamical system, and the other the observed
-distribution. Our work quantifies the predictability of a single
-observed time series using weighted permutation entropy and makes no
-assumptions about the generating process.  More closely related is the
-work by Boffetta et al.~\cite{boffetta02}.  \alert{[[Sentences from
-      Ryan explaining exactly what the difference is.]]]
+ours~\cite{haven2005}, but there are only superficial similarities between the
+two research projects. Haven \emph{et al.} utilize the relative entropy to
+quantify the difference in predictability between two distributions: one evolved
+from a small ensemble of past states using the known dynamical system, and the
+other the observed distribution. Our work quantifies the predictability of a
+single observed time series using weighted permutation entropy and makes no
+assumptions about the generating process. More closely related is the work by
+Boffetta et al.~\cite{boffetta02}. There, they investigate the scaling behavior
+of finite-size Lyapunov exponents (FSLE) and $\epsilon$-entropy for a wide
+variety of systems. The scaling of these measures acts as a general proxy for
+the theoretical ability to predict the future behavior of those systems. Though
+in broad strokes our work is similar to theirs, it differs in two fundamental
+ways. First, we consider with real-world data, e.g. stock prices, where we do
+not necessarily have the ability to perturb or otherwise interact with the
+system of interest, nor can we get an arbitrary amount of data which might be
+needed to estimate the $\epsilon$-entropy for small $\epsilon$. Second, we tie
+information, in the form of the weighted permutation entropy, directly to
+prediction error via MASE values. Though FSLE and $\epsilon$-entropy allow for
+the comparison of predictability between systems, they do not directly provide
+an estimate of prediction error.

### 2014-07-16 response.tex
revisions after Trident

@@
 discussed in a later section of this response.)  The image below is an
-augmented version of our \alert{revised Figure 6 [[This needs to be
-      regenerated with the new error bars]]} that includes all of
-these examples.
+augmented version of our revised Figure 7 that includes all of these
+examples.
 
 \begin{center}
-    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extra_autolog}
+    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras}
 \end{center}
@@
 
-We have \alert{Someone needs to do this: [[changed the term used in
-      the paper to AutoARIMA]]} in order to make it clear that this
-refers not to the general class of these models, but to a particular
-member of that class chosen via a particular procedure.
+We have changed the term used in the paper to {\tt auto.arima} in
+order to make it clear that this refers not to the general class of
+these models, but to a particular member of that class chosen via a
+particular procedure.
 
@@
 set.  Perhaps as a result of this, wordlength choices are rarely
-justified in the permutation entropy literature.  \alert{[[Can you
-      guys please extract a few representatively bad quotes from some
-      of the PE papers and put them in here?]]}
+justified in the permutation entropy literature.
+
+The extent of the discussion in [Bandt \& Pompe, {\sl PRL}, 2002], for
+instance---the paper in which PE was proposed---is ``Our entropies are
+calculated for different [$\ell$] but we do not to determine a limit
+for large [$\ell$], although this is an interesting theoretical
+problem.  ``For practical purposes, we recommend 3\dots7...''  The
+corresponding discussion in [Cao {\sl et al.}, {\sl PRE}, 2004] reads
+``In their paper, Bandt \& Pompe recommended [$\ell$] of 3-7.  We
+often found that 3 and 4 may still be too small, and a value of 5, 6,
+or 7 seems to be the most suitable....''
 
@@
 dynamical behaviors.  For all of these examples---as well as for the
-three additional ones constructed for this rebuttal letter (Henon,
-Lorenz, SFI A)---the pattern (Figure 6) was consistent.  This strongly
-suggests, but of course does not prove, that our approach is general.
-As the reviewer suggested, this will require extensive study and
-evaluation in future work.
+39 additional ones\footnote{Henon, Lorenz, SFI A, random walk,
+  nonlinear transformations of the 30 [[svdtwo]] and [[svdsix]]
+  signals} constructed for this rebuttal letter---the pattern in
+Figure 7 remained virtually unchanged, as described in the following
+entry.  This strongly suggests, but of course does not prove, that our
+approach is general.  As the reviewer suggested, this will require
+extensive study and evaluation in future work.
 
@@
 
-\alert{response and modification of text both are pending experiments}
+\alert{Liz to fix following text, now that we have these experiments}
 
@@
   complementary way}.  Ryan will do $x(t)$ and $5e^{x(t)}$ and show
-that it's on the curve.  This is consistent with our claim of
-generality; this is just another time series.  A xform like this
-effectively changes the generating process, but our model-free result
-handles it.
+that it's on the curve.
+
+% Ryan will insert figure here with x(t) = svdone and svdsix
+% \begin{center}
+%     \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras}
+% \end{center}
+
+This is consistent with our claim of generality; this is just another
+time series.  A xform like this effectively changes the generating
+process, but our model-free result handles it.
 
@@
 of the paper is not related to the crispness of the fit.
+
+\alert{Need the change and the confidence stuff for this: Fit changed
+  by [[what]] when we redid it with these 34 new points.}  It now
+reflects a map, a flow, a purely stochastic signal, a variety of
+real-world data from two different systems (laser, computer), and
+nonlinear transformations of that real-world data.  Seems pretty
+compelling.
 

### 2014-07-16 intro.tex
fixing autoarima command and formatting bugs
formatting errors
revisions after Trident

@@
-\section{Introduction  }\label{sec:intro}
+\section{Introduction}\label{sec:intro}
 %****NOTES START HERE****
@@
 (which exhibits high entropy) as purely complex, whereas a low-entropy
-periodic signal is on the low end of the complexity spectrum.
-\alert{This differs from the notion of complexity used by
-  e.g. \cite{Shalizi2008}, which would consider a time series without
-  any statistical regularities to be non-complex.  Liz moved this
-  here.  Is there a better place for it?}  We argue that an extension
-of \emph{permutation entropy}~\cite{bandt2002per}---a method for
+periodic signal is on the low end of the complexity spectrum.  This
+differs from the notion of complexity used by e.g. \cite{Shalizi2008},
+which would consider a time series without any statistical
+regularities to be non-complex.  We argue that an extension of
+\emph{permutation entropy}~\cite{bandt2002per}---a method for
 approximating the entropy through ordinal analysis---is an effective
@@
 %
- \begin{figure}[htbp]
+ \begin{figure}[htp]
     \centering

### 2014-07-16 model.tex
revisions after Trident

--- 2014-07-16 model.tex
@@
 Another weakness of this method is prediction horizon: an ARIMA
-forecast is guaranteed to converge to the mean after some number of
-predictions, depending on model order.  To sidestep this issue, we
-build forecasts in a stepwise fashion: i.e., fit the \arima model
-to the existing data, use that model to perform a one-step prediction,
-rebuild it using the latest observations, and iterate until the
-desired prediction horizon is reached.  (For consistency, we take the
-same approach with the other three models in this study as well, even
-though doing so amounts to artificially hobbling LMA.)
+forecast is guaranteed to converge (to the mean, to a constant value,
+or to a linear trend) after some number of predictions, depending on
+model order.  To sidestep this issue, we build forecasts in a stepwise
+fashion: i.e., fit the \arima model to the existing data, use that
+model to perform a one-step prediction, rebuild it using the latest
+observations, and iterate until the desired prediction horizon is
+reached.  For consistency, we take the same approach with the other
+three models in this study as well, even though doing so amounts to
+artificially hobbling LMA.
 

### 2014-07-16 results.tex
revisions after Trident

@@
   \centering
-  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_autolog}
+  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy}
   \caption{Weighted permutation entropy vs. mean absolute scaled error
@@
 those from {\tt dgesdd$_1$}, which we have excluded for reasons
-explained in the text.  [[Say what the colored region is and why we
-    included it.]]}
+explained in the text.  The shaded region boundaries reflect the
+standard deviation of the model in its parameter space.}
   \label{fig:wpe_vs_mase_best}
@@
   for reasons described below} captures this relationship between WPE
-and prediction error\footnote{The logarithmic fit is appropriate here
-  because any signal with zero entropy should be perfectly
-  predictable, and so should ideally have a MASE value of zero.}.
+and prediction error quite nicely.  The logarithmic shape of this
+curve makes sense here because any signal with zero entropy should be
+perfectly predictable, and so should ideally have a MASE value of
+zero.  The shaded region shows the standard deviation of the model in
+its parameter space: i.e., $y = a \log(b x + c)$ with $\pm$ one
+standard deviation on each of the three parameters.
+
 This is not a formal result.  The forecast methods and data sets used
@@
 Numerical values, including means and standard deviations of the
-errors, can be found in Table~\ref{tab:error}.  The dashed curve [[and
-    shaded regions?  need to check this throughout the text and in all
-    captions after we finalize the plot]] is the same as in the
-previous figure.  }
+errors, can be found in Table~\ref{tab:error}.  The dashed curve and
+shaded region is the same as in the previous figure.  }
     \label{fig:wpe_vs_mase_all}
@@
 the three plots.  The mean MASE scores of predictions generated with
-the nonlinear LMA method are generally closer to the [[dashed curve]];
-the \arima method clusters are more widely spread, the \naive ~clusters
+the nonlinear LMA method are generally closer to the dashed curve; the
+\arima method clusters are more widely spread, the \naive ~clusters
 even more so.  This geometric validation of the second conjecture in
@@
 Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
-the right of the [[dashed curve]].  \emph{This indicates that these
+the right of the dashed curve.  \emph{This indicates that these
   methods are not leveraging the available information in the signal}.
@@
 
-The [[dashed curve]] in Figures~\ref{fig:wpe_vs_mase_best}
+The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
 and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
@@
 predictive structure than the method is able to use.  When the
-predictions are well below [[that curve]] (as in the top and middle
-images in Figure~\ref{fig:wpe_vs_mase_all}), one should try a
-different method.  Conversely, the position of the LMA cluster---well
-to the left of the other two, near the [[dashed curve]]---reflects
-this method's ability to capture and exploit the structure that is
-present in this signal.  [[Liz to wordsmith: Again, this curve was
-    determined from a finite set of methods and data.  But we worked
-    hard to make these experiments representative and comprehensive:
-    forecast methods from simple to sophisticated; time-series data
-    from a system whose behavior spans the dynamical behavior space.
-    While we are cautiously opmistic about its generality, more
-    exploration will be required to firm up that conclusion.  Say that
-    Henon, Lorenz and SFI A fall within the colored region too, and
-    that we're working on a broader study.]]
+predictions are well below that curve (as in the top and middle images
+in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
+method.  Conversely, the position of the LMA cluster---well to the
+left of the other two, near the dashed curve---reflects this method's
+ability to capture and exploit the structure that is present in this
+signal.  [[Liz to wordsmith: Again, this curve was determined from a
+    finite set of methods and data.  But we worked hard to make these
+    experiments representative and comprehensive: forecast methods
+    from simple to sophisticated; time-series data from a system whose
+    behavior spans the dynamical behavior space.  While we are
+    cautiously opmistic about its generality, more exploration will be
+    required to firm up that conclusion.  Say that Henon, Lorenz and
+    SFI A fall within the colored region too, as do nonlinear
+    transformations of \svdtwo and \svdsix.  Add here how much the fit
+    changed with those additions.  As we add examples to this list,
+    this empirical result will firm up.]]
 

### 2014-07-16 related.tex
few edits to end of related.tex

@@
 an estimate of prediction error.
+
+They don't do real predictions and they don't use real-world data.
+
+Also they had models so could generate as much data as needed to
+perform their calculations.  Not practical in real data.
+
+They never actually do prediction, just use FTLE as proxy.  This makes
+general sense if you've got a deterministic system, but not stochastic
+systems, where FTLEs don't reflect predictability.

### 2014-07-18 results.tex
adding alert to update text based on table typo

@@
 transmits very little information forward in time and provides almost
-no structure for prediction methods to work with.  Here, the \naive
+no structure for prediction methods to work with.
+\alert{This needs to be reworded to reflect the typo, \naive is no longer the winner.
+MASE is $1.797 \pm 0.010$, rw actually wins $MASE = 1.1381 \pm 0.011$  Here, the \naive
 ~method performs best ($MASE=0.951 \pm 0.001$), simply because its
@@
 (cf.,~\cite{mytkowicz09}), and LMA is designed to capture and exploit
-that kind of structure.
+that kind of structure.}
 

### 2014-07-22 results.tex
update table, fit equations

@@
 The dashed line in the figure, a least-squares fit of $y = a \log(b x
-+ c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
++ 1) + c x^16$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
   for reasons described below} captures this relationship between WPE
@@
 zero.  The shaded region shows the standard deviation of the model in
-its parameter space: i.e., $y = a \log(b x + c)$ with $\pm$ one
+its parameter space: i.e., $y = a \log(b x + 1) + c x^16$ with $\pm$ one
 standard deviation on each of the three parameters.
@@
 %    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
-  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
+  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy4a}
 \caption{WPE vs. MASE for all trials, methods, and systems---with the
   exception of \svdone, \svdthree, and \svdfive are omitted from the
-  top plot for scale reasons.
+  top plot for scale reasons. \alert{there's also a version with all plots
+  stacked (4x1) instead of (2x2)}
 %

### 2014-07-22 response.tex
update table, fit equations
Liz editing doc and response wrt findings

@@
 points from the other time series fall within a narrow band around the
-fit.  Note that this fit ($y = a \log(b x + c)$) is different than in
+fit.  Note that this fit ($y = a \log(b x + 1) + c x^16$) is different than in
 the submitted version.  A careful examination of the data, prompted by
@@
 
-The original wording of this finding did indeed make it sound trivial.
-Our intent, as we have modified the text to reflect, is that the
-appropriateness of a predictor can be discovered \emph{empirically},
-via comparison of its MASE value with the WPE.  Previous explorations
-of predictability have used clean data and/or systems where the
-generating process is known.  Our work makes neither of these
-assumptions.  \alert{[[Sentence from Josh here about why that's
-      important: given a time series from an unknown system, there's
-      no way a priori to know what model is right.  Point to the col
-      example in the text as an example of this.  But need to check
-      after inserting it to make sure that this point has not been
-      belabored too much already in this letter.]]}  The paragraph
-that we have added to the introduction (immediately following these
-two claims) should hopefully make this more clear.
+The reviewer is absolutely right.  This finding, as it was originally
+stated, is indeed trivial.  Our intent, as we have modified the text
+to reflect, is that the appropriateness of a predictor can be
+discovered \emph{empirically}, via comparison of its MASE value with
+its WPE value.  Previous explorations of predictability have used
+clean data and/or systems where the generating process is known.  Our
+work makes neither of these assumptions.
+
+Given a linear stochastic system, a linear stochastic prediction model
+should certainly be more productive than a nonlinear deterministic
+prediction model.  But in real-world data, the generating process is
+almost never known, and it can be very hard to evaluate a model that
+one has constructed.  Standard methods like the AIC approach used in
+ARIMA modeling make a choice from among a large class of possible
+models.  These selection algorithms rarely have a ``none of these are
+appropriate'' output; they simply choose the model that best satisfies
+their built-in criteria---even if none of the models in the class are
+really appropriate.
+
+The challenge for the practitioner is to know whether there is a
+mismatch between a particular time series and a particular forecast
+model: i.e., whether one can do better and should try another model,
+or whether the time series is so complicated that the model is doing
+as well as one could expect.  The WPE vs. MASE curve in our paper is
+an empirical heuristic that should help practitioners do that.
+
+We appreciate the reviewer alerting us to our muddy wording and
+thinking about these issues.  We have reworked the findings to clarify
+matters and also added an explanatory paragraph directly following
+those findings.  While this lengthened the introduction, we feel that
+it has focused and clarified the contribution of the work.
 
@@
 % Ryan will insert figure here with x(t) = svdone and svdsix
-% \begin{center}
-%     \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras}
-% \end{center}
+\begin{center}
+    \includegraphics[width=0.8\columnwidth]{figs/nonlinear_transform_data}
+\end{center}
 
@@
 
+Because of the depth and extent of the revisions that we made in
+response to this reviewer's comments, we did not feel that it was
+effective to highlight the changes in a different color in the
+manuscript; please let us know if a more-detailed description of the
+changes would make the re-reading easier.
+
 \end{document}

### 2014-07-22 intro.tex
Liz editing doc and response wrt findings

@@
 conclude that that method is inadequate to the task and that one
-should seek another method.  The goal of this paper is to explore and
-formalize that strategy.
+should seek another method.  The goal of this paper is to develop
+effective, practical heuristics for that strategy.
 
@@
 
-We focus on real-valued, scalar, time-series data.
+We focus on real-valued, scalar, time-series data from physical experiments.
 %
@@
 relationship between complexity, predictive structure, and
-predictability, we generate forecasts for a variety of time-series
-datasets using four different prediction methods, then compare the
-accuracy of those predictions to the permutation entropy of the
-associated signals.  This results in two primary findings:
+predictability, we generate forecasts for a variety of experimental
+time-series datasets using four different prediction methods, then
+compare the accuracy of those predictions to the permutation entropy
+of the associated signals.  This results in two primary findings:
+% 
 \begin{enumerate}
-\item The complexity of a noisy real-valued time series is \emph{(i)}
-  quantifiable by permutation entropy and \emph{(ii)} correlated with
-  the accuracy of an appropriate predictor.
-\item The way information is generated and processed internally by a
-  system plays a crucial role in the success of different forecasting
-  schema---and in the choice of which one is appropriate for a given
-  empirical time series.
+% \item The complexity of a noisy real-valued time series can be
+%   effectively quantifiable by computing its permutation entropy.
+
+\item The permutation entropy of a noisy real-valued time series is
+  correlated with the accuracy of an appropriate predictor.
+
+\item The relationship between permutation entropy and prediction
+  accuracy is a useful empirical heuristic for identifying when there
+  is a mismatch between a time series from an unknown system and a
+  given prediction model.
 \end{enumerate}
+% 
 There has, of course, been a great deal of good work on different ways
 to measure the complexity of data, and previous explorations have
-confirmed repeatedly that complexity is a challenge to prediction, but
-this collection of issues has not been properly explored in the
-context of noisy, poorly sampled, real-world data from unknown
-systems.  That is the contribution of this paper.
+confirmed repeatedly that complexity is a challenge to prediction.  It
+is well known that the way information is generated and processed
+internally by a system plays a crucial role in the success of
+different forecasting methods---and in the choice of which method is
+appropriate for a given time series.  
+This constellation of issues has not been properly explored, however,
+in the context of noisy, poorly sampled, real-world data from unknown
+systems.  That exploration---and the development of strategies for
+putting its results into effective practice---is the primary
+contribution of this work.  The empirical results in the later
+sections of this paper not only elucidate the relationship between
+complexity and predictability.  The strategy that we derive from these
+results can also aid practitioners in choosing an appropriate
+prediction model for a given real-world noisy time series from an
+unknown system---a challenging task for which little guidance is
+currently available.
+
+\alert{With all the extra material above, the following paragraph now
+  feels repetitive.  Given that, and the reviewer's comment that the
+  intro is too long, I suggest that we merge it into a later section:}
 
@@
 
-For the purposes of this study, we require a broad array of
+\alert{This paragraph, too, might well be moved to the beginning of
+  section 3:}
+
+For the purposes of this study, we required a broad array of
 time-series datasets from across the complexity spectrum.  We chose to

### 2014-07-22 related.tex
Liz editing doc and response wrt findings

--- 2014-07-16 related.tex
@@
 moves away from those restrictions.  Our approach falls into this
-class.  Our goal is to measure the inherent complexity of an empirical
-time series, then correlate that complexity with the predictive
-accuracy of forecasts made using a number of different methods.
+class: we wish to measure the inherent complexity of an empirical time
+series, then correlate that complexity with the predictive accuracy of
+forecasts made using a number of different methods.
 
@@
 the generating system is stochastic.
-
-The notion of complexity used here differs from that considered by
-e.g. \cite{Shalizi2008}, which would consider a time series without
-any statistical regularities to be non-complex.
 
@@
 
-As a final note: there has been prior work under a similar title to
+There has been prior work under a similar title to
 ours~\cite{haven2005}, but there are only superficial similarities
@@
 observed time series using weighted permutation entropy and makes no
-assumptions about the generating process.  More closely related is the
-work by Boffetta et al.~\cite{boffetta02}.  \alert{[[Sentences from
-      Ryan explaining exactly what the difference is.]]]
+assumptions about the generating process.
+
+More closely related is the work by Boffetta \emph{et
+  al.}~\cite{boffetta02}, who investigate the scaling behavior of
+finite-time Lyapunov exponents (FTLE) and $\epsilon$-entropy for a
+wide variety of deterministic systems with known dynamics.  While the
+scaling of these measures acts as a general proxy for predictability
+bounds, this approach differs from our work in a number of fundamental
+ways.  First, we focus on real-world time-series data, where one does
+not necessarily have the ability to perturb or otherwise interact with
+the system of interest, nor can one obtain or manufacture the
+(possibly large) number of points that might be needed to estimate the
+$\epsilon$-entropy for small $\epsilon$.  Second, we tie
+information---in the form of the weighted permutation
+entropy---directly to prediction error via MASE values.  Though FTLE
+and $\epsilon$-entropy allow for the comparison of predictability
+between systems, they do not directly provide an estimate of
+prediction error.  [[Finally, our approach also holds for
+    \emph{stochastic} systems, where the computation of FTLEs makes no
+    sense and/or doesn't reflect predictability.  I remember someone
+    saying this but didn't see it here, so I added it.  Boffetta et
+    al. just do dynamical systems, yes?  And lambda of a stochastic
+    system wouldnt' make sense, yes?]]

### 2014-07-23 intro.tex
couple questions

@@
 %
-We do not make any knowledge of the generating process or its
+We do not make\alert{should this be assume instead of make} any knowledge of the generating process or its
 properties: whether it is linear, nonlinear, deterministic,

### 2014-07-23 related.tex
couple questions

@@
 theoretical assessment or comparison of existing methods.  Our goals
-are focused more on the \emph{practice} of predication: \emph{(i)} to
+are focused more on the \emph{practice} of predication\alert{should this be prediction}: \emph{(i)} to
 empirically quantify the predictive structure that is present in a

### 2014-07-23 intro.tex
moved paragraphs out of intro

--- 2014-07-22 intro.tex
@@
 %
-We do not make any knowledge of the generating process or its
+We do not assume any knowledge of the generating process or its
 properties: whether it is linear, nonlinear, deterministic,
@@
 different forecasting methods---and in the choice of which method is
-appropriate for a given time series.  
-This constellation of issues has not been properly explored, however,
-in the context of noisy, poorly sampled, real-world data from unknown
-systems.  That exploration---and the development of strategies for
-putting its results into effective practice---is the primary
-contribution of this work.  The empirical results in the later
-sections of this paper not only elucidate the relationship between
-complexity and predictability.  The strategy that we derive from these
-results can also aid practitioners in choosing an appropriate
-prediction model for a given real-world noisy time series from an
-unknown system---a challenging task for which little guidance is
-currently available.
+appropriate for a given time series.  This constellation of issues has
+not been properly explored, however, in the context of noisy, poorly
+sampled, real-world data from unknown systems.  That exploration---and
+the development of strategies for putting its results into effective
+practice---is the primary contribution of this work.  The empirical
+results in the later sections of this paper not only elucidate the
+relationship between complexity and predictability.  The strategy that
+we derive from these results can also aid practitioners in choosing an
+appropriate prediction model for a given real-world noisy time series
+from an unknown system---a challenging task for which little guidance
+is currently available.
 
-\alert{With all the extra material above, the following paragraph now
-  feels repetitive.  Given that, and the reviewer's comment that the
-  intro is too long, I suggest that we merge it into a later section:}
+% % Moved this to the beginning of forecasting section:
+% 
+% The forecast methods used in this study were chosen to be a
+% representative sampling of standard prediction strategies, but they do
+% not, of course, cover that space exhaustively.  Our goal here is an
+% empirical assessment of the relationship between predictability and
+% complexity, not formal results about a ``best'' predictor for a given
+% time series.  From a practitioner's standpoint, it would be useful to
+% know \emph{ a priori} if a time series contains enough predictive
+% structure to make it worth spending the time and effort searching for
+% a good forecast method.  It would also be useful to know if a given
+% method is inadequate---that is, if an alternative method could do
+% better and so one should continue searching.
 
-The forecast methods used in this study were chosen to be a
-representative sampling of standard prediction strategies, but they do
-not, of course, cover that space exhaustively.  Our goal here is an
-empirical assessment of the relationship between predictability and
-complexity, not formal results about a ``best'' predictor for a given
-time series.  From a practitioner's standpoint, it would be useful to
-know \emph{ a priori} if a time series contains enough predictive
-structure to make it worth spending the time and effort searching for
-a good forecast method.  It would also be useful to know if a given
-method is inadequate---that is, if an alternative method could do
-better and so one should continue searching.
-
-%[Introduce and make it clear why we choose comp. perf. as test bed.]
-
-\alert{This paragraph, too, might well be moved to the beginning of
-  section 3:}
-
-For the purposes of this study, we required a broad array of
-time-series datasets from across the complexity spectrum.  We chose to
-study sensor data from a computer-performance experiment.  While this
-is not a common laboratory experiment, it is a highly appropriate
-choice here.  Computers are extremely complicated systems and their
-dynamics is surprisingly rich.
-% Modern microprocessor chips contain multiple processing units and
-% multi-layer memories, for instance, and they use complicated
-% hardware/software strategies to maximize performance by moving data
-% and threads of computation across those resources.  As a result, the
-The processor and memory loads during the execution of even a very
-simple program can exhibit dynamical chaos, for
-instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
-example: a performance trace of a short program that repeatedly
-initializes the upper triangle of a matrix in column-major order.
+% % Moved this to the beginning of methods section:
 % 
- \begin{figure}[htp]
-    \centering
-    \includegraphics[width=\columnwidth]{figs/colshortts}
-    % where an .eps filename suffix will be assumed under latex,
-    % and a .pdf suffix will be assumed for pdflatex
-    \caption{A computer performance trace: the processor load of \col,
-      a simple C program that repeatedly initializes a matrix in
-      column-major order, running on an Intel
-      i7\textsuperscript{\textregistered}-based machine.}
-   \label{fig:col-ipc}
-  \end{figure}
-%
-A small change in the code can cause the dynamics to bifurcate to a
-periodic regime.  
-% 
-% Indeed, many computer performance traces exhibit
-% interesting regime changes as a program moves through the different
-% phases of its operation.
-By running different programs on the same computer, then, we can
-produce traces that span the whole range of the complexity spectrum,
-from completely predictable to completely unstructured---which makes
-this an ideal testbed for this study\footnote{Predicting the
-  \emph{state} of a computer, of course, would amount to solving the
-  halting problem.  What we are doing here is predicting computer
-  \emph{performance}, which does not violate the Rice-Shapiro
-  theorem~\cite{hopcroft2007}.}.
-% footnote to appease the theoretician in Josh
+% For the purposes of this study, we required a broad array of
+% time-series datasets from across the complexity spectrum.  We chose to
+% study sensor data from a computer-performance experiment.  While this
+% is not a common laboratory experiment, it is a highly appropriate
+% choice here.  Computers are extremely complicated systems and their
+% dynamics is surprisingly rich.
+% % Modern microprocessor chips contain multiple processing units and
+% % multi-layer memories, for instance, and they use complicated
+% % hardware/software strategies to maximize performance by moving data
+% % and threads of computation across those resources.  As a result, the
+% The processor and memory loads during the execution of even a very
+% simple program can exhibit dynamical chaos, for
+% instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
+% example: a performance trace of a short program that repeatedly
+% initializes the upper triangle of a matrix in column-major order.
+% %
+%  \begin{figure}[htp]
+%     \centering
+%     \includegraphics[width=\columnwidth]{figs/colshortts}
+%     % where an .eps filename suffix will be assumed under latex,
+%     % and a .pdf suffix will be assumed for pdflatex
+%     \caption{A computer performance trace: the processor load of \col,
+%       a simple C program that repeatedly initializes a matrix in
+%       column-major order, running on an Intel
+%       i7\textsuperscript{\textregistered}-based machine.}
+%    \label{fig:col-ipc}
+%   \end{figure}
+% %
+% A small change in the code can cause the dynamics to bifurcate to a
+% periodic regime.  
+% % 
+% % Indeed, many computer performance traces exhibit
+% % interesting regime changes as a program moves through the different
+% % phases of its operation.
+% By running different programs on the same computer, then, we can
+% produce traces that span the whole range of the complexity spectrum,
+% from completely predictable to completely unstructured---which makes
+% this an ideal testbed for this study\footnote{Predicting the
+%   \emph{state} of a computer, of course, would amount to solving the
+%   halting problem.  What we are doing here is predicting computer
+%   \emph{performance}, which does not violate the Rice-Shapiro
+%   theorem~\cite{hopcroft2007}.}.
+% % footnote to appease the theoretician in Josh
 

### 2014-07-23 methods.tex
second round of LizB edits
moved paragraphs out of intro

@@
 % \subsection{Time Series Collection}
+
+For the purposes of this study, we required a broad array of
+time-series datasets from across the complexity spectrum.  We chose to
+study sensor data from a computer-performance experiment.  While this
+is not a common laboratory experiment, it is a highly appropriate
+choice here.  Computers are extremely complicated systems and their
+dynamics is surprisingly rich.
+% Modern microprocessor chips contain multiple processing units and
+% multi-layer memories, for instance, and they use complicated
+% hardware/software strategies to maximize performance by moving data
+% and threads of computation across those resources.  As a result, the
+The processor and memory loads during the execution of even a very
+simple program can exhibit dynamical chaos, for
+instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
+example: a short segment of a performance trace of a short program
+that repeatedly initializes the upper triangle of a matrix in
+column-major order.
+%
+ \begin{figure}[htp]
+    \centering
+    \includegraphics[width=\columnwidth]{figs/colshortts}
+    % where an .eps filename suffix will be assumed under latex,
+    % and a .pdf suffix will be assumed for pdflatex
+    \caption{A short segment of a computer performance trace: the
+      instructions per CPU clock cycle (IPC) during the execution of
+      \col, a simple C program that repeatedly initializes a matrix in
+      column-major order, running on an Intel
+      i7\textsuperscript{\textregistered}-based machine.  Each point
+      is the average IPC in a 100,000 instruction period.}
+   \label{fig:col-ipc}
+  \end{figure}
+%
+A small change in the code can cause the dynamics to bifurcate to a
+periodic regime.  
+% 
+% Indeed, many computer performance traces exhibit
+% interesting regime changes as a program moves through the different
+% phases of its operation.
+By running different programs on the same computer, then, we can
+produce traces that span the whole range of the complexity spectrum,
+from completely predictable to completely unstructured---which makes
+this an ideal testbed for this study\footnote{Predicting the
+  \emph{state} of a computer, of course, would amount to solving the
+  halting problem.  What we are doing here is predicting computer
+  \emph{performance}, which does not violate the Rice-Shapiro
+  theorem~\cite{hopcroft2007}.}.
+% footnote to appease the theoretician in Josh
 
@@
 
-% \subsection{The Programs: [[Maybe The generating processes]]{\color{blue} EDITABLE}}
+% \subsection{The Programs: [Maybe The generating processes]{\color{blue} EDITABLE}}
 % We collect performance traces of processor efficiency (IPC) of three
-% exemplar programs, [[If we change the title of the section, add this:
-%     which act as our generating processes]]---a simple
+% exemplar programs, [If we change the title of the section, add this:
+%     which act as our generating processes]---a simple
 % microkernel(\col) and two complex programs: one from the SPEC 2006CPU
@@
 
-\col is a simple C program that repeatedly initializes the upper
-triangle of a 2048 $\times$ 2048 matrix in column-major order.
+% Commented this out because the addition of the paragraph from the
+% introduction seemed like it could replace it
 % 
-% by looping over the following three lines of code:
-% \begin{lstlisting}
-%   for(i=0;i<2048;i++)
-%   	for(j=i;j<2048;j++)
-%   		data[j][i] = 0;
-% \end{lstlisting}
-% full listing
-% #define SIZE (2048)
+% \col is a simple C program that repeatedly initializes the upper
+% triangle of a 2048 $\times$ 2048 matrix in column-major order.
+% % 
+% % by looping over the following three lines of code:
+% % \begin{lstlisting}
+% %   for(i=0;i<2048;i++)
+% %   	for(j=i;j<2048;j++)
+% %   		data[j][i] = 0;
+% % \end{lstlisting}
+% % full listing
+% % #define SIZE (2048)
+% % 
+% % int main(int argc, char ** argv)
+% % {
+% %   int i, j, n;
+% %   static int data[SIZE][SIZE];
+% %   
+% %   for(n=0;n<numInits;n++){  
+% %   for(i=0;i<SIZE;i++)
+% %   	for(j=i;j<SIZE;j++)
+% %   		data[j][i] = 0;
+% %   }
+% %   return 0;
+% % }
+% %
+% As mentioned in Section~\ref{sec:intro}, this simple program exhibits
+% surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
+% of the processor performance during the execution of this program is
+% shown in Figure~\ref{fig:col-ts}.
+% \begin{figure}[htbp]
+%   \centering
+%     \includegraphics[width=\columnwidth]{figs/colFullTS}
+%     \caption{The instructions executed per CPU clock cycle (IPC) as
+%       the \col program runs.  Each point is the average IPC in a
+%       100,000 instruction period. Figure~\ref{fig:col-ipc} is a
+%       closeup of the first part of this signal.  }
+%     \label{fig:col-ts}
 %   
-% int main(int argc, char ** argv)
-% {
-%   int i, j, n;
-%   static int data[SIZE][SIZE];
-%   
-%   for(n=0;n<numInits;n++){  
-%   for(i=0;i<SIZE;i++)
-%   	for(j=i;j<SIZE;j++)
-%   		data[j][i] = 0;
-%   }
-%   return 0;
-% }
-%
-As mentioned in Section~\ref{sec:intro}, this simple program exhibits
-surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
-of the processor performance during the execution of this program is
-shown in Figure~\ref{fig:col-ts}.
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=\columnwidth]{figs/colFullTS}
-    \caption{The instructions executed per CPU clock cycle (IPC) as
-      the \col program runs.  Each point is the average IPC in a
-      100,000 instruction period. Figure~\ref{fig:col-ipc} is a
-      closeup of the first part of this signal.  }
-    \label{fig:col-ts}
-
-  \end{figure}
+%   \end{figure}
 
@@
     \includegraphics[width=\columnwidth]{figs/gccfullts}
-    \caption{The instructions per CPU clock cycle (IPC) during the
-      execution of \gcc. Each point is the average IPC in a 100,000
-      instruction period.}
+    \caption{An IPC trace during the execution of \gcc.}
     \label{fig:gcc-ts}
@@
     \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
-    \caption{The instructions per CPU clock cycle (IPC) during the
-      execution of \svd. Each point is the average IPC in a 100,000
-      instruction period.  The colors identify the different
-      \emph{segments} of the signal that are discussed in the text.}
+    \caption{An IPC trace during the execution of \gcc.  The colors
+      identify the different \emph{segments} of the signal that are
+      discussed in the text.}
     \label{fig:svd-ts-colored}
@@
 different phases of a single time series.  The complexity of \col and
-\gcc in Figures~\ref{fig:col-ts} and \ref{fig:gcc-ts} appears to be
+\gcc in Figures~\ref{fig:col-ipc} and \ref{fig:gcc-ts} appears to be
 far more consistent over time---probably the result of a single
@@
 
-%[[Orphan citation?]] For regimes \cite{cao2004det}
+%[Orphan citation?] For regimes \cite{cao2004det}

### 2014-07-23 related.tex
moved paragraphs out of intro

--- 2014-07-22 related.tex
@@
 theoretical assessment or comparison of existing methods.  Our goals
-are focused more on the \emph{practice} of predication: \emph{(i)} to
+are focused more on the \emph{practice} of prediction: \emph{(i)} to
 empirically quantify the predictive structure that is present in a
@@
   al.}~\cite{boffetta02}, who investigate the scaling behavior of
-finite-time Lyapunov exponents (FTLE) and $\epsilon$-entropy for a
+finite-size Lyapunov exponents (FSLE) and $\epsilon$-entropy for a
 wide variety of deterministic systems with known dynamics.  While the
@@
 information---in the form of the weighted permutation
-entropy---directly to prediction error via MASE values.  Though FTLE
+entropy---directly to prediction error via MASE values.  Though FSLE
 and $\epsilon$-entropy allow for the comparison of predictability
 between systems, they do not directly provide an estimate of
-prediction error.  [[Finally, our approach also holds for
+prediction error.  \alert{Finally, our approach also holds for
   \emph{stochastic} systems, where the computation of FTLEs makes no
-    sense and/or doesn't reflect predictability.  I remember someone
+  sense and/or doesn't reflect predictability.  [[I remember someone
       saying this but didn't see it here, so I added it.  Boffetta et
       al. just do dynamical systems, yes?  And lambda of a stochastic
-    system wouldnt' make sense, yes?]]
+      system wouldnt' make sense, yes?]]}

### 2014-07-24 concl.tex
airplane edits

@@
 \section{ Conclusions \& Future Work }\label{sec:conc}
+
+[[Moved this out of intro but it was redundant in the model section
+    and some of it is repeated in the results section.  Maybe here?
+    The forecast methods used in this study were chosen to be a
+    representative sampling of standard prediction strategies, but
+    they do not, of course, cover that space exhaustively.  Our goal
+    here is an empirical assessment of the relationship between
+    predictability and complexity, not formal results about a ``best''
+    predictor for a given time series.  From a practitioner's
+    standpoint, it would be useful to know \emph{ a priori} if a time
+    series contains enough predictive structure to make it worth
+    spending the time and effort searching for a good forecast method.
+    It would also be useful to know if a given method is
+    inadequate---that is, if an alternative method could do better and
+    so one should continue searching.]]
 
@@
 higher the WPE, in general, the higher the prediction error.  The
-relationship is roughly logarithmic.  [[This is just an observed result.
-It makes sense because, firstly, any signal with zero entropy should
-be perfectly predictable, and so should ideally have a MASE value of
-zero.  As the entropy rises, the MASE value rise: slowly at first
-because the signal still has lots of info, then more rapidly because
-[[why]], and finally saturating at [[what]] because of [[why]].
-Supporting our hypothesis, forecast errors for the \col and \svdfive
-signals are higher than the corresponding WPEs suggest---except for
-the nonlinear LMA predictor.  There are some exceptions: \svdone is
-easier to predict than its WPE value suggests using any of the three
-forecast strategies studied here.  On the whole, though, signals with
-higher weighted permutation entropy are harder to predict.
+relationship is roughly logarithmic.
+
+[[Maybe put (or reiterate) the ``range of examples'' material here.
+    This is just an observed result.  It makes sense because any
+    signal with zero entropy should be perfectly predictable, and so
+    should ideally have a MASE value of zero.  As the entropy rises,
+    the MASE value rise: slowly at first---where the signal is not
+    very complex---and then more rapidly after WPE $\approx$ 0.5.  We
+    are in the process of exploring the theory behind this shape, as
+    well as testing other shapes.  Supporting our hypothesis, forecast
+    errors for the \col and \svdfive signals are higher than the
+    corresponding WPEs suggest---except for the nonlinear LMA
+    predictor.  There are some exceptions: \svdone is easier to
+    predict than its WPE value suggests using any of the three
+    forecast strategies studied here.  On the whole, though, signals
+    with higher weighted permutation entropy are harder to predict.
 
@@
 guideline for knowing when a model is not well-matched to the task at
-hand: [[a MASE score that is significantly higher than $3.58 \cdot
-    \mathrm{WPE} - 1.91$, where $\mathrm{WPE}$ is that of the
-    corresponding signal]] [[need to reword]], is an indication that
-that time series has more predictive structure than that forecast
-model can capture and exploit.
+hand: a MASE score that is significantly\footnote{more than one
+  $\sigma$ out} higher than [[$3.58 \cdot \mathrm{WPE} - 1.91$]],
+where $\mathrm{WPE}$ is that of the corresponding signal, is an
+indication that that time series has more predictive structure than
+that forecast model can capture and exploit.
 
@@
 Preliminary results of this study show that MASE vs. WPE points for
-data from the SFI prediction competition \cite{weigend-book} fall in the shaded
-region on Figure [[what]]\footnote{Synthetic data from paradigmatic
-  systems like the H\'{e}non map \cite{henon} and the Lorenz system
-  \cite{lorenz} also fit this pattern.}.
+data from the SFI prediction competition \cite{weigend-book} fall in
+the shaded region on Figure [[what]]\footnote{Synthetic data from
+  paradigmatic systems like the H\'{e}non map \cite{henon} and the
+  Lorenz system \cite{lorenz} also fit this pattern.}.
 

### 2014-07-24 intro.tex
airplane edits

@@
 should seek another method.  The goal of this paper is to develop
-effective, practical heuristics for that strategy.
+effective heuristics to put that strategy into practice.
 
@@
 
-\item The permutation entropy of a noisy real-valued time series is
-  correlated with the accuracy of an appropriate predictor.
+\item The permutation entropy of a noisy real-valued time series from
+  an unknown system is correlated with the accuracy of an appropriate
+  predictor.
 
 \item The relationship between permutation entropy and prediction
-  accuracy is a useful empirical heuristic for identifying when there
-  is a mismatch between a time series from an unknown system and a
-  given prediction model.
+  accuracy is a useful empirical heuristic for identifying mismatches
+  between prediction models and time-series data.
+
 \end{enumerate}

### 2014-07-24 methods.tex
airplane edits

@@
 from the SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK
-linear algebra package (\svd).  In these experiments, the scalar
-observation $x_i$ was a measurement of the processor performance at
-time $i$ during the execution of each program.  
+linear algebra package (\svd).  In all of these experiments, the
+scalar observation $x_i$ was a measurement of the processor
+performance at time $i$ during the execution of each program.
 % 
@@
 % 
-For an in-depth description of this experimental setup,
+% These traces, and the processes that generated them, are described
+% briefly in the rest of this section.
+% 
+For statistical validation, we collected 15 performance traces from
+each of the three programs.  For an in-depth description of this
+experimental setup,
 % including discussion of the implications of the sampling interval,
-please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
-validation, we collected 15 performance traces from each of the three
-programs.  These traces, and the processes that generated them, are
-described in more depth in the rest of this section.
+please see \cite{zach-IDA10,mytkowicz09,todd-phd}.
 
@@
 The behavior of this program as it computes the singular values of
-this matrix is very interesting, as shown in Figure
+this matrix is very interesting, as is clearly visible in Figure
 \ref{fig:svd-ts-colored}.  As the code moves though its different
@@
 
-This wide range of behaviors provides a unique advantage, for the
-purposes of this paper, in that a number of different generating
-processes---with a wide range of complexities---are at work in
-different phases of a single time series.  The complexity of \col and
-\gcc in Figures~\ref{fig:col-ipc} and \ref{fig:gcc-ts} appears to be
-far more consistent over time---probably the result of a single
-generating process.  \svd, in contrast, has multiple regimes, each the
-result of a different generating process.  To take advantage of this,
-we split the signal into six different segments, thereby obtaining an
-array of examples for the analyses in the following sections.  For
-notational convenience, we refer to these 90 time-series data
-sets\footnote{15 runs, each with six regimes} as {\tt dgesdd$_i$},
-with $i \in \{1\dots6\}$ where $i$ corresponds to one of the six
-segments of the signal, ordered from left to right.  These segments,
-which were determined visually, are shown in different colors in
+The wide range of behaviors in Figure~\ref{fig:svd-ts-colored}
+provides a unique advantage, for the purposes of this paper, in that a
+number of different generating processes---with a wide range of
+complexities---are at work in different phases of a single time
+series.  The complexity of \col and \gcc in Figures~\ref{fig:col-ipc}
+and \ref{fig:gcc-ts} appears to be far more consistent over
+time---probably the result of a single generating process.  \svd, in
+contrast, has multiple regimes, each the result of a different
+generating process.  To take advantage of this, we split the signal
+into six different segments, thereby obtaining an array of examples
+for the analyses in the following sections.  For notational
+convenience, we refer to these 90 time-series data sets\footnote{15
+  runs, each with six regimes} as {\tt dgesdd$_i$}, with $i \in
+\{1\dots6\}$ where $i$ corresponds to one of the six segments of the
+signal, ordered from left to right.  These segments, which were
+determined visually, are shown in different colors in
 Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,

### 2014-07-24 model.tex
fixing a table entry
update table, fit equations
airplane edits

@@
     \caption{3D projections of delay-coordinate embeddings of the
-       traces from (a) Figure~\ref{fig:col-ts} (b)
+       traces from (a) Figure~\ref{fig:col-ipc} (b)
        Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
@@
 
-Since LMA does not rest on an assumption of linearity (as ARIMA does),
-it can handle both linear and nonlinear processes.  If the underlying
-generating process is nondeterministic, however, it can perform
-poorly.  In Figure~\ref{fig:embedding}b, for instance, little
-structure is visible, so LMA may not work well.  More structure
-appears to be present in~Figure~\ref{fig:embedding}c, but this
-reconstruction also appears to contain some noise.  The question as to
-how much structure is present in a reconstruction---and how much of
-that structure can be captured and used by LMA---is apropos of the
-central question treated in this paper.  It may be that \gcc has some
-redundancy that LMA cannot exploit, or that the structure in \svdfive
-is effectively obfuscated, from the standpoint of the LMA method, by
-noise.  Without any knowledge of the generating process, answers to
-these questions can only be derived from the data, with all of the
-attendant problems (noise, sampling issues, etc.)  By quantifying the
-balance between redundancy (that is, predictive structure) and entropy
-for these real-valued time series, as shown in
-Section~\ref{sec:results}, we can begin to answer these questions
-in an effective and practical manner.
+Since LMA does not rest on an assumption of linearity (as ARIMA models
+do), it can handle both linear and nonlinear processes.  If the
+underlying generating process is nondeterministic, however, it can
+perform poorly.  Figure~\ref{fig:embedding}b, for instance, appears to
+contain little structure, so LMA may not work well on this signal.
+More structure appears to be present in Figure~\ref{fig:embedding}c,
+but this reconstruction also appears to contain some noise.  The
+question as to how much structure is present in a reconstruction---and
+how much of that structure can be captured and used by LMA---is
+apropos of the central question treated in this paper.  It may be that
+\gcc has some redundancy that LMA cannot exploit, or that the
+structure in \svdfive is effectively obfuscated, from the standpoint
+of the LMA method, by noise.  Without any knowledge of the generating
+process, answers to these questions can only be derived from the data,
+with all of the attendant problems (noise, sampling issues, etc.)  By
+quantifying the balance between redundancy (that is, predictive
+structure) and entropy for these real-valued time series, as shown in
+Section~\ref{sec:results}, we can begin to answer these questions in
+an effective and practical manner.
 
@@
 %  %\end{subfigure}
-
 %\end{figure*}
-
 \newcolumntype{C}{>{\centering\arraybackslash} m{0.6\columnwidth} }
@@
 is the average in-sample forecast error for a random-walk prediction
-over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$ means that the
-prediction error in question was, on the average, smaller than the
-error of a random-walk forecast on the same data.  Analogously,
-$MASE>1$ means that the corresponding prediction method did
-\emph{worse}, on average, than the random-walk method.  We chose this
-error metric because it allows for fair comparison across varying
-methods, prediction horizons, and signal scales.
+over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$
+means that the prediction error in question was, on the average,
+smaller than the error of a random-walk forecast on the same data.
+Analogously, $MASE>1$ means that the corresponding prediction method
+did \emph{worse}, on average, than the random-walk method.  We chose
+this error metric because it allows for fair comparison across varying
+methods, prediction horizons, and signal scales---attributes that are
+key to any broad study of predictability.
 
@@
   \begin{center}
-  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
-  \hline\hline Signal & LMA MASE  & \arima MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
+  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
+  \hline\hline Signal & LMA MASE & \arima MASE & na\"{i}ve MASE & RW MASE & WPE \\
   \hline
- \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\
 
-\gcc           & $ 1.530\pm 0.021$ & $1.837 \pm0.016 $ & $0.951 \pm 0.001$ & $0.943 \pm 0.001$ \\
+  \col       & $0.050 \pm 0.002$ & $0.599 \pm 0.211$ & $0.571 \pm 0.002$  & $1.001 \pm 0.002$ & $0.513 \pm 0.003$ \\
 
-\svdone     & $ 0.827\pm 0.076$ & $ 0.714\pm 0.075 $ & $2.676\pm4.328$&  $0.957 \pm 0.016$ \\
+  \gcc       & $1.530 \pm 0.021$ & $1.837 \pm 0.016$ & $1.797 \pm 0.010$  & $1.138 \pm 0.011$ & $0.943 \pm 0.001$ \\
 
- \svdtwo    & $1.279 \pm0.020 $ & $2.163 \pm0.027 $ &  $3.054\pm0.040$ &   $0.846 \pm0.004$ \\
+  \svdone    & $0.827 \pm 0.076$ & $0.714 \pm 0.075$ & $2.676 \pm 4.328$  & $0.933 \pm 0.095$ & $0.957 \pm 0.016$ \\
 
-\svdthree     & $0.619 \pm0.021 $ & $0.713 \pm 0.010 $ & $31.386\pm 0.282$ &  $0.716 \pm 0.006$ \\
- \svdfour     & $ 0.779\pm0.036 $ & $0.979 \pm0.032 $ & $2.661\pm0.074$ & $0.825 \pm 0.008$ \\
+  \svdtwo    & $1.279 \pm 0.020$ & $2.163 \pm 0.027$ & $3.054 \pm 0.040$  & $1.125 \pm 0.012$ & $0.846 \pm 0.004$ \\
 
-\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
- \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
+  \svdthree  & $0.619 \pm 0.021$ & $0.713 \pm 0.010$ & $31.386 \pm 0.282$ & $0.707 \pm 0.009$ & $0.716 \pm 0.006$ \\
+
+  \svdfour   & $0.779 \pm 0.036$ & $0.979 \pm 0.032$ & $2.661 \pm 0.074$  & $1.034 \pm 0.035$ & $0.825 \pm 0.008$ \\
+
+  \svdfive   & $0.718 \pm 0.048$ & $2.370 \pm 0.051$ & $20.870 \pm 0.192$ & $1.001 \pm 0.047$ & $0.678 \pm 0.007$ \\
+
+  \svdsix    & $0.739 \pm 0.068$ & $1.438 \pm 0.061$ & $2.197 \pm 0.083$  & $1.060 \pm 0.055$ & $0.748 \pm 0.011$ \\
+
   \hline\hline
@@
   \end{center}
- \label{default}
  \label{tab:error}

### 2014-07-24 results.tex
airplane edits

@@
 
-In this section, we offer an empirical validation of the two
-conjectures offered in Section \ref{sec:intro}, namely:
+In this section, we offer an empirical validation of the two findings
+introduced in Section \ref{sec:intro}, namely:
 
 \begin{enumerate}
+% \item The complexity of a noisy real-valued time series can be
+%   effectively quantifiable by computing its permutation entropy.
 
-\item That the weighted permutation entropy (WPE) of a noisy
-  real-valued time series from an unknown system is correlated with
-  prediction accuracy---i.e., that the predictable structure in an
-  empirical time-series data set can be quantified by its WPE.
+\item The weighted permutation entropy (WPE) of a noisy real-valued
+  time series from an unknown system is correlated with prediction
+  accuracy---i.e., the predictable structure in an empirical
+  time-series data set can be quantified by its WPE.
 
-%\subitem Simply reminders not to be included]]WPE can quantify when
-%a noisy real-valued time series is predictable.I am unsatisfied
-%with predictable in this sentence, need a better word to say "better
-%than random walk" or ``able to be forecast effectively" or ``has the
-%structural capacity to transmit information in a way that the time
-%series can be effectively forecast"
-
-%\item The existence of complex structure in a noisy real-valued time
-%series is quantifiable and this type of complexity is directly
-%correlated with predictability
-
-\item That the way information is generated and processed by an
-  unknown system is correlated with the effectiveness of a given
-  predictor on empirical time-series data from that system
-
-%\subitem Simply reminders not to be included]]We will have shown
-%that the existence whether linear or nonlinear is picked up on with
-%WPE but this point gets at whether the prediction model can use the
-%structure or not (linear can't use nonlinear structure). The right to
-%left shifts in \col and some of the \svd regimes and the lack of
-%shift in \gcc illustrate this nicely.
-
-%\subitem Simply reminders not to be included]]This is the linear vs
-%nonliear vs random we see with the right to left shifts with lower
-%complexity time series.
+\item The relationship between WPE and MASE is a useful empirical
+  heuristic for identifying mismatches between prediction models and
+  time-series data---i.e., there is structure in the data that the
+  model is unable to exploit.
 
@@
 Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
-corresponding MASE values of the best prediction of each time series.
+corresponding MASE values of the best prediction of each of the
+\alert{[[108?]]} time series in this study
 \begin{figure}[htbp]
@@
 %
-The dashed curve is a least-squares log fit of all the points except
-those from {\tt dgesdd$_1$}, which we have excluded for reasons
-explained in the text.  The shaded region boundaries reflect the
-standard deviation of the model in its parameter space.}
+The dashed curve is a least-squares \alert{[[log]]} fit of all the
+points except those from {\tt dgesdd$_1$}, which we have excluded for
+reasons explained in the text.  The shaded region boundaries reflect
+the standard deviation of the model in its parameter space.}
   \label{fig:wpe_vs_mase_best}
@@
 As the plot shows, the best-case prediction error for each trace from
-these eight systems is roughly proportional to the log of the weighted
-permutation entropy, which is consistent with our first conjecture.
-The dashed line in the figure, a least-squares fit of $y = a \log(b x
-+ 1) + c x^16$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
-  for reasons described below} captures this relationship between WPE
-and prediction error quite nicely.  The logarithmic shape of this
-curve makes sense here because any signal with zero entropy should be
+these eight systems is roughly proportional to the \alert{[[log]]} of
+the weighted permutation entropy, which is consistent with our first
+conjecture.  \alert{We need to explain and justify this baroque curve
+  verrry carefully.  Occam's razor it is not.}  The dashed line in the
+figure, a least-squares fit of $y = a \log(b x + 1) + c x^{16}$ to
+these errors\footnote{with the exception of {\tt dgesdd$_1$}, for
+  reasons described below} captures this relationship between WPE and
+prediction error quite nicely.  The logarithmic shape of this curve
+makes sense here because any signal with zero entropy should be
 perfectly predictable, and so should ideally have a MASE value of
 zero.  The shaded region shows the standard deviation of the model in
-its parameter space: i.e., $y = a \log(b x + 1) + c x^16$ with $\pm$ one
-standard deviation on each of the three parameters.
+its parameter space: i.e., $y = a \log(b x + 1) + c x^{16}$ with $\pm$
+one standard deviation on each of the three parameters.
 
@@
 and the range of dynamical behaviors, but they do not cover those
-spaces exhaustively.  Our goal here is an\emph{empirical} assessment
+spaces exhaustively.  Our goal here is an \emph{empirical} assessment
 of the relationship between predictability and complexity, not formal

### 2014-07-24 wpe.tex
airplane edits

@@
 worse yet, they require knowledge of the underlying system.  This is
-not useful in our context, where we want to measure the entropy to
-quantify predictability of an arbitrary time-series.
+not useful in our context, where the target of study is experimental
+time-series data.
 
@@
 result converges, but data length issues can intrude before that
-convergence is reached.  We used this approach to choose $\ell = 6$
-for the experiments in the following section.  This value represents a
+convergence is reached.  We used this approach, which represents a
 good balance between accurate ordinal statistics and finite-data
-effects.
+effects, to choose $\ell$ values for the experiments in the following
+section.
 

### 2014-07-25 concl.tex
synching up

@@
 \section{ Conclusions \& Future Work }\label{sec:conc}
-
-[[Moved this out of intro but it was redundant in the model section
-    and some of it is repeated in the results section.  Maybe here?
-    The forecast methods used in this study were chosen to be a
-    representative sampling of standard prediction strategies, but
-    they do not, of course, cover that space exhaustively.  Our goal
-    here is an empirical assessment of the relationship between
-    predictability and complexity, not formal results about a ``best''
-    predictor for a given time series.  From a practitioner's
-    standpoint, it would be useful to know \emph{ a priori} if a time
-    series contains enough predictive structure to make it worth
-    spending the time and effort searching for a good forecast method.
-    It would also be useful to know if a given method is
-    inadequate---that is, if an alternative method could do better and
-    so one should continue searching.]]
 

### 2014-07-25 methods.tex
synching up

@@
 complexities---are at work in different phases of a single time
-series.  The complexity of \col and \gcc in Figures~\ref{fig:col-ipc}
-and \ref{fig:gcc-ts} appears to be far more consistent over
-time---probably the result of a single generating process.  \svd, in
-contrast, has multiple regimes, each the result of a different
-generating process.  To take advantage of this, we split the signal
-into six different segments, thereby obtaining an array of examples
-for the analyses in the following sections.  For notational
+series.  The \col and \gcc traces in Figures~\ref{fig:col-ipc} and
+\ref{fig:gcc-ts} appear to be far more consistent over time---probably
+the result of a single generating process with consistent complexity.
+\svd, in contrast, has multiple regimes, each the result of a
+different generating process.  To take advantage of this, we split the
+signal into six different segments, thereby obtaining an array of
+examples for the analyses in the following sections.  For notational
 convenience, we refer to these 90 time-series data sets\footnote{15

### 2014-07-25 model.tex
synching up

@@
 signals, as discussed below.
+
+These four forecast methods were chosen to be a representative
+sampling of standard prediction strategies, but they do not, of
+course, cover that space exhaustively.  Our goal here is an empirical
+assessment of the relationship between predictability and complexity,
+not formal results about a ``best'' predictor for a given time series.
+From a practitioner's standpoint, it would be useful to know \emph{ a
+  priori} if a time series contains enough predictive structure to
+make it worth spending the time and effort searching for a good
+forecast method.  It would also be useful to know if a given method is
+inadequate---that is, if an alternative method could do better and so
+one should continue searching.  \alert{[[need to check that this is not
+    redundant, given the other text in this section (and elsewhere in
+    the paper).  it may be better to put this in the conclusion.]]}
 
@@
 
+\alert{We need to go through this section and the results section,
+  rewriting as necessary to account for our addition of RW to the
+  table}
+
 To study the relationship between predictability and complexity, we
@@
   \end{tabular}
-  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts of \col,
-     \gcc, and \svdfive and all four prediction strategies.
-  }
+  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts
+    of \col, \gcc, and \svdfive made using each of the four 
+    strategies studied here.  }
   \label{fig:forecast-example}
@@
 \begin{table*}
-\caption{Mean
-absolute scaled error (MASE) scores and weighted permutation entropies for all eight
-   processes studied in this paper}
+\caption{Mean absolute scaled error (MASE) scores and weighted
+  permutation entropies for all eight processes studied in this paper.
+  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.}
   \begin{center}

### 2014-07-28 related.tex
update boffetta stuff

@@
 (possibly large) number of points that might be needed to estimate the
-$\epsilon$-entropy for small $\epsilon$.  Second, we tie
-information---in the form of the weighted permutation
-entropy---directly to prediction error via MASE values.  Though FSLE
-and $\epsilon$-entropy allow for the comparison of predictability
-between systems, they do not directly provide an estimate of
-prediction error.  \alert{Finally, our approach also holds for
-  \emph{stochastic} systems, where the computation of FTLEs makes no
-  sense and/or doesn't reflect predictability.  [[I remember someone
-      saying this but didn't see it here, so I added it.  Boffetta et
-      al. just do dynamical systems, yes?  And lambda of a stochastic
-      system wouldnt' make sense, yes?]]}
+$\epsilon$-entropy for small $\epsilon$.  Second, we tie information
+---in the form of the weighted permutation entropy---directly to
+prediction error via MASE values.  Though FSLE and $\epsilon$-entropy
+allow for the comparison of predictability between systems, they do
+not directly provide an estimate of prediction error.  Finally, our
+approach also holds for \emph{stochastic} systems, where the
+computation of FSLEs makes no sense and/or doesn't reflect
+predictability, and Boffetta et al.'s analysis requires knowledge of
+the noise and its interaction with the system.

### 2014-07-29 intro.tex
working through

@@
 example, fall in this class.  On the opposite end of this spectrum are
-signals that are what one can call \emph{fully complex}, where the
+signals that are what one could call \emph{fully complex}, where the
 underlying generating process transmits no information at all from the
@@
 This leads naturally to an important and challenging question: given a
-noisy real-valued empirical time series, does there exist a forecast
-model that can leverage the information (if any) that is being
-transmitted from past to future by the underlying generating process?
-A first step in answering this question is to reliably quantify where
-on the complexity spectrum a given time series falls; a second step is
-to determine how complexity and predictability are related in these
-kinds of data sets.  With these answers in hand, one can develop a
-practical strategy for assessing appropriateness of forecast methods
-for a given time series.  If the forecast produced by a particular
-method is poor, for instance, but the time series contains a
-significant amount of predictive structure, one can reasonably
-conclude that that method is inadequate to the task and that one
-should seek another method.  The goal of this paper is to develop
-effective heuristics to put that strategy into practice.
+noisy real-valued time series from an unknown system, does there exist
+a forecast model that can leverage the information (if any) that is
+being transmitted from past to future by the underlying generating
+process?  A first step in answering this question is to reliably
+quantify where on the complexity spectrum a given time series falls; a
+second step is to determine how complexity and predictability are
+related in these kinds of data sets.  With these answers in hand, one
+can develop a practical strategy for assessing appropriateness of
+forecast methods for a given time series.  If the forecast produced by
+a particular method is poor, for instance, but the time series
+contains a significant amount of predictive structure, one can
+reasonably conclude that that method is inadequate to the task and
+that one should seek another method.  The goal of this paper is to
+develop effective heuristics to put that strategy into practice.
 
@@
 is well known that the way information is generated and processed
-internally by a system plays a crucial role in the success of
+internally by a system plays a critical role in the success of
 different forecasting methods---and in the choice of which method is
@@
 not been properly explored, however, in the context of noisy, poorly
-sampled, real-world data from unknown systems.  That exploration---and
+sampled, real-world data from unknown systems.  That exploration, and
 the development of strategies for putting its results into effective
-practice---is the primary contribution of this work.  The empirical
+practice, is the primary contribution of this work.  The empirical
 results in the later sections of this paper not only elucidate the
@@
 situates this work in that context. Section~\ref{sec:methods} covers
-the experimental setup and methods used to collect each time
-series. Section~\ref{sec:model} describes the prediction models used
-in this study.  Section~\ref{sec:meaComplex} reviews permutation
-entropy, the technique that we use to measure complexity.  In
+the experimental setup and methods used to collect the time-series
+data. Section~\ref{sec:model} describes the prediction models used in
+this study.  Section~\ref{sec:meaComplex} reviews permutation entropy,
+the technique that we use to measure complexity.  In
 Section~\ref{sec:results}, we estimate the complexity of each

### 2014-07-29 methods.tex
working through

@@
 programs---the simple \col loop whose performance is depicted in
-Figure~\ref{fig:col-ipc}, as well as two more-complex programs: one
-from the SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK
-linear algebra package (\svd).  In all of these experiments, the
-scalar observation $x_i$ was a measurement of the processor
-performance at time $i$ during the execution of each program.
+Figure~\ref{fig:col-ipc} and two more-complex programs: one from the
+SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK linear
+algebra package (\svd).  In all of these experiments, the scalar
+observation $x_i$ was a measurement of the processor performance at
+time $i$ during the execution of each program.
 % 
@@
 For statistical validation, we collected 15 performance traces from
-each of the three programs.  For an in-depth description of this
-experimental setup,
+each of the three programs.  For an in-depth description of the
+experimental setup used to gather these data,
 % including discussion of the implications of the sampling interval,
@@
 It is a \emph{compiler}: a program that translates code written in a
-high-level language (C, in the case of \gcc) into a lower-level format
-that can be executed by the processor chip.  Its behavior is far more
-complicated than that of \col, as is clear from
-Figure~\ref{fig:gcc-ts}.
+high-level language 
+% (C, in the case of \gcc) 
+into a lower-level format that can be executed by the processor chip.
+Its behavior is far more complicated than that of \col, as is clear
+from Figure~\ref{fig:gcc-ts}.
   \begin{figure}[t]
@@
 \cite{lapack}.  It calculates the singular value decomposition of a
-rectangular $M$ by $N$ matrix with real-valued entries.  For our experiments, we
-choose $M=750$ and $N=1000$ and generate the matrix entries randomly.
+rectangular $M$ by $N$ matrix with real-valued entries.  For our
+experiments, we chose $M=750$ and $N=1000$ and generated the matrix
+entries randomly.
 % 
@@
 this matrix is very interesting, as is clearly visible in Figure
-\ref{fig:svd-ts-colored}.  As the code moves though its different
-phases---diagonalizing the matrix, computing its transpose,
-multiplying, etc.---the processor utilization patterns change quite radically,
-as shown in Figure~\ref{fig:svd-ts-colored}.
+\ref{fig:svd-ts-colored}.  
 \begin{figure}[t]
@@
   \end{figure}
-For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
-instructions), roughly 1.8 instructions are executed per cycle, on the
-average, by the eight processing units on this chip.  After that, the
-IPC moves through a number of different oscillatory regimes, which we
-have color-coded in the figure in order to make textual
-cross-references easy to track.
+As the code moves though its different phases---diagonalizing the
+matrix, computing its transpose, multiplying, etc.---the processor
+utilization patterns change quite radically.  For the first
+$\sim$21,000 measurements (21,000 $\times$ 100,000 instructions),
+roughly 1.8 instructions are executed per cycle, on the average, by
+the eight processing units on this chip.  After that, the IPC moves
+through a number of different oscillatory regimes, which we have
+color-coded in the figure in order to make textual cross-references
+easy to track.
 
 The wide range of behaviors in Figure~\ref{fig:svd-ts-colored}
-provides a unique advantage, for the purposes of this paper, in that a
-number of different generating processes---with a wide range of
+provides a distinct advantage, for the purposes of this paper, in that
+a number of different generating processes---with a wide range of
 complexities---are at work in different phases of a single time

### 2014-07-29 model.tex
working through

@@
 Of course, methods that could capture and exploit the geometry of the
-data, or its temporal patterns, would be far more effective in the
+data and/or its temporal patterns would be far more effective in the
 situations described in the previous paragraph.  The \arima and LMA
@@
 A simple and yet powerful way to capture and exploit the structure of
-data is to fit a hyperplane to the dataset and then use it to make
-predictions.  The roots of this approach date back to the original
-autoregressive schema~\cite{weigend93}, which forecasts the next time
-step through a weighted average of past observations: $$p_i =
-\sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
+data is to fit a hyperplane to the known points and then use it to
+make predictions.  The roots of this approach date back to the
+original autoregressive schema~\cite{weigend93}, which forecasts the
+next time step through a weighted average of past observations: $$p_i
+= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
 generally computed using either an ordinary least squares approach, or
@@
 automated fitting techniques described in~\cite{autoARIMA} to
-accomplish this, producing what we call an \emph{\arima model}.
-This procedure uses sophisticated methods---KPSS unit-root
-tests~\cite{KPSSunit}, a customization of the Canova-Hansen
+accomplish this, producing what we will call an ``\arima model''in the
+rest of this paper.  This procedure uses sophisticated methods---KPSS
+unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
 test~\cite{Canova1995}, and the Akaike information
@@
 the model fitted to the detrended data---to select good values for the
-free parameters of the \arima model.
+free parameters of the ARIMA model.
 
@@
 fashion: i.e., fit the \arima model to the existing data, use that
-model to perform a one-step prediction, rebuild it using the latest
-observations, and iterate until the desired prediction horizon is
-reached.  For consistency, we take the same approach with the other
-three models in this study as well, even though doing so amounts to
-artificially hobbling LMA.
+model to perform a one-step prediction, rebuild the \arima model using
+the latest observations, and iterate until the desired prediction
+horizon is reached.  For consistency, we take the same approach with
+the other three models in this study as well, even though doing so
+amounts to artificially hobbling LMA.
 
@@
 from a $d$-dimensional smooth compact manifold $M$ to
-$\mathbb{R}^{2d+1}$ is a diffeomorphism on $M$---in other words, that
+$\mathbb{R}^{2d+1}$ is a diffeomorphism on $M$: in other words, that
 the reconstructed dynamics and the true (hidden) dynamics have the
-same topology.  This is an extremely powerful result: among other
+same topology.  This is an extremely powerful result; among other
 things, it means that one can model the full system dynamics, up to
-diffeomorphism, without measuring---or even knowing---every one of its
-state variables.
+diffeomorphism, without measuring---or even knowing---anything about
+the state variables.
 
@@
 an estimate of $\tau$~\cite{fraser-swinney} and using the
-false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.
-Some example plots of data from
-Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following
+false-near(est)-neighbor method of~\cite{KBA92} to estimate $m$.  Some
+example plots of data from
+Figures~\ref{fig:col-ipc}-\ref{fig:svd-ts-colored}, embedded following
 this procedure, are shown in Figure~\ref{fig:embedding}.
@@
 underlying generating process is nondeterministic, however, it can
-perform poorly.  Figure~\ref{fig:embedding}b, for instance, appears to
-contain little structure, so LMA may not work well on this signal.
-More structure appears to be present in Figure~\ref{fig:embedding}c,
+perform poorly.  Figure~\ref{fig:embedding}(b), for instance, appears
+to contain little structure, so LMA may not work well on this signal.
+More structure appears to be present in Figure~\ref{fig:embedding}(c),
 but this reconstruction also appears to contain some noise.  The
-question as to how much structure is present in a reconstruction---and
-how much of that structure can be captured and used by LMA---is
-apropos of the central question treated in this paper.  It may be that
-\gcc has some redundancy that LMA cannot exploit, or that the
-structure in \svdfive is effectively obfuscated, from the standpoint
-of the LMA method, by noise.  Without any knowledge of the generating
-process, answers to these questions can only be derived from the data,
-with all of the attendant problems (noise, sampling issues, etc.)  By
-quantifying the balance between redundancy (that is, predictive
-structure) and entropy for these real-valued time series, as shown in
-Section~\ref{sec:results}, we can begin to answer these questions in
+question as to how much structure is present in a reconstruction, and
+how much of that structure can be captured and used by LMA, is apropos
+of the central question treated in this paper.  It may be that \gcc
+has some redundancy that LMA cannot exploit, or that the structure in
+\svdfive is effectively obfuscated, from the standpoint of the LMA
+method, by noise.  Without any knowledge of the generating process,
+answers to these questions can only be derived from the data, with all
+of the attendant problems (noise, sampling issues, and so on).  By
+quantifying the balance between redundancy, predictive structure, and
+entropy for these real-valued time series---as shown in
+Section~\ref{sec:results}---we can begin to answer these questions in
 an effective and practical manner.
@@
 \label{sec:accuracy}
-
-\alert{We need to go through this section and the results section,
-  rewriting as necessary to account for our addition of RW to the
-  table}
 
@@
 generates a very accurate prediction of the \col data, while \arima
-does not.  Horizontal lines result when a constant predictor (e.g.,
-\naive) is used on a non-constant signal.  Point clouds reflect the
-structure of the distribution of the errors---roughly normally
+does not.  Horizontal lines result when a constant predictor like the
+\naive ~method is used on a non-constant signal.  Point clouds reflect
+the structure of the distribution of the errors---roughly normally
 distributed, for example, in the case of LMA on \gcc.  Note that the
-shapes of some of the plots in Figure~\ref{fig:forecast-example}, random walk and \arima on \col for example,  are
-reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
-For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
-two-dimensional embedding with $\tau=1$.  For \arima, the
-correspondence is not quite as simple, since the $p_j$ values are
+shapes of some of the plots in Figure~\ref{fig:forecast-example}
+(e.g., random walk and \arima on \col) are reminiscent of the
+projected embedding in Figure~\ref{fig:embedding}(a).  Indeed, for a
+random-walk predictor, a $p_j$ vs. $c_j$ plot is technically
+equivalent to a two-dimensional embedding with $\tau=1$.  For \arima,
+the correspondence is not quite as simple, since the $p_j$ values are
 linear combinations of a number of past values of the $c_j$, but the
@@
 applicable measurement of forecast accuracy without the problems seen
-in the other measurements."  MASE is a normalized measure: the scaling
-term in the denominator
+in the other measurements."  Like many error metrics, MASE is a
+normalized measure: the scaling term in the denominator
 %
@@
 Analogously, $MASE>1$ means that the corresponding prediction method
-did \emph{worse}, on average, than the random-walk method.  We chose
-this error metric because it allows for fair comparison across varying
-methods, prediction horizons, and signal scales---attributes that are
-key to any broad study of predictability.
+did \emph{worse}, on average, than the random-walk method.  While its
+comparative nature is somewhat different than traditional metrics like
+normalized root mean squared error, MASE has the significant advantage
+of allowing one to make a fair comparison across varying methods,
+prediction horizons, and signal scales---attributes that are key to
+any broad study of predictability.
 
-MASE scores for all 360 experiments are tabulated in the three
-middle columns in Table~\ref{tab:error}.
+MASE scores for all 360 experiments are tabulated in the four middle
+columns in Table~\ref{tab:error}.
 \begin{table*}
@@
   permutation entropies for all eight processes studied in this paper.
-  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.}
+  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
+  \alert{please reorder the columns from left to right to match the
+    order in which the methods were introduced, rather than in the
+    reverse of that order}}
   \begin{center}
@@
   \end{table*}%
-Comparing these with the geometry of the plots in
-Figure~\ref{fig:forecast-example}, one can see some obvious
-correspondences.  The average LMA MASE score for the \col signals was
-$0.050$, for instance, while \arima scored much worse ($0.599$).
-That is, LMA performed roughly 20 times better on \col signals than a
-random-walk predictor, while \arima only outperformed random walk
-by a factor of 1.7.  This is in accordance with the visual appearance
-of the corresponding images in Figure~\ref{fig:forecast-example}.
-While the correspondence between MASE score and the visual appearance
-of these kinds of plots is clear cut in this case, that is not always
-so.  The plots of LMA predictions of \col and \svdfive both lie near
-the diagonal, for instance, but the corresponding MASE scores were
-very different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20
-and 1.4 times better than random-walk forecasts of the same signals).
-This is a result of the normalization in the MASE score calculation.
-Recall from Section~\ref{sec:simple} that the random-walk method
-performs especially poorly on signals that oscillate rapidly.  \col
-fits this description, so the random-walk error on this signal---the
-denominator of the three MASE scores in the first row of the
-Table---is pathologically high, which skews those scores down.
-Random-walk prediction is very effective for the \svdfive signal, on
-the other hand, so the denominator is small and the MASE scores are
-skewed upwards.  Normalization is, as is often the case, a
-double-edged sword.  Here, it facilitates comparisons across signals
-of different types and lengths, but its particular pathologies must be
-taken into account, as discussed at more length in
+In view of the discussion at the end of the previous paragraph, the
+fact that the values in the second-from-right column are not
+identically equal to 1.00 may be somewhat surprising.  \alert{[[Need
+      to add a sentence here...presumably it's because not all RW
+      forecasts are the same?  Why is this more of an issue in
+      \svdthree?]]}  Comparing the values in Table~\ref{tab:error}
+with the geometry of the plots in Figure~\ref{fig:forecast-example},
+one can see some obvious correspondences.  The average LMA MASE score
+for the \col signals was $0.050$, for instance, while \arima scored
+much worse ($0.599$).  That is, LMA performed roughly 20 times better
+on \col signals than a random-walk predictor, while \arima only
+outperformed random walk by a factor of 1.7.  This is in accordance
+with the visual appearance of the corresponding images in
+Figure~\ref{fig:forecast-example}.  While the correspondence between
+MASE score and the visual appearance of these kinds of plots is clear
+cut in this case, that is not always so.  The plots of LMA predictions
+of \col and \svdfive both lie near the diagonal, for instance, but the
+corresponding MASE scores were very different: $0.050$ for \col and
+$0.718$ for \svdfive (resp., 20 and 1.4 times better than random-walk
+forecasts of the same signals).  This is a result of the normalization
+in the MASE score calculation.  Recall from Section~\ref{sec:simple}
+that the random-walk method performs especially poorly on signals that
+oscillate rapidly.  \col fits this description, so the random-walk
+error on this signal---the denominator of the four MASE scores in the
+first row of the Table---is pathologically high, which skews those
+scores down.  Random-walk prediction is very effective for the
+\svdfive signal, on the other hand, so the denominator is small and
+the MASE scores are skewed upwards.  Normalization is, as is often the
+case, a double-edged sword.  Here, it facilitates comparisons across
+signals of different types and lengths, but its particular pathologies
+must be taken into account, as discussed at more length in
 Section~\ref{sec:results}.

### 2014-07-29 related.tex
working through

@@
 generating process and knowledge about what will happen to the error
-if those assumptions hold or fail.  Model-free information analysis
-moves away from those restrictions.  Our approach falls into this
-class: we wish to measure the inherent complexity of an empirical time
-series, then correlate that complexity with the predictive accuracy of
-forecasts made using a number of different methods.
+if those assumptions hold or fail.  Model-\emph{free} information
+analysis moves away from those restrictions.  Our approach falls into
+this class: we wish to measure the inherent complexity of an empirical
+time series, then correlate that complexity with the predictive
+accuracy of forecasts made using a number of different methods.
 
@@
 another.  On the other hand, deterministic systems---including chaotic
-ones---have high redundancy that is maximal in the infinite limit,
-and thus they can be perfectly predicted if observed for long
-enough~\cite{weigend93}.  In practice, it is
-quite difficult to estimate the redundancy of an arbitrary,
-real-valued time series.  Doing so requires knowing either the
-Kolmogorov-Sinai entropy or the values of all positive Lyapunov
-exponents of the system.  Both of these calculations are
-difficult---the latter particularly so if the data are very noisy or
-the generating system is stochastic.
+ones---have high redundancy that is maximal in the infinite limit, and
+thus they can be perfectly predicted if observed for long
+enough~\cite{weigend93}.  In practice, it is quite difficult to
+estimate the redundancy of an arbitrary, real-valued time series.
+Doing so requires knowing either the Kolmogorov-Sinai entropy or the
+values of all positive Lyapunov exponents of the system.  Both of
+these calculations are difficult, the latter particularly so if the
+data are very noisy or the generating system is stochastic.
 
@@
 approach applicable to noisy real-valued time series from all classes
-of systems.
+of systems, deterministic and stochastic.
 
-There has been prior work under a similar title to
+There has been prior work under a very similar title to
 ours~\cite{haven2005}, but there are only superficial similarities
@@
 
-More closely related is the work by Boffetta \emph{et
+More closely related is the work of Boffetta \emph{et
   al.}~\cite{boffetta02}, who investigate the scaling behavior of
 finite-size Lyapunov exponents (FSLE) and $\epsilon$-entropy for a
-wide variety of deterministic systems with known dynamics.  While the
-scaling of these measures acts as a general proxy for predictability
-bounds, this approach differs from our work in a number of fundamental
-ways.  First, we focus on real-world time-series data, where one does
-not necessarily have the ability to perturb or otherwise interact with
-the system of interest, nor can one obtain or manufacture the
-(possibly large) number of points that might be needed to estimate the
-$\epsilon$-entropy for small $\epsilon$.  Second, we tie information
----in the form of the weighted permutation entropy---directly to
-prediction error via MASE values.  Though FSLE and $\epsilon$-entropy
-allow for the comparison of predictability between systems, they do
-not directly provide an estimate of prediction error.  Finally, our
-approach also holds for \emph{stochastic} systems, where the
-computation of FSLEs makes no sense and/or doesn't reflect
-predictability, and Boffetta et al.'s analysis requires knowledge of
-the noise and its interaction with the system.
+wide variety of deterministic systems with known dynamics and additive
+noise.  While the scaling of these measures acts as a general proxy
+for predictability bounds, this approach differs from our work in a
+number of fundamental ways.  First, we focus on real-world time-series
+data, where one does not necessarily have the ability to perturb or
+otherwise interact with the system of interest, nor can one obtain or
+manufacture the (possibly large) number of points that might be needed
+to estimate the $\epsilon$-entropy for small $\epsilon$.  Second, we
+do not require \emph{a priori} knowledge about the noise and its
+interaction with the system.  Third, we tie information---in the form
+of the weighted permutation entropy---directly to prediction error via
+MASE values.  Though FSLE and $\epsilon$-entropy allow for the
+comparison of predictability between systems, they do not directly
+provide an estimate of prediction error.  Finally, our approach also
+holds for stochastic systems, where neither the FLSEs nor their
+relationship to predictability is well defined.

### 2014-07-29 results.tex
synching up
working through

@@
 
+\alert{Are there any updates that we need to do besides the curve
+  shape and the addition of random walk results?}
+
 In this section, we offer an empirical validation of the two findings
@@
   heuristic for identifying mismatches between prediction models and
-  time-series data---i.e., there is structure in the data that the
-  model is unable to exploit.
+  time-series data---i.e., when there is structure in the data that
+  the model is unable to exploit.
 
@@
 %
-The dashed curve is a least-squares \alert{[[log]]} fit of all the
-points except those from {\tt dgesdd$_1$}, which we have excluded for
-reasons explained in the text.  The shaded region boundaries reflect
-the standard deviation of the model in its parameter space.}
+The solid curve is a least-squares log fit of these points.
+% 
+% of all the points except those from {\tt dgesdd$_1$}, which we have
+% excluded for reasons explained in the text.
+% 
+The dashed curves reflect the standard deviation of the model in its
+parameter space.  The meaning of the shaded region is explained in the
+text.}
   \label{fig:wpe_vs_mase_best}
 \end{figure}
-As the plot shows, the best-case prediction error for each trace from
-these eight systems is roughly proportional to the \alert{[[log]]} of
-the weighted permutation entropy, which is consistent with our first
-conjecture.  \alert{We need to explain and justify this baroque curve
-  verrry carefully.  Occam's razor it is not.}  The dashed line in the
-figure, a least-squares fit of $y = a \log(b x + 1) + c x^{16}$ to
-these errors\footnote{with the exception of {\tt dgesdd$_1$}, for
-  reasons described below} captures this relationship between WPE and
-prediction error quite nicely.  The logarithmic shape of this curve
-makes sense here because any signal with zero entropy should be
-perfectly predictable, and so should ideally have a MASE value of
-zero.  The shaded region shows the standard deviation of the model in
-its parameter space: i.e., $y = a \log(b x + 1) + c x^{16}$ with $\pm$
-one standard deviation on each of the three parameters.
+\alert{As the plot shows, the best-case prediction error for each trace from
+these eight systems is roughly proportional to the log of the weighted
+permutation entropy, which is consistent with our first conjecture.}
+The dashed line in the figure, a least-squares fit of $y = a \log(b x
++ 1)$ to these errors,
+% 
+% \footnote{with the exception of {\tt
+%     dgesdd$_1$}, for reasons described below} 
+% 
+captures this relationship between WPE and prediction error quite
+nicely.  The logarithmic shape of this curve makes sense here because
+any signal with zero entropy should be perfectly predictable, and so
+should ideally have a MASE value of zero.  The shaded region shows the
+standard deviation of the model in its parameter space: i.e., $y = a
+\log(b x + 1) + c x^{16}$ with $\pm$ one standard deviation on each of
+the three parameters.
 

### 2014-07-29 wpe.tex
working through

@@
 information in the present that is also in the past.  The second part
-is the aforementioned \emph{Shannon entropy rate}.  It seems obvious
-that the more redundancy in a signal, the more predictable it is.
+is the aforementioned Shannon entropy rate.  It seems obvious that the
+more redundancy in a signal, the more predictable it should be.
 Redundancy is critical to prediction, in general.  And the specific
-\emph{form} of the redundancy dictates whether a particular prediction
-method will work well or poorly on the corresponding signal.  A linear
-method cannot capture or make use of nonlinear redundancy, for
-instance.  This issue---which has been explored extensively using
-clean data and/or systems where the generating process is known, but
-not so much with empirical data from unknown systems---is central to
-the claims in this paper and the discussion in the following section.
+\emph{form} of the redundancy should dictate whether a particular
+prediction method will work well or poorly on the corresponding
+signal.  A linear method cannot capture or make use of nonlinear
+redundancy, for instance.  This issue---which has been explored
+extensively using clean data and/or systems where the generating
+process is known, but not so much with empirical data from unknown
+systems---is central to the claims in this paper and the discussion in
+the following section.
 
@@
 worse yet, they require knowledge of the underlying system.  This is
-not useful in our context, where the target of study is experimental
-time-series data.
+inappropriate in our context, where the target of study is
+experimental time-series data.
 

### 2014-07-30 model.tex
pushing changes midstream, in case of lossage

@@
 inadequate---that is, if an alternative method could do better and so
-one should continue searching.  \alert{[[need to check that this is not
+one should continue searching.  \alert{need to check that this is not
   redundant, given the other text in this section (and elsewhere in
-    the paper).  it may be better to put this in the conclusion.]]}
+  the paper).  it may be better to put this in the conclusion.}
 
@@
 To study the relationship between predictability and complexity, we
-use the four methods outlined above to generate predictions of all 360
-traces described in Section~\ref{sec:methods}, then calculate the
-error of the predictions with respect to the true continuations.
+use the four methods outlined above to generate predictions of all 120
+\alert{we had 360 there.  seems like it should be 8 times 15 which is
+  120.}  traces described in Section~\ref{sec:methods}, then calculate
+the error of the predictions with respect to the true continuations.
 Specifically, we split each time series into two pieces: the first
@@
 is the average in-sample forecast error for a random-walk prediction
-over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$
+over the initial training signal $\{x_i\}^n_{i=1}$.  That is, MASE$<1$
 means that the prediction error in question was, on the average,
-smaller than the error of a random-walk forecast on the same data.
-Analogously, $MASE>1$ means that the corresponding prediction method
-did \emph{worse}, on average, than the random-walk method.  While its
-comparative nature is somewhat different than traditional metrics like
-normalized root mean squared error, MASE has the significant advantage
-of allowing one to make a fair comparison across varying methods,
-prediction horizons, and signal scales---attributes that are key to
-any broad study of predictability.
+smaller than the average error of a random-walk forecast on the
+training data.  Analogously, MASE$>1$ means that the corresponding
+prediction method did \emph{worse}, on average, than the random-walk
+method.  While its comparative nature is somewhat different than
+traditional metrics like normalized root mean squared error, MASE has
+the significant advantage of allowing one to make a fair comparison
+across varying methods, prediction horizons, and signal
+scales---attributes that are key to any broad study of predictability.
 
@@
 In view of the discussion at the end of the previous paragraph, the
-fact that the values in the second-from-right column are not
-identically equal to 1.00 may be somewhat surprising.  \alert{[[Need
-      to add a sentence here...presumably it's because not all RW
-      forecasts are the same?  Why is this more of an issue in
-      \svdthree?]]}  Comparing the values in Table~\ref{tab:error}
-with the geometry of the plots in Figure~\ref{fig:forecast-example},
-one can see some obvious correspondences.  The average LMA MASE score
-for the \col signals was $0.050$, for instance, while \arima scored
-much worse ($0.599$).  That is, LMA performed roughly 20 times better
-on \col signals than a random-walk predictor, while \arima only
-outperformed random walk by a factor of 1.7.  This is in accordance
-with the visual appearance of the corresponding images in
-Figure~\ref{fig:forecast-example}.  While the correspondence between
-MASE score and the visual appearance of these kinds of plots is clear
-cut in this case, that is not always so.  The plots of LMA predictions
-of \col and \svdfive both lie near the diagonal, for instance, but the
-corresponding MASE scores were very different: $0.050$ for \col and
-$0.718$ for \svdfive (resp., 20 and 1.4 times better than random-walk
-forecasts of the same signals).  This is a result of the normalization
-in the MASE score calculation.  Recall from Section~\ref{sec:simple}
-that the random-walk method performs especially poorly on signals that
-oscillate rapidly.  \col fits this description, so the random-walk
-error on this signal---the denominator of the four MASE scores in the
-first row of the Table---is pathologically high, which skews those
-scores down.  Random-walk prediction is very effective for the
-\svdfive signal, on the other hand, so the denominator is small and
-the MASE scores are skewed upwards.  Normalization is, as is often the
-case, a double-edged sword.  Here, it facilitates comparisons across
-signals of different types and lengths, but its particular pathologies
-must be taken into account, as discussed at more length in
-Section~\ref{sec:results}.
+fact that the values in the second-from-\alert{left} column are not
+identically equal to 1.00 may be somewhat surprising.  This can happen
+due to differences between in-sample and out-of-sample forecasting;
+the last 10\% of \svdthree, for instance---the out-of-sample
+signal---was easier, on the average, for RW to predict than the first
+90\%. \alert{[[Liz to do: add pointer to Josh's paragraph for deeper
+      analysis]]}.
+
+Comparing the values in Table~\ref{tab:error} with the geometry of the
+plots in Figure~\ref{fig:forecast-example}, one can see some obvious
+correspondences.  The average LMA MASE score for the \col signals was
+$0.050$, for instance, while \arima scored much worse ($0.599$).  That
+is, LMA performed roughly 20 times better on \col signals than a
+random-walk predictor, while \arima only outperformed random walk by a
+factor of 1.7.  This is in accordance with the visual appearance of
+the corresponding images in Figure~\ref{fig:forecast-example}.
+
+In other cases, the correspondence between MASE score and the visual
+appearance of these kinds of plots is not so clear cut.  The plots of
+LMA predictions of \col and \svdfive both lie near the diagonal, for
+instance, but the corresponding MASE scores were very different:
+$0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times
+better than random-walk forecasts of the same signals).  This is a
+result of the normalization in the MASE score calculation.  Recall
+from Section~\ref{sec:simple} that the random-walk method performs
+especially poorly on signals that oscillate rapidly.  \col fits this
+description, so the random-walk error on this signal---the denominator
+of the four MASE scores in the first row of the Table---is
+pathologically high, which skews those scores down.  Random-walk
+prediction is very effective for the \svdfive signal, on the other
+hand, so the denominator is small and the MASE scores are skewed
+upwards.
+
+Normalization is, as is often the case, a double-edged sword.  Here,
+it facilitates comparisons across signals of different types and
+lengths, but its particular pathologies must be taken into account, as
+discussed at more length in Section~\ref{sec:results}.
 

### 2014-07-30 results.tex
pushing changes midstream, in case of lossage

@@
 
-\alert{Are there any updates that we need to do besides the curve
-  shape and the addition of random walk results?}
-
 In this section, we offer an empirical validation of the two findings
@@
 
-The experiments below involve three different prediction methods
+The experiments below involve four different prediction methods
 applied to time-series data from eight different systems: {\tt
-  col\_major}, {\tt 403.gcc}, and the six different regimes of the
+  col\_major}, {\tt 403.gcc}, and the six different segments of the
 {\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective
@@
 Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
-corresponding MASE values of the best prediction of each of the
-\alert{[[108?]]} time series in this study
-\begin{figure}[htbp]
+corresponding MASE values of the \emph{best} prediction for each of
+the 120 time series in this study.
+\begin{figure}
   \centering
-  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy}
+  \includegraphics[width=1.1\columnwidth]{figs/new_prediction_vs_entropy}
   \caption{Weighted permutation entropy vs. mean absolute scaled error
-    (MASE) of the best prediction of each time series.
-%
-% (across all trials and all prediction methods)
-%
-The solid curve is a least-squares log fit of these points.
+    (MASE) of the best prediction of each time series.  The solid
+    curve is a least-squares log fit of these points.
 % 
@@
 \end{figure}
-\alert{As the plot shows, the best-case prediction error for each trace from
-these eight systems is roughly proportional to the log of the weighted
-permutation entropy, which is consistent with our first conjecture.}
-The dashed line in the figure, a least-squares fit of $y = a \log(b x
-+ 1)$ to these errors,
 % 
-% \footnote{with the exception of {\tt
-%     dgesdd$_1$}, for reasons described below} 
+% As the plot shows, the best-case prediction error for each trace from
+% these eight systems is roughly proportional to the log of the weighted
+% permutation entropy, which is consistent with our first conjecture.
 % 
-captures this relationship between WPE and prediction error quite
-nicely.  The logarithmic shape of this curve makes sense here because
-any signal with zero entropy should be perfectly predictable, and so
-should ideally have a MASE value of zero.  The shaded region shows the
-standard deviation of the model in its parameter space: i.e., $y = a
-\log(b x + 1) + c x^{16}$ with $\pm$ one standard deviation on each of
-the three parameters.
+There is an obvious upward trend, which is consistent with our first
+conjecture.  However, a simple linear fit is a bad idea here.  First,
+any signal with zero entropy should be perfectly predictable (i.e.,
+MASE $\approx 0$), so any curve fitted to these data should pass
+through the origin.  Moreover, WPE does not grow without bound, so one
+would expect the patterns in the WPE-MASE pairs to reach some sort of
+asymptote.  For these reasons, we chose to fit a function of the form
+$y = a \log(b x + 1)$ to these points, with $y =$ WPE and $x=$ MASE.
+\alert{can we cite a figure of merit for this fit?  we also need to
+  say what $a$ and $b$ were.}  The solid curve in the figure shows
+this fit; the dashed curves show the standard deviation of this model
+in its parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one
+standard deviation on each of the two parameters.  Points that fall
+within this deviation volume \alert{(light grey)} correspond to
+predictions that are comparable to the best ones found in this study;
+points that fall \emph{above} that volume \alert{(dark grey)} are
+better still.  We chose to truncate the shaded region because of a
+subtle point about the MASE of an ideal predictor, which should not be
+larger than 1 unless the training and test signals are different.
+This is discussed at more length below.
+%[[make sure that's true]]
 
-This is not a formal result.  The forecast methods and data sets used
-here were chosen to span the space of standard prediction strategies
-and the range of dynamical behaviors, but they do not cover those
-spaces exhaustively.  Our goal here is an \emph{empirical} assessment
-of the relationship between predictability and complexity, not formal
-results about a ``best'' predictor for a given time series.  There may
-be other methods that produce lower MASE values than those in
-Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
-above and below the curve in this plot strongly suggests a pattern of
-(inverse) correlation between the underlying predictability of a time
-series and its WPE.  The rest of this section describes our results in
-more detail---including the measures taken to assure meaningful
-comparisons across methods, trials, and programs---and elaborates on
-the meaning of the dashed curve in the figure.
+The curves and regions in Figure~\ref{fig:wpe_vs_mase_best} are a
+graphical representation of the correlation in the first finding.  It
+is, we believe, a useful heuristic for determining whether a given
+prediction method is well matched to a particular time series.  It is
+not, of course, a formal result.  The forecast methods and data sets
+used here were chosen to span the space of standard prediction
+strategies and the range of dynamical behaviors, but they do not cover
+those spaces exhaustively.  Our goal here is an \emph{empirical}
+assessment of the relationship between predictability and complexity,
+not formal results about a ``best'' predictor for a given time series.
+There may be other methods that produce lower MASE values than those
+in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
+above and below the one-$\sigma$ region about the dashed curve in this
+plot strongly suggests a pattern of correlation between the underlying
+predictability of a time series and its WPE.  The rest of this section
+describes these results and claims in more detail---including the
+measures taken to assure meaningful comparisons across methods,
+trials, and programs---and elaborates on the meaning of the different
+curves and limits in the figure.
 
-Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
-set of experiments.
-\begin{figure}
+\alert{Needs to get added somewhere here: The positions of the WPE
+  vs. error points, and the reasons to choose a particular function to
+  fit to them, will be different if you use another error metric.
+  This one is just good for error $=$ MASE.  But MASE is good for
+  global comparisons and we got lots of other time-series data to fall
+  on it (list), so we believe it.}
+
+Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the
+\emph{full} set of experiments.
+\begin{figure*}
   \centering
@@
 %    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
-  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy4a}
+  \includegraphics[width=1.7\columnwidth]{figs/new_predictions_vs_entropy4a}
 \caption{WPE vs. MASE for all trials, methods, and systems---with the
   exception of \svdone, \svdthree, and \svdfive are omitted from the
-  top plot for scale reasons. \alert{there's also a version with all plots
-  stacked (4x1) instead of (2x2)}
+  top plot for scale reasons.
 %
@@
 errors, can be found in Table~\ref{tab:error}.  The dashed curve and
-shaded region is the same as in the previous figure.  }
+shaded regions are the same as in the previous figure.  }
     \label{fig:wpe_vs_mase_all}
-\end{figure}
+\end{figure*}
 There are 15 points in each cluster, one for each trial.  (The points
@@
 MASE variance is larger, which spreads out the clusters horizontally.
+The mean MASE scores of predictions generated with the nonlinear LMA
+method, for instance, are generally closer to the dashed curve; the
+\arima method clusters are more widely spread, the \naive ~clusters
+even more so.
+
+The main thing to note here, however, is not the geometric details of
+the clusters, but the fact that many of them are to the right of
+and/or below the dashed curve that identifies the boundary of the
+shaded region.  That is, these predictions are not as good as our
+heuristic suggests they could be.  Focusing in on any cluster of
+points makes this clear: LMA works best for \svdsix, for instance,
+followed by the random-walk prediction method, then \arima and \naive.
+The idea behind the shaded region is to identify when there is
+predictive structure in the data that a given method is not exploiting
+(viz., \arima on \svdsix)---i.e., that one could do better with
+another forecast strategy.
+
 The distribution of the cluster \emph{positions} is quite different in
-the three plots.  The mean MASE scores of predictions generated with
-the nonlinear LMA method are generally closer to the dashed curve; the
-\arima method clusters are more widely spread, the \naive ~clusters
-even more so.  This geometric validation of the second conjecture in
-this paper is discussed in more detail in the following paragraphs.
+the three plots.
 
@@
 transmits very little information forward in time and provides almost
-no structure for prediction methods to work with.
-\alert{This needs to be reworded to reflect the typo, \naive is no longer the winner.
-MASE is $1.797 \pm 0.010$, rw actually wins $MASE = 1.1381 \pm 0.011$  Here, the \naive
-~method performs best ($MASE=0.951 \pm 0.001$), simply because its
-calculation of the mean filters out the noise in the signal.  Since
-fitting a hyperplane using least squares should have similar effects,
-the fact that LMA outperforms \arima ($1.530 \pm 0.021$ vs. $1.837
-\pm 0.016$) is somewhat counterintuitive.  However, the small amount
-of predictive structure that is present in this signal is nonlinear
-(cf.,~\cite{mytkowicz09}), and LMA is designed to capture and exploit
-that kind of structure.}
+no structure for prediction methods to work with.  \alert{This needs
+  to be reworded to reflect the typo, \naive is no longer the
+  winner. MASE is $1.797 \pm 0.010$, rw actually wins MASE$=1.1381 \pm
+  0.011$ Here, the \naive ~method performs best (MASE$=0.951 \pm
+  0.001$), simply because its calculation of the mean filters out the
+  noise in the signal.  Since fitting a hyperplane using least squares
+  should have similar effects, the fact that LMA outperforms \arima
+  ($1.530 \pm 0.021$ vs. $1.837 \pm 0.016$) is somewhat
+  counterintuitive.  However, the small amount of predictive structure
+  that is present in this signal is nonlinear
+  (cf.,~\cite{mytkowicz09}), and LMA is designed to capture and
+  exploit that kind of structure.}
 
@@
 top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
-score of 0.6, which is better than LMA and \arima.  In five of the
-15 \svdone trials, however, there were step changes in the signal.
-The \naive ~method has a very difficult time with signals like this,
+score of 0.6, which is better than LMA and \arima.  In five of the 15
+\svdone trials, however, there were step changes in the signal.  The
+\naive ~method has a very difficult time with signals like this,
 particularly if there are multiple step changes.  This raised the MASE
@@
   $\approx 2.5$, as well as two points that are beyond the domain of
-  the graph, at MASE 11.2 and 14.8.}, and in turn raising both the
-mean and variance of this set of trials.
+  the graph, at MASE $\approx 11.2--14.8$.}, and in turn raising both
+the mean and variance of this set of trials.
 

### 2014-07-30 results.tex
new draft of intro through results

@@
 % 
-There is an obvious upward trend, which is consistent with our first
-conjecture.  However, a simple linear fit is a bad idea here.  First,
-any signal with zero entropy should be perfectly predictable (i.e.,
-MASE $\approx 0$), so any curve fitted to these data should pass
-through the origin.  Moreover, WPE does not grow without bound, so one
-would expect the patterns in the WPE-MASE pairs to reach some sort of
-asymptote.  For these reasons, we chose to fit a function of the form
-$y = a \log(b x + 1)$ to these points, with $y =$ WPE and $x=$ MASE.
-\alert{can we cite a figure of merit for this fit?  we also need to
-  say what $a$ and $b$ were.}  The solid curve in the figure shows
-this fit; the dashed curves show the standard deviation of this model
-in its parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one
-standard deviation on each of the two parameters.  Points that fall
-within this deviation volume \alert{(light grey)} correspond to
-predictions that are comparable to the best ones found in this study;
-points that fall \emph{above} that volume \alert{(dark grey)} are
-better still.  We chose to truncate the shaded region because of a
-subtle point about the MASE of an ideal predictor, which should not be
-larger than 1 unless the training and test signals are different.
-This is discussed at more length below.
+There is an obvious upward trend, which is consistent with the notion
+that there is a pattern in the WPE-MASE relationship.  However, a
+simple linear fit is a bad idea here.  First, any signal with zero
+entropy should be perfectly predictable (i.e., MASE $\approx 0$), so
+any curve fitted to these data should pass through the origin.
+Moreover, WPE does not grow without bound, so one would expect the
+patterns in the WPE-MASE pairs to reach some sort of asymptote.  For
+these reasons, we chose to fit a function of the form $y = a \log(b x
++ 1)$ to these points, with $y =$ WPE and $x=$ MASE.  \alert{can we
+  cite a figure of merit for this fit?  we also need to say what $a$
+  and $b$ were.}  The solid curve in the figure shows this fit; the
+dashed curves show the standard deviation of this model in its
+parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
+deviation on each of the two parameters.  Points that fall within this
+deviation volume \alert{(light grey)} correspond to predictions that
+are comparable to the best ones found in this study; points that fall
+\emph{above} that volume \alert{(dark grey)} are better still.  We
+chose to truncate the shaded region because of a subtle point
+regarding the MASE of an ideal predictor, which should not be larger
+than 1 unless the training and test signals are different.  This is
+discussed at more length below.
 %[[make sure that's true]]
@@
 The curves and regions in Figure~\ref{fig:wpe_vs_mase_best} are a
-graphical representation of the correlation in the first finding.  It
+graphical representation of the first finding.  This representation
 is, we believe, a useful heuristic for determining whether a given
@@
 
-\alert{Needs to get added somewhere here: The positions of the WPE
-  vs. error points, and the reasons to choose a particular function to
-  fit to them, will be different if you use another error metric.
-  This one is just good for error $=$ MASE.  But MASE is good for
-  global comparisons and we got lots of other time-series data to fall
-  on it (list), so we believe it.}
-
-Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the
-\emph{full} set of experiments.
+Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
+set of experiments.
 \begin{figure*}
@@
   exception of \svdone, \svdthree, and \svdfive are omitted from the
-  top plot for scale reasons.
+  top-left plot for scale reasons, as described in the text.
 %
@@
 Numerical values, including means and standard deviations of the
-errors, can be found in Table~\ref{tab:error}.  The dashed curve and
-shaded regions are the same as in the previous figure.  }
+errors, can be found in Table~\ref{tab:error}.  The curves and shaded
+regions are the same as in the previous figure.  }
     \label{fig:wpe_vs_mase_all}
@@
 in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
-for the corresponding trace in any of the three plots in
+for the corresponding trace in any of the four plots in
 Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
@@
 \arima method clusters are more widely spread, the \naive ~clusters
-even more so.
+even more so.  A few of the clusters have very high variance; these
+are discussed later in this section.
 
-The main thing to note here, however, is not the geometric details of
-the clusters, but the fact that many of them are to the right of
-and/or below the dashed curve that identifies the boundary of the
-shaded region.  That is, these predictions are not as good as our
-heuristic suggests they could be.  Focusing in on any cluster of
+The main thing to note here, however, is not the details of the shapes
+of the clusters, but rather their positions in the four plots:
+specifically, the fact that many of them are to the right of and/or
+below the dashed curve that identifies the boundary of the shaded
+region.  \emph{These predictions are not as good as our heuristic
+  suggests they could be.}  Focusing in on any single cluster of
 points makes this clear: LMA works best for \svdsix, for instance,
 followed by the random-walk prediction method, then \arima and \naive.
-The idea behind the shaded region is to identify when there is
-predictive structure in the data that a given method is not exploiting
-(viz., \arima on \svdsix)---i.e., that one could do better with
-another forecast strategy.
-
-The distribution of the cluster \emph{positions} is quite different in
-the three plots.
+When a practioner calculates an WPE vs. MASE value that is outside the
+shaded region, that should suggest to her that the prediction method
+is not well matched to the task at hand: that is, the time series has
+more predictive structure than the method is able to use.
+% 
+% The idea behind the shaded region is to help practitioners identify
+% when there is predictive structure in the data that a given method is
+% not exploiting (viz., \arima on \svdsix)---i.e., that one could do
+% better with another forecast strategy.
+% 
+In this case---e.g., \arima on \svdsix---she should try a different
+method.  The position of the LMA cluster for \svdsix, on the other
+hand, reflects this method's ability to capture and exploit the
+structure that is present in this signal and suggests to the
+practitioner that [[she really can't do any better]].  The following
+discussion addresses the details that underlie these claims.
+% 
+% The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
+% and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
+% argument at the end of the previous paragraph.  
+%
 
 Though \col is a very simple program, its dynamics are actually quite
-complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
-graphical representations in the left-hand column of
-Figure~\ref{fig:forecast-example} that the \naive ~and \arima
-prediction methods do not perform as well on this signal as the
-nonlinear LMA method.  The MASE scores of the \naive ~and \arima
-predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
-across all 15 trials.  That is, these methods are performing only
-$\approx 1.7$ times better than the random-walk method, a primitive
-strategy that simply uses the current value as the prediction.
-However, the WPE value for the \col trials is $0.513 \pm 0.003$, which
-is in the center of the complexity spectrum described in
-Section~\ref{sec:intro}.
+complicated, as discussed in Section~\ref{sec:intro}.  Recall from
+Figure~\ref{fig:forecast-example} and Table~\ref{tab:error} that the
+\naive, \arima, and (especially) random-walk prediction methods do not
+perform very well on this signal.  The MASE scores of these
+predictions are $0.571 \pm 0.002$, $1.001 \pm 0.002$, and $0.599 \pm
+0.211$, respectively, across all 15 trials.  That is, \naive ~and
+\arima are performing only $\approx 1.7$ times better than the
+random-walk method, a primitive strategy that simply uses the current
+value as the prediction.  However, the WPE value for the \col trials
+is $0.513 \pm 0.003$, which is in the center of the complexity
+spectrum described in Section~\ref{sec:intro}.
 
@@
 information transfer in the signal, but predictions with comparatively
-poor MASE scores---is obvious in the geometry of the top two images in
-Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
-the right of the dashed curve.  \emph{This indicates that these
-  methods are not leveraging the available information in the signal}.
-The dynamics of \col may be complicated, but they are not
-unstructured.  This signal is nonlinear and
-deterministic\cite{mytkowicz09}, and if one uses a prediction
+poor MASE scores---is obvious in the geometry of the two left-hand
+images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
+are far to the right of and/or below the dashed curve.  Again, this
+indicates that these methods are not leveraging the available
+information in the signal.  The dynamics of \col may be complicated,
+but they are not unstructured.  This signal is nonlinear and
+deterministic~\cite{mytkowicz09}, and if one uses a prediction
 technique that is based a nonlinear model (LMA)---rather than a method
-that simply predicts the running mean (\naive) or one that uses a
-linear model (\arima)---the MASE score is much better: $0.050 \pm
-0.001$.  This prediction is 20 times more accurate than a random-walk
-forecast, which is more in line with the level of predictive structure
-that the low WPE value suggests is present in the signal.
-
-The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
-and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
-argument at the end of the previous paragraph.  It can allow
-practitioners to recognize when a prediction method is not well
-matched to the task at hand: that is, when the time series has more
-predictive structure than the method is able to use.  When the
-predictions are well below that curve (as in the top and middle images
-in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
-method.  Conversely, the position of the LMA cluster---well to the
-left of the other two, near the dashed curve---reflects this method's
-ability to capture and exploit the structure that is present in this
-signal.  [[Liz to wordsmith: Again, this curve was determined from a
-    finite set of methods and data.  But we worked hard to make these
-    experiments representative and comprehensive: forecast methods
-    from simple to sophisticated; time-series data from a system whose
-    behavior spans the dynamical behavior space.  While we are
-    cautiously opmistic about its generality, more exploration will be
-    required to firm up that conclusion.  Say that Henon, Lorenz and
-    SFI A fall within the colored region too, as do nonlinear
-    transformations of \svdtwo and \svdsix.  Add here how much the fit
-    changed with those additions.  As we add examples to this list,
-    this empirical result will firm up.]]
+that simply predicts the running mean (\naive) or the previous value
+(random walk), or one that uses a linear model (\arima)---the MASE
+score is much better: $0.050 \pm 0.001$.  This prediction is 20 times
+more accurate than a random-walk forecast, which is more in line with
+the level of predictive structure that the low WPE value suggests is
+present in the signal.  The random-walk MASE scores for \col are all
+$\approx 1$---as one would expect---pushing those points well below
+the shaded region.  Clearly the stationarity assumption on which this
+method is based does not hold for this signal.
 
 Note that the \col points (blue {\color{blue}$+$} icons) are clustered
-very tightly in the lower left quadrant of the LMA and \naive ~MASE
-vs. WPE plots, but spread out horizontally in the \arima plot.
-This is because of the way that these models are built
-\cite{autoARIMA}.  If a KPSS test of the time series in question
-indicates that it is nonstationary, the \arima recipe adds an
-integration term to the model.  This test gives mixed results in the
-case of the \col process, flagging five of the 15 trials as stationary
-and ten as nonstationary.  ARIMA models without an integration term
-perform more poorly on this signal, which increases the error, thereby
-spreading out the points\footnote{We tested this hypothesis by forcing
-  the inclusion of an integration term in the five cases where a KPSS
-  test indicated that such a term was not needed.  This action removed
-  the spread, pushing all 15 of the \col \arima points in
-  Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.}.
-\alert{This paragraph probably needs more, in response to the ref's
-  gritch about fitting.  We certainly want to call his attention to
-  it, either way.}
+very tightly in the lower left quadrant of the \naive, random-walk,
+and LMA plots in Figure~\ref{fig:wpe_vs_mase_all}, but spread out
+horizontally in the \arima plot.  This is because of the way the
+\arima process builds ARIMA models \cite{autoARIMA}.  If a KPSS test
+of the time series in question indicates that it is nonstationary, the
+\arima recipe adds an integration term to the model.  This test gives
+mixed results in the case of the \col process, flagging five of the 15
+trials as stationary and ten as nonstationary.  ARIMA models without
+an integration term perform more poorly on this signal, which
+increases the error, thereby spreading out the points.  We tested this
+hypothesis by forcing the inclusion of an integration term in the five
+cases where a KPSS test indicated that such a term was not needed.
+This action removed the spread, pushing all 15 of the \col ~ \arima
+points in Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.
+\alert{This paragraph needs more, in response to the ref's gritch
+  about fitting: the stuff about \arima being an automated recipe that
+  can give bad results, and the fact that our technique can tell when
+  it does.  And we should call his attention to it in that part of the
+  response letter.}
 
@@
 structure that can, in theory, be used to predict the future course of
-the time series.  As mentioned above, though, the effectiveness of any
-prediction strategy depends on how well it leverages the information
-that is available in the data: methods that use linear models, for
-instance, may fail when applied to nonlinear processes.
-%
-% (This is the basis for the second conjecture above.)
-%
-This match between model and task is an issue in the \svdfive
-experiments as well.  The MASE scores of the \naive ~and \arima
-predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
-0.051$, respectively: that is, 20.87 and 2.37 times worse than a
-simple random walk forecast of the same signals\footnote{The \naive
-  ~MASE score is large because of the bimodal nature of the
-  distribution of the values of the signal, which makes guessing the
-  mean a particularly bad strategy.  The same thing is true of the
-  \svdthree signal.}.  As before, the positions of these points on a
-WPE vs. MASE plot---significantly below and to the right of the shaded
-region---should suggest to a practitioner that there is more structure
-in this signal than the corresponding method is able to leverage, and
-that a different method might do better.  Indeed, for \svdfive, the
-LMA method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster
-of results that is much closer to the dashed curve on the WPE-MASE
-plot.  Again, this validates our second conjecture: the LMA method can
-capture and reproduce the way in which the \svdfive system processes
+the time series.  
+% As mentioned above, though, the effectiveness of any prediction
+% strategy depends on how well it leverages the information that is
+% available in the data: methods that use linear models, for instance,
+% may fail when applied to nonlinear processes.  (This is the basis
+% for the second conjecture above.)
+In the \svdfive results, this match between model and task manifests
+quite clearly.  The MASE scores of the \naive ~and \arima predictions
+for this system are $20.870 \pm 0.192$ and $2.370 \pm 0.051$,
+respectively: that is, 20.87 and 2.37 times worse than a simple random
+walk forecast of the same signals\footnote{The \naive ~MASE score is
+  large because of the bimodal nature of the distribution of the
+  values of the signal, which makes guessing the mean a particularly
+  bad strategy.  The same thing is true of the \svdthree signal.}.  As
+before, the positions of these points on a WPE vs. MASE
+plot---significantly below and to the right of the shaded
+region---should suggest to a practitioner that
+%  there is more structure in this signal than the corresponding
+% method is able to leverage, and that
+a different method might do better.  Indeed, for \svdfive, the LMA
+method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster of
+results that largely within the shaded region on the WPE-MASE plot.
+This is consistent with our second finding: the LMA method can capture
+and reproduce the way in which the \svdfive system processes
 information, but the \naive ~and \arima prediction methods cannot.
@@
 transmits very little information forward in time and provides almost
-no structure for prediction methods to work with.  \alert{This needs
-  to be reworded to reflect the typo, \naive is no longer the
-  winner. MASE is $1.797 \pm 0.010$, rw actually wins MASE$=1.1381 \pm
-  0.011$ Here, the \naive ~method performs best (MASE$=0.951 \pm
-  0.001$), simply because its calculation of the mean filters out the
-  noise in the signal.  Since fitting a hyperplane using least squares
-  should have similar effects, the fact that LMA outperforms \arima
-  ($1.530 \pm 0.021$ vs. $1.837 \pm 0.016$) is somewhat
+no structure for prediction methods to work with.  Here, the
+random-walk predictor is the winner, with a MASE of $1.1381 \pm
+0.011$.  This makes sense; in a fully complex signal, the random-walk
+will always be best, assuming that the training signal accurately
+reflects the overall behavior of the out-of-sample time series.
+\alert{feels like we should justify this a bit}
+% 
+% The \naive ~method works fairly well too: $1.797 \pm 0.010$, simply
+% because its calculation of the mean filters out the noise in the
+% signal.
+% 
+Since fitting a hyperplane using least squares should filter out some
+of the noise in the signal, the fact that LMA outperforms \arima
+($1.530 \pm 0.021$ vs. $1.837 \pm 0.016$) may be somewhat
 counterintuitive.  However, the small amount of predictive structure
-  that is present in this signal is nonlinear
-  (cf.,~\cite{mytkowicz09}), and LMA is designed to capture and
-  exploit that kind of structure.}
+that is present in this signal is nonlinear (cf.,~\cite{mytkowicz09}),
+and LMA is designed to capture and exploit that kind of structure.
+\alert{this is still true, no?  or should we just say ``see, LMA does
+  better as one would expect because the signal is nonlinear?''}
 
@@
 other seven systems in this study.  Though its weighted permutation
-entropy is very high ($0.957 \pm 0.016$), two of the three prediction
-methods do quite well on this signal, yielding MASE scores of $0.827
-\pm 0.076$ (LMA) and $0.714 \pm 0.075$ (\arima).  This pushes the
-corresponding clusters of points in the bottom two images in
-Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
-other seven signals.  The MASE scores of the predictions that were
-produced by the \naive ~method for this system, however, are highly
-inconsistent.  The majority of the blue diamond-shaped points on the
-top plot in Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE
-score of 0.6, which is better than LMA and \arima.  In five of the 15
+entropy is very high ($0.957 \pm 0.016$), three of the four prediction
+methods do quite well on this signal, yielding mean MASE scores of
+0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
+corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
+well above the trend followed by the other seven signals.  \alert{we
+  should add a sentence saying why we think this is so.}  The MASE
+scores of the predictions that were produced by the \naive ~method for
+this system, however, are highly inconsistent.  The majority of the
+blue diamond-shaped points on the top-left plot in
+Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
+0.6, which is better than the other three methods.  In five of the 15
 \svdone trials, however, there were step changes in the signal.  The
@@
   $\approx 2.5$, as well as two points that are beyond the domain of
-  the graph, at MASE $\approx 11.2--14.8$.}, and in turn raising both
+  the graph, at MASE $\approx 11.2-14.8$.}, and in turn raising both
 the mean and variance of this set of trials.
@@
 by the way MASE is calculated.  Recall that MASE scores are scaled
-\emph{relative to a random-walk forecast}, which uses the current
-value as the prediction.  This strategy works very badly on signals
-with frequent, large, rapid transitions.  Consider a signal that
-oscillates from one end of its range to the other at every step.  A
-signal like this will have a low WPE, much like {\tt col\_major}.
+\emph{relative to a random-walk forecast of the training set}.  There
+are two issues here.  First, random-walk prediction works very badly
+on signals with frequent, large, rapid transitions.  Consider a signal
+that oscillates from one end of its range to the other at every step.
+A signal like this will have a low WPE, much like {\tt col\_major}.
 However, a random-walk forecast of this signal will be 180 degrees out
@@
 leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
-clusters for \arima and LMA are above the dashed curve in
-Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
-closeup in Figure~\ref{fig:svdone-ts},
+clusters in Figure~\ref{fig:wpe_vs_mase_all} are above the dashed
+curve.  This time series, which is shown in closeup in
+Figure~\ref{fig:svdone-ts},
 \begin{figure}[htbp]
@@
 effect of the skew in the scaling factor and exacerbates the spread in
-the \svdone MASE values.  For all of these reasons, we view \svdone as
-an outlier and exclude it from the fit calculation of the dashed curve
-in Figure~\ref{fig:wpe_vs_mase_best}.
+the \svdone MASE values.
 
-[[Liz to wordsmith: Even though MASE has weaknesses, we concur with
-    \cite{MASE} that it is an effective way to compare prediction
-    error across time series of different lengths and different
-    scales.  Point to that paper for extensive evaluation and
-    comparison wrt other metrics. We are in the process of repeating
-    our own study with other metrics, other forecast methods, and
-    time-series data from other systems.  Segue to next section.]]
+% For all of these reasons, we view \svdone as an outlier and exclude it
+% from the fit calculation of the dashed curve in
+% Figure~\ref{fig:wpe_vs_mase_best}.
+
+The second effect that can skew MASE scores is nonstationarity.  Since
+this metric is normalized by the error of a random-walk forecast
+\emph{on the training signal}, differences between the test signal and
+training signal can create issues.  This is why the MASE values in
+Table~\ref{tab:errors} are $\neq 1.00$ for every random-walk forecast
+of every time series: the last 10\% of these signals is significantly
+different from the first 90\%.  The deviation from 1.00 will depend on
+the process (whether it has multiple regimes, what those regimes look
+like, and how it switches between them) as well as the experimental
+setup (e.g., sensor precision and data length).  Nonetheless, this
+effect suggests another potential heuristic for time-series
+prediction: if a practitioner calculates a WPE $>$ \alert{(lower
+  bound)} for a time series, and a MASE significantly more than 1,
+\alert{why not just $MASE \neq 1.0$?}  this could be an indication
+that the signal is nonstationary.  We are in the process of exploring
+this idea.
+
+Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
+and~\ref{fig:wpe_vs_mase_best} were determined from a finite set of
+methods and data.  [[But we worked hard to make these experiments
+    representative and comprehensive: forecast methods from simple to
+    sophisticated; time-series data from a system whose behavior spans
+    the dynamical behavior space.]]  While we are cautiously optmistic
+about its generality, more exploration will be required to firm up
+that conclusion.  Our preliminary work along those lines shows that
+data from the Henon map \cite{henon}, the Lorenz system \cite{lorenz},
+and the SFI A data set \cite{sfi-data} all fall within the one-$sigma$
+volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
+and~\ref{fig:wpe_vs_mase_best} region, as do various nonlinear
+transformations of \svdtwo and \svdsix.
+
+Of course, the geometry of the curves and bounds do not hold for other
+error metrics.  The positions of the WPE vs. error points---and the
+reasons to choose a particular function to fit to them, or a
+theoretical bound on the meaning of the results---may change if one
+calculates error in a different way.  The results presented here only
+hold for the MASE error metric.  Even though this metric has its
+weaknesses, we concur with \cite{MASE} that it is an effective way to
+compare prediction error across time-series data of different lengths
+and different scales.  [[Point to that paper for extensive evaluation
+    and comparison wrt other metrics.]]  We are in the process of
+repeating our own study with other metrics, other forecast methods,
+and time-series data from other systems.  [[Segue to next section.]]
 

### 2014-07-30 concl.tex
added sentence to concl about nonstationarity
few minor fixes on results and concl

@@
 \section{ Conclusions \& Future Work }\label{sec:conc}
+
+[[To address the shortcomings mentioned in the previous section, we
+    are in the process of repeating our own study with other metrics,
+    other forecast methods, and time-series data from other systems,
+    as described in the following section.]]
+
+[[Add sentence about ``this curve doesn't hold for nonstationary data,''
+talk about what effects that has, tie back to the points to the right
+of the MASE wall, and opine about how one might know that this was
+happening (and what one might do if it did).]]
 

### 2014-07-30 results.tex
few minor fixes on results and concl

@@
 % for the second conjecture above.)
-In the \svdfive results, this match between model and task manifests
-quite clearly.  The MASE scores of the \naive ~and \arima predictions
-for this system are $20.870 \pm 0.192$ and $2.370 \pm 0.051$,
-respectively: that is, 20.87 and 2.37 times worse than a simple random
-walk forecast of the same signals\footnote{The \naive ~MASE score is
-  large because of the bimodal nature of the distribution of the
-  values of the signal, which makes guessing the mean a particularly
-  bad strategy.  The same thing is true of the \svdthree signal.}.  As
-before, the positions of these points on a WPE vs. MASE
-plot---significantly below and to the right of the shaded
-region---should suggest to a practitioner that
+The MASE scores of the \naive ~and \arima predictions for this system
+are $20.870 \pm 0.192$ and $2.370 \pm 0.051$, respectively: that is,
+20.87 and 2.37 times worse than a simple random walk forecast of the
+same signals\footnote{The \naive ~MASE score is large because of the
+  bimodal nature of the distribution of the values of the signal,
+  which makes guessing the mean a particularly bad strategy.  The same
+  thing is true of the \svdthree signal.}.  As before, the positions
+of these points on a WPE vs. MASE plot---significantly below and to
+the right of the shaded region---should suggest to a practitioner that
 %  there is more structure in this signal than the corresponding
@@
 training signal can create issues.  This is why the MASE values in
-Table~\ref{tab:errors} are $\neq 1.00$ for every random-walk forecast
-of every time series: the last 10\% of these signals is significantly
-different from the first 90\%.  The deviation from 1.00 will depend on
-the process (whether it has multiple regimes, what those regimes look
-like, and how it switches between them) as well as the experimental
-setup (e.g., sensor precision and data length).  Nonetheless, this
-effect suggests another potential heuristic for time-series
-prediction: if a practitioner calculates a WPE $>$ \alert{(lower
-  bound)} for a time series, and a MASE significantly more than 1,
-\alert{why not just $MASE \neq 1.0$?}  this could be an indication
-that the signal is nonstationary.  We are in the process of exploring
-this idea.
+Table~\ref{tab:error} are not identically one for every random-walk
+forecast of every time series: the last 10\% of these signals is
+significantly different from the first 90\%.  The deviation from 1.00
+will depend on the process---whether it has multiple regimes, what
+those regimes look like, and how it switches between them---as well as
+the experimental setup (e.g., sensor precision and data length).
+Nonetheless, this effect suggests another potential heuristic for
+time-series prediction: if a practitioner calculates a WPE $>$
+\alert{(lower bound)} for a time series, and a MASE significantly more
+than 1, \alert{why not just $MASE \neq 1.0$?}  this could be an
+indication that the signal is nonstationary.  We are in the process of
+exploring this idea.
 
@@
 data from the Henon map \cite{henon}, the Lorenz system \cite{lorenz},
-and the SFI A data set \cite{sfi-data} all fall within the one-$sigma$
+and the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
 volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
-and~\ref{fig:wpe_vs_mase_best} region, as do various nonlinear
+and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
 transformations of \svdtwo and \svdsix.
@@
 and different scales.  [[Point to that paper for extensive evaluation
-    and comparison wrt other metrics.]]  We are in the process of
-repeating our own study with other metrics, other forecast methods,
-and time-series data from other systems.  [[Segue to next section.]]
+    and comparison wrt other metrics.]]
 

### 2014-07-30 model.tex
more cleanup

@@
 the last 10\% of \svdthree, for instance---the out-of-sample
-signal---was easier, on the average, for RW to predict than the first
-90\%. \alert{[[Liz to do: add pointer to Josh's paragraph for deeper
-      analysis]]}.
+signal---was more amenable to random-walk prediction, on the average,
+than the first 90\%.  See Section~\ref{sec:results} for a deeper
+discussion of this effect.
 
@@
 it facilitates comparisons across signals of different types and
-lengths, but its particular pathologies must be taken into account, as
-discussed at more length in Section~\ref{sec:results}.
+lengths, but its particular pathologies must be taken into account
+when analyzing the results, as discussed at more length in
+Section~\ref{sec:results}.
 

### 2014-07-30 results.tex
more cleanup

@@
 discussed at more length below.
-%[[make sure that's true]]
 
@@
 hand, reflects this method's ability to capture and exploit the
-structure that is present in this signal and suggests to the
-practitioner that [[she really can't do any better]].  The following
-discussion addresses the details that underlie these claims.
+structure that is present in this signal.  WPE vs. MASE values like
+this, which fall in the shaded region, should suggest to the
+practitioner that the prediction method is well-suited to the task.
+The following discussion lays out the details that underlie these
+claims.
 % 
@@
 \alert{this is still true, no?  or should we just say ``see, LMA does
-  better as one would expect because the signal is nonlinear?''}
+  better as one would expect because the signal is nonlinear?''  also,
+  we need to talk about the fact that the best of these points, and of
+  \svdtwo, is past the MASE wall and not in the shaded region.}
 
@@
 and~\ref{fig:wpe_vs_mase_best} were determined from a finite set of
-methods and data.  [[But we worked hard to make these experiments
-    representative and comprehensive: forecast methods from simple to
-    sophisticated; time-series data from a system whose behavior spans
-    the dynamical behavior space.]]  While we are cautiously optmistic
-about its generality, more exploration will be required to firm up
-that conclusion.  Our preliminary work along those lines shows that
-data from the Henon map \cite{henon}, the Lorenz system \cite{lorenz},
-and the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
+methods and data.  We put a lot of thought and effort into making
+these experiments upon which these results are based both
+representative and comprehensive.  The forecast methods involved range
+from the simple to the sophisticated; the time-series data was sampled
+from a system whose behavior spans the dynamical behavior space.
+While we are cautiously optimistic about the generality of our
+conclusions, more exploration will be required before we can make any
+such claim.  Our preliminary work along those lines shows that data
+from the Henon map \cite{henon}, the Lorenz system \cite{lorenz}, and
+the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
 volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
@@
 
-Of course, the geometry of the curves and bounds do not hold for other
-error metrics.  The positions of the WPE vs. error points---and the
-reasons to choose a particular function to fit to them, or a
-theoretical bound on the meaning of the results---may change if one
-calculates error in a different way.  The results presented here only
-hold for the MASE error metric.  Even though this metric has its
-weaknesses, we concur with \cite{MASE} that it is an effective way to
-compare prediction error across time-series data of different lengths
-and different scales.  [[Point to that paper for extensive evaluation
-    and comparison wrt other metrics.]]
+Of course, the geometry of these curves and bounds do not necessarily
+extend to other error metrics.  The positions of the WPE vs. error
+points---and the reasons to choose a particular function to fit to
+them or impute theoretical bounds on the meaning of the results---may
+change if one calculates error using a procedure other than the one
+described in Section~\ref{sec:accuracy}.  The MASE error metric, as
+discussed above, has its weaknesses.  Even so, we concur with
+\cite{MASE} that it is an effective way to compare prediction error
+across time-series data of different lengths and different scales, and
+we are convinced by the extensive evaluations that are offered in that
+paper, as well as the comparisons of MASE to other metrics, that its
+strengths far outweigh its weaknesses.
 

### 2014-07-30 results.tex
add random walk to list of predictors

@@
 Working from the first 90\% of each signal, we generated a prediction
-of the last 10\% using the \naive, \arima, and LMA prediction methods,
+of the last 10\% using the \naive, random walk, \arima, and LMA prediction methods,
 as described in Section~\ref{sec:accuracy}, then calculated the MASE

### 2014-07-30 model.tex
working

@@
 use the four methods outlined above to generate predictions of all 120
-\alert{we had 360 there.  seems like it should be 8 times 15 which is
-  120.}  traces described in Section~\ref{sec:methods}, then calculate
-the error of the predictions with respect to the true continuations.
+traces described in Section~\ref{sec:methods}, then calculate the
+error of the predictions with respect to the true continuations.
 Specifically, we split each time series into two pieces: the first

### 2014-07-30 results.tex
working

@@
 these reasons, we chose to fit a function of the form $y = a \log(b x
-+ 1)$ to these points, with $y =$ WPE and $x=$ MASE.  \alert{can we
-  cite a figure of merit for this fit?  we also need to say what $a$
-  and $b$ were.}  The solid curve in the figure shows this fit; the
-dashed curves show the standard deviation of this model in its
-parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
-deviation on each of the two parameters.  Points that fall within this
-deviation volume \alert{(light grey)} correspond to predictions that
-are comparable to the best ones found in this study; points that fall
-\emph{above} that volume \alert{(dark grey)} are better still.  We
-chose to truncate the shaded region because of a subtle point
-regarding the MASE of an ideal predictor, which should not be larger
-than 1 unless the training and test signals are different.  This is
-discussed at more length below.
++ 1)$ to these points, with $y =$ WPE and $x=$ MASE\footnote{The
+  specific values of the coefficiencts are $a=7.97 \times 10^{-2}$ and
+  $b=1.52 \times 10^3$.}.  The solid curve in the figure shows this
+fit; the dashed curves show the standard deviation of this model in
+its parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one
+standard deviation on each of the two parameters.  Points that fall
+within this deviation volume (light grey) correspond to predictions
+that are comparable to the best ones found in this study; points that
+fall \emph{above} that volume (dark grey) are better still.  We chose
+to truncate the shaded region because of a subtle point regarding the
+MASE of an ideal predictor, which should not be larger than 1 unless
+the training and test signals are different.  This is discussed at
+more length below.
 
@@
 the level of predictive structure that the low WPE value suggests is
-present in the signal.  The random-walk MASE scores for \col are all
-$\approx 1$---as one would expect---pushing those points well below
-the shaded region.  Clearly the stationarity assumption on which this
-method is based does not hold for this signal.
+present in the signal.  The MASE scores of random-walk predictions of
+\col are all $\approx 1$---as one would expect---pushing those points
+well below the shaded region.  Clearly the stationarity assumption on
+which this method is based does not hold for this signal.
 
@@
 no structure for prediction methods to work with.  Here, the
-random-walk predictor is the winner, with a MASE of $1.1381 \pm
-0.011$.  This makes sense; in a fully complex signal, the random-walk
-will always be best, assuming that the training signal accurately
-reflects the overall behavior of the out-of-sample time series.
-\alert{feels like we should justify this a bit}
+random-walk predictor is the best of the methods used here.  This
+makes sense; in a fully complex signal, where there is no predictive
+structure to utilize, methods that depend on exploiting that
+structure---like ARIMA and LMA---will be outperformed by methods that
+do not rely on that structure.
 % 
-% The \naive ~method works fairly well too: $1.797 \pm 0.010$, simply
-% because its calculation of the mean filters out the noise in the
-% signal.
+% in a fully complex signal, the MASE of a random-walk forecast will
+% always be $\approx 1$, assuming that the training signal accurately
+% reflects the overall behavior of the out-of-sample time series
 %
@@
 and LMA is designed to capture and exploit that kind of structure.
-\alert{this is still true, no?  or should we just say ``see, LMA does
-  better as one would expect because the signal is nonlinear?''  also,
-  we need to talk about the fact that the best of these points, and of
-  \svdtwo, is past the MASE wall and not in the shaded region.}
+Note that all four \gcc clusters are outside the shaded region; in the
+case of the random-walk prediction, for instance, the MASE value is
+$1.1381 \pm 0.011$.  This is due to nonstationarity in the signal: in
+particular, differences between the training and test signals.
+\svdtwo fits the same patterns, for the same reasons---and visibly so,
+in the red segment of Figure~\ref{fig:svd-ts-colored}, where the
+period and amplitude of the oscillations are decreasing.
 

### 2014-07-30 results.tex
end of weds mtg

@@
 points in Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.
-\alert{This paragraph needs more, in response to the ref's gritch
-  about fitting: the stuff about \arima being an automated recipe that
-  can give bad results, and the fact that our technique can tell when
-  it does.  And we should call his attention to it in that part of the
-  response letter.}
+\alert{Josh to do: this paragraph needs more, in response to the ref's
+  gritch about fitting: the stuff about \arima being an automated
+  recipe that can give bad results, and the fact that our technique
+  can tell when it does.  Look at the response letter for context and
+  ammunition---and also put something in there that calls his
+  attention to this paragraph.}
 
@@
 and LMA is designed to capture and exploit that kind of structure.
-Note that all four \gcc clusters are outside the shaded region; in the
-case of the random-walk prediction, for instance, the MASE value is
-$1.1381 \pm 0.011$.  This is due to nonstationarity in the signal: in
-particular, differences between the training and test signals.
-\svdtwo fits the same patterns, for the same reasons---and visibly so,
-in the red segment of Figure~\ref{fig:svd-ts-colored}, where the
-period and amplitude of the oscillations are decreasing.
+Note that all four \gcc clusters in Figure~\ref{fig:wpe_vs_mase_all}
+are outside the shaded region; in the case of the random-walk
+prediction, for instance, the MASE value is $1.1381 \pm 0.011$.  This
+is due to nonstationarity in the signal: in particular, differences
+between the training and test signals.  \svdtwo fits the same
+patterns, for the same reasons---and visibly so, in the red segment of
+Figure~\ref{fig:svd-ts-colored}, where the period and amplitude of the
+oscillations are decreasing.
 
@@
 corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
-well above the trend followed by the other seven signals.  \alert{we
-  should add a sentence saying why we think this is so.}  The MASE
-scores of the predictions that were produced by the \naive ~method for
-this system, however, are highly inconsistent.  The majority of the
-blue diamond-shaped points on the top-left plot in
+well above the trend followed by the other seven signals.  The reasons
+for this are discussed in the following paragraph.  The MASE scores of
+the predictions that were produced by the \naive ~method for this
+system, however, are highly inconsistent.  The majority of the blue
+diamond-shaped points on the top-left plot in
 Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
@@
 the experimental setup (e.g., sensor precision and data length).
-Nonetheless, this effect suggests another potential heuristic for
-time-series prediction: if a practitioner calculates a WPE $>$
-\alert{(lower bound)} for a time series, and a MASE significantly more
-than 1, \alert{why not just $MASE \neq 1.0$?}  this could be an
-indication that the signal is nonstationary.  We are in the process of
-exploring this idea.
+Nonetheless, this effect suggests another potentially useful
+heuristic: if the MASE of a random-walk prediction of a time series is
+significantly different from 1, this could be an indication that the
+signal is nonstationary.  \alert{Should we point back to the obvious
+  MASE wall in the figure to add weight to this?}  We are in the
+process of exploring this idea.
 
 Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
-and~\ref{fig:wpe_vs_mase_best} were determined from a finite set of
+and~\ref{fig:wpe_vs_mase_all} were determined from a finite set of
 methods and data.  We put a lot of thought and effort into making

### 2014-07-31 response.tex
fixing table and adding paragraph

@@
 points from the other time series fall within a narrow band around the
-fit.  Note that this fit ($y = a \log(b x + 1) + c x^16$) is different than in
+fit.  Note that this fit ($y = a \log(b x + 1)$) is different than in
 the submitted version.  A careful examination of the data, prompted by
@@
 included them, however, because they are used (sometimes to great
-success) in practice and because that made for a more comprehensive
-exploration.
+success) in practice and because that made for a more comprehensive exploration.
 
-The ``AutoARIMA'' method that was used to build all of the ARIMA
+The {\tt auto.arima} method that was used to build all of the ARIMA
 models in this paper---another ``off the shelf'' method in the
 modeling and prediction communities---follows the procedure outlined
-in Section \alert{[[blah]]} to choose values for the model parameters.
+in Section V B to choose values for the model parameters.
 Sometimes the choices made by that procedure (e.g., setting a

### 2014-07-31 model.tex
fixing table and adding paragraph

@@
 columns in Table~\ref{tab:error}.
+%\begin{table*}
+%\caption{Mean absolute scaled error (MASE) scores and weighted
+%  permutation entropies for all eight processes studied in this paper.
+%  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
+%  \alert{please reorder the columns from left to right to match the
+%    order in which the methods were introduced, rather than in the
+%    reverse of that order}}
+%  \begin{center}
+%  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
+%  \hline\hline Signal & LMA MASE & \arima MASE & na\"{i}ve MASE & RW MASE & WPE \\
+%  \hline
+%
+%  \col       & $0.050 \pm 0.002$ & $0.599 \pm 0.211$ & $0.571 \pm 0.002$  & $1.001 \pm 0.002$ & $0.513 \pm 0.003$ \\
+%
+%  \gcc       & $1.530 \pm 0.021$ & $1.837 \pm 0.016$ & $1.797 \pm 0.010$  & $1.138 \pm 0.011$ & $0.943 \pm 0.001$ \\
+%
+%  \svdone    & $0.827 \pm 0.076$ & $0.714 \pm 0.075$ & $2.676 \pm 4.328$  & $0.933 \pm 0.095$ & $0.957 \pm 0.016$ \\
+%
+%  \svdtwo    & $1.279 \pm 0.020$ & $2.163 \pm 0.027$ & $3.054 \pm 0.040$  & $1.125 \pm 0.012$ & $0.846 \pm 0.004$ \\
+%
+%  \svdthree  & $0.619 \pm 0.021$ & $0.713 \pm 0.010$ & $31.386 \pm 0.282$ & $0.707 \pm 0.009$ & $0.716 \pm 0.006$ \\
+%
+%  \svdfour   & $0.779 \pm 0.036$ & $0.979 \pm 0.032$ & $2.661 \pm 0.074$  & $1.034 \pm 0.035$ & $0.825 \pm 0.008$ \\
+%
+%  \svdfive   & $0.718 \pm 0.048$ & $2.370 \pm 0.051$ & $20.870 \pm 0.192$ & $1.001 \pm 0.047$ & $0.678 \pm 0.007$ \\
+%
+%  \svdsix    & $0.739 \pm 0.068$ & $1.438 \pm 0.061$ & $2.197 \pm 0.083$  & $1.060 \pm 0.055$ & $0.748 \pm 0.011$ \\
+%
+%  \hline\hline
+%  \end{tabular*}
+%  \end{center}
+% \label{tab:error}
+%  \end{table*}%
+%  
  \begin{table*}
@@
   \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
-  \hline\hline Signal & LMA MASE & \arima MASE & na\"{i}ve MASE & RW MASE & WPE \\
+  \hline\hline 
+Signal & RW MASE & na\"{i}ve MASE & \arima MASE & LMA MASE & WPE \\
 \hline
-
-  \col       & $0.050 \pm 0.002$ & $0.599 \pm 0.211$ & $0.571 \pm 0.002$  & $1.001 \pm 0.002$ & $0.513 \pm 0.003$ \\
-
-  \gcc       & $1.530 \pm 0.021$ & $1.837 \pm 0.016$ & $1.797 \pm 0.010$  & $1.138 \pm 0.011$ & $0.943 \pm 0.001$ \\
-
-  \svdone    & $0.827 \pm 0.076$ & $0.714 \pm 0.075$ & $2.676 \pm 4.328$  & $0.933 \pm 0.095$ & $0.957 \pm 0.016$ \\
-
-  \svdtwo    & $1.279 \pm 0.020$ & $2.163 \pm 0.027$ & $3.054 \pm 0.040$  & $1.125 \pm 0.012$ & $0.846 \pm 0.004$ \\
-
-  \svdthree  & $0.619 \pm 0.021$ & $0.713 \pm 0.010$ & $31.386 \pm 0.282$ & $0.707 \pm 0.009$ & $0.716 \pm 0.006$ \\
-
-  \svdfour   & $0.779 \pm 0.036$ & $0.979 \pm 0.032$ & $2.661 \pm 0.074$  & $1.034 \pm 0.035$ & $0.825 \pm 0.008$ \\
-
-  \svdfive   & $0.718 \pm 0.048$ & $2.370 \pm 0.051$ & $20.870 \pm 0.192$ & $1.001 \pm 0.047$ & $0.678 \pm 0.007$ \\
-
-  \svdsix    & $0.739 \pm 0.068$ & $1.438 \pm 0.061$ & $2.197 \pm 0.083$  & $1.060 \pm 0.055$ & $0.748 \pm 0.011$ \\
-
+  \col       & $1.001 \pm 0.002$ & $0.571 \pm 0.002$  & $0.599 \pm 0.211$ & $0.050 \pm 0.002$ & $0.513 \pm 0.003$ \\
+  \gcc       & $1.138 \pm 0.011$ & $1.797 \pm 0.010$  & $1.837 \pm 0.016$ & $1.530 \pm 0.021$ & $0.943 \pm 0.001$ \\
+  \svdone    & $0.933 \pm 0.095$ & $2.676 \pm 4.328$  & $0.714 \pm 0.075$ & $0.827 \pm 0.076$ & $0.957 \pm 0.016$ \\
+  \svdtwo    & $1.125 \pm 0.012$ & $3.054 \pm 0.040$  & $2.163 \pm 0.027$ & $1.279 \pm 0.020$ & $0.846 \pm 0.004$ \\
+  \svdthree  & $0.707 \pm 0.009$ & $31.386 \pm 0.282$ & $0.713 \pm 0.010$ & $0.619 \pm 0.021$ & $0.716 \pm 0.006$ \\
+  \svdfour   & $1.034 \pm 0.035$ & $2.661 \pm 0.074$  & $0.979 \pm 0.032$ & $0.779 \pm 0.036$ & $0.825 \pm 0.008$ \\
+  \svdfive   & $1.001 \pm 0.047$ & $20.870 \pm 0.192$ & $2.370 \pm 0.051$ & $0.718 \pm 0.048$ & $0.678 \pm 0.007$ \\
+  \svdsix    & $1.060 \pm 0.055$ & $2.197 \pm 0.083$  & $1.438 \pm 0.061$ & $0.739 \pm 0.068$ & $0.748 \pm 0.011$ \\  
     \hline\hline
@@
   \end{table*}%
+
 In view of the discussion at the end of the previous paragraph, the

### 2014-07-31 pecomplexity.tex
bunch of formatting issues
fixing autoarima command and formatting bugs
working through
fixing table and adding paragraph

@@
 %\usepackage{caption}
-\usepackage{subcaption}
+%\usepackage{subcaption}
+\usepackage[caption=false]{subfig}
 \usepackage{microtype}
@@
 \newcommand{\svdsix}{{\tt dgesdd$_6$}\xspace}
+\newcommand{\arima}{{\tt auto.arima}\xspace}
+
 \newcommand{\naive}{na\"ive}
@@
 %\\
-Quantifying Time-Series Predictability through Structural Complexity
+\alert{Quantifying Time-Series Predictability through Structural Complexity}
 %\\
@@
 \emph{weighted permutation entropy} is an effective way to estimate
-that complexity.  To validate these conjectures, we study 120
+that redundancy.  To validate these conjectures, we study 120
 different time-series data sets.  For each time series, we construct

### 2014-07-31 results.tex
fixing table and adding paragraph

@@
 
+The discussion in the previous paragraph brings to light an interesting feature of our proposed heuristic. The \arima method that was used to build all of the ARIMA
+models in this paper, follows the procedure outlined
+in Section V B to choose values for the model parameters.
+Sometimes the choices made by that procedure produce a sub-optimal predictor, however there is nothing to say whether a different model from the ARIMA class would work better.
+Inappropriate MASE scores (compared to
+what the WPE value might suggest) may suggest an inappropriateness in the order selection and parameter estimation procedure.
+During the fitting process of \arima many entire subclasses of ARIMA models are eliminated based on various tests, such as the KPSS test.
+If these tests give erroneous results then entire subclasses of ARIMA models are never even considered, even if they would have been more likely.
+Our heuristic may be able to flag when this has occured.
+By examining the MASE vs WPE scores of \col, for example, we were able to see that, while all signal complexities were the same some of the signals were being forecast much worse by \arima.
+This suggested to us that the signals that were being forecast badly could be forecast better, sure enough by adjusting a single parameter in the fitting process we received on average predictions that was twice as good for these signals.
+
 The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt

### 2014-07-31 results.tex
editing the ARIMA fit paragraph

--- 2014-07-30 results.tex
@@
 points in Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.
-\alert{Josh to do: this paragraph needs more, in response to the ref's
-  gritch about fitting: the stuff about \arima being an automated
-  recipe that can give bad results, and the fact that our technique
-  can tell when it does.  Look at the response letter for context and
-  ammunition---and also put something in there that calls his
-  attention to this paragraph.}
+%\alert{Josh to do: this paragraph needs more, in response to the %ref's
+%  gritch about fitting: the stuff about \arima being an automated
+%  recipe that can give bad results, and the fact that our %technique
+%  can tell when it does.  Look at the response letter for context %and
+%  ammunition---and also put something in there that calls his
+%  attention to this paragraph.}
+
+The discussion in the previous paragraph brings to light an interesting feature of our proposed heuristic. The \arima algorithm that was used to build all of the ARIMA
+models in this paper, follows the procedure outlined
+in Section V B to choose values for the model parameters.  
+As discussed previously, this algorithm employs sophisticated tests, such as the KPSS test, to intelligently traverse and reduce model space and then uses well-known criterion for choosing the most-likely of the remaining models.
+However, it is quite possible that this procedure can choose a sub-optimal fit.
+Specifically, if the initial space of models being searched is not broad enough, or if one of the preliminary tests gives an erroneous result, then the best ARIMA model may be excluded or eliminated before likelihood is even examined.
+This circumstance is hard to detect.
+Maximal-likelihood  tests, such as AIC used in \arima, can only conclude which of the models provided is ``most likely"---but cannot conclude that \emph{none} of the models are likely.
+In contrast,
+%Sometimes the choices made by that procedure produce a sub-optimal predictor, however there is nothing to say whether a different model from the ARIMA class would work better. 
+inappropriate MASE scores (compared to
+what the WPE value might suggest) may suggest an inappropriateness in the order selection and parameter estimation procedure, which would flag to a practitioner to reevaluate the appropriateness of fit.
+For example,
+%During the fitting process of \arima many subclasses of ARIMA models are eliminated based on various tests, such as the KPSS test.
+If these tests give erroneous results then entire subclasses of ARIMA models are never even considered.
+Our heuristic may be able to flag when this has occurred.
+by examining the \arima MASE scores vs.
+WPE values of \col, we were able to see that, while all signal complexities were the same, some of the signals had much higher MASE values.
+This suggested to us inappropriate parameter selection for these time series.
+We then went back and manually assisted the automated fitting procedure, resulting in models with consistent MASE vs WPE score, and improved forecasts.
 

### 2014-07-31 results.tex
swap two plots, update text to reflect

@@
   exception of \svdone, \svdthree, and \svdfive are omitted from the
-  top-left plot for scale reasons, as described in the text.
+  top-right plot for scale reasons, as described in the text.
 %
@@
 followed by the random-walk prediction method, then \arima and \naive.
-When a practioner calculates an WPE vs. MASE value that is outside the
+When a practitioner calculates an WPE vs. MASE value that is outside the
 shaded region, that should suggest to her that the prediction method
@@
 poor MASE scores---is obvious in the geometry of the two left-hand
-images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
+and the upper-right hand images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
 are far to the right of and/or below the dashed curve.  Again, this
@@
 system, however, are highly inconsistent.  The majority of the blue
-diamond-shaped points on the top-left plot in
+diamond-shaped points on the top-right plot in
 Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
@@
 and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
-transformations of \svdtwo and \svdsix.
+transformations of \svdtwo, \svdfive and \svdsix.
 

### 2014-07-31 results.tex
swap two words

@@
 Working from the first 90\% of each signal, we generated a prediction
-of the last 10\% using the \naive, random walk, \arima, and LMA prediction methods,
+of the last 10\% using the random-walk, \naive, \arima, and LMA prediction methods,
 as described in Section~\ref{sec:accuracy}, then calculated the MASE

### 2014-08-01 response.tex
editing the ARIMA fit paragraph
add bit about fit change, fix henon

@@
 SFI prediction competition, for instance.  (It also holds for various
-paradigmatic systems, such as the Henon map and the Lorenz system, as
+paradigmatic systems, such as the H\'{e}non map and the Lorenz system, as
 discussed in a later section of this response.)  The image below is an
@@
 relationship posited in our paper does indeed hold for paradigmatic
-systems like the Henon map and the Lorenz equations.  We had not
+systems like the H\'{e}non map and the Lorenz equations.  We had not
 included these examples in our paper because they simply confirmed
@@
 chosen parameter values were suboptimal---or even that no member of
-the ARIMA family will work and one should try a nonlinear model.
+the ARIMA family will work and one should try a nonlinear model. A discussion to this effect has been added to Section
+VI in hopes of clarifying this completely valid concern and to illustrate that this is a feature of our proposed heuristic.
 
@@
 for large [$\ell$], although this is an interesting theoretical
-problem.  ``For practical purposes, we recommend 3\dots7...''  The
+problem. For practical purposes, we recommend 3\dots7...''  The
 corresponding discussion in [Cao {\sl et al.}, {\sl PRE}, 2004] reads
@@
 dynamical behaviors.  For all of these examples---as well as for the
-39 additional ones\footnote{Henon, Lorenz, SFI A, random walk,
+39 additional ones\footnote{H\'{e}non, Lorenz, SFI A, random walk,
   nonlinear transformations of the 30 [[svdtwo]] and [[svdsix]]
@@
 
-The fit of the regression line was not a function of the small number
-of points; when we included all 105 WPE-MASE pairs, rather than just a
-representative from each program, the slope of the fit changed by 2\%
-and the y intercept by 1\%.
-
 We have revised the Figure, and the associated text, to include all
-105 points.  We have also moved from a linear fit to a log fit in this
+120 points.  We have also moved from a linear fit to a log fit in this
 curve to better match the theoretical relationship between MASE and
@@
 of the paper is not related to the crispness of the fit.
+
+Furthermore, we have recomputed the fit using both the 120 computer performance
+data time-series as well as the 35 nonlinear transforms and the H\'{e}non, Lorenz,
+random walk, and SFI Data A time-series. Not only does the fit line hardly
+change, but the error bounds are contained completely within those provided by
+just the computer performance data. This can be seen in the following figure
+where the blue fit and bounds are those from the computer performance traces and
+the magenta ones are those including the additional 39 time-series.
+
+\begin{center}
+    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras_with_nonlinear_all_points}
+\end{center}
 

### 2014-08-01 model.tex
removing an alert and fixing a HEAD issue
friday round I

@@
   LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
-  \alert{please reorder the columns from left to right to match the
-    order in which the methods were introduced, rather than in the
-    reverse of that order}}
+  }
   \begin{center}
@@
 In view of the discussion at the end of the previous paragraph, the
-fact that the values in the second-from-\alert{left} column are not
+fact that the values in the second-from-left column are not
 identically equal to 1.00 may be somewhat surprising.  This can happen

### 2014-08-01 results.tex
friday round I

@@
 
-The discussion in the previous paragraph brings to light an interesting feature of our proposed heuristic. The \arima algorithm that was used to build all of the ARIMA
-models in this paper, follows the procedure outlined
-in Section V B to choose values for the model parameters.
-As discussed previously, this algorithm employs sophisticated tests, such as the KPSS test, to intelligently traverse and reduce model space and then uses well-known criterion for choosing the most-likely of the remaining models.
-However, it is quite possible that this procedure can choose a sub-optimal fit.
-Specifically, if the initial space of models being searched is not broad enough, or if one of the preliminary tests gives an erroneous result, then the best ARIMA model may be excluded or eliminated before likelihood is even examined.
-This circumstance is hard to detect.
-Maximal-likelihood  tests, such as AIC used in \arima, can only conclude which of the models provided is ``most likely"---but cannot conclude that \emph{none} of the models are likely.
-In contrast,
-%Sometimes the choices made by that procedure produce a sub-optimal predictor, however there is nothing to say whether a different model from the ARIMA class would work better.
-inappropriate MASE scores (compared to
-what the WPE value might suggest) may suggest an inappropriateness in the order selection and parameter estimation procedure, which would flag to a practitioner to reevaluate the appropriateness of fit.
-For example,
-%During the fitting process of \arima many subclasses of ARIMA models are eliminated based on various tests, such as the KPSS test.
+The discussion in the previous paragraph highlights the second finding
+of this paper: the ability of our proposed heuristic to flag
+inappropriate models.  \arima is an automated, mechanical procedure
+for choosing parameters for an ARIMA model of a given data set.  While
+the tests and criteria employed by this algorithm are sophisticated
+(see Section~\ref{sec:arima}), the results can still be
+sub-optimal---if the initial space of models being searched is not
+broad enough, for instance, or if one of the preliminary tests gives
+an erroneous result.  \arima \emph{always} returns a model, and it can
+be very hard to detect when that model is bad.  Our results suggest a
+way to do so: if the MASE score of an \arima model is out of line with
+the WPE value, that can be an indication of inappropriateness in the
+order selection and parameter estimation procedure.  By examining the
+\arima MASE scores vs. WPE values of \col, we were able to see that,
+while all signal complexities were the same, some of the signals had
+much higher MASE values. This suggested to us inappropriate parameter
+selection for these time series. We then went back and manually
+assisted the automated fitting procedure, resulting in models with
+consistent MASE vs WPE score, and improved forecasts.
+
+% The discussion in the previous paragraph brings to light an
+% interesting feature of our proposed heuristic. The \arima algorithm
+% that was used to build all of the ARIMA models in this paper follows
+% the procedure outlined in Section V B to choose values for the model
+% parameters.  As discussed previously, this algorithm employs
+% sophisticated tests, such as the KPSS test, to intelligently traverse
+% and reduce model space and then uses well-known criterion for choosing
+% the most-likely of the remaining models. However, it is quite possible
+% that this procedure can choose a sub-optimal fit. Specifically, if the
+% initial space of models being searched is not broad enough, or if one
+% of the preliminary tests gives an erroneous result, then the best
+% ARIMA model may be excluded or eliminated before likelihood is even
+% examined. This circumstance is hard to detect. Maximal-likelihood
+% tests, such as AIC used in \arima, can only conclude which of the
+% models provided is ``most likely"---but cannot conclude that
+% \emph{none} of the models are likely. In contrast,
+% %Sometimes the choices made by that procedure produce a sub-optimal predictor, however there is nothing to say whether a different model from the ARIMA class would work better.
+% inappropriate MASE scores (compared to what the WPE value might
+% suggest) may suggest an inappropriateness in the order selection and
+% parameter estimation procedure, which would flag to a practitioner to
+% reevaluate the appropriateness of fit. For example,
+% %During the fitting process of \arima many subclasses of ARIMA models are eliminated based on various tests, such as the KPSS test.
 If these tests give erroneous results then entire subclasses of ARIMA models are never even considered.
 Our heuristic may be able to flag when this has occurred.
-by examining the \arima MASE scores vs.
-WPE values of \col, we were able to see that, while all signal complexities were the same, some of the signals had much higher MASE values.
-This suggested to us inappropriate parameter selection for these time series.
-We then went back and manually assisted the automated fitting procedure, resulting in models with consistent MASE vs WPE score, and improved forecasts.
+% by examining the \arima MASE scores vs. WPE values of \col, we were
+% able to see that, while all signal complexities were the same, some of
+% the signals had much higher MASE values. This suggested to us
+% inappropriate parameter selection for these time series. We then went
+% back and manually assisted the automated fitting procedure, resulting
+% in models with consistent MASE vs WPE score, and improved forecasts.
 
@@
 those regimes look like, and how it switches between them---as well as
-the experimental setup (e.g., sensor precision and data length).
-Nonetheless, this effect suggests another potentially useful
-heuristic: if the MASE of a random-walk prediction of a time series is
-significantly different from 1, this could be an indication that the
-signal is nonstationary.  \alert{Should we point back to the obvious
-  MASE wall in the figure to add weight to this?}  We are in the
-process of exploring this idea.
+the experimental setup (e.g., sensor precision and data length).  For
+the processes studied here, these effects do not cause the MASE values
+to exceed 1.15, but pathological situations (e.g., a huge switch in
+scale right at the training/test signal boundary) could produce higher
+values.  This suggests another potentially useful heuristic: if the
+MASE of a random-walk prediction of a time series is significantly
+different from 1, this could be an indication that the signal is
+nonstationary.  We are in the process of exploring this idea.
 
@@
 from the Henon map \cite{henon}, the Lorenz system \cite{lorenz}, and
-the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
-volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
+the SFI A data set \cite{weigend-book,sfi-data} all fall within the
+one-$\sigma$ volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
 and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear

### 2014-08-01 results.tex
fixed end of JG paragraph

@@
 be very hard to detect when that model is bad.  Our results suggest a
-way to do so: if the MASE score of an \arima model is out of line with
-the WPE value, that can be an indication of inappropriateness in the
-order selection and parameter estimation procedure.  By examining the
-\arima MASE scores vs. WPE values of \col, we were able to see that,
-while all signal complexities were the same, some of the signals had
-much higher MASE values. This suggested to us inappropriate parameter
-selection for these time series. We then went back and manually
-assisted the automated fitting procedure, resulting in models with
-consistent MASE vs WPE score, and improved forecasts.
+way to do so: if the MASE score of an auto-fitted model like \arima is
+out of line with the WPE value of the data, that can be an indication
+of inappropriateness in the order selection and parameter estimation
+procedure.  A close examination of the \arima MASE scores and WPE
+values of the 15 \col traces, for instance, brought out such a
+situation: all of these signals had the same complexity, but some (the
+outliers in the blue cluster of $+$ points in the bottom-left panel of
+Figure~\ref{fig:what}) had much higher \arima MASE values, indicating
+a failure in the \arima process.  We then went back and manually
+assisted the automated fitting procedure---forcing inclusion of a term
+that is appropriate for a nonstationary signal.  In all cases, this
+produced ARIMA models with MASE scores in line with the others in that
+cluster.
 

### 2014-08-01 response.tex
fix numbers in response

@@
 Furthermore, we have recomputed the fit using both the 120 computer performance
-data time-series as well as the 35 nonlinear transforms and the H\'{e}non, Lorenz,
+data time-series as well as the 45 nonlinear transforms and the H\'{e}non, Lorenz,
 random walk, and SFI Data A time-series. Not only does the fit line hardly
@@
 where the blue fit and bounds are those from the computer performance traces and
-the magenta ones are those including the additional 39 time-series.
+the magenta ones are those including the additional 49 time-series.
 

### 2014-08-01 response.tex
some edits on response

--- 2014-08-01 response.tex
@@
 of discussion of this cluster of issues in [Hyndman and Koehler, 2006]
-(our reference \alert{[[fill in]])}.  If the reviewer thinks that it
-would strengthen our paper, we can recapitulate some of that material
-in our Section IVD.
+(our reference [21]).  If the reviewer thinks that it would strengthen
+our paper, we can recapitulate some of that material in our Section
+IVD.
+
+Note that the fit in the revised version does not omit any points.
+
+\smallskip
 
@@
 
+\smallskip
+
 \emph{Page 10: `error between the predictions and the true continuations'
@@
 
-We have modified the text as the reviewer suggested, and agree that it reads
-more clearly now. We have also modified the figure as suggested.
+We have modified the text as the reviewer suggested and agree that it
+reads more clearly now. We have also modified the figure as suggested.
+
+\smallskip
 
@@
 Thank you for catching our mistake; we have corrected the equation.
+
+\smallskip
 
@@
 We thank the reviewer for taking the time to read our paper and
-provide us with valuable feedback.
+provide us with valuable feedback.  The questions raised in this
+review caused us to completely rethink the results, the fits, the
+findings, and the conclusions---hence the long delay in resubmitting
+this new version.
+
+\smallskip
 
@@
 clearly.
+
+\smallskip
 
@@
 fit.  Note that this fit ($y = a \log(b x + 1)$) is different than in
-the submitted version.  A careful examination of the data, prompted by
-the reviewer's questions, suggested that this was a better match to
-the data; theoretically, too, a log fit is more appropriate because
-any signal with zero entropy should be perfectly predictable, and so
-should ideally have a MASE value of zero.  Finally, note that there
-are a lot more points on this plot than in our original version; this
-is discussed further below.
+the submitted version.  A careful examination of the data and some
+careful thought about the meaning and theoretical bounds of
+MASE---prompted by the reviewer's questions---suggested that this was
+a better match to the data.  This analysis and consideration is
+chronicled in the third paragraph of the results section.  Finally,
+note that there are a lot more points on this plot than in our
+original version; this is discussed further below.
 
@@
 practical criterion for best-case prediction error in a model-free
-way, and to test it in the context of empirical data.
+way, and to test it in the context of empirical data.  We have also
+changed the title to line up with this goal.
 
@@
 These claims have indeed been explored extensively in theory and in
-toy models.  Our specific claim, which was not well explained in the
-submitted version, is that those claims also hold---and can be
+toy models.  Our specific goal, which was not well explained in the
+submitted version, is to show that those claims also hold---and can be
 practically applied---in real-world time-series data that may contain
@@
 definitions of complexity to the introduction and related-work
-section, and clarified the working definition that we use in this
+sections and clarified the working definition that we use in this
 paper.
@@
 The challenge for the practitioner is to know whether there is a
-mismatch between a particular time series and a particular forecast
-model: i.e., whether one can do better and should try another model,
-or whether the time series is so complicated that the model is doing
-as well as one could expect.  The WPE vs. MASE curve in our paper is
-an empirical heuristic that should help practitioners do that.
+mismatch between a particular time series from an unknown system and a
+particular forecast model: i.e., whether one can do better and should
+try another model, or whether the time series is so complicated that
+the model under consideration is doing as well as one could expect.
+The curves and shaded regions on the WPE vs. MASE plots in our paper
+are an empirical heuristic that should help practitioners do that.
 
@@
 matters and also added an explanatory paragraph directly following
-those findings.  While this lengthened the introduction, we feel that
-it has focused and clarified the contribution of the work.
+those findings.
 
@@
 
-Complicated, unknown systems are where prediction really
-matters---especially systems where a good prediction can make a
-practical difference (e.g., reallocation of resources in a computer to
-suit the dynamically changing needs of a program).  Well-known systems
-are certainly good testcases.  As is also mentioned above, the
-relationship posited in our paper does indeed hold for paradigmatic
-systems like the H\'{e}non map and the Lorenz equations.  We had not
-included these examples in our paper because they simply confirmed
-well-known results, and because our focus was on confirming those
-results in the context of noisy, real-world data.  \alert{[[Liz to do,
-      then finish this wording: We have added a sentence in the
-      conclusion (or maybe end of discussion) about the fact that the
-      WPE vs. MASE relationship for these synthetic examples follows
-      the empirically determined pattern in our paper.]]}
+Well-known systems are certainly good testcases, but complicated,
+unknown systems are where prediction really matters---especially
+systems where a good prediction can make a practical difference (e.g.,
+reallocation of resources in a computer to suit the dynamically
+changing needs of a program).  As mentioned above, the relationship
+posited in our paper does indeed hold for paradigmatic systems like
+the H\'{e}non map and the Lorenz equations.  We had not included these
+examples in our paper because they simply confirmed well-known
+results, and because our focus was on confirming those results in the
+context of noisy, real-world data.
+
+\alert{[[Liz to do, then finish
+      this wording: We have added a sentence in the conclusion (or
+      maybe end of discussion) about the fact that the WPE vs. MASE
+      relationship for these synthetic examples follows the
+      empirically determined pattern in our paper.]]}
 

### 2014-08-01 response.tex
response edits

@@
 results, and because our focus was on confirming those results in the
-context of noisy, real-world data.
-
-\alert{[[Liz to do, then finish
-      this wording: We have added a sentence in the conclusion (or
-      maybe end of discussion) about the fact that the WPE vs. MASE
-      relationship for these synthetic examples follows the
-      empirically determined pattern in our paper.]]}
+context of noisy, real-world data.  The penultimate paragraph of the
+results section now explains that the WPE vs. MASE relationship for
+these two synthetic examples (as well as for the Santa Fe ``A'' data
+set and for various nonlinear transformations of our original data)
+all follow the empirically determined pattern in our paper.
 
@@
 included them, however, because they are used (sometimes to great
-success) in practice and because that made for a more comprehensive exploration.
+success) in practice and because that made for a more comprehensive
+exploration.
 
@@
 modeling and prediction communities---follows the procedure outlined
-in Section V B to choose values for the model parameters.
-Sometimes the choices made by that procedure (e.g., setting a
-particular parameter to zero) produce a naive or random-walk
-predictor.  Our method can help practioners know whether this is a
-good or bad thing.  If a particular model performs poorly (compared to
-what the WPE value might suggest), it may be an indication that the
-chosen parameter values were suboptimal---or even that no member of
-the ARIMA family will work and one should try a nonlinear model. A discussion to this effect has been added to Section
-VI in hopes of clarifying this completely valid concern and to illustrate that this is a feature of our proposed heuristic.
+in Section V B to choose values for the model parameters.  Sometimes
+the choices made by that procedure (e.g., setting a particular
+parameter to zero) produce a naive or random-walk predictor.  Our
+method can help practioners know whether this is a good or bad thing.
+If a particular model performs poorly, compared to what the WPE value
+might suggest, it may be an indication that the chosen parameter
+values were suboptimal---or even that no member of the ARIMA family
+will work and one should try a different class of model.  A discussion
+of these issues has been added to Section VI (page 18, left-hand
+column, ``The discussion in the previous paragraph...'') in the hopes
+of clarifying this completely valid concern and illustrating that this
+is a feature of our proposed heuristic.
 
@@
 This short paragraph in this paper, which explains the efficiencies
-and inefficiencies of these schemes, was present for two purposes:
-first, to explain why we have included these simple prediction schemes
-and not just the more-complicated ones; second, to help in explaining
-the specific results described later in the paper.
+and inefficiencies of these schemes, was intended to serve two
+purposes: first, to explain why we have included these simple
+prediction schemes and not just the more-complicated ones; second, to
+help in explaining the specific results described later in the paper.
 
@@
 
-\alert{Liz to do: We have extended the justification for our use of
-  the MASE metric.}  The paper in which MASE was introduced---[Hyndman
-  and Koehler, 2006], our reference \alert{[[fill in]]}---includes an
-extensive evaluation of this metric, as well as a comprehensive
-comparison to other metrics.  If the reviewer thinks that it would
-strengthen our paper, we can recapitulate more of that material in our
-Section IVD. As part of our current work on this topic, we are
-evaluating how other choices of metric affect the shape of our curves,
-but we feel that a comprehensive study of this issue is best left as a
-sequel to this work.
+We have extended the justification for our use of the MASE metric; see
+the last paragraph in the results section.  As mentioned there, the
+paper in which MASE was introduced---[Hyndman and Koehler, 2006], our
+reference [21]---includes an extensive evaluation of this metric, as
+well as a comprehensive comparison to other metrics.  If the reviewer
+thinks that it would strengthen our paper, we can recapitulate more of
+that material in our paper.  As part of our current work on this
+topic, we are evaluating how other choices of metric affect the shape
+of our curves, but we feel that a comprehensive study of this issue is
+best left as a sequel to this work.
 
@@
 WPE values depend not only on wordlength, but also on \emph{data}
-length---since one needs more data to get good statistics on the words
+length, since one needs more data to get good statistics on the words
 for the PE calculation.  These two interacting knobs are not easy to
 set.  Perhaps as a result of this, wordlength choices are rarely
-justified in the permutation entropy literature.
+justified in the permutation entropy literature.  The extent of the
+discussion in [Bandt \& Pompe, {\sl PRL}, 2002], for instance---the
+paper in which PE was proposed---is ``Our entropies are calculated for
+different [$\ell$] but we do not to determine a limit for large
+[$\ell$], although this is an interesting theoretical problem. For
+practical purposes, we recommend 3\dots7...''  The corresponding
+discussion in [Cao {\sl et al.}, {\sl PRE}, 2004] reads ``In their
+paper, Bandt \& Pompe recommended [$\ell$] of 3-7.  We often found
+that 3 and 4 may still be too small, and a value of 5, 6, or 7 seems
+to be the most suitable....''
 
-The extent of the discussion in [Bandt \& Pompe, {\sl PRL}, 2002], for
-instance---the paper in which PE was proposed---is ``Our entropies are
-calculated for different [$\ell$] but we do not to determine a limit
-for large [$\ell$], although this is an interesting theoretical
-problem. For practical purposes, we recommend 3\dots7...''  The
-corresponding discussion in [Cao {\sl et al.}, {\sl PRE}, 2004] reads
-``In their paper, Bandt \& Pompe recommended [$\ell$] of 3-7.  We
-often found that 3 and 4 may still be too small, and a value of 5, 6,
-or 7 seems to be the most suitable....''
-
-The right thing to do, as we argued in Section [[what]], is to compute
-PE across a \emph{range} of wordlengths ($\ell$) and look for a flat
-spot in the curve of PE versus $\ell$.  We used that method to choose
-individual $\ell$ values for each time series studied here.  We
-conjecture that a systematic approach like sill mitigate any variation
-in the results, thereby preserving the shape of the WPE vs. MASE
-curve.  We have not yet confirmed that conjecture, but we are in the
-process of doing so.
+The right thing to do, as we argued in the last paragraph of Section
+V, is to compute PE across a \emph{range} of wordlengths ($\ell$) and
+look for a flat spot in the curve of PE versus $\ell$.  We applied
+that approach \emph{individually} to each of the time series studied
+here to choose a $\ell$ value that was appropriate to each one.  All
+of those values ended up being 6, which is consistent with the
+informal arguments in previous papers, but that was not an \emph{a
+  priori} or global choice.  We conjecture that a systematic approach
+like this will mitigate any variation in the results, thereby
+preserving the shape of the WPE vs. MASE curve.  We have not yet
+confirmed that conjecture, but we are in the process of doing so.
 
@@
   nonlinear transformations of the 30 [[svdtwo]] and [[svdsix]]
-  signals} constructed for this rebuttal letter---the pattern in
-Figure 7 remained virtually unchanged, as described in the following
-entry.  This strongly suggests, but of course does not prove, that our
-approach is general.  As the reviewer suggested, this will require
-extensive study and evaluation in future work.
+  signals} constructed as a result of the reviewer's questions---the
+pattern in Figure 7 remained virtually unchanged, as described in the
+following entry.  This strongly suggests, but of course does not
+prove, that our approach is general.  As the reviewer suggested, this
+will require extensive study and evaluation in future work.
 

### 2014-08-01 response.tex
almost-final response letter

@@
 \newcommand{\alert}[1]{{\color{red}#1}}
+\newcommand{\svdone}{{\tt dgesdd}$_1$\xspace}
+\newcommand{\svdtwo}{{\tt dgesdd$_2$}\xspace}
+\newcommand{\svdthree}{{\tt dgesdd$_3$}\xspace}
+\newcommand{\svdfour}{{\tt dgesdd$_4$}\xspace}
+\newcommand{\svdfive}{{\tt dgesdd$_5$}\xspace}
+\newcommand{\svdsix}{{\tt dgesdd$_6$}\xspace}
 
@@
 dynamical behaviors.  For all of these examples---as well as for the
-39 additional ones\footnote{H\'{e}non, Lorenz, SFI A, random walk,
-  nonlinear transformations of the 30 [[svdtwo]] and [[svdsix]]
-  signals} constructed as a result of the reviewer's questions---the
-pattern in Figure 7 remained virtually unchanged, as described in the
-following entry.  This strongly suggests, but of course does not
-prove, that our approach is general.  As the reviewer suggested, this
-will require extensive study and evaluation in future work.
+49 additional ones
+% \footnote{H\'{e}non, Lorenz, SFI A, 
+%  nonlinear transformations of the 45 svdone, svdtwo and signals}
+constructed as a result of the reviewer's questions and described in
+various parts of this response letter---the pattern in Figure 7
+remained virtually unchanged.  This strongly suggests, but of course
+does not prove, that our approach is general.  As the reviewer
+suggested, this will require extensive study and evaluation in future
+work.
 
@@
 
-\alert{Liz to fix following text, now that we have these experiments}
+Regarding the second part of this point: we have revised the fits in
+Figures 6 and 7 in the paper, and the associated text, to include all
+120 points.  As mentioned above, we have also moved from a linear fit
+to a log fit of these points in order to better match the theoretical
+relationship between MASE and WPE.  We now show and discuss the
+one-$\sigma$ volume around that fit in an effort to emphasize that the
+fundamental claim of the paper is not related to the crispness of the
+fit.
 
-PE is invariant under an invertible nonlinear transformation of the
-values in the time series, but WPE is not.  But MASE varies \emph{in a
-  complementary way}.  Ryan will do $x(t)$ and $5e^{x(t)}$ and show
-that it's on the curve.
+Regarding the first part of this point: while PE is invariant under an
+invertible nonlinear transformation of the values in the time series,
+WPE is not---which amplifies the potential concerns about this issue.
+To explore this, we performed the invertible nonlinear transformation
+$f(x)=5e^{x(t)}$ on several of our traces ($x(t)=$ {\tt dgesdd}$_2$,
+{\tt dgesdd}$_5$, and {\tt dgesdd}$_6$).  The image below shows WPE
+vs. MASE results of the best predictor on these three traces, along
+with those on the untransformed originals:
 
-% Ryan will insert figure here with x(t) = svdone and svdsix
 \begin{center}
@@
 
-This is consistent with our claim of generality; this is just another
-time series.  A xform like this effectively changes the generating
-process, but our model-free result handles it.
+The WPE vs. MASE points of the transformed data move---upwards,
+generally---but they remain within the shaded region.  Indeed, the
+values for the transformed version of {\tt dgesdd}$_5$ (dark-green
+icons) move {\sl closer} to the curve.  When these 45 new points are
+included in the $y = a \log(b x + 1)$ fit, the coefficients
+\alert{barely change: $a$ by n\% and $b$ by n\%.  Ryan, what are these
+  numbers?}
 
-We have revised the Figure, and the associated text, to include all
-120 points.  We have also moved from a linear fit to a log fit in this
-curve to better match the theoretical relationship between MASE and
-WPE, as mentioned above.  Finally, we now show a ``fuzzy'' version of
-the fitted curve in an effort to emphasize that the fundamental claim
-of the paper is not related to the crispness of the fit.
-
-Furthermore, we have recomputed the fit using both the 120 computer performance
-data time-series as well as the 45 nonlinear transforms and the H\'{e}non, Lorenz,
-random walk, and SFI Data A time-series. Not only does the fit line hardly
-change, but the error bounds are contained completely within those provided by
-just the computer performance data. This can be seen in the following figure
-where the blue fit and bounds are those from the computer performance traces and
-the magenta ones are those including the additional 49 time-series.
+This supports our claims of generality: a transformation like this
+effectively changes the generating process, but does not affect the
+performance of our model-free heuristic.  And the curves that capture
+that heuristic---which were derived from data from a single
+computer-performance experiment---do not only hold for traces from
+that experiment and transformed versions of them.  The preliminary
+results of our broader study, currently underway, indicate that they
+also hold for data from synthetic examples (the H\'{e}non map, the
+Lorenz equations, and the output of a purely stochastic random-walk
+process) and for real-world data from a different system (SFI data set
+A).  The image below shows some of these results:
 
@@
 
-\alert{Need the change and the confidence stuff for this: Fit changed
-  by [[what]] when we redid it with these 34 new points.}  It now
-reflects a map, a flow, a purely stochastic signal, a variety of
-real-world data from two different systems (laser, computer), and
-nonlinear transformations of that real-world data.  Seems pretty
-compelling.
+The blue fit lines here are derived from the original 120 traces; the
+magenta one is fit to those traces plus the nonlinear transformations
+of {\tt dgesdd}$_2$, {\tt dgesdd}$_5$, and {\tt dgesdd}$_6$), a
+10,000-point orbit from the H\'{e}non map with $a=1.4$ and $b=0.3$,
+the $x$ coordinate of a 10,000-step 4th-order Runge-Kutta solution of
+the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$ and $\Delta
+t=0.01$, an \alert{nnn-point} random-walk time series generated
+\alert{how, Ryan?}, and data set ``A'' from the Santa Fe Institute
+time-series competition website.
+
+Including these additional 49 points hardly changed the main central
+curve of the fit.  Indeed, doing so caused the one-$\sigma$ volume of
+that fit to shrink (viz., the dashed magenta curves inside the dashed
+blue curves).  That is, the additional points firmed up the findings.
+The magenta curve reflects a map, a flow, a purely stochastic signal,
+a variety of real-world data from two different systems (laser,
+computer), and nonlinear transformations of that real-world data
+
+We have added a sentence about these preliminary results to the
+penultimate paragraph in Section VI.
+
+\smallskip
 
@@
 
-Because of the depth and extent of the revisions that we made in
-response to this reviewer's comments, we did not feel that it was
-effective to highlight the changes in a different color in the
-manuscript; please let us know if a more-detailed description of the
-changes would make the re-reading easier.
-
 \end{document}

### 2014-08-01 related.tex
pushing changes midstream, in case of lossage
added comment abotu boffetta

@@
 for predictability bounds, this approach differs from our work in a
-number of fundamental ways.  First, we focus on real-world time-series
-data, where one does not necessarily have the ability to perturb or
-otherwise interact with the system of interest, nor can one obtain or
-manufacture the (possibly large) number of points that might be needed
-to estimate the $\epsilon$-entropy for small $\epsilon$.  Second, we
-do not require \emph{a priori} knowledge about the noise and its
-interaction with the system.  Third, we tie information---in the form
-of the weighted permutation entropy---directly to prediction error via
-MASE values.  Though FSLE and $\epsilon$-entropy allow for the
-comparison of predictability between systems, they do not directly
-provide an estimate of prediction error.  Finally, our approach also
-holds for stochastic systems, where neither the FLSEs nor their
-relationship to predictability is well defined.
+number of fundamental ways.  First, Boffetta \emph{et al.} is a
+theoretical study; they do not work with any real predictions.  We
+focus on real-world time-series data, where one does not necessarily
+have the ability to perturb or otherwise interact with the system of
+interest, nor can one obtain or manufacture the (possibly large)
+number of points that might be needed to estimate the
+$\epsilon$-entropy for small $\epsilon$.  Second, we do not require
+\emph{a priori} knowledge about the noise and its interaction with the
+system.  Third, we tie information---in the form of the weighted
+permutation entropy---directly to prediction error via calculated
+values of a specific error metric.  Though FSLE and $\epsilon$-entropy
+allow for the comparison of predictability between systems, they do
+not directly provide an estimate of prediction error.  Finally, our
+approach also holds for stochastic systems, where neither the FLSEs
+nor their relationship to predictability is well defined.

### 2014-08-01 response.tex
synching everything

@@
 findings, and the conclusions---hence the long delay in resubmitting
-this new version.
+this new version, and the length and depth of this response letter.
 
@@
 We have shortened the introduction and modified the manuscript
-throughout to make it more clear that our goal is to derive a
+throughout to make it more clear that our goal was to derive a
 practical criterion for best-case prediction error in a model-free
@@
 toy models.  Our specific goal, which was not well explained in the
-submitted version, is to show that those claims also hold---and can be
-practically applied---in real-world time-series data that may contain
-dynamical and/or observational noise.  This has not, to our knowledge,
-received much attention in the literature.  We have added a paragraph
-to the introduction that hopefully makes all of this more clear.
+submitted version, was to show that those claims also hold---and can
+be practically applied---in real-world time-series data that may
+contain dynamical and/or observational noise.  This has not, to our
+knowledge, received much attention in the literature.  We have added a
+paragraph to the introduction that hopefully makes all of this more
+clear.
 
 We appreciate being informed about the Boffetta et al.  paper, and we
-have included a discussion of it at the end of the related-work
-section.
+have included a discussion of it at the end of Section II.
 
@@
 We have added a short discussion about the differences in notions and
-definitions of complexity to the introduction and related-work
-sections and clarified the working definition that we use in this
-paper.
+definitions of complexity to Sections I and II and clarified the
+working definition that we use in our work.
 
@@
 stated, is indeed trivial.  Our intent, as we have modified the text
-to reflect, is that the appropriateness of a predictor can be
-discovered \emph{empirically}, via comparison of its MASE value with
-its WPE value.  Previous explorations of predictability have used
-clean data and/or systems where the generating process is known.  Our
-work makes neither of these assumptions.
+to reflect, is that the appropriateness of a predictor for a given
+time series can be discovered \emph{empirically}, via comparison of
+its MASE value with the WPE value of the data.  Previous explorations
+of predictability have used clean data and/or systems where the
+generating process is known.  Our work makes neither of these
+assumptions.
 
@@
 
-The challenge for the practitioner is to know whether there is a
-mismatch between a particular time series from an unknown system and a
-particular forecast model: i.e., whether one can do better and should
-try another model, or whether the time series is so complicated that
-the model under consideration is doing as well as one could expect.
-The curves and shaded regions on the WPE vs. MASE plots in our paper
-are an empirical heuristic that should help practitioners do that.
+The challenge that this poses for a practitioner is to know whether
+there is a mismatch between a particular time series from an unknown
+system and a particular forecast model: i.e., whether one can do
+better and should try another model, or whether the time series is so
+complicated that the model under consideration is doing as well as one
+could expect.  The curves and shaded regions on the WPE vs. MASE plots
+in our paper are an empirical heuristic that should help practitioners
+do that.
 
@@
 
-Well-known systems are certainly good testcases, but complicated,
+Well-known systems are certainly good test cases, but complicated,
 unknown systems are where prediction really matters---especially
@@
 reallocation of resources in a computer to suit the dynamically
-changing needs of a program).  As mentioned above, the relationship
-posited in our paper does indeed hold for paradigmatic systems like
-the H\'{e}non map and the Lorenz equations.  We had not included these
-examples in our paper because they simply confirmed well-known
-results, and because our focus was on confirming those results in the
-context of noisy, real-world data.  The penultimate paragraph of the
-results section now explains that the WPE vs. MASE relationship for
-these two synthetic examples (as well as for the Santa Fe ``A'' data
-set and for various nonlinear transformations of our original data)
-all follow the empirically determined pattern in our paper.
+changing needs of a program).  The relationship posited in our paper
+does indeed hold for paradigmatic systems like the H\'{e}non map and
+the Lorenz equations.  We had not included these examples in our paper
+because they simply confirmed well-known results, and because our
+focus was on confirming those results in the context of noisy,
+real-world data.  The penultimate paragraph of the results section now
+explains that the WPE vs. MASE relationship for these two synthetic
+examples (as well as for the Santa Fe ``A'' data set, for a
+white-noise process, and for various nonlinear transformations of our
+original data) all follow the empirically determined pattern in our
+paper.
 
@@
 
-The primary goal of this paper is to explore the relationship between
-prediction error and WPE in noisy, real-world systems, and the
-practical applicability of this relationship.  Accordingly, we have
-cut down on the level of detail regarding the benchmark system and
-pointed interested readers to our other papers for the engineering
-details.  We did not feel that removing all of this material would
-be effective, however, because it helps the reader understand why
-this is a good testbed for exploring the problem at hand.
+The primary goal of this paper was to explore the relationship between
+prediction error and WPE in noisy, real-world systems, and to assess
+the practical applicability of this relationship.  Accordingly, we
+have cut down on the level of detail regarding the benchmark system
+and pointed interested readers to our other papers for the engineering
+details.  We did not feel that removing all of this material would be
+effective, however, because it helps the reader understand why this is
+a good testbed for exploring the problem at hand.
 
@@
 included them, however, because they are used (sometimes to great
-success) in practice and because that made for a more comprehensive
-exploration.
+success) in practice, because that made for a more-comprehensive
+exploration, and because of a subtle issue regarding model building
+procedures.
 
@@
 the last paragraph in the results section.  As mentioned there, the
-paper in which MASE was introduced---[Hyndman and Koehler, 2006], our
-reference [21]---includes an extensive evaluation of this metric, as
-well as a comprehensive comparison to other metrics.  If the reviewer
-thinks that it would strengthen our paper, we can recapitulate more of
-that material in our paper.  As part of our current work on this
-topic, we are evaluating how other choices of metric affect the shape
-of our curves, but we feel that a comprehensive study of this issue is
-best left as a sequel to this work.
+paper in which MASE was introduced---our reference [21]---includes an
+extensive evaluation of this metric, as well as a comprehensive
+comparison to other metrics.  If the reviewer thinks that it would
+strengthen our paper, we can recapitulate more of that material in our
+paper.  As part of our current work on this topic, we are evaluating
+how other choices of metric affect the shape of our curves, but we
+feel that a comprehensive study of this issue is best left as a sequel
+to this work.
 
@@
 length, since one needs more data to get good statistics on the words
-for the PE calculation.  These two interacting knobs are not easy to
-set.  Perhaps as a result of this, wordlength choices are rarely
-justified in the permutation entropy literature.  The extent of the
-discussion in [Bandt \& Pompe, {\sl PRL}, 2002], for instance---the
-paper in which PE was proposed---is ``Our entropies are calculated for
-different [$\ell$] but we do not to determine a limit for large
-[$\ell$], although this is an interesting theoretical problem. For
-practical purposes, we recommend 3\dots7...''  The corresponding
-discussion in [Cao {\sl et al.}, {\sl PRE}, 2004] reads ``In their
-paper, Bandt \& Pompe recommended [$\ell$] of 3-7.  We often found
-that 3 and 4 may still be too small, and a value of 5, 6, or 7 seems
-to be the most suitable....''
+for the PE calculation if those words are longer.  These two
+interacting knobs are not easy to set.  Perhaps as a result of this,
+wordlength choices are rarely justified in the permutation entropy
+literature.  The extent of the discussion in [Bandt \& Pompe, {\sl
+    PRL}, 2002], for instance---the paper in which PE was
+proposed---is ``Our entropies are calculated for different [$\ell$]
+but we do not determine a limit for large [$\ell$], although this is
+an interesting theoretical problem. For practical purposes, we
+recommend 3\dots7...''  The corresponding discussion in [Cao {\sl et
+    al.}, {\sl PRE}, 2004] reads ``In their paper, Bandt \& Pompe
+recommended [$\ell$] of 3-7.  We often found that 3 and 4 may still be
+too small, and a value of 5, 6, or 7 seems to be the most
+suitable....''
 
@@
 look for a flat spot in the curve of PE versus $\ell$.  We applied
-that approach \emph{individually} to each of the time series studied
-here to choose a $\ell$ value that was appropriate to each one.  All
-of those values ended up being 6, which is consistent with the
+that approach individually to each of the time series studied here to
+choose a $\ell$ value that was appropriate to that particular trace.
+(All of those values ended up being 6, which is consistent with the
 informal arguments in previous papers, but that was not an \emph{a
-  priori} or global choice.  We conjecture that a systematic approach
+  priori} or global choice.)  We conjecture that a systematic approach
 like this will mitigate any variation in the results, thereby
@@
 constructed as a result of the reviewer's questions and described in
-various parts of this response letter---the pattern in Figure 7
+different parts of this response letter---the pattern in Figure 7
 remained virtually unchanged.  This strongly suggests, but of course
@@
 
-This supports our claims of generality: a transformation like this
-effectively changes the generating process, but does not affect the
-performance of our model-free heuristic.  And the curves that capture
-that heuristic---which were derived from data from a single
-computer-performance experiment---do not only hold for traces from
-that experiment and transformed versions of them.  The preliminary
-results of our broader study, currently underway, indicate that they
-also hold for data from synthetic examples (the H\'{e}non map, the
-Lorenz equations, and the output of a purely stochastic random-walk
-process) and for real-world data from a different system (SFI data set
-A).  The image below shows some of these results:
+This series of experiments supports our claims of generality: a
+transformation like this effectively changes the generating process,
+but does not affect the performance of our model-free heuristic.  And
+the curves that capture that heuristic---which were derived from data
+from a single computer-performance experiment---do not only hold for
+traces from that experiment and transformed versions of them.  The
+preliminary results of our broader study, currently underway, indicate
+that they also hold for data from synthetic examples (the H\'{e}non
+map, the Lorenz equations, and the output of a purely stochastic
+random-walk process) and for real-world data from a different system
+(SFI data set A).  The image below shows some of these results:
 
@@
 magenta one is fit to those traces plus the nonlinear transformations
-of {\tt dgesdd}$_2$, {\tt dgesdd}$_5$, and {\tt dgesdd}$_6$), a
-10,000-point orbit from the H\'{e}non map with $a=1.4$ and $b=0.3$,
-the $x$ coordinate of a 10,000-step 4th-order Runge-Kutta solution of
-the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$ and $\Delta
-t=0.01$, an \alert{nnn-point} random-walk time series generated
+of {\tt dgesdd}$_2$, {\tt dgesdd}$_5$, and {\tt dgesdd}$_6$, the $x$
+coordinate of a 10,000-point orbit from the H\'{e}non map with $a=1.4$
+and $b=0.3$, the $x$ coordinate of a 10,000-step 4th-order Runge-Kutta
+solution of the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$
+and $\Delta t=0.01$, a 10,000-point random-walk time series generated
 \alert{how, Ryan?}, and data set ``A'' from the Santa Fe Institute
@@
 
-Including these additional 49 points hardly changed the main central
-curve of the fit.  Indeed, doing so caused the one-$\sigma$ volume of
-that fit to shrink (viz., the dashed magenta curves inside the dashed
-blue curves).  That is, the additional points firmed up the findings.
-The magenta curve reflects a map, a flow, a purely stochastic signal,
-a variety of real-world data from two different systems (laser,
-computer), and nonlinear transformations of that real-world data
+Note that the inclusion of these additional 49 points hardly changed
+the main central curve of the fit.  Indeed, doing so caused the
+one-$\sigma$ volume of that fit to shrink (viz., the dashed magenta
+curves inside the dashed blue curves).  That is, the additional points
+firmed up the findings.  The magenta curve reflects a map, a flow, a
+purely stochastic signal, a variety of real-world data from two
+different systems (laser, computer), and nonlinear transformations of
+that real-world data---again, supporting the generality of our
+heuristic.
 
-We have added a sentence about these preliminary results to the
-penultimate paragraph in Section VI.
+We are in the process of broadening this study to other systems,
+transformations, and situations; in the meantime, we have added a
+sentence about these preliminary results to the penultimate paragraph
+in Section VI.
 

### 2014-08-01 results.tex
synching everything

@@
 for choosing parameters for an ARIMA model of a given data set.  While
-the tests and criteria employed by this algorithm are sophisticated
-(see Section~\ref{sec:arima}), the results can still be
+the tests and criteria employed by this algorithm
+(Section~\ref{sec:arima}) are sophisticated , the results can still be
 sub-optimal---if the initial space of models being searched is not
@@
 situation: all of these signals had the same complexity, but some (the
-outliers in the blue cluster of $+$ points in the bottom-left panel of
-Figure~\ref{fig:what}) had much higher \arima MASE values, indicating
-a failure in the \arima process.  We then went back and manually
-assisted the automated fitting procedure---forcing inclusion of a term
-that is appropriate for a nonstationary signal.  In all cases, this
-produced ARIMA models with MASE scores in line with the others in that
-cluster.
+outliers in the blue cluster of {\color{blue}$+$} points in the
+bottom-left panel of Figure~\ref{fig:wpe_vs_mase_all}) had much higher
+MASE values, indicating a failure in the \arima process.  We then went
+back to the data, noticed some shifts of the baseline in those
+signals, and manually assisted the automated fitting procedure to
+force the inclusion of a term that is appropriate for a nonstationary
+signal.  In all cases, this produced ARIMA models with MASE scores in
+line with the others in that cluster.
 
@@
 to exceed 1.15, but pathological situations (e.g., a huge switch in
-scale right at the training/test signal boundary) could produce higher
-values.  This suggests another potentially useful heuristic: if the
-MASE of a random-walk prediction of a time series is significantly
-different from 1, this could be an indication that the signal is
-nonstationary.  We are in the process of exploring this idea.
+scale right at the training/test signal boundary, or a signal that
+simply grows exponentially) could produce higher values.  This
+suggests another potentially useful heuristic: if the MASE of a
+random-walk prediction of a time series is significantly different
+from 1, this could be an indication that the signal is nonstationary.
+We are in the process of exploring this idea.
 

### 2014-08-01 results.tex
added mention of DVS fits

@@
 parameter space.  The meaning of the shaded region is explained in the
-text.}
+text. \alert{Do you think we should explain that here, rather than
+  punting to the text?  It is, after all, the primary claim of the
+  paper...}}
   \label{fig:wpe_vs_mase_best}
@@
 from 1, this could be an indication that the signal is nonstationary.
-We are in the process of exploring this idea.
+We are in the process of exploring this idea and its relationship to
+the DVS fits mentioned in Section~\ref{sec:related}.
 

### 2014-08-02 response.tex
add details about random walk

@@
-\documentclass[12pt]{article}
+e\documentclass[12pt]{article}
 
@@
 solution of the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$
-and $\Delta t=0.01$, a 10,000-point random-walk time series generated
-\alert{how, Ryan?}, and data set ``A'' from the Santa Fe Institute
-time-series competition website.
+and $\Delta t=0.01$, a 10,000-step random-walk time series generated
+by taking the cumulative sum of $\mu=0$, $\sigma=1$ Gaussian random 
+variables, and data set ``A'' from the Santa Fe Institute time-series
+competition website.
 

### 2014-08-03 response.tex
response is done except alert for Ryan

--- 2014-08-01 response.tex
@@
 findings, and the conclusions---hence the long delay in resubmitting
-this new version, and the length and depth of this response letter.
+this new version, the amount of rewritten material in the manuscript,
+and the length of this response letter.
 
@@
 holds for a time series of voltages from a chaotic laser used in the
-SFI prediction competition, for instance.  (It also holds for various
-paradigmatic systems, such as the H\'{e}non map and the Lorenz system, as
-discussed in a later section of this response.)  The image below is an
-augmented version of our revised Figure 7 that includes all of these
-examples.
+SFI prediction competition, for instance.  It also holds for various
+paradigmatic systems, such as the H\'{e}non map, the Lorenz system,
+and a white-noise process.  The image below is an augmented version of
+our revised Figure 7 that includes all of these examples; details
+about each series and about the effects upon the fit are described
+later in this letter.
 
@@
 points from the other time series fall within a narrow band around the
-fit.  Note that this fit ($y = a \log(b x + 1)$) is different than in
-the submitted version.  A careful examination of the data and some
-careful thought about the meaning and theoretical bounds of
-MASE---prompted by the reviewer's questions---suggested that this was
-a better match to the data.  This analysis and consideration is
-chronicled in the third paragraph of the results section.  Finally,
-note that there are a lot more points on this plot than in our
-original version; this is discussed further below.
+fit.
+
+Note that this fit ($y = a \log(b x + 1)$) is different than in the
+submitted version.  A careful examination of the data and some careful
+thought about the meaning and theoretical bounds of MASE---prompted by
+the reviewer's questions---suggested that this was a better match to
+the data.  This analysis and consideration is chronicled in the third
+paragraph of the results section.  Finally, note that there are a lot
+more points on this plot than in our original version; this is
+discussed further below.
 
@@
 reallocation of resources in a computer to suit the dynamically
-changing needs of a program).  The relationship posited in our paper
-does indeed hold for paradigmatic systems like the H\'{e}non map and
-the Lorenz equations.  We had not included these examples in our paper
-because they simply confirmed well-known results, and because our
-focus was on confirming those results in the context of noisy,
-real-world data.  The penultimate paragraph of the results section now
-explains that the WPE vs. MASE relationship for these two synthetic
-examples (as well as for the Santa Fe ``A'' data set, for a
-white-noise process, and for various nonlinear transformations of our
-original data) all follow the empirically determined pattern in our
-paper.
+changing needs of a program).  The relationship posited in our paper,
+as mentioned above, does indeed hold for paradigmatic systems like the
+H\'{e}non map and the Lorenz equations.  We had not included these
+examples in our paper because they simply confirmed well-known
+results, and because our focus was on confirming those results in the
+context of noisy, real-world data.  The penultimate paragraph of the
+results section now explains that the WPE vs. MASE relationship for
+these two synthetic examples (as well as for the Santa Fe ``A'' data
+set, for a white-noise process, and for various nonlinear
+transformations of our original data) all follow the empirically
+determined pattern in our paper.
 
@@
 
-We have changed the term used in the paper to {\tt auto.arima} in
+We have also changed the term used in the paper to {\tt auto.arima} in
 order to make it clear that this refers not to the general class of
@@
 We have extended the justification for our use of the MASE metric; see
-the last paragraph in the results section.  As mentioned there, the
-paper in which MASE was introduced---our reference [21]---includes an
-extensive evaluation of this metric, as well as a comprehensive
-comparison to other metrics.  If the reviewer thinks that it would
-strengthen our paper, we can recapitulate more of that material in our
-paper.  As part of our current work on this topic, we are evaluating
-how other choices of metric affect the shape of our curves, but we
-feel that a comprehensive study of this issue is best left as a sequel
-to this work.
+the last paragraph in Section VI.  As mentioned there, the paper in
+which MASE was introduced---our reference [21]---includes an extensive
+evaluation of this metric, as well as a comprehensive comparison to
+other metrics.  If the reviewer thinks that it would strengthen our
+paper, we can recapitulate more of that material in our paper.  As
+part of our current work on this topic, we are evaluating how other
+choices of metric affect the shape of our curves, but we feel that a
+comprehensive study of this issue is best left as a sequel to this
+work.
 
@@
 preserving the shape of the WPE vs. MASE curve.  We have not yet
-confirmed that conjecture, but we are in the process of doing so.
+confirmed that conjecture, but we are working on it.
 
@@
 values for the transformed version of {\tt dgesdd}$_5$ (dark-green
-icons) move {\sl closer} to the curve.  When these 45 new points are
-included in the $y = a \log(b x + 1)$ fit, the coefficients
-\alert{barely change: $a$ by n\% and $b$ by n\%.  Ryan, what are these
-  numbers?}
+icons) move {\sl closer} to the curve.
 
@@
 random-walk process) and for real-world data from a different system
-(SFI data set A).  The image below shows some of these results:
+(SFI data set A).  The image below shows these results:
 
@@
 the main central curve of the fit.  Indeed, doing so caused the
-one-$\sigma$ volume of that fit to shrink (viz., the dashed magenta
-curves inside the dashed blue curves).  That is, the additional points
-firmed up the findings.  The magenta curve reflects a map, a flow, a
-purely stochastic signal, a variety of real-world data from two
-different systems (laser, computer), and nonlinear transformations of
-that real-world data---again, supporting the generality of our
+one-$\sigma$ volume of the fit to shrink (viz., the dashed magenta
+curves inside the dashed blue curves).  \emph{That is, the additional
+  points firmed up the findings.}  The magenta curve reflects a map, a
+flow, a purely stochastic signal, a variety of real-world data from
+two different systems (laser, computer), and nonlinear transformations
+of that real-world data---again, supporting the generality of our
 heuristic.

### 2014-08-04 response.tex
remove erroneous 'e'
added two alerts to response

@@
-e\documentclass[12pt]{article}
+\documentclass[12pt]{article}
 
@@
 of discussion of this cluster of issues in [Hyndman and Koehler, 2006]
-(our reference [21]).  If the reviewer thinks that it would strengthen
-our paper, we can recapitulate some of that material in our Section
-IVD.
+\alert{(our reference [21])}.  If the reviewer thinks that it would
+strengthen our paper, we can recapitulate some of that material in our
+Section IVD.
 
@@
 the last paragraph in Section VI.  As mentioned there, the paper in
-which MASE was introduced---our reference [21]---includes an extensive
-evaluation of this metric, as well as a comprehensive comparison to
-other metrics.  If the reviewer thinks that it would strengthen our
-paper, we can recapitulate more of that material in our paper.  As
-part of our current work on this topic, we are evaluating how other
-choices of metric affect the shape of our curves, but we feel that a
-comprehensive study of this issue is best left as a sequel to this
-work.
+which MASE was introduced---\alert{our reference [21]}---includes an
+extensive evaluation of this metric, as well as a comprehensive
+comparison to other metrics.  If the reviewer thinks that it would
+strengthen our paper, we can recapitulate more of that material in our
+paper.  As part of our current work on this topic, we are evaluating
+how other choices of metric affect the shape of our curves, but we
+feel that a comprehensive study of this issue is best left as a sequel
+to this work.
 

### 2014-08-04 concl.tex
added mention of DVS fits
Liz final read through round I

--- 2014-07-25 concl.tex
@@
 \section{ Conclusions \& Future Work }\label{sec:conc}
+
+% To address the shortcomings mentioned in the previous section, we are
+% in the process of repeating our own study with other metrics, other
+% forecast methods, and time-series data from other systems.
 
@@
 converge to the true entropy value. Using a variety of forecast models
-and more than a hundred time-series data sets from a
-computer-performance experiment, we have shown that prediction
-accuracy is indeed correlated with weighted permutation entropy: the
-higher the WPE, in general, the higher the prediction error.  The
-relationship is roughly logarithmic.
-
-[[Maybe put (or reiterate) the ``range of examples'' material here.
-    This is just an observed result.  It makes sense because any
-    signal with zero entropy should be perfectly predictable, and so
-    should ideally have a MASE value of zero.  As the entropy rises,
-    the MASE value rise: slowly at first---where the signal is not
-    very complex---and then more rapidly after WPE $\approx$ 0.5.  We
-    are in the process of exploring the theory behind this shape, as
-    well as testing other shapes.  Supporting our hypothesis, forecast
-    errors for the \col and \svdfive signals are higher than the
-    corresponding WPEs suggest---except for the nonlinear LMA
-    predictor.  There are some exceptions: \svdone is easier to
-    predict than its WPE value suggests using any of the three
-    forecast strategies studied here.  On the whole, though, signals
-    with higher weighted permutation entropy are harder to predict.
-
-% Further, the persistent WPE values of 0.5--0.6 for the {\tt
-%   col\_major} trace are consistent with dynamical chaos, further
-% corroborating the results of~\cite{mytkowicz09}.
+and more than 150 time-series data sets from experiments and
+simulations, we have shown that prediction accuracy is indeed
+correlated with weighted permutation entropy: the higher the WPE, in
+general, the higher the prediction error.  The relationship is roughly
+logarithmic, which makes theoretical sense, given the nature of WPE,
+predictability, and MASE.
 
@@
 %  \naive ~wins on \svdone with the exception to the 5 outliers.
-The dashed curve and shaded region in
-Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
-operationalize the discussion in this paragraph.  These geometric
-features are a preliminary, but potentially useful, empirical
-guideline for knowing when a model is not well-matched to the task at
-hand: a MASE score that is significantly\footnote{more than one
-  $\sigma$ out} higher than [[$3.58 \cdot \mathrm{WPE} - 1.91$]],
-where $\mathrm{WPE}$ is that of the corresponding signal, is an
-indication that that time series has more predictive structure than
-that forecast model can capture and exploit.
 
-[[This paragraph duplicates material in the last paragraph of this
-    section.  Some of it needs to be here rather than there, judging
-    by the ref's reactions.  But the later paragraph has a nice
-    concluding tone, so need to figure out what goes where.]]
+The curves and shaded regions in Figures~\ref{fig:wpe_vs_mase_best}
+and~\ref{fig:wpe_vs_mase_all} generalize and operationalize the
+discussion in the previous paragraph.  These geometric features are a
+preliminary, but potentially useful, empirical heurists for knowing
+when a model is not well-matched to the task at hand: a point that is
+below and/or to the right of the shaded regions on a plot like
+Figure~\ref{fig:wpe_vs_mase_best} indicates the time series has more
+predictive structure than the forecast model can capture and
+exploit---and that one would be well advised to try another method.
 
-This curve was determined empirically using a finite set of forecast
-methods and data sets---the latter drawn from a single experimental
-system.  Computer dynamics spans the full range of behavior reported
-in the nonlinear dynamics literature.  Standard nonlinear time-series
-analysis of data from these systems has confirmed the existence of
-fixed points, periodic orbits of different periods, chaotic
-trajectories with various Lyapunov exponents, etc. in their dynamics.
-While it is certainly possible that the nonlinear dynamics of computer
-performance is subtly different from the nonlinear dynamics of other
-systems---in a manner that cannot be detected by traditional nonlinear
-time-series analysis---we believe that our results will hold for time
-series drawn from other physical systems.  We are certainly aware that
-this claim may not be true, however, and we are in the process of a
-follow-up study exploring a broader range of applications.
-Preliminary results of this study show that MASE vs. WPE points for
-data from the SFI prediction competition \cite{weigend-book} fall in
-the shaded region on Figure [[what]]\footnote{Synthetic data from
-  paradigmatic systems like the H\'{e}non map \cite{henon} and the
-  Lorenz system \cite{lorenz} also fit this pattern.}.
+These curves were determined empirically using a specific error metric
+and a finite set of forecast methods and time series traces.  If one
+uses a different error metric, the geometry---and the heuristic---will
+no longer apply.  And while the methods and traces were chosen to be
+representative of the practice, they are of course not completely
+comprehensive.  It is certainly possible, for instance, that the
+nonlinear dynamics of computer performance is subtly different from
+the nonlinear dynamics of other systems.  Our preliminary results on
+other systems (Henon, Lorenz, white noise, SFI ``A'', transformations
+of the computer performance data) lead us to believe that our results
+will generalize beyond the examples described in this paper.  We are
+in the process of following up on that exploration with a broader
+study of data, forecast methods, and error metrics.
 
-Given that information is a fundamental limitation in predictability,
-then gathering and using more information is an obvious next step.
-But there is an equally obvious tension here between data length and
-prediction speed: a forecast that requires half a second to compute is
-not useful for the purposes of real-time control of a computer system
-with a MHz clock rate, for instance.  Another alternative is to sample
-several system variables simultaneously and build multivariate models.
-This is a particular challenge in nonlinear LMA-type models, since
-multivariate delay-coordinate embedding (e.g.,
+\alert{I'm not sure this paragraph adds anything.  Can/should we
+  remove it?  Given that information is a fundamental limitation in
+  predictability, then gathering and using more information is an
+  obvious next step.  But there is an equally obvious tension here
+  between data length and prediction speed: a forecast that requires
+  half a second to compute is not useful for the purposes of real-time
+  control of a computer system with a MHz clock rate, for instance.
+  Another alternative is to sample several system variables
+  simultaneously and build multivariate models.  This is a particular
+  challenge in nonlinear LMA-type models, since multivariate
+  delay-coordinate embedding (e.g.,
   \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
   computationally prohibitive.  We are working on alternative methods
-that sidestep that complexity.
+  that sidestep that complexity.}
 
@@
 Nonstationarity is a serious challenge in any time-series modeling
-problem.  Indeed, one of the first applications of permutation entropy
-was to recognize the regime shift in brainwave data that occurs when
-someone has a seizure~\cite{cao2004det}.  Recall that the signal in
-Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
-this paper because it contained a number of different regimes.  We
-segmented this signal visually, but one could imagine using
-permutation entropy to do so instead.  Automating regime-shift
+problem.  [[Can destroy our heuristic: regime shifts can cause the
+    test signal to look different from the training signal, which can
+    skew the MASE score (and the points on the curve).  A RW MASE
+    score larger than 1, however, as mentioned at end of prev section,
+    can flag when this is happening.]]  Detecting regime shifts---and
+adapting prediction models accordingly---is an important area of
+future work.  Indeed, one of the first applications of permutation
+entropy was to recognize the regime shift in brainwave data that
+occurs when someone has a seizure~\cite{cao2004det}.  Recall that the
+signal in Figure~\ref{fig:svd-ts-colored} was especially useful for
+the study in this paper because it contained a number of different
+regimes.  We segmented this signal visually, but one could imagine
+using permutation entropy to do so instead.  Automating regime-shift
 detection would be an important step towards a fully adaptive modeling
@@
 with regime-shift detection, but also suggest what kind of model might
-work well in each new regime.
-
-All of the experiments described in this paper involve a particular
-physical system.  We chose that system because its behavior spans a
-broad spectrum of classical time series patterns.  Since these
-patterns (dynamical chaos, periodicities, noise, etc.) are generic, we
-believe that our results will generalize beyond this particular
-system.  Testing that claim with different examples is another
-important next step---and one that will be critical to validating our
-claims about the utility of the shaded region around the dashed curve
-in Figure [[what]] as a heuristic for determining whether a particular
-forecast method is well-matched to a particular task.  Of particular
-interest would be the class of so-called \emph{hybrid
-  systems}~\cite{hybrid}, which exhibit discrete transitions between
-different continuous regimes---e.g., a lathe that has an intermittent
-instability or traffic at an internet router, whose characteristic
-normal traffic patterns shift radically during an attack.  Effective
-modeling and prediction of these kinds of systems is quite difficult;
-doing so adaptively and automatically---in the manner that is alluded
-to at the end of the previous paragraph---would be an even more
-important challenge.
-
-% The way structure, information and complexity are processed by a
-% generating process play a crucial role in the success of a given
-% prediction scheme.
+work well in each new regime.  Of particular interest would be the
+class of so-called \emph{hybrid systems}~\cite{hybrid}, which exhibit
+discrete transitions between different continuous regimes---e.g., a
+lathe that has an intermittent instability or traffic at an internet
+router, whose characteristic normal traffic patterns shift radically
+during an attack.  Effective modeling and prediction of these kinds of
+systems is quite difficult; doing so adaptively and automatically---in
+the manner that is alluded to at the end of the previous
+paragraph---would be an even more important challenge.

### 2014-08-04 methods.tex
Liz final read through round I

@@
 instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
-example: a short segment of a performance trace of a short program
-that repeatedly initializes the upper triangle of a matrix in
+example: a short segment of a performance trace of a four-line C
+program that repeatedly initializes the upper triangle of a matrix in
 column-major order.
@@
       instructions per CPU clock cycle (IPC) during the execution of
-      \col, a simple C program that repeatedly initializes a matrix in
-      column-major order, running on an Intel
-      i7\textsuperscript{\textregistered}-based machine.  Each point
-      is the average IPC in a 100,000 instruction period.}
+      \col, a simple program that repeatedly initializes a matrix in
+      column-major order.  Each point is the average IPC in a 100,000
+      instruction period.}
    \label{fig:col-ipc}
@@
 the result of a single generating process with consistent complexity.
-\svd, in contrast, has multiple regimes, each the result of a
-different generating process.  To take advantage of this, we split the
-signal into six different segments, thereby obtaining an array of
-examples for the analyses in the following sections.  For notational
+\svd, in contrast, has multiple regimes, each the result of different
+generating processes.  To take advantage of this, we split the signal
+into six different segments, thereby obtaining an array of examples
+for the analyses in the following sections.  For notational
 convenience, we refer to these 90 time-series data sets\footnote{15

### 2014-08-04 model.tex
added mention of DVS fits
Liz final read through round I

@@
 \item The \emph{ARIMA} (auto-regressive integrated moving average)
-  method, a common linear forecast strategy, built using the
+  method, a common linear forecast strategy, instatiated via the
   \emph{\arima} procedure \cite{autoARIMA}, and
@@
 inadequate---that is, if an alternative method could do better and so
-one should continue searching.  \alert{need to check that this is not
-  redundant, given the other text in this section (and elsewhere in
-  the paper).  it may be better to put this in the conclusion.}
+one should continue searching.  \alert{Liz to do: check that this is
+  not redundant, given the other text in this section (and elsewhere
+  in the paper).  it may be better to put this in the conclusion.}
 
@@
 automated fitting techniques described in~\cite{autoARIMA} to
-accomplish this, producing what we will call an ``\arima model''in the
-rest of this paper.  This procedure uses sophisticated methods---KPSS
-unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
-test~\cite{Canova1995}, and the Akaike information
+accomplish this, producing what we will call an ``\arima model'' in
+the rest of this paper.  This procedure uses sophisticated
+methods---KPSS unit-root tests~\cite{KPSSunit}, a customization of the
+Canova-Hansen test~\cite{Canova1995}, and the Akaike information
 criterion~\cite{akaike1974}, conditioned on the maximum likelihood of
@@
 space of the given point, then taking that neighbor's path as the
-prediction---provides accurate forecasts when the generating process
-is a deterministic dynamical system.
+prediction---provides remarkably accurate forecasts when the
+generating process is a deterministic dynamical system.
 
@@
 perform poorly.  Figure~\ref{fig:embedding}(b), for instance, appears
-to contain little structure, so LMA may not work well on this signal.
-More structure appears to be present in Figure~\ref{fig:embedding}(c),
-but this reconstruction also appears to contain some noise.  The
-question as to how much structure is present in a reconstruction, and
-how much of that structure can be captured and used by LMA, is apropos
-of the central question treated in this paper.  It may be that \gcc
-has some redundancy that LMA cannot exploit, or that the structure in
-\svdfive is effectively obfuscated, from the standpoint of the LMA
-method, by noise.  Without any knowledge of the generating process,
-answers to these questions can only be derived from the data, with all
-of the attendant problems (noise, sampling issues, and so on).  By
-quantifying the balance between redundancy, predictive structure, and
-entropy for these real-valued time series---as shown in
+to contain little structure, so one might not expect LMA to work well
+on this signal.  More structure appears to be present in
+Figure~\ref{fig:embedding}(c), but this reconstruction also appears to
+contain some noise.  The question as to how much structure is present
+in a reconstruction, and how much of that structure can be captured
+and used by LMA, is apropos of the central question treated in this
+paper.  It may be that \gcc has some redundancy that LMA cannot
+exploit, or that the structure in \svdfive is effectively obfuscated,
+from the standpoint of the LMA method, by noise.
+
+Without any knowledge of the generating process, answers to these
+questions can only be derived from the data, with all of the attendant
+problems (noise, sampling issues, and so on).  By quantifying the
+balance between redundancy, predictive structure, and entropy for
+these real-valued time series---as shown in
 Section~\ref{sec:results}---we can begin to answer these questions in
@@
 Specifically, we split each time series into two pieces: the first
-90\%, referred to as the ``initial training" signal and denoted
+90\%, referred to here as the ``initial training" signal and denoted
 $\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
@@
   \end{table*}%
-
 In view of the discussion at the end of the previous paragraph, the
@@
 LMA predictions of \col and \svdfive both lie near the diagonal, for
-instance, but the corresponding MASE scores were very different:
+instance, but the corresponding MASE scores are very different:
 $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times

### 2014-08-04 pecomplexity.tex
removing an alert and fixing a HEAD issue
editing the ARIMA fit paragraph
friday round I
Liz final read through round I

@@
 %\\
-\alert{Quantifying Time-Series Predictability through Structural Complexity}
+% A model-free strategy for quantifying the predictability of
+% experimental time-series data 
+Model-free quantification of time-series predictability
+% \alert{Quantifying Time-Series Predictability through Structural Complexity}
 %\\
@@
 
-This paper provides insight into when, why, and how forecast
-strategies fail when they are applied to complicated time series. We
-conjecture that the inherent complexity of real-world time-series
+\alert{This paper provides insight into when, why, and how forecast
+  strategies fail when they are applied to complicated time series.}
+We conjecture that the inherent complexity of real-world time-series
 data---which results from the dimension, nonlinearity, and

### 2014-08-04 related.tex
Liz final read through round I

@@
 series~\cite{Smith199250}.  This approach can be used to show that
-different portions of a time series can exhibit varying levels of
-local predictive uncertainty.  We expand on this finding later in this
+different portions of a time series exhibit varying levels of
+local predictive uncertainty.  We expand on this idea later in this
 paper with a time series that exhibits interesting regime shifts.
@@
 building an autoregressive model, for example, it is common to
-calculate forecast errors and verify that they are normally distributed.
-If they are not, that suggests that there is structure in the time
-series that the model-building process was unable to capture and use.
-The problem with this approach is lack of generality.
+calculate forecast errors and verify that they are normally
+distributed.  If they are not, that suggests that there is structure
+in the time series that the model-building process was unable to
+recognize, capture, and exploit.  The problem with this approach is
+lack of generality.
 \label{page:normal-errors}
@@
 this class: we wish to measure the inherent complexity of an empirical
-time series, then correlate that complexity with the predictive
-accuracy of forecasts made using a number of different methods.
+time series, then study the correlation of that complexity with the
+predictive accuracy of forecasts made using a number of different
+methods.
 
@@
 for predictability bounds, this approach differs from our work in a
-number of fundamental ways.  First, Boffetta \emph{et al.} is a
-theoretical study; they do not work with any real predictions.  We
-focus on real-world time-series data, where one does not necessarily
-have the ability to perturb or otherwise interact with the system of
-interest, nor can one obtain or manufacture the (possibly large)
-number of points that might be needed to estimate the
-$\epsilon$-entropy for small $\epsilon$.  Second, we do not require
-\emph{a priori} knowledge about the noise and its interaction with the
-system.  Third, we tie information---in the form of the weighted
-permutation entropy---directly to prediction error via calculated
-values of a specific error metric.  Though FSLE and $\epsilon$-entropy
-allow for the comparison of predictability between systems, they do
-not directly provide an estimate of prediction error.  Finally, our
-approach also holds for stochastic systems, where neither the FLSEs
-nor their relationship to predictability is well defined.
+number of fundamental ways.  First, \cite{boffetta02} is a theoretical
+study that does not involve any actual predictions.  We focus on
+real-world time-series data, where one does not necessarily have the
+ability to perturb or otherwise interact with the system of interest,
+nor can one obtain or manufacture the (possibly large) number of
+points that might be needed to estimate the $\epsilon$-entropy for
+small $\epsilon$.  Second, we do not require \emph{a priori} knowledge
+about the noise and its interaction with the system.  Third, we tie
+information---in the form of the weighted permutation
+entropy---directly to prediction error via calculated values of a
+specific error metric.  Though FSLE and $\epsilon$-entropy allow for
+the comparison of predictability between systems, they do not directly
+provide an estimate of prediction error.  Finally, our approach also
+holds for stochastic systems, where neither the FLSEs nor their
+relationship to predictability is well defined.

### 2014-08-04 results.tex
Liz final read through round I

--- 2014-08-01 results.tex
@@
 
-\item The relationship between WPE and MASE is a useful empirical
-  heuristic for identifying mismatches between prediction models and
-  time-series data---i.e., when there is structure in the data that
-  the model is unable to exploit.
+\item The relationship between WPE and mean absolute scaled error
+  (MASE) is a useful empirical heuristic for identifying mismatches
+  between prediction models and time-series data---i.e., when there is
+  structure in the data that the model is unable to exploit.
 
@@
 Working from the first 90\% of each signal, we generated a prediction
-of the last 10\% using the random-walk, \naive, \arima, and LMA prediction methods,
-as described in Section~\ref{sec:accuracy}, then calculated the MASE
-value of those predictions.  We also calculated the WPE of each time
-series using a wordlength chosen via the procedure described in
-Section~\ref{sec:meaComplex}.  In order to assess the run-to-run
-variability of these results, we repeated all of these calculations on
-15 separate trials: i.e., 15 different runs of each program.
+of the last 10\% using the random-walk, \naive, \arima, and LMA
+prediction methods, as described in Section~\ref{sec:accuracy}, then
+calculated the MASE value of those predictions.  We also calculated
+the WPE of each time series using a wordlength chosen via the
+procedure described at the end of Section~\ref{sec:meaComplex}.  In
+order to assess the run-to-run variability of these results, we
+repeated all of these calculations on 15 separate trials: i.e., 15
+different runs of each program.
 
@@
 parameter space.  The meaning of the shaded region is explained in the
-text.}
+text. \alert{[[Do you think we should explain that here, rather than
+      punting to the text?  It is, after all, the primary claim of the
+      paper...]]}}
   \label{fig:wpe_vs_mase_best}
@@
 these reasons, we chose to fit a function of the form $y = a \log(b x
-+ 1)$ to these points, with $y =$ WPE and $x=$ MASE\footnote{The
-  specific values of the coefficiencts are $a=7.97 \times 10^{-2}$ and
-  $b=1.52 \times 10^3$.}.  The solid curve in the figure shows this
-fit; the dashed curves show the standard deviation of this model in
-its parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one
-standard deviation on each of the two parameters.  Points that fall
-within this deviation volume (light grey) correspond to predictions
-that are comparable to the best ones found in this study; points that
-fall \emph{above} that volume (dark grey) are better still.  We chose
-to truncate the shaded region because of a subtle point regarding the
-MASE of an ideal predictor, which should not be larger than 1 unless
-the training and test signals are different.  This is discussed at
-more length below.
++ 1)$ to these points\footnote{The specific values of the coefficients
+  are $a=7.97 \times 10^{-2}$ and $b=1.52 \times 10^3$.}, with $y =$
+WPE and $x=$ MASE.  The solid curve in the figure shows this fit; the
+dashed curves show the standard deviation of this model in its
+parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
+deviation on each of the two parameters.  Points that fall within this
+deviation volume (light grey) \alert{[[should we consider
+      crosshatching or something else, given the printing
+      wierdness?]]} correspond to predictions that are comparable to
+the best ones found in this study; points that fall \emph{above} that
+volume (dark grey) are better still.  We chose to truncate the shaded
+region because of a subtle point regarding the MASE of an ideal
+predictor, which should not be larger than 1 unless the training and
+test signals are different.  This is discussed at more length below.
 
@@
 The mean MASE scores of predictions generated with the nonlinear LMA
-method, for instance, are generally closer to the dashed curve; the
-\arima method clusters are more widely spread, the \naive ~clusters
-even more so.  A few of the clusters have very high variance; these
-are discussed later in this section.
+method are generally closer to the dashed curve; the \arima method
+clusters are more widely spread, the \naive ~clusters even more so.  A
+few of the clusters have very high variance; these are discussed later
+in this section.
 
@@
 region.  \emph{These predictions are not as good as our heuristic
-  suggests they could be.}  Focusing in on any single cluster of
-points makes this clear: LMA works best for \svdsix, for instance,
-followed by the random-walk prediction method, then \arima and \naive.
-When a practitioner calculates an WPE vs. MASE value that is outside the
-shaded region, that should suggest to her that the prediction method
-is not well matched to the task at hand: that is, the time series has
-more predictive structure than the method is able to use.
+  suggests they could be.}  Focusing in on any single signal makes
+this clear: LMA works best for \svdsix, for instance, followed by the
+random-walk prediction method, then \arima and \naive.  This provides
+some practical leverage: if one calculates an WPE vs. MASE value that
+is outside the shaded region, that should suggest that the prediction
+method is not well matched to the task at hand: that is, the time
+series has more predictive structure than the method is able to use.
 %
@@
 %
-In this case---e.g., \arima on \svdsix---she should try a different
-method.  The position of the LMA cluster for \svdsix, on the other
-hand, reflects this method's ability to capture and exploit the
+In the case of \arima on \svdsix, for instance, one should try a
+different method.  The position of the LMA cluster for \svdsix, on the
+other hand, reflects this method's ability to capture and exploit the
 structure that is present in this signal.  WPE vs. MASE values like
@@
 0.211$, respectively, across all 15 trials.  That is, \naive ~and
-\arima are performing only $\approx 1.7$ times better than the
-random-walk method, a primitive strategy that simply uses the current
-value as the prediction.  However, the WPE value for the \col trials
-is $0.513 \pm 0.003$, which is in the center of the complexity
-spectrum described in Section~\ref{sec:intro}.
+\arima perform only $\approx 1.7$ times better than the random-walk
+method, a primitive strategy that simply uses the current value as the
+prediction.  However, the WPE value for the \col trials is $0.513 \pm
+0.003$, a signal that is in the center of the complexity spectrum
+described in Section~\ref{sec:intro}.
 
@@
 information transfer in the signal, but predictions with comparatively
-poor MASE scores---is obvious in the geometry of the two left-hand
-and the upper-right hand images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
+poor MASE scores---is obvious in the geometry of three of the four
+images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
 are far to the right of and/or below the dashed curve.  Again, this
@@
 (random walk), or one that uses a linear model (\arima)---the MASE
-score is much better: $0.050 \pm 0.001$.  This prediction is 20 times
-more accurate than a random-walk forecast, which is more in line with
-the level of predictive structure that the low WPE value suggests is
-present in the signal.  The MASE scores of random-walk predictions of
-\col are all $\approx 1$---as one would expect---pushing those points
-well below the shaded region.  Clearly the stationarity assumption on
-which this method is based does not hold for this signal.
+score is much improved: $0.050 \pm 0.001$.  This prediction is 20
+times more accurate than a random-walk forecast, which is more in line
+with the level of predictive structure that the low WPE value suggests
+is present in the \col signal.  The MASE scores of random-walk
+predictions of this signal are all $\approx 1$---as one would
+expect---pushing those points well below the shaded region.  Clearly
+the stationarity assumption on which this method is based does not
+hold for this signal.
 
-Note that the \col points (blue {\color{blue}$+$} icons) are clustered
-very tightly in the lower left quadrant of the \naive, random-walk,
-and LMA plots in Figure~\ref{fig:wpe_vs_mase_all}, but spread out
-horizontally in the \arima plot.  This is because of the way the
-\arima process builds ARIMA models \cite{autoARIMA}.  If a KPSS test
-of the time series in question indicates that it is nonstationary, the
-\arima recipe adds an integration term to the model.  This test gives
-mixed results in the case of the \col process, flagging five of the 15
-trials as stationary and ten as nonstationary.  ARIMA models without
-an integration term perform more poorly on this signal, which
-increases the error, thereby spreading out the points.  We tested this
-hypothesis by forcing the inclusion of an integration term in the five
-cases where a KPSS test indicated that such a term was not needed.
-This action removed the spread, pushing all 15 of the \col ~ \arima
-points in Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.
+The \col example brings out some of the shortcomings of automated
+model-building processes.  Note that the {\color{blue}$+$} points are
+clustered very tightly in the lower left quadrant of the \naive,
+random-walk, and LMA plots in Figure~\ref{fig:wpe_vs_mase_all}, but
+spread out horizontally in the \arima plot.  This is because of the
+way the \arima process builds ARIMA models \cite{autoARIMA}.  If a
+KPSS test of the time series in question indicates that it is
+nonstationary, the \arima recipe adds an integration term to the
+model.  This test gives mixed results in the case of the \col process,
+flagging five of the 15 trials as stationary and ten as nonstationary.
+We conjectured that ARIMA models without an integration term perform
+more poorly on these five signals, which increases the error and
+thereby spreads out the points.  We tested this hypothesis by forcing
+the inclusion of an integration term in the five cases where a KPSS
+test indicated that such a term was not needed.  This action removed
+the spread, pushing all 15 of the \col ~ \arima points in
+Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.
 %\alert{Josh to do: this paragraph needs more, in response to the %ref's
@@
 the tests and criteria employed by this algorithm
-(Section~\ref{sec:arima}) are sophisticated , the results can still be
+(Section~\ref{sec:arima}) are sophisticated, the results can still be
 sub-optimal---if the initial space of models being searched is not
@@
 are $20.870 \pm 0.192$ and $2.370 \pm 0.051$, respectively: that is,
-20.87 and 2.37 times worse than a simple random walk forecast of the
-same signals\footnote{The \naive ~MASE score is large because of the
-  bimodal nature of the distribution of the values of the signal,
-  which makes guessing the mean a particularly bad strategy.  The same
-  thing is true of the \svdthree signal.}.  As before, the positions
-of these points on a WPE vs. MASE plot---significantly below and to
-the right of the shaded region---should suggest to a practitioner that
+20.87 and 2.37 times worse than a simple random-walk forecast of the
+training set portions of the same signals\footnote{The \naive ~MASE
+  score is large because of the bimodal nature of the distribution of
+  the values of the signal, which makes guessing the mean a
+  particularly bad strategy.  The same thing is true of the \svdthree
+  signal.}.  As before, the positions of these points on a WPE
+vs. MASE plot---significantly below and to the right of the shaded
+region---should suggest to a practitioner that
 %  there is more structure in this signal than the corresponding
@@
 structure to utilize, methods that depend on exploiting that
-structure---like ARIMA and LMA---will be outperformed by methods that
-do not rely on that structure.
+structure---like ARIMA and LMA---cannot get any traction on those
+signals.
 %
@@
 is due to nonstationarity in the signal: in particular, differences
-between the training and test signals.  \svdtwo fits the same
-patterns, for the same reasons---and visibly so, in the red segment of
-Figure~\ref{fig:svd-ts-colored}, where the period and amplitude of the
-oscillations are decreasing.
+between the training and test signals.  The same effect is at work in
+the \svdtwo results, for the same reasons---and visibly so, in the red
+segment of Figure~\ref{fig:svd-ts-colored}, where the period and
+amplitude of the oscillations are decreasing.
 
@@
 0.6, which is better than the other three methods.  In five of the 15
-\svdone trials, however, there were step changes in the signal.  The
-\naive ~method has a very difficult time with signals like this,
-particularly if there are multiple step changes.  This raised the MASE
-scores of these trials, pushing the corresponding points to the
-right\footnote{This includes the cluster of three points near MASE
-  $\approx 2.5$, as well as two points that are beyond the domain of
-  the graph, at MASE $\approx 11.2-14.8$.}, and in turn raising both
-the mean and variance of this set of trials.
+\svdone trials, however, there were step changes in the signal.
+\alert{[[Seems like same thing as we pointed out and discussed in
+      \col.  We should either point out the correspondence or explain
+      why it's different.]]}  The \naive ~method has a very difficult
+time with signals like this, particularly if there are multiple step
+changes.  This raised the MASE scores of these trials, pushing the
+corresponding points to the right\footnote{This includes the cluster
+  of three points near MASE $\approx 2.5$, as well as two points that
+  are beyond the domain of the graph, at MASE $\approx 11.2-14.8$.},
+and in turn raising both the mean and variance of this set of trials.
 
@@
 by the way MASE is calculated.  Recall that MASE scores are scaled
-\emph{relative to a random-walk forecast of the training set}.  There
-are two issues here.  First, random-walk prediction works very badly
-on signals with frequent, large, rapid transitions.  Consider a signal
-that oscillates from one end of its range to the other at every step.
-A signal like this will have a low WPE, much like {\tt col\_major}.
-However, a random-walk forecast of this signal will be 180 degrees out
-of phase with the true continuation.  Since random-walk error appears
-in the denominator of the MASE score, this effect can shift points
-leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
-clusters in Figure~\ref{fig:wpe_vs_mase_all} are above the dashed
-curve.  This time series, which is shown in closeup in
+\emph{relative to a random-walk forecast}.  There are two issues here.
+First, random-walk prediction works very badly on signals with
+frequent, large, rapid transitions.  Consider a signal that oscillates
+from one end of its range to the other at every step.  A signal like
+this will have a low WPE, much like {\tt col\_major}.  However, a
+random-walk forecast of this signal will be 180 degrees out of phase
+with the true continuation.  Since random-walk error appears in the
+denominator of the MASE score, this effect can shift points leftwards
+on a WPE vs. MASE plot, and that is exactly why the \svdone clusters
+in Figure~\ref{fig:wpe_vs_mase_all} are above the dashed curve.  This
+time series, part of which is shown in closeup in
 Figure~\ref{fig:svdone-ts},
@@
 from 1, this could be an indication that the signal is nonstationary.
-We are in the process of exploring this idea.
+We are in the process of exploring this idea and its relationship to
+the DVS fits mentioned in the fourth paragraph of
+Section~\ref{sec:related}.
 
@@
 
-Of course, the geometry of these curves and bounds do not necessarily
-extend to other error metrics.  The positions of the WPE vs. error
-points---and the reasons to choose a particular function to fit to
-them or impute theoretical bounds on the meaning of the results---may
-change if one calculates error using a procedure other than the one
-described in Section~\ref{sec:accuracy}.  The MASE error metric, as
-discussed above, has its weaknesses.  Even so, we concur with
-\cite{MASE} that it is an effective way to compare prediction error
-across time-series data of different lengths and different scales, and
-we are convinced by the extensive evaluations that are offered in that
-paper, as well as the comparisons of MASE to other metrics, that its
-strengths far outweigh its weaknesses.
+Of course, the geometry of the curves and bounds in these figures do
+not necessarily extend to other error metrics.  The positions of the
+WPE vs. error points---and the reasons to choose a particular function
+to fit to them or impute theoretical bounds on the meaning of the
+results---may change if one calculates error using a procedure other
+than the one described in Section~\ref{sec:accuracy}.  The MASE error
+metric, as discussed above, has its weaknesses.  Even so, we concur
+with \cite{MASE} that it is an effective way to compare prediction
+error across time-series data of different lengths and different
+scales, and we are convinced by the extensive evaluations that are
+offered in that paper, as well as the comparisons of MASE to other
+metrics, that its strengths far outweigh its weaknesses.
 

### 2014-08-04 wpe.tex
pushing changes midstream, in case of lossage
Liz final read through round I

@@
 Estimating the entropy of an arbitrary, real-valued time series is a
-challenging problem.  Our approach to this problem, which is the topic
-of this section, draws upon methods and results from a variety of
-fields including time-series analysis, dynamical systems, and
-stochastic processes.
+significant challenge.  Our approach to this problem, which is the topic of
+this section, draws upon methods and results from a variety of fields
+including time-series analysis, dynamical systems, and stochastic
+processes.
 
@@
 the past (aka ``the conditional entropy of the present given the
-past'').  The first part is known as the \emph{redundancy}, as it is
+past'').  The first part is known as the \emph{redundancy}:
 information in the present that is also in the past.  The second part
 is the aforementioned Shannon entropy rate.  It seems obvious that the
-more redundancy in a signal, the more predictable it should be.
-Redundancy is critical to prediction, in general.  And the specific
-\emph{form} of the redundancy should dictate whether a particular
-prediction method will work well or poorly on the corresponding
-signal.  A linear method cannot capture or make use of nonlinear
-redundancy, for instance.  This issue---which has been explored
-extensively using clean data and/or systems where the generating
-process is known, but not so much with empirical data from unknown
-systems---is central to the claims in this paper and the discussion in
-the following section.
+more redundancy in a signal, the more predictable it should be.  And
+the specific \emph{form} of the redundancy should dictate whether a
+particular prediction method will work well or poorly on the
+corresponding signal.  A linear method cannot capture or make use of
+nonlinear redundancy, for instance.  This issue---which has been
+explored extensively using clean data and/or systems where the
+generating process is known, but not so much with empirical data from
+unknown systems---is central to the claims in this paper and the
+discussion in the following section.
 
@@
 technique, the \emph{weighted permutation entropy} (WPE), which was
-introduced in~\cite{fadlallah2013}.  The intent behind the weighting
-is to correct for observational noise that is larger than the trends
-in the data, but smaller than the larger-scale features.  Consider,
-for example, a signal that switches between two fixed points and
-contains some additive noise. The PE of such a signal will be
-dominated by the noise about the fixed points, driving it to $\approx
-1$---which in some sense hides the fact that the signal is actually
-quite structured.  In this situation, the terms of the weighted
-permutation entropy are dominated by the switching rather than the
-stochastic fluctuations. To accomplish this, the \emph{weight} of a
-permutation is taken into account:
+introduced in~\cite{fadlallah2013}.  \alert{We never actually say WHY
+  we use it.}  The intent behind the weighting is to correct for
+observational noise that is larger than the trends in the data, but
+smaller than the larger-scale features.  Consider, for example, a
+signal that switches between two fixed points and contains some
+additive noise. The PE of such a signal will be dominated by the noise
+about the fixed points, driving it to $\approx 1$---which in some
+sense hides the fact that the signal is actually quite structured.  In
+this situation, the terms of the weighted permutation entropy are
+dominated by the switching rather than the stochastic fluctuations. To
+accomplish this, the \emph{weight} of a permutation is taken into
+account:
 \begin{align*}
@@
 result converges, but data length issues can intrude before that
-convergence is reached.  We used this approach, which represents a
-good balance between accurate ordinal statistics and finite-data
-effects, to choose $\ell$ values for the experiments in the following
-section.
+convergence is reached.  We used this approach, which we believe
+strikes a good balance between accurate ordinal statistics and
+finite-data effects, to choose $\ell$ values for the experiments in
+the following section.
 

### 2014-08-04 model.tex
Liz final read-through round II

@@
 
-These four forecast methods were chosen to be a representative
-sampling of standard prediction strategies, but they do not, of
-course, cover that space exhaustively.  Our goal here is an empirical
-assessment of the relationship between predictability and complexity,
-not formal results about a ``best'' predictor for a given time series.
-From a practitioner's standpoint, it would be useful to know \emph{ a
-  priori} if a time series contains enough predictive structure to
-make it worth spending the time and effort searching for a good
-forecast method.  It would also be useful to know if a given method is
-inadequate---that is, if an alternative method could do better and so
-one should continue searching.  \alert{Liz to do: check that this is
-  not redundant, given the other text in this section (and elsewhere
-  in the paper).  it may be better to put this in the conclusion.}
+% These four forecast methods were chosen to be a representative
+% sampling of standard prediction strategies, but they do not, of
+% course, cover that space exhaustively.  Our goal here is an empirical
+% assessment of the relationship between predictability and complexity,
+% not formal results about a ``best'' predictor for a given time series.
+% From a practitioner's standpoint, it would be useful to know \emph{ a
+%   priori} if a time series contains enough predictive structure to
+% make it worth spending the time and effort searching for a good
+% forecast method.  It would also be useful to know if a given method is
+% inadequate---that is, if an alternative method could do better and so
+% one should continue searching.  \alert{Liz to do: check that this is
+%   not redundant, given the other text in this section (and elsewhere
+%   in the paper).  it may be better to put this in the conclusion.}
 

### 2014-08-04 results.tex
Liz final read-through round II

@@
 of inappropriateness in the order selection and parameter estimation
-procedure.  A close examination of the \arima MASE scores and WPE
-values of the 15 \col traces, for instance, brought out such a
-situation: all of these signals had the same complexity, but some (the
-outliers in the blue cluster of {\color{blue}$+$} points in the
-bottom-left panel of Figure~\ref{fig:wpe_vs_mase_all}) had much higher
-MASE values, indicating a failure in the \arima process.  We then went
-back to the data, noticed some shifts of the baseline in those
-signals, and manually assisted the automated fitting procedure to
-force the inclusion of a term that is appropriate for a nonstationary
-signal.  In all cases, this produced ARIMA models with MASE scores in
-line with the others in that cluster.
+procedure.
+
+% A close examination of the \arima MASE scores and WPE values of the 15
+% \col traces, for instance, brought out such a situation: all of these
+% signals had the same complexity, but some (the outliers in the blue
+% cluster of {\color{blue}$+$} points in the bottom-left panel of
+% Figure~\ref{fig:wpe_vs_mase_all}) had much higher MASE values,
+% indicating a failure in the \arima process.  We then went back to the
+% data, noticed some shifts of the baseline in those signals, and
+% manually assisted the automated fitting procedure to force the
+% inclusion of a term that is appropriate for a nonstationary signal.
+% In all cases, this produced ARIMA models with MASE scores in line with
+% the others in that cluster.
 
@@
 
-Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
-and~\ref{fig:wpe_vs_mase_all} were determined from a finite set of
+The curves in Figures~\ref{fig:wpe_vs_mase_best}
+and~\ref{fig:wpe_vs_mase_all} were determined from finite sets of
 methods and data.  We put a lot of thought and effort into making
-these experiments upon which these results are based both
-representative and comprehensive.  The forecast methods involved range
-from the simple to the sophisticated; the time-series data was sampled
-from a system whose behavior spans the dynamical behavior space.
-While we are cautiously optimistic about the generality of our
-conclusions, more exploration will be required before we can make any
-such claim.  Our preliminary work along those lines shows that data
-from the Henon map \cite{henon}, the Lorenz system \cite{lorenz}, and
-the SFI A data set \cite{weigend-book,sfi-data} all fall within the
-one-$\sigma$ volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
-and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
-transformations of \svdtwo, \svdfive and \svdsix.
+these sets representative and comprehensive.  The forecast methods
+involved range from the simple to the sophisticated; the time-series
+data analyzed in this section was sampled from a system whose behavior
+spans the dynamical behavior space.  While we are cautiously
+optimistic about the generality of our conclusions, more exploration
+will be required before we can make any such claim.  Our preliminary
+work along those lines shows that data from the Henon map
+\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set
+\cite{weigend-book,sfi-data}, and a white-noise process all fall
+within the one-$\sigma$ volume of the fit in
+Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
+region, as do various nonlinear transformations of \svdtwo, \svdfive
+and \svdsix.
 

### 2014-08-04 response.tex
alerts in response

@@
 of discussion of this cluster of issues in [Hyndman and Koehler, 2006]
-\alert{(our reference [21])}.  If the reviewer thinks that it would
-strengthen our paper, we can recapitulate some of that material in our
-Section IVD.
+\alert{[[(our reference [21])]]}.  If the reviewer thinks that it
+would strengthen our paper, we can recapitulate some of that material
+in our Section IVD.
 
@@
 the last paragraph in Section VI.  As mentioned there, the paper in
-which MASE was introduced---\alert{our reference [21]}---includes an
-extensive evaluation of this metric, as well as a comprehensive
+which MASE was introduced---\alert{[[our reference [21]]]}---includes
+an extensive evaluation of this metric, as well as a comprehensive
 comparison to other metrics.  If the reviewer thinks that it would
@@
 and $\Delta t=0.01$, a 10,000-step random-walk time series generated
-by taking the cumulative sum of $\mu=0$, $\sigma=1$ Gaussian random
-variables, and data set ``A'' from the Santa Fe Institute time-series
-competition website.
+by taking the cumulative sum of $\mu=0$, $\sigma=1$ \alert{[[generated
+      how?  Gaussian random variables,]]} and data set ``A'' from the
+Santa Fe Institute time-series competition website.
 

### 2014-08-04 response.tex
minor changes to response

@@
 way, and to test it in the context of empirical data.  We have also
-changed the title to line up with this goal.
+changed the title and abstract to line up with this goal.
 
@@
 preliminary results of our broader study, currently underway, indicate
-that they also hold for data from synthetic examples (the H\'{e}non
-map, the Lorenz equations, and the output of a purely stochastic
-random-walk process) and for real-world data from a different system
-(SFI data set A).  The image below shows these results:
+that they also hold for data from synthetic examples and for
+real-world data from a different system.  The image below shows some
+of these results:
 
@@
 two different systems (laser, computer), and nonlinear transformations
-of that real-world data---again, supporting the generality of our
+of the computer-performance data---again, supporting the generality of our
 heuristic.

### 2014-08-04 concl.tex
adding some comments

@@
 the nonlinear dynamics of other systems.  Our preliminary results on
-other systems (Henon, Lorenz, white noise, SFI ``A'', transformations
+other systems (H\'enon, Lorenz, white noise, SFI ``A'', transformations
 of the computer performance data) lead us to believe that our results
@@
 
-\alert{I'm not sure this paragraph adds anything.  Can/should we
+{\color{blue}agreed, I don't think it addds anything}\alert{I'm not sure this paragraph adds anything.  Can/should we
   remove it?  Given that information is a fundamental limitation in

### 2014-08-04 related.tex
adding some comments

@@
 theoretical assessment or comparison of existing methods.  Our goals
-are focused more on the \emph{practice} of prediction: \emph{(i)} to
+are focused more on the \emph{practice} of prediction: \emph{(i)} \alert{to
 empirically quantify the predictive structure that is present in a
-real-valued scalar time series and \emph{(ii)} to explore how the
+real-valued scalar time series} and \emph{(ii)} to explore how the
 performance of prediction methods is related to that inherent

### 2014-08-04 results.tex
adding some comments

@@
 will be required before we can make any such claim.  Our preliminary
-work along those lines shows that data from the Henon map
-\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set
-\cite{weigend-book,sfi-data}, and a white-noise process all fall
+work along those lines shows that data from the H\'enon map
+\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set\footnote{\alert{This dataset can be retrieved from \url{http://www-psych.stanford.edu/~andreas/Time-Series/SantaFe.html}.}}
+\cite{weigend-book,sfi-data},\alert{maybe make the website a footnote instead} and a random-walk process all fall
 within the one-$\sigma$ volume of the fit in

### 2014-08-04 response.tex
add description of random-walk generation

@@
 solution of the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$
-and $\Delta t=0.01$, a 10,000-step random-walk time series generated
-by taking the cumulative sum of $\mu=0$, $\sigma=1$ \alert{[[generated
-      how?  Gaussian random variables,]]} and data set ``A'' from the
-Santa Fe Institute time-series competition website.
+and $\Delta t=0.01$, a 10,000-step random-walk time series generated by
+taking the cumulative sum of $\mu=0$, $\sigma=1$ Gaussian random
+variables using Numpy ({\tt numpy.cumsum(numpy.random.normal(0.0, 1.0, size=10000))}),
+and data set ``A'' from the Santa Fe Institute time-series competition
+website.
 

### 2014-08-04 intro.tex
few more comments

@@
 predictive structure, i.e, signals whose future values can be
-successfully predicted from past values.  Signals like this can be
+successfully\alert{perfectly?} predicted from past values.  Signals like this can be
 viewed as the product of an underlying process that generates
@@
 future, regardless of what model one chooses to use. Signals in the
-midrange of this spectrum---e.g., deterministic chaos---pose
-interesting challenges from a modeling perspective.  In these signals,
-enough information is being transmitted from the past to the future
-that an \emph{ideal} model---one that captures the generating
-process---can forecast the future behavior of the observed system with
-high accuracy.
+midrange of this spectrum, e.g., deterministic chaos, pose interesting
+challenges from a modeling perspective.  In these signals, enough
+information is being transmitted from the past to the future that an
+\emph{ideal} model---one that captures the generating process---can
+forecast the future behavior of the observed system with high
+accuracy.
 
@@
 a forecast model that can leverage the information (if any) that is
-being transmitted from past to future by the underlying generating
+being transmitted forward in time by the underlying generating
 process?  A first step in answering this question is to reliably
@@
 properties: whether it is linear, nonlinear, deterministic,
-stochastic, stationary, non-stationary, etc.  To explore the
-relationship between complexity, predictive structure, and
-predictability, we generate forecasts for a variety of experimental
-time-series datasets using four different prediction methods, then
-compare the accuracy of those predictions to the permutation entropy
-of the associated signals.  This results in two primary findings:
+stochastic, etc.  To explore the relationship between complexity,
+predictive structure, and actual predictability, we generate forecasts
+for a variety of experimental time-series datasets using four
+different prediction methods, then compare the accuracy of those
+predictions to the permutation entropy of the associated signals.
+This results in two primary findings:
 % 

### 2014-08-04 concl.tex
rest

@@
 
-{\color{blue}agreed, I don't think it addds anything}\alert{I'm not sure this paragraph adds anything.  Can/should we
-  remove it?  Given that information is a fundamental limitation in
-  predictability, then gathering and using more information is an
-  obvious next step.  But there is an equally obvious tension here
-  between data length and prediction speed: a forecast that requires
-  half a second to compute is not useful for the purposes of real-time
-  control of a computer system with a MHz clock rate, for instance.
-  Another alternative is to sample several system variables
-  simultaneously and build multivariate models.  This is a particular
-  challenge in nonlinear LMA-type models, since multivariate
-  delay-coordinate embedding (e.g.,
-  \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
-  computationally prohibitive.  We are working on alternative methods
-  that sidestep that complexity.}
+% {\color{blue}agreed, I don't think it addds anything}\alert{I'm not
+%   sure this paragraph adds anything.  Can/should we remove it?  Given
+%   that information is a fundamental limitation in predictability, then
+%   gathering and using more information is an obvious next step.  But
+%   there is an equally obvious tension here between data length and
+%   prediction speed: a forecast that requires half a second to compute
+%   is not useful for the purposes of real-time control of a computer
+%   system with a MHz clock rate, for instance.  Another alternative is
+%   to sample several system variables simultaneously and build
+%   multivariate models.  This is a particular challenge in nonlinear
+%   LMA-type models, since multivariate delay-coordinate embedding
+%   (e.g., \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
+%   computationally prohibitive.  We are working on alternative methods
+%   that sidestep that complexity.}
+%
 

### 2014-08-04 results.tex
rest

--- 2014-08-04 results.tex
@@
 The dashed curves reflect the standard deviation of the model in its
-parameter space.  The meaning of the shaded region is explained in the
-text. \alert{[[Do you think we should explain that here, rather than
-      punting to the text?  It is, after all, the primary claim of the
-      paper...]]}}
+parameter space.  Points that lie below and to the right of the shaded
+region indicate that the time series has more predictive structure
+than the forecast strategy is able to utilize.}
   \label{fig:wpe_vs_mase_best}
@@
 deviation on each of the two parameters.  Points that fall within this
-deviation volume (light grey) \alert{[[should we consider
-      crosshatching or something else, given the printing
-      wierdness?]]} correspond to predictions that are comparable to
-the best ones found in this study; points that fall \emph{above} that
-volume (dark grey) are better still.  We chose to truncate the shaded
-region because of a subtle point regarding the MASE of an ideal
-predictor, which should not be larger than 1 unless the training and
-test signals are different.  This is discussed at more length below.
+deviation volume (light grey) correspond to predictions that are
+comparable to the best ones found in this study; points that fall
+\emph{above} that volume (dark grey) are better still.  We chose to
+truncate the shaded region because of a subtle point regarding the
+MASE of an ideal predictor, which should not be larger than 1 unless
+the training and test signals are different.  This is discussed at
+more length below.
 
@@
 0.6, which is better than the other three methods.  In five of the 15
-\svdone trials, however, there were step changes in the signal.
-\alert{[[Seems like same thing as we pointed out and discussed in
-      \col.  We should either point out the correspondence or explain
-      why it's different.]]}  The \naive ~method has a very difficult
+\svdone trials, however, there were step changes in the signal.  This
+is a different nonstationarity than in the case of \col---large jump
+discontinuities rather than small shifts in the baseline---and not one
+that we were able to handle by simply forcing the ARIMA model to
+include a particular term.  The \naive ~method has a very difficult
 time with signals like this, particularly if there are multiple step
@@
 will be required before we can make any such claim.  Our preliminary
-work along those lines shows that data from the Henon map
-\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set
-\cite{weigend-book,sfi-data}, and a white-noise process all fall
-within the one-$\sigma$ volume of the fit in
-Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
-region, as do various nonlinear transformations of \svdtwo, \svdfive
-and \svdsix.
+work along those lines shows that data from the H\'enon map
+\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data
+set\footnote{This dataset can be retrieved from
+  \url{http://www-psych.stanford.edu/~andreas/Time-Series/SantaFe.html}.}
+\cite{weigend-book}, and a random-walk process all fall within the
+one-$\sigma$ volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
+and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
+transformations of \svdtwo, \svdfive and \svdsix.
 

### 2014-08-04 wpe.tex
rest

@@
 technique, the \emph{weighted permutation entropy} (WPE), which was
-introduced in~\cite{fadlallah2013}.  \alert{We never actually say WHY
-  we use it.}  The intent behind the weighting is to correct for
-observational noise that is larger than the trends in the data, but
-smaller than the larger-scale features.  Consider, for example, a
-signal that switches between two fixed points and contains some
-additive noise. The PE of such a signal will be dominated by the noise
-about the fixed points, driving it to $\approx 1$---which in some
-sense hides the fact that the signal is actually quite structured.  In
-this situation, the terms of the weighted permutation entropy are
-dominated by the switching rather than the stochastic fluctuations. To
-accomplish this, the \emph{weight} of a permutation is taken into
-account:
+introduced in~\cite{fadlallah2013}.  The intent behind the weighting
+is to correct for observational noise that is larger than the trends
+in the data, but smaller than the larger-scale features.  Consider,
+for example, a signal that switches between two fixed points and
+contains some additive noise. The PE of such a signal will be
+dominated by the noise about the fixed points, driving it to $\approx
+1$---which in some sense hides the fact that the signal is actually
+quite structured.  In this situation, the terms of the weighted
+permutation entropy are dominated by the switching rather than the
+stochastic fluctuations.  The amplitude of the switches plays a key
+role in predictability, so weighting the scale of ordinal changes is
+important for quantifying predictive structure accurately.  To
+accomplish this, WPE takes into account the \emph{weight} of a
+permutation:
 \begin{align*}

### 2014-08-04 response.tex
done?

--- 2014-08-04 response.tex
@@
 of discussion of this cluster of issues in [Hyndman and Koehler, 2006]
-\alert{[[(our reference [21])]]}.  If the reviewer thinks that it
-would strengthen our paper, we can recapitulate some of that material
-in our Section IVD.
+(our reference [40]).  If the reviewer thinks that it would strengthen
+our paper, we can recapitulate some of that material in our Section
+IVD.
 
@@
 the last paragraph in Section VI.  As mentioned there, the paper in
-which MASE was introduced---\alert{[[our reference [21]]]}---includes
-an extensive evaluation of this metric, as well as a comprehensive
-comparison to other metrics.  If the reviewer thinks that it would
-strengthen our paper, we can recapitulate more of that material in our
-paper.  As part of our current work on this topic, we are evaluating
-how other choices of metric affect the shape of our curves, but we
-feel that a comprehensive study of this issue is best left as a sequel
-to this work.
+which MASE was introduced---our reference [40]---includes an extensive
+evaluation of this metric, as well as a comprehensive comparison to
+other metrics.  If the reviewer thinks that it would strengthen our
+paper, we can recapitulate more of that material in our paper.  As
+part of our current work on this topic, we are evaluating how other
+choices of metric affect the shape of our curves, but we feel that a
+comprehensive study of this issue is best left as a sequel to this
+work.
 
@@
 and $\Delta t=0.01$, a 10,000-step random-walk time series generated
-by taking the cumulative sum of $\mu=0$, $\sigma=1$ \alert{[[generated
-      how?  Gaussian random variables,]]} and data set ``A'' from the
-Santa Fe Institute time-series competition website.
+by taking the cumulative sum of $\mu=0$, $\sigma=1$ Gaussian random
+variables using Python's {\tt numpy} library\footnote{{\tt
+    numpy.cumsum(numpy.random.normal(0.0, 1.0, size=10000))}}, and
+  data set ``A'' from the Santa Fe Institute time-series competition
+  website.
 

### 2014-08-04 concl.tex
conclusion hacking

@@
 % forecast methods, and time-series data from other systems.
+
+\alert{add FW stuff from section VI}
 
@@
 
-In this paper, we have argued that redundancy~\cite{crutchfield2003}
-is a useful definition of the inherent predictability of an empirical
-time series.  To operationalize that definition, we use an
-approximation of the Kolmogorov-Sinai entropy~\cite{lind95}, estimated
-using a weighted version of the permutation entropy
-of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation
-of forward information transfer in a time series---is ideal for our
-purposes because it works with real-valued data and is known to
-converge to the true entropy value. Using a variety of forecast models
-and more than 150 time-series data sets from experiments and
-simulations, we have shown that prediction accuracy is indeed
-correlated with weighted permutation entropy: the higher the WPE, in
-general, the higher the prediction error.  The relationship is roughly
-logarithmic, which makes theoretical sense, given the nature of WPE,
-predictability, and MASE.
+In this paper, we have argued that redundancy is a useful definition
+of the inherent predictability of an empirical time series.  To
+operationalize that definition, we use an approximation of the
+Kolmogorov-Sinai entropy, estimated using a weighted version of the
+permutation entropy of~\cite{bandt2002per}.  This WPE technique---an
+ordinal calculation of forward information transfer in a time
+series---is ideal for our purposes because it works with real-valued
+data and is known to converge to the true entropy value. Using a
+variety of forecast models and more than 150 time-series data sets
+from experiments and simulations, we have shown that prediction
+accuracy is indeed correlated with weighted permutation entropy: the
+higher the WPE, in general, the higher the prediction error.  The
+relationship is roughly logarithmic, which makes theoretical sense,
+given the nature of WPE, predictability, and MASE.
 
@@
 the nonlinear dynamics of other systems.  Our preliminary results on
-other systems (H\'enon, Lorenz, white noise, SFI ``A'', transformations
-of the computer performance data) lead us to believe that our results
-will generalize beyond the examples described in this paper.  We are
-in the process of following up on that exploration with a broader
-study of data, forecast methods, and error metrics.
+other systems (H\'enon, Lorenz, a white-noise process, SFI ``A'',
+transformations of the computer performance data) lead us to believe
+that our results will generalize beyond the examples described in this
+paper.  We are in the process of following up on that exploration with
+a broader study of data, forecast methods, and error metrics.
 
@@
 Nonstationarity is a serious challenge in any time-series modeling
-problem.  [[Can destroy our heuristic: regime shifts can cause the
-    test signal to look different from the training signal, which can
-    skew the MASE score (and the points on the curve).  A RW MASE
-    score larger than 1, however, as mentioned at end of prev section,
-    can flag when this is happening.]]  Detecting regime shifts---and
-adapting prediction models accordingly---is an important area of
-future work.  Indeed, one of the first applications of permutation
-entropy was to recognize the regime shift in brainwave data that
-occurs when someone has a seizure~\cite{cao2004det}.  Recall that the
-signal in Figure~\ref{fig:svd-ts-colored} was especially useful for
-the study in this paper because it contained a number of different
-regimes.  We segmented this signal visually, but one could imagine
-using permutation entropy to do so instead.  Automating regime-shift
-detection would be an important step towards a fully adaptive modeling
-strategy, where old models are discarded and new ones are rebuilt
-whenever the time series enters a new regime.  WPE would be
-particularly powerful in this scenario, as its value can not only help
-with regime-shift detection, but also suggest what kind of model might
-work well in each new regime.  Of particular interest would be the
-class of so-called \emph{hybrid systems}~\cite{hybrid}, which exhibit
-discrete transitions between different continuous regimes---e.g., a
-lathe that has an intermittent instability or traffic at an internet
-router, whose characteristic normal traffic patterns shift radically
-during an attack.  Effective modeling and prediction of these kinds of
-systems is quite difficult; doing so adaptively and automatically---in
-the manner that is alluded to at the end of the previous
-paragraph---would be an even more important challenge.
+problem.  Any regime shift that causes a change in the predictive
+structure between the training signal and the test signal, for
+instance, will skew the MASE score (and the points on the curve).
+This can destroy the utility of our heuristic.  \alert{Conversely,
+  though,} a MASE score of a random-walk prediction that is larger
+than 1---as mentioned at end of the previous section---can potentially
+be a good indicator that a signal is nonstationary.  Detecting regime
+shifts---and adapting prediction models accordingly---is an important
+area of future work.  Indeed, one of the first applications of
+permutation entropy was to recognize the regime shift in brainwave
+data that occurs when someone has a seizure~\cite{cao2004det}.  Recall
+that the signal in Figure~\ref{fig:svd-ts-colored} was especially
+useful for the study in this paper because it contained a number of
+different regimes.  We segmented this signal visually, but one could
+imagine using permutation entropy to do so instead.  Automating
+regime-shift detection would be an important step towards a fully
+adaptive modeling strategy, where old models are discarded and new
+ones are rebuilt whenever the time series enters a new regime.  WPE
+would be particularly powerful in this scenario, as its value can not
+only help with regime-shift detection, but also suggest what kind of
+model might work well in each new regime.  Of particular interest
+would be the class of so-called \emph{hybrid systems}~\cite{hybrid},
+which exhibit discrete transitions between different continuous
+regimes---e.g., a lathe that has an intermittent instability or
+traffic at an internet router, whose characteristic normal traffic
+patterns shift radically during an attack.  Effective modeling and
+prediction of these kinds of systems is quite difficult; doing so
+adaptively and automatically---in the manner that is alluded to at the
+end of the previous paragraph---would be an interesting challenge.

### 2014-08-04 concl.tex
maybe done

@@
 % forecast methods, and time-series data from other systems.
-
-\alert{add FW stuff from section VI}
 
@@
 
-In this paper, we have argued that redundancy is a useful definition
-of the inherent predictability of an empirical time series.  To
-operationalize that definition, we use an approximation of the
+In this paper, we have argued that redundancy is a useful proxy for
+the inherent predictability of an empirical time series.  To
+operationalize that relationship, we use an approximation of the
 Kolmogorov-Sinai entropy, estimated using a weighted version of the
@@
 discussion in the previous paragraph.  These geometric features are a
-preliminary, but potentially useful, empirical heurists for knowing
-when a model is not well-matched to the task at hand: a point that is
-below and/or to the right of the shaded regions on a plot like
-Figure~\ref{fig:wpe_vs_mase_best} indicates the time series has more
-predictive structure than the forecast model can capture and
+preliminary, but potentially useful, heuristic for knowing when a
+model is not well-matched to the task at hand: a point that is below
+and/or to the right of the shaded regions on a plot like
+Figure~\ref{fig:wpe_vs_mase_all} indicates that the time series has
+more predictive structure than the forecast model can capture and
 exploit---and that one would be well advised to try another method.
@@
 These curves were determined empirically using a specific error metric
-and a finite set of forecast methods and time series traces.  If one
-uses a different error metric, the geometry---and the heuristic---will
-no longer apply.  And while the methods and traces were chosen to be
+and a finite set of forecast methods and time-series traces.  If one
+uses a different error metric, the geometry of the heuristic will be
+different---and may not even make sense, if one uses a metric that
+does not support comparison across different time series.  And while
+the methods and traces used in this study were chosen to be
 representative of the practice, they are of course not completely
@@
 structure between the training signal and the test signal, for
-instance, will skew the MASE score (and the points on the curve).
-This can destroy the utility of our heuristic.  \alert{Conversely,
-  though,} a MASE score of a random-walk prediction that is larger
-than 1---as mentioned at end of the previous section---can potentially
-be a good indicator that a signal is nonstationary.  Detecting regime
+instance, will skew the MASE score (and the points on the curve),
+obviating the utility of our heuristic.  \alert{Conversely, though,} a
+MASE score of a random-walk prediction that is larger than 1---as
+mentioned at end of the previous section---could potentially be a good
+indicator that a signal is nonstationary.  Detecting regime
 shifts---and adapting prediction models accordingly---is an important

### 2014-08-04 concl.tex
last paragraph

@@
 structure between the training signal and the test signal, for
-instance, will skew the MASE score (and the points on the curve),
-obviating the utility of our heuristic.  \alert{Conversely, though,} a
-MASE score of a random-walk prediction that is larger than 1---as
-mentioned at end of the previous section---could potentially be a good
-indicator that a signal is nonstationary.  Detecting regime
-shifts---and adapting prediction models accordingly---is an important
-area of future work.  Indeed, one of the first applications of
-permutation entropy was to recognize the regime shift in brainwave
-data that occurs when someone has a seizure~\cite{cao2004det}.  Recall
-that the signal in Figure~\ref{fig:svd-ts-colored} was especially
-useful for the study in this paper because it contained a number of
-different regimes.  We segmented this signal visually, but one could
-imagine using permutation entropy to do so instead.  Automating
-regime-shift detection would be an important step towards a fully
-adaptive modeling strategy, where old models are discarded and new
-ones are rebuilt whenever the time series enters a new regime.  WPE
-would be particularly powerful in this scenario, as its value can not
-only help with regime-shift detection, but also suggest what kind of
-model might work well in each new regime.  Of particular interest
-would be the class of so-called \emph{hybrid systems}~\cite{hybrid},
-which exhibit discrete transitions between different continuous
-regimes---e.g., a lathe that has an intermittent instability or
-traffic at an internet router, whose characteristic normal traffic
-patterns shift radically during an attack.  Effective modeling and
-prediction of these kinds of systems is quite difficult; doing so
-adaptively and automatically---in the manner that is alluded to at the
-end of the previous paragraph---would be an interesting challenge.
+instance, may skew the MASE score and thereby obviate the utility of
+our heuristic.  Conversely, though, a MASE score of a random-walk
+prediction that is significantly different from 1.0---as mentioned at
+end of the previous section---could potentially be a good indicator of
+nonstationary.
+
+Detecting regime shifts---and adapting prediction models
+accordingly---is an important area of future work.  Indeed, one of the
+first applications of permutation entropy was to recognize the regime
+shift in brainwave data that occurs when someone has a
+seizure~\cite{cao2004det}.  Recall that the signal in
+Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
+this paper because it contained a number of different regimes.  We
+segmented this signal visually, but one could imagine using some
+combination of WPE and MASE to do so instead (e.g., in a sliding
+window).  Automating regime-shift detection would be an important step
+towards a fully adaptive modeling strategy, where old models are
+discarded and new ones are rebuilt whenever the time series enters a
+new regime.  WPE would be particularly powerful in this scenario, as
+its value can not only help with regime-shift detection, but also
+suggest what kind of model might work well in each new regime.  Of
+particular interest would be the class of so-called \emph{hybrid
+  systems}~\cite{hybrid}, which exhibit discrete transitions between
+different continuous regimes---e.g., a lathe that has an intermittent
+instability or traffic at an internet router, whose characteristic
+normal traffic patterns shift radically during an attack.  Effective
+modeling and prediction of these kinds of systems is quite difficult;
+doing so adaptively and automatically---in the manner that is alluded
+to at the end of the previous paragraph---would be an interesting
+challenge.

### 2014-08-05 concl.tex
Liz is done

--- 2014-08-04 concl.tex
@@
 structure between the training signal and the test signal, for
-instance, will skew the MASE score (and the points on the curve),
-obviating the utility of our heuristic.  \alert{Conversely, though,} a
-MASE score of a random-walk prediction that is larger than 1---as
-mentioned at end of the previous section---could potentially be a good
-indicator that a signal is nonstationary.  Detecting regime
-shifts---and adapting prediction models accordingly---is an important
-area of future work.  Indeed, one of the first applications of
-permutation entropy was to recognize the regime shift in brainwave
-data that occurs when someone has a seizure~\cite{cao2004det}.  Recall
-that the signal in Figure~\ref{fig:svd-ts-colored} was especially
-useful for the study in this paper because it contained a number of
-different regimes.  We segmented this signal visually, but one could
-imagine using permutation entropy to do so instead.  Automating
-regime-shift detection would be an important step towards a fully
-adaptive modeling strategy, where old models are discarded and new
-ones are rebuilt whenever the time series enters a new regime.  WPE
-would be particularly powerful in this scenario, as its value can not
-only help with regime-shift detection, but also suggest what kind of
-model might work well in each new regime.  Of particular interest
-would be the class of so-called \emph{hybrid systems}~\cite{hybrid},
-which exhibit discrete transitions between different continuous
+instance, may skew the MASE score and thereby affecting the utility of
+our heuristic.  Conversely, though, a MASE score of a random-walk
+prediction that is significantly different from 1.0---as mentioned at
+end of the previous section---could potentially be a good indicator of
+nonstationarity.
+
+Detecting regime shifts---and adapting prediction models
+accordingly---is an important area of future work.  Indeed, one of the
+first applications of permutation entropy was to recognize the regime
+shift in brainwave data that occurs when someone has a
+seizure~\cite{cao2004det}.  Recall that the signal in
+Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
+this paper because it contained a number of different regimes.  We
+segmented this signal visually, but one could imagine using some
+combination of WPE and MASE to do so instead (e.g., in a sliding
+window across the time series).  Automating regime-shift detection
+would be an important step towards a fully adaptive modeling strategy,
+where old models are discarded and new ones are rebuilt whenever the
+time series enters a new regime.  Our WPE vs. MASE results could be
+particularly powerful in this scenario, as their values could not only
+help with regime-shift detection, but also suggest what kind of model
+might work well in each new regime.  Of particular interest would be
+the class of so-called \emph{hybrid systems}~\cite{hybrid}, which
+exhibit discrete transitions between different continuous
 regimes---e.g., a lathe that has an intermittent instability or

### 2014-08-05 concl.tex
white nosie to RW

@@
 the nonlinear dynamics of other systems.  Our preliminary results on
-other systems (H\'enon, Lorenz, a white-noise process, SFI ``A'',
+other systems (H\'enon, Lorenz, a random-walk process, SFI ``A'',
 transformations of the computer performance data) lead us to believe

### 2014-08-05 concl.tex
add 'nonlinear'

@@
 other systems (H\'enon, Lorenz, a random-walk process, SFI ``A'',
-transformations of the computer performance data) lead us to believe
+nonlinear transformations of the computer performance data) lead us to believe
 that our results will generalize beyond the examples described in this

### 2014-10-14 FinalSubmissiontoPRE/pecomplexity.tex
adding final submission directory

\documentclass[%
pre,
reprint,
superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose,
%preprint,
showpacs,
%preprintnumbers,
nofootinbib,
nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-1}
%\documentclass{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
%\usepackage{hyperref}
\usepackage[sort&compress]{natbib}
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage[caption=false]{subfig}
\usepackage{microtype}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{amssymb}
%\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{amsthm}
\usepackage{xspace}
\usepackage{pifont} % check and x marks
\usepackage{color}
\usepackage{listings}
\usepackage{rotating}
\usepackage{array,booktabs}
\lstset{language=C}

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\newcommand{\gcc}{{\tt 403.gcc}\xspace}
\newcommand{\svd}{{\tt dgesdd}\xspace}
\newcommand{\svdone}{{\tt dgesdd}$_1$\xspace}
\newcommand{\svdtwo}{{\tt dgesdd$_2$}\xspace}
\newcommand{\svdthree}{{\tt dgesdd$_3$}\xspace}
\newcommand{\svdfour}{{\tt dgesdd$_4$}\xspace}
\newcommand{\svdfive}{{\tt dgesdd$_5$}\xspace}
\newcommand{\svdsix}{{\tt dgesdd$_6$}\xspace}
\newcommand{\arima}{{\tt auto.arima}\xspace}

\newcommand{\naive}{na\"ive}
\newcommand{\row}{{\tt row\_major}\xspace}
\newcommand{\col}{{\tt col\_major}\xspace}
\newcommand{\alert}[1]{{\color{red}#1}}
%\newcommand{\color{red}}{{\color[rgb]{1,0,0}}}

%\parskip=3pt

%\usepackage{authblk}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newtheorem*{mydef}{Definition}

%Version 1
%\author{
%  \IEEEauthorblockN{Joshua Garland}
%  \IEEEauthorblockA{Dept. of Computer Science\\
%    University of Colorado at Boulder\\
%    Colorado, USA\\
%    Email: joshua.garland@colorado.edu}
%  \and
%  \IEEEauthorblockN{Ryan G.~James}
%  \IEEEauthorblockA{Complexity Sciences Center \& Dept. of Physics\\
%    University of California at Davis\\
%    California, USA\\
%    Email: rgjames@ucdavis.edu}
%    \and
%      \IEEEauthorblockN{Elizabeth Bradley}
%         \IEEEauthorblockA{Santa Fe Institute \\
%    New Mexico, USA\\
%    }
%  \IEEEauthorblockA{Dept. of Computer Science\\
%    University of Colorado at Boulder\\
%    Colorado, USA\\
%    Email: lizb@colorado.edu}
%  }

 %\author{
 %  \IEEEauthorblockN{
 %    Joshua Garland\IEEEauthorrefmark{1},
 %    Ryan G.~James\IEEEauthorrefmark{1} and
 %    Elizabeth Bradley\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}}
 %  \IEEEauthorblockA{
 %    \IEEEauthorrefmark{1}Dept. of Computer Science
 %    University of Colorado, Boulder, Colorado 80309-0430 USA\\
 %    Email: joshua.garland@colorado.edu, ryan.james@colorado.edu}
 %  \IEEEauthorblockA{
 %    \IEEEauthorrefmark{2}Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, New Mexico 87501  USA\\
 %    Email: lizb@colorado.edu}
 %  % \IEEEauthorblockA{
 %  %   \IEEEauthorrefmark{3}Complexity Sciences Center \& Physics Dept., University of California, Davis, %California 95616 USA\\
%   %   Email: rgjames@ucdavis.edu}
% }

\begin{document}

\preprint{APS/123-QED}

%\title{Manuscript Title}% Force line breaks with \\
%\thanks{A footnote to the article title}%

\title{
%Determinism, Complexity, and Predictability in Computer Performance\\
%Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
%\\
% A model-free strategy for quantifying the predictability of
% experimental time-series data 
Model-free quantification of time-series predictability
% \alert{Quantifying Time-Series Predictability through Structural Complexity}
%\\
%Quantifying Predictability using Structural Complexity
%\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
%\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
}

%\author[1]{Joshua Garland %\thanks{joshua.garland@colorado.edu}}
%\author[1]{Ryan James \thanks{ryan.james@colorado.edu}}
%\author[1,2]{Elizabeth Bradley \thanks{lizb@colorado.edu}}
%\affil[1]{Department of Computer Science\\
%  University of Colorado at Boulder\\
%  Colorado, USA
%}
%\affil[2]{Santa Fe Institute\\
%  New Mexico, USA
%}

\author{Joshua Garland}
\email{joshua.garland@colorado.edu}
\affiliation{Department of Computer Science, University of Colorado at Boulder. Colorado, USA}

\author{Ryan James}
\email{ryan.james@colorado.edu}
\affiliation{Department of Computer Science, University of Colorado at Boulder. Colorado, USA}

\author{Elizabeth Bradley}
\email{lizb@colorado.edu}
\affiliation{Department of Computer Science, University of Colorado at Boulder. Colorado, USA}
\affiliation{Santa Fe Institute, New Mexico, USA}

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified

\begin{abstract}

This paper provides insight into when, why, and how forecast
strategies fail when they are applied to complicated time series.  We
conjecture that the inherent complexity of real-world time-series
data---which results from the dimension, nonlinearity, and
non-stationarity of the generating process, as well as from
measurement issues like noise, aggregation, and finite data
length---is both \emph{empirically} quantifiable and directly
correlated with predictability.  In particular, we argue that
\emph{redundancy} is an effective way to measure complexity and
predictive structure in an experimental time series and that
\emph{weighted permutation entropy} is an effective way to estimate
that redundancy.  To validate these conjectures, we study 120
different time-series data sets.  For each time series, we construct
predictions using a wide variety of forecast models, then compare the
accuracy of the predictions with the permutation entropy of that time
series.  We use the results to develop a model-free heuristic that can
help practitioners recognize when a particular prediction method is
not well matched to the task at hand: that is, when the time series
has more predictive structure than that method can capture and
exploit.

\vspace{0.1in}
\noindent
{\bf Keywords}: entropy, permutation entropy, weighted permutation
entropy, time-series analysis, predictability

\preprint{Santa Fe Institute Working Paper XX-XX-XXX}
\preprint{arxiv.org:XXXX.XXXX [nlin.CD]}

\end{abstract}
\pacs{
05.45.Tp    % Time series analysis (nonlinear dynamics)
89.70.-a    % Information theory
%89.70.Hj    %   communication complexity
%89.70.Cf    %   entropy
%05.45.-a    % Nonlinear dynamics
89.75.Kd    %   Patterns (complex systems)
%95.75.Wx    % Time series analysis, time variability
95.75.Pq    % Mathematical procedures and computer techniques
%07.05.-t    % Computers in experimental physics
}

\maketitle

\section{Introduction}\label{sec:intro}
%****NOTES START HERE****

%BRAINSTORMING
%\begin{itemize}
%\item Complexity need not be hard to predict (can %point at the simple predictions paper) 
%\item random walk for example is best predicted by %guess what just happend
%\item The kind of complexity present matters, i.e., %that is whether the complexity is structured or not.%

%\item \col brings about the point nicely that some %prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working.

%\end{itemize}

%\begin{it}
%Paragraph on computer performance, including citations to Todd paper
%and summary of the results that indicate that they're deterministic
%nonlinear dynamical systems.  Given that, we should be able to
%predict.  What benefits would accrue if we could do so: power mgmt,
%end world hunger [this is my primary goal everyday :)], etc.
%\end{it}
%Things to add to introduction
%\begin{enumerate}

%\item \cmark Make an argument that Computer Performance %is a great testing ground as it gives signals that completely cover the spectrum of complexity \col \dots \gcc

%\item When deterministic structure even complex structure exists that structure can be utilized for prediction.
%\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard.
%\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex. Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not. Maybe even having 4 predictions. top being ARIMA of the above signals and bottom being LMA of the above signals. Show that one improved and one did not. Is it that we used the wrong method to predict or is it that we simply can't predict the signal better than a random walk due to high levels of internal signal complexity.

%\item\cmark Introduce the two main contributions of the paper which are outlined at the begining of the results section

%\end{enumerate}
%\noindent ***NOTES END HERE****
%
%\noindent ****SECTION START****

%<<<<<<< HEAD
%Complicated time-series data are ubiquitous in modern scientific research.  The
%complexity of these data spans a wide %range.  On the low end of this spectrum
%are time series that exhibit perfect %predictive structure, i.e, signals whose
%future values can be successfully %predicted from past values.  Signals like %this
%can be viewed as the product of an %underlying process that generates %information
%and/or transmits it from the past to the %future in a perfectly predictable
%fashion.  Constant or periodic signals, %for example, fall in this class.  On the
%opposite end of this spectrum are signals %that are what we call \emph{fully
%complex}, where the underlying generating %process transmits no information at
%all from the past to the future.  White %noise processes fall in this class.  In
%fully complex signals, knowledge of the %past gives no insight into the future,
%regardless of what model one chooses to %use. Signals in the midrange of this
%spectrum---e.g., deterministic chaos---%pose interesting challenges from a
%modeling perspective.  In these signals, %enough information is being transmitted
%%from the past to the future that an \emph{ideal} model---one that captures the
%generating process---can forecast the future behavior of the observed system
%with high accuracy. Though this notion of complexity differs from that
%considered by e.g. \cite{Shalizi2008}, which would consider a time series
%without any statistical regularities to be non-complex, it has been extensively
%considered by Boffetta et al~\cite{boffetta02}.
%
%As a corollary of the undecidability of the halting
%problem~\cite{halting-problem}, however, no single forecasting schema
%is ideal for all noise-free deterministic
%signals~\cite{weigend93}---let alone all real-world time-series
%data sets.  This leads naturally to an important and challenging
%question: given a complicated real-valued time series, does there
%exist a forecast model that can leverage the information (if any) that
%is being transmitted from past to future by the underlying generating
%process?  A first step in answering this question is to reliably
%quantify where on the complexity spectrum a given time series falls; a
%second step is to determine how complexity and predictability are
%related.  These are the goals of this paper.
%
%In particular, we wish to determine---without knowing anything about
%the generating process---whether a time series is too complex to successfully
%predict.  An important practical corollary to this is a strategy for
%assessing appropriateness of forecast methods.  If the forecast
%produced by a particular method is poor but the time series contains a
%significant amount of predictive structure, one can reasonably conclude that that method is inadequate
%to the task and that one should seek another method.
%=======
Complicated time-series data are ubiquitous in modern scientific
research.  The complexity of these data spans a wide range.  On the
low end of this spectrum are time series that exhibit perfect
predictive structure, i.e, signals whose future values can be
successfully predicted from past values.  Signals like this can be
viewed as the product of an underlying process that generates
information and/or transmits it from the past to the future in a
perfectly predictable fashion.  Constant or periodic signals, for
example, fall in this class.  On the opposite end of this spectrum are
signals that are what one could call \emph{fully complex}, where the
underlying generating process transmits no information at all from the
past to the future.  White noise processes fall in this class.  In
fully complex signals, knowledge of the past gives no insight into the
future, regardless of what model one chooses to use. Signals in the
midrange of this spectrum, e.g., deterministic chaos, pose interesting
challenges from a modeling perspective.  In these signals, enough
information is being transmitted from the past to the future that an
\emph{ideal} model---one that captures the generating process---can
forecast the future behavior of the observed system with high
accuracy.

% As a corollary of the undecidability of the halting
% problem~\cite{halting-problem}, however, no single forecasting schema
% is ideal for all noise-free deterministic
% signals~\cite{weigend-book}---let alone all real-world time-series
% data sets.

This leads naturally to an important and challenging question: given a
noisy real-valued time series from an unknown system, does there exist
a forecast model that can leverage the information (if any) that is
being transmitted forward in time by the underlying generating
process?  A first step in answering this question is to reliably
quantify where on the complexity spectrum a given time series falls; a
second step is to determine how complexity and predictability are
related in these kinds of data sets.  With these answers in hand, one
can develop a practical strategy for assessing appropriateness of
forecast methods for a given time series.  If the forecast produced by
a particular method is poor, for instance, but the time series
contains a significant amount of predictive structure, one can
reasonably conclude that that method is inadequate to the task and
that one should seek another method.  The goal of this paper is to
develop effective heuristics to put that strategy into practice.

% We have two goals in this paper: to determine---without knowing
% anything about the generating process---whether an empirical time
% series is too complex to successfully predict.
%>>>>>>> %1574864f4a0a007a6917eac812688ba02a5de1d5

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation~\cite{crutchfield2003}.
\label{page:redundancy}
Our approach exploits this decomposition in order to assess how much
predictive structure is present in a signal---i.e., where it falls on
the complexity spectrum mentioned above.  We define \emph{complexity}
as a particular approximation of Kolmogorov-Sinai
entropy~\cite{lind95}.  That is, we view a random-walk time series
(which exhibits high entropy) as purely complex, whereas a low-entropy
periodic signal is on the low end of the complexity spectrum.  This
differs from the notion of complexity used by e.g. \cite{Shalizi2008},
which would consider a time series without any statistical
regularities to be non-complex.  We argue that an extension of
\emph{permutation entropy}~\cite{bandt2002per}---a method for
approximating the entropy through ordinal analysis---is an effective
way to assess the complexity of a given time series.  Permutation
entropy is ideal for our purposes because it works with real-valued
data and is known to converge to the true entropy value. Other
existing techniques either require specific knowledge of the
generating process or produce biased values of the
entropy~\cite{bollt2001}.

We focus on real-valued, scalar, time-series data from physical experiments.
%
% that appears complicated: i.e., not periodic, linear, constant, etc.
%
We do not assume any knowledge of the generating process or its
properties: whether it is linear, nonlinear, deterministic,
stochastic, etc.  To explore the relationship between complexity,
predictive structure, and actual predictability, we generate forecasts
for a variety of experimental time-series datasets using four
different prediction methods, then compare the accuracy of those
predictions to the permutation entropy of the associated signals.
This results in two primary findings:
% 
\begin{enumerate}
% \item The complexity of a noisy real-valued time series can be
%   effectively quantifiable by computing its permutation entropy.

\item The permutation entropy of a noisy real-valued time series from
  an unknown system is correlated with the accuracy of an appropriate
  predictor.

\item The relationship between permutation entropy and prediction
  accuracy is a useful empirical heuristic for identifying mismatches
  between prediction models and time-series data.

\end{enumerate}
% 
There has, of course, been a great deal of good work on different ways
to measure the complexity of data, and previous explorations have
confirmed repeatedly that complexity is a challenge to prediction.  It
is well known that the way information is generated and processed
internally by a system plays a critical role in the success of
different forecasting methods---and in the choice of which method is
appropriate for a given time series.  This constellation of issues has
not been properly explored, however, in the context of noisy, poorly
sampled, real-world data from unknown systems.  That exploration, and
the development of strategies for putting its results into effective
practice, is the primary contribution of this work.  The empirical
results in the later sections of this paper not only elucidate the
relationship between complexity and predictability.  The strategy that
we derive from these results can also aid practitioners in choosing an
appropriate prediction model for a given real-world noisy time series
from an unknown system---a challenging task for which little guidance
is currently available.

% % Moved this to the beginning of forecasting section:
% 
% The forecast methods used in this study were chosen to be a
% representative sampling of standard prediction strategies, but they do
% not, of course, cover that space exhaustively.  Our goal here is an
% empirical assessment of the relationship between predictability and
% complexity, not formal results about a ``best'' predictor for a given
% time series.  From a practitioner's standpoint, it would be useful to
% know \emph{ a priori} if a time series contains enough predictive
% structure to make it worth spending the time and effort searching for
% a good forecast method.  It would also be useful to know if a given
% method is inadequate---that is, if an alternative method could do
% better and so one should continue searching.

% % Moved this to the beginning of methods section:
% 
% For the purposes of this study, we required a broad array of
% time-series datasets from across the complexity spectrum.  We chose to
% study sensor data from a computer-performance experiment.  While this
% is not a common laboratory experiment, it is a highly appropriate
% choice here.  Computers are extremely complicated systems and their
% dynamics is surprisingly rich.
% % Modern microprocessor chips contain multiple processing units and
% % multi-layer memories, for instance, and they use complicated
% % hardware/software strategies to maximize performance by moving data
% % and threads of computation across those resources.  As a result, the
% The processor and memory loads during the execution of even a very
% simple program can exhibit dynamical chaos, for
% instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
% example: a performance trace of a short program that repeatedly
% initializes the upper triangle of a matrix in column-major order.
% %
%  \begin{figure}[htp]
%     \centering
%     \includegraphics[width=\columnwidth]{figs/colshortts}
%     % where an .eps filename suffix will be assumed under latex,
%     % and a .pdf suffix will be assumed for pdflatex
%     \caption{A computer performance trace: the processor load of \col,
%       a simple C program that repeatedly initializes a matrix in
%       column-major order, running on an Intel
%       i7\textsuperscript{\textregistered}-based machine.}
%    \label{fig:col-ipc}
%   \end{figure}
% %
% A small change in the code can cause the dynamics to bifurcate to a
% periodic regime.  
% % 
% % Indeed, many computer performance traces exhibit
% % interesting regime changes as a program moves through the different
% % phases of its operation.
% By running different programs on the same computer, then, we can
% produce traces that span the whole range of the complexity spectrum,
% from completely predictable to completely unstructured---which makes
% this an ideal testbed for this study\footnote{Predicting the
%   \emph{state} of a computer, of course, would amount to solving the
%   halting problem.  What we are doing here is predicting computer
%   \emph{performance}, which does not violate the Rice-Shapiro
%   theorem~\cite{hopcroft2007}.}.
% % footnote to appease the theoretician in Josh

%The computer systems community has applied a variety of prediction
%strategies to traces like this, most of which employ regression.  An
%appealing alternative builds on the recently established fact that
%computers can be effectively modeled as deterministic nonlinear
%dynamical systems \cite{mytkowicz09}.  This result %implies the
%existence of a deterministic forecast rule for those %dynamics.  In
%particular, one can use \emph{delay-coordinate %embedding} to
%reconstruct the underlying dynamics of computer %performance, then use
%the resulting model to forecast the future values of computer
%performance metrics such as memory or processor loads
%\cite{josh-ida2011}.  In the case of simple microkernels like the one
%that produced the trace in Figure~\ref{fig:ipc}, this deterministic
%modeling and forecast strategy works very well.  In more-complicated
%programs, however, such as numerical software or compilers,
%this forecast strategy---as well as the traditional methods---break
%down some of the time, but work fine others.

%We argue that \emph{permutation entropy}
%\cite{bandt2002per}, a method for measuring the %entropy of a
%real-valued-finite-length time series through ordinal %analysis, is an
%effective way to explore that conjecture.  %{\color{red} Probably remove this and move a%s it is in the experimental methods or point to that section for more details}We study three
%examples---a simple microkernel and two complex %programs: one from the
%SPEC 2006CPU benchmark suite, and one from LAPACK---%running on an Intel i7-based machine.  For
%each program, we calculate the permutation entropy of %the processor
%load (instructions per cycle), then compare that to %the prediction accuracy attainable for
%that trace using a series of deterministic models.

The rest of the paper is organized as follows.
Section~\ref{sec:related} discusses previous results on generating
partitions, local modeling, and error distribution analysis, and
situates this work in that context. Section~\ref{sec:methods} covers
the experimental setup and methods used to collect the time-series
data. Section~\ref{sec:model} describes the prediction models used in
this study.  Section~\ref{sec:meaComplex} reviews permutation entropy,
the technique that we use to measure complexity.  In
Section~\ref{sec:results}, we estimate the complexity of each
empirical time series and compare that complexity to the accuracy of
predictions produced by the methods of Section~\ref{sec:model},
operating on that time series.  In Section~\ref{sec:conc}, we discuss
these results and their implications, and consider future areas of
research.

\section{Related Work }\label{sec:related}

Modeling time-series data for the purposes of prediction dates back at
least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
then, hundreds, if not thousands, of strategies have been developed
for a wide variety of prediction tasks.  The purpose of this paper is
not to add a new weapon to this arsenal, nor to do any sort of
theoretical assessment or comparison of existing methods.  Our goals
are focused more on the \emph{practice} of prediction: \emph{(i)} to
empirically quantify the predictive structure that is present in a
real-valued scalar time series and \emph{(ii)} to explore how the
performance of prediction methods is related to that inherent
complexity.  It would, of course, be neither practical nor interesting
to report results for every existing forecast method; instead, we
choose a representative set, as described in Section~\ref{sec:model}.

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most of the corresponding
solutions fall into two categories that we call model-based error
analysis and model-free information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
  used in different ways in the forecasting community.  Here, we
  distinguish those terms by the part of the time series that is the
  focus of the prediction: the observed data for the former and the
  unknown future for the latter.  In-sample forecasts---comparisons of
  predictions generated from \emph{part} of the observed time
  series---are useful for assessing model error and prediction
  horizons, among other things.}.
%
These distributions are then used as estimates of out-of-sample
forecast errors in those regions.  For example, Smith \emph{et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series exhibit varying levels of
local predictive uncertainty.  We expand on this idea later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate forecast errors and verify that they are normally
distributed.  If they are not, that suggests that there is structure
in the time series that the model-building process was unable to
recognize, capture, and exploit.  The problem with this approach is
lack of generality.
\label{page:normal-errors}
Normally distributed errors indicate that a model has captured the
structure in the data insofar as is possible, \emph{given the
  formulation of that particular model} (viz., the best possible
linear fit to a nonlinear dataset).  This gives no indication as to
whether another modeling strategy might do better.

A practice known as deterministic vs. stochastic
modeling~\cite{weigend93, Casdagli92dvsplots} bridges the gap
between local and global approaches to model-based error analysis.
The basic idea is to construct a series of local linear fits,
beginning with a few points and working up to a global linear fit that
includes all known points, and then analyze how the average
out-of-sample forecast error changes as a function of number of points
in the fit. The shape of such a ``DVS" graph indicates the amounts of
determinism and stochasticity present in a time series.

The model-based error analysis methods described in the previous three
paragraphs are based on specific assumptions about the underlying
generating process and knowledge about what will happen to the error
if those assumptions hold or fail.  Model-\emph{free} information
analysis moves away from those restrictions.  Our approach falls into
this class: we wish to measure the inherent complexity of an empirical
time series, then study the correlation of that complexity with the
predictive accuracy of forecasts made using a number of different
methods.

We build on the notion of \emph{redundancy} that was introduced on
page~\pageref{page:redundancy}, which formally quantifies how
information propagates forward through a time series:
% how much information prior observations of a system lend to future
% values
i.e., the mutual information between the past $n$ observations and the
current one.
% took this out because DCE hasn't been introduced yet:
%
% For example,
% this would be the amount of information that delay coordinate
% embedding captures.
The redundancy of i.i.d. random processes, for instance, is zero,
since all observations in such a process are independent of one
another.  On the other hand, deterministic systems---including chaotic
ones---have high redundancy that is maximal in the infinite limit, and
thus they can be perfectly predicted if observed for long
enough~\cite{weigend93}.  In practice, it is quite difficult to
estimate the redundancy of an arbitrary, real-valued time series.
Doing so requires knowing either the Kolmogorov-Sinai entropy or the
values of all positive Lyapunov exponents of the system.  Both of
these calculations are difficult, the latter particularly so if the
data are very noisy or the generating system is stochastic.

Using entropy and redundancy to quantify the inherent predictability
of a time series is not a new idea.  Past methods for this, however,
(e.g.,~\cite{Shannon1951, mantegna1994linguistic}) have hinged on
knowledge of the \emph{generating partition} of the underlying
process, which lets one transform real-valued observations into
symbols in a way that preserves the underlying dynamics~\cite{lind95}.
Using a projection that is not a generating partition---e.g., simply
binning the data---can introduce spurious complexity into the
resulting symbolic sequence and thus misrepresent the entropy of the
underlying system~\cite{bollt2001}.  Generating partitions are
luxuries that are rarely, if ever, afforded to an analyst, since one
needs to know the underlying dynamics in order to construct one.  And
even if the dynamics are known, these partitions are difficult to
compute and often have fractal boundaries~\cite{eisele1999}.

We sidestep these issues by using a variant of the \emph{permutation
  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate the
value of the Kolmogorov-Sinai entropy of a real-valued time
series---and thus the redundancy in that data, which our results
confirm to be an effective proxy for predictability.  This differs
from existing approaches in a number of ways.  It does not rely on
generating partitions---and thus does not introduce bias into the
results if one does not know the dynamics or cannot compute the
partition.  Permutation entropy makes no assumptions about, and
requires no knowledge of, the underlying generating process: linear,
nonlinear, the Lyapunov spectrum, etc.  These features make our
approach applicable to noisy real-valued time series from all classes
of systems, deterministic and stochastic.

There has been prior work under a very similar title to
ours~\cite{haven2005}, but there are only superficial similarities
between the two research projects. Haven \emph{et al.} utilize the
relative entropy to quantify the difference in predictability between
two distributions: one evolved from a small ensemble of past states
using the known dynamical system, and the other the observed
distribution. Our work quantifies the predictability of a single
observed time series using weighted permutation entropy and makes no
assumptions about the generating process.

More closely related is the work of Boffetta \emph{et
  al.}~\cite{boffetta02}, who investigate the scaling behavior of
finite-size Lyapunov exponents (FSLE) and $\epsilon$-entropy for a
wide variety of deterministic systems with known dynamics and additive
noise.  While the scaling of these measures acts as a general proxy
for predictability bounds, this approach differs from our work in a
number of fundamental ways.  First, \cite{boffetta02} is a theoretical
study that does not involve any actual predictions.  We focus on
real-world time-series data, where one does not necessarily have the
ability to perturb or otherwise interact with the system of interest,
nor can one obtain or manufacture the (possibly large) number of
points that might be needed to estimate the $\epsilon$-entropy for
small $\epsilon$.  Second, we do not require \emph{a priori} knowledge
about the noise and its interaction with the system.  Third, we tie
information---in the form of the weighted permutation
entropy---directly to prediction error via calculated values of a
specific error metric.  Though FSLE and $\epsilon$-entropy allow for
the comparison of predictability between systems, they do not directly
provide an estimate of prediction error.  Finally, our approach also
holds for stochastic systems, where neither the FLSEs nor their
relationship to predictability is well defined.
\section{Experimental Methods}\label{sec:methods}
% \begin{enumerate}
% \item \cmark  Experimental methods: (how we collect the %time series and what the times series areThis should be HPM %PAPI, which programs we model
%\item \cmark program descriptions and relevant citations
%\subitem \cmark description of \col
%\subitem \cmark description of \gcc
%\subitem \cmark description of \svd 
%\subitem \cmark description of svd regimes
% \end{enumerate}

% \subsection{Time Series Collection}

For the purposes of this study, we required a broad array of
time-series datasets from across the complexity spectrum.  We chose to
study sensor data from a computer-performance experiment.  While this
is not a common laboratory experiment, it is a highly appropriate
choice here.  Computers are extremely complicated systems and their
dynamics is surprisingly rich.
% Modern microprocessor chips contain multiple processing units and
% multi-layer memories, for instance, and they use complicated
% hardware/software strategies to maximize performance by moving data
% and threads of computation across those resources.  As a result, the
The processor and memory loads during the execution of even a very
simple program can exhibit dynamical chaos, for
instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
example: a short segment of a performance trace of a four-line C
program that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htp]
    \centering
    \includegraphics[width=\columnwidth]{colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A short segment of a computer performance trace: the
      instructions per CPU clock cycle (IPC) during the execution of
      \col, a simple program that repeatedly initializes a matrix in
      column-major order.  Each point is the average IPC in a 100,000
      instruction period.}
   \label{fig:col-ipc}
  \end{figure}
%
A small change in the code can cause the dynamics to bifurcate to a
periodic regime.  
% 
% Indeed, many computer performance traces exhibit
% interesting regime changes as a program moves through the different
% phases of its operation.
By running different programs on the same computer, then, we can
produce traces that span the whole range of the complexity spectrum,
from completely predictable to completely unstructured---which makes
this an ideal testbed for this study\footnote{Predicting the
  \emph{state} of a computer, of course, would amount to solving the
  halting problem.  What we are doing here is predicting computer
  \emph{performance}, which does not violate the Rice-Shapiro
  theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

The time-series data sets for these experiments were collected on an
Intel Core\textsuperscript{\textregistered} i7-2600-based machine.
% 
% running the 2.6.38-8 Linux kernel.  This particular microprocessor
% chip has eight processing units, a clock rate of 3.40 GHz, and a cache
% size of 8192 KB.  
% 
We gathered performance traces during the execution of three different
programs---the simple \col loop whose performance is depicted in
Figure~\ref{fig:col-ipc} and two more-complex programs: one from the
SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK linear
algebra package (\svd).  In all of these experiments, the scalar
observation $x_i$ was a measurement of the processor performance at
time $i$ during the execution of each program.
% 
% To record these measurements, we used the {\tt libpfm4} library, via
% PAPI
% % 
% % \footnote{Performance Application Programming Interface}
% % 
% 5.2~\cite{papi}, to stop program execution at 100,000-instruction
% intervals---the unit of time in this paper---and read the contents of
% the CPU's onboard hardware performance monitors\footnote{These
%   specialty registers are built into modern microprocessors for the
%   purpose of monitoring and storage of performance data.  They can be
%   programmed to measure a wide array of processor and memory metrics
%   besides the one used here.}, which we had programmed to count how
% many instructions were executed in each clock cycle (IPC).  
% 
% These traces, and the processes that generated them, are described
% briefly in the rest of this section.
% 
For statistical validation, we collected 15 performance traces from
each of the three programs.  For an in-depth description of the
experimental setup used to gather these data,
% including discussion of the implications of the sampling interval,
please see
\cite{josh-ida2011,josh-ida2013,zach-IDA10,mytkowicz09,todd-phd}.

% Obviously using a system to measure itself can cause
% interference--i.e, are you measuring the generating process cleanly or
% are you adding noise to the process?---but due diligence was put into
% monitoring the generating process of these time series without
% interfering with it in a significant way.

% \subsection{The Programs: [Maybe The generating processes]{\color{blue} EDITABLE}}
% We collect performance traces of processor efficiency (IPC) of three
% exemplar programs, [If we change the title of the section, add this:
%     which act as our generating processes]---a simple
% microkernel(\col) and two complex programs: one from the SPEC 2006CPU
% benchmark suite (\gcc), and one from LAPACK (\svd). In this section we
% provide a quick explanation of each program as well as an example
% performance trace (Seen in Figures
% \ref{fig:col-ts}-\ref{fig:svd-ts-colored}.

% Commented this out because the addition of the paragraph from the
% introduction seemed like it could replace it
% 
% \col is a simple C program that repeatedly initializes the upper
% triangle of a 2048 $\times$ 2048 matrix in column-major order.
% % 
% % by looping over the following three lines of code:
% % \begin{lstlisting}
% %   for(i=0;i<2048;i++)
% %   	for(j=i;j<2048;j++)
% %   		data[j][i] = 0;
% % \end{lstlisting}
% % full listing
% % #define SIZE (2048)
% % 
% % int main(int argc, char ** argv)
% % {
% %   int i, j, n;
% %   static int data[SIZE][SIZE];
% %   
% %   for(n=0;n<numInits;n++){  
% %   for(i=0;i<SIZE;i++)
% %   	for(j=i;j<SIZE;j++)
% %   		data[j][i] = 0;
% %   }
% %   return 0;
% % }
% %
% As mentioned in Section~\ref{sec:intro}, this simple program exhibits
% surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
% of the processor performance during the execution of this program is
% shown in Figure~\ref{fig:col-ts}.
% \begin{figure}[htbp]
%   \centering
%     \includegraphics[width=\columnwidth]{figs/colFullTS}
%     \caption{The instructions executed per CPU clock cycle (IPC) as
%       the \col program runs.  Each point is the average IPC in a
%       100,000 instruction period. Figure~\ref{fig:col-ipc} is a
%       closeup of the first part of this signal.  }
%     \label{fig:col-ts}
%   
%   \end{figure}

The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
complicated programs that are used in the computer-science community
to assess and compare the performance of different computers.  \gcc is
a member of that suite.
% 
% This program is written by Richard Stallman and is based on Version
% 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with
% many of its optimization flags enabled, compiling a series of input
% files and generating x86-64 assembly code files intended for execution
% on AMD Opteron processors\cite{spec}. 
% 
It is a \emph{compiler}: a program that translates code written in a
high-level language 
% (C, in the case of \gcc) 
into a lower-level format that can be executed by the processor chip.
Its behavior is far more complicated than that of \col, as is clear
from Figure~\ref{fig:gcc-ts}.
  \begin{figure}[t]
  \centering
    \includegraphics[width=\columnwidth]{gccfullts}
    \caption{An IPC trace during the execution of \gcc.}
    \label{fig:gcc-ts}
  \end{figure}
Unlike \col, where the processor utilization is quite structured,
\gcc's performance appears almost random.

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ matrix with real-valued entries.  For our
experiments, we chose $M=750$ and $N=1000$ and generated the matrix
entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as is clearly visible in Figure
\ref{fig:svd-ts-colored}.  
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{SVD1RegimesColored}
    \caption{An IPC trace during the execution of \gcc.  The colors
      identify the different \emph{segments} of the signal that are
      discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
As the code moves though its different phases---diagonalizing the
matrix, computing its transpose, multiplying, etc.---the processor
utilization patterns change quite radically.  For the first
$\sim$21,000 measurements (21,000 $\times$ 100,000 instructions),
roughly 1.8 instructions are executed per cycle, on the average, by
the eight processing units on this chip.  After that, the IPC moves
through a number of different oscillatory regimes, which we have
color-coded in the figure in order to make textual cross-references
easy to track.

The wide range of behaviors in Figure~\ref{fig:svd-ts-colored}
provides a distinct advantage, for the purposes of this paper, in that
a number of different generating processes---with a wide range of
complexities---are at work in different phases of a single time
series.  The \col and \gcc traces in Figures~\ref{fig:col-ipc} and
\ref{fig:gcc-ts} appear to be far more consistent over time---probably
the result of a single generating process with consistent complexity.
\svd, in contrast, has multiple regimes, each the result of different
generating processes.  To take advantage of this, we split the signal
into six different segments, thereby obtaining an array of examples
for the analyses in the following sections.  For notational
convenience, we refer to these 90 time-series data sets\footnote{15
  runs, each with six regimes} as {\tt dgesdd$_i$}, with $i \in
\{1\dots6\}$ where $i$ corresponds to one of the six segments of the
signal, ordered from left to right.  These segments, which were
determined visually, are shown in different colors in
Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
of course, particularly since the regimes exhibit some fractal
structure.  Thus, it may be the case that more than one generating
process is at work in each of our segments.  This is a factor in the
discussion of Section~\ref{sec:results}.

%[Orphan citation?] For regimes \cite{cao2004det}

\section{Modeling }\label{sec:model}
%
% {\color{blue} EDITABLE}
%
% Section outline:
%
% \begin{enumerate}
% \item \cmark section intro
% \item \cmark fix order from simplest to most complex
% \item Description of DCE and parameter estimation
% \item\cmark Description of auto ARIMA
% \item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
% \item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.
%
% \end{enumerate}

In this section, we describe the four different forecasting methods
used in this study, as well as the error metric used to evaluate their
predictive accuracy.  These methods include:
\begin{itemize}
\item The \emph{random-walk} method, which uses the previous value in
  the observed signal as the forecast,

\item The \emph{\naive} method, which uses the mean of the
  observed signal as the forecast,

\item The \emph{ARIMA} (auto-regressive integrated moving average)
  method, a common linear forecast strategy, instatiated via the
  \emph{\arima} procedure \cite{autoARIMA}, and

\item The \emph{LMA} (Lorenz method of analogues) method, which uses a
  near-neighbor forecast strategy on a dynamical reconstruction of the
  signal.
\end{itemize}
ARIMA models are based on standard linear techniques.  LMA is designed
to capture and exploit the deterministic structure of a signal from a
nonlinear dynamical system.  The \naive ~and random-walk methods,
somewhat surprisingly, often outperform these more-sophisticated
prediction strategies in the case of highly complex signals, as
discussed below.

% These four forecast methods were chosen to be a representative
% sampling of standard prediction strategies, but they do not, of
% course, cover that space exhaustively.  Our goal here is an empirical
% assessment of the relationship between predictability and complexity,
% not formal results about a ``best'' predictor for a given time series.
% From a practitioner's standpoint, it would be useful to know \emph{ a
%   priori} if a time series contains enough predictive structure to
% make it worth spending the time and effort searching for a good
% forecast method.  It would also be useful to know if a given method is
% inadequate---that is, if an alternative method could do better and so
% one should continue searching.  \alert{Liz to do: check that this is
%   not redundant, given the other text in this section (and elsewhere
%   in the paper).  it may be better to put this in the conclusion.}

\subsection{Two Simple Prediction Strategies}
\label{sec:simple}

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the predicted value $p_i$ at time $i$ is
calculated using the following relation: $$p_i = x_{i-1}$$ The
prediction strategy that we refer to using the term ``\naive''
averages the prior observations to generate the forecast: $$p_i =
\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
simplistic, they are not without merit.  For a time series near the
high end of the complexity spectrum---i.e., one that possesses very
little predictive structure---these two methods can actually be the
best choice.  In forecasting currency exchange rates, for instance,
sophisticated econometrics-based prediction models fail to
consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
These signals are constantly changing, noisy, and possess very little
predictive structure, but their variations are not---on the
average---very large, so the random-walk method's strategy of simply
guessing the last known value is not a bad choice.  If a signal has a
unimodal distribution with low variance, the \naive ~prediction
strategy will perform quite well---even if the signal is highly
complex---simply because the mean is a good approximation of the
future behavior.  Moreover, the \naive ~prediction strategy's temporal
average effects a low-pass filtering operation, which can  mitigate the
complexity in signals with very little predictive structure.

Both of these methods have significant weaknesses, however.  Because
they do not model the temporal patterns in the data, or even the
distribution of its values, they cannot track changes in that
structure.  This causes them to fail in a number of important
situations.  Random-walk strategies are a particularly bad choice for
time series that change significantly at every time step.  In the
worst case---a large-amplitude square wave whose period is equivalent
to twice the sample time---a random-walk prediction would be exactly
180 degrees out of phase with the true continuation.  The \naive
~method would be a better choice in this situation, since it would
always split the difference.  It would, however, perform poorly when a
signal has a number of long-lived regimes that have significantly
different means.  In this situation, the inertia of the \naive
~method's accumulating mean is a liability and the agility of the
random-walk method is an advantage, since it can respond quickly to
regime shifts.

Of course, methods that could capture and exploit the geometry of the
data and/or its temporal patterns would be far more effective in the
situations described in the previous paragraph.  The \arima and LMA
methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
designed to do exactly that.
% Note, though, that periodic patterns
%appear only in signals that are at the low end of the %complexity
%spectrum.
However, if a signal contains little predictive structure, forecast
strategies like ARIMA and LMA have nothing to work with and thus will
often be outperformed by the two simple strategies described in this
section.  This effect is explored further in Sections~\ref{sec:accuracy}
and~\ref{sec:results}.

\subsection{A Regression-Based Prediction Strategy}
\label{sec:arima}
%\begin{enumerate}
%\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
%\item\cmark define rigorously, all the backshift operator garbage

%$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = c + \Theta(B^m)\theta(B)\epsilon_i$$

%$$B^mX_i  = X_{i-m}$$
%\item\cmark discuss that this method needs linear structure to work correctly
%\item \cmark maybe talk about converging to zero and limits on prediction horizon
%\end{enumerate}

%For this \cite{davislinearts}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% here is the old version of this material, for use in other documents...
%
%Perhaps the simplest way to capture and exploit the structure of data
%is to fit a hyperplane to the dataset and then use that it to forecast
%new data points.  The roots of this date back to Yule's 1927 invention
%of the autoregressive schema~\cite{weigend93}, which forecasts the
%next time step through a weighted average of past observations: $$p_i
%= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
%that is fit to the past $p+1$ [???] values of $x_i$ via the
%backshift operator $B$ \footnote{I changed the superscript to $k$ to
%  make this definition general---not connected to any of the terms
%  that have specific meanings later in this section.}:
%$$B^k x_i = x_{i-k}$$
%%
%with $k=1...p$.  To account for noise in the data, one can add a
%so-called ``moving average'' term to the model:
%$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
%degree $q$ and $\epsilon$ is white noise with the same mean and
%variance as the data.
%% if you leave this in, you'll need to explain a lot about the data
%% and the calculation
%% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.
%To remove nonstationarities in the data, one can detrend it using a
%differencing operation: $(1-B^d)x_i$.
%
%A strategy that incorporates all three of these features is called a
%\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
% $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%%
%{\color{red} You had superscripts on some of the $B$s and not on
%  others.  I added them but I may have gotten them wrong.  I also
%  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
%  the first use of that construction.  Please make sure things are
%  consistent, here and throughout this section.}  Here, $p$, $d$ and
%$q$ correspond to the orders of the autoregressive, detrending, and
%moving average terms, respectively.
%
%If periodic structure is present in the data, a \emph{seasonal ARIMA}
%model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%%
%$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
%\Theta(B^Q)\theta(B^q)\epsilon_i$$
%%
%Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
%and $Q$, respectively [which accomplish what purpose?  why?  how?
%    where did $D$ come from and what does it do?].  $m$ is the
%seasonal frequency [shouldn't this be ``period''?].  [Explain why
%    one should choose $P=Q=m$, since that's what the equation below
%    implicitly does.]  To use a seasonal ARIMA model, one first uses
%the detrending term to remove nonstationarity: $$\hat{x_i} =
%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $p_i$ by
%evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
%\Theta(B^m)\theta(B^q)\epsilon_i$$
% ...down to here
%%%%%%%%%%%%%%%%%%%%%%%%

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the known points and then use it to
make predictions.  The roots of this approach date back to the
original autoregressive schema~\cite{weigend93}, which forecasts the
next time step through a weighted average of past observations: $$p_i
= \sum_{j=1}^{i-1} a_j x_j$$ The weighting coefficients $a_j$ are
generally computed using either an ordinary least squares approach, or
with the method of moments using the Yule-Walker equations.  To
account for noise in the data, one can add a so-called ``moving
average'' term to the model; to remove nonstationarities, one can
detrend the data using a differencing operation.  A strategy that
incorporates all three of these features is called a \emph{nonseasonal
  ARIMA model}.  If evidence of periodic structure is present in the
data, a \emph{seasonal ARIMA model}, which adds a sampling operation
that filters out periodicities, can be a good choice.

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms.  We employ the
automated fitting techniques described in~\cite{autoARIMA} to
accomplish this, producing what we will call an ``\arima model'' in
the rest of this paper.  This procedure uses sophisticated
methods---KPSS unit-root tests~\cite{KPSSunit}, a customization of the
Canova-Hansen test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximum likelihood of
the model fitted to the detrended data---to select good values for the
free parameters of the ARIMA model.

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts can be inaccurate.
Another weakness of this method is prediction horizon: an ARIMA
forecast is guaranteed to converge (to the mean, to a constant value,
or to a linear trend) after some number of predictions, depending on
model order.  To sidestep this issue, we build forecasts in a stepwise
fashion: i.e., fit the \arima model to the existing data, use that
model to perform a one-step prediction, rebuild the \arima model using
the latest observations, and iterate until the desired prediction
horizon is reached.  For consistency, we take the same approach with
the other three models in this study as well, even though doing so
amounts to artificially hobbling LMA.

\subsection{A Nonlinear Prediction Strategy}
\label{sec:lma}

When the temporal progressions in a time series are produced by a
deterministic nonlinear process, one can use a technique called
delay-coordinate embedding
%
%% Note: claiming that we can reconstruct the dynamics of the
%% underlying generating process isn't right.  That would be
%% equivalent to solving the system identification problem!  We're
%% not reconstructing the underlying system, just its output.
%
to model the structure of the information generation and transmission
occurring in the underlying process, then use that reconstruction to
generate forecasts.  This section discusses the theory and
implementation of a prediction strategy that is based on this idea.

Delay-coordinate embedding~\cite{packard80,Sauer:1991lr,Takens:1981uq}
allows one to reconstruct a dynamical system's full state-space
dynamics from a scalar time-series measurement---provided that some
conditions hold regarding those data.  Specifically, if the underlying
dynamics and the measurement function---the mapping from the unknown
state vector $\vec{X}$ to the observed value $x_i$---are both smooth
and generic, Takens~\cite{Takens:1981uq} formally proves that the
delay-coordinate map
\[
F(\tau,m)(\vec{X}) = ([x_{i} ~ x_{i+\tau} ~ \dots ~x_{i+m\tau}])
\]
from a $d$-dimensional smooth compact manifold $M$ to
$\mathbb{R}^{2d+1}$ is a diffeomorphism on $M$: in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.  This is an extremely powerful result; among other
things, it means that one can model the full system dynamics, up to
diffeomorphism, without measuring---or even knowing---anything about
the state variables.

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as
an estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor method of~\cite{KBA92} to estimate $m$.  Some
example plots of data from
Figures~\ref{fig:col-ipc}-\ref{fig:svd-ts-colored}, embedded following
this procedure, are shown in Figure~\ref{fig:embedding}.
% \begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colipc3d}
%    \caption{\col }
%    \label{fig:colEmbedding}
%  \end{subfigure}%  \\
%
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gcc3dipc}
%    \caption{\gcc}
%    \label{fig:gccEmbedding}
%  \end{subfigure}
%  \\
%  \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svd53dipc2}
%    \caption{\svdfive}
%    \label{fig:svdfiveEmbedding}
%  \end{subfigure}%
%%
%
%
%     %\includegraphics[width=\textwidth]{colipc3d}
%     \caption{3D projections of delay-coordinate embeddings of the
%       traces from (a) Figure~\ref{fig:col-ts} (b)
%       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
%       Figure~\ref{fig:svd-ts-colored}.}
% \label{fig:embedding}
% \end{figure}
 \begin{figure}[!ht]
    \subfloat[\col\label{fig:colEmbedding}]{%
      \includegraphics[width=\columnwidth]{colipc3d}
    }
    \hfill
    \subfloat[\gcc\label{fig:gccEmbedding}]{%
      \includegraphics[width=\columnwidth]{gcc3dipc}
    }
        \hfill
    \subfloat[\svdfive \label{fig:svdfiveEmbedding}]{%
      \includegraphics[width=\columnwidth]{svd53dipc2}
    }
    \caption{3D projections of delay-coordinate embeddings of the
       traces from (a) Figure~\ref{fig:col-ipc} (b)
       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
       Figure~\ref{fig:svd-ts-colored}.}
    \label{fig:embedding}
  \end{figure}
%% can cut for space if need be:
%The coordinates of each point in these plots are differently delayed
%elements of the instructions per cycle time series $X_{i,obs}$: that
%is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
%$X_{i+2\tau,obs}$ on the third, and so on.
%

Geometric structure in these kinds of plots is an indication of
structure in the information generation/transmission process that
produced the time series.  The dynamical systems community has
developed a number of methods that leverage this structure to generate
predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
One of the most straightforward of these is the \emph{Lorenz method of
  analogues} (LMA), which is essentially nearest-neighbor prediction
in the embedded\footnote{Lorenz's original formulation used the full
  system state space;
%
%According to \cite{kantz97},
%
this method was first extended to embedded dynamics by
Pikovsky~\cite{pikovsky86-sov}, but is also related to the prediction
work of Sugihara \& May~\cite{sugihara90}}
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
prediction---provides remarkably accurate forecasts when the
generating process is a deterministic dynamical system.

Since LMA does not rest on an assumption of linearity (as ARIMA models
do), it can handle both linear and nonlinear processes.  If the
underlying generating process is nondeterministic, however, it can
perform poorly.  Figure~\ref{fig:embedding}(b), for instance, appears
to contain little structure, so one might not expect LMA to work well
on this signal.  More structure appears to be present in
Figure~\ref{fig:embedding}(c), but this reconstruction also appears to
contain some noise.  The question as to how much structure is present
in a reconstruction, and how much of that structure can be captured
and used by LMA, is apropos of the central question treated in this
paper.  It may be that \gcc has some redundancy that LMA cannot
exploit, or that the structure in \svdfive is effectively obfuscated,
from the standpoint of the LMA method, by noise.

Without any knowledge of the generating process, answers to these
questions can only be derived from the data, with all of the attendant
problems (noise, sampling issues, and so on).  By quantifying the
balance between redundancy, predictive structure, and entropy for
these real-valued time series---as shown in
Section~\ref{sec:results}---we can begin to answer these questions in
an effective and practical manner.

% If enough predictive structure is present in these embeddings, that
% structure can be used to build an effective forecast model.

\subsection{Assessing Prediction Accuracy}
\label{sec:accuracy}

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 120
traces described in Section~\ref{sec:methods}, then calculate the
error of the predictions with respect to the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to here as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, whose ability to generate accurate predictions is limited
only by the positive Lyapunov exponents of the system.  However, the
performance of the other three methods used here will degrade severely
if the associated models are not periodically rebuilt.  In order to
make the comparison fair, we used an iterative one-step prediction
schema \emph{for all four methods}.  This has the slightly confusing
effect of causing the ``test'' signal to be used both to assess the
accuracy of each model and for periodic refitting.

Figure~\ref{fig:forecast-example} shows example forecasts made using
all four methods for the \col, \gcc, and \svdfive time series.
%\begin{figure*}[htbp]
%  \centering
%
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colRWForecast}
%    \caption{\col\\ random walk }
%    \label{fig:colRW}
%  \end{subfigure}%
%   \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
%    \caption{\gcc\\ random walk }
%    \label{fig:gccRW}
%  \end{subfigure}%
%     \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
%    \caption{\svdfive\\ random walk}
%    \label{fig:svd5RW}
%  \end{subfigure}%
%  \\
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
%    \caption{\col\\ na\"ive }
%    \label{fig:colMEAN}
%  \end{subfigure}%
%   \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
%    \caption{\gcc\\ na\"ive }
%    \label{fig:gccMEAN}
%  \end{subfigure}%
%     \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
%    \caption{\svdfive\\ na\"ive }
%    \label{fig:svd5MEAN}
%  \end{subfigure}%
%  \\
%    \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
%    \caption{\col \\ \arima}
%    \label{fig:colARIMA}
%  \end{subfigure}
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
%    \caption{\gcc \\ \arima }
%    \label{fig:gccARIMA}
%  \end{subfigure}%
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
%    \caption{\svdfive \\ \arima }
%    \label{fig:svd5ARIMA}
%  \end{subfigure}%
%  \\
%
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
%    \caption{\col \\ LMA}
%    \label{fig:colLMA}
%  \end{subfigure}
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
%    \caption{\gcc \\ LMA}
%    \label{fig:gccLMA}
%  \end{subfigure}
%        \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
%    \caption{\svdfive \\ LMA}
%    \label{fig:svd5LMA}
%  \end{subfigure}
%  %\begin{subfigure}{0.5\textwidth}
%  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
%   \caption{Predicted ($p_j$) versus true values ($c_j$) for
%     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
%% On this type of plot, a perfect prediction lies exactly on the
%% diagonal.
%%
%%
%}
%\label{fig:forecast-example}
%  %\end{subfigure}
%\end{figure*}
\newcolumntype{C}{>{\centering\arraybackslash} m{0.6\columnwidth} }
\begin{figure*}
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{m{0.1\columnwidth}|CCC}
    & {\large \col} & {\large \gcc} & {\large \svdfive} \\
    & & & \\
            \hline \\
    \rotatebox{90}{\large{random walk}} &
    \includegraphics[width=0.6\columnwidth]{colRWForecast} &
    \includegraphics[width=0.6\columnwidth]{gccRWForecast} &
    \includegraphics[width=0.6\columnwidth]{svdfiveRWForecast} \\
    \rotatebox{90}{\large \naive} &
    \includegraphics[width=0.6\columnwidth]{colMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{gccMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{svdfiveMeanForecast} \\
    \rotatebox{90}{\large \arima} &
    \includegraphics[width=0.6\columnwidth]{colARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{gccARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{svdfiveARIMAForecast} \\
    \rotatebox{90}{\large LMA} &
    \includegraphics[width=0.6\columnwidth]{colLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{gccLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{svdfiveLMAForecast}
  \end{tabular}
  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts
    of \col, \gcc, and \svdfive made using each of the four 
    strategies studied here.  }
  \label{fig:forecast-example}
\end{figure*}
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while \arima
does not.  Horizontal lines result when a constant predictor like the
\naive ~method is used on a non-constant signal.  Point clouds reflect
the structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of some of the plots in Figure~\ref{fig:forecast-example}
(e.g., random walk and \arima on \col) are reminiscent of the
projected embedding in Figure~\ref{fig:embedding}(a).  Indeed, for a
random-walk predictor, a $p_j$ vs. $c_j$ plot is technically
equivalent to a two-dimensional embedding with $\tau=1$.  For \arima,
the correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the \arima $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error (MASE) between the true and predicted signals:
%
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
%
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  Like many error metrics, MASE is a
normalized measure: the scaling term in the denominator
%
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$
%
is the average in-sample forecast error for a random-walk prediction
over the initial training signal $\{x_i\}^n_{i=1}$.  That is, MASE$<1$
means that the prediction error in question was, on the average,
smaller than the average error of a random-walk forecast on the
training data.  Analogously, MASE$>1$ means that the corresponding
prediction method did \emph{worse}, on average, than the random-walk
method.  While its comparative nature is somewhat different than
traditional metrics like normalized root mean squared error, MASE has
the significant advantage of allowing one to make a fair comparison
across varying methods, prediction horizons, and signal
scales---attributes that are key to any broad study of predictability.

MASE scores for all 360 experiments are tabulated in the four middle
columns in Table~\ref{tab:error}.
%\begin{table*}
%\caption{Mean absolute scaled error (MASE) scores and weighted
%  permutation entropies for all eight processes studied in this paper.
%  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
%  \alert{please reorder the columns from left to right to match the
%    order in which the methods were introduced, rather than in the
%    reverse of that order}}
%  \begin{center}
%  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
%  \hline\hline Signal & LMA MASE & \arima MASE & na\"{i}ve MASE & RW MASE & WPE \\
%  \hline
%
%  \col       & $0.050 \pm 0.002$ & $0.599 \pm 0.211$ & $0.571 \pm 0.002$  & $1.001 \pm 0.002$ & $0.513 \pm 0.003$ \\
%
%  \gcc       & $1.530 \pm 0.021$ & $1.837 \pm 0.016$ & $1.797 \pm 0.010$  & $1.138 \pm 0.011$ & $0.943 \pm 0.001$ \\
%
%  \svdone    & $0.827 \pm 0.076$ & $0.714 \pm 0.075$ & $2.676 \pm 4.328$  & $0.933 \pm 0.095$ & $0.957 \pm 0.016$ \\
%
%  \svdtwo    & $1.279 \pm 0.020$ & $2.163 \pm 0.027$ & $3.054 \pm 0.040$  & $1.125 \pm 0.012$ & $0.846 \pm 0.004$ \\
%
%  \svdthree  & $0.619 \pm 0.021$ & $0.713 \pm 0.010$ & $31.386 \pm 0.282$ & $0.707 \pm 0.009$ & $0.716 \pm 0.006$ \\
%
%  \svdfour   & $0.779 \pm 0.036$ & $0.979 \pm 0.032$ & $2.661 \pm 0.074$  & $1.034 \pm 0.035$ & $0.825 \pm 0.008$ \\
%
%  \svdfive   & $0.718 \pm 0.048$ & $2.370 \pm 0.051$ & $20.870 \pm 0.192$ & $1.001 \pm 0.047$ & $0.678 \pm 0.007$ \\
%
%  \svdsix    & $0.739 \pm 0.068$ & $1.438 \pm 0.061$ & $2.197 \pm 0.083$  & $1.060 \pm 0.055$ & $0.748 \pm 0.011$ \\
%
%  \hline\hline
%  \end{tabular*}
%  \end{center}
% \label{tab:error}
%  \end{table*}%
%  
 \begin{table*}
\caption{Mean absolute scaled error (MASE) scores and weighted
  permutation entropies for all eight processes studied in this paper.
  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
  }
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
  \hline\hline 
Signal & RW MASE & na\"{i}ve MASE & \arima MASE & LMA MASE & WPE \\
\hline
  \col       & $1.001 \pm 0.002$ & $0.571 \pm 0.002$  & $0.599 \pm 0.211$ & $0.050 \pm 0.002$ & $0.513 \pm 0.003$ \\
  \gcc       & $1.138 \pm 0.011$ & $1.797 \pm 0.010$  & $1.837 \pm 0.016$ & $1.530 \pm 0.021$ & $0.943 \pm 0.001$ \\
  \svdone    & $0.933 \pm 0.095$ & $2.676 \pm 4.328$  & $0.714 \pm 0.075$ & $0.827 \pm 0.076$ & $0.957 \pm 0.016$ \\
  \svdtwo    & $1.125 \pm 0.012$ & $3.054 \pm 0.040$  & $2.163 \pm 0.027$ & $1.279 \pm 0.020$ & $0.846 \pm 0.004$ \\
  \svdthree  & $0.707 \pm 0.009$ & $31.386 \pm 0.282$ & $0.713 \pm 0.010$ & $0.619 \pm 0.021$ & $0.716 \pm 0.006$ \\
  \svdfour   & $1.034 \pm 0.035$ & $2.661 \pm 0.074$  & $0.979 \pm 0.032$ & $0.779 \pm 0.036$ & $0.825 \pm 0.008$ \\
  \svdfive   & $1.001 \pm 0.047$ & $20.870 \pm 0.192$ & $2.370 \pm 0.051$ & $0.718 \pm 0.048$ & $0.678 \pm 0.007$ \\
  \svdsix    & $1.060 \pm 0.055$ & $2.197 \pm 0.083$  & $1.438 \pm 0.061$ & $0.739 \pm 0.068$ & $0.748 \pm 0.011$ \\  
    \hline\hline
  \end{tabular*}
  \end{center}
 \label{tab:error}
  \end{table*}%
In view of the discussion at the end of the previous paragraph, the
fact that the values in the second-from-left column are not
identically equal to 1.00 may be somewhat surprising.  This can happen
due to differences between in-sample and out-of-sample forecasting;
the last 10\% of \svdthree, for instance---the out-of-sample
signal---was more amenable to random-walk prediction, on the average,
than the first 90\%.  See Section~\ref{sec:results} for a deeper
discussion of this effect.

Comparing the values in Table~\ref{tab:error} with the geometry of the
plots in Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals was
$0.050$, for instance, while \arima scored much worse ($0.599$).  That
is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while \arima only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.

In other cases, the correspondence between MASE score and the visual
appearance of these kinds of plots is not so clear cut.  The plots of
LMA predictions of \col and \svdfive both lie near the diagonal, for
instance, but the corresponding MASE scores are very different:
$0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times
better than random-walk forecasts of the same signals).  This is a
result of the normalization in the MASE score calculation.  Recall
from Section~\ref{sec:simple} that the random-walk method performs
especially poorly on signals that oscillate rapidly.  \col fits this
description, so the random-walk error on this signal---the denominator
of the four MASE scores in the first row of the Table---is
pathologically high, which skews those scores down.  Random-walk
prediction is very effective for the \svdfive signal, on the other
hand, so the denominator is small and the MASE scores are skewed
upwards.

Normalization is, as is often the case, a double-edged sword.  Here,
it facilitates comparisons across signals of different types and
lengths, but its particular pathologies must be taken into account
when analyzing the results, as discussed at more length in
Section~\ref{sec:results}.

% Table \ref{tab:error} provides the MASE distributions
% {\color{red}[Joshua: Ryan, Is this the right word? we give mean $\pm$
%       std. dev but some have very skewed right tails]} for each of
% the eight signals and three prediction strategies, these are averaged
% over 15 runs of each signal-method combination. For later comparison,
% Table \ref{tab:error} also has the complexity measure we introduce in
% Section \ref{sec:meaComplex}.

%[and cherry pick a few examples of \gcc and \col to put in the text ]

%---works quite well on the trace in Figure~%\ref{fig:ipc}, as%
%shown in Figure~\ref{fig:lmacol}.
%
%\begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\col }
%    \label{fig:lmacol}
%  \end{subfigure}%
%
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\gcc}
%    \label{fig:lmagcc}
%  \end{subfigure}
%  \begin{subfigure}{\columnwidth}
%%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\svdfive}
%    \label{fig:lmasvdfive}
%  \end{subfigure}%
%  \caption{{\color{red} [If Liz thinks we should %include these I actually need to generate this %figure]}An LMA-based forecast of
%       processor-efficiency performance trace from the%\col ,\gcc and \svdfive generating processes.  Red %circles and blue $\times$s are the true and
%       predicted values, respectively; vertical bars %show where these
%       values differ.}\label{fig:lmapredictions}
% \end{figure}
%
%\begin{figure}[htbp]
%  \centering
%   \includegraphics[width=\textwidth]{colPredShortTS}
%    \caption{A forecast of the last 4,000 points of the %signal in
%      Figure~\ref{fig:ipc} using an LMA-based strategy on %the
%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
%      $\times$s are the true and predicted values, %respectively;
%      vertical bars show where these values differ.}
%\label{fig:lmacol}
%\end{figure}
%
%\begin{figure}[htbp]
%  \centering
%    \includegraphics[width=\textwidth]{colPredShortTS}
%     \caption{{\color{red} [actually need to generate this %figure]}An LMA-based forecast of the last 4,000 points of a
%       processor-load performance trace from the \gcc
%       benchmark.  Red circles and blue $\times$s are the %true and
%       predicted values, respectively; vertical bars show %where these
%       values differ.}
%\label{fig:lmagcc}
%\end{figure}
%
%On the other hand, if we use the same approach to %forecast the
%processor efficiency (IPC) of the \gcc time series, %the
%prediction is far less accurate; see Figure~%\ref{fig:lmagcc}.
%Figure~\ref{fig:lmasvdfive} is clearly a better %prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.

%This gets at the utility of the contribution of this %paper: These time series all come from the same system---computers---but they are not equally predictable (at this point by LMA). For a practitioner is it the case that
%  Our
%conjecture is that while they come from similar systems \gcc produces processor load traces on the top of the complexity spectrum whereas \col produces processor loads with complexity in the mid to low region of the complexity spectrum. If this is the case then \gcc might be much better predicted using a method like Random Walk or \naive which can effectively predict in the presence of complexity. This is explored more rigorously in the results Section of this paper (Section \ref{sec:results}. So that we don't have to compare the forecast accuarcy by visual comparing the predictions we calculate a figure of merit to compare the predictions.

\section{Measuring Structural Complexity }\label{sec:meaComplex}

Estimating the entropy of an arbitrary, real-valued time series is a
significant challenge.  Our approach to this problem, which is the topic of
this section, draws upon methods and results from a variety of fields
including time-series analysis, dynamical systems, and stochastic
processes.

For the purposes of this paper, we view the Shannon entropy---in
particular its growth rate with respect to word length (the
\emph{Shannon entropy rate})---as a measure of complexity and
unpredictability in a time series.  Time-series consisting of i.i.d.
random variables, such as white noise, have maximal entropy rates,
whereas highly structured time-series, for example periodic, have very
low (or zero) entropy rates. A time series with a high entropy rate is
almost completely unpredictable; conversely, one with low entropy rate
is often quite predictable. This can be made more rigorous: Pesin's
relation \cite{pesin77} states that in chaotic dynamical systems, the
Kolmogorov-Sinai (KS) entropy is equal to the sum of the positive
Lyapunov exponents, $\lambda_i$.  The Lyapunov exponents directly
quantify the rate at which nearby states of the system diverge with
time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta
x(0) \right|$.  The faster the divergence, the larger the entropy.
The KS entropy is defined as the supremum of the Shannon entropy rates
of all partitions~\cite{petersen1989}. The partition that achieves
this supremum is the \emph{generating partition} discussed in
Section~\ref{sec:related}.

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}:
information in the present that is also in the past.  The second part
is the aforementioned Shannon entropy rate.  It seems obvious that the
more redundancy in a signal, the more predictable it should be.  And
the specific \emph{form} of the redundancy should dictate whether a
particular prediction method will work well or poorly on the
corresponding signal.  A linear method cannot capture or make use of
nonlinear redundancy, for instance.  This issue---which has been
explored extensively using clean data and/or systems where the
generating process is known, but not so much with empirical data from
unknown systems---is central to the claims in this paper and the
discussion in the following section.

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems are, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
inappropriate in our context, where the target of study is
experimental time-series data.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a
``natural complexity measure for time series''
\cite{bandt2002per}. The permutation entropy employs a method of
discretizing real-valued time series that follows the intrinsic
behavior of the system under examination.  It has many advantages,
including being robust to observational noise, and its application
does not require any knowledge of the underlying mechanisms.  Rather
than looking at the statistics of sequences of values, as is done when
computing the Shannon entropy, permutation entropy looks at the
statistics of the \emph{orderings} of sequences of values using
ordinal analysis. Ordinal analysis of a time series is the process of
mapping successive time-ordered elements of a time series to their
value-ordered permutation of the same size.  By way of example, if
$(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern},
$\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  The
ordinal pattern of the permutation $(x_1, x_2, x_3) = (9, 7, 1)$ is
$321$.

\begin{mydef}[Permutation Entropy]

  Given a time series $\{x_i\}_{i = 1,\dots,N}$. Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$. For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_i\}_{i = 1,\dots,N}$:
  \begin{align*}
    P(\pi) = \frac{\left|\{i|i \leq N-\ell,\phi(x_{i+1},\dots,x_{i+\ell}) = \pi\}\right|}{N-\ell+1}
  \end{align*}
  where $P(\pi)$ quantifies the probability of an ordinal and
  $|\cdot|$ is set cardinality. The \emph{permutation entropy} of
  order $\ell \ge 2$ is defined as
  \begin{align*}
    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P(\pi) \log_2 P(\pi)
  \end{align*}

\end{mydef}

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.
With this in mind, it is common in the literature to normalize
permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With
this convention, ``low'' PE is close to 0 and ``high'' PE is close to
1. Finally, it should be noted that the permutation entropy has been
shown to be identical to the Kolmolgorov-Sinai entropy for many large
classes of systems \cite{amigo2012permutation}, as long as
observational noise is sufficiently small. As mentioned before, this
is equal to the Shannon entropy rate of a generating partition of the
system. This transitive chain of equalities, from permutation entropy
to Shannon entropy rate via the KS entropy, allows us to approximate
the redundancy---being the dual of the Shannon entropy rate---of a
signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger-scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE of such a signal will be
dominated by the noise about the fixed points, driving it to $\approx
1$---which in some sense hides the fact that the signal is actually
quite structured.  In this situation, the terms of the weighted
permutation entropy are dominated by the switching rather than the
stochastic fluctuations.  The amplitude of the switches plays a key
role in predictability, so weighting the scale of ordinal changes is
important for quantifying predictive structure accurately.  To
accomplish this, WPE takes into account the \emph{weight} of a
permutation:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i + 1}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

The weighted probability of a permutation is defined as:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell) \cdot \delta(\phi(x_{i+1}^\ell), \pi) }{\displaystyle \sum_{i \le N - \ell} w(x_{i+1}^\ell)}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively,
this weighted probability enhances permutations that are involved in
``large'' features and de-emphasizes permutations that are small in
amplitude relative to the features of the time series. Using the
standard form of an entropy, the weighted permutation entropy is:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, making $0
\le \textrm{WPE} \le 1$.  This normalization is used in all the
results that follow.

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration in that choice is that the value be large enough
that forbidden ordinals are discovered, yet small enough that
reasonable statistics over the ordinals are gathered.  If an average
of 100 counts per ordinal is considered to be sufficient, for
instance, then $\ell = \operatornamewithlimits{argmax}_{\hat{\ell}} \{
N \gtrapprox 100 \hat{\ell}! \}$.  In the literature, $3 \le \ell \le
6$ is a standard choice---generally without any formal justification.
In theory, the permutation entropy should reach an asymptote with
increasing $\ell$, but that can require an arbitrarily long time
series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $\ell$ until the
result converges, but data length issues can intrude before that
convergence is reached.  We used this approach, which we believe
strikes a good balance between accurate ordinal statistics and
finite-data effects, to choose $\ell$ values for the experiments in
the following section.

% was a good choice, because with $\ell \geq 7$ many signals had
% entropies plummet due to finite data effects and so with a word length
% of 6 we are capturing as much structure as possible without
% sacraficing the accuracy of ordinal statistics.

%\subsection{Regime Selection:Detecting complexity shift in a time series}\label{sec:wpeRegime}
%
%{\color{red} this is from a previous paper I wrote, if this is used we need to update the plots to use wpe and windowed wpe instead of standard pe, also not sure how much of this is necessary but added as starting point}
%
%Permutation entropy is a measure of the complexity of time series as a whole, but it can also be used to calculate a dynamical shift in a time series as illustrated in \cite{cao2004det}. To detect dynamical changes in a time series we follow the methods of \cite{cao2004det}. Partition the time series  $\{x_t\}_{t = 1,\dots,T}$ into overlapping\footnote{Following \cite{cao2004det} we use maximal overlapping block. (A window shift of 1.)} (or nonoverlapping) blocks. Calculate $H(n)$ for each block, and treat $H(n)$ as a function of time.  According to \cite{cao2004det} drastic changes in permutation entropy can indicate change in the underlying model.
%
%In addition to introducing this block method, \cite{cao2004det} introduce the use of a \emph{lag} when calculating permutation entropy. This is identical to the lag used in delay coordinate embedding and often the same language is used. Cao et al. describe the process of symbolization in \cite{bandt2002per} as embedding the time series in an $m$ dimensional symbol space. When we symbolize a scalar time series,$\{x_t\}_{t = 1,\dots,T}$, we first embed the scalar time series into $m$ dimensional vectors before matching this to a symbol in $\mathcal{S}_m$. So pre-symbolization our time series changes from scalar to objects of the form:  $$[x(i),\,x(i+1),\,\dots,x(i+(m-1)]$$ This vector is then mapped by $\phi$ to a symbol in $\mathcal{S}_m$. Cao et al. argue that we can view this as embedding the time series in an $m$ dimensional space with lag 1. Cao et al. then state ``Since in practice the optimal $L$[lag] may be different from 1, we shall present the idea for any $m$[embedding dimension] and $L$."\footnote{No further explanation is given about why this is the case.} To not confuse notation we define $\tau$ to be the lag parameter. The process of symbolization then starts with transforming the scalar time series into a set of vectors of the form $$[x(i),\,x(i+\tau),\,\dots,x(i+(m-1)\tau]$$ where $\tau\ge1$.
%
%\subsubsection{Example: Concatenated Time Series}\label{sec:concatenate}
%As an example of this detection, consider Figure \ref{fig:ts-banded} (a). This is a plot of a concatenated time series comprised of several orbits of the Logistic map with varying parameters as well as intermittent noise. Each concatenated band (either an orbit of the Logistic map or noise) is 50,000 points. The bands of noise are generated with Matlab's \verb|randsample|, using the chaotic bands as the sampling population. Each Logistic map trajectory begins at $x=0.6$. The period-two trajectories have an $R$ of 3.2, the period-three trajectory has an R of 3.838, and the chaotic trajectories have an $R$ of 3.65.
%
%The concatenated time series is comprised of two period-two trajectories($t=0-50,000$ and $t=250,000-300,000$), one period-three trajectory($t=100,000-150,000$), two chaotic trajectories ($t=50,000-100,000$ and $t=200,000-250,000$, and two different bands of noise($t=150,000-200,000$ and $t=300,000-350,000$). Visually the bands of noise and the chaotic trajectory are indistinguishable. If this were an observed time series from a real process, it would be impossible to visually identify which bands were generated deterministically and which were simply noise. If we calculate block permutation entropy as discussed in \cite{cao2004det} and plot this as a function of time it becomes very clear which bands have structure and which do not.
%
%In Figure \ref{fig:ts-banded} (b) we plot the concatenated time series, as well as permutation entropy as a function of time(red, blue and green curves). The red, blue and green curves are the permutation entropy calculated using $m=5$ and $\tau = 1,2,3$ (respectively). Using these curves it is clear that bands 2 and 5 ($t=50,000-100,000$ and $t=200,000-250,000$) are chaotic because they have structure\footnote{Permutation entropy less than one.}. On the other hand, bands 4 and 7 ($t=150,000-200,000$ and $t=300,000-350,000$) have no structure\footnote{Permutation entropy of one.} and are the noise bands. This is in fact the case when comparing to the way the time series was constructed.
%
%The main advantage we can see to the use of lags other than 1 is for analysis of periodicity. This is also the major advantage listed in \cite{cao2004det}. Notice in the period 2 bands the blue curve (permutation entropy calculated with a lag of 2) is zero. This means that with a lag of 2 the period 2 orbit is perfectly structured\footnote{Equivalently, not enough permutations are being realized to impact the calculation of $H(n)$}. If we increase the lag to 3 (the green curve) the two period-two bands have a little bit of structure(as some permutations will be realized) but the period 3 band has zero permutation entropy.
%
%%TSLogistic = [period2Logistic;chaosLogistic;period3Logistic;randomSampleFromMap2;chaosLogistic;period2Logistic;randomSampleFromMap3];
%
%%\begin{figure}
%%  \centering
%%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-alone}}
%%    {(a)}
%%
%%  \subfigure{\includegraphics[width=\columnwidth]{unused-figs/ts-banded-pe}}
%%  {(b)}
%%  \caption{ (a) The concatenated time series discussed in Section~\ref{sec:concatenate}. (b) Plot of the concatenated time series from Section~\ref{sec:concatenate}, as well as permutation entropy as a function of time(blue,green and red curves). Each permutation entropy curve shown is calculated using $m=5$, however, the red,blue and green curve use a $\tau = 1,2,3$ respectively.}\label{fig:ts-banded}
%%\end{figure}
%
%
%
%
%
%
%and what is it good for with forward pointer to the other sections
%
%
%
%\subsubsection{Example: The transient Logistic map}
%In Section~\ref{sec:concatenate} we discussed a time series which stayed in a parameter regime for long amounts of time and then abruptly changed. As we showed, permutation was a great proxy to detect this change. Another type of parameter change that may occur in a physical system is constant parameter drift, i.e., some parameter slowly changes over time. To explore this type of parameter drift we construct the same time series as was used in \cite{cao2004det}. Iterate the logistic map, starting with $R=2.8$, then at every iteration increase $R$ by $10^{-5}$ until $R=4$. The time series, which can be seen in Figure~\ref{fig:translogistic} appears very similar to the standard bifurcation diagram for the Logistic map. The curves in red and blue are the permutation entropy calculated with $m=5$ and $\tau = 1,2$ (respectively).
%
% As we can see, as bifurcations occur in the dynamics the permutation entropy changes as well. An interesting feature of block calculations are the spikes that occur at bifurcation points. As the block begins to encompass multiple regimes, e.g., fixed point and  period two, permutations are being realized for both regimes. This overlap causes an increase in the time series complexity, and thus a spike in the permutation entropy. The dips you see in the second half correspond to bands of periodicity which occur between the chaos. It is worth noting, that the blue curve (permutation entropy with lag 2) does not go to zero in the period 2 parameter range. This is because this is not a true period 2 orbit but transient that is moving across \emph{several} period 2 orbits, thus it maintains a higher level of complexity, albeit still very low.
%
% This illustrates that permutation entropy can not only distinguish between instant (and drastic) parameter shifts but also very slow and gradual parameter drift.
%
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\columnwidth]{unused-figs/tstranslog-perm}
%  \caption{Transient Logistic map time series and permutation entropy curves. Both permutation entropy curves are calculated with $m=5$, the red and blue correspond to $\tau = 1,2$ respectively.}
%  \label{fig:translogistic}
%\end{figure}
%%\begin{enumerate}
%%\item do with henon bifurcation. what happens near the broken chaotic window
%
%%\end{enumerate}
%
%
%The weighted permutation entropy for the \svd program is given in
%Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
%the time series. Within each of those windows, the statistics over words of
%length 4 are computed and the WPE is calculated. The gray bands denote regions
%where the 5,000 value window overlapped visually-distinct regimes. It can be
%seen that the behaviors of the weighted permutation entropy vary between
%regimes. [I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist. Maybe point at some of the predicting predictability papers.]
%
%%\begin{figure}[htbp]
%%  \centering
%%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
%%  \caption{[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]The weighted permutation entropy of %one run of SVD. The gray bands
%%    are regions where the window overlaps regimes. The window size used is
%%    $5,000 \times 100,000$ instructions and the word length is $4$.}
%%  \label{fig:wwpe}
%%\end{figure}
%
%
%%%FIGURE INTENT
%%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen.
%%%%%%%%%%%%%%%%%%
%\begin{figure}[htbp]
%  \centering
%  %\begin{subfigure}{0.3\textwidth}
%  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
%  %  \caption{The instructions per cycle of \svd. Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}. From left to right each change in color represents a change in regime for 6 regimes in total. }
% %   \label{fig:svd_ts}
% % \end{subfigure}%
%  %\\
% % \begin{subfigure}{\textwidth}
%
% {\color{red}TODO: Put this image back in.}
%%Liz commented this image out temporarily so that her mac doesn't hang
%%when she scrolls past p12
%% We decided to skip the figure completely, since we took out
%% all stuff about PE as regime-shift detection.
%% [should add that as future work, though.]
%  \includegraphics[width=\columnwidth]{figs/SVD_wwpe}
%  \caption{The weighted permutation entropy of one run of SVD. The gray bands are regions where the window overlaps regimes. The window size used is 5,000 $\times$ 100,000 instructions and the word length is $6$. For reference the instructions per cycle of \svd are plotted as a ghost behind this plot. Each color on the ghosted time series corresonds to the different regimes as determined by visual inspection. From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
%\end{figure}
%

 \section{Predictability, Complexity, and Permutation Entropy %{\color{blue}  EDITABLE}
 } % (fold)
 \label{sec:results}
% I changed the title of this section because there are lots of
% results in the modeling section too.  The results HERE are about the
% relationship between predictability and WPE

%OUTLINE:
%\begin{enumerate}
%\item \cmark~Introduce MASE.
%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
%its WPE.
%\item \cmark~Talk about structure analysis. Just because %there is forward information
%transfer, does not mean that linear predictors can get at %this.
%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
%side plot.
%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
%\begin{enumerate}
%\item \cmark~clusters are distributed differently
%\item \cmark~clusters are shaped differently---tight or not
%\item \cmark~clusters move differently between LMA and ARIMA
%\item finally, the diagonal line is important. If you're %below it, you could do
%better.
%\end{enumerate}
%\end{enumerate}

%BRAINSTORMING
%\begin{itemize}
%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.
%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE

%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction

%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working.

%\end{itemize}

In this section, we offer an empirical validation of the two findings
introduced in Section \ref{sec:intro}, namely:

\begin{enumerate}
% \item The complexity of a noisy real-valued time series can be
%   effectively quantifiable by computing its permutation entropy.

\item The weighted permutation entropy (WPE) of a noisy real-valued
  time series from an unknown system is correlated with prediction
  accuracy---i.e., the predictable structure in an empirical
  time-series data set can be quantified by its WPE.

\item The relationship between WPE and mean absolute scaled error
  (MASE) is a useful empirical heuristic for identifying mismatches
  between prediction models and time-series data---i.e., when there is
  structure in the data that the model is unable to exploit.

\end{enumerate}

%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)
%%%%%%%%%%%%%%%%%%%%%%%%%

%\item First paragraph: WPE is a good measure of predictability.  Figure:
%best athlete MASE vs. its WPE.

%\item Quantifying structured and unstructured complexity is
%nontrivial in the case of real-valued noisy time series but WPE does
%this. talk about again here but justify in information theory
%section]]

The experiments below involve four different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different segments of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective
of these experiments was to explore how prediction accuracy is related
to WPE.
%
% , and how that relationship depends on the generating process
% and the prediction method.
%
Working from the first 90\% of each signal, we generated a prediction
of the last 10\% using the random-walk, \naive, \arima, and LMA
prediction methods, as described in Section~\ref{sec:accuracy}, then
calculated the MASE value of those predictions.  We also calculated
the WPE of each time series using a wordlength chosen via the
procedure described at the end of Section~\ref{sec:meaComplex}.  In
order to assess the run-to-run variability of these results, we
repeated all of these calculations on 15 separate trials: i.e., 15
different runs of each program.

% Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these
% predictions for each trial of each system: that is, the lowest error
% over all three methods for each of the eight programs.  The WPE is
% plotted against the corresponding MASE value here in order to bring
% out the correlation between these two quantities.

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the \emph{best} prediction for each of
the 120 time series in this study.
\begin{figure}
  \centering
  \includegraphics[width=1.1\columnwidth]{new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.  The solid
    curve is a least-squares log fit of these points.
%
% of all the points except those from {\tt dgesdd$_1$}, which we have
% excluded for reasons explained in the text.
%
The dashed curves reflect the standard deviation of the model in its
parameter space.  Points that lie below and to the right of the shaded
region indicate that the time series has more predictive structure
than the forecast strategy is able to utilize.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
%
% As the plot shows, the best-case prediction error for each trace from
% these eight systems is roughly proportional to the log of the weighted
% permutation entropy, which is consistent with our first conjecture.
%
There is an obvious upward trend, which is consistent with the notion
that there is a pattern in the WPE-MASE relationship.  However, a
simple linear fit is a bad idea here.  First, any signal with zero
entropy should be perfectly predictable (i.e., MASE $\approx 0$), so
any curve fitted to these data should pass through the origin.
Moreover, WPE does not grow without bound, so one would expect the
patterns in the WPE-MASE pairs to reach some sort of asymptote.  For
these reasons, we chose to fit a function of the form $y = a \log(b x
+ 1)$ to these points\footnote{The specific values of the coefficients
  are $a=7.97 \times 10^{-2}$ and $b=1.52 \times 10^3$.}, with $y =$
WPE and $x=$ MASE.  The solid curve in the figure shows this fit; the
dashed curves show the standard deviation of this model in its
parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
deviation on each of the two parameters.  Points that fall within this
deviation volume (light grey) correspond to predictions that are
comparable to the best ones found in this study; points that fall
\emph{above} that volume (dark grey) are better still.  We chose to
truncate the shaded region because of a subtle point regarding the
MASE of an ideal predictor, which should not be larger than 1 unless
the training and test signals are different.  This is discussed at
more length below.

The curves and regions in Figure~\ref{fig:wpe_vs_mase_best} are a
graphical representation of the first finding.  This representation
is, we believe, a useful heuristic for determining whether a given
prediction method is well matched to a particular time series.  It is
not, of course, a formal result.  The forecast methods and data sets
used here were chosen to span the space of standard prediction
strategies and the range of dynamical behaviors, but they do not cover
those spaces exhaustively.  Our goal here is an \emph{empirical}
assessment of the relationship between predictability and complexity,
not formal results about a ``best'' predictor for a given time series.
There may be other methods that produce lower MASE values than those
in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the one-$\sigma$ region about the dashed curve in this
plot strongly suggests a pattern of correlation between the underlying
predictability of a time series and its WPE.  The rest of this section
describes these results and claims in more detail---including the
measures taken to assure meaningful comparisons across methods,
trials, and programs---and elaborates on the meaning of the different
curves and limits in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure*}
  \centering
%    \includegraphics[width=\columnwidth]{ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{LMA_prediction_vs_entropy}
  \includegraphics[width=1.7\columnwidth]{new_predictions_vs_entropy4a}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top-right plot for scale reasons, as described in the text.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The curves and shaded
regions are the same as in the previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure*}
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding trace in any of the four plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---\arima predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters horizontally.
The mean MASE scores of predictions generated with the nonlinear LMA
method are generally closer to the dashed curve; the \arima method
clusters are more widely spread, the \naive ~clusters even more so.  A
few of the clusters have very high variance; these are discussed later
in this section.

The main thing to note here, however, is not the details of the shapes
of the clusters, but rather their positions in the four plots:
specifically, the fact that many of them are to the right of and/or
below the dashed curve that identifies the boundary of the shaded
region.  \emph{These predictions are not as good as our heuristic
  suggests they could be.}  Focusing in on any single signal makes
this clear: LMA works best for \svdsix, for instance, followed by the
random-walk prediction method, then \arima and \naive.  This provides
some practical leverage: if one calculates an WPE vs. MASE value that
is outside the shaded region, that should suggest that the prediction
method is not well matched to the task at hand: that is, the time
series has more predictive structure than the method is able to use.
%
% The idea behind the shaded region is to help practitioners identify
% when there is predictive structure in the data that a given method is
% not exploiting (viz., \arima on \svdsix)---i.e., that one could do
% better with another forecast strategy.
%
In the case of \arima on \svdsix, for instance, one should try a
different method.  The position of the LMA cluster for \svdsix, on the
other hand, reflects this method's ability to capture and exploit the
structure that is present in this signal.  WPE vs. MASE values like
this, which fall in the shaded region, should suggest to the
practitioner that the prediction method is well-suited to the task.
The following discussion lays out the details that underlie these
claims.
%
% The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
% and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
% argument at the end of the previous paragraph.
%

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from
Figure~\ref{fig:forecast-example} and Table~\ref{tab:error} that the
\naive, \arima, and (especially) random-walk prediction methods do not
perform very well on this signal.  The MASE scores of these
predictions are $0.571 \pm 0.002$, $1.001 \pm 0.002$, and $0.599 \pm
0.211$, respectively, across all 15 trials.  That is, \naive ~and
\arima perform only $\approx 1.7$ times better than the random-walk
method, a primitive strategy that simply uses the current value as the
prediction.  However, the WPE value for the \col trials is $0.513 \pm
0.003$, a signal that is in the center of the complexity spectrum
described in Section~\ref{sec:intro}.

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of three of the four
images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
are far to the right of and/or below the dashed curve.  Again, this
indicates that these methods are not leveraging the available
information in the signal.  The dynamics of \col may be complicated,
but they are not unstructured.  This signal is nonlinear and
deterministic~\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or the previous value
(random walk), or one that uses a linear model (\arima)---the MASE
score is much improved: $0.050 \pm 0.001$.  This prediction is 20
times more accurate than a random-walk forecast, which is more in line
with the level of predictive structure that the low WPE value suggests
is present in the \col signal.  The MASE scores of random-walk
predictions of this signal are all $\approx 1$---as one would
expect---pushing those points well below the shaded region.  Clearly
the stationarity assumption on which this method is based does not
hold for this signal.

The \col example brings out some of the shortcomings of automated
model-building processes.  Note that the {\color{blue}$+$} points are
clustered very tightly in the lower left quadrant of the \naive,
random-walk, and LMA plots in Figure~\ref{fig:wpe_vs_mase_all}, but
spread out horizontally in the \arima plot.  This is because of the
way the \arima process builds ARIMA models \cite{autoARIMA}.  If a
KPSS test of the time series in question indicates that it is
nonstationary, the \arima recipe adds an integration term to the
model.  This test gives mixed results in the case of the \col process,
flagging five of the 15 trials as stationary and ten as nonstationary.
We conjectured that ARIMA models without an integration term perform
more poorly on these five signals, which increases the error and
thereby spreads out the points.  We tested this hypothesis by forcing
the inclusion of an integration term in the five cases where a KPSS
test indicated that such a term was not needed.  This action removed
the spread, pushing all 15 of the \col ~ \arima points in
Figure~\ref{fig:wpe_vs_mase_all} into a tight cluster.
%\alert{Josh to do: this paragraph needs more, in response to the %ref's
%  gritch about fitting: the stuff about \arima being an automated
%  recipe that can give bad results, and the fact that our %technique
%  can tell when it does.  Look at the response letter for context %and
%  ammunition---and also put something in there that calls his
%  attention to this paragraph.}

The discussion in the previous paragraph highlights the second finding
of this paper: the ability of our proposed heuristic to flag
inappropriate models.  \arima is an automated, mechanical procedure
for choosing parameters for an ARIMA model of a given data set.  While
the tests and criteria employed by this algorithm
(Section~\ref{sec:arima}) are sophisticated, the results can still be
sub-optimal---if the initial space of models being searched is not
broad enough, for instance, or if one of the preliminary tests gives
an erroneous result.  \arima \emph{always} returns a model, and it can
be very hard to detect when that model is bad.  Our results suggest a
way to do so: if the MASE score of an auto-fitted model like \arima is
out of line with the WPE value of the data, that can be an indication
of inappropriateness in the order selection and parameter estimation
procedure.

% A close examination of the \arima MASE scores and WPE values of the 15
% \col traces, for instance, brought out such a situation: all of these
% signals had the same complexity, but some (the outliers in the blue
% cluster of {\color{blue}$+$} points in the bottom-left panel of
% Figure~\ref{fig:wpe_vs_mase_all}) had much higher MASE values,
% indicating a failure in the \arima process.  We then went back to the
% data, noticed some shifts of the baseline in those signals, and
% manually assisted the automated fitting procedure to force the
% inclusion of a term that is appropriate for a nonstationary signal.
% In all cases, this produced ARIMA models with MASE scores in line with
% the others in that cluster.

% The discussion in the previous paragraph brings to light an
% interesting feature of our proposed heuristic. The \arima algorithm
% that was used to build all of the ARIMA models in this paper follows
% the procedure outlined in Section V B to choose values for the model
% parameters.  As discussed previously, this algorithm employs
% sophisticated tests, such as the KPSS test, to intelligently traverse
% and reduce model space and then uses well-known criterion for choosing
% the most-likely of the remaining models. However, it is quite possible
% that this procedure can choose a sub-optimal fit. Specifically, if the
% initial space of models being searched is not broad enough, or if one
% of the preliminary tests gives an erroneous result, then the best
% ARIMA model may be excluded or eliminated before likelihood is even
% examined. This circumstance is hard to detect. Maximal-likelihood
% tests, such as AIC used in \arima, can only conclude which of the
% models provided is ``most likely"---but cannot conclude that
% \emph{none} of the models are likely. In contrast,
% %Sometimes the choices made by that procedure produce a sub-optimal predictor, however there is nothing to say whether a different model from the ARIMA class would work better.
% inappropriate MASE scores (compared to what the WPE value might
% suggest) may suggest an inappropriateness in the order selection and
% parameter estimation procedure, which would flag to a practitioner to
% reevaluate the appropriateness of fit. For example,
% %During the fitting process of \arima many subclasses of ARIMA models are eliminated based on various tests, such as the KPSS test. If these tests give erroneous results then entire subclasses of ARIMA models are never even considered. Our heuristic may be able to flag when this has occurred.
% by examining the \arima MASE scores vs. WPE values of \col, we were
% able to see that, while all signal complexities were the same, some of
% the signals had much higher MASE values. This suggested to us
% inappropriate parameter selection for these time series. We then went
% back and manually assisted the automated fitting procedure, resulting
% in models with consistent MASE vs WPE score, and improved forecasts.

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be used to predict the future course of
the time series.
% As mentioned above, though, the effectiveness of any prediction
% strategy depends on how well it leverages the information that is
% available in the data: methods that use linear models, for instance,
% may fail when applied to nonlinear processes.  (This is the basis
% for the second conjecture above.)
The MASE scores of the \naive ~and \arima predictions for this system
are $20.870 \pm 0.192$ and $2.370 \pm 0.051$, respectively: that is,
20.87 and 2.37 times worse than a simple random-walk forecast of the
training set portions of the same signals\footnote{The \naive ~MASE
  score is large because of the bimodal nature of the distribution of
  the values of the signal, which makes guessing the mean a
  particularly bad strategy.  The same thing is true of the \svdthree
  signal.}.  As before, the positions of these points on a WPE
vs. MASE plot---significantly below and to the right of the shaded
region---should suggest to a practitioner that
%  there is more structure in this signal than the corresponding
% method is able to leverage, and that
a different method might do better.  Indeed, for \svdfive, the LMA
method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster of
results that largely within the shaded region on the WPE-MASE plot.
This is consistent with our second finding: the LMA method can capture
and reproduce the way in which the \svdfive system processes
information, but the \naive ~and \arima prediction methods cannot.

The WPE of \gcc is higher still: $0.943 \pm 0.001$.  This system
transmits very little information forward in time and provides almost
no structure for prediction methods to work with.  Here, the
random-walk predictor is the best of the methods used here.  This
makes sense; in a fully complex signal, where there is no predictive
structure to utilize, methods that depend on exploiting that
structure---like ARIMA and LMA---cannot get any traction on those
signals.
%
% in a fully complex signal, the MASE of a random-walk forecast will
% always be $\approx 1$, assuming that the training signal accurately
% reflects the overall behavior of the out-of-sample time series
%
Since fitting a hyperplane using least squares should filter out some
of the noise in the signal, the fact that LMA outperforms \arima
($1.530 \pm 0.021$ vs. $1.837 \pm 0.016$) may be somewhat
counterintuitive.  However, the small amount of predictive structure
that is present in this signal is nonlinear (cf.,~\cite{mytkowicz09}),
and LMA is designed to capture and exploit that kind of structure.
Note that all four \gcc clusters in Figure~\ref{fig:wpe_vs_mase_all}
are outside the shaded region; in the case of the random-walk
prediction, for instance, the MASE value is $1.1381 \pm 0.011$.  This
is due to nonstationarity in the signal: in particular, differences
between the training and test signals.  The same effect is at work in
the \svdtwo results, for the same reasons---and visibly so, in the red
segment of Figure~\ref{fig:svd-ts-colored}, where the period and
amplitude of the oscillations are decreasing.

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), three of the four prediction
methods do quite well on this signal, yielding mean MASE scores of
0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
well above the trend followed by the other seven signals.  The reasons
for this are discussed in the following paragraph.  The MASE scores of
the predictions that were produced by the \naive ~method for this
system, however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top-right plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6, which is better than the other three methods.  In five of the 15
\svdone trials, however, there were step changes in the signal.  This
is a different nonstationarity than in the case of \col---large jump
discontinuities rather than small shifts in the baseline---and not one
that we were able to handle by simply forcing the ARIMA model to
include a particular term.  The \naive ~method has a very difficult
time with signals like this, particularly if there are multiple step
changes.  This raised the MASE scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near MASE $\approx 2.5$, as well as two points that
  are beyond the domain of the graph, at MASE $\approx 11.2-14.8$.},
and in turn raising both the mean and variance of this set of trials.

The effects described in the previous paragraph are also exacerbated
by the way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}.  There are two issues here.
First, random-walk prediction works very badly on signals with
frequent, large, rapid transitions.  Consider a signal that oscillates
from one end of its range to the other at every step.  A signal like
this will have a low WPE, much like {\tt col\_major}.  However, a
random-walk forecast of this signal will be 180 degrees out of phase
with the true continuation.  Since random-walk error appears in the
denominator of the MASE score, this effect can shift points leftwards
on a WPE vs. MASE plot, and that is exactly why the \svdone clusters
in Figure~\ref{fig:wpe_vs_mase_all} are above the dashed curve.  This
time series, part of which is shown in closeup in
Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure}
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in Figure~\ref{fig:svdone-ts}),
punctuated by short excursions above 1.9.  In the former regime, which
makes up more than 80\% of the signal, there are frequent dips to 1.82
and occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.

% For all of these reasons, we view \svdone as an outlier and exclude it
% from the fit calculation of the dashed curve in
% Figure~\ref{fig:wpe_vs_mase_best}.

The second effect that can skew MASE scores is nonstationarity.  Since
this metric is normalized by the error of a random-walk forecast
\emph{on the training signal}, differences between the test signal and
training signal can create issues.  This is why the MASE values in
Table~\ref{tab:error} are not identically one for every random-walk
forecast of every time series: the last 10\% of these signals is
significantly different from the first 90\%.  The deviation from 1.00
will depend on the process---whether it has multiple regimes, what
those regimes look like, and how it switches between them---as well as
the experimental setup (e.g., sensor precision and data length).  For
the processes studied here, these effects do not cause the MASE values
to exceed 1.15, but pathological situations (e.g., a huge switch in
scale right at the training/test signal boundary, or a signal that
simply grows exponentially) could produce higher values.  This
suggests another potentially useful heuristic: if the MASE of a
random-walk prediction of a time series is significantly different
from 1, this could be an indication that the signal is nonstationary.
We are in the process of exploring this idea and its relationship to
the DVS graphs mentioned in the fourth paragraph of
Section~\ref{sec:related}.

The curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from finite sets of
methods and data.  We put a lot of thought and effort into making
these sets representative and comprehensive.  The forecast methods
involved range from the simple to the sophisticated; the time-series
data analyzed in this section was sampled from a system whose behavior
spans the dynamical behavior space.  While we are cautiously
optimistic about the generality of our conclusions, more exploration
will be required before we can make any such claim.  Our preliminary
work along those lines shows that data from the H\'enon map
\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data
set\footnote{This dataset can be retrieved from
  \url{http://www-psych.stanford.edu/~andreas/Time-Series/SantaFe.html}.}
\cite{weigend-book}, and a random-walk process all fall within the
one-$\sigma$ volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo, \svdfive and \svdsix.

Of course, the geometry of the curves and bounds in these figures do
not necessarily extend to other error metrics.  The positions of the
WPE vs. error points---and the reasons to choose a particular function
to fit to them or impute theoretical bounds on the meaning of the
results---may change if one calculates error using a procedure other
than the one described in Section~\ref{sec:accuracy}.  The MASE error
metric, as discussed above, has its weaknesses.  Even so, we concur
with \cite{MASE} that it is an effective way to compare prediction
error across time-series data of different lengths and different
scales, and we are convinced by the extensive evaluations that are
offered in that paper, as well as the comparisons of MASE to other
metrics, that its strengths far outweigh its weaknesses.

%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
%\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.
%%%%%%%%%%%%%%%%%%%%%%%%%

%In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly compare the performance of the LMA and ARIMA prediction methods (respectively) to the value of the weighted permutation entropy for all runs of each program under consideration. The LMA MASE values are largely similar to those of the best predictions, primarily because LMA often performed superior to ARIMA (and the na\"ive method). On the other hand, the ARIMA MASE values are largely uncorrelated with WPE values. As was hinted at while discussing the previous results, the fact that ARIMA is uncorrelated with WPE brings about an interesting perspective on information transfer and WPE. The WPE is sensitive to both linear and nonlinear structure. When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure that LMA can handle but ARIMA cannot. So while ARIMA is consistently out performed by random walk,  there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.

%is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MAS>1$). How do we determine if the prediction strategy is not adequate for the prediction task, or if the signal is simply too complex to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.

\section{ Conclusions \& Future Work }\label{sec:conc}

% To address the shortcomings mentioned in the previous section, we are
% in the process of repeating our own study with other metrics, other
% forecast methods, and time-series data from other systems.

Forecast strategies that are designed to capture predictive structure
are ineffective when signal complexity outweighs information
redundancy.  This poses a number of serious challenges in practice.
Without knowing anything about the generating process, it is difficult
to determine how much predictive structure is present in a noisy,
real-world time series.  And even if predictive structure exists, a
given forecast method may not work, simply because it cannot exploit
the structure that is present (e.g., a linear model of a nonlinear
process).  If a forecast model is not producing good results, a
practitioner needs to know why: is the reason that the data contain no
predictive structure---i.e., that no model will work---or is the model
that s/he is using simply not good enough?

In this paper, we have argued that redundancy is a useful proxy for
the inherent predictability of an empirical time series.  To
operationalize that relationship, we use an approximation of the
Kolmogorov-Sinai entropy, estimated using a weighted version of the
permutation entropy of~\cite{bandt2002per}.  This WPE technique---an
ordinal calculation of forward information transfer in a time
series---is ideal for our purposes because it works with real-valued
data and is known to converge to the true entropy value. Using a
variety of forecast models and more than 150 time-series data sets
from experiments and simulations, we have shown that prediction
accuracy is indeed correlated with weighted permutation entropy: the
higher the WPE, in general, the higher the prediction error.  The
relationship is roughly logarithmic, which makes theoretical sense,
given the nature of WPE, predictability, and MASE.

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, \arima, and the \naive
~method on the \gcc signal, which is extremely complex---i.e.,
extremely low redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.

The curves and shaded regions in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} generalize and operationalize the
discussion in the previous paragraph.  These geometric features are a
preliminary, but potentially useful, heuristic for knowing when a
model is not well-matched to the task at hand: a point that is below
and/or to the right of the shaded regions on a plot like
Figure~\ref{fig:wpe_vs_mase_all} indicates that the time series has
more predictive structure than the forecast model can capture and
exploit---and that one would be well advised to try another method.

These curves were determined empirically using a specific error metric
and a finite set of forecast methods and time-series traces.  If one
uses a different error metric, the geometry of the heuristic will be
different---and may not even make sense, if one uses a metric that
does not support comparison across different time series.  And while
the methods and traces used in this study were chosen to be
representative of the practice, they are of course not completely
comprehensive.  It is certainly possible, for instance, that the
nonlinear dynamics of computer performance is subtly different from
the nonlinear dynamics of other systems.  Our preliminary results on
other systems (H\'enon, Lorenz, a random-walk process, SFI ``A'',
nonlinear transformations of the computer performance data) lead us to believe
that our results will generalize beyond the examples described in this
paper.  We are in the process of following up on that exploration with
a broader study of data, forecast methods, and error metrics.

% {\color{blue}agreed, I don't think it addds anything}\alert{I'm not
%   sure this paragraph adds anything.  Can/should we remove it?  Given
%   that information is a fundamental limitation in predictability, then
%   gathering and using more information is an obvious next step.  But
%   there is an equally obvious tension here between data length and
%   prediction speed: a forecast that requires half a second to compute
%   is not useful for the purposes of real-time control of a computer
%   system with a MHz clock rate, for instance.  Another alternative is
%   to sample several system variables simultaneously and build
%   multivariate models.  This is a particular challenge in nonlinear
%   LMA-type models, since multivariate delay-coordinate embedding
%   (e.g., \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
%   computationally prohibitive.  We are working on alternative methods
%   that sidestep that complexity.}
%

%Paragraphs on the issues that come up, including one about the "amount
%of info" one: if one could sample more variables, for instance, one
%might be able to do a better job of predicting more-complex traces.
%Segue to some handwaving about multivariable LMA models; tie this back
%to the "computers are NLD systems" stuff in the intro.  This is a real
%challenge; current approaches to this modelling problem have the major
%issue of taking way too long to build.  And that's a big issue if
%you're trying not just to classify, but to predict.  In a system that
%runs at MHz speeds, a prediction that takes milliseconds to compute is
%not useful.
%\end{it}

Nonstationarity is a serious challenge in any time-series modeling
problem.  Any regime shift that causes a change in the predictive
structure between the training signal and the test signal, for
instance, may skew the MASE score and thereby affecting the utility of
our heuristic.  Conversely, though, a MASE score of a random-walk
prediction that is significantly different from 1.0---as mentioned at
end of the previous section---could potentially be a good indicator of
nonstationarity.

Detecting regime shifts---and adapting prediction models
accordingly---is an important area of future work.  Indeed, one of the
first applications of permutation entropy was to recognize the regime
shift in brainwave data that occurs when someone has a
seizure~\cite{cao2004det}.  Recall that the signal in
Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
this paper because it contained a number of different regimes.  We
segmented this signal visually, but one could imagine using some
combination of WPE and MASE to do so instead (e.g., in a sliding
window across the time series).  Automating regime-shift detection
would be an important step towards a fully adaptive modeling strategy,
where old models are discarded and new ones are rebuilt whenever the
time series enters a new regime.  Our WPE vs. MASE results could be
particularly powerful in this scenario, as their values could not only
help with regime-shift detection, but also suggest what kind of model
might work well in each new regime.  Of particular interest would be
the class of so-called \emph{hybrid systems}~\cite{hybrid}, which
exhibit discrete transitions between different continuous
regimes---e.g., a lathe that has an intermittent instability or
traffic at an internet router, whose characteristic normal traffic
patterns shift radically during an attack.  Effective modeling and
prediction of these kinds of systems is quite difficult; doing so
adaptively and automatically---in the manner that is alluded to at the
end of the previous paragraph---would be an interesting challenge.

\section*{Acknowledgment}
This work was partially supported by NSF grant \#CMMI-1245947 and ARO
grant \#W911NF-12-1-0288.  We would also like to acknowledge the Santa
Fe Institute whose interdisciplinary halls created a place for this
work to be born, and the anonymous reviewers whose comments
strengthened the presentation.

%\bibliographystyle{aipauth4-1}
\bibliographystyle{apsrev4-1}
\bibliography{bibliofile}

\end{document}

### 2014-10-14 FinalSubmissiontoPRE/pecomplexity.tex
adding Color Online

@@
     % and a .pdf suffix will be assumed for pdflatex
-    \caption{A short segment of a computer performance trace: the
+    \caption{(Color online) A short segment of a computer performance trace: the
       instructions per CPU clock cycle (IPC) during the execution of
@@
     \includegraphics[width=\columnwidth]{gccfullts}
-    \caption{An IPC trace during the execution of \gcc.}
+    \caption{(Color online) An IPC trace during the execution of \gcc.}
     \label{fig:gcc-ts}
@@
     \includegraphics[width=\columnwidth]{SVD1RegimesColored}
-    \caption{An IPC trace during the execution of \gcc.  The colors
+    \caption{(Color online) An IPC trace during the execution of \gcc.  The colors
       identify the different \emph{segments} of the signal that are
@@
     }
-    \caption{3D projections of delay-coordinate embeddings of the
+    \caption{(Color online) 3D projections of delay-coordinate embeddings of the
        traces from (a) Figure~\ref{fig:col-ipc} (b)
@@
   \end{tabular}
-  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts
+  \caption{(Color online) Predicted ($p_j$) versus true values ($c_j$) for forecasts
     of \col, \gcc, and \svdfive made using each of the four 
@@
   \includegraphics[width=1.1\columnwidth]{new_prediction_vs_entropy}
-  \caption{Weighted permutation entropy vs. mean absolute scaled error
+  \caption{(Color online) Weighted permutation entropy vs. mean absolute scaled error
     (MASE) of the best prediction of each time series.  The solid
@@
   \includegraphics[width=1.7\columnwidth]{new_predictions_vs_entropy4a}
-\caption{WPE vs. MASE for all trials, methods, and systems---with the
+\caption{(Color online) WPE vs. MASE for all trials, methods, and systems---with the
   exception of \svdone, \svdthree, and \svdfive are omitted from the
@@
     \includegraphics[width=\columnwidth]{svdonets2}
-\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
+\caption{(Color online) A small portion of the \svdone time series}\label{fig:svdone-ts}
 \end{figure}

### 2014-10-15 FinalSubmissiontoPRE/pecomplexity.tex
fixing figure 4
fixing the table figure

@@
     \includegraphics[width=\columnwidth]{SVD1RegimesColored}
-    \caption{(Color online) An IPC trace during the execution of \gcc.  The colors
+    \caption{(Color online) An IPC trace during the execution of \gcc.  The colors (also separated by vertical dashed lines)
       identify the different \emph{segments} of the signal that are
@@
 signal, ordered from left to right.  These segments, which were
-determined visually, are shown in different colors in
+determined visually, are shown in different colors (also separated by vertical dashed lines) in
 Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
@@
  \begin{figure}[!ht]
-    \subfloat[\col\label{fig:colEmbedding}]{%
+    \subfloat[\col instructions per cycle\label{fig:colEmbedding}]{%
       \includegraphics[width=\columnwidth]{colipc3d}
@@
     \hfill
-    \subfloat[\gcc\label{fig:gccEmbedding}]{%
+    \subfloat[\gcc instructions per cycle\label{fig:gccEmbedding}]{%
       \includegraphics[width=\columnwidth]{gcc3dipc}
@@
         \hfill
-    \subfloat[\svdfive \label{fig:svdfiveEmbedding}]{%
+    \subfloat[\svdfive instructions per cycle\label{fig:svdfiveEmbedding}]{%
       \includegraphics[width=\columnwidth]{svd53dipc2}

### 2014-10-15 FinalSubmissiontoPRE/pecomplexity.tex
fixing some wording for greyscale

@@
 
-% Commented this out because the addition of the paragraph from the
-% introduction seemed like it could replace it
-% 
-% \col is a simple C program that repeatedly initializes the upper
-% triangle of a 2048 $\times$ 2048 matrix in column-major order.
-% % 
-% % by looping over the following three lines of code:
-% % \begin{lstlisting}
-% %   for(i=0;i<2048;i++)
-% %   	for(j=i;j<2048;j++)
-% %   		data[j][i] = 0;
-% % \end{lstlisting}
-% % full listing
-% % #define SIZE (2048)
-% % 
-% % int main(int argc, char ** argv)
-% % {
-% %   int i, j, n;
-% %   static int data[SIZE][SIZE];
-% %   
-% %   for(n=0;n<numInits;n++){  
-% %   for(i=0;i<SIZE;i++)
-% %   	for(j=i;j<SIZE;j++)
-% %   		data[j][i] = 0;
-% %   }
-% %   return 0;
-% % }
-% %
-% As mentioned in Section~\ref{sec:intro}, this simple program exhibits
-% surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
-% of the processor performance during the execution of this program is
-% shown in Figure~\ref{fig:col-ts}.
-% \begin{figure}[htbp]
-%   \centering
-%     \includegraphics[width=\columnwidth]{figs/colFullTS}
-%     \caption{The instructions executed per CPU clock cycle (IPC) as
-%       the \col program runs.  Each point is the average IPC in a
-%       100,000 instruction period. Figure~\ref{fig:col-ipc} is a
-%       closeup of the first part of this signal.  }
-%     \label{fig:col-ts}
-%   
-%   \end{figure}
-
 The SPEC CPU2006 benchmark suite \cite{spec} is a collection of
@@
 utilization patterns change quite radically.  For the first
-$\sim$21,000 measurements (21,000 $\times$ 100,000 instructions),
+$\sim$21,000 (in units of $10^5$ instructions) measurements,
 roughly 1.8 instructions are executed per cycle, on the average, by
@@
 through a number of different oscillatory regimes, which we have
-color-coded in the figure in order to make textual cross-references
+color-coded in the figure (as well as separated by vertical dashed lines) in order to make textual cross-references
 easy to track.
@@
 
-\svdone---the dark blue segment of
+\svdone---the dark blue (first) segment of
 Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
